### 2011-03-29 papers/wflp2011/paper.tex
Paper on compiler added

\documentclass{llncs}

\usepackage{times}
\usepackage{url}
\def\UrlFont{\tt}
\usepackage{latexsym}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{color}

\usepackage{listings}
\lstset{aboveskip=1.0ex,
        belowskip=1.0ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        basewidth=0.55em,
        basicstyle=\small\ttfamily}

\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
       }

\lstnewenvironment{curry}{\lstset{backgroundcolor=\color[rgb]{0.9,0.9,0.9}}}{}
\lstnewenvironment{haskell}{}{}
\newcommand{\listline}{\vrule width0pt depth1.75ex}

\newcommand{\code}[1]{\mbox{\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\ol}[1]{\overline{#1}}

% ------------------------------------------------------------------
% SYMBOLS & FORMULAS

\newcommand{\Cc}{{\cal{C}}} % constructors
\newcommand{\Fc}{{\cal{F}}} % defined functions
\newcommand{\Var}{{\cal V}ar} % variables in an object
\newcommand{\Dom}{{\cal D}om} % domain of a substitution
\newcommand{\Rc}{{\cal{R}}} % rewrite system
\newcommand{\Tc}{{\cal{T}}} % algebra of terms
\newcommand{\Xc}{{\cal{X}}} % set of all variables
\renewcommand{\emptyset}{\varnothing}
\newcommand{\suc}{\mathit{success}}  % success
\newcommand{\Suc}{\mathtt{Success}}  % success type
\newcommand{\funset}{\ensuremath{_{\cal S}}}
\newcommand{\seq}{\mathrel{\code{=:=}}} % strict equality of math mode

\begin{document}
\pagestyle{plain}
\sloppy

\title{KiCS2: A New Approach to Compile Curry to Haskell}

\author{
Bernd Bra{\ss}el
\kern1em
Michael Hanus
\kern1em
Bj{\"o}rn Peem{\"o}ller
\kern1em
Fabian Reck
}
\institute{
Institut f\"ur Informatik, CAU Kiel, D-24098 Kiel, Germany \\
\email{\{bbr|mh|bjp|fre\}@informatik.uni-kiel.de}
}

\maketitle

\begin{abstract}
bla bla
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:Introduction}

Functional logic languages integrate the most important
features of functional and logic languages
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they combine higher-oder functions and demand-driven
evaluation from functional programming with logic programming features
like nondeterministic search and computing with partial information
(logic variables).
The combination of these features
has led to new design patterns \cite{AntoyHanus02FLOPS}
and better abstractions for application programming,
e.g., as shown for programming with databases
\cite{BrasselHanusMueller08PADL,Fischer05},
GUI programming \cite{Hanus00PADL},
web programming \cite{Hanus01PADL,Hanus06PPDP,HanusKoschnicke10PADL},
or string parsing \cite{CaballeroLopez99}.

The implementation of functional logic languages is challenging
since a reasonable implementation has to support the operational
features mentioned above.
One possible approach
is the design of new abstract machines appropriately supporting
these operational features and implementing them in some (typically,
imperative) language, like C \cite{Lux99FLOPS}
or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
Another approach is the reuse of already existing implementations
of some of these features by translating
functional logic programs into logic or functional languages.
For instance, if one compiles into Prolog, one can reuse
the existing backtracking implementation for nondeterministic
search and logic variables and unification for computing with partial
information. However, one has to implement demand-driven evaluation
and higher-order functions \cite{AntoyHanus00FROCOS}.
A disadvantage of this approach is the commitment to a fixed
search strategy (backtracking).

If one compiles into a non-strict functional language like Haskell,
one can reuse the implementation
of lazy evaluation and higher-order functions, but one has
to implement nondeterministic evaluations
\cite{BrasselFischerHanusReck11,BrasselHuch07}.
Although Haskell offers list comprehensions for modelling
backtracking \cite{Wadler85}, this cannot be exploited
due to the specific semantical requirements of the combination
of non-strict and nondeterministic operations
\cite{GonzalezEtAl99}. Thus, additional implementation efforts
are necessary like implementation of shared
nondeterministic computations \cite{FischerKiselyovShan09}.

Nevertheless, the translation of functional logic languages
into other high-level languages is attractive:
it limits the implementation efforts compared to an implementation
from scratch and one can exploit the existing implementation technologies
provided that the efforts to implement the missing features
are reasonable.

In this paper we describe an implementation that is based
on the latter principles. We present a method to compile
programs written in the functional logic language Curry \cite{Hanus06Curry}
into Haskell programs
The difficulty of such an implementation is the fact
that nondeterministic results can occur in any place of a
computation. Thus, one cannot separate logic computations
by the use of list comprehensions \cite{Wadler85},
but any outcome of some operation could be potentially nondeterministic,
i.e., it might have more than one result value.
We solve this problem by an explicit representation of
nondeterministic values, i.e., we extend each data type by
another constructor to represent the choice between several values.
This idea is also the basis of Curry implementation KiCS
\cite{BrasselHuch07,BrasselHuch09}.
However, KiCS is based on unsafe features of Haskell
that inhibit the use of optimizations provided by Haskell compilers
like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
In contrast, our implementation is not based on unsafe features.
In addition, we also support more flexible search strategies
and new features to encapsulate nondeterministic computations.

The general objective of our approach is the support
of flexible strategies to explore the search space
resulting from nondeterministic computations.
In contrast to Prolog-based implementations
that use backtracking and, therefore, are incomplete,
we also want to support complete strategies like breadth-first search,
iterative deepening or parallel search (in order to exploit
multi-core architectures). We achieve this goal by an
explicit representation of the search space as data
that can be traversed by various operations.
Moreover, purely deterministic computations
are implemented as purely functional programs so that
they are executed with almost the same efficiency
of their purely functional counterparts.

In the next section, we sketch are source language Curry
and introduce a normalized form of Curry programs that is the
basis of our translation scheme.
Section~\ref{sec:Compilation} presents the basic ideas
of this translation scheme.
Further aspects related to our compiler are discussed in
Section~\ref{sec:FurtherAspects}.
We show benchmarks of our initial implementation in
Section~\ref{sec:Benchmarks}
before we conclude in Section~\ref{sec:Conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Curry Programs}
\label{sec:Curry}

The syntax the functional logic language Curry \cite{Hanus06Curry}
is close to Haskell \cite{PeytonJones03Haskell}, i.e.,
variables and function names usually
start with lowercase letters, the names of type and data constructors
start with an uppercase letter, and the application of $f$
to $e$ is denoted by juxtaposition (``$f~e$'').
In addition, Curry allows free (logic) 
variables in conditions and right-hand sides of defining rules.
In contrast to functional programming and similarly to logic programming,
operations can be defined by overlapping rules so that
they might yield more than one result on the same input.
Such operations are also called \emph{nondeterministic}.
For instance, Curry offers a \emph{choice} operation that is predefined by
the following rules:
%
\begin{curry}
  x ? _ = x
  _ ? y = y
\end{curry}
%
Thus, we can define a nondeterministic operation \code{aBool} by
\label{ex:aBool}
%
\begin{curry}
  aBool = True ? False
\end{curry}
%
so that the expression \ccode{aBool} has two values:
\code{True} and \code{False}.

If nondeterministic operations are used as arguments in other operations,
a semantical ambiguity might occur. Consider the operations
%
\begin{curry}
  xor True  False = True
  xor True  True  = False
  xor False x     = x

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{lstlisting}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool 
                 $\to~$  xor True False      $\to~$  True
\end{lstlisting}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid such dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed the
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and nondeterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics is the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

Due to these considerations, an implementation of Curry
has to support lazy evaluation where operations can have multiple results
and unevaluated arguments must be shared.
Since this is a complex task, if we try to implement it
directly on the source program,
we perform some simplifications before we generate the target code.

First of all, we assume that our programs are \emph{first-order}.
Higher-order programs can be transformed into first-order programs
by a technique called ``defunctionalization'' \cite{Reynolds72}
which is often used in Prolog-based implementations of
functional logic languages \cite{AntoyHanus00FROCOS}.
However, we will later use the higher-order features of Haskell
instead of defunctionalization.

\label{sec:no-logic-vars}
Furthermore, we assume that our programs \emph{do not contain logic variables}.
This assumption can be made since it has been shown
\cite{AntoyHanus06ICLP} that logic variables can be replaced
by nondeterministic ``generators,''
i.e., operations that evaluate to all possible values of the type of
the logic variable. For instance, a Boolean logic variable
can be replaced by the generator \code{aBool} defined above.

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\
    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
                                            & \mbox{let binding}\\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & \ol{D_k}
\end{array}
\]
\caption{Uniform programs}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.
There, %$\ol{o_n}$ denotes the object sequence $o_1,\ldots,o_n$,
the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules where only one constructor occurs in the same argument
of all rules.\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following sections.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Compilation to Haskell: The Basics}
\label{sec:Compilation}

\subsection{Representing Nondeterministic Computations}

As mentioned above, our implementation is based on the explicit
representation of nondeterministic results of computation
as values. This could be easily achieved by adding a constructor
to each data type to represent a choice between two values.
For instance, one can redefine the data type for Boolean values
as follows:
\begin{haskell}
  data Bool = False | True | Choice Bool Bool
\end{haskell}
Thus, we can implement the nondeterministic operation \code{aBool}
defined in Section~\ref{ex:aBool} as follows:
\begin{haskell}
  aBool = Choice True False
\end{haskell}
If operations can deliver nondeterministic values,
we have to extend the rules for operation defined by
pattern matching so that they do not fail on nondeterministic argument
values. Instead, they move the nondeterministic choice on level above,
i.e., a choice in some argument leads to a choice in any result
of this operation (this is also called a ``pull-tab'' step in
\cite{AlqaddoumiAntoyFischerReck10}). For instance,
the rules of the uniform operation \code{xor} shown above
are extended as follows:
%
\begin{curry}
  xor True           x = xor' x
  xor False          x = x
  xor (Choice x1 x2) x = Choice (xor x1 x) (xor x2 x)

  xor' False          = True
  xor' True           = False
  xor' (Choice x1 x2) = Choice (xor' x1) (xor' x2)
\end{curry}
%
The operation \code{xorSelf} is not defined by a pattern matching rule
and, thus, need not be changed.
If we evaluate the expression \ccode{xorSelf aBool}, we get the result
\begin{lstlisting}
  Choice (Choice False True) (Choice True False)
\end{lstlisting}
How can we interpret this result?
In principle, the choices represent different possible values.
Thus, if we want to show the different values of an expression
(which is usually the task of a top-level ``read-eval-print'' loop),
we enumerate all values contained in the choices.
These are \code{False}, \code{True}, \code{True}, and \code{False}
in the result above.
Unfortunately, this is not conform with the call-time choice semantics
discussed in Section~\ref{ctc-semantics} which should not yield
a value like \code{True}.
The call-time choice semantics requires that the choice
of a value made for
the initial expression \code{aBool} should be consistent
in the entire computation.
For instance, if we select the value \code{False} for the
expression \code{aBool}, this selection should be made
at all other places where this expression might have been copied
during the computation. However, our initial implementation
duplicates the initially single \code{Choice} into finally three
occurrences of \code{Choice}.

We can correct this wrong behavior of our implementation
by identifying different \code{Choice} occurrences that our duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{curry}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{curry}
If the expression \code{aBool} assigns the number \code{1}
to its choice, we obtain the result
\begin{lstlisting}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{lstlisting}
when evaluating the expression \ccode{xorSelf aBool}.
If we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. This, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields to only
value \code{False}.

Note that each \code{Choice} occurring for the first time in a computation
has to get its own unique identifier.
For instance, if we evaluate the expression \ccode{xor aBool aBool},
the two occurrences of \code{aBool} assign different identifiers
to their \code{Choice} constructor (e.g., \code{1} for the left
and \code{2} for the right \code{aBool} argument) so that this evaluates to
\begin{lstlisting}
  Choice 1 (Choice 2 False True) (Choice 2 True False)
\end{lstlisting}
Here we can make different selections for the outermost and inner
\code{Choice} constructors so that this nondeterministic result represent
four values.

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each nondeterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a nondeterministic argument
yields nondeterministic results for each of the argument's values.
\item Each choice occurring in a computation gets its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques for pure functional programming
and makes the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we want to avoid unsafe features in our implementation
so that we have to thread some sort of global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program.

\code{thisID} creates a fresh identifier from a current
identifier supply, and
\code{leftSupply} and \code{rightSupply} create
new distinct identifier supplies from a given one.
The latter are used when an operation based on two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the other operations must be supplied with their individual distinct
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
There are various implementations of \code{IDSupply}.
The simplest implementation uses integers:
\begin{haskell}
  type IDSupply = Integer
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
  thisID      n = n
\end{haskell}
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
he most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

\subsection{The Basic Translation Scheme}

Functional logic computations can also fail, e.g.,
due to partially defined operations.
Computing with failures is a typical programming technique
and provides for specific programming patterns \cite{AntoyHanus02FLOPS}.
Hence, in contrast to functional programming,
a failing computation should not abort the complete evaluation
but it should be considered as some part of a computation that
does not produce a meaningful result.
In order to implement this behavior, we extend
each datatype by a further constructor \code{Fail}
and complete each function with matching rules
by a final rule that matches everything and returns the value \code{Fail}.
For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is extended as follows:\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice (List a) (List a) | Fail
\end{haskell}
The operation is extended by an identifier supply and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
Note that the final rule returns \code{Fail} if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation, i.e., it also propagates failures.

As already discussed above, an occurrence of \ccode{?}
in the right-hand side is translated into a \code{Choice} supplied
with a fresh identifier by the operation \code{thisID}.
In order to ensure that each occurrence of \ccode{?} in the source program
get its own identifier, all choices and all operations in the right-hand
side of a rule get their own identier supplies via appropriate
applications of \code{leftSupply} and \code{rightSupply}
to the supply of the defined operations.
For instance, a rule like
\begin{curry}
  main2 = xor aBool (False ? True)
\end{curry}
is translated into
\begin{haskell}
  main2 s = let s1 = leftSupply  s
                s2 = rightSupply s
                s3 = leftSupply  s2
                s4 = rightSupply s2
             in xor (aBool s3) (Choice (thisID s4) False True) s1
\end{haskell}
An obvious optimization, performed by our compiler,
is a \emph{determinism analysis}.
If an operation does not call, directly or indirectly through
other operations, the choice operation \ccode{?},
then it is not necessary to pass a supply for identifiers.
In this case, the \code{IDSupply} argument can be omitted
so that the generated code is nearly identical to a corresponding
functional program (apart from the additional rules to match
the constructors \code{Choice} and \code{Fail}).

As mentioned in Section~\ref{sec:no-logic-vars},
our compiler translates occurrences of logic variables
into generators. Since these generators are standard
nondeterministic operations, they are translated like any other operation.
For instance, the operation \code{aBool} is a generator for
Boolean values and its translation into Haskell has been presented above.

The correctness of this transformation from nondeterministic source programs
into deterministic target programs has been formally shown
in \cite{Brassel11Thesis}.

\subsection{Extracting Values}

So far, our generated operations compute all the nondeterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail = Fail
  try x = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation support various options together with various
implementations of \code{IDSupply}.
For instance, a simple implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are just memory cells in Haskell:\footnote{%
In this case, integers cannot be used for \code{IDSupply}.
Nevertheless,
this type can be implemented as an infinite tree containing
all memory cells required in a computation that are created as
needed with initial value \code{NoChoice}.}
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
The implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

    newChoice ch x = do setChoice i ch
                        printValsDFS (try x)
                        setChoice i NoChoice
\end{haskell}
This operation prints a computed value and ignores failures.
If there is some choice, it checks whether a choice for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoChoice}).
If a choice has been made, it follows this choice.
Otherwise, the left choice is made and stored. After printing
all the values w.r.t.\ this choice, the choice is undone (like in backtracking)
and the right choice is made and stored.

For instance, to print all values of the expression \code{main}
defined in Section~\ref{sec:xor-main},
we evaluate the Haskell expression \ccode{printValsDFS (try (main inits))}
where \code{inits} is the initial value for \code{IDSupply}
(where all references are initialized with \code{NoChoice})
and obtain the output
\begin{haskell}
  False
  False
\end{haskell}
Of course, printing all values via depth-first search
is only one option which is not sufficient in case of infinite search
spaces. For instance, one can easily define an operation
that prints only the first solution. Due to the lazy evaluation
strategy of Haskell, such an operation can be also applied to
infinite choice structures.
In order to abstract from these different printing options,
our implementation contains a more general
approach by translating choice structures into monadic structures
w.r.t.\ various strategies (depth-first search, breadth-first search,
iterative deepening, parallel search).
This allows an independent processing of the resulting monadic structures,
e.g.,
by an interactive loop where the user can request the individual values.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Benchmarks}
\label{sec:Benchmarks}

In this section we evaluate our compiler by comparing the
efficiency of the generated Haskell programs to various
other systems, in particular, other implementations of Curry.
For our comparison with other Curry implementations,
we consider PAKCS \cite{Hanus10PAKCS} (Version 1.9.2) which is based on
compiling Curry into Prolog \cite{AntoyHanus00FROCOS}
(based on SICStus-Prolog 4.1.2, but a SWI-Prolog 5.10 back end is also
available
but much slower). PAKCS has been used for a number of practical
applications of Curry.
Another mature implementation which we consider is MCC \cite{Lux99FLOPS}
(Version 0.9.10) which compiles Curry into C.

The functional logic language Toy \cite{Lopez-FraguasSanchez-Hernandez99}
has many similarities to Curry and the Toy system compiles Toy programs
into Prolog programs. However, we have not included a comparison
in this paper since \cite{AntoyHanus00FROCOS} contains benchmarks
showing that the implementation of sharing used in PAKCS produces
more efficient programs.

Our compiler has been executed with the Glasgow Haskell Compiler
(GHC 6.12.1, option -O2). All benchmarks were executed on a Linux machine
(Ubuntu 10.10) with an Intel Core i5 (2.53GHz) processor.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark.

\begin{figure}
\centering
\begin{tabular}{|l|r|r|r|r|}
\hline
System  & Reverse & ReversePrim &  Tak  & TakPeano \\
\hline
KiCS2   &    0.13 &        0.11 &  0.24 &     0.82 \\
 \hline
PAKCS   &    2.07 &        1.91 & 39.19 &    60.69 \\
\hline
MCC     &    0.37 &        0.38 &  1.03 &     3.99 \\
\hline
GHC     &    0.11 &        0.10 &  0.04 &     0.50 \\
\hline
SICStus &    0.39 &        0.28 &  0.50 &     4.82 \\
\hline
SWI     &    1.33 &        1.14 &  1.87 &    12.72 \\
\hline
\end{tabular}
\caption{Benchmarks: first-order functional programs}
 \label{fig:bench-first-order}
\end{figure}
%
The first collection of benchmarks (Fig.~\ref{fig:bench-first-order})
are purely first-order functional programs.
The Prolog (SICStus, SWI) and Haskell (GHC) programs have been rewritten
according to the Curry formulation.
``Reverse'' is the naive reverse program applied to a list of 4096 elements.
``ReversePrim'' is the same but with user-defined lists instead of
the predefined lists. ``Tak'' is highly recursive tak function applied
to arguments (27,16,8) and ``TakPeano'' is the same but one
user-defined natural numbers in Peano representation.
Note that the Prolog programs uses a strict evaluation strategy
in contrast to all others. Thus, the difference between PAKCS and SICStus
shows the overhead to implement lazy evaluation in Prolog.

- higher-order

- nondeterministic

- logic variables

- sharing over nondet for free!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Further Aspects}
\label{sec:FurtherAspects}

- discuss search strategies

- encapsulated search

- free variables and unification

- set functions??

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Related Work}
\label{sec:Conclusions}

better program analysis to approximate purely functional and
higher-order computations

\bibliographystyle{plain}
\bibliography{mh}

\end{document}

% LocalWords:  Curry

### 2011-03-29 IDSupplyIORef.hs

@@ -1,70 +1,0 @@
-----------------
--- IDSupply
-----------------
-
-newtype ID = ID (IORef Choice) deriving Eq
-
-instance Show ID where show _ = "r"
-
-mkInt :: ID -> Integer
-mkInt = error "IDSupplyIORef.mkInt" 
-
-data IDSupply = IDSupply ID IDSupply IDSupply
-
-thisID :: IDSupply -> ID
-thisID (IDSupply i _ _) = i
-
-thisIDForFree :: IDSupply -> ID
-thisIDForFree = error "free variable must be used with other supply library"
-
-bind :: ID -> ID -> Solved
-bind i j = undefined
-
-(=:=) :: Try a -> Try a -> C_Success
-(=:=) = undefined
-
-leftID, rightID :: ID -> ID
-leftID  _ = undefined
-rightID _ = undefined
-
-nfChoiceIO :: (NormalFormIO a,NonDet a) => (a -> IO b) -> ID -> a -> a -> IO b
-nfChoiceIO = undefined
-
-narrow :: NonDet a => ID -> a -> a -> a
-narrow = choiceCons
-
-nfChoice :: (NormalForm a,NonDet b) => (a -> b) -> ID -> a -> a -> b
-nfChoice cont i x1 x2 = choiceCons i (cont $!! x1) (cont $!! x2)
-
-tryChoice :: ID -> a -> a -> Try a
-tryChoice = Choice
-
-leftSupply, rightSupply :: IDSupply -> IDSupply
-leftSupply  (IDSupply _ s _) = s
-rightSupply (IDSupply _ _ s) = s
-
-initIDSupply :: IO IDSupply
-initIDSupply = getPureSupply NoChoice
-
-{-# NOINLINE getPureSupply #-}
-getPureSupply :: Choice -> IO IDSupply
-getPureSupply def = do
-    s1 <- unsafeInterleaveIO (getPureSupply def)
-    s2 <- unsafeInterleaveIO (getPureSupply def)
-    n  <- unsafeInterleaveIO  (newIORef def)
-    return (IDSupply (ID n) s1 s2)
-
-
-lookupChoice :: ID -> IO Choice
-lookupChoice (ID r) = readIORef r
-
-setChoice :: ID -> Choice -> IO ()
-setChoice (ID r) c = writeIORef r c
-
-showsChoice :: Show a => Int -> ID -> a -> a -> ShowS
-showsChoice d r x1 x2 = 
[…]

### 2011-03-30 ID.hs
Module structure of ID and IDImpls improved; examples for function patterns added

@@ -1,14 +1,84 @@
 module ID
-  ( Choice (..), ID (..), Constraint (..), IDSupply
+  ( Constraint (..), Choice (..), defaultChoice, isDefaultChoice
+  , ID (..), IDSupply
   , mkInt, leftID, rightID, narrowID
   , initSupply, leftSupply, rightSupply, thisID, freeID
-  , lookupChoice, setChoice, setUnsetChoice
+  , lookupChoice, lookupChoiceID, setChoice, setUnsetChoice
   ) where
 
+import Control.Monad (liftM, when)
 import IDImpl
 
-leftID, rightID :: ID -> ID
+-- ---------------------------------------------------------------------------
+-- Constraint
+-- ---------------------------------------------------------------------------
+
+data Constraint
+  = ID :=: Choice
+  | Failed
+  | ConstraintChoice ID [Constraint] [Constraint]
+    deriving (Eq, Show)
+
+-- ---------------------------------------------------------------------------
+-- Choice
+-- ---------------------------------------------------------------------------
+
+-- Type to encode the selection taken in a Choice(s) structure
+data Choice
+    -- No choice has been made so far
+  = NoChoice
+  | ChooseLeft
+  | ChooseRight
+    -- ChooseN consIdx argCnt is the choice for the constructor with the
+    -- index consIdx which has argCnt arguments
+  | ChooseN Int Int
+    -- a free or narrowed variable is bound to the free variable with the
+    -- given id; the bindings of the IDs for the arguments have not been
+    -- propagated yet
+  | BindTo ID
+    -- a free or narrowed variable is bound to the variable with the given ID;
+    -- the bindings for the n arguments have also been propagated
+  | BoundTo ID Int
+  | LazyBind [Constraint]
+    deriving Show
+
+instance Eq Choice where
+  NoChoice    == NoChoice    = True
+  ChooseLeft  == ChooseLeft  = True
+  ChooseRight == ChooseRight = True
+  ChooseN c _ == ChooseN d _ = c == d
+  BindTo  i   == BindTo  j   = error "ID.Choice.(==) on BindTo"
+  BoundTo i _ == BoundTo j _ = error "ID.Choice.(==) on BoundTo"
+  LazyBind _  == LazyBind _  = error "ID.Choice.(==) on LazyBind"
[…]

### 2011-03-30 idsupplyinteger/IDImpl.hs

@@ -2,78 +2,27 @@
 -- | ID implementation using Integers
 -- ---------------------------------------------------------------------------
 module IDImpl
-  ( Choice (..), ID (..), Constraint (..), IDSupply
-  , mkInt, initSupply, leftSupply, rightSupply, thisID
-  , lookupChoiceRaw, setChoice
-  , store
+  ( Ref, mkIntRef, IDSupply, initSupply, leftSupply, rightSupply, thisRef
+  , lookupChoiceRef, setChoiceRef
   ) where
 
 import Data.IORef
-import qualified Data.Map
-import System.IO.Unsafe
+import qualified Data.Map as Map
+import System.IO.Unsafe (unsafePerformIO)
 
--- Type to encode the selection taken in a Choice structure
-data Choice
-  = NoChoice        -- No choice has been made so far
-  | ChooseLeft
-  | ChooseRight
-  | ChooseN Int Int -- ChooseN consIdx argCnt is the choice for the
-                    -- constructor with the index consIdx which has argCnt
-                    -- arguments
-  | BindTo ID       -- a free or narrowed variable is bound to the
-                    -- free variable with the given id; the bindings of the
-                    -- IDs for the arguments have not been propagated yet
-  | BoundTo ID Int  -- a free or narrowed variable is bound to the variable
-                    -- with the given ID; the bindings for the n arguments
-                    -- have also been propagated
-  | LazyBind [Constraint]
-    deriving Show
-
-instance Eq Choice where
-  NoChoice    == NoChoice    = True
-  ChooseLeft  == ChooseLeft  = True
-  ChooseRight == ChooseRight = True
-  ChooseN c _ == ChooseN d _ = c == d
-  BindTo  i   == BindTo  j   = i == j
-  BoundTo i _ == BoundTo j _ = i == j
-  LazyBind cs == LazyBind ds = cs == ds
-  _           == _           = False
-
--- Type to identify different Choice structures in a non-deterministic result.
--- Here we implement it as integer values.
-data ID
-  = ID Ref
-  | FreeID IDSupply
-  | Narrowed IDSupply
-    deriving (Eq, Ord)
[…]

### 2011-03-30 idsupplyioref/IDImpl.hs

@@ -2,53 +2,23 @@
 -- | ID implementation using IORefs
 -- ---------------------------------------------------------------------------
 module IDImpl
-  ( Choice (..), ID (..), IDSupply
-  , mkInt, initSupply, leftSupply, rightSupply, thisID
-  , lookupChoiceRaw, setChoice
+  ( Ref, mkIntRef, IDSupply, initSupply, leftSupply, rightSupply, thisRef
+  , lookupChoiceRef, setChoiceRef
   ) where
 
 import Data.IORef
-import System.IO.Unsafe
+import System.IO.Unsafe (unsafeInterleaveIO)
 
--- Type to encode the selection taken in a Choice structure
-data Choice
-  = NoChoice        -- No choice has been made so far
-  | ChooseLeft
-  | ChooseRight
-  | ChooseN Int Int -- ChooseN consIdx argCnt is the choice for the
-                    -- constructor with the index consIdx which has argCnt
-                    -- arguments
-  | BindTo ID       -- a free or narrowed variable is bound to the
-                    -- free variable with the given id; the bindings of the
-                    -- IDs for the arguments have not been propagated yet
-  | BoundTo ID Int  -- a free or narrowed variable is bound to the variable
-                    -- with the given ID; the bindings for the n arguments
-                    -- have also been propagated
-  deriving (Eq, Show)
+import {-# SOURCE #-} ID (Choice, defaultChoice, isDefaultChoice)
 
--- Type to identify different Choice structures in a non-deterministic result.
--- Here we implement it as IO references
-data ID
-  = ID Ref
-  | FreeID IDSupply
-  | Narrowed IDSupply
-    deriving Eq
+type Ref = (IORef Choice)
 
-type Ref = IORef Choice
-
-instance Show ID where
-  show (ID _)       = "ID"
-  show (FreeID _)   = "Free"
-  show (Narrowed _) = "Narrowed"
+instance Show (IORef Choice) where
+  show _ = ""
 
 -- Conversion of ID into integer not possible for this implementation
-mkInt :: ID -> Integer
-mkInt = error "IDSupplyIORef.mkInt"
-
-ref :: ID -> Ref
-ref (ID r) = r
[…]

### 2011-03-30 papers/wflp2011/paper.tex

@@
 
-\newcommand{\code}[1]{\mbox{\texttt{#1}}}
+\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
 \newcommand{\ccode}[1]{``\code{#1}''}
@@
 
-\newcommand{\Cc}{{\cal{C}}} % constructors
-\newcommand{\Fc}{{\cal{F}}} % defined functions
-\newcommand{\Var}{{\cal V}ar} % variables in an object
-\newcommand{\Dom}{{\cal D}om} % domain of a substitution
-\newcommand{\Rc}{{\cal{R}}} % rewrite system
-\newcommand{\Tc}{{\cal{T}}} % algebra of terms
-\newcommand{\Xc}{{\cal{X}}} % set of all variables
-\renewcommand{\emptyset}{\varnothing}
-\newcommand{\suc}{\mathit{success}}  % success
-\newcommand{\Suc}{\mathtt{Success}}  % success type
 \newcommand{\funset}{\ensuremath{_{\cal S}}}
-\newcommand{\seq}{\mathrel{\code{=:=}}} % strict equality of math mode
 
@@
 \begin{abstract}
-bla bla
+In this paper we present our first results of a new system
+to compile functional logic programs of the source language Curry
+into purely functional Haskell programs.
+Our implementation is based on the idea to represent
+nondeterministic result values of an expression as a data structure.
+This enables the application of various search strategies
+to extract the different values from the search space.
+In addition to the increased functionality of our system,
+we demonstrate by a several benchmarks that our implementation
+can compete with or outperforms other existing implementations of Curry.
 \end{abstract}
@@
 another constructor to represent the choice between several values.
-This idea is also the basis of Curry implementation KiCS
+This idea is also the basis of the Curry implementation KiCS
 \cite{BrasselHuch07,BrasselHuch09}.
@@
 like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
-In contrast, our implementation is not based on unsafe features.
+In contrast, our implementation, called KiCS2, is not based on unsafe features.
 In addition, we also support more flexible search strategies
@@
 of this translation scheme.
-Further aspects related to our compiler are discussed in
-Section~\ref{sec:FurtherAspects}.
-We show benchmarks of our initial implementation in
-Section~\ref{sec:Benchmarks}
+Benchmarks of our initial implementation of this scheme
+are presented in Section~\ref{sec:Benchmarks}.
+Further features of our system are sketched in
+Section~\ref{sec:FurtherAspects}
 before we conclude in Section~\ref{sec:Conclusions}.
@@
     &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
-    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\
-    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
-                                            & \mbox{let binding}\\[1ex]
+    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
+%    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
+%                                            & \mbox{let binding}\\[1ex]
   D & ::= & f(x_1,\ldots,x_n) = e
@@
 
-We can correct this wrong behavior of our implementation
-by identifying different \code{Choice} occurrences that our duplicates
+We can correct this unintended behavior of our implementation
+by identifying different \code{Choice} occurrences that are duplicates
 of some single \code{Choice}. This can be easily done by attaching
@@
 There are various implementations of \code{IDSupply}.
-The simplest implementation uses integers:
+The simplest implementation uses unbounded integers:
 \begin{haskell}
@@
 \end{haskell}
+There are other more sophisticated implementations available
+\cite{AugustssonRittriSynek94}.
 Actually, our compilation system is parameterized over different
@@
 in order to perform some experiments and choose
-he most appropiate for a given application.
+the most appropiate for a given application.
 Each implementation must ensure that, if $s$ is a value of type
@@
 in a depth-first manner by the following operation:
+\label{sec:printValsDFS}
 \begin{haskell}
@@
 \centering
-\begin{tabular}{|l|r|r|r|r|}
+\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
 \hline
-System  & Reverse & ReversePrim &  Tak  & TakPeano \\
-\hline
+System  & ReverseUser & Reverse &  Tak  & TakPeano \\\hline
 KiCS2   &        0.13 &    0.11 &  0.24 &     0.82 \\
- \hline
 PAKCS   &        2.07 &    1.91 & 39.19 &    60.69 \\
-\hline
 MCC     &        0.37 &    0.38 &  1.03 &     3.99 \\
-\hline
 GHC     &        0.11 &    0.10 &  0.04 &     0.50 \\
-\hline
 SICStus &        0.39 &    0.28 &  0.50 &     4.82 \\
-\hline
 SWI     &        1.33 &    1.14 &  1.87 &    12.72 \\
@@
 according to the Curry formulation.
-``Reverse'' is the naive reverse program applied to a list of 4096 elements.
-``ReversePrim'' is the same but with user-defined lists instead of
-the predefined lists. ``Tak'' is highly recursive tak function applied
-to arguments (27,16,8) and ``TakPeano'' is the same but one
+``ReverseUser'' is the naive reverse program applied to a list of 4096 elements,
+where all data (lists, numbers) are user-defined.
+``Reverse'' is the same but with built-in lists.
+``Tak'' is a highly recursive function on naturals applied
+to arguments (27,16,8) and ``TakPeano'' is the same but with
 user-defined natural numbers in Peano representation.
@@
 
-- higher-order
+One can deduce from these results that one of the initial
+goals for this compiler is satisfied since functional Curry programs
+are executed almost with the same speed as their Haskell equivalents.
+An overhead is visible if one uses built-in numbers
+(due to the potential nondeterministic values, KiCS2 cannot
+directly map operations on numbers directly into the Haskell primitives)
+where the GHC can apply particular optimizations.
 
-- nondeterministic
+\begin{figure}
+\centering
+\begin{tabular}{|@{~~}l@{~~}|*{5}{@{~~}r@{~~}|}}
+\hline
+System  & ReverseHO & Primes & PrimesPeano & Queens & QueensUser \\\hline
+KiCS2   &     32.82 &   1.12 &        0.29 &   9.77 &     12.41  \\
+KiCS2HO &      0.21 &   0.09 &        0.27 &   0.63 &      0.72  \\
+PAKCS   &      7.24 &  13.55 &       22.74 &  78.52 &     81.98  \\
+MCC     &      0.22 &   0.32 &        1.26 &   2.65 &      2.93  \\
+GHC     &      0.21 &   0.06 &        0.21 &   0.05 &      0.12  \\
+\hline
+\end{tabular}
+\caption{Benchmarks: higher-order functional programs}
+ \label{fig:bench-higher-order}
+\end{figure}
+%
+The next collection of benchmarks (Fig.~\ref{fig:bench-higher-order})
+considers higher-order functional programs so that we drop
+the comparison to first-order Prolog systems.
+``Reverse'' reverses a list in linear time using higher-order functions
+like \code{foldl} and \code{flip} applied to a list with one million
+elements.
+``Primes'' computes the the 2000.\ prime number via the
+sieve of Eratosthenes using higher-order functions,
+and ``PrimesPeano'' computes the 256.\ prime number but
+with Peano numbers and user-defined lists.
+Finally, ``Queens'' (and ``QueensUser'' with user-defined lists)
+computes the number of safe positions of queens on a $11 \times 11$
+chess board.
 
-- logic variables
+As discussed above, our compiler performs an optimization
+when all operations are deterministic. However, in the case
+of higher-order functions, this optimization cannot be performed
+since any operation, i.e., also a nondeterministic operation,
+can be passed as an argument. As shown in the first line of this table,
+this considerably reduces the overall performance.
+To improve this situation, our compiler generates two versions
+of a higher-order function: a general version applicable to any
+argument and a specialized version where all higher-order arguments
+are assumed to be deterministic operations.
+Moreover, we implemented a program analysis
+to approximate those operations that call higher-order functions
+with deterministic operations so that one can use their
+specialized versions. The result of this improvement
+is shown as ``KiCS2HO'' and demonstrates its usefulness so that
+it is always used in the subsequent benchmarks.
 
-- sharing over nondet for free!
+\begin{figure}
+\centering
+\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
+\hline
+System  & PermSort & PermSortPeano &  Last \\\hline
+KiCS2   &     3.23 &          1.26 &  0.26 \\
+PAKCS   &    26.08 &         21.75 &  2.58 \\
+MCC     &     1.23 &          1.28 &  0.08 \\
+\hline
+\end{tabular}
+\caption{Benchmarks: nondeterministic functional logic programs}
+ \label{fig:bench-nondet}
+\end{figure}
+%
+To evaluate the efficiency of nondeterministic computations
+(Fig.~\ref{fig:bench-nondet}),
+we sorted a list containing 15 elements by enumerating all permutations
+and selecting the sorted ones (``PermSort'' and ``PermSortPeano''
+for Peano numbers and user-defined lists with 14 elements)
+and computed the last element \code{x}
+of a list \code{xs} containing 100,000 elements
+by solving the equation \ccode{ys++[x] =:= xs}
+(the implementation of unification and variable bindings
+require some additional machinery that is sketched in
+Section~\ref{sec:unification}).
+Although the results shows that our implementation cannot compete
+with MCC, it is superior to PAKCS which exploits Prolog features
+like backtracking and logic variables for these benchmarks.
+
+Since our implementation represents nondeterministic values
+as Haskell data structures, we get, in contrast to most
+other implementations of Curry, one interesting improvement for free:
+deterministic subcomputations are shared even if they occur
+in different nondeterministic computations.
+To show this effect of our implementation, consider the
+nondeterministic sort operation \code{psort} (permutation sort)
+and the infinite list of all prime numbers \code{primes},
+as used in the previous benchmarks, and the following definitions:
+\begin{curry}
+  goal1 = [primes!!803, primes!!802, primes!!801, primes!!800]
+  goal2 = psort [6173,6163,6151,6143]
+  goal3 = psort [primes!!803, primes!!802, primes!!801, primes!!800]
+\end{curry}
+In principle, one would expect that the sum of the execution times
+of \code{goal1} and \code{goal2} is equal to the time to execute
+\code{goal3}. However, implementations based on backtracking
+evaluate the primes occurring in \code{goal3} multiple times,
+as can be seen by the run times for PAKCS and MCC shown
+in Fig~\ref{fig:bench-sharing-over-nondet}.
+%
+\begin{figure}
+\centering
+\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
+\hline
+System  &   goal1  &   goal2 & goal3 \\\hline
+KiCS2   &    0.32  &   0.00  &  0.32 \\
+PAKCS   &    8.88  &   0.02  & 91.68 \\
+MCC     &    0.20  &   0.00  &  1.94 \\
+\hline
+\end{tabular}
+\caption{Benchmarks: sharing over nondeterminism}
+ \label{fig:bench-sharing-over-nondet}
+\end{figure}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Further Aspects}
+\section{Further Features}
 \label{sec:FurtherAspects}
 
-- discuss search strategies
+In this section we sketch some additional features of our implementation.
+Due to lack of space, we cannot discuss all details of the implementation.
 
-- encapsulated search
+\subsection{Search Strategies}
 
-- free variables and unification
+Due to the fact that we represent nondeterministic results
+in a data structure rather as a computation as in implementations
+based on backtracking, we can provide different methods to
+explore the search space containing the different result values.
+We have already seen in Section~\ref{sec:printValsDFS}
+how we can explore this search space to print all values in
+depth-first order.
+Apart from this simple approach,
+our implementation contains various strategies
+(depth-first, breadth-first, iterative deepening, parallel search)
+to transform a choice structure into a list of results
+that can be printed in different ways (e.g., all solutions, only the first
+solutions, or one after another by user requests).
+Actually, the user can set options to select the search strategy and
+the printing method.
 
-- set functions??
+\begin{figure}
+\centering
+\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
+\hline
+Method               & PermSort & PermSortPeano &  NDNums \\\hline
+\code{printValsDFS}  &    3.24  &         1.26  & $\infty$ \\
+depth-first search   &    3.63  &         1.40  & $\infty$ \\
+breadth-first search &   18.30  &         6.85  &    11.52 \\
+iterative deepening  &    5.94  &         1.20  &     3.22 \\
+%parallel (1 proc.)   &    5.84  &         2.09  & $\infty$ \\
+%parallel (2 proc.)   &    3.46  &         1.25  & $\infty$ \\
+\hline
+\end{tabular}
+\caption{Benchmarks: comparing different search strategies}
+ \label{fig:search-strategies}
+\end{figure}
+%
+In order to compare the various search strategies,
+Fig.~\ref{fig:search-strategies} contains corresponding
+benchmarks where ``PermSort'' and ``PermSortPeano'' are the programs
+discussed in Fig.~\ref{fig:bench-nondet} and ``NDNums'' is the program
+\begin{curry}
+  f n = f (n+1) ? n
+\end{curry}
+where we look for the first solution of \ccode{f 0 == 25000}
+(obviously, depth-first search strategies do not terminate on this
+expression).
+All strategies except for the ``direct print'' method \code{printValsDFS}
+translate choice structures into monadic list structures
+in order to print them according to the user options.
+The benchmarks show that the overhead of this transformation
+is negligible so that this more flexible approach is the default one.
+
+We also made initial benchmarks with the parallel strategy
+where nondeterministic choices are explored via GHC's \code{par}
+construct. For the permutation sort we obtained a speedup of 1.7
+when executing the same program on two processors, but no essential
+speedup is obtained for more than two processors. 
+Better results require a careful analysis of the synchronization
+caused by the global structure to manage the state of choices
+which will be done in future work.
+
+\subsection{Encapsulated Search}
+
+Instead of choosing different search strategies to evaluate
+top-level expressions, our system also contains a primitive operation
+\begin{curry}
+ searchTree :: a -> SearchTree a
+\end{curry}
+to translate the search space caused by the evaluation of its argument
+into tree structure of the form
+\begin{curry}
+ data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
+\end{curry}
+so that the programmer can define its own search strategy or to
+collect all nondeterministic values into a list structure
+for further processing.
+
+% TODO: set functions??
+
+\subsection{Logic Variables and Unification}
+\label{sec:unification}
+
+Although our implementation is based on eliminating all logic
+variables from the source program by introducing generators,
+many functional logic programs contain equational constraints
+(e.g., compare the example ``Last'' of Fig.~\ref{fig:bench-nondet})
+to put conditions one computed results.
+Solving such conditions by generating all values is not always
+a reasonable approach. For instance, if \code{xs} and \code{ys}
+are free variables of type \code{[Bool]}, the equational
+constraint \ccode{xs=:=ys} has an infinite number of solutions.
+Instead of enumerating all these solutions, it is preferable to
+delay this enumeration but remember the condition that both
+\code{xs} and \code{ys} must always be evaluated to the same value.
+This demands for extending the representation of nondeterministic
+values by the possibility to add equational constraints between
+different choice identifiers.
+Due to lack of space, we have to omit the detailed description
+of this extension.
+However, it should be noted that the example
+``Last'' of Fig.~\ref{fig:bench-nondet}
+shows that unification can be supported with a reasonable efficiency.
 
@@
 
-better program analysis to approximate purely functional and
-higher-order computations
+We have presented a new system to compile functional logic programs
+into purely functional programs.
+In order to be consistent with the call-time choice semantics
+of functional logic languages like Curry or Toy,
+we represent nondeterministic values
+in choice structures where each choice has an identification.
+Values for such choice identifiers are passed through
+nondeterministic operations so that fresh identifiers are available
+when a new choice needs to be created.
+The theoretical justification of this implementation technique
+is provided in \cite{Brassel11Thesis}.
+Apart from the parser, where we reused an existing parser
+implemented in Haskell, the compiler is completely written in Curry.
+
+Due to the representation of nondeterministic values as data,
+our system easily supports various search strategies
+in constrast to Prolog-based implementations.
+Since we compile Curry programs into Haskell,
+we can exploit the implementation efforts done
+for functional programming.
+Hence, purely functional parts of functional logic programs
+can be executed with almost the same efficiency as Haskell programs.
+Our benchmarks show that even the execution of the
+nondeterministic parts can compete with other
+implementations of Curry.
+Furthermore, we can also exploit advanced developments
+in the implementation of Haskell, like the parallel evaluation of
+programs, which is an interesting topic for future work.
+Finally, our implementation has many opportunities
+for optimization, like
+better program analyses to approximate purely deterministic computations.
 

### 2011-03-30 Basics.hs
Bugfix

@@ -724,6 +724,9 @@ solved = return (Just (return ()))
 unsolvable :: Solved
 unsolvable = return Nothing
 
+mkSolved :: IO (IO ()) -> Solved
+mkSolved mkReset = mkReset >>= return . Just 
+
 (>>>) :: Solved -> Solved -> Solved
 a >>> b = do
   mra <- a
@@ -755,14 +758,14 @@ solve (i :=: cc) = lookupChoice i >>= choose cc
   where
     -- 1.: the Choice which should be stored for i
     -- 2.: the Choice for i in the store
-    choose (LazyBind cs) NoChoice      = setUnsetChoice i cc
-    choose _             (LazyBind cs) = setUnsetChoice i NoChoice >>> solves cs >>> solve (i :=: cc)
+    choose (LazyBind cs) NoChoice      = mkSolved (setUnsetChoice i cc)
+    choose _             (LazyBind cs) = mkSolved (setUnsetChoice i NoChoice) >>> solves cs >>> solve (i :=: cc)
     choose (LazyBind cs) _             = solves cs
 {-    choose (LazyBind cs) (LazyBind cs2) = solves cs >>> solves cs2
     choose (LazyBind cs) (ChooseN _ _)  = solves cs
     choose (ChooseN _ _) (LazyBind cs) = (setUnsetChoice i NoChoice >>> solves cs) >>> solve (i :=: cc)-}
     choose (BindTo j) ci       = lookupChoice j >>= check j ci
-    choose c          NoChoice = setUnsetChoice i c
+    choose c          NoChoice = mkSolved (setUnsetChoice i c)
     choose c          x | c==x = solved
     choose c          ci       = unsolvable
 
@@ -773,10 +776,10 @@ solve (i :=: cc) = lookupChoice i >>= choose cc
 --     check j (LazyBind cs) (ChooseN _ _)  = solves cs >>> solve (i :=: cc) -- cc = BindTo j
 --     check j c@(ChooseN _ _) (LazyBind cs) = (setUnsetChoice j NoChoice >>> solves cs) >>> solve (j :=: c)
 
-    check j NoChoice NoChoice = setUnsetChoice i (BindTo j)
+    check j NoChoice NoChoice = mkSolved (setUnsetChoice i (BindTo j))
 
-    check _ NoChoice y        = setUnsetChoice i y
-    check j x        NoChoice = setUnsetChoice j x
[…]

### 2011-03-31 papers/wflp2011/paper.tex

@@
 This enables the application of various search strategies
-to extract the different values from the search space.
-In addition to the increased functionality of our system,
-we demonstrate by a several benchmarks that our implementation
+to extract values from the search space.
+We show by a several benchmarks that our implementation
 can compete with or outperforms other existing implementations of Curry.
@@
 
+A more detailed discussion of this translation scheme can be found
+in the original proposal \cite{BrasselFischer08IFL}.
 The correctness of this transformation from nondeterministic source programs
@@
 (due to the potential nondeterministic values, KiCS2 cannot
-directly map operations on numbers directly into the Haskell primitives)
-where the GHC can apply particular optimizations.
+directly map operations on numbers into the Haskell primitives)
+where GHC can apply specific optimizations.
 
@@
 the comparison to first-order Prolog systems.
-``Reverse'' reverses a list in linear time using higher-order functions
-like \code{foldl} and \code{flip} applied to a list with one million
-elements.
+``ReverseHO'' reverses a list with one million
+elements in linear time using higher-order functions
+like \code{foldl} and \code{flip}.
 ``Primes'' computes the the 2000.\ prime number via the
@@
 \centering
-\begin{tabular}{|@{~~}l@{~~}|*{3}{@{~~}r@{~~}|}}
+\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
 \hline
-System  & PermSort & PermSortPeano &  Last \\\hline
-KiCS2   &     3.23 &          1.26 &  0.26 \\
-PAKCS   &    26.08 &         21.75 &  2.58 \\
-MCC     &     1.23 &          1.28 &  0.08 \\
+System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
+KiCS2   &     3.34 &         3.99  &  0.26 &   0.63 \\
+PAKCS   &    26.08 &        67.51  &  2.58 &  12.46 \\
+MCC     &     1.23 &         4.00  &  0.08 &   0.46 \\
 \hline
@@
 and selecting the sorted ones (``PermSort'' and ``PermSortPeano''
-for Peano numbers and user-defined lists with 14 elements)
-and computed the last element \code{x}
+for Peano numbers),
+computed the last element \code{x}
 of a list \code{xs} containing 100,000 elements
@@
 require some additional machinery that is sketched in
-Section~\ref{sec:unification}).
-Although the results shows that our implementation cannot compete
-with MCC, it is superior to PAKCS which exploits Prolog features
+Section~\ref{sec:unification}),
+and matched a regular expression in a string of length 200,000
+following the nondeterministic specification of \code{grep} shown
+in \cite{AntoyHanus10CACM}.
+Although the results show that our high-level implementation is not far
+from the efficiency of MCC,
+and it is superior to PAKCS which exploits Prolog features
 like backtracking and logic variables for these benchmarks.
@@
 \begin{curry}
-  goal1 = [primes!!803, primes!!802, primes!!801, primes!!800]
-  goal2 = psort [6173,6163,6151,6143]
-  goal3 = psort [primes!!803, primes!!802, primes!!801, primes!!800]
+goal1 = [primes!!1003, primes!!1002, primes!!1001, primes!!1000]
+goal2 = psort [7949,7937,7933,7927]
+goal3 = psort [primes!!1003, primes!!1002, primes!!1001, primes!!1000]
 \end{curry}
@@
 \hline
-System  &   goal1  &   goal2 & goal3 \\\hline
-KiCS2   &    0.32  &   0.00  &  0.32 \\
-PAKCS   &    8.88  &   0.02  & 91.68 \\
-MCC     &    0.20  &   0.00  &  1.94 \\
+System  & \code{goal1} & \code{goal2} & \code{goal3} \\\hline
+KiCS2   &        0.46  &        0.00  &        0.46  \\
+PAKCS   &       13.70  &        0.02  &      139.55  \\
+MCC     &        0.30  &        0.00  &        3.06  \\
 \hline
@@
 Method               & PermSort & PermSortPeano &  NDNums \\\hline
-\code{printValsDFS}  &    3.24  &         1.26  & $\infty$ \\
-depth-first search   &    3.63  &         1.40  & $\infty$ \\
-breadth-first search &   18.30  &         6.85  &    11.52 \\
-iterative deepening  &    5.94  &         1.20  &     3.22 \\
-%parallel (1 proc.)   &    5.84  &         2.09  & $\infty$ \\
-%parallel (2 proc.)   &    3.46  &         1.25  & $\infty$ \\
+\code{printValsDFS}  &    3.34  &         3.99  & $\infty$ \\
+depth-first search   &    3.61  &         4.39  & $\infty$ \\
+breadth-first search &   18.34  &        21.78  &    10.96 \\
+iterative deepening  &    6.16  &         7.60  &     3.10 \\
+%parallel (1 proc.)   &    5.84  &         6.54  & $\infty$ \\
+%parallel (2 proc.)   &    3.46  &         3.79  & $\infty$ \\
 \hline
@@
 Better results require a careful analysis of the synchronization
-caused by the global structure to manage the state of choices
-which will be done in future work.
+caused by the global structure to manage the state of choices.
+This is a topic for future work.
 
@@
 to translate the search space caused by the evaluation of its argument
-into tree structure of the form
+into a tree structure of the form
 \begin{curry}
@@
 collect all nondeterministic values into a list structure
-for further processing.
+for further processing \cite{BrasselHuch07}.
 

### 2011-04-01 papers/wflp2011/paper.tex
Paper updated

@@
 
-\title{KiCS2: A New Approach to Compile Curry to Haskell}
+\title{KiCS2: A New Compiler from Curry to Haskell}
 
@@
 Our implementation is based on the idea to represent
-nondeterministic result values of an expression as a data structure.
+nondeterministic result values of an expression in a single data structure.
 This enables the application of various search strategies
 to extract values from the search space.
-We show by a several benchmarks that our implementation
+We show by several benchmarks that our implementation
 can compete with or outperforms other existing implementations of Curry.
@@
 \cite{BrasselFischerHanusReck11,BrasselHuch07}.
-Although Haskell offers list comprehensions for modelling
+Although Haskell offers list comprehensions to model
 backtracking \cite{Wadler85}, this cannot be exploited
@@
 programs written in the functional logic language Curry \cite{Hanus06Curry}
-into Haskell programs
+into Haskell programs based on the ideas shown
+in \cite{BrasselFischer08IFL}.
 The difficulty of such an implementation is the fact
@@
 by the use of list comprehensions \cite{Wadler85},
-but any outcome of some operation could be potentially nondeterministic,
+but the outcome of any operation could be potentially nondeterministic,
 i.e., it might have more than one result value.
@@
 In addition, we also support more flexible search strategies
-and new features to encapsulate nondeterministic computations.
+and new features to encapsulate nondeterministic computations
+(which are not described in detail in this paper due to lack of space).
 
@@
 
-In the next section, we sketch are source language Curry
+In the next section, we sketch the source language Curry
 and introduce a normalized form of Curry programs that is the
@@
 
-The syntax the functional logic language Curry \cite{Hanus06Curry}
-is close to Haskell \cite{PeytonJones03Haskell}, i.e.,
-variables and function names usually
-start with lowercase letters, the names of type and data constructors
-start with an uppercase letter, and the application of $f$
-to $e$ is denoted by juxtaposition (``$f~e$'').
+The syntax of the functional logic language Curry \cite{Hanus06Curry}
+is close to Haskell \cite{PeytonJones03Haskell}.
 In addition, Curry allows free (logic) 
@@
 and exclude such unintended results,
-Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed the
+Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
 the rewriting logic CRWL as a logical
@@
 Since this is a complex task, if we try to implement it
-directly on the source program,
-we perform some simplifications before we generate the target code.
+directly on the level of source programs,
+we perform some simplifications on programs before we generate the target code.
 
@@
 As mentioned above, our implementation is based on the explicit
-representation of nondeterministic results of computation
-as values. This could be easily achieved by adding a constructor
+representation of nondeterministic results in a data structure.
+This could be easily achieved by adding a constructor
 to each data type to represent a choice between two values.
@@
 pattern matching so that they do not fail on nondeterministic argument
-values. Instead, they move the nondeterministic choice on level above,
+values. Instead, they move the nondeterministic choice one level above,
 i.e., a choice in some argument leads to a choice in any result
@@
 \code{leftSupply} and \code{rightSupply} create
-new distinct identifier supplies from a given one.
-The latter are used when an operation based on two\footnote{The
+new different identifier supplies (without the one obtained
+by \code{thisID}) from a given one.
+The latter are used when an operation calls two\footnote{The
 extension to more than two is straightforward.}
@@
 each datatype by a further constructor \code{Fail}
-and complete each function with matching rules
+and complete each function containing matching rules
 by a final rule that matches everything and returns the value \code{Fail}.
@@
 \end{haskell}
-The operation is extended by an identifier supply and further matching rules:
+The operation \code{hread} is extended by an identifier supply
+and further matching rules:
 \begin{haskell}
@@
 applications of \code{leftSupply} and \code{rightSupply}
-to the supply of the defined operations.
+to the supply of the defined operation.
 For instance, a rule like
@@
 To provide a common interface for such operations, we introduce
-a data type to represent the general outcome of a computation
+a data type to represent the general outcome of a computation,
 \begin{haskell}
@@
 this type can be implemented as an infinite tree containing
-all memory cells required in a computation that are created as
-needed with initial value \code{NoChoice}.}
+all memory cells required in a computation that are created
+on demand with initial value \code{NoChoice}.}
 \begin{haskell}
@@
 where \code{inits} is the initial value for \code{IDSupply}
-(where all references are initialized with \code{NoChoice})
-and obtain the output
+(where all references are initialized with \code{NoChoice}).
+Thus, we obtain the output
 \begin{haskell}
@@
 that prints only the first solution. Due to the lazy evaluation
-strategy of Haskell, such an operation can be also applied to
+strategy of Haskell, such an operation can also be applied to
 infinite choice structures.
@@
 For our comparison with other Curry implementations,
-we consider PAKCS \cite{Hanus10PAKCS} (Version 1.9.2) which is based on
-compiling Curry into Prolog \cite{AntoyHanus00FROCOS}
+we consider PAKCS \cite{Hanus10PAKCS} (Version 1.9.2) which compiles
+Curry into Prolog \cite{AntoyHanus00FROCOS}
 (based on SICStus-Prolog 4.1.2, but a SWI-Prolog 5.10 back end is also
@@
 user-defined natural numbers in Peano representation.
-Note that the Prolog programs uses a strict evaluation strategy
+Note that the Prolog programs use a strict evaluation strategy
 in contrast to all others. Thus, the difference between PAKCS and SICStus
@@
 One can deduce from these results that one of the initial
-goals for this compiler is satisfied since functional Curry programs
+goals for this compiler is satisfied, since functional Curry programs
 are executed almost with the same speed as their Haskell equivalents.
@@
 specialized versions. The result of this improvement
-is shown as ``KiCS2HO'' and demonstrates its usefulness so that
-it is always used in the subsequent benchmarks.
+is shown as ``KiCS2HO'' and demonstrates its usefulness.
+Therefore, it is always used in the subsequent benchmarks.
 
@@
 System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
-KiCS2   &     3.34 &         3.99  &  0.26 &   0.63 \\
+KiCS2   &     2.61 &         3.99  &  0.18 &   0.50 \\
 PAKCS   &    26.08 &        67.51  &  2.58 &  12.46 \\
@@
 In this section we sketch some additional features of our implementation.
-Due to lack of space, we cannot discuss all details of the implementation.
+Due to lack of space, we cannot discuss all details of them.
 
@@
 Method               & PermSort & PermSortPeano &  NDNums \\\hline
-\code{printValsDFS}  &    3.34  &         3.99  & $\infty$ \\
-depth-first search   &    3.61  &         4.39  & $\infty$ \\
-breadth-first search &   18.34  &        21.78  &    10.96 \\
-iterative deepening  &    6.16  &         7.60  &     3.10 \\
+\code{printValsDFS}  &    2.61  &         3.99  & $\infty$ \\
+depth-first search   &    5.10  &         7.62  & $\infty$ \\
+breadth-first search &   18.54  &        20.23  &    10.96 \\
+iterative deepening  &    8.33  &        10.01  &     3.10 \\
 %parallel (1 proc.)   &    5.84  &         6.54  & $\infty$ \\
@@
 In order to compare the various search strategies,
-Fig.~\ref{fig:search-strategies} contains corresponding
-benchmarks where ``PermSort'' and ``PermSortPeano'' are the programs
+Fig.~\ref{fig:search-strategies} contains some corresponding
+benchmarks.
+``PermSort'' and ``PermSortPeano'' are the programs
 discussed in Fig.~\ref{fig:bench-nondet} and ``NDNums'' is the program
@@
 The benchmarks show that the overhead of this transformation
-is negligible so that this more flexible approach is the default one.
+is acceptable so that this more flexible approach is the default one.
 
@@
 \end{curry}
-so that the programmer can define its own search strategy or to
+With this primitive,
+the programmer can define its own search strategy or to
 collect all nondeterministic values into a list structure
@@
 (e.g., compare the example ``Last'' of Fig.~\ref{fig:bench-nondet})
-to put conditions one computed results.
+to put conditions on computed results.
 Solving such conditions by generating all values is not always
@@
 of this extension.
-However, it should be noted that the example
-``Last'' of Fig.~\ref{fig:bench-nondet}
+However, it should be noted that the examples
+``Last'' and ``RegExp'' of Fig.~\ref{fig:bench-nondet}
 shows that unification can be supported with a reasonable efficiency.
@@
 is provided in \cite{Brassel11Thesis}.
-Apart from the parser, where we reused an existing parser
+Apart from the parser, where we reused an existing one
 implemented in Haskell, the compiler is completely written in Curry.
@@
 in the implementation of Haskell, like the parallel evaluation of
-programs, which is an interesting topic for future work.
+expressions, which is an interesting topic for future work.
 Finally, our implementation has many opportunities

### 2011-04-05 papers/wflp2011/paper.tex

@@
 %
-The first collection of benchmarks (Fig.~\ref{fig:bench-first-order})
+The first collection of benchmarks\footnote{All benchmarks are available
+at \url{http://www.informatik.uni-kiel.de/~mh/repos/kics2/benchmarks.tgz}.}
+(Fig.~\ref{fig:bench-first-order})
 are purely first-order functional programs.
@@
 \code{printValsDFS}  &    2.61  &         3.99  & $\infty$ \\
-depth-first search   &    5.10  &         7.62  & $\infty$ \\
-breadth-first search &   18.54  &        20.23  &    10.96 \\
+depth-first search   &    4.78  &         5.90  & $\infty$ \\
+breadth-first search &   18.35  &        20.23  &    10.96 \\
 iterative deepening  &    8.33  &        10.01  &     3.10 \\

### 2011-04-08 papers/wflp2011/paper.tex
Typos and small corrections in the wflp paper

@@
 (see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
-In particular, they combine higher-oder functions and demand-driven
+In particular, they combine higher-order functions and demand-driven
 evaluation from functional programming with logic programming features
@@
 the existing backtracking implementation for nondeterministic
-search and logic variables and unification for computing with partial
+search as well as logic variables and unification for computing with partial
 information. However, one has to implement demand-driven evaluation
@@
 it limits the implementation efforts compared to an implementation
-from scratch and one can exploit the existing implementation technologies
+from scratch and one can exploit the existing implementation technologies,
 provided that the efforts to implement the missing features
@@
 In this paper we describe an implementation that is based
-on the latter principles. We present a method to compile
+on the latter principle. We present a method to compile
 programs written in the functional logic language Curry \cite{Hanus06Curry}
@@
 If operations can deliver nondeterministic values,
-we have to extend the rules for operation defined by
+we have to extend the rules for operations defined by
 pattern matching so that they do not fail on nondeterministic argument
@@
 \code{Choice}, we also have to select the right branch in
-its selected argument \code{(Choice 1 True False)} which yields to only
+its selected argument \code{(Choice 1 True False)} which yields the only
 value \code{False}.
@@
 Here we can make different selections for the outermost and inner
-\code{Choice} constructors so that this nondeterministic result represent
+\code{Choice} constructors so that this nondeterministic result represents
 four values.
@@
 
-\code{thisID} creates a fresh identifier from a current
+\code{thisID} creates an identifier from a current
 identifier supply, and
 \code{leftSupply} and \code{rightSupply} create
-new different identifier supplies (without the one obtained
+new different identifier supplies (where both do not supply the ID obtained
 by \code{thisID}) from a given one.
@@
 \end{haskell}
-The operation \code{hread} is extended by an identifier supply
+The operation \code{head} is extended by an identifier supply
 and further matching rules:
@@
 get its own identifier, all choices and all operations in the right-hand
-side of a rule get their own identier supplies via appropriate
+side of a rule get their own identifier supplies via appropriate
 applications of \code{leftSupply} and \code{rightSupply}
@@
 is a \emph{determinism analysis}.
-If an operation does not call, directly or indirectly through
+If an operation does not call, neither directly nor indirectly through
 other operations, the choice operation \ccode{?},
@@
 using updatable values, i.e., the Haskell type \code{IORef}.
-In this case, choice identifiers are just memory cells in Haskell:\footnote{%
-In this case, integers cannot be used for \code{IDSupply}.
-Nevertheless,
-this type can be implemented as an infinite tree containing
+In this case, choice identifiers are memory cells instead of integers:\footnote{%
+Consequently, integers cannot be used for \code{IDSupply}s. Nevertheless,
+\code{IDSupply}s can be implemented as an infinite tree containing
 all memory cells required in a computation that are created
@@
 iterative deepening, parallel search).
-This allows an independent processing of the resulting monadic structures,
+This allows for an independent processing of the resulting monadic structures,
 e.g.,
@@
 Curry into Prolog \cite{AntoyHanus00FROCOS}
-(based on SICStus-Prolog 4.1.2, but a SWI-Prolog 5.10 back end is also
+(based on SICStus-Prolog 4.1.2, a SWI-Prolog 5.10 back end is also
 available
@@
 like \code{foldl} and \code{flip}.
-``Primes'' computes the the 2000.\ prime number via the
+``Primes'' computes the the 2000th prime number via the
 sieve of Eratosthenes using higher-order functions,
-and ``PrimesPeano'' computes the 256.\ prime number but
+and ``PrimesPeano'' computes the 256th prime number but
 with Peano numbers and user-defined lists.
 Finally, ``Queens'' (and ``QueensUser'' with user-defined lists)
-computes the number of safe positions of queens on a $11 \times 11$
+computes the number of safe positions of $11$ queens on a $11 \times 11$
 chess board.
@@
 in \cite{AntoyHanus10CACM}.
-Although the results show that our high-level implementation is not far
+The results show that our high-level implementation is not far
 from the efficiency of MCC,
@@
 Due to the fact that we represent nondeterministic results
-in a data structure rather as a computation as in implementations
+in a data structure rather than as a computation as in implementations
 based on backtracking, we can provide different methods to

### 2011-04-08 Basics.hs
Small refactorings

@@ -19,6 +19,9 @@ import Solver (Solution, SolutionTree (..), solves)
 nonAsciiChr :: Int# -> Char#
 nonAsciiChr i = chr# i
 
+-- TODO: Rename Frees to Free and Choices to Narrowed
+-- METATODO: Reason about pros and cons of reusing Choices for free, narrowed, (?)
+
 -- |Data type to wrap values in a generic structure
 data Try a
   = Val a                 -- ^Value in head normal form (HNF)
@@ -133,17 +136,14 @@ class (Show a, NonDet a) => NormalForm a where
   searchNF = error "searchNF not implemented"
 
 
+-- Auxiliaries for $!!
+
 -- Auxiliary function to create a binary Choice and apply a continuation to
 -- the normal forms of its two alternatives
 nfChoice :: (NormalForm a, NonDet b) => (a -> b) -> ID -> a -> a -> b
 nfChoice cont i@(ID _) x1 x2 = choiceCons i (cont $!! x1) (cont $!! x2)
 nfChoice _ _ _ _ = error "Basics.nfChoice: no ID"
 -- nfChoice cont i@(FreeID _) x1 x2 = cont (choiceCons i x1 x2)
-
-gnfChoice :: (NormalForm a, NonDet b) => (a -> b) -> ID -> a -> a -> b
-gnfChoice cont i@(ID _) x1 x2 = choiceCons i (cont $## x1) (cont $## x2)
-gnfChoice _ _ _ _ = error "Basics.gnfChoice: no ID"
-
 
 -- Auxiliary function to create a n-ary Choice and apply a continuation to
 -- the normal forms of its alternatives
@@ -152,9 +152,18 @@ nfChoices _      (ID _)       _  = error "Basics.nfChoices: ID"
 nfChoices cont i@(FreeID _)   xs = cont (choicesCons i xs)
 nfChoices cont i@(Narrowed _) xs = choicesCons i (map (cont $!!) xs)
 
+
+-- Auxiliaries for $##
+
+gnfChoice :: (NormalForm a, NonDet b) => (a -> b) -> ID -> a -> a -> b
+gnfChoice cont i@(ID _) x1 x2 = choiceCons i (cont $## x1) (cont $## x2)
+gnfChoice _ _ _ _ = error "Basics.gnfChoice: no ID"
+
 gnfChoices :: (NormalForm a, NonDet b) => (a -> b) -> ID -> [a] -> b
 gnfChoices cont i xs = narrows i (map (cont $##) xs)
 
+
[…]

### 2011-04-08 Solver.hs

@@ -7,6 +7,7 @@ module Solver
 
 import ID
 
+-- |Type for an action to reset a choice made earlier
 type Reset = IO ()
 
 type Solution = IO SolutionTree

### 2011-04-11 papers/wflp2011/paper.tex
Paper updated, SearchTree module changed

@@
 \documentclass{llncs}
+
+%\addtolength{\textheight}{0.2cm}
+%\addtolength{\topmargin}{-0.1cm}
+%\addtolength{\textwidth}{0.2cm}
+%\addtolength{\oddsidemargin}{-0.1cm}
+%\addtolength{\evensidemargin}{-0.1cm}
 
@@
 \newcommand{\us}{\char95\xspace} % underscore
+\newcommand{\dollar}{\code{\char36}} % dollar in programs
 \newcommand{\Choice}[2]{#1 \mathop{?} #2}
-\newcommand{\ol}[1]{\overline{#1}}
 
@@
 and unevaluated arguments must be shared.
-Since this is a complex task, if we try to implement it
-directly on the level of source programs,
-we perform some simplifications on programs before we generate the target code.
-
-First of all, we assume that our programs are \emph{first-order}.
-Higher-order programs can be transformed into first-order programs
-by a technique called ``defunctionalization'' \cite{Reynolds72}
-which is often used in Prolog-based implementations of
-functional logic languages \cite{AntoyHanus00FROCOS}.
-However, we will later use the higher-order features of Haskell
-instead of defunctionalization.
+This is a complex task, especially if we try to implement it directly
+on the level of source programs. Therefore, we perform some
+simplifications on programs before we generate the target code.
 
 \label{sec:no-logic-vars}
-Furthermore, we assume that our programs \emph{do not contain logic variables}.
+First of fall, we assume that our programs
+\emph{do not contain logic variables}.
 This assumption can be made since it has been shown
@@
 can be replaced by the generator \code{aBool} defined above.
+
+Furthermore, we discuss our translation scheme only for \emph{first-order}
+programs for the sake of simplicity.
+However, our implementation also supports higher-order features
+(see Section~\ref{sec:Benchmarks}) by exploiting the corresponding
+features of Haskell.
 
@@
     &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
-  P & ::= & \ol{D_k}
+  P & ::= & D_1 \ldots D_k
 \end{array}
 \]
-\caption{Uniform programs}
+\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
 \label{fig:uniform}
@@
 according to the definition in Fig.~\ref{fig:uniform}.
-There, %$\ol{o_n}$ denotes the object sequence $o_1,\ldots,o_n$,
-the variables in the left-hand sides of each rule are pairwise
+There, the variables in the left-hand sides of each rule are pairwise
 different, and
@@
 \end{haskell}
+In general, one has to propagate all choices to the top level
+of a computation before printing the results.
+This can be obtained by computing the normal form of the main
+expression before passing it to the operation \code{try}.
+Thus, the result values of \code{main} can be
+printed by evaluating
+\ccode{printValsDFS (try (id $\dollar$!! main inits))}
+where \ccode{$f$ \$!! $x$} %$
+denotes the application of the operation $f$ to the normal form
+of its argument $x$.
+
 Of course, printing all values via depth-first search
@@
 ``Reverse'' is the same but with built-in lists.
-``Tak'' is a highly recursive function on naturals applied
-to arguments (27,16,8) and ``TakPeano'' is the same but with
+``Tak'' is a highly recursive function on naturals \cite{Partain93}
+applied to arguments (27,16,8) and ``TakPeano'' is the same but with
 user-defined natural numbers in Peano representation.
@@
 \begin{curry}
- searchTree :: a -> SearchTree a
+ searchTree :: a -> IO (SearchTree a)
 \end{curry}

### 2011-04-11 lib/External_SearchTree.hs

@@ -15,6 +15,6 @@ instance MonadPlus C_SearchTree where
   mzero = C_Fail
   mplus = C_Or
 
-external_d_C_searchTree :: (NormalForm a,NonDet a) => a -> C_SearchTree a 
-external_d_C_searchTree x = searchMPlus Data.Map.empty (try (id $!! x))
+external_d_C_getSearchTree :: (NormalForm a,NonDet a) => a -> C_SearchTree a 
+external_d_C_getSearchTree x = searchMPlus Data.Map.empty (try (id $!! x))
 

### 2011-04-13 papers/wflp2011/paper.tex
Monadic compiler benchmarks added

@@
 
-%\addtolength{\textheight}{0.2cm}
-%\addtolength{\topmargin}{-0.1cm}
-%\addtolength{\textwidth}{0.2cm}
-%\addtolength{\oddsidemargin}{-0.1cm}
-%\addtolength{\evensidemargin}{-0.1cm}
+\addtolength{\textheight}{0.2cm}
+\addtolength{\topmargin}{-0.1cm}
+\addtolength{\textwidth}{0.2cm}
+\addtolength{\oddsidemargin}{-0.1cm}
+\addtolength{\evensidemargin}{-0.1cm}
 
@@
 \label{sec:Benchmarks}
+% Benchmark from April 11 on climens
 
@@
 (Version 0.9.10) which compiles Curry into C.
+MonC \cite{BrasselFischerHanusReck11} is a compiler
+from Curry into Haskell. It is based on a monadic representation
+of nondeterministic computations where sharing is explicitly
+managed by the technique proposed in \cite{FischerKiselyovShan09}.
+Since this compiler
+is in an experimental state, we could not execute all benchmarks with
+MonC (these are marked by ``n/a'').
 
@@
 (GHC 6.12.1, option -O2). All benchmarks were executed on a Linux machine
-(Ubuntu 10.10) with an Intel Core i5 (2.53GHz) processor.
+running Debian 5.0.7 with an Intel Core 2 Duo (3.0GHz) processor.
 The timings were performed with the time command measuring the
-execution time (in seconds) of a compiled executable for each benchmark.
+execution time (in seconds) of a compiled executable for each benchmark
+as a mean of three runs. ``oom'' denotes a memory overflow in a computation.
 
@@
 System  & ReverseUser & Reverse &  Tak  & TakPeano \\\hline
-KiCS2   &        0.13 &    0.11 &  0.24 &     0.82 \\
-PAKCS   &        2.07 &    1.91 & 39.19 &    60.69 \\
-MCC     &        0.37 &    0.38 &  1.03 &     3.99 \\
-GHC     &        0.11 &    0.10 &  0.04 &     0.50 \\
-SICStus &        0.39 &    0.28 &  0.50 &     4.82 \\
-SWI     &        1.33 &    1.14 &  1.87 &    12.72 \\
+KiCS2   &      0.12   &   0.12  &  0.21 &    0.79  \\
+PAKCS   &      2.05   &   1.88  & 39.80 &   62.43  \\
+MCC     &      0.43   &   0.47  &  1.21 &    5.49  \\
+MonC    &     23.39   &  22.00  & 20.37 &     oom  \\
+GHC     &      0.12   &   0.12  &  0.04 &    0.49  \\
+SICStus &      0.39   &   0.29  &  0.49 &    5.20  \\
+SWI     &      1.63   &   1.39  &  1.84 &   11.66  \\
 \hline
@@
 System  & ReverseHO & Primes & PrimesPeano & Queens & QueensUser \\\hline
-KiCS2   &     32.82 &   1.12 &        0.29 &   9.77 &     12.41  \\
-KiCS2HO &      0.21 &   0.09 &        0.27 &   0.63 &      0.72  \\
-PAKCS   &      7.24 &  13.55 &       22.74 &  78.52 &     81.98  \\
-MCC     &      0.22 &   0.32 &        1.26 &   2.65 &      2.93  \\
-GHC     &      0.21 &   0.06 &        0.21 &   0.05 &      0.12  \\
+KiCS2   &      oom  &   1.22 &       0.30  &  10.02 &     13.08  \\
+KiCS2HO &     0.24  &   0.09 &       0.27  &   0.65 &      0.73  \\
+PAKCS   &     7.97  &  14.52 &      23.08  &  81.72 &     81.98  \\
+MCC     &     0.27  &   0.32 &       1.77  &   3.25 &      3.62  \\
+MonC    &      oom  &  16.74 &        oom  &    oom &       oom  \\
+GHC     &     0.24  &   0.06 &       0.22  &   0.06 &      0.11  \\
 \hline
@@
 System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
-KiCS2   &     2.61 &         3.99  &  0.18 &   0.50 \\
-PAKCS   &    26.08 &        67.51  &  2.58 &  12.46 \\
-MCC     &     1.23 &         4.00  &  0.08 &   0.46 \\
+KiCS2   &    2.83  &        3.68   &  0.14 &   0.49 \\
+PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\
+MCC     &    1.46  &        5.74   &  0.09 &   0.57 \\
+MonC    &   48.15  &      916.61   &   n/a &    n/a \\
 \hline
@@
 System  & \code{goal1} & \code{goal2} & \code{goal3} \\\hline
-KiCS2   &        0.46  &        0.00  &        0.46  \\
-PAKCS   &       13.70  &        0.02  &      139.55  \\
-MCC     &        0.30  &        0.00  &        3.06  \\
+KiCS2   &        0.34  &        0.00  &        0.34  \\
+PAKCS   &       14.90  &        0.02  &      153.65  \\
+MCC     &        0.33  &        0.00  &        3.46  \\
 \hline
@@
 Method               & PermSort & PermSortPeano &  NDNums \\\hline
-\code{printValsDFS}  &    2.61  &         3.99  & $\infty$ \\
-depth-first search   &    4.78  &         5.90  & $\infty$ \\
-breadth-first search &   18.35  &        20.23  &    10.96 \\
-iterative deepening  &    8.33  &        10.01  &     3.10 \\
-%parallel (1 proc.)   &    5.84  &         6.54  & $\infty$ \\
-%parallel (2 proc.)   &    3.46  &         3.79  & $\infty$ \\
+\code{printValsDFS}  &    2.82  &         3.66  & $\infty$ \\
+depth-first search   &    5.33  &         6.09  & $\infty$ \\
+breadth-first search &   26.16  &        29.25  &    34.00 \\
+iterative deepening  &    9.16  &        10.91  &     0.16 \\
+%parallel (1 proc.)   &    5.46  &         6.29  & $\infty$ \\
+%parallel (2 proc.)   &    3.16  &         3.49  & $\infty$ \\
 \hline
@@
 With this primitive,
-the programmer can define its own search strategy or to
+the programmer can define its own search strategy or
 collect all nondeterministic values into a list structure

### 2011-04-15 papers/badhonnef2011/abstract.tex

\documentclass{scrartcl}

\begin{document}

\title{KiCS2: A New Compiler from Curry to Haskell}

\author{
Bernd Bra\ss{}el
\kern1em
Michael Hanus
\kern1em
 Bj{\"o}rn Peem{\"o}ller
\kern1em
Fabian Reck \\
Institut f{\"u}r Informatik, CAU Kiel, D-24098 Kiel, Germany \\
\texttt{\{bbr|mh|bjp|fre\}@informatik.uni-kiel.de}}
\date{}
\maketitle
\thispagestyle{empty}

\begin{abstract}
We present our first steps towards a new system to compile functional logic
programs of the source language Curry into purely functional Haskell programs.
This system is based on a translation scheme where non-deterministic values
are explicitly represented, i.e., by extending each type with two additional
constructors for failed computations and for a choice between several values.
Logic variables are replaced by generator functions which preserve a
call-time-choice semantics. This enables the application of various search
strategies (e.g., depth-frist, breadth-first, parallel) to extract values from
the search space, as well as the encapsulation of non-deterministic
computations. Furthermore, deterministic computations are translated into purely
functional programs so that they are executed with almost the same efficiency as
their purely functional equivalents. Several benchmarks show that our
implementation can compete with or outperforms other existing implementations
of Curry.
\end{abstract}
\end{document}

### 2011-04-18 Basics.hs
Centralized trace function

@@ -14,6 +14,7 @@ import System.IO (Handle)
 import ID
 import MonadList
 import Solver (Solution, SolutionTree (..), solves)
+import Debug
 
 
 nonAsciiChr :: Int# -> Char#
@@ -714,7 +715,7 @@ prdfs mainexp = initSupply >>= \s -> printValsDFS False print (id $!! (mainexp s
 
 printValsDFS :: (Show a, NormalForm a) => Bool -> (a -> IO ()) -> a -> IO ()
 printValsDFS fb cont a = do
---   putStrLn $ "\n\n\n-- \nprintValsDFS " ++ take 200 (show a)
+  trace $ "\n\n-- \n\nprintValsDFS " ++ take 200 (show a)
   printValsDFS' fb cont (try a)
 
 printValsDFS' :: (Show a, NormalForm a) => Bool -> (a -> IO ()) -> Try a -> IO ()

### 2011-04-18 ID.hs

@@ -8,14 +8,12 @@ module ID
   , initSupply, leftSupply, rightSupply, thisID, freeID
     -- * Choice management
   , lookupChoice, lookupID, lookupChoiceID, setChoice, setUnsetChoice
-  , setChoiceRaw, nextNIDs
+  , nextNIDs
   ) where
 
 import Control.Monad (liftM, when, zipWithM_)
 import IDImpl
-
-trace :: Bool
-trace = False
+import Debug
 
 -- ---------------------------------------------------------------------------
 -- Constraint
@@ -179,9 +177,9 @@ lookupID i = snd `liftM` lookupChoiceID i
 -- |Lookup the 'Choice' and the 'ID' an 'ID' ultimately is bound to
 lookupChoiceID :: ID -> IO (Choice, ID)
 lookupChoiceID i = do
-  when trace $ putStrLn $ "lookupChoiceID " ++ show i
+  trace $ "lookupChoiceID " ++ show i
   r <- lookupChoiceRaw i >>= unchain
-  when trace $ putStrLn $ "lookupChoiceID returned " ++ take 200 (show r)
+  trace $ "lookupChoiceID returned " ++ take 200 (show r)
   return r
   where
     -- For BindTo, we shorten chains of multiple BindTos by directly binding
@@ -195,12 +193,21 @@ lookupChoiceID i = do
         _             -> error $ "ID.lookupChoiceID: " ++ show c
       return retVal
       where
-        shortenTo lastId = when (j /= lastId) $ setChoiceRaw i (BindTo lastId)
+        shortenTo lastId = when (j /= lastId) $ do
+          trace $ "shorten " ++ show i ++ " to " ++ show lastId
+          setChoiceRaw i (BindTo lastId)
 
     -- For BoundTo, the chains should already be shortened since the Choice
     -- "BoundTo j" is only set if the variable j has been set to a "ChooseN"
     -- and therefore could not have been changed in between.
-    unchain (BoundTo j _) = lookupChoiceID j
+    unchain (BoundTo j num) = do
+      retVal@(c, lastId) <- lookupChoiceID j
+      case c of
+        NoChoice       -> return ()
+        ChooseN _ num' -> checkPropagation i j num num'
+        LazyBind _     -> return ()
+        _              -> error $ "ID.lookupChoiceID: " ++ show c
+      return retVal
 
     -- For all other choices, there are no chains at all
[…]

### 2011-04-18 papers/badhonnef2011/abstract.tex
Sentence about logic variables removed
Updates w.r.t. new naming scheme for aux. files, idci -> kics2

@@
 are explicitly represented, i.e., by extending each type with two additional
-constructors for failed computations and for a choice between several values.
-Logic variables are replaced by generator functions which preserve a
-call-time-choice semantics. This enables the application of various search
+constructors to represent failed computations and
+a choice between several values.
+This enables the application of various search
 strategies (e.g., depth-frist, breadth-first, parallel) to extract values from
 the search space, as well as the encapsulation of non-deterministic
-computations. Furthermore, deterministic computations are translated into purely
+computations.
+Logic variables are represented by non-deterministic
+operations that generate all possible values for those variables.
+To preserve the call-time semantics of functional logic languages,
+non-deterministic choices are decorated with unique identifiers
+to implement the sharing of such choices.
+Our implementation has the property that
+deterministic computations are translated into purely
 functional programs so that they are executed with almost the same efficiency as

### 2011-04-19 papers/badhonnef2011/abstract.tex
Missing word added

@@
 operations that generate all possible values for those variables.
-To preserve the call-time semantics of functional logic languages,
+To preserve the call-time-choice semantics of functional logic languages,
 non-deterministic choices are decorated with unique identifiers

### 2011-04-19 papers/wflp2011/paper.tex
Slight correction
WFLP paper updated

@@
         flexiblecolumns=false,
-        basewidth=0.55em,
+        basewidth=0.52em,
         basicstyle=\small\ttfamily}
@@
 Our implementation is based on the idea to represent
-nondeterministic result values of an expression in a single data structure.
+non-deterministic result values of an expression in a single data structure.
 This enables the application of various search strategies
@@
 evaluation from functional programming with logic programming features
-like nondeterministic search and computing with partial information
+like non-deterministic search and computing with partial information
 (logic variables).
@@
 For instance, if one compiles into Prolog, one can reuse
-the existing backtracking implementation for nondeterministic
+the existing backtracking implementation for non-deterministic
 search as well as logic variables and unification for computing with partial
@@
 of lazy evaluation and higher-order functions, but one has
-to implement nondeterministic evaluations
+to implement non-deterministic evaluations
 \cite{BrasselFischerHanusReck11,BrasselHuch07}.
@@
 due to the specific semantical requirements of the combination
-of non-strict and nondeterministic operations
+of non-strict and non-deterministic operations
 \cite{GonzalezEtAl99}. Thus, additional implementation efforts
 are necessary like implementation of shared
-nondeterministic computations \cite{FischerKiselyovShan09}.
+non-deterministic computations \cite{FischerKiselyovShan09}.
 
@@
 The difficulty of such an implementation is the fact
-that nondeterministic results can occur in any place of a
+that non-deterministic results can occur in any place of a
 computation. Thus, one cannot separate logic computations
 by the use of list comprehensions \cite{Wadler85},
-but the outcome of any operation could be potentially nondeterministic,
+but the outcome of any operation could be potentially non-deterministic,
 i.e., it might have more than one result value.
 We solve this problem by an explicit representation of
-nondeterministic values, i.e., we extend each data type by
+non-deterministic values, i.e., we extend each data type by
 another constructor to represent the choice between several values.
@@
 In addition, we also support more flexible search strategies
-and new features to encapsulate nondeterministic computations
+and new features to encapsulate non-deterministic computations
 (which are not described in detail in this paper due to lack of space).
@@
 of flexible strategies to explore the search space
-resulting from nondeterministic computations.
+resulting from non-deterministic computations.
 In contrast to Prolog-based implementations
@@
 they might yield more than one result on the same input.
-Such operations are also called \emph{nondeterministic}.
+Such operations are also called \emph{non-deterministic}.
 For instance, Curry offers a \emph{choice} operation that is predefined by
@@
 %
-Thus, we can define a nondeterministic operation \code{aBool} by
+Thus, we can define a non-deterministic operation \code{aBool} by
 \label{ex:aBool}
@@
 
-If nondeterministic operations are used as arguments in other operations,
+If non-deterministic operations are used as arguments in other operations,
 a semantical ambiguity might occur. Consider the operations
@@
 where arguments are evaluated before the function calls.
-In order to avoid such dependencies on the evaluation strategies
+In order to avoid dependencies on the evaluation strategies
 and exclude such unintended results,
@@
 (execution- and strategy-independent) foundation for declarative
-programming with non-strict and nondeterministic operations.  This
+programming with non-strict and non-deterministic operations.  This
 logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
@@
 on the level of source programs. Therefore, we perform some
-simplifications on programs before we generate the target code.
+simplifications on programs before the target code is generated.
 
@@
 \cite{AntoyHanus06ICLP} that logic variables can be replaced
-by nondeterministic ``generators,''
+by non-deterministic ``generators,''
 i.e., operations that evaluate to all possible values of the type of
@@
 
-\subsection{Representing Nondeterministic Computations}
+\subsection{Representing Non-deterministic Computations}
 
 As mentioned above, our implementation is based on the explicit
-representation of nondeterministic results in a data structure.
+representation of non-deterministic results in a data structure.
 This could be easily achieved by adding a constructor
@@
 \end{haskell}
-Thus, we can implement the nondeterministic operation \code{aBool}
+Thus, we can implement the non-deterministic operation \code{aBool}
 defined in Section~\ref{ex:aBool} as follows:
@@
 \end{haskell}
-If operations can deliver nondeterministic values,
+If operations can deliver non-deterministic values,
 we have to extend the rules for operations defined by
-pattern matching so that they do not fail on nondeterministic argument
-values. Instead, they move the nondeterministic choice one level above,
+pattern matching so that they do not fail on non-deterministic argument
+values. Instead, they move the non-deterministic choice one level above,
 i.e., a choice in some argument leads to a choice in any result
@@
 \end{curry}
-If the expression \code{aBool} assigns the number \code{1}
-to its choice, we obtain the result
+If we evaluate the expression \ccode{xorSelf aBool} and
+assign the number \code{1}
+to the choice of \code{aBool}, we obtain the result
 \begin{lstlisting}
@@
 \end{lstlisting}
-when evaluating the expression \ccode{xorSelf aBool}.
-If we show the values contained in this result,
+When we show the values contained in this result,
 we have to make \emph{consistent} selections in choices with
@@
 \end{lstlisting}
-Here we can make different selections for the outermost and inner
-\code{Choice} constructors so that this nondeterministic result represents
+Here we can make different selections for the outer and inner
+\code{Choice} constructors so that this non-deterministic result represents
 four values.
@@
 \begin{enumerate}
-\item Each nondeterministic choice is represented by a \code{Choice}
+\item Each non-deterministic choice is represented by a \code{Choice}
 constructor with a unique identifier.
@@
 moved to the result of this operation with the same identifier,
-i.e., a nondeterministic argument
-yields nondeterministic results for each of the argument's values.
+i.e., a non-deterministic argument
+yields non-deterministic results for each of the argument's values.
 \item Each choice occurring in a computation gets its own unique
@@
 is required. Unfortunately, unsafe features inhibit the use
-of optimization techniques for pure functional programming
-and makes the application of advanced evaluation and search
+of optimization techniques developed for purely functional programs
+and make the application of advanced evaluation and search
 strategies (e.g., parallel strategies) more complex.
-Therefore, we want to avoid unsafe features in our implementation
-so that we have to thread some sort of global information
+Therefore, we avoid unsafe features in our implementation.
+Instead, we thread some global information
 through our program in order to supply fresh references
@@
 into generators. Since these generators are standard
-nondeterministic operations, they are translated like any other operation.
+non-deterministic operations, they are translated like any other operation.
 For instance, the operation \code{aBool} is a generator for
@@
 in the original proposal \cite{BrasselFischer08IFL}.
-The correctness of this transformation from nondeterministic source programs
-into deterministic target programs has been formally shown
+The correctness of this transformation from non-deterministic source programs
+into deterministic target programs is formally shown
 in \cite{Brassel11Thesis}.
@@
 
-So far, our generated operations compute all the nondeterministic
+So far, our generated operations compute all the non-deterministic
 values of an expression represented by a structure containing
@@
 from choice identifiers to some value of type \code{Choice}.
-Our implementation support various options together with various
+Our implementation support various options together with different
 implementations of \code{IDSupply}.
@@
 \label{sec:Benchmarks}
-% Benchmark from April 11 on climens
+% Benchmarks from April 11 on climens
 
@@
 from Curry into Haskell. It is based on a monadic representation
-of nondeterministic computations where sharing is explicitly
+of non-deterministic computations where sharing is explicitly
 managed by the technique proposed in \cite{FischerKiselyovShan09}.
@@
 An overhead is visible if one uses built-in numbers
-(due to the potential nondeterministic values, KiCS2 cannot
+(due to the potential non-deterministic values, KiCS2 cannot
 directly map operations on numbers into the Haskell primitives)
@@
 when all operations are deterministic. However, in the case
-of higher-order functions, this optimization cannot be performed
-since any operation, i.e., also a nondeterministic operation,
+of higher-order functions, this determinism optimization cannot be performed
+since any operation, i.e., also a non-deterministic operation,
 can be passed as an argument. As shown in the first line of this table,
@@
 to approximate those operations that call higher-order functions
-with deterministic operations so that one can use their
-specialized versions. The result of this improvement
+with deterministic operations so that their specialized versions
+are used. The result of this improvement
 is shown as ``KiCS2HO'' and demonstrates its usefulness.
@@
 \end{tabular}
-\caption{Benchmarks: nondeterministic functional logic programs}
+\caption{Benchmarks: non-deterministic functional logic programs}
  \label{fig:bench-nondet}
@@
 %
-To evaluate the efficiency of nondeterministic computations
+To evaluate the efficiency of non-deterministic computations
 (Fig.~\ref{fig:bench-nondet}),
-we sorted a list containing 15 elements by enumerating all permutations
+we sort a list containing 15 elements by enumerating all permutations
 and selecting the sorted ones (``PermSort'' and ``PermSortPeano''
 for Peano numbers),
-computed the last element \code{x}
+compute the last element \code{x}
 of a list \code{xs} containing 100,000 elements
@@
 Section~\ref{sec:unification}),
-and matched a regular expression in a string of length 200,000
-following the nondeterministic specification of \code{grep} shown
+and match a regular expression in a string of length 200,000
+following the non-deterministic specification of \code{grep} shown
 in \cite{AntoyHanus10CACM}.
@@
 and it is superior to PAKCS which exploits Prolog features
-like backtracking and logic variables for these benchmarks.
-
-Since our implementation represents nondeterministic values
+like backtracking, logic variables and unification for these benchmarks.
+
+Since our implementation represents non-deterministic values
 as Haskell data structures, we get, in contrast to most
@@
 deterministic subcomputations are shared even if they occur
-in different nondeterministic computations.
+in different non-deterministic computations.
 To show this effect of our implementation, consider the
-nondeterministic sort operation \code{psort} (permutation sort)
+non-deterministic sort operation \code{psort} (permutation sort)
 and the infinite list of all prime numbers \code{primes},
@@
 \end{tabular}
-\caption{Benchmarks: sharing over nondeterminism}
+\caption{Benchmarks: sharing over non-determinism}
  \label{fig:bench-sharing-over-nondet}
@@
 
-Due to the fact that we represent nondeterministic results
+Due to the fact that we represent non-deterministic results
 in a data structure rather than as a computation as in implementations
@@
 We have already seen in Section~\ref{sec:printValsDFS}
-how we can explore this search space to print all values in
+how this search space can be explored to print all values in
 depth-first order.
@@
 that can be printed in different ways (e.g., all solutions, only the first
-solutions, or one after another by user requests).
+solution, or one after another by user requests).
 Actually, the user can set options to select the search strategy and
@@
 (obviously, depth-first search strategies do not terminate on this
-expression).
+equation).
 All strategies except for the ``direct print'' method \code{printValsDFS}
@@
 
-We also made initial benchmarks with the parallel strategy
-where nondeterministic choices are explored via GHC's \code{par}
+We also made initial benchmarks with a parallel strategy
+where non-deterministic choices are explored via GHC's \code{par}
 construct. For the permutation sort we obtained a speedup of 1.7
@@
 the programmer can define its own search strategy or
-collect all nondeterministic values into a list structure
+collect all non-deterministic values into a list structure
 for further processing \cite{BrasselHuch07}.
@@
 \code{xs} and \code{ys} must always be evaluated to the same value.
-This demands for extending the representation of nondeterministic
+This demands for extending the representation of non-deterministic
 values by the possibility to add equational constraints between
@@
 ``Last'' and ``RegExp'' of Fig.~\ref{fig:bench-nondet}
-shows that unification can be supported with a reasonable efficiency.
+show that unification can be supported with a reasonable efficiency.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions and Related Work}
+\section{Conclusions}
 \label{sec:Conclusions}
@@
 of functional logic languages like Curry or Toy,
-we represent nondeterministic values
+we represent non-deterministic values
 in choice structures where each choice has an identification.
 Values for such choice identifiers are passed through
-nondeterministic operations so that fresh identifiers are available
+non-deterministic operations so that fresh identifiers are available
 when a new choice needs to be created.
@@
 
-Due to the representation of nondeterministic values as data,
+Due to the representation of non-deterministic values as data,
 our system easily supports various search strategies
@@
 Our benchmarks show that even the execution of the
-nondeterministic parts can compete with other
+non-deterministic parts can compete with other
 implementations of Curry.

### 2011-04-20 lib/External_Distribution.hs
Libraries adapted to new naming scheme

@@ -1,22 +1,23 @@
 import Installation
+import qualified Curry_Prelude as CP
 
-external_d_C_curryCompiler :: C_String
+external_d_C_curryCompiler :: CP.C_String
 external_d_C_curryCompiler = toCurry "kics2"
 
-external_d_C_curryCompilerMajorVersion :: C_Int
+external_d_C_curryCompilerMajorVersion :: CP.C_Int
 external_d_C_curryCompilerMajorVersion = toCurry (majorVersion :: Int)
 
-external_d_C_curryCompilerMinorVersion :: C_Int
+external_d_C_curryCompilerMinorVersion :: CP.C_Int
 external_d_C_curryCompilerMinorVersion = toCurry (minorVersion ::Int)
 
-external_d_C_curryRuntime :: C_String
+external_d_C_curryRuntime :: CP.C_String
 external_d_C_curryRuntime = toCurry "ghc"
 
-external_d_C_curryRuntimeMajorVersion :: C_Int
+external_d_C_curryRuntimeMajorVersion :: CP.C_Int
 external_d_C_curryRuntimeMajorVersion = toCurry (6::Int)
 
-external_d_C_curryRuntimeMinorVersion :: C_Int
+external_d_C_curryRuntimeMinorVersion :: CP.C_Int
 external_d_C_curryRuntimeMinorVersion = toCurry (12::Int)
 
-external_d_C_installDir :: C_String
+external_d_C_installDir :: CP.C_String
 external_d_C_installDir = toCurry installDir

### 2011-04-20 lib/External_Float.hs

@@ -1,46 +1,48 @@
-external_d_C_prim_Float_plus :: C_Float -> C_Float -> C_Float
+import qualified Curry_Prelude as CP
+
+external_d_C_prim_Float_plus :: CP.C_Float -> CP.C_Float -> CP.C_Float
 external_d_C_prim_Float_plus x1 x2 =
   toCurry ((fromCurry x1 + fromCurry x1) :: Float)
 
-external_d_C_prim_Float_minus :: C_Float -> C_Float -> C_Float
+external_d_C_prim_Float_minus :: CP.C_Float -> CP.C_Float -> CP.C_Float
 external_d_C_prim_Float_minus x1 x2 =
   toCurry ((fromCurry x1 - fromCurry x1) :: Float)
 
-external_d_C_prim_Float_times :: C_Float -> C_Float -> C_Float
+external_d_C_prim_Float_times :: CP.C_Float -> CP.C_Float -> CP.C_Float
 external_d_C_prim_Float_times x1 x2 =
   toCurry ((fromCurry x1 * fromCurry x1) :: Float)
 
-external_d_C_prim_Float_div :: C_Float -> C_Float -> C_Float
+external_d_C_prim_Float_div :: CP.C_Float -> CP.C_Float -> CP.C_Float
 external_d_C_prim_Float_div x1 x2 =
   toCurry ((fromCurry x1 / fromCurry x1) :: Float)
 
-external_d_C_prim_i2f :: C_Int -> C_Float
+external_d_C_prim_i2f :: CP.C_Int -> CP.C_Float
 external_d_C_prim_i2f x = toCurry (fromInteger (fromCurry x) :: Float)
 
-external_d_C_prim_truncate :: C_Float -> C_Int
+external_d_C_prim_truncate :: CP.C_Float -> CP.C_Int
 external_d_C_prim_truncate x = toCurry (truncate (fromCurry x :: Float) :: Int)
 
-external_d_C_prim_round :: C_Float -> C_Int
+external_d_C_prim_round :: CP.C_Float -> CP.C_Int
 external_d_C_prim_round x = toCurry (round (fromCurry x :: Float) :: Int)
 
-external_d_C_prim_sqrt :: C_Float -> C_Float
+external_d_C_prim_sqrt :: CP.C_Float -> CP.C_Float
 external_d_C_prim_sqrt x = toCurry (sqrt (fromCurry x :: Float))
 
-external_d_C_prim_log :: C_Float -> C_Float
[…]

### 2011-04-20 lib/External_IO.hs

@@ -1,5 +1,6 @@
 import System.IO
 import Control.Concurrent
+import qualified Curry_Prelude as CP
 
 type C_Handle = PrimData CurryHandle
 
@@ -33,26 +34,28 @@ external_d_C_stdout = PrimData (OneHandle stdout)
 external_d_C_stderr :: C_Handle
 external_d_C_stderr = PrimData (OneHandle stderr)
 
-external_d_C_prim_openFile :: OP_List C_Char -> C_IOMode -> C_IO C_Handle
+external_d_C_prim_openFile :: CP.OP_List CP.C_Char -> C_IOMode
+                           -> CP.C_IO C_Handle
 external_d_C_prim_openFile =
   fromHaskellIO2 (\s m -> openFile s m >>= return . OneHandle)
 
-external_d_C_prim_hClose :: C_Handle -> C_IO OP_Unit
+external_d_C_prim_hClose :: C_Handle -> CP.C_IO CP.OP_Unit
 external_d_C_prim_hClose = fromHaskellIO1
   (\ch -> case ch of OneHandle h       -> hClose h
                      InOutHandle h1 h2 -> hClose h1 >> hClose h2)
 
-external_d_C_prim_hFlush :: C_Handle -> C_IO OP_Unit
+external_d_C_prim_hFlush :: C_Handle -> CP.C_IO CP.OP_Unit
 external_d_C_prim_hFlush = fromHaskellIO1 (hFlush . outputHandle)
 
-external_d_C_prim_hIsEOF :: C_Handle -> C_IO C_Bool
+external_d_C_prim_hIsEOF :: C_Handle -> CP.C_IO CP.C_Bool
 external_d_C_prim_hIsEOF = fromHaskellIO1 (hIsEOF . inputHandle)
 
-external_d_C_prim_hSeek :: C_Handle -> C_SeekMode -> C_Int -> C_IO OP_Unit
+external_d_C_prim_hSeek :: C_Handle -> C_SeekMode -> CP.C_Int
+                        -> CP.C_IO CP.OP_Unit
 external_d_C_prim_hSeek = fromHaskellIO3 (\h -> hSeek (inputHandle h))
 
 
-external_d_C_prim_hWaitForInput :: C_Handle -> C_Int -> C_IO C_Bool
+external_d_C_prim_hWaitForInput :: C_Handle -> CP.C_Int -> CP.C_IO CP.C_Bool
[…]

### 2011-04-20 lib/External_IOExts.hs

@@ -3,14 +3,15 @@ import System.IO.Unsafe(unsafePerformIO) -- for global associations
 import System.Process(runInteractiveCommand)
 import Control.Concurrent(forkIO)
 import System.IO
+import qualified Curry_Prelude as CP
 
-external_d_C_prim_execCmd :: C_String
-                          -> C_IO (OP_Tuple3 C_Handle C_Handle C_Handle)
+external_d_C_prim_execCmd :: CP.C_String
+ -> CP.C_IO (CP.OP_Tuple3 Curry_IO.C_Handle Curry_IO.C_Handle Curry_IO.C_Handle)
 external_d_C_prim_execCmd = fromHaskellIO1
   (\s -> do (h1,h2,h3,_) <- runInteractiveCommand s
             return (OneHandle h1, OneHandle h2, OneHandle h3))
 
-external_d_C_prim_connectToCmd :: C_String -> C_IO C_Handle
+external_d_C_prim_connectToCmd :: CP.C_String -> CP.C_IO Curry_IO.C_Handle
 external_d_C_prim_connectToCmd = fromHaskellIO1
   (\s -> do (hin,hout,herr,_) <- runInteractiveCommand s
             forkIO (forwardError herr)
@@ -32,12 +33,12 @@ type Assocs = [(String,String)]
 assocs :: IORef Assocs
 assocs = unsafePerformIO (newIORef [])
 
-external_d_C_prim_setAssoc :: C_String -> C_String -> C_IO OP_Unit
+external_d_C_prim_setAssoc :: CP.C_String -> CP.C_String -> CP.C_IO CP.OP_Unit
 external_d_C_prim_setAssoc = fromHaskellIO2 
   (\key val -> do as <- readIORef assocs
                   writeIORef assocs ((key,val):as))
 
-external_d_C_prim_getAssoc :: C_String -> C_IO (C_Maybe (C_String))
+external_d_C_prim_getAssoc :: CP.C_String -> CP.C_IO (CP.C_Maybe (CP.C_String))
 external_d_C_prim_getAssoc = fromHaskellIO1
   (\key -> do as <- readIORef assocs
               return (lookup key as))
@@ -79,7 +80,7 @@ instance Unifiable (C_IORef a) where
   (=.=) _ _ = error "(=.=) for C_IORef"
   bind i (Choice_C_IORef j@(FreeID _) _ _) = [i :=: (BindTo j)]
 
[…]

### 2011-04-20 lib/External_Profile.hs

@@ -1,16 +1,19 @@
 import System.Mem (performGC)
 import System.CPUTime
+import qualified Curry_Prelude as CP
 
-external_d_C_getProcessInfos :: C_IO (OP_List (OP_Tuple2 C_ProcessInfo C_Int))
+external_d_C_getProcessInfos ::
+            CP.C_IO (CP.OP_List (CP.OP_Tuple2 C_ProcessInfo CP.C_Int))
 external_d_C_getProcessInfos = fromIO $ do
   t <- getCPUTime
-  return (OP_Cons (OP_Tuple2 C_RunTime (toCurry (t `div` (10^9)))) OP_List)
+  return (CP.OP_Cons (CP.OP_Tuple2 C_RunTime (toCurry (t `div` (10^9))))
+                     CP.OP_List)
 
-external_d_C_garbageCollectorOff :: C_IO OP_Unit
+external_d_C_garbageCollectorOff :: CP.C_IO CP.OP_Unit
 external_d_C_garbageCollectorOff = fromHaskellIO0 (return ()) -- not supported
 
-external_d_C_garbageCollectorOn :: C_IO OP_Unit
+external_d_C_garbageCollectorOn :: CP.C_IO CP.OP_Unit
 external_d_C_garbageCollectorOn = fromHaskellIO0 (return ()) -- not supported
 
-external_d_C_garbageCollect :: C_IO OP_Unit
+external_d_C_garbageCollect :: CP.C_IO CP.OP_Unit
 external_d_C_garbageCollect = fromHaskellIO0 performGC

### 2011-04-20 lib/External_ReadShowTerm.hs

@@ -1,18 +1,21 @@
 import GHC.Exts (Char (C#))
+import qualified Curry_Prelude as CP
 
-external_d_C_prim_showTerm :: Show a => a -> C_String
+external_d_C_prim_showTerm :: Show a => a -> CP.C_String
 external_d_C_prim_showTerm t = toCurry (show t)
 
-external_d_C_prim_showQTerm :: Show a => a -> C_String
+external_d_C_prim_showQTerm :: Show a => a -> CP.C_String
 external_d_C_prim_showQTerm t = toCurry (show t)
 
-external_d_C_prim_readsUnqualifiedTerm :: Read a => OP_List C_String
-                                -> C_String -> OP_List (OP_Tuple2 a C_String)
+external_d_C_prim_readsUnqualifiedTerm ::
+  Read a => CP.OP_List CP.C_String -> CP.C_String
+         -> CP.OP_List (CP.OP_Tuple2 a CP.C_String)
 external_d_C_prim_readsUnqualifiedTerm _ = external_d_C_prim_readsQTerm
 
-external_d_C_prim_readsQTerm :: Read a => C_String
-                                       -> OP_List (OP_Tuple2 a C_String)
+external_d_C_prim_readsQTerm :: Read a => CP.C_String
+                             -> CP.OP_List (CP.OP_Tuple2 a CP.C_String)
 external_d_C_prim_readsQTerm s = toCurryPairs (reads (fromCurry s))
   where 
-   toCurryPairs [] = OP_List
-   toCurryPairs ((v,s):xs) = OP_Cons (OP_Tuple2 v (toCurry s)) (toCurryPairs xs)
+   toCurryPairs [] = CP.OP_List
+   toCurryPairs ((v,s):xs) = CP.OP_Cons (CP.OP_Tuple2 v (toCurry s))
+                                        (toCurryPairs xs)

### 2011-04-20 lib/External_Prelude.hs
Improved: unifiable integers

@@ -75,7 +75,7 @@ instance Show C_Int where
   showsPrec d (Guard_C_Int c e) = showsGuard d c e
   showsPrec _ Fail_C_Int = showChar '!'
   showsPrec d (C_Int x1) = shows (I# x1)
-  showsPrec d (C_CurryInt x1) = showsPrec d (integer2int x1)
+  showsPrec d (C_CurryInt x1) = (\x -> shows (I# (curryint2primint x))) $!! x1
 
 instance Read C_Int where
   readsPrec d s = map readInt (readsPrec d s) where readInt (I# i, s) = (C_Int i, s)
@@ -115,22 +115,24 @@ instance NormalForm C_Int where
   searchNF search cont (C_CurryInt x1) = search (\y1 -> cont (C_CurryInt y1)) x1
 
 instance Unifiable C_Int where
+  (=.=) (C_Int      x1) (C_Int      y1) = if (x1 ==# y1) then C_Success else Fail_C_Success
+  (=.=) (C_Int      x1) (C_CurryInt y1) = (primint2curryint x1) =:= y1
+  (=.=) (C_CurryInt x1) (C_Int      y1) = x1 =:= (primint2curryint y1)
   (=.=) (C_CurryInt x1) (C_CurryInt y1) = x1 =:= y1
-  (=.=) (C_Int x1) x2 = (int2integer x1) =.= x2
-  (=.=) x1 (C_Int x2) = x1 =.= (int2integer x2)
   (=.=) _ _ = Fail_C_Success
+  (=.<=) (C_Int      x1) (C_Int      y1) = if (x1 ==# y1) then C_Success else Fail_C_Success
+  (=.<=) (C_Int      x1) (C_CurryInt y1) = (primint2curryint x1) =:<= y1
+  (=.<=) (C_CurryInt x1) (C_Int      y1) = x1 =:<= (primint2curryint y1)
   (=.<=) (C_CurryInt x1) (C_CurryInt y1) = x1 =:<= y1
-  (=.<=) (C_Int x1) x2 = (int2integer x1) =.<= x2
-  (=.<=) x1 (C_Int x2) = x1 =.<= (int2integer x2)
   (=.<=) _ _ = Fail_C_Success
-  bind i (C_Int x2) = bind i (int2integer x2)
+  bind i (C_Int x2) = bind i (C_CurryInt (primint2curryint x2))
[…]

### 2011-04-20 papers/wflp2011/paper.tex
Paper refinements

@@
 \begin{abstract}
-In this paper we present our first results of a new system
+In this paper we present our first results of a new system % first steps towards ?
 to compile functional logic programs of the source language Curry
@@
 Our implementation is based on the idea to represent
-non-deterministic result values of an expression in a single data structure.
+non-deterministic values in the respective data structure.
 This enables the application of various search strategies
@@
 by the use of list comprehensions \cite{Wadler85},
-but the outcome of any operation could be potentially non-deterministic,
+as the outcome of any operation could be potentially non-deterministic,
 i.e., it might have more than one result value.
@@
 like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
-In contrast, our implementation, called KiCS2, is not based on unsafe features.
+In contrast, our implementation, called KiCS2, avoids such unsafe features.
 In addition, we also support more flexible search strategies
@@
 operations are shared. Hence, we can evaluate the expression above
-lazily provided that all occurrences of \code{aBool}
+lazily, provided that all occurrences of \code{aBool}
 are shared so that all of them reduce either to \code{True} or to \code{False}.
-The requirements of this call-time choice semantics is the
+The requirements of this call-time choice semantics are the
 reason why it is not simply possible to use list comprehensions
@@
 \label{sec:no-logic-vars}
-First of fall, we assume that our programs
+First of all, we assume that our programs
 \emph{do not contain logic variables}.
@@
 either a function is defined by a single rule without pattern matching,
-or it is defined by rules where only one constructor occurs in the same argument
-of all rules.\footnote{For simplicity, we require in
+or it is defined by rules with only one constructor in each rule, and
+in the same argument for all rules.
+\footnote{For simplicity, we require in
 Fig.~\ref{fig:uniform} that the matching argument is always the
@@
 representation of non-deterministic results in a data structure.
-This could be easily achieved by adding a constructor
+This can easily be achieved by adding a constructor
 to each data type to represent a choice between two values.
@@
 Thus, we can implement the non-deterministic operation \code{aBool}
-defined in Section~\ref{ex:aBool} as follows:
+defined in Section~\ref{ex:aBool} as:
 \begin{haskell}
@@
 In this case, choice identifiers are memory cells instead of integers:\footnote{%
-Consequently, integers cannot be used for \code{IDSupply}s. Nevertheless,
+Consequently, integers cannot be used for \code{IDSupply}s. Instead,
 \code{IDSupply}s can be implemented as an infinite tree containing
@@
 System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
-KiCS2   &    2.83  &        3.68   &  0.14 &   0.49 \\
+KiCS2HO &    2.83  &        3.68   &  0.14 &   0.49 \\
 PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\
@@
 System  & \code{goal1} & \code{goal2} & \code{goal3} \\\hline
-KiCS2   &        0.34  &        0.00  &        0.34  \\
+KiCS2HO &        0.34  &        0.00  &        0.34  \\
 PAKCS   &       14.90  &        0.02  &      153.65  \\

### 2011-04-20 papers/wflp2011/paper.tex
Final changes for submitted WFLP paper

@@
 \begin{abstract}
-In this paper we present our first results of a new system % first steps towards ?
+In this paper we present our first steps towards a new system
 to compile functional logic programs of the source language Curry
@@
 either a function is defined by a single rule without pattern matching,
-or it is defined by rules with only one constructor in each rule, and
-in the same argument for all rules.
+or it is defined by rules with only one constructor in the left-hand
+side of each rule, and in the same argument for all rules.%
 \footnote{For simplicity, we require in

### 2011-05-01 talks/talk.tex

\documentclass[utf8,handout]{beamer}

\usepackage{graphicx}
\usepackage{color}
\usepackage{listings}
\usetheme{Boadilla}

\setbeamertemplate{footline}%{infolines theme}
{
\leavevmode%
\hbox{%
\begin{beamercolorbox}[wd=.333333\paperwidth,ht=2.25ex,dp=1ex,center]{author in head/foot}%
\usebeamerfont{author in head/foot}\insertshortauthor%~~(\insertshortinstitute)
\end{beamercolorbox}%
\begin{beamercolorbox}[wd=.333333\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
\usebeamerfont{title in head/foot}\insertshorttitle
\end{beamercolorbox}%
\begin{beamercolorbox}[wd=.333333\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
\usebeamerfont{date in head/foot}\insertshortdate{}\hspace*{2em}
\insertframenumber{} / \inserttotalframenumber\hspace*{2ex}
\end{beamercolorbox}}%
\vskip0pt%
}

\begin{document}

\title[KiCS2]{KiCS2: A New Compiler from Curry to Haskell}

\date{May 04, 2011}

\author[B. Braßel, M. Hanus, B. Peemöller, F. Reck]
 {Bernd Braßel \and Michael Hanus \\
   Björn Peemöller \and Fabian Reck\\
  \texttt{\{bbr, mh, bjp, fre\}@informatik.uni-kiel.de}}

\institute{Kiel University}

\begin{frame}
  \titlepage
\end{frame}

\section{Introduction}

\begin{frame}[fragile]
\frametitle{Curry}
\begin{columns}
\column{.45\textwidth}
\begin{itemize}
\item Lazy functional logic programing language.
\item Haskell-like syntax.
\item Extended by non-determinism and free variables. %TODO: failure?
\end{itemize}
\column{.45\textwidth}
\begin{block}{Running example}
\begin{semiverbatim}
data Bool = True | False

aBool :: Bool
aBool = True ? False

not True  = False
not False = True

xor True  x = not x
xor False x = x

xorSelf x = xor x x
\end{semiverbatim}
\end{block}
\end{columns}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example Call}
\begin{semiverbatim}
> xorSelf aBool
False
False
No more solutions.
\end{semiverbatim}
\begin{itemize}
\item The call of \verb!aBool! yields two solutions
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{KiCS2}
% Compilation strategies
%   - compile to logic language (Pakcs compiles to Prolog)
%   - compile to functional language (KiCS compiles to Haskell)
%   - compile to abstract machine (MCC compiles to C)
\begin{block}{Approach}
\begin{itemize}
\item Compile Curry programs into Haskell
\item Reuse expresiveness of Haskell (lazy evaluation, higher-order functions)
\item Benefit from mature Haskell compiler (GHC)
\end{itemize}
\end{block}

\begin{block}{Goals}
\begin{itemize}
\item Efficient execution of purely functional programs
\item Avoid unsafe features to enable optimization (in contrast to KiCS)
\item Support for different search strategies (including parallel strategies)
\end{itemize}
\end{block}
\end{frame}

\section{Compilation scheme}

\begin{frame}[fragile]
\frametitle{Representation of Non-Determinism}
\begin{itemize}
  \item No built-in non-determinism in Haskell.
  \item Idea: Explicit representation in data types as constructors.
\end{itemize}

\begin{block}{Curry}
\begin{semiverbatim}
data Bool = True | False

aBool :: Bool
aBool = True ? False
\end{semiverbatim}
\end{block}

\begin{block}{Haskell}
\begin{semiverbatim}
data Bool = True | False | Choice_Bool Bool Bool

aBool :: Bool
aBool = Choice_Bool True False
\end{semiverbatim}
\end{block}
\end{frame}

\begin{frame}[fragile]
\frametitle{Extension of the Operations}
\begin{itemize}
  \item Operations have to deal with non-deterministic values.
  \item Non-determinism in arguments is propagated.
%   \item Purely functional Curry programs are Haskell programs (no runtime overhead) (first-order)
\end{itemize}

\begin{block}{Haskell}
\begin{semiverbatim}
not True                = False
not False               = True
not (Choice_Bool x1 x2) = Choice_Bool (not x1) (not x2)

xor True                x = not x
xor False               x = x
xor (Choice_Bool x1 x2) x = Choice_Bool (xor x1 x)
                                        (xor x2 x)

xorSelf x = xor x x
\end{semiverbatim}
\end{block}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example calls}
\begin{example}
\verb!not aBool!
$\leadsto$ \verb!not (Choice_Bool True False)!
$\leadsto$ \verb!Choice_Bool (not True) (not False)!
$\leadsto$ \verb!Choice_Bool False True!
\end{example}

\begin{example}
\verb!xorSelf aBool!
$\leadsto$ \verb!xor aBool aBool!
$\leadsto$ \verb!xor (Choice_Bool True False) aBool!
$\leadsto$ \verb!Choice_Bool (xor True aBool) (xor False aBool)!
$\leadsto$ \verb!Choice_Bool (not aBool) aBool!
$\leadsto$ \verb!Choice_Bool (Choice_Bool False True) (Choice_Bool True False)!
\end{example}

\begin{itemize}
  \item Yiels four results: \verb!False, True, True, False!
  \item Only \verb!False, False! are reasonable.
\end{itemize}

\end{frame}

\begin{frame}[fragile]
\frametitle{Run-Time Choice vs. Call-Time Choice}
\begin{block}{Curry}
\begin{lstlisting}
xorSelf x = xor x x
\end{lstlisting}
\end{block}
\pause
\begin{block}{Run-Time Choice}
\begin{itemize}
\item Variables denote non-deterministic computations
\item {\tt x}'s may evaluate to different values in the same non-deterministic branch
\end{itemize}
\end{block}
\pause
\begin{block}{Call-Time Choice}
\begin{itemize}
\item Variables denote values
\item Both {\tt x} evaluate to the same value in each non-deterministic branch
\end{itemize}
\end{block}
\pause
\begin{itemize}
\item Call-Time Choice is often more intuitive
\item Curry has Call-Time Choice semantics 
\item Unrestricted propagation violates Call-Time Choice
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Identifying Choices}
\begin{itemize}
\item Choices are annotated with unique identifiers
\item Functions that may introduce non-determinism are supplied with an additional parameter to obtain fresh \verb!ID!s
\item Decisions made for different occurrences of the same Choice have to be consistent
\end{itemize}

\begin{block}{Choice Identifier}
\begin{semiverbatim}
type ID = Integer

data Bool = False | True | Choice_Bool ID Bool Bool

aBool :: ID -> Bool
aBool i = Choice_Bool i True False

xor (Choice_Bool i x1 x2) x = Choice_Bool i (xor x1 x) 
                                            (xor x2 x)
\end{semiverbatim}
\end{block}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example call}
% Mention that (aBool 1) is computed only once due to sharing
% Mention the origin of the initial ID 1
\begin{example}
\verb!xorSelf (aBool 1)!
$\leadsto$ \verb!xor (aBool 1) (aBool 1)!
$\leadsto$ \verb!xor (Choice_Bool 1 True False) (Choice_Bool 1 True False)!
$\leadsto$ $\dots$
$\leadsto$ \verb!Choice_Bool 1 (Choice_Bool 1 False True) (Choice_Bool 1 True False)!
\end{example}

\begin{itemize}
\item The results \verb!True, True! are identified as invalid
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Representing Failure}
\begin{itemize}
 \item Computations with failures is a common programming pattern in FLP
 \item Failures do not abort the computation, but are ignored
 \item Thus, failures have to be represented as well.
\end{itemize}

\begin{block}{Curry}
\begin{semiverbatim}
data Bool = True | False

ensureTrue True = True
\end{semiverbatim}
\end{block}

\begin{block}{Haskell}
\begin{semiverbatim}
data Bool = True | False | Choice_Bool Bool Bool | Fail_Bool

ensureTrue True                  = True
ensureTrue (Choice_Bool i x1 x2) = ...
ensureTrue _                     = Fail_Bool
\end{semiverbatim}
\end{block}
\end{frame}

\begin{frame}[fragile]
\frametitle{Higher Order Functions}
\begin{itemize}
\item Up to now only first-order programs were considered
\item Deterministic higher-order functions can remain unchanged
\end{itemize}
\begin{columns}
\column{.45\textwidth}
\begin{block}{Curry}
\begin{semiverbatim}
apply :: (a -> b) -> a -> b
apply f x = f x

> apply not True
False
\end{semiverbatim}
\end{block}
\column{.45\textwidth}
\begin{block}{Haskell}
\begin{semiverbatim}
apply :: (a -> b) -> a -> b
apply f x = f x

> apply not True
False
\end{semiverbatim}
\end{block}
\end{columns}
\end{frame}

\begin{frame}[fragile]
\frametitle{Higher Order Functions}
\begin{itemize}
\item Non-deterministic functions need an additional parameter
\item An identifier is passed to functional argument when applied
\end{itemize}
\begin{columns}
\column{.45\textwidth}
\begin{block}{Curry}
\begin{semiverbatim}
apply :: (a -> b) -> a -> b
apply f x = f x

ndNot x = x ? not x

> apply ndNot True
True
False
\end{semiverbatim}
\end{block}
\column{.45\textwidth}
\begin{block}{Haskell}
\begin{semiverbatim}
apply :: (a -> ID -> b) 
      -> a -> ID -> b
apply f x i = f x i

ndNot x i = 
  Choice_Bool i x (not x)

> apply ndNot True 1
Choice_Bool 1 True False
\end{semiverbatim}
\end{block}
\end{columns}
\end{frame}

\begin{frame}[fragile]
\frametitle{Higher-order optimization}
Deterministic functions have to be wrapped
\begin{block}{Curry}
\begin{semiverbatim}
> apply not True
False
\end{semiverbatim}
\end{block}
\begin{block}{Haskell}
\begin{semiverbatim}
> apply (\\x _ -> not x) True 1
False
\end{semiverbatim}
\end{block}
\begin{block}{Optimization}
\begin{itemize}
\item Higher-order functions are translated into
  \begin{itemize}
   \item a deterministic variant
   \item a non-deterministic variant
  \end{itemize}
\item Benchmarks show enormously improved performance
\end{itemize}
\end{block}
Non-deterministic choice between functions is also possible.
\end{frame}

\section{Benchmarks}

\begin{frame}
\frametitle{Benchmarks}
\end{frame}

\section{Conclusion}

\begin{frame}
\frametitle{Additional Features}
\begin{itemize}

\item Free Variables
\item Unification
\item Function Patterns
\item Encapsulated Search

\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusion}
\begin{itemize}
\item A transformation of a functional logic program into a functional program.
\item The resulting program can be optimized by the Haskell compiler
\item A formal description of our approach can be found in our paper.
\end{itemize}
\end{frame}

\begin{frame}
\begin{center}

 {\huge Thank You!}\\[3ex]
 \pause
 {\huge Questions?}

\end{center}
\end{frame}

\end{document}

### 2011-05-03 papers/wflp2011/paper.tex
Paper update w.r.t. reviewers' comments

@@
 \documentclass{llncs}
-
-\addtolength{\textheight}{0.2cm}
-\addtolength{\topmargin}{-0.1cm}
-\addtolength{\textwidth}{0.2cm}
-\addtolength{\oddsidemargin}{-0.1cm}
-\addtolength{\evensidemargin}{-0.1cm}
 
@@
 \newcommand{\Choice}[2]{#1 \mathop{?} #2}
-
-% ------------------------------------------------------------------
-% SYMBOLS & FORMULAS
-
 \newcommand{\funset}{\ensuremath{_{\cal S}}}
@@
 \begin{document}
-\pagestyle{plain}
+%\pagestyle{plain}
 \sloppy
@@
 Our implementation is based on the idea to represent
-non-deterministic values in the respective data structure.
+non-deterministic results as values in the data types corresponding
+to the results.
 This enables the application of various search strategies
@@
     &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
-%    &  |  & \mbox{\code{let}}~\{x_1=e_1;\ldots;x_n = e_n\}~\code{in}~e
-%                                            & \mbox{let binding}\\[1ex]
   D & ::= & f(x_1,\ldots,x_n) = e
@@
 Thus, we assume that all programs are \emph{uniform}
-according to the definition in Fig.~\ref{fig:uniform}.
+according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
+Uniform programs have been also considered in
+\cite{MorenoEtAl90ALP} to define lazy narrowing strategies
+for functional logic programs. Although the motivations are similar,
+our notion of uniform program is more restrictive since we
+allow only a single non-variable argument in each left-hand side
+of a rule.}
 There, the variables in the left-hand sides of each rule are pairwise
@@
 in the result above.
-Unfortunately, this is not conform with the call-time choice semantics
+Unfortunately, this does not conform to the call-time choice semantics
 discussed in Section~\ref{ctc-semantics} which should not yield
@@
 yields non-deterministic results for each of the argument's values.
-\item Each choice occurring in a computation gets its own unique
+\item Each choice introduced in a computation gets its own unique
 identifier.
@@
 \begin{haskell}
+  initSupply  :: IO IDSupply
   thisID      :: IDSupply -> ID
@@
 and add a new argument of type \code{IDSupply} to each operation
-of the source program.
-
-\code{thisID} creates an identifier from a current
-identifier supply, and
-\code{leftSupply} and \code{rightSupply} create
-new different identifier supplies (where both do not supply the ID obtained
-by \code{thisID}) from a given one.
-The latter are used when an operation calls two\footnote{The
+of the source program, i.e., a Curry operation of type
+\begin{curry}
+  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$->$~\tau$
+\end{curry}
+is translated into a Haskell function of type
+\begin{haskell}
+  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$-> IDSupply ->$~\tau$
+\end{haskell}
+%
+Conceptually, one can consider \code{IDSupply}
+as an infinite set of identifiers that is created at the beginning
+of an evaluation by the operation \code{initSupply}.
+The operation \code{thisID} takes some identifier from this set, and
+\code{leftSupply} and \code{rightSupply} splits this set
+into two disjoint subsets without the identifier
+obtained by \code{thisID}.
+The split operations \code{leftSupply} and \code{rightSupply}
+are used when an operation calls two\footnote{The
 extension to more than two is straightforward.}
 other operations in the right-hand side of a rule. In this case,
-the other operations must be supplied with their individual distinct
+the other operations must be supplied with their individual disjoint
 identifier supplies. For instance, the operation \code{main}
@@
 \begin{curry}
+  main :: Bool
   main = xorSelf aBool
@@
 \begin{haskell}
+  main :: IDSupply -> Bool
   main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
@@
   type IDSupply = Integer
+  initSupply    = return 1
+  thisID      n = n
   leftSupply  n = 2*n
   rightSupply n = 2*n+1
-  thisID      n = n
 \end{haskell}
@@
 from choice identifiers to some value of type \code{Choice}.
-Our implementation support various options together with different
+Our implementation supports various options together with different
 implementations of \code{IDSupply}.
-For instance, a simple implementation can be obtained by
+For instance, a simple but efficient implementation can be obtained by
 using updatable values, i.e., the Haskell type \code{IORef}.
-In this case, choice identifiers are memory cells instead of integers:\footnote{%
-Consequently, integers cannot be used for \code{IDSupply}s. Instead,
-\code{IDSupply}s can be implemented as an infinite tree containing
-all memory cells required in a computation that are created
-on demand with initial value \code{NoChoice}.}
+In this case, choice identifiers are memory cells instead of integers:
 \begin{haskell}
@@
 \end{haskell}
-The implementation of the lookup and set operations is straightforward:
+Consequently, an implementation of \code{IDSupply} requires
+an infinite set of memory cells which can represented as a tree structure:
+\begin{haskell}
+  data IDSupply = IDSupply ID IDSupply IDSupply
+  thisID      (IDSupply r _ _) = r
+  leftSupply  (IDSupply _ s _) = s
+  rightSupply (IDSupply _ _ s) = s
+  \end{haskell}
+The infinite tree of memory cells (with initial value \code{NoChoice})
+can be constructed as follows,
+where the operation \code{unsafeInterleaveIO} is used to construct
+the tree on demand:
+\begin{haskell}
+  initSupply = getIDTree
+  getIDTree = do s1 <- unsafeInterleaveIO getIDTree
+                 s2 <- unsafeInterleaveIO getIDTree
+                 r  <- unsafeInterleaveIO (newIORef NoChoice)
+                 return (IDSupply (ID r) s1 s2)
+  \end{haskell}
+Using memory cells,
+the implementation of the lookup and set operations is straightforward:
 \begin{haskell}
@@
 defined in Section~\ref{sec:xor-main},
-we evaluate the Haskell expression \ccode{printValsDFS (try (main inits))}
-where \code{inits} is the initial value for \code{IDSupply}
-(where all references are initialized with \code{NoChoice}).
+we evaluate the Haskell expression
+\begin{haskell}
+  initSupply >>= \s -> printValsDFS (try (main s))
+\end{haskell}
 Thus, we obtain the output
@@
 \end{haskell}
-In general, one has to propagate all choices to the top level
+In general, one has to propagate all choices and failures to the top level
 of a computation before printing the results.
-This can be obtained by computing the normal form of the main
+Otherwise, the operation \code{try} applied to an expression
+like \ccode{Just aBool} would return a \code{Val}-structure
+instead of a \code{Choice}
+so that the main operation \code{printValsDFS} would miss
+the non-determinism of the result value.
+Therefore, we have to compute the normal form of the main
 expression before passing it to the operation \code{try}.
-Thus, the result values of \code{main} can be
-printed by evaluating
-\ccode{printValsDFS (try (id $\dollar$!! main inits))}
+Hence, the result values of \code{main} are printed by evaluating
+\begin{haskell}
+  initSupply >>= \s -> printValsDFS (try (id $\dollar$!! main s))
+\end{haskell}
 where \ccode{$f$ \$!! $x$} %$
@@
 of its argument $x$.
+This has the effect that a choice or failure occurring somewhere in
+a computation will be moved (by the operation \ccode{\$!!})
+to the root of the main expression
+so that the corresponding search strategy can process it.
+This ensures that, after the computation to a normal form,
+an expression without a \code{Choice} or \code{Fail} at the root
+is a value, i.e., it does not contain a \code{Choice} or \code{Fail}.
 
@@
 \label{sec:Benchmarks}
-% Benchmarks from April 11 on climens
 
@@
 like \code{foldl} and \code{flip}.
-``Primes'' computes the the 2000th prime number via the
+``Primes'' computes the 2000th prime number via the
 sieve of Eratosthenes using higher-order functions,
@@
 In this section we sketch some additional features of our implementation.
-Due to lack of space, we cannot discuss all details of them.
+Due to lack of space, we cannot discuss them in detail.
 
@@
 iterative deepening  &    9.16  &        10.91  &     0.16 \\
-%parallel (1 proc.)   &    5.46  &         6.29  & $\infty$ \\
-%parallel (2 proc.)   &    3.16  &         3.49  & $\infty$ \\
 \hline
@@
 for further processing \cite{BrasselHuch07}.
-
-% TODO: set functions??
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
+\section{Conclusions and Related Work}
 \label{sec:Conclusions}
@@
 implementations of Curry.
-Furthermore, we can also exploit advanced developments
+
+We have already discussed in the introduction the various
+efforts to implement functional logic languages,
+like the construction of abstract machines
+\cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP,Lux99FLOPS} and
+the compilation into Prolog \cite{AntoyHanus00FROCOS}
+or Haskell \cite{BrasselFischerHanusReck11,BrasselHuch07,BrasselHuch09}.
+Our benchmarks show that an efficient implementation
+by compiling into
+a functional language depends on carefully handling
+the sharing of non-deterministic choices. For instance, our previous
+implementation \cite{BrasselFischerHanusReck11}, where sharing
+is explicitly managed by the monadic techniques proposed
+in \cite{FischerKiselyovShan09}, has not satisfied the expectations
+that came from the benchmarks reported in \cite{FischerKiselyovShan09}.
+Due to these experiences, we used in our new compiler the
+compilation scheme initially proposed in \cite{BrasselFischer08IFL}
+which produces much faster code, as shown in our benchmarks.
+
+If non-deterministic results are collected in data structures,
+one has more fine-grained control over non-deterministic steps.
+For instance, \cite{AlqaddoumiAntoyFischerReck10} proposes
+pull-tab steps to move non-determinism from arguments to
+the result position of a function. Antoy \cite{Antoy11ICLP}
+shows that single pull-tab steps are semantics-preserving.
+Thus, it is not necessary to move each choice to the root
+of an expression, as done in our implementation,
+but one could also perform further local computations in the arguments
+of a choice before moving it up. This might be a reasonable strategy
+if all non-deterministic values are required but many computations fail.
+However, the general effects of such refinements need further
+investigations.
+
+Our implementation has many opportunities for optimization, like
+better program analyses to approximate purely deterministic computations.
+We can also exploit advanced developments
 in the implementation of Haskell, like the parallel evaluation of
-expressions, which is an interesting topic for future work.
-Finally, our implementation has many opportunities
-for optimization, like
-better program analyses to approximate purely deterministic computations.
-
-\bibliographystyle{plain}
-\bibliography{mh}
+expressions. These are interesting topics for future work.
+
+\begin{thebibliography}{10}
+
+\bibitem{AlbertHanusHuchOliverVidal05}
+E.~Albert, M.~Hanus, F.~Huch, J.~Oliver, and G.~Vidal.
+\newblock Operational semantics for declarative multi-paradigm languages.
+\newblock {\em Journal of Symbolic Computation}, 40(1):795--829, 2005.
+
+\bibitem{AlqaddoumiAntoyFischerReck10}
+A.~Alqaddoumi, S.~Antoy, S.~Fischer, and F.~Reck.
+\newblock The pull-tab transformation.
+\newblock In {\em Proc. of the Third International Workshop on Graph
+  Computation Models}, pages 127--132. Enschede, The Netherlands, 2010.
+\newblock Available at
+  http://gcm-events.org/gcm2010/pages/gcm2010-preproceedings.pdf.
+
+\bibitem{Antoy01PPDP}
+S.~Antoy.
+\newblock Constructor-based conditional narrowing.
+\newblock In {\em Proc.\ of the 3rd International ACM SIGPLAN Conference on
+  Principles and Practice of Declarative Programming (PPDP 2001)}, pages
+  199--206. ACM Press, 2001.
+
+\bibitem{Antoy11ICLP}
+S.~Antoy.
+\newblock On the correctness of the pull-tab transformation.
+\newblock In {\em To appear in Proceedings of the 27th International Conference
+  on Logic Programming (ICLP 2011)}, 2011.
+
+\bibitem{AntoyHanus00FROCOS}
+S.~Antoy and M.~Hanus.
+\newblock Compiling multi-paradigm declarative programs into {Prolog}.
+\newblock In {\em Proc. International Workshop on Frontiers of Combining
+  Systems (FroCoS'2000)}, pages 171--185. Springer LNCS 1794, 2000.
+
+\bibitem{AntoyHanus02FLOPS}
+S.~Antoy and M.~Hanus.
+\newblock Functional logic design patterns.
+\newblock In {\em Proc.\ of the 6th International Symposium on Functional and
+  Logic Programming (FLOPS 2002)}, pages 67--87. Springer LNCS 2441, 2002.
+
+\bibitem{AntoyHanus06ICLP}
+S.~Antoy and M.~Hanus.
+\newblock Overlapping rules and logic variables in functional logic programs.
+\newblock In {\em Proceedings of the 22nd International Conference on Logic
+  Programming (ICLP 2006)}, pages 87--101. Springer LNCS 4079, 2006.
+
+\bibitem{AntoyHanus10CACM}
+S.~Antoy and M.~Hanus.
+\newblock Functional logic programming.
+\newblock {\em Communications of the ACM}, 53(4):74--85, 2010.
+
+\bibitem{AntoyHanusLiuTolmach05}
+S.~Antoy, M.~Hanus, J.~Liu, and A.~Tolmach.
+\newblock A virtual machine for functional logic computations.
+\newblock In {\em Proc.\ of the 16th International Workshop on Implementation
+  and Application of Functional Languages (IFL 2004)}, pages 108--125. Springer
+  LNCS 3474, 2005.
+
+\bibitem{AugustssonRittriSynek94}
+L.~Augustsson, M.~Rittri, and D.~Synek.
+\newblock On generating unique names.
+\newblock {\em Journal of Functional Programming}, 4(1):117--123, 1994.
+
+\bibitem{Brassel11Thesis}
+B.~Bra{\ss}el.
+\newblock {\em Implementing Functional Logic Programs by Translation into
+  Purely Functional Programs}.
+\newblock PhD thesis, Christian-Albrechts-Universit{\"a}t zu Kiel, 2011.
+
+\bibitem{BrasselFischer08IFL}
+B.~Bra{\ss}el and S.~Fischer.
+\newblock From functional logic programs to purely functional programs
+  preserving laziness.
+\newblock In {\em Pre-Proceedings of the 20th Workshop on Implementation and
+  Application of Functional Languages (IFL 2008)}, 2008.
+
+\bibitem{BrasselFischerHanusReck11}
+B.~Bra{\ss}el, S.~Fischer, M.~Hanus, and F.~Reck.
+\newblock Transforming functional logic programs into monadic functional
+  programs.
+\newblock In {\em Proc. of the 19th International Workshop on Functional and
+  (Constraint) Logic Programming (WFLP 2010)}, pages 30--47. Springer LNCS
+  6559, 2011.
+
+\bibitem{BrasselHanusMueller08PADL}
+B.~Bra{\ss}el, M.~Hanus, and M.~M{\"u}ller.
+\newblock High-level database programming in {Curry}.
+\newblock In {\em Proc. of the Tenth International Symposium on Practical
+  Aspects of Declarative Languages (PADL'08)}, pages 316--332. Springer LNCS
+  4902, 2008.
+
+\bibitem{BrasselHuch07}
+B.~Bra{\ss}el and F.~Huch.
+\newblock On a tighter integration of functional and logic programming.
+\newblock In {\em Proc. APLAS 2007}, pages 122--138. Springer LNCS 4807, 2007.
+
+\bibitem{BrasselHuch09}
+B.~Bra{\ss}el and F.~Huch.
+\newblock The {Kiel} {Curry} {System} {KiCS}.
+\newblock In {\em Applications of Declarative Programming and Knowledge
+  Management}, pages 195--205. Springer LNAI 5437, 2009.
+
+\bibitem{CaballeroLopez99}
+R.~Caballero and F.J. L{\'o}pez-Fraguas.
+\newblock A functional-logic perspective of parsing.
+\newblock In {\em Proc. 4th Fuji International Symposium on Functional and
+  Logic Programming (FLOPS'99)}, pages 85--99. Springer LNCS 1722, 1999.
+
+\bibitem{Fischer05}
+S.~Fischer.
+\newblock A functional logic database library.
+\newblock In {\em Proc. of the ACM SIGPLAN 2005 Workshop on Curry and
+  Functional Logic Programming (WCFLP 2005)}, pages 54--59. ACM Press, 2005.
+
+\bibitem{FischerKiselyovShan09}
+S.~Fischer, O.~Kiselyov, and C.~Shan.
+\newblock Purely functional lazy non-deterministic programming.
+\newblock In {\em Proceeding of the 14th ACM SIGPLAN International Conference
+  on Functional Programming (ICFP 2009)}, pages 11--22. ACM, 2009.
+
+\bibitem{GonzalezEtAl99}
+J.C. Gonz{\'a}lez-Moreno, M.T. Hortal{\'a}-Gonz{\'a}lez, F.J.
+  L{\'o}pez-Fraguas, and M.~Rodr{\'\i}guez-Artalejo.
+\newblock An approach to declarative programming based on a rewriting logic.
+\newblock {\em Journal of Logic Programming}, 40:47--87, 1999.
+
+\bibitem{Hanus00PADL}
+M.~Hanus.
+\newblock A functional logic programming approach to graphical user interfaces.
+\newblock In {\em International Workshop on Practical Aspects of Declarative
+  Languages (PADL'00)}, pages 47--62. Springer LNCS 1753, 2000.
+
+\bibitem{Hanus01PADL}
+M.~Hanus.
+\newblock High-level server side web scripting in {Curry}.
+\newblock In {\em Proc.\ of the Third International Symposium on Practical
+  Aspects of Declarative Languages (PADL'01)}, pages 76--92. Springer LNCS
+  1990, 2001.
+
+\bibitem{Hanus06PPDP}
+M.~Hanus.
+\newblock Type-oriented construction of web user interfaces.
+\newblock In {\em Proceedings of the 8th ACM SIGPLAN International Conference
+  on Principles and Practice of Declarative Programming (PPDP'06)}, pages
+  27--38. ACM Press, 2006.
+
+\bibitem{Hanus07ICLP}
+M.~Hanus.
+\newblock Multi-paradigm declarative languages.
+\newblock In {\em Proceedings of the International Conference on Logic
+  Programming (ICLP 2007)}, pages 45--75. Springer LNCS 4670, 2007.
+
+\bibitem{Hanus10PAKCS}
+M.~Hanus, S.~Antoy, B.~Bra{\ss}el, M.~Engelke, K.~H{\"o}ppner, J.~Koj,
+  P.~Niederau, R.~Sadre, and F.~Steiner.
+\newblock {PAKCS}: The {P}ortland {A}achen {K}iel {C}urry {S}ystem.
+\newblock Available at \url{http://www.informatik.uni-kiel.de/~pakcs/}, 2010.
+
+\bibitem{HanusKoschnicke10PADL}
+M.~Hanus and S.~Koschnicke.
+\newblock An {ER-based} framework for declarative web programming.
+\newblock In {\em Proc. of the 12th International Symposium on Practical
+  Aspects of Declarative Languages (PADL 2010)}, pages 201--216. Springer LNCS
+  5937, 2010.
+
+\bibitem{HanusSadre99JFLP}
+M.~Hanus and R.~Sadre.
+\newblock An abstract machine for curry and its concurrent implementation in
+  java.
+\newblock {\em Journal of Functional and Logic Programming}, 1999(6), 1999.
+
+\bibitem{Hanus06Curry}
+M.~Hanus~(ed.).
+\newblock Curry: An integrated functional logic language (vers.\ 0.8.2).
+\newblock Available at \url{http://www.curry-language.org}, 2006.
+
+\bibitem{Hussmann92}
+H.~Hussmann.
+\newblock Nondeterministic algebraic specifications and nonconfluent term
+  rewriting.
+\newblock {\em Journal of Logic Programming}, 12:237--255, 1992.
+
+\bibitem{Lopez-FraguasSanchez-Hernandez99}
+F.~L\'opez-Fraguas and J.~S\'anchez-Hern\'andez.
+\newblock {TOY}: A multiparadigm declarative system.
+\newblock In {\em Proc. of RTA'99}, pages 244--247. Springer LNCS 1631, 1999.
+
+\bibitem{LopezRodriguezSanchez07}
+F.J. L{\'o}pez-Fraguas, J.~Rodr{\'\i}guez-Hortal{\'a}, and
+  J.~S{\'a}nchez-Hern{\'a}ndez.
+\newblock A simple rewrite notion for call-time choice semantics.
+\newblock In {\em Proceedings of the 9th ACM SIGPLAN International Conference
+  on Principles and Practice of Declarative Programming (PPDP'07)}, pages
+  197--208. ACM Press, 2007.
+
+\bibitem{Lux99FLOPS}
+W.~Lux.
+\newblock Implementing encapsulated search for a lazy functional logic
+  language.
+\newblock In {\em Proc. 4th Fuji International Symposium on Functional and
+  Logic Programming (FLOPS'99)}, pages 100--113. Springer LNCS 1722, 1999.
+
+\bibitem{MorenoEtAl90ALP}
+J.J. M{oreno-Navarro}, H.~Kuchen, R.~Loogen, and M.~R{odr{\'\i}guez-Artalejo}.
+\newblock Lazy narrowing in a graph machine.
+\newblock In {\em Proc. Second International Conference on Algebraic and Logic
+  Programming}, pages 298--317. Springer LNCS 463, 1990.
+
+\bibitem{Partain93}
+W.~Partain.
+\newblock The nofib benchmark suite of {H}askell programs.
+\newblock In {\em Proceedings of the 1992 Glasgow Workshop on Functional
+  Programming}, pages 195--202. Springer, 1993.
+
+\bibitem{PeytonJones03Haskell}
+S.~Peyton~Jones, editor.
+\newblock {\em Haskell 98 Language and Libraries---The Revised Report}.
+\newblock Cambridge University Press, 2003.
+
+\bibitem{Wadler85}
+P.~Wadler.
+\newblock How to replace failure by a list of successes.
+\newblock In {\em Functional Programming and Computer Architecture}, pages
+  113--128. Springer LNCS 201, 1985.
+
+\end{thebibliography}
 

### 2011-05-05 papers/wflp2011/paper.tex
Definition "keepfiles" added to rc file, WFLP paper updated

@@
 according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
-Uniform programs have been also considered in
+A variant of uniform programs have been considered in
 \cite{MorenoEtAl90ALP} to define lazy narrowing strategies
@@
 allow only a single non-variable argument in each left-hand side
-of a rule.}
+of a rule.
+Uniform programs have been also applied in
+\cite{Zartmann97SAS} to define a denotational analysis
+of functional logic programs.}
 There, the variables in the left-hand sides of each rule are pairwise
@@
 language for our translation into Haskell which will be presented
-in the following sections.
+in the following.
 
@@
 can be constructed as follows,
-where the operation \code{unsafeInterleaveIO} is used to construct
+where \code{unsafeInterleaveIO} is used to construct
 the tree on demand:
 \begin{haskell}
-  initSupply = getIDTree
+  initSupply = getIDTree$\listline$
   getIDTree = do s1 <- unsafeInterleaveIO getIDTree
@@
 
-The functional logic language Toy \cite{Lopez-FraguasSanchez-Hernandez99}
-has many similarities to Curry and the Toy system compiles Toy programs
+The functional logic language TOY \cite{Lopez-FraguasSanchez-Hernandez99}
+has many similarities to Curry and the TOY system compiles TOY programs
 into Prolog programs. However, we have not included a comparison
@@
 The first collection of benchmarks\footnote{All benchmarks are available
-at \url{http://www.informatik.uni-kiel.de/~mh/repos/kics2/benchmarks.tgz}.}
+at \url{http://www-ps.informatik.uni-kiel.de/kics2/benchmarks/}.}
 (Fig.~\ref{fig:bench-first-order})
@@
 In order to be consistent with the call-time choice semantics
-of functional logic languages like Curry or Toy,
+of functional logic languages like Curry or TOY,
 we represent non-deterministic values
@@
 
+\bibitem{Zartmann97SAS}
+F.~Zartmann.
+\newblock Denotational Abstract Interpretation of Functional Logic Programs.
+\newblock In {\em Proc. of the 4th International Symposium on Static Analysis
+  (SAS'97)}, pp. 141--156. Springer LNCS 1302, 1997.
+
 \end{thebibliography}

### 2011-05-06 papers/wflp2011/paper.tex

@@
 we could have the reduction
-\begin{lstlisting}
+\begin{haskell}
   xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                  $\to~$  xor True False      $\to~$  True
-\end{lstlisting}
+\end{haskell}
 leading to the unintended result \code{True}.
@@
 \cite{AntoyHanus06ICLP} that logic variables can be replaced
-by non-deterministic ``generators,''
+by non-deterministic ``generators'',
 i.e., operations that evaluate to all possible values of the type of
@@
 according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
-A variant of uniform programs have been considered in
+A variant of uniform programs has been considered in
 \cite{MorenoEtAl90ALP} to define lazy narrowing strategies
 for functional logic programs. Although the motivations are similar,
-our notion of uniform program is more restrictive since we
+our notion of uniform programs is more restrictive since we
 allow only a single non-variable argument in each left-hand side
 of a rule.
-Uniform programs have been also applied in
+Uniform programs have also been applied in
 \cite{Zartmann97SAS} to define a denotational analysis
@@
 also called LOIS (limited overlapping inductively sequential) system.
-Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
+Furthermore, Bra{\ss}el \cite{Brassel11Thesis} showed the semantical
 equivalence of narrowing computations in LOIS systems
@@
 %
-\begin{curry}
+\begin{haskell}
   xor True           x = xor' x
@@
   xor' (Choice x1 x2) = Choice (xor' x1) (xor' x2)
-\end{curry}
+\end{haskell}
 %
@@
 If we evaluate the expression \ccode{xorSelf aBool}, we get the result
-\begin{lstlisting}
+\begin{haskell}
   Choice (Choice False True) (Choice True False)
-\end{lstlisting}
+\end{haskell}
 How can we interpret this result?
@@
 Unfortunately, this does not conform to the call-time choice semantics
-discussed in Section~\ref{ctc-semantics} which should not yield
+discussed in Section~\ref{ctc-semantics} which excludes
 a value like \code{True}.
@@
 the identifiers will be kept, e.g.,
-\begin{curry}
+\begin{haskell}
   xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
-\end{curry}
+\end{haskell}
 If we evaluate the expression \ccode{xorSelf aBool} and
@@
 to the choice of \code{aBool}, we obtain the result
-\begin{lstlisting}
+\begin{haskell}
   Choice 1 (Choice 1 False True) (Choice 1 True False)
-\end{lstlisting}
+\end{haskell}
 When we show the values contained in this result,
@@
 and \code{2} for the right \code{aBool} argument) so that this evaluates to
-\begin{lstlisting}
+\begin{haskell}
   Choice 1 (Choice 2 False True) (Choice 2 True False)
-\end{lstlisting}
+\end{haskell}
 Here we can make different selections for the outer and inner
@@
 yields non-deterministic results for each of the argument's values.
-\item Each choice introduced in a computation gets its own unique
+\item Each choice introduced in a computation is supplied with its own unique
 identifier.
@@
 The operation \code{thisID} takes some identifier from this set, and
-\code{leftSupply} and \code{rightSupply} splits this set
+\code{leftSupply} and \code{rightSupply} split this set
 into two disjoint subsets without the identifier
@@
 other operations in the right-hand side of a rule. In this case,
-the other operations must be supplied with their individual disjoint
+the called operations must be supplied with their individual disjoint
 identifier supplies. For instance, the operation \code{main}
@@
 \end{haskell}
-There are various implementations of \code{IDSupply}.
+The type \code{IDSupply} can be implemented in various ways.
 The simplest implementation uses unbounded integers:
@@
 \end{haskell}
-Consequently, an implementation of \code{IDSupply} requires
+Consequently, the implementation of \code{IDSupply} requires
 an infinite set of memory cells which can represented as a tree structure:
@@
 
-Instead of choosing different search strategies to evaluate
+In addition to the different search strategies to evaluate
 top-level expressions, our system also contains a primitive operation
@@
 
-We have already discussed in the introduction the various
+In the introduction we already discussed the various
 efforts to implement functional logic languages,
@@
 that came from the benchmarks reported in \cite{FischerKiselyovShan09}.
-Due to these experiences, we used in our new compiler the
+Due to these experiences, in our new compiler we use the
 compilation scheme initially proposed in \cite{BrasselFischer08IFL}

### 2011-05-06 tests/MainFunctionPattern.hs
addapted tests to new structure

@@ -7,5 +7,5 @@ import Curry_FunctionPattern
 
 main =getArgs >>= test . head
 
-test "goal1" = prdfs nd_C_goal1
+test "goal1" = prdfs print nd_C_goal1
 test _       = error "unknown goal"

### 2011-05-09 Basics.hs
fixed problem with the matching of nats

@@ -289,6 +289,7 @@ class ConvertCurryHaskell ctype htype where -- needs MultiParamTypeClasses
 -- Matching for Integers
 ------------------------------------------------------------------------------
 
+-- TODO: use unboxed int
 
 matchInteger :: NonDet a => [(Int,a)] -> C_Integer -> a
 matchInteger rules (C_Neg nat)              = 
@@ -302,9 +303,11 @@ matchInteger rules (Choices_C_Integer i cs) =
 matchInteger rules Fail_C_Integer           = failCons
 matchInteger rules (Guard_C_Integer cs int) = guardCons cs (matchInteger rules int)
 
-matchNat [] _ = failCons
+matchNat []    _                    = failCons
 matchNat rules C_IHi                = maybe failCons id $ lookup 1 rules
-matchNat rules (C_O nat)            = matchNat (map halfKey $ filter (even.fst) rules) nat
+matchNat rules (C_O nat)            = matchNat (map halfKey $ filter (evenPos.fst) rules) nat
+  where
+   evenPos n = even n && (0 < n)
 matchNat rules (C_I nat)            = matchNat (map halfKey $ filter (odd.fst) rules) nat
 matchNat rules (Choice_C_Nat i l r) = narrow i (matchNat rules l) (matchNat rules r)
 matchNat rules (Choices_C_Nat i cs) = narrows i $ map (matchNat rules) cs

### 2011-05-09 papers/wflp2011/paper.tex
Paper fine-tuning

@@
 Our implementation is based on the idea to represent
-non-deterministic results as values in the data types corresponding
+non-deterministic results as values of the data types corresponding
 to the results.
@@
 We show by several benchmarks that our implementation
-can compete with or outperforms other existing implementations of Curry.
+can compete with or outperform other existing implementations of Curry.
 \end{abstract}
@@
 of some of these features by translating
-functional logic programs into logic or functional languages.
+functional logic programs into either logic or functional languages.
 For instance, if one compiles into Prolog, one can reuse
@@
 they are executed with almost the same efficiency
-of their purely functional counterparts.
+as their purely functional counterparts.
 
@@
 Note that this result cannot be obtained if we use a strict strategy
-where arguments are evaluated before the function calls.
+where arguments are evaluated before the operation calls.
 In order to avoid dependencies on the evaluation strategies
@@
 are shared so that all of them reduce either to \code{True} or to \code{False}.
-The requirements of this call-time choice semantics are the
+The requirements of the call-time choice semantics are the
 reason why it is not simply possible to use list comprehensions
@@
 we assume that each case expression is transformed into
-a new operation in order to avoid complications by translating
-nested case expressions.
+a new operation in order to avoid complications arising from the translation
+of nested case expressions.
 Thus, we assume that all programs are \emph{uniform}
@@
 we have to make \emph{consistent} selections in choices with
-same identifiers. This, if we select the left branch as the value
+same identifiers. Thus, if we select the left branch as the value
 of the outermost \code{Choice}, we also have to select the left branch
@@
 \code{Choice}, we also have to select the right branch in
-its selected argument \code{(Choice 1 True False)} which yields the only
+its selected argument \code{(Choice 1 True False)} which yields the sole
 value \code{False}.
@@
 In order to implement this behavior, we extend
-each datatype by a further constructor \code{Fail}
-and complete each function containing matching rules
+each data type by a further constructor \code{Fail}
+and complete each operation containing matching rules
 by a final rule that matches everything and returns the value \code{Fail}.
@@
 applications of Curry.
-Another mature implementation which we consider is MCC \cite{Lux99FLOPS}
+Another mature implementation we consider is MCC \cite{Lux99FLOPS}
 (Version 0.9.10) which compiles Curry into C.
@@
 Our compiler has been executed with the Glasgow Haskell Compiler
-(GHC 6.12.1, option -O2). All benchmarks were executed on a Linux machine
+(GHC 6.12.3, option -O2). All benchmarks were executed on a Linux machine
 running Debian 5.0.7 with an Intel Core 2 Duo (3.0GHz) processor.

### 2011-06-09 lib/External_Prelude.hs
catch/catchFail extended to catch all exceptions

@@ -1,6 +1,7 @@
 {-# LANGUAGE MagicHash, MultiParamTypeClasses #-}
 
-import GHC.IO.Exception (IOException (..))
+import qualified Control.Exception as C
+
 
 -- ATTENTION: Do not introduce line breaks in import declarations as these
 -- are not recognized!
@@ -707,8 +708,8 @@ external_d_C_prim_appendFile :: C_String -> C_String -> C_IO OP_Unit
 external_d_C_prim_appendFile = fromHaskellIO2 appendFile
 
 external_d_C_catchFail :: C_IO a -> C_IO a -> C_IO a
-external_d_C_catchFail act err = fromIO $ catch (toIOWithFailCheck act) handle
-  where handle ioErr = print ioErr >> (toIO err)
+external_d_C_catchFail act err = fromIO $ C.catch (toIOWithFailCheck act) handle
+  where handle e = print (e :: C.SomeException) >> (toIO err)
         toIOWithFailCheck act =
           case act of Fail_C_IO -> ioError (userError "I/O action failed")
                       _         -> toIO act
@@ -759,12 +760,12 @@ external_nd_C_apply :: NonDet b => Func a b -> a -> IDSupply -> b
 external_nd_C_apply = nd_apply
 
 external_d_C_catch :: C_IO a -> (C_IOError -> C_IO a) -> C_IO a
-external_d_C_catch act cont = fromIO $ catch (toIO act) handle where
-  handle = toIO . cont . C_IOError . toCurry . ioe_description
+external_d_C_catch act cont = fromIO $ C.catch (toIO act) handle where
+  handle e = toIO $ cont $ C_IOError $ toCurry $ show (e :: C.SomeException)
 
 external_nd_C_catch :: C_IO a -> Func C_IOError (C_IO a) -> IDSupply -> C_IO a
-external_nd_C_catch act cont s = C_IO $ catch (toIO act) handle where
-  handle e = toIO (nd_apply cont (C_IOError (toCurry (ioe_description e))) s)
+external_nd_C_catch act cont s = fromIO $ C.catch (toIO act) handle where
+  handle e = toIO $ nd_apply cont (C_IOError $ toCurry $ show (e :: C.SomeException)) s
 
 -- TODO: Support non-deterministic IO ?
 external_d_OP_gt_gt_eq :: (Curry t0, Curry t1) => C_IO t0 -> (t0 -> C_IO t1) -> C_IO t1

### 2011-06-14 papers/ciclops2011/paper.tex
CICLOPS'11 paper added

--- 2011-05-09 papers/wflp2011/paper.tex
@@
 
-\title{KiCS2: A New Compiler from Curry to Haskell}
+\title{Haskell for Logic Programmers:\\
+ Compiling Functional Logic Programs to Haskell\thanks{%
+A slightly modified version of this paper will appear in
+Proc.\ of WFLP 2011.}}
 
@@
 \begin{abstract}
-In this paper we present our first steps towards a new system
-to compile functional logic programs of the source language Curry
-into purely functional Haskell programs.
-Our implementation is based on the idea to represent
-non-deterministic results as values of the data types corresponding
-to the results.
-This enables the application of various search strategies
-to extract values from the search space.
-We show by several benchmarks that our implementation
-can compete with or outperform other existing implementations of Curry.
+In this paper we present recent results about a new system to compile
+functional logic programs of the source language Curry into purely
+functional Haskell programs. The implementation is based on the idea
+to represent non-determinism as data in Haskell. This enables the
+application of various search strategies to extract values from the
+search space. We show by several benchmarks that our implementation
+can compete with or outperforms other existing implementations of
+Curry.
 \end{abstract}
@@
 The combination of these features
-has led to new design patterns \cite{AntoyHanus02FLOPS}
+has led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
 and better abstractions for application programming,
@@
 The implementation of functional logic languages is challenging
-since a reasonable implementation has to support the operational
-features mentioned above.
+since the operational features mentioned above have to be supported.
 One possible approach
-is the design of new abstract machines appropriately supporting
+is the design of abstract machines appropriately supporting
 these operational features and implementing them in some (typically,
@@
 non-deterministic computations \cite{FischerKiselyovShan09}.
-
-Nevertheless, the translation of functional logic languages
-into other high-level languages is attractive:
-it limits the implementation efforts compared to an implementation
-from scratch and one can exploit the existing implementation technologies,
-provided that the efforts to implement the missing features
-are reasonable.
 
@@
 another constructor to represent the choice between several values.
-This idea is also the basis of the Curry implementation KiCS
-\cite{BrasselHuch07,BrasselHuch09}.
-However, KiCS is based on unsafe features of Haskell
-that inhibit the use of optimizations provided by Haskell compilers
-like GHC.\footnote{\url{http://www.haskell.org/ghc/}}
-In contrast, our implementation, called KiCS2, avoids such unsafe features.
-In addition, we also support more flexible search strategies
-and new features to encapsulate non-deterministic computations
-(which are not described in detail in this paper due to lack of space).
-
-The general objective of our approach is the support
-of flexible strategies to explore the search space
-resulting from non-deterministic computations.
-In contrast to Prolog-based implementations
-that use backtracking and, therefore, are incomplete,
-we also want to support complete strategies like breadth-first search,
+Hence, each expression evaluates into an explicit representation
+of its search space. Therefore, our implementation is not
+committed to backtracking (in contrast to Prolog-based implementations)
+but also supports complete strategies like breadth-first search,
 iterative deepening or parallel search (in order to exploit
-multi-core architectures). We achieve this goal by an
-explicit representation of the search space as data
-that can be traversed by various operations.
+multi-core architectures).
 Moreover, purely deterministic computations
@@
 The syntax of the functional logic language Curry \cite{Hanus06Curry}
-is close to Haskell \cite{PeytonJones03Haskell}.
+is close to Haskell \cite{PeytonJones03Haskell}, i.e.,
+(type) variables and function names usually
+start with lowercase letters and the names of type and data constructors
+start with an uppercase letter. The application of $f$
+to $e$ is denoted by juxtaposition (``$f~e$'').
 In addition, Curry allows free (logic)
@@
 \cite{AugustssonRittriSynek94}.
-Actually, our compilation system is parameterized over different
+Actually, our compiler is parameterized over different
 implementations of \code{IDSupply}
-in order to perform some experiments and choose
-the most appropiate for a given application.
+in order to choose the most appropiate for a given application.
 Each implementation must ensure that, if $s$ is a value of type
-\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
+\code{IDSupply} and $o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$,
+then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
 and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
 are different identifiers provided that
-$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
-and $o_1\cdots o_n \neq o'_1\cdots o'_m$.
+$o_1\cdots o_n \neq o'_1\cdots o'_m$.
 
@@
 
-Functional logic computations can also fail, e.g.,
+Similarly to logic programming,
+functional logic computations can also fail, e.g.,
 due to partially defined operations.
@@
 To provide a common interface for such operations, we introduce
-a data type to represent the general outcome of a computation,
+a data type to represent the general outcome of a potentially
+non-deterministic computation,
 \begin{haskell}
@@
 by an interactive loop where the user can request the individual values.
+Therefore, our implementation separates the logic
+(what are the values of an expression?) from the control
+(in which order should the values be computed?),
+in contrast to Prolog-based implementations,
+like PAKCS \cite{Hanus10PAKCS} or TOY \cite{Lopez-FraguasSanchez-Hernandez99},
+that are committed to a fixed search strategy (backtracking).
 
@@
 in a data structure rather than as a computation as in implementations
-based on backtracking, we can provide different methods to
+based on backtracking, we can provide various methods to
 explore the search space containing the different result values.
@@
 \begin{curry}
- searchTree :: a -> IO (SearchTree a)
+ getSearchTree :: a -> IO (SearchTree a)
 \end{curry}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions and Related Work}
+\section{Conclusions}
 \label{sec:Conclusions}
@@
 is provided in \cite{Brassel11Thesis}.
-Apart from the parser, where we reused an existing one
-implemented in Haskell, the compiler is completely written in Curry.
 
@@
 our system easily supports various search strategies
-in constrast to Prolog-based implementations.
+in contrast to Prolog-based implementations.
+Actually, our implementation can also be used as a testbed
+to explore different strategies for non-deterministic computations.
+For instance, \cite{AlqaddoumiAntoyFischerReck10} proposes
+pull-tab steps to move non-determinism from arguments to
+the result position of a function. Antoy \cite{Antoy11ICLP}
+shows that single pull-tab steps are semantics-preserving.
+Thus, it is not necessary to move each choice to the root
+of an expression, as done in our current implementation,
+but one could also perform further local computations in the arguments
+of a choice before moving it up. This might be a reasonable strategy
+if all non-deterministic values are required but many computations fail.
+However, the general effects of such refinements need further
+investigation that can be performed by slightly changing our
+compilation scheme.
+
 Since we compile Curry programs into Haskell,
@@
 Hence, purely functional parts of functional logic programs
-can be executed with almost the same efficiency as Haskell programs.
+are executed with almost the same efficiency as Haskell programs.
 Our benchmarks show that even the execution of the
@@
 implementations of Curry.
+Apart from the parser, where we reused an existing one
+implemented in Haskell, our compiler is completely written in Curry.
+The first version of the compiler was implemented
+with the Prolog-based Curry implementation PAKCS \cite{Hanus10PAKCS}.
+After bootstrapping this compiler, it became approximately 20 times
+faster. This shows that the results of the small benchmarks
+presented in this paper are also valid for larger applications.
 
@@
 
-If non-deterministic results are collected in data structures,
-one has more fine-grained control over non-deterministic steps.
-For instance, \cite{AlqaddoumiAntoyFischerReck10} proposes
-pull-tab steps to move non-determinism from arguments to
-the result position of a function. Antoy \cite{Antoy11ICLP}
-shows that single pull-tab steps are semantics-preserving.
-Thus, it is not necessary to move each choice to the root
-of an expression, as done in our implementation,
-but one could also perform further local computations in the arguments
-of a choice before moving it up. This might be a reasonable strategy
-if all non-deterministic values are required but many computations fail.
-However, the general effects of such refinements need further
-investigations.
-
 Our implementation has many opportunities for optimization, like
@@
 
+%\bibliographystyle{plain}
+%\bibliography{mh}
 \begin{thebibliography}{10}
@@
 S.~Antoy.
-\newblock On the correctness of the pull-tab transformation.
+\newblock On the correctness of pull-tabbing.
 \newblock In {\em To appear in Proceedings of the 27th International Conference
@@
 \newblock {\em Communications of the ACM}, 53(4):74--85, 2010.
+
+\bibitem{AntoyHanus11WFLP}
+S.~Antoy and M.~Hanus.
+\newblock New functional logic design patterns.
+\newblock In {\em Proc.\ of the 20th International Workshop on Functional and
+  (Constraint) Logic Programming (WFLP 2011)}. Springer LNCS (to appear), 2011.
 
@@
 F.~Zartmann.
-\newblock Denotational Abstract Interpretation of Functional Logic Programs.
+\newblock Denotational abstract interpretation of functional logic programs.
 \newblock In {\em Proc. of the 4th International Symposium on Static Analysis
-  (SAS'97)}, pp. 141--156. Springer LNCS 1302, 1997.
+  (SAS'97)}, pages 141--156. Springer LNCS 1302, 1997.
 

### 2011-06-16 talks/badhonnef2011/talk.tex
Talk for Bad Honnef moved

--- 2011-05-01 talks/talk.tex
@@
-\documentclass[utf8,handout]{beamer}
+\documentclass[utf8]{beamer}
 
@@
 
+\newenvironment{program}{\begin{semiverbatim}\small}{\end{semiverbatim}}
+
 \begin{document}
@@
 
-\author[B. Braßel, M. Hanus, B. Peemöller, F. Reck]
+\author[B. Braßel, M. Hanus, \underline{B. Peemöller}, F. Reck]
  {Bernd Braßel \and Michael Hanus \\
-   Björn Peemöller \and Fabian Reck\\
+   \underline{Björn Peemöller} \and Fabian Reck\\
   \texttt{\{bbr, mh, bjp, fre\}@informatik.uni-kiel.de}}
@@
 \frametitle{Curry}
-\begin{columns}
+\begin{columns}[t]
 \column{.45\textwidth}
@@
 \item Haskell-like syntax.
-\item Extended by non-determinism and free variables. %TODO: failure?
+\item Extended by non-determinism, free variables, constraints, unification.
 \end{itemize}
+\begin{example}<2>
+\begin{program}
+> xorSelf aBool
+False
+False
+\end{program}
+\end{example}
 \column{.45\textwidth}
-\begin{block}{Running example}
-\begin{semiverbatim}
+\begin{block}{Running Code Example}
+\begin{program}
 data Bool = True | False
@@
 xorSelf x = xor x x
-\end{semiverbatim}
+\end{program}
 \end{block}
 \end{columns}
-\end{frame}
-
-\begin{frame}[fragile]
-\frametitle{Example Call}
-\begin{semiverbatim}
-> xorSelf aBool
-False
-False
-No more solutions.
-\end{semiverbatim}
-\begin{itemize}
-\item The call of \verb!aBool! yields two solutions
-\end{itemize}
 \end{frame}
@@
 \item Compile Curry programs into Haskell
-\item Reuse expresiveness of Haskell (lazy evaluation, higher-order functions)
+\item Reuse features of Haskell (lazy evaluation, higher-order functions)
 \item Benefit from mature Haskell compiler (GHC)
@@
 \end{block}
-
+\pause
 \begin{block}{Goals}
@@
 \item Efficient execution of purely functional programs
-\item Avoid unsafe features to enable optimization (in contrast to KiCS)
+\item Avoid unsafe operations to enable optimization (in contrast to KiCS)
 \item Support for different search strategies (including parallel strategies)
@@
 \end{itemize}
-
+\pause
 \begin{block}{Curry}
-\begin{semiverbatim}
+\begin{program}
 data Bool = True | False
@@
 aBool = True ? False
-\end{semiverbatim}
+\end{program}
 \end{block}
@@
 \begin{block}{Haskell}
-\begin{semiverbatim}
-data Bool = True | False | Choice_Bool Bool Bool
+\begin{program}
+data Bool = True | False | Choice Bool Bool
 
 aBool :: Bool
-aBool = Choice_Bool True False
-\end{semiverbatim}
+aBool = Choice True False
+\end{program}
 \end{block}
@@
 \begin{frame}[fragile]
-\frametitle{Extension of the Operations}
+\frametitle{Extension of Operations}
 \begin{itemize}
@@
 \end{itemize}
-
+\pause
 \begin{block}{Haskell}
-\begin{semiverbatim}
+\begin{program}
 not True           = False
 not False          = True
-not (Choice_Bool x1 x2) = Choice_Bool (not x1) (not x2)
+not (Choice x1 x2) = Choice (not x1) (not x2)
 
@@
 xor False          x = x
-xor (Choice_Bool x1 x2) x = Choice_Bool (xor x1 x)
-                                        (xor x2 x)
+xor (Choice x1 x2) x = Choice (xor x1 x) (xor x2 x)
 
 xorSelf x = xor x x
-\end{semiverbatim}
+\end{program}
 \end{block}
@@
 \begin{example}
-\verb!not aBool!
-$\leadsto$ \verb!not (Choice_Bool True False)!
-$\leadsto$ \verb!Choice_Bool (not True) (not False)!
-$\leadsto$ \verb!Choice_Bool False True!
+\verb!not aBool!\\
+$\leadsto$ \verb!not (Choice True False)!\\
+$\leadsto$ \verb!Choice (not True) (not False)!\\
+$\leadsto$ \verb!Choice False True!
 \end{example}
-
+\pause
 \begin{example}
-\verb!xorSelf aBool!
-$\leadsto$ \verb!xor aBool aBool!
-$\leadsto$ \verb!xor (Choice_Bool True False) aBool!
-$\leadsto$ \verb!Choice_Bool (xor True aBool) (xor False aBool)!
-$\leadsto$ \verb!Choice_Bool (not aBool) aBool!
-$\leadsto$ \verb!Choice_Bool (Choice_Bool False True) (Choice_Bool True False)!
+\verb!xorSelf aBool!\\
+$\leadsto$ \verb!xor aBool aBool!\\
+$\leadsto$ \verb!xor (Choice True False) aBool!\\
+$\leadsto$ \verb!Choice (xor True aBool) (xor False aBool)!\\
+$\leadsto$ \verb!Choice (not aBool) aBool!\\
+$\leadsto$ \verb!Choice (Choice False True) (Choice True False)!
 \end{example}
-
 \begin{itemize}
-  \item Yiels four results: \verb!False, True, True, False!
-  \item Only \verb!False, False! are reasonable.
+  \item Yields two results \verb!True! and \verb!False!
+  \item Only \verb!False! is correct w.r.t. the semantics of Curry.
 \end{itemize}
-
 \end{frame}
@@
 \begin{frame}[fragile]
-\frametitle{Run-Time Choice vs. Call-Time Choice}
+\frametitle{Call-Time Choice vs. Run-Time Choice}
 \begin{block}{Curry}
-\begin{lstlisting}
+\begin{program}
 xorSelf x = xor x x
-\end{lstlisting}
+\end{program}
+\end{block}
+\pause
+\begin{block}{Call-Time Choice}
+\begin{itemize}
+\item Variables denote values
+\item Both occurrences of {\tt x} are substituted with the same value in each non-deterministic branch
+\end{itemize}
 \end{block}
@@
 \item Variables denote non-deterministic computations
-\item {\tt x}'s may evaluate to different values in the same non-deterministic branch
-\end{itemize}
-\end{block}
-\pause
-\begin{block}{Call-Time Choice}
-\begin{itemize}
-\item Variables denote values
-\item Both {\tt x} evaluate to the same value in each non-deterministic branch
+\item Substitutions of {\tt x} may evaluate to different values in the same non-deterministic branch
 \end{itemize}
@@
 \end{itemize}
-
+\pause
 \begin{block}{Choice Identifier}
-\begin{semiverbatim}
+\begin{program}
 type ID = Integer
 
-data Bool = False | True | Choice_Bool ID Bool Bool
+data Bool = False | True | Choice ID Bool Bool
 
 aBool :: ID -> Bool
-aBool i = Choice_Bool i True False
+aBool i = Choice i True False
 
-xor (Choice_Bool i x1 x2) x = Choice_Bool i (xor x1 x) 
-                                            (xor x2 x)
-\end{semiverbatim}
+xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
+\end{program}
 \end{block}
@@
 \begin{example}
-\verb!xorSelf (aBool 1)!
-$\leadsto$ \verb!xor (aBool 1) (aBool 1)!
-$\leadsto$ \verb!xor (Choice_Bool 1 True False) (Choice_Bool 1 True False)!
-$\leadsto$ $\dots$
-$\leadsto$ \verb!Choice_Bool 1 (Choice_Bool 1 False True) (Choice_Bool 1 True False)!
+\verb!xorSelf (aBool 1)!\\
+$\leadsto$ \verb!xor (aBool 1) (aBool 1)!\\
+$\leadsto$ \verb!xor (Choice 1 True False) (Choice 1 True False)!\\
+$\leadsto$ $\dots$\\
+$\leadsto$ \verb!Choice 1 (Choice 1 False True) (Choice 1 True False)!
 \end{example}
-
 \begin{itemize}
-\item The results \verb!True, True! are identified as invalid
+\item The paths to \verb!True! are identified as invalid
 \end{itemize}
@@
 
-\begin{frame}[fragile]
-\frametitle{Representing Failure}
-\begin{itemize}
- \item Computations with failures is a common programming pattern in FLP
- \item Failures do not abort the computation, but are ignored
- \item Thus, failures have to be represented as well.
-\end{itemize}
-
-\begin{block}{Curry}
-\begin{semiverbatim}
-data Bool = True | False
-
-ensureTrue True = True
-\end{semiverbatim}
-\end{block}
-
-\begin{block}{Haskell}
-\begin{semiverbatim}
-data Bool = True | False | Choice_Bool Bool Bool | Fail_Bool
-
-ensureTrue True                  = True
-ensureTrue (Choice_Bool i x1 x2) = ...
-ensureTrue _                     = Fail_Bool
-\end{semiverbatim}
-\end{block}
-\end{frame}
+% \begin{frame}[fragile]
+% \frametitle{Representing Failure}
+% \begin{itemize}
+%  \item Computing with failure is a common programming pattern in FLP
+%  \item Failures do not abort the computation, but are ignored
+%  \item Thus, failures have to be represented as well.
+% \end{itemize}
+% \begin{block}{Curry}
+% \begin{program}
+% data Bool = True | False
+% 
+% ensureTrue True = True
+% \end{program}
+% \end{block}
+% \begin{block}{Haskell}
+% \begin{program}
+% data Bool = True | False | Choice Bool Bool | Fail
+% 
+% ensureTrue True             = True
+% ensureTrue (Choice i x1 x2) = ...
+% ensureTrue _                = Fail
+% \end{program}
+% \end{block}
+% \end{frame}
 
 \begin{frame}[fragile]
-\frametitle{Higher Order Functions}
+\frametitle{Higher-Order Functions}
 \begin{itemize}
@@
 \begin{block}{Curry}
-\begin{semiverbatim}
+\begin{program}
 apply :: (a -> b) -> a -> b
@@
 False
-\end{semiverbatim}
+\end{program}
 \end{block}
 \column{.45\textwidth}
-\begin{block}{Haskell}
-\begin{semiverbatim}
+\begin{block}{Haskell (first approach)}
+\begin{program}
 apply :: (a -> b) -> a -> b
@@
 False
-\end{semiverbatim}
+\end{program}
 \end{block}
@@
 \begin{frame}[fragile]
-\frametitle{Higher Order Functions}
+\frametitle{Higher-Order Functions (cont.)}
 \begin{itemize}
@@
 \end{itemize}
-\begin{columns}
+\begin{columns}[t]
 \column{.45\textwidth}
 \begin{block}{Curry}
-\begin{semiverbatim}
-apply :: (a -> b) -> a -> b
+\begin{program}
+apply :: (a -> b) 
+      -> a -> b
 apply f x = f x
 
-ndNot x = x ? not x
+to be = be ? not (to be)
 
-> apply ndNot True
+> apply to True
 True
 False
-\end{semiverbatim}
+...
+\end{program}
 \end{block}
-\column{.45\textwidth}
+\column{.5\textwidth}
 \begin{block}{Haskell}
-\begin{semiverbatim}
+\begin{program}
 apply :: (a -> ID -> b) 
@@
 
-ndNot x i = 
-  Choice_Bool i x (not x)
+to be i = 
+  Choice il be (not (to be ir))
+  where il = leftID  i
+        ir = rightID i
 
-> apply ndNot True 1
-Choice_Bool 1 True False
-\end{semiverbatim}
+> apply to True 1
+Choice 2 True (
+Choice 6 False (
+Choice 14 ...
+\end{program}
 \end{block}
@@
 \begin{frame}[fragile]
-\frametitle{Higher-order optimization}
-Deterministic functions have to be wrapped
+\frametitle{Higher-Order Optimization}
+
+\begin{itemize}
+\item Deterministic functions have to be wrapped
+\item This causes an unnecessary overhead
+\end{itemize}
+
+\begin{columns}[t]
+\column{.45\textwidth}
 \begin{block}{Curry}
-\begin{semiverbatim}
+\begin{program}
 > apply not True
 False
-\end{semiverbatim}
+\end{program}
 \end{block}
+\column{.45\textwidth}
 \begin{block}{Haskell}
-\begin{semiverbatim}
+\begin{program}
 > apply (\\x _ -> not x) True 1
 False
-\end{semiverbatim}
+\end{program}
 \end{block}
-\begin{block}{Optimization}
+\end{columns}
+\vspace{3ex}
 \begin{itemize}
-\item Higher-order functions are translated into
+\item Thus, higher-order functions are translated into
   \begin{itemize}
@@
 \item Benchmarks show enormously improved performance
+% \item Non-deterministic choice between functions is also possible.
 \end{itemize}
-\end{block}
-Non-deterministic choice between functions is also possible.
 \end{frame}
@@
 \begin{frame}
-\frametitle{Benchmarks}
+\frametitle{Benchmark: Reverse of a List}
+\begin{itemize}
+\item Functional programs run with almost no overhead
+\end{itemize}
+\begin{center}
+\includegraphics[width=9cm]{reverse}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Benchmark: Reverse using foldr}
+\begin{itemize}
+\item Higher-order optimization dramatically improves performance
+\end{itemize}
+\begin{center}
+\includegraphics[width=9cm]{reverseho}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Benchmark: Non-Deterministic Permutation Sort}
+\begin{itemize}
+\item For non-deterministic programs, KiCS2 can compete with existing compilers
+\end{itemize}
+\begin{center}
+\includegraphics[width=9cm]{permsort}
+\end{center}
+\end{frame}
+
+\begin{frame}[fragile]
+\frametitle{Benchmark: Halve a Peano Number}
+\begin{itemize}
+\item However, there is still much room for improvements
+\end{itemize}
+\begin{center}
+\includegraphics[width=9cm]{half}
+\begin{program}
+        half y | equal (add x x) y = x where x free
+\end{program}
+\end{center}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Additional Features}
+\frametitle{Status quo}
 \begin{itemize}
-
-\item Free Variables
-\item Unification
+\item Additional Features (work in progress)
+\begin{itemize}
+\item Free variables and unification
 \item Function Patterns
-\item Encapsulated Search
-
+\item Sharing over non-determinism
+\item Different search strategies (depth-first, breadth-first, iterative deepening, parallel)
+\item Encapsulated search
+\end{itemize}
+\item Future work
+\begin{itemize}
+\item More detailed analysis to decrease overhead of logic features
+\item Set Functions
+\item First release (coming soon)
+\end{itemize}
 \end{itemize}
@@
 \begin{itemize}
-\item A transformation of a functional logic program into a functional program.
-\item The resulting program can be optimized by the Haskell compiler
-\item A formal description of our approach can be found in our paper.
+\item Transformation of a functional logic program into a functional program.
+\item Resulting program can be optimized by the Haskell compiler.
+\item Our Approach can compete with or outperform existing implementations.
+\item A formal description of our approach can be found in our paper (WFLP 2011).
 \end{itemize}
@@
 \begin{center}
-
 {\huge Thank You!}\\[3ex]
@@
 {\huge Questions?}
-
 \end{center}

### 2011-06-16 src/bootstrap.sh
modyfied bootstrap process; added TODOs

@@ -1,28 +1,0 @@
-#!/bin/sh
-LOGFILE=bootstrap.log
-
-# Bootstrap KiCS2 compiler
-
-# make everything clean:
-make clean
-
-# start logging
-echo "Bootstrapping started at `date`" > $LOGFILE
-
-# Create kics2 via PAKCS (stage 1)
-echo "Compiling stage 1" >> $LOGFILE
-make Compile.state 2>&1 | tee -a $LOGFILE
-
-# Create kics2 via kics2 (stage 2)
-echo "Compiling stage 2" >> $LOGFILE
-make Compile 2>&1 | tee -a $LOGFILE
-
-# Create REPL via KiCS2, stage 2
-echo "Compiling REPL" >> $LOGFILE
-make REPL 2>&1 | tee -a $LOGFILE
-
-echo "Bootstrapping finished at  `date`" >> $LOGFILE
-
-# Now, kics2 should work with the bootstrapped compiler, i.e., you can
-# install the complete system:
-make

### 2011-06-16 src/mkPrimTypes.sh
Clean up

@@ -106,4 +106,4 @@ function replace ()
 }
 
 rm -f $DEST
-./idc -i./lib $SRC && replace $DEST
+../bin/idc -i./lib $SRC && replace $DEST

### 2011-06-28 runtime/Search.hs
resolved TODO in prdfs

@@ -157,7 +157,7 @@ printValsDFS' fb cont (Guard cs e) = solves cs >>= traverse fb
       mapM_ (>>= traverse True) cs
       reset
     traverse False (ChoicesST _     cs) = do
-      mapM_ (>>= traverse True) cs -- TODO: do not reset at last alternative
+      mapMButLast_ (>>= traverse True) (>>= traverse False) cs
 
 --     traverse Nothing = return ()
 --     traverse (Just reset) =  if fb then (printValsDFS fb . try) $!< e >> reset
@@ -180,6 +180,14 @@ zipWithButLast3 _ lastf (a:[]) (b:_ ) (c:_)  = lastf a b c : []
 zipWithButLast3 _ lastf (a:_ ) (b:[]) (c:_)  = lastf a b c : []
 zipWithButLast3 _ lastf (a:_ ) (b:_ ) (c:[]) = lastf a b c : []
 zipWithButLast3 f lastf (a:as) (b:bs) (c:cs) = f a b c : zipWithButLast3 f lastf as bs cs
+
+mapMButLast_ :: (a -> IO b) -> (a -> IO b) -> [a] -> IO ()
+mapMButLast_ f lastf = sequence_ . mapButLast f lastf
+
+mapButLast :: (a -> b) -> (a -> b) -> [a] -> [b]
+mapButLast _ _     []     = []
+mapButLast _ lastf [x]    = [lastf x]
+mapButLast f lastf (x:xs) = f x : mapButLast f lastf xs
 
 -- Attempt to gain more abstraction during search
 

### 2011-06-28 papers/wlp2011/paper.tex
WLP paper added

\documentclass{llncs}

\usepackage{times}
\usepackage{url}
\def\UrlFont{\tt}
\usepackage{latexsym}
\usepackage{xspace}
\usepackage{hyperref}
\usepackage{color}

\usepackage{listings}
\lstset{aboveskip=1.0ex,
        belowskip=1.0ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        basewidth=0.52em,
        basicstyle=\small\ttfamily}

\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
       }

\lstnewenvironment{curry}{\lstset{backgroundcolor=\color[rgb]{0.9,0.9,0.9}}}{}
\lstnewenvironment{haskell}{}{}
\newcommand{\listline}{\vrule width0pt depth1.75ex}

\newcommand{\code}[1]{\mbox{\small\texttt{#1}}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92\xspace} % backslash
\newcommand{\us}{\char95\xspace} % underscore
\newcommand{\dollar}{\code{\char36}} % dollar in programs
\newcommand{\Choice}[2]{#1 \mathop{?} #2}
\newcommand{\funset}{\ensuremath{_{\cal S}}}

\begin{document}
\pagestyle{plain}
\sloppy

\title{Implementing Equational Constraints\\ in a Functional Language}

\author{
Bernd Bra{\ss}el
\kern1em
Michael Hanus
\kern1em
Bj{\"o}rn Peem{\"o}ller
\kern1em
Fabian Reck
}
\institute{
Institut f\"ur Informatik, CAU Kiel, D-24098 Kiel, Germany \\
\email{\{bbr|mh|bjp|fre\}@informatik.uni-kiel.de}
}

\maketitle

\begin{abstract}
KiCS2 is a new system
to compile functional logic programs of the source language Curry
into purely functional Haskell programs.
The implementation is based on the idea to represent
the search space as a data structure and logic variables
as operations that generate their values.
This has the advantage that one can apply various,
in particular, complete search strategies to compute solutions.
However, the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
To overcome this drawback, we propose new techniques
to implement equational constraints in this framework.
In particular, we show how unification modulo function evaluation
and functional patterns can be added without sacrificing
the efficiency of the kernel implementation.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:Introduction}

Functional logic languages combine the most important
features of functional and logic programming in a single language
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they provide higher-order functions and demand-driven
evaluation from functional programming together with logic programming features
like non-deterministic search and computing with partial information
(logic variables).
This combination
has led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
and better abstractions for application programming,
but it put new demands on the implementation side.

Previous implementations of functional logic languages
can be classified into three categories:
(1) designing new abstract machines appropriately supporting
these operational features and implementing them in some (typically,
imperative) language, like C \cite{Lux99FLOPS}
or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP},
(2) compilation into logic languages (Prolog) and reuse
the existing backtracking implementation for non-deterministic
search as well as logic variables and unification for computing with partial
information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}, or
(3) compilation into non-strict functional languages like Haskell
and reuse the implementation
of lazy evaluation and higher-order functions
\cite{BrasselHuch07,BrasselHuch09}.
The latter approach requires the implementation
of non-deterministic evaluations but has the advantage
that the explicit handling of non-determinism provides for
various search strategies like depth-first, breadth-first, parallel,
or iterative deepening instead of committing to a fixed (incomplete)
strategy like backtracking \cite{BrasselHuch07}.

In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselFischerHanusReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
Actually, it has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same results.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation in functional logic languages).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvement
without sacrificing the efficiency of the kernel implementation.

In the next section, we review the source language Curry
and the features that are considered in this paper.
Section~\ref{sec:Compilation} sketches the implementation scheme
of KiCS2.
Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
discuss the extensions to implement unification modulo
functional evaluation and functional patterns, respectively.
Benchmarks demonstrating the usefulness of this scheme
are presented in Section~\ref{sec:Benchmarks}
before we conclude in Section~\ref{sec:Conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Curry Programs}
\label{sec:Curry}

The syntax of the functional logic language Curry \cite{Hanus06Curry}
is close to Haskell \cite{PeytonJones03Haskell},
i.e., type variables and names of defined operations usually
start with lowercase letters and the names of type and data constructors
start with an uppercase letter. The application of $f$
to $e$ is denoted by juxtaposition (``$f~e$'').
In addition to Haskell, Curry allows free (logic)
variables in conditions and right-hand sides of defining rules.
Hence, an operation is defined by conditional rewrite rules of the form:
%
\begin{equation}
\label{rule}
f~t_1 \ldots t_n \code{~|~} c \code{~=~} e \code{~~where~} vs \code{~free}
\end{equation}
where the \emph{condition} $c$ is optional and
$vs$ is the list of variables occurring in $c$ or $e$ but not in the
\emph{left-hand side} $f~t_1 \ldots t_n$.

In contrast to functional programming and similarly to logic programming,
operations can be defined by overlapping rules so that
they might yield more than one result on the same input.
Such operations are also called \emph{non-deterministic}.
For instance, Curry offers a \emph{choice} operation that is predefined by
the following rules:
%
\begin{curry}
  x ? _ = x
  _ ? y = y
\end{curry}
%
Thus, we can define a non-deterministic operation \code{aBool} by
\label{ex:aBool}
%
\begin{curry}
  aBool = True ? False
\end{curry}
%
so that the expression \ccode{aBool} has two values:
\code{True} and \code{False}.

If non-deterministic operations are used as arguments in other operations,
a semantical ambiguity might occur. Consider the operations
%
\begin{curry}
  xor True  False = True
  xor True  True  = False
  xor False x     = x

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

The condition $c$ in rule (\ref{rule}) is typically a
conjunction of \emph{equational constraints}
of the form \code{$e_1$\,=:=\,$e_2$}.
Such a constraint is satisfiable if
both sides $e_1$ and $e_2$ are evaluable to unifiable data terms.
For instance, if the symbol ``\code{++}'' denotes the usual list
concatenation operation, we can define an operation \code{last}
that computes the last element \code{e} of a non-empty list \code{xs}
as follows:
%
\begin{curry}
  last xs | ys++[e]=:=xs = e   where ys,e free  
\end{curry}
%
As in Haskell, the rules defining most functions are
\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule}) $t_1
\ldots t_n$ are made of variables and/or data constructor symbols
only.  However, Curry also allows \emph{functional pattern}
\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
defined operations.  For instance, we can define \code{last} also
as:
%
\begin{curry}
  last (xs++[e]) = e 
\end{curry}
%
Here, the functional pattern \code{(xs++[e])}
states that \code{last} is reducible
to \code{e} provided that the argument can be matched against
some value of \code{(xs++[e])} where \code{xs} and \code{e} are
free variables.
By instantiating \code{xs} to arbitrary lists, the value
of \code{(xs++[e])} is any list having \code{e} as its last element.
Functional patterns are
a powerful feature to express arbitrary selections in term structures.
For instance, functional patterns supports a straightforward
processing of XML data with incompletely specified
or evolving formats \cite{Hanus11ICLP}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Compilation of Scheme of KiCS2}
\label{sec:Compilation}

In this section we sketch the translation of Curry programs
into Haskell programs as performed by KiCS2,
since this is necessary to understand the extensions
described in the subsequent sections.
More details about this translation scheme can be found in
\cite{BrasselFischer08IFL,BrasselFischerHanusReck11}.

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
The first argument of each \code{Choice} constructor of type \code{ID}
is used to implement the call-time choice semantics
discussed in Section~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, are either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice}. The difficulty is to get a unique identifier
when needed, i.e., when some operation evaluates to a \code{Choice}.
Since we want to compile into \code{purely} functional programs
(in order to enable powerful program optimizations),
we cannot use unsafe features with side effects to generate
such identifiers.
Hence, we pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick an identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available \cite{AugustssonRittriSynek94}
(see below for a simple implementation) and our implementation
is parametric over concrete implementations of \code{IDSupply}.

When translating Curry to Haskell, KiCS2 adds to each operation
an additional argument of type \code{IDSupply}.
For instance, the operation \code{aBool}
defined in Section~\ref{ex:aBool} is translated into:
\begin{haskell}
  aBool :: IDSupply -> Bool
  aBool s = Choice (thisID s) True False
\end{haskell}
Similarly, the operation
\begin{curry}
  main :: Bool
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main :: IDSupply -> Bool
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
so that the set \code{s} is split into a set \code{(leftSupply s)}
containing identifier for the evaluation of \code{aBool}
and a set \code{(rightSupply s)} containing identifiers
for the evaluation of operation \code{xorSelf}.

Since all data types are extended by additional constructors,
we must also extend the definition of operations performing
pattern matching. For instance, consider the definition of lists
\begin{curry}
  data List a = Nil | Cons a (List a)
\end{curry}
and an operation to extract the first element of a non-empty list:
\begin{curry}
  head :: List a -> a
  head (Cons x xs) = x
\end{curry}
The type definition is extended as follows:
\begin{haskell}
  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
\end{haskell}
The operation \code{head} is extended by an identifier supply
and further matching rules:
\begin{haskell}
  head :: List a -> IDSupply -> a
  head (Cons x xs)      s = x
  head (Choice i x1 x2) s = Choice i (head x1 s) (head x2 s)
  head _                s = Fail
\end{haskell}
The second rule transforms a non-deterministic argument
into a non-deterministic result and
the final rule returns \code{Fail} in all other cases,
i.e., if \code{head} is
applied to the empty list as well as if the matching argument
is already a failed computation (failure propagation).

To show a concrete example, we use the following implementation
of \code{IDSupply} based on unbounded integers:
%
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
\end{haskell}
%
If we apply the same transformation to the rules defining \code{xor}
(actually, \code{xor} performs pattern matching on two arguments
which is avoided in KiCS2 by introducing an intermediate operation,
see \cite{BrasselFischerHanusReck11}) and evaluate the main expression
\code{(main 1)},
we obtain the result
\begin{haskell}
  Choice 2 (Choice 2 False True) (Choice 2 True False)
\end{haskell}
Thus, the result is non-deterministic and contains several choices,
but all of them have the same identifier.
This identifier is used when the values contained in this result
are shown: one has to make \emph{consistent} selections in choices with
same identifiers. Thus, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 2 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 2 True False)} which yields the only
value \code{False}.
Thus, unintended value \code{True} is not produced.

In order to extract values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right) choices.
The latter requirement can be implemented by storing
the choices already made during the traversal.
For this purpose, we use the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we assume operations to lookup the current choice
for a given identifier or change its choice (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
  printValsDFS v                = print v
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

    newChoice ch x = do setChoice i ch
                        printValsDFS (try x)
                        setChoice i NoChoice
\end{haskell}
This operation prints ignores failures and prints values
that are not a \code{Choice}. For a \code{Choice},
it checks whether a choice for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoChoice}).
If a choice has been made, it follows this choice.
Otherwise, the left choice is made and stored. After printing
all the values w.r.t.\ this choice, the choice is undone (like in backtracking)
and the right choice is made and stored.

In general, this operation is applied to the normal form
of the main expression (where \code{initSupply} is used to
compute an initial identifier supply passed to this expression).
The normal form computation is necessary for structured data like lists
so that a failure or choice in some part of the data is moved to
the root. Other search strategies, like
breadth-first search, iterative deepening, or parallel search,
can be obtained by different implementations of this main operation
to print all values. Instead of printing all values,
one can also collect all values in a tree-like data structure
so that the programmer can implement his own search strategies
(this corresponds to encapsulating search \cite{BrasselHanusHuch04JFLP}).
Finally, one can also have other options rather than printing all values
of the main expression.
For instance, one can easily define operations
to prints only the first solution or one by one upon user requests.
Due to the lazy evaluation strategy of Haskell,
such operations can also be applied to infinite choice structures.

An important optimization, performed by our compiler,
is a \emph{determinism analysis}.
If an operation does not call, neither directly nor indirectly through
other operations, the choice operation \ccode{?},
then it is not necessary to pass an identifier supply
so that the generated code is nearly identical to a corresponding
functional program.
Actually, the benchmarks presented in \cite{BrasselFischerHanusReck11}
show that this implementation outperforms all other Curry implementations
for deterministic operations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

As mentioned in the introduction,
our compiler translates occurrences of logic variables
into generators. For instance, each Boolean logic variable
is replaced by a generator defined like \code{aBool}.
Since such generators are standard
non-deterministic operations, they are translated like any other operation.
However, if equational constraints should be solved,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Unification}
\label{sec:Unification}

Motivate: discuss == vs. =:=

infinite search space for variable unification

distinguish free variables from choices

extend unification

add binding constraints

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Functional Patterns}
\label{sec:FuncPatterns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Benchmarks}
\label{sec:Benchmarks}

Benchmarks for Last with ==, =:= and func. patterns

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Related Work}
\label{sec:Conclusions}

\bibliographystyle{plain}
\bibliography{mh}

\end{document}

% LocalWords:  Curry KiCS

### 2011-06-30 papers/wlp2011/paper.tex

@@
 \newcommand{\Choice}[2]{#1 \mathop{?} #2}
-\newcommand{\funset}{\ensuremath{_{\cal S}}}
+\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
 
@@
 %
+\label{ex:last}
 \begin{curry}
@@
 we have to traverse it, compute all possible choices
-but consider the choice identifiers to make consistent (left/right) choices.
+but consider the choice identifiers to make consistent (left/right)
+decisions.
 The latter requirement can be implemented by storing
-the choices already made during the traversal.
+the decision already made for some choices during the traversal.
 For this purpose, we use the type
 \begin{haskell}
-  data Choice = NoChoice | ChooseLeft | ChooseRight
+  data Decision = NoDecision | ChooseLeft | ChooseRight
 \end{haskell}
-where \code{NoChoice} represents the fact that a choice has not yet been made.
-Furthermore, we assume operations to lookup the current choice
-for a given identifier or change its choice (depending on the implementation
+where \code{NoDecision} represents the fact that the value of a choice
+is yet undecided.
+Furthermore, we assume operations to lookup the current decision
+for a given identifier or change its decision (depending on the implementation
 of \code{IDSupply}, KiCS2 supports various implementations
@@
 \begin{haskell}
-  lookupChoice :: ID -> IO Choice
-  setChoice    :: ID -> Choice -> IO ()
+  lookupChoice :: ID -> IO Decision
+  setChoice    :: ID -> Decision -> IO ()
 \end{haskell}
@@
     choose ChooseRight = printValsDFS (try x2)
-    choose NoChoice    = do newChoice ChooseLeft  x1
-                            newChoice ChooseRight x2
+    choose NoDecision  = do newDecision ChooseLeft  x1
+                            newDecision ChooseRight x2
 
-    newChoice ch x = do setChoice i ch
+    newDecision ch x = do setChoice i ch
                           printValsDFS (try x)
-                        setChoice i NoChoice
+                          setChoice i NoDecision
 \end{haskell}
@@
 that are not a \code{Choice}. For a \code{Choice},
-it checks whether a choice for
+it checks whether a decision for
 this identifier has already been made (note that the initial value
-for all identifiers is \code{NoChoice}).
-If a choice has been made, it follows this choice.
-Otherwise, the left choice is made and stored. After printing
-all the values w.r.t.\ this choice, the choice is undone (like in backtracking)
-and the right choice is made and stored.
+for all identifiers is \code{NoDecision}).
+If a decision has been made for this choice, it follows this decision.
+Otherwise, the left alternative is used and this decision is stored.
+After printing
+all the values w.r.t.\ this decision,
+the decision is undone (like in backtracking)
+and the right alternative is used and stored.
 
@@
 As mentioned in the introduction,
-our compiler translates occurrences of logic variables
-into generators. For instance, each Boolean logic variable
-is replaced by a generator defined like \code{aBool}.
+KiCS2 translates occurrences of logic variables
+into generators. For instance, consider the operation \code{not}
+defined by the rules:
+\begin{curry}
+  not True  = False
+  not False = True
+\end{curry}
+Then the expression \ccode{not x}, where \code{x} is a logic variable,
+is translated into \ccode{not aBool} (we omit the \code{IDSupply} arguments
+here). The latter expression is evaluated by reducing the
+argument \code{aBool} to a choice between \code{True} or \code{False}
+followed by applying \code{not} to this choice.
+This is similar to a narrowing step on \ccode{not x}
+that instantiates the variable \code{x} to \code{True} or \code{False}.
+Actually, it has been shown in \cite{AntoyHanus06ICLP}
+that narrowing expressions with logic variables
+is equivalent to rewriting with generators.
 Since such generators are standard
-non-deterministic operations, they are translated like any other operation.
+non-deterministic operations, they are translated like any other operation
+so that one does not need additional run-time support for them.
 However, if equational constraints should be solved,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Unification}
+\section{Equational Constraints and Unification}
 \label{sec:Unification}
 
-Motivate: discuss == vs. =:=
+As known from logic programming, predicates or constraints
+are important to restrict the set of intended values
+in a non-deterministic computation.
+Apart from user-defined predicates,
+equational constraints of the form \code{$e_1$\,=:=\,$e_2$}
+are the most important kind of constraints.
+We have already seen in the first rule defining operation
+\code{last} in Section~\ref{ex:last} a typical application
+of an equational constraint.
 
-infinite search space for variable unification
+Due to the presence of non-terminating operations
+and infinite data structures,
+\ccode{=:=} is interpreted as the \emph{strict equality} on terms
+\cite{GiovannettiLeviMoisoPalamidessi91},
+i.e., the equation \code{$e_1$\,=:=\,$e_2$} is satisfied iff
+$e_1$ and $e_2$ are reducible to the same ground constructor term.
+In particular, expressions that do not have a value
+are not equal w.r.t.\ \ccode{=:=}, e.g.,
+the equational constraint \ccode{head [] =:= head []}
+is not satisfiable.\footnote{From now on, we use the
+standard notation for lists, i.e., \code{[]} denotes the empty list
+and \code{(x:xs)} denotes a list with head element \code{x}
+and tail \code{xs}.}
 
-distinguish free variables from choices
+Due to this constructive definition, \ccode{=:=}
+can be considered as a binary function defined by the
+following rules (here we show only the rules for the Boolean
+and list types, where \code{Success} denotes the only constructor
+of the type \code{Success} of constraints):
+\begin{curry}
+  True   =:= True    =  Success
+  False  =:= False   =  Success
+  []     =:= []      =  Success
+  (x:xs) =:= (y:ys)  =  x =:= y & xs =:= ys
 
-extend unification
+  Success & c  =  c
+\end{curry}
+%
+If we translate these operations into Haskell by the scheme
+shown in Section~\ref{sec:Compilation},
+the following rules are added to these rules
+in order to propagate choices and failures:
+\begin{haskell}
+  Choice i l r =:= y             =  Choice i (l =:= y) (r =:= y)
+  x            =:= Choice i l r  =  Choice i (x =:= l) (x =:= r)
+  _            =:= _             =  Fail
 
-add binding constraints
+  Choice i l r & c  =  Choice i (l & c) (r & c)
+  _            & _  =  Fail
+
+\end{haskell}
+%
+Although this is a correct implementation of equational constraints,
+it might lead to an unnecessarily large search space
+when it is applied to generators representing logic variables.
+For instance, consider the following generator for Boolean lists:
+\begin{curry}
+  aBList = [] ? (aBool : aBList)
+\end{curry}
+This is translated into Haskell as follows:
+\begin{haskell}
+  aBList :: IDSupply -> [Bool]
+  aBList s = Choice (thisID s) [] (aBool (leftSupply s)
+                                   : aBList (rightSupply s))
+\end{haskell}
+Now consider the equational constraint \ccode{x =:= [True]}.
+If the logic variable \code{x} is replaced by \code{aBList},
+the translated expression \ccode{aBList s =:= [True]}
+creates a search space when evaluating its first argument,
+although there is no search required since there is only one possible
+binding for \code{x}.
+Furthermore and even worse, there is an infinite search space
+when unifying two logic variables. For instance,
+the expression \ccode{xs=:=ys \& xs++ys=:=[True]}
+results in an infinite search space when the logic variables
+\code{xs} and \code{ys} are replaced by generators.
+
+To avoid these problems, we have to implement the idea
+of the well-known unification principle \cite{Robinson65}.
+Instead of enumerating all values for logic variables
+occurring in an equational constraint,
+we \emph{bind} them to another variable or term.
+Since we compile into a purely functional language,
+the binding cannot be performed by some side effect.
+Instead, we add binding constraints to the computed results.
+
+To implement this idea, we have to distinguish free variables
+from ``standard choices'' (corresponding to occurrences of \ccode{?})
+in the target code. For this purpose, we refine the definition
+of the type \code{ID} as follows:\footnote{For the sake
+of simplicity, we consider in the following the
+implementation of \code{IDSupply} as unbounded integers.}
+\begin{haskell}
+  data ID = ID Integer | FreeID Integer
+\end{haskell}
+The new constructor \code{FreeID} marks a choice corresponding to
+a free variable, e.g., the generator for Boolean variables is
+redefined as
+\begin{haskell}
+  aBool s = Choice (FreeID (thisID s)) True False
+\end{haskell}
+%
+If an operation is applied to a free variable and requires its value,
+the free variable is transformed into a standard choice.
+For this purpose, we define a simple operation to perform
+this transformation:
+\begin{haskell}
+  narrow :: ID -> ID
+  narrow (FreeID i) = ID i
+  narrow x          = x
+\end{haskell}
+Furthermore, we use this operation in narrowing steps,
+i.e., in all rules operating on \code{Choice} constructors.
+For instance, in the implementation of the operation \code{not}
+we replace the rule
+\begin{haskell}
+  not (Choice i x1 x2) s = Choice i (not x1 s) (not x2 s)
+\end{haskell}
+by the rule
+\begin{haskell}
+  not (Choice i x1 x2) s = Choice (narrow i) (not x1 s) (not x2 s)
+\end{haskell}
+With this small change, the run-time system can distinguish
+free variables (choices with an identifier of the form \code{(FreeID$\ldots$)})
+from other values. This is quite useful at the top-level
+to print values containing logic variables in a specific form
+rather than enumerating all values, similarly to logic programming systems.
+For instance, KiCS2 evaluates the application of \code{head}
+to an unknown list as follows:
+\begin{haskell}
+  Prelude> head xs where xs free
+  {xs = (_x2:_x3)} _x2
+\end{haskell}
+Thus, free variables are prefixed by \code{\us{}x}.
+
+As mentioned above, the consideration of free variables
+is relevant in equational constraints where binding constraints are
+generated. For this purpose, we introduce a type to represent
+a \emph{binding constraints} as pair of a choice identifier
+and a decision for this choice:
+\begin{haskell}
+  data Constraint = ID :=: Decision
+\end{haskell}
+Furthermore, we extend each data type by the possibility to
+add some constraints:
+\begin{haskell}
+  data Bool   = $\ldots$ | Guard [Constraint] Bool
+  data List a = $\ldots$ | Guard [Constraint] (List a)
+\end{haskell}
+Thus, \code{(Guard $cs$ $v$)} represents a \code{constrained value},
+i.e., the value $v$ is only valid if the constraints $cs$ are
+consistent. The consistency is checked when a value is extracted
+from a choice structure, e.g., by the operation \code{printValsDFS},
+whereas the binding constraints are created by
+equational constraints: if a free variable
+should be bound to a constructor, we create the same
+decision as it would be done by the (successful) branch
+of the generator. In case of the Boolean values,
+this can be implemented by the following additional rules
+for \ccode{=:=}:
+\begin{haskell}
+  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
+  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
+\end{haskell}
+Hence, the binding of a variable to some known value
+is implemented as a binding constraint for the choice identifier
+for this variable. However, if we want to bind a variable
+to another variable, we cannot store a concrete decision.
+Instead, we store the information that the decisions for
+both variables, when they are made to extract values,
+must be identical. For this purpose we extend the decision
+type to cover this information:
+\begin{haskell}
+  data Decision = $\ldots$ | BindTo ID
+\end{haskell}
+Furthermore, we add the rule that an equational constraint
+between two variables yields a binding for these variables:
+\begin{haskell}
+  Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
+    =  Guard [i :=: BindTo j] Success
+\end{haskell}
+%
+As mentioned above, the checking of constraints
+for consistency is done when values are extracted.
+For this purpose, we extend the definition of the corresponding
+operations by calling a constraint solver for the constraints.
+For instance, the definition of \code{printValsDFS} is extended by
+a rule handling constrained values:
+%
+\begin{haskell}
+  $\ldots$
+  printValsDFS (Guard constraints x) = solve constraints (printValsDFS x)
+  $\ldots$
+\end{haskell}
+The operation \code{solve} checks the consistency of the constraints
+given in the first arguments,
+stores further decisions made by the constraints, e.g.,
+if variables are bound together,
+and evaluates its second argument.
+Furthermore, the operations \code{lookupChoice} and \code{setChoice}
+are extended to deal with decisions of type \code{BindTo},
+i.e., they follow chains of variable bindings.
+
+Due to the extensions described in this section,
+we obtain the implementation of a unification procedure
+that integrates the demand-driven evaluation of operations
+occurring in the arguments.
+Benchmarks evaluating the efficiency of this approach are
+shown in Section~\ref{sec:Benchmarks}.
+
+\todo{Mention accumulation of constraints in \&?}
 
@@
 
+also short comparison of =:= with PAKCS and MCC
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \label{sec:Conclusions}
+
+Future work: adding other constraint systems
 
@@
 
-% LocalWords:  Curry KiCS
+% LocalWords:  Curry KiCS equational

### 2011-06-30 runtime/Solver.hs
fixed bug with LazyBind in Narrowed ConstraintChoices

@@ -62,11 +62,12 @@ solve (ConstraintChoice i lcs rcs) = lookupChoice i >>= chooseCC
       (mkSolution (setUnsetChoice i ChooseRight) >>> solves rcs)
     chooseCC c           = error $ "Solver.solve.chooseCC: " ++ show c
 
-solve (ConstraintChoices i@(Narrowed pns _) css) = lookupChoice i >>= chooseCCs
+solve cc@(ConstraintChoices i@(Narrowed pns _) css) = lookupChoice i >>= chooseCCs
   where
     chooseCCs (ChooseN c _) = solves (css !! c)
     chooseCCs NoChoice      = return $
       ChoicesST (return ()) $ zipWith3 mkChoice [0 ..] css pns
+    chooseCCs (LazyBind cs) = solves cs >>> solve cc 
     chooseCCs c           = error $ "Solver.solve.chooseCCs: " ++ show c
 
     mkChoice n cs pn = mkSolution (setUnsetChoice i (ChooseN n pn)) >>> solves cs

### 2011-06-30 papers/wlp2011/paper.tex
Functional pattern section started

@@
 As in Haskell, the rules defining most functions are
-\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule}) $t_1
-\ldots t_n$ are made of variables and/or data constructor symbols
+\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule})
+$t_1,\ldots,t_n$ are made of variables and/or data constructor symbols
 only.  However, Curry also allows \emph{functional pattern}
 \cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
-defined operations.  For instance, we can define \code{last} also
+defined operations.  For instance, we can define the last element
+of a list also by
 as:
@@
 \begin{curry}
-  last (xs++[e]) = e 
+  last' (xs++[e]) = e 
 \end{curry}
@@
 Here, the functional pattern \code{(xs++[e])}
-states that \code{last} is reducible
+states that \code{last'} is reducible
 to \code{e} provided that the argument can be matched against
@@
 are the most important kind of constraints.
-We have already seen in the first rule defining operation
+We have already seen in the operation
 \code{last} in Section~\ref{ex:last} a typical application
@@
 Due to the extensions described in this section,
-we obtain the implementation of a unification procedure
+we obtain an implementation of a unification procedure
 that integrates the demand-driven evaluation of operations
@@
 \label{sec:FuncPatterns}
+
+A well-known disadvantage of equational constraints
+is the fact that \ccode{=:=} is interpreted as strict equality.
+Thus, if one uses equational constraints to express
+requirements on arguments, the resulting operations might be too strict.
+For instance, the equational constraint in the condition
+defining \code{last} (see Section~\ref{ex:last})
+requires that \code{ys++[e]} as well as \code{xs}
+must be reducible to identical values so that the input list \code{xs}
+is completely evaluated.
+Hence, if \code{failed} denotes an operation whose evaluation fails,
+the evaluation of \code{last [failed,2]} has no result.
+On the other hand, the evaluation of \code{last' [failed,2]} yields
+the value \code{2}, i.e., the definition of \code{last'} is less strict.
+This is due to the fact that a functional pattern like
+\code{(xs++[e])} abbreviates all values to which it can be evaluated
+(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
+Conceptually, the rule defining \code{last'}
+abbreviates the following (infinite) set of rules:
+\begin{curry}
+  last' [e] = e
+  last' [x1,e] = e
+  last' [x1,x2,e] = e
+  $\ldots$
+\end{curry}
+As a further example,
+consider an operation that returns the first duplicated element in
+a list.
+Using equational constraint, we can define it as follows:
+\begin{curry}
+  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
+            = e    where ys,zs,e free
+\end{curry}
+The first equational constraint is used to split the input list \code{xs}
+into three sublists.
+The last constraint ensures that the first sublist
+\code{ys} does not contain duplicated elements
+(the library operation removes all duplicates from a list)
+and the second constraint ensures that the first element after \code{ys}
+occurs in \code{ys}.
+Although this implementation is concise, it cannot be applied
+to infinite lists due to the strict interpretation of \ccode{=:=}.
+This is not the case if we define this operation by a functional pattern:
+\begin{curry}
+  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
+                        = e
+\end{curry}
+%
+Functional patterns are quite useful in situations
+where one wants to select parts of some structure
+at arbitrary positions. For instance,
+functional patterns have been applied to support
+the high-level programming of querying or transforming
+XML data \cite{Hanus11ICLP}.
+
+Obviously, one cannot implement functional patterns by
+a transformation into an infinite set of rules.
+Fortunately, functional patterns can be implemented by
+a specific \emph{lazy unification} procedure \ccode{=:<=}
+\cite{AntoyHanus05LOPSTR}.
+For instance, the definition of \code{last'} is transformed into
+\begin{curry}
+  last' ys | (xs++[e]) =:<= ys
+           = e   where xs,e free
+\end{curry}
+The behavior of \ccode{=:<=} is similar to \ccode{=:=}
+except for the case that a variable in the left argument
+should be bound to some expression: instead of evaluating
+the expression to some value and bind the variable to it,
+the variable is bound to the \emph{unevaluated} expression
+(see \cite{AntoyHanus05LOPSTR} for more details).
+Due to this slight change, failures or infinite structures
+in actual arguments do not cause problems in the matching
+of functional patterns.
+
+This discussion shows the requirements on an implementation
+of functional patterns. The general structure
+is quite similar to equational constraints
+with the exception that variables could be also bound
+to unevaluated expressions. If such variables are later
+accessed, they are evaluated.
+This can be achieved by adding a further alternative
+to the type of decisions:
+\begin{haskell}
+data Decision = $\ldots$ | LazyBind [Constraint]
+\end{haskell}
+%
+The implementation of the lazy unification \ccode{=:<=}
+is almost identical to the strict unification \ccode{=:=}
 

### 2011-06-30 papers/wlp2011/paper.tex
Added Benchmark table
Section 5: first version finished

@@
 Hence, if \code{failed} denotes an operation whose evaluation fails,
-the evaluation of \code{last [failed,2]} has no result.
-On the other hand, the evaluation of \code{last' [failed,2]} yields
-the value \code{2}, i.e., the definition of \code{last'} is less strict.
+the evaluation of \code{last [failed,True]} has no result.
+On the other hand, the evaluation of \code{last' [failed,True]} yields
+the value \code{True}, i.e., the definition of \code{last'} is less strict.
 This is due to the fact that a functional pattern like
@@
 \begin{curry}
-  last' ys | (xs++[e]) =:<= ys
-           = e   where xs,e free
+  last' ys | (xs++[e]) =:<= ys  = e   where xs,e free
 \end{curry}
@@
 %
-The implementation of the lazy unification \ccode{=:<=}
-is almost identical to the strict unification \ccode{=:=}
+The implementation of the lazy unification operation \ccode{=:<=}
+is almost identical to the strict unification operation \ccode{=:=}
+as shown in Section~\ref{sec:Unification}.
+The only change is in the rules where a free variable occurs
+in the left argument. All these rules are replaced by the single
+rule
+\begin{haskell}
+  Choice (FreeID i) _ _ =:<= x
+    = Guard [i :=: LazyBind (lazyBind i x)] Success
+\end{haskell}
+where the auxiliary operation \code{lazyBind}
+implements the demand-driven evaluation of the right argument \code{x}:
+\begin{haskell}
+  lazyBind :: ID -> a -> [Constraint]
+  lazyBind i True  = [i :=: ChooseLeft ]
+  lazyBind i False = [i :=: ChooseRight]
+\end{haskell}
+This has the effect that the argument \code{x} is stored
+without evaluation (due to the lazy evaluation strategy of the
+target language Haskell) as a binding constraint.
+However, it is evaluated by \code{lazyBind} when its binding
+is required by another part of the computation.
+Similarly to equational constraints,
+lazy bindings are processed by a solver when values are extracted.
+In particular, if a variable has more than one lazy binding constraint
+(which is possible if a functional pattern evaluates to a non-linear term),
+the corresponding expressions are evaluated and unified
+according to the semantics of functional patterns
+\cite{AntoyHanus05LOPSTR}.
+
+In order to demonstrate the operational behavior of our implementation,
+we sketch the evaluation of the lazy unification constraint
+\code{xs++[e]=:<=[failed,True]} that occurs when the expression
+\code{last' [failed,True]} is evaluated (here we omit failed branches
+and some other details; note that logic variables are replaced
+by generators):
+\begin{haskell}
+  aBList 2 ++ [aBool 3] =:<= [failed,True]
+  $\leadsto$ [aBool 4, aBool 3] =:<= [failed, True]
+  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
+  $\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed)
+           , 3 :=: LazyBind (lazyBind 3 True)] Success
+\end{haskell}
+If the binding of identifier \code{3} is later required,
+\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
+which corresponds to the value \code{True} of the generator \code{aBool}.
+Note that the variable with identifier \code{4} does not occur
+anywhere else so that the binding \code{(lazyBind 4 failed)}
+will never be evaluated, as intended.
 

### 2011-06-30 papers/wlp2011/paper.tex
WLP-Paper: Section 1-4 revised

--- 2011-06-30 papers/wlp2011/paper.tex
@@
 This has the advantage that one can apply various,
-in particular, complete search strategies to compute solutions.
+and in particular, complete search strategies to compute solutions.
 However, the generation of all values for logic variables
 might be inefficient for applications that
-exploit constraints on partial values.
+exploit constraints on partially known values.
 To overcome this drawback, we propose new techniques
@@
 has led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
-and better abstractions for application programming,
-but it put new demands on the implementation side.
+and better abstractions for application programming, but it also gave rise
+to new challenges of implementation.
 
@@
 can be classified into three categories:
-(1) designing new abstract machines appropriately supporting
+\begin{enumerate}
+ \item designing new abstract machines appropriately supporting
 these operational features and implementing them in some (typically,
@@
 or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP},
-(2) compilation into logic languages (Prolog) and reuse
+
+ \item compilation into logic languages (Prolog) and reusing
 the existing backtracking implementation for non-deterministic
@@
 information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}, or
-(3) compilation into non-strict functional languages like Haskell
-and reuse the implementation
+
+ \item compilation into non-strict functional languages like Haskell
+and reusing the implementation
 of lazy evaluation and higher-order functions
 \cite{BrasselHuch07,BrasselHuch09}.
+\end{enumerate}
+
 The latter approach requires the implementation
 of non-deterministic evaluations but has the advantage
-that the explicit handling of non-determinism provides for
+that the explicit handling of non-determinism allows for
 various search strategies like depth-first, breadth-first, parallel,
@@
 all possible ground values of the type of the logic variable.
-Actually, it has been shown \cite{AntoyHanus06ICLP}
+It has been shown \cite{AntoyHanus06ICLP}
 that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
@@
 can be implemented in this framework.
-We show that both extensions lead to efficiency improvement
+We show that both extensions lead to efficiency improvements
 without sacrificing the efficiency of the kernel implementation.
@@
 Note that this result cannot be obtained if we use a strict strategy
-where arguments are evaluated before the function calls.
+where arguments are evaluated prior to the function calls.
 In order to avoid dependencies on the evaluation strategies
@@
 lazily, provided that all occurrences of \code{aBool}
-are shared so that all of them reduce either to \code{True} or to \code{False}.
-The requirements of this call-time choice semantics are the
+are shared so that all of them reduce either to \code{True} or to \code{False}
+consistently. The requirements of this call-time choice semantics are the
 reason why it is not simply possible to use list comprehensions
@@
 Such a constraint is satisfiable if
-both sides $e_1$ and $e_2$ are evaluable to unifiable data terms.
+both sides $e_1$ and $e_2$ are reducible to unifiable data terms.
 For instance, if the symbol ``\code{++}'' denotes the usual list
@@
 \begin{curry}
-  last xs | ys++[e]=:=xs = e   where ys,e free  
+  last xs | ys++[e] =:= xs = e   where ys, e free
 \end{curry}
 %
-As in Haskell, the rules defining most functions are
-\emph{constructor-based} \cite{ODonnell85}, in (\ref{rule})
-$t_1,\ldots,t_n$ are made of variables and/or data constructor symbols
-only.  However, Curry also allows \emph{functional pattern}
-\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might contain calls to
+Like in Haskell, most rules defining functions are \emph{constructor-based} 
+\cite{ODonnell85}, i.e., in (\ref{rule})
+$t_1,\ldots,t_n$ consist of variables and/or data constructor symbols
+only.  However, Curry also allows \emph{functional patterns}
+\cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
 defined operations. For instance, we can define the last element
-of a list also by
-as:
+of a list also by:
 %
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Compilation of Scheme of KiCS2}
+\section{The Compilation Scheme of KiCS2}
 \label{sec:Compilation}
@@
 that both duplicates, which later evaluate to a non-deterministic
-choice between two values, are either \code{True} or \code{False}.
+choice between two values, yield either \code{True} or \code{False}.
 This is obtained by assigning a unique identifier (of type \code{ID})
@@
 when needed, i.e., when some operation evaluates to a \code{Choice}.
-Since we want to compile into \code{purely} functional programs
+Since we want to compile into \emph{purely} functional programs
 (in order to enable powerful program optimizations),
@@
 so that the set \code{s} is split into a set \code{(leftSupply s)}
-containing identifier for the evaluation of \code{aBool}
+containing identifiers for the evaluation of \code{aBool}
 and a set \code{(rightSupply s)} containing identifiers
-for the evaluation of operation \code{xorSelf}.
+for the evaluation of the operation \code{xorSelf}.
 
@@
 \end{curry}
-The type definition is extended as follows:
+The type definition is then extended as follows:
 \begin{haskell}
@@
   thisID      n = n
-  leftSupply  n = 2*n
-  rightSupply n = 2*n+1
+  leftSupply  n = 2 * n
+  rightSupply n = 2 * n + 1
 \end{haskell}
@@
 This identifier is used when the values contained in this result
-are shown: one has to make \emph{consistent} selections in choices with
+are searched for: one has to make \emph{consistent} selections in choices with
 same identifiers. Thus, if we select the left branch as the value
@@
 value \code{False}.
-Thus, unintended value \code{True} is not produced.
+Thus, the unintended value \code{True} is not produced.
 
-In order to extract values from a \code{Choice} structure,
+In general, to extract all values from a \code{Choice} structure,
 we have to traverse it, compute all possible choices
@@
 The latter requirement can be implemented by storing
-the decision already made for some choices during the traversal.
-For this purpose, we use the type
+the decisions already made for some choices during the traversal.
+For this purpose, we introduce the type
 \begin{haskell}
@@
 \begin{haskell}
-  lookupChoice :: ID -> IO Decision
-  setChoice    :: ID -> Decision -> IO ()
+  lookupDecision :: ID -> IO Decision
+  setDecision    :: ID -> Decision -> IO ()
 \end{haskell}
@@
   printValsDFS Fail             = return ()
-  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
+  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
   printValsDFS v                = print v
    where
-    choose ChooseLeft  = printValsDFS (try x1)
-    choose ChooseRight = printValsDFS (try x2)
-    choose NoDecision  = do newDecision ChooseLeft  x1
+    follow ChooseLeft  = printValsDFS x1
+    follow ChooseRight = printValsDFS x2
+    follow NoDecision  = do newDecision ChooseLeft  x1
                             newDecision ChooseRight x2
 
-    newDecision ch x = do setChoice i ch
-                          printValsDFS (try x)
-                          setChoice i NoDecision
+    newDecision d x = do setDecision i d
+                         printValsDFS x
+                         setDecision i NoDecision
 \end{haskell}
-This operation prints ignores failures and prints values
-that are not a \code{Choice}. For a \code{Choice},
-it checks whether a decision for
+This operation ignores failures and prints values that are no \code{Choice}.
+For a \code{Choice}, it checks whether a decision for
 this identifier has already been made (note that the initial value
@@
 For instance, one can easily define operations
-to prints only the first solution or one by one upon user requests.
+to print either the first solution only or one by one upon user request.
 Due to the lazy evaluation strategy of Haskell,
@@
 
-An important optimization, performed by our compiler,
-is a \emph{determinism analysis}.
-If an operation does not call, neither directly nor indirectly through
-other operations, the choice operation \ccode{?},
-then it is not necessary to pass an identifier supply
-so that the generated code is nearly identical to a corresponding
-functional program.
+To avoid an unnecessary growth of the search space represented by
+\code{Choice} constructors, out compiler performs an optimization for
+deterministic operations. If an operation does not call, 
+neither directly nor indirectly through
+other operations, the choice operation \ccode{?},\footnote{%
+Note that before applying this compilation scheme, overlapping rules in
+user-defined operations are transformed into calls of the choice operation
+\ccode{?}.}
+then it is not necessary to pass an identifier supply to the operation.
+In consequence, only the matching rules are extended so that the generated code
+is nearly identical to a corresponding functional program.
 Actually, the benchmarks presented in \cite{BrasselFischerHanusReck11}
@@
 Then the expression \ccode{not x}, where \code{x} is a logic variable,
-is translated into \ccode{not aBool} (we omit the \code{IDSupply} arguments
+is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
 here). The latter expression is evaluated by reducing the
@@
 that instantiates the variable \code{x} to \code{True} or \code{False}.
-Actually, it has been shown in \cite{AntoyHanus06ICLP}
-that narrowing expressions with logic variables
-is equivalent to rewriting with generators.
+% Actually, it has been shown in \cite{AntoyHanus06ICLP}
+% that narrowing expressions with logic variables
+% is equivalent to rewriting with generators.
 Since such generators are standard
@@
 are the most important kind of constraints.
-We have already seen in the operation
-\code{last} in Section~\ref{ex:last} a typical application
-of an equational constraint.
+We have already seen a typical application
+of an equational constraint in the operation
+\code{last} in Section~\ref{ex:last}.
 
@@
   False  =:= False   =  Success
+
   []     =:= []      =  Success
@@
 \begin{haskell}
+  Fail         =:= _             =  Fail
+  _            =:= Fail          =  Fail
   Choice i l r =:= y             =  Choice i (l =:= y) (r =:= y)
@@
 
+  Fail         & _     =  Fail
+  _            & Fail  =  Fail
   Choice i l r & c     =  Choice i (l & c) (r & c)
@@
 \begin{curry}
-  aBList = [] ? (aBool : aBList)
+  aBoolList = [] ? (aBool : aBoolList)
 \end{curry}
@@
 \begin{haskell}
-  aBList :: IDSupply -> [Bool]
-  aBList s = Choice (thisID s) [] (aBool (leftSupply s)
-                                   : aBList (rightSupply s))
+  aBoolList :: IDSupply -> [Bool]
+  aBoolList s = Choice (thisID s) [] (aBool (leftSupply s)
+                                      : aBoolList (rightSupply s))
 \end{haskell}
@@
 If the logic variable \code{x} is replaced by \code{aBList},
-the translated expression \ccode{aBList s =:= [True]}
+the translated expression \ccode{aBoolList s =:= [True]}
 creates a search space when evaluating its first argument,
@@
 when unifying two logic variables. For instance,
-the expression \ccode{xs=:=ys \& xs++ys=:=[True]}
+the expression \ccode{xs =:= ys \& xs++ys =:= [True]}
 results in an infinite search space when the logic variables
@@
 
-To implement this idea, we have to distinguish free variables
-from ``standard choices'' (corresponding to occurrences of \ccode{?})
+To implement unification, we have to distinguish free variables
+from ``standard choices'' (introduced by the operation \ccode{?})
 in the target code. For this purpose, we refine the definition
@@
 \begin{haskell}
-  data ID = ID Integer | FreeID Integer
+  data ID = ChoiceID Integer | FreeID Integer
 \end{haskell}
@@
   narrow :: ID -> ID
-  narrow (FreeID i) = ID i
+  narrow (FreeID i) = ChoiceID i
   narrow x          = x
@@
 With this small change, the run-time system can distinguish
-free variables (choices with an identifier of the form \code{(FreeID$\ldots$)})
+free variables (choices with an identifier of the form \code{(FreeID $\ldots$)})
 from other values. This is quite useful at the top-level
@@
 As mentioned above, the consideration of free variables
-is relevant in equational constraints where binding constraints are
+is relevant in equational constraints where \emph{binding constraints} are
 generated. For this purpose, we introduce a type to represent
-a \emph{binding constraints} as pair of a choice identifier
+a binding constraint as a pair of a choice identifier
 and a decision for this choice:
@@
 Furthermore, we extend each data type by the possibility to
-add some constraints:
+add such a constraint:
 \begin{haskell}
-  data Bool   = $\ldots$ | Guard [Constraint] Bool
-  data List a = $\ldots$ | Guard [Constraint] (List a)
+  data Bool   = $\ldots$ | Guard Constraint Bool
+  data List a = $\ldots$ | Guard Constraint (List a)
 \end{haskell}
-Thus, \code{(Guard $cs$ $v$)} represents a \code{constrained value},
-i.e., the value $v$ is only valid if the constraints $cs$ are
-consistent. The consistency is checked when a value is extracted
-from a choice structure, e.g., by the operation \code{printValsDFS},
-whereas the binding constraints are created by
-equational constraints: if a free variable
-should be bound to a constructor, we create the same
-decision as it would be done by the (successful) branch
-of the generator. In case of the Boolean values,
+Thus, \code{(Guard $c$ $v$)} represents a \emph{constrained value},
+i.e., the value $v$ is only valid if the constraint $c$ is
+consistent with the decisions previously made during search.
+The binding constraints are created by the
+equational constraint operation \ccode{=:=}: if a free variable
+should be bound to a constructor, we make the same
+decision as it would be done in the successful branch
+of the generator. In case of Boolean values,
 this can be implemented by the following additional rules
@@
 \begin{haskell}
-  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
-  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
+  Choice (FreeID i) _ _ =:= True   =  Guard (i :=: ChooseLeft ) Success
+  Choice (FreeID i) _ _ =:= False  =  Guard (i :=: ChooseRight) Success
 \end{haskell}
@@
 both variables, when they are made to extract values,
-must be identical. For this purpose we extend the decision
+must be identical. For this purpose, we extend the \code{Decision}
 type to cover this information:
@@
   Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
-    =  Guard [i :=: BindTo j] Success
+    =  Guard (i :=: BindTo j) Success
 \end{haskell}
 %
-As mentioned above, the checking of constraints
-for consistency is done when values are extracted.
-For this purpose, we extend the definition of the corresponding
+
+The consistency of constraints is checked when values are extracted
+from a choice structure, e.g., by the operation \code{printValsDFS}.
+For this purpose, we extend the definition of the corresponding search
 operations by calling a constraint solver for the constraints.
@@
   $\ldots$
-  printValsDFS (Guard constraints x) = solve constraints (printValsDFS x)
+  printValsDFS (Guard c x) = do consistent <- add c
+                                if consistent then do printValsDFS x
+                                                      remove c
+                                              else return ()
   $\ldots$
 \end{haskell}
-The operation \code{solve} checks the consistency of the constraints
-given in the first arguments,
-stores further decisions made by the constraints, e.g.,
-if variables are bound together,
-and evaluates its second argument.
-Furthermore, the operations \code{lookupChoice} and \code{setChoice}
-are extended to deal with decisions of type \code{BindTo},
-i.e., they follow chains of variable bindings.
+The operation \code{add} checks the consistency of the constraint \code{c}
+with the decisions made so far and stores the decision made by the constraint
+if consistency is sustained. In this case, the constrained value is evaluated
+before the contstraint is removed to allow backtracking.
+Furthermore, the operations \code{lookupDecision} and \code{setDecision}
+are extended to deal with bindings between two variables,
+i.e., they follow variable chains in case of \code{BindTo} constructors.
 
@@
 shown in Section~\ref{sec:Benchmarks}.
-
-\todo{Mention accumulation of constraints in \&?}
 

### 2011-06-30 papers/wlp2011/paper.tex
Benchmark chapter extended

--- 2011-06-30 papers/wlp2011/paper.tex
@@
 Hence, if \code{failed} denotes an operation whose evaluation fails,
-the evaluation of \code{last [failed,True]} has no result.
-On the other hand, the evaluation of \code{last' [failed,True]} yields
-the value \code{True}, i.e., the definition of \code{last'} is less strict.
+the evaluation of \code{last [failed,2]} has no result.
+On the other hand, the evaluation of \code{last' [failed,2]} yields
+the value \code{2}, i.e., the definition of \code{last'} is less strict.
 This is due to the fact that a functional pattern like
@@
 \begin{curry}
-  last' ys | (xs++[e]) =:<= ys  = e   where xs,e free
+  last' ys | (xs++[e]) =:<= ys
+           = e   where xs,e free
 \end{curry}
@@
 %
-The implementation of the lazy unification operation \ccode{=:<=}
-is almost identical to the strict unification operation \ccode{=:=}
-as shown in Section~\ref{sec:Unification}.
-The only change is in the rules where a free variable occurs
-in the left argument. All these rules are replaced by the single
-rule
-\begin{haskell}
-  Choice (FreeID i) _ _ =:<= x
-    = Guard [i :=: LazyBind (lazyBind i x)] Success
-\end{haskell}
-where the auxiliary operation \code{lazyBind}
-implements the demand-driven evaluation of the right argument \code{x}:
-\begin{haskell}
-  lazyBind :: ID -> a -> [Constraint]
-  lazyBind i True  = [i :=: ChooseLeft ]
-  lazyBind i False = [i :=: ChooseRight]
-\end{haskell}
-This has the effect that the argument \code{x} is stored
-without evaluation (due to the lazy evaluation strategy of the
-target language Haskell) as a binding constraint.
-However, it is evaluated by \code{lazyBind} when its binding
-is required by another part of the computation.
-Similarly to equational constraints,
-lazy bindings are processed by a solver when values are extracted.
-In particular, if a variable has more than one lazy binding constraint
-(which is possible if a functional pattern evaluates to a non-linear term),
-the corresponding expressions are evaluated and unified
-according to the semantics of functional patterns
-\cite{AntoyHanus05LOPSTR}.
-
-In order to demonstrate the operational behavior of our implementation,
-we sketch the evaluation of the lazy unification constraint
-\code{xs++[e]=:<=[failed,True]} that occurs when the expression
-\code{last' [failed,True]} is evaluated (here we omit failed branches
-and some other details; note that logic variables are replaced
-by generators):
-\begin{haskell}
-  aBList 2 ++ [aBool 3] =:<= [failed,True]
-  $\leadsto$ [aBool 4, aBool 3] =:<= [failed, True]
-  $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
-  $\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed)
-           , 3 :=: LazyBind (lazyBind 3 True)] Success
-\end{haskell}
-If the binding of identifier \code{3} is later required,
-\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
-which corresponds to the value \code{True} of the generator \code{aBool}.
-Note that the variable with identifier \code{4} does not occur
-anywhere else so that the binding \code{(lazyBind 4 failed)}
-will never be evaluated, as intended.
+The implementation of the lazy unification \ccode{=:<=}
+is almost identical to the strict unification \ccode{=:=}
 
@@
 
-Benchmarks for Last with ==, =:= and func. patterns
+Unlike the Curry implementations PAKCS and MCC our approach does
+not manipulate a global state to unify a logic variable. Nevertheless,
+our benchmarks show that it can compete or even outperform these 
+implementations.
+The programs used for the benchmarks are primarily the same as in \cite{AntoyHanus05LOPSTR}.
 
-also short comparison of =:= with PAKCS and MCC
+\begin{tabular}{l l}
+\bf{last}& Computes the last element of a list using strict equality\\
+\bf{simplify}& Simplification of symbolic arithmetic expressions\\
+\bf{varInExp}& Nondeterministically returns a variable occuring
+                in a symbolic arithmetic expression \\
+\bf{half}& Computes the half of a peano number using logic variables\\
+\end{tabular}
+
+The table below shows a comparison of the curry systems where
+strict equality (\code{=:=}) is used for unification:
+\footnote{\ccode{inc x n} is a naive addition that adds 1 \code{n} times
+          to its argument x }
+
+\begin{tabular}{|l|c|c|c|}
+\hline
+Expression: & KiCS2 & PAKCS & MCC \\
+\code{last (map (inc 0) [1..100000])} & &  & \\
+\code{simplify}                       &&&\\
+\code{varsInExp} & &  &\\
+\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
+\hline
+\end{tabular}
+
+All of our examples can also be defined with function patterns, i.e.,
+the lazy equality operator \code{=:<=} is used. The sole other
+Curry system featuring function patterns is PAKCS:
+
+\begin{tabular}{|l|c|c|c|}
+\hline
+Expression: & KiCS2 & PAKCS \\
+\code{last (map (inc 0) [1..100000])} & &  & \\
+\code{simplify}                       &&&\\
+\code{varsInExp} & &  &\\
+\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
+\hline
+\end{tabular}
+
+In Section \ref{sec:Unification} we stated that equational
+constraits can be implemented using ordinary pattern matching.
+We also mentioned that such an implementation would lead to
+a large search space that has to be traversed to extract
+the results of a computation. To show the benificial effects
+of our equality constraint implementations we compare them to
+the results using the equality operator \code{==} that
+produces such a search space:
+
+\begin{tabular}{|l|c|c|c|}
+\hline
+Expression: &\code{==} & \code{=:=} & \code{=:<=}\\
+\hline 
+\code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
+\code{simplify}                       &10.34&10.33&7.27\\
+\code{varsInExp} & 0.49& 84.92 & 0.09\\
+\code{toPeano 5000 == halfS (toPeano 10000)} & 23.3 & 1.81 & no Fun Pat\\
+\hline
+\end{tabular}
 

### 2011-06-30 papers/wlp2011/paper.tex

@@
   fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
-            = e    where ys,zs,e free
+            = e    where ys, zs, e free
 \end{curry}
@@
 \code{ys} does not contain duplicated elements
-(the library operation removes all duplicates from a list)
+(the library operation \code{nub} removes all duplicates from a list)
 and the second constraint ensures that the first element after \code{ys}
@@
 Obviously, one cannot implement functional patterns by
-a transformation into an infinite set of rules.
-Fortunately, functional patterns can be implemented by
-a specific \emph{lazy unification} procedure \ccode{=:<=}
+a transformation into an infinite set of rules. Instead, they are
+implemented by a specific \emph{lazy unification} procedure \ccode{=:<=}
 \cite{AntoyHanus05LOPSTR}.
@@
 \begin{curry}
-  last' ys | (xs++[e]) =:<= ys  = e   where xs,e free
+  last' ys | (xs++[e]) =:<= ys  = e   where xs, e free
 \end{curry}
@@
 should be bound to some expression: instead of evaluating
-the expression to some value and bind the variable to it,
+the expression to some value and binding the variable to the value,
 the variable is bound to the \emph{unevaluated} expression
@@
 
-This discussion shows the requirements on an implementation
-of functional patterns. The general structure
+The following discussion shows the requirements on an 
+implementation of functional patterns.
+The general structure
 is quite similar to equational constraints
 with the exception that variables could be also bound
-to unevaluated expressions. If such variables are later
-accessed, they are evaluated.
+to unevaluated expressions. Only if such variables are later
+accessed, the expressions they are bound to are evaluated.
 This can be achieved by adding a further alternative
@@
 \begin{haskell}
-  data Decision = $\ldots$ | LazyBind [Constraint]
+  data Decision = $\ldots$ | LazyBind Constraint
 \end{haskell}
@@
 as shown in Section~\ref{sec:Unification}.
-The only change is in the rules where a free variable occurs
+The only difference is in the rules where a free variable occurs
 in the left argument. All these rules are replaced by the single
@@
   Choice (FreeID i) _ _ =:<= x
-    = Guard [i :=: LazyBind (lazyBind i x)] Success
+    = Guard (i :=: LazyBind (lazyBind i x)) Success
 \end{haskell}
@@
 \begin{haskell}
-  lazyBind :: ID -> a -> [Constraint]
-  lazyBind i True  = [i :=: ChooseLeft ]
-  lazyBind i False = [i :=: ChooseRight]
+  lazyBind :: ID -> a -> Constraint
+  lazyBind i True  = i :=: ChooseLeft
+  lazyBind i False = i :=: ChooseRight
 \end{haskell}
-This has the effect that the argument \code{x} is stored
-without evaluation (due to the lazy evaluation strategy of the
-target language Haskell) as a binding constraint.
+The use of the additional \code{LazyBind} constructor allows the argument 
+\code{x} to be stored in a binding constraint without evaluation 
+(due to the lazy evaluation strategy of the
+target language Haskell).
 However, it is evaluated by \code{lazyBind} when its binding
@@
 we sketch the evaluation of the lazy unification constraint
-\code{xs++[e]=:<=[failed,True]} that occurs when the expression
+\code{xs++[e] =:<= [failed,True]} that occurs when the expression
 \code{last' [failed,True]} is evaluated (here we omit failed branches
@@
 \begin{haskell}
-  aBList 2 ++ [aBool 3] =:<= [failed,True]
-  $\leadsto$ [aBool 4, aBool 3] =:<= [failed, True]
+  aBoolList 2 ++ [aBool 3] =:<= [failed,True]
+  $\leadsto$ [aBool 4,aBool 3] =:<= [failed,True]
   $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
-  $\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed)
-           , 3 :=: LazyBind (lazyBind 3 True)] Success
+  $\leadsto$ Guard (4 :=: LazyBind (lazyBind 4 failed))
+           (Guard (3 :=: LazyBind (lazyBind 3 True)) Success)
 \end{haskell}
@@
 
-Unlike the Curry implementations PAKCS and MCC our approach does
-not manipulate a global state to unify a logic variable. Nevertheless,
-our benchmarks show that it can compete or even outperform these 
-implementations.
-The programs used for the benchmarks are primarily the same as in \cite{AntoyHanus05LOPSTR}.
-
-\begin{tabular}{l l}
-\bf{last}& Computes the last element of a list using strict equality\\
-\bf{simplify}& Simplification of symbolic arithmetic expressions\\
-\bf{varInExp}& Nondeterministically returns a variable occuring
-                in a symbolic arithmetic expression \\
-\bf{half}& Computes the half of a peano number using logic variables\\
-\end{tabular}
-
-The table below shows a comparison of the curry systems where
-strict equality (\code{=:=}) is used for unification:
-\footnote{\ccode{inc x n} is a naive addition that adds 1 \code{n} times
-          to its argument x }
-
-\begin{tabular}{|l|c|c|c|}
-\hline
-Expression: & KiCS2 & PAKCS & MCC \\
-\code{last (map (inc 0) [1..100000])} & &  & \\
-\code{simplify}                       &&&\\
-\code{varsInExp} & &  &\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
-\hline
-\end{tabular}
-
-All of our examples can also be defined with function patterns, i.e.,
-the lazy equality operator \code{=:<=} is used. The sole other
-Curry system featuring function patterns is PAKCS:
-
-\begin{tabular}{|l|c|c|c|}
-\hline
-Expression: & KiCS2 & PAKCS \\
-\code{last (map (inc 0) [1..100000])} & &  & \\
-\code{simplify}                       &&&\\
-\code{varsInExp} & &  &\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
-\hline
-\end{tabular}
-
-In Section \ref{sec:Unification} we stated that equational
-constraits can be implemented using ordinary pattern matching.
-We also mentioned that such an implementation would lead to
-a large search space that has to be traversed to extract
-the results of a computation. To show the benificial effects
-of our equality constraint implementations we compare them to
-the results using the equality operator \code{==} that
-produces such a search space:
-
 \begin{tabular}{|l|c|c|c|}
@@
 \hline
+\code{last (take 100000 (repeat failed) ++ [1])} & \code{failed} & \code{failed} & 0.14\\ 
 \code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
@@
 \end{tabular}
+
+The table above shows the evaluation of our approach. Primarily we
+used the same Benchmarks as in \cite{}
+
+Benchmarks for Last with ==, =:= and func. patterns
+
+also short comparison of =:= with PAKCS and MCC
 

### 2011-07-01 papers/wlp2011/paper.tex
added more benchmarks

--- 2011-06-30 papers/wlp2011/paper.tex
@@
 not manipulate a global state to unify a logic variable. Nevertheless,
-our benchmarks show that it can compete or even outperform these 
-implementations.
+our benchmarks show that it can in most cases
+compete or even outperform these implementations.
 The programs used for the benchmarks are primarily the same as in \cite{AntoyHanus05LOPSTR}.
@@
 Expression: & KiCS2 & PAKCS & MCC \\
-\code{last (map (inc 0) [1..100000])} & &  & \\
-\code{simplify}                       &&&\\
-\code{varsInExp} & &  &\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
+\code{last (map (inc 0) [1..100000])} &0.52 & 4.27&0.15 \\
+\code{grep} & 0.13& 2.83& 0.15\\
+\code{simplify} &10.33&0.15&0.01\\
+\code{varsInExp} & 86.27 & 0.35  & 0.03\\
+\code{fromPeano (half (toPeano 10000))} & &  & \\
+\code{palindrom} & 13.45 &32.82 & 1.04\\
+\code{horseman} & & &\\
 \hline
@@
 
-All of our examples can also be defined with function patterns, i.e.,
+Some of our examples can also be defined with function patterns, i.e.,
 the lazy equality operator \code{=:<=} is used. The sole other
@@
 Expression: & KiCS2 & PAKCS \\
-\code{last (map (inc 0) [1..100000])} & &  & \\
-\code{simplify}                       &&&\\
-\code{varsInExp} & &  &\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & &  &\\
+\code{last (map (inc 0) [1..100000])} & 0.20& 3.93\\
+\code{simplify}  &7.09&0.26&\\
+\code{varsInExp} & 0.14 & 0.69\\
+\code{fromPeano (half (toPeano 10000))} & &  \\
+\code{palindrom} & 17.55 & $\infty$\\
+\code{horseman} & & \\
 \hline
@@
 \hline 
-\code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
-\code{simplify}                       &10.34&10.33&7.27\\
-\code{varsInExp} & 0.49& 84.92 & 0.09\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & 23.3 & 1.81 & no Fun Pat\\
+\code{last (map (inc 0) [1..100000])} & 118.33 & 0.52 & 1.40 \\
+\code{grep} & 0.94 & 0.13 &\\
+\code{simplify} & 9.18& 10.33 & 7.09\\
+\code{varsInExp} & 0.50& 86.27 & 0.14\\
+\code{fromPeano (half (toPeano 10000))} &  &  & \\
+\code{palindrom} & 31.12&13.45 & 17.55\\
+\code{horseman} & & & \\
 \hline

### 2011-07-01 papers/wlp2011/paper.tex

--- 2011-06-30 papers/wlp2011/paper.tex
@@
 
+Unlike the Curry implementations PAKCS and MCC our approach does
+not manipulate a global state to unify a logic variable. Nevertheless,
+our benchmarks show that it can in most cases
+compete or even outperform these implementations.
+The programs used for the benchmarks are primarily the same as in \cite{AntoyHanus05LOPSTR}.
+
+\begin{tabular}{l l}
+\bf{last}& Computes the last element of a list using strict equality\\
+\bf{simplify}& Simplification of symbolic arithmetic expressions\\
+\bf{varInExp}& Nondeterministically returns a variable occuring
+                in a symbolic arithmetic expression \\
+\bf{half}& Computes the half of a peano number using logic variables\\
+\end{tabular}
+
+The table below shows a comparison of the curry systems where
+strict equality (\code{=:=}) is used for unification:
+\footnote{\ccode{inc x n} is a naive addition that adds 1 \code{n} times
+          to its argument x }
+
+\begin{tabular}{|l|c|c|c|}
+\hline
+Expression: & KiCS2 & PAKCS & MCC \\
+\code{last (map (inc 0) [1..100000])} &0.52 & 4.27&0.15 \\
+\code{grep} & 0.13& 2.83& 0.15\\
+\code{simplify} &10.33&0.15&0.01\\
+\code{varsInExp} & 86.27 & 0.35  & 0.03\\
+\code{fromPeano (half (toPeano 10000))} & &  & \\
+\code{palindrom} & 13.45 &32.82 & 1.04\\
+\code{horseman} & & &\\
+\hline
+\end{tabular}
+
+Some of our examples can also be defined with function patterns, i.e.,
+the lazy equality operator \code{=:<=} is used. The sole other
+Curry system featuring function patterns is PAKCS:
+
+\begin{tabular}{|l|c|c|c|}
+\hline
+Expression: & KiCS2 & PAKCS \\
+\code{last (map (inc 0) [1..100000])} & 0.20& 3.93\\
+\code{simplify}  &7.09&0.26&\\
+\code{varsInExp} & 0.14 & 0.69\\
+\code{fromPeano (half (toPeano 10000))} & &  \\
+\code{palindrom} & 17.55 & $\infty$\\
+\code{horseman} & & \\
+\hline
+\end{tabular}
+
+In Section \ref{sec:Unification} we stated that equational
+constraits can be implemented using ordinary pattern matching.
+We also mentioned that such an implementation would lead to
+a large search space that has to be traversed to extract
+the results of a computation. To show the benificial effects
+of our equality constraint implementations we compare them to
+the results using the equality operator \code{==} that
+produces such a search space:
+
 \begin{tabular}{|l|c|c|c|}
@@
 \hline 
-\code{last (take 100000 (repeat failed) ++ [1])} & \code{failed} & \code{failed} & 0.14\\ 
-\code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
-\code{simplify}                       &10.34&10.33&7.27\\
-\code{varsInExp} & 0.49& 84.92 & 0.09\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & 23.3 & 1.81 & no Fun Pat\\
+\code{last (map (inc 0) [1..100000])} & 118.33 & 0.52 & 1.40 \\
+\code{grep} & 0.94 & 0.13 &\\
+\code{simplify} & 9.18& 10.33 & 7.09\\
+\code{varsInExp} & 0.50& 86.27 & 0.14\\
+\code{fromPeano (half (toPeano 10000))} &  &  & \\
+\code{palindrom} & 31.12&13.45 & 17.55\\
+\code{horseman} & & & \\
 \hline
 \end{tabular}
-
-The table above shows the evaluation of our approach. Primarily we
-used the same Benchmarks as in \cite{}
-
-Benchmarks for Last with ==, =:= and func. patterns
-
-also short comparison of =:= with PAKCS and MCC
 

### 2011-07-01 papers/wlp2011/paper.tex

--- 2011-06-30 papers/wlp2011/paper.tex
@@
 \newcommand{\todo}[1]{\fbox{\sc To do: #1}}
+
+% Redefine description env as equation list to gain key alignment
+\usepackage{eqlist}
+\let\description=\eqlist
+\let\enddescription=\endeqlist
+\let\eqlistlabel\descriptionlabel
 
@@
 
+In Section \ref{sec:Unification} we stated that equational
+constraints could also be implemented using pattern matching, but that
+this implementation would increase the search space that has to be traversed 
+to extract the values of a computation. To show the benificial effects
+of our equality constraint implementations, we compare them to
+the results using the equality operator \code{==} that
+produces such a search space. The programs used for the benchmarks 
+are partially taken from \cite{AntoyHanus05LOPSTR}.
+%
+\begin{quote}
+\begin{description}
+ \item[last]      Computes the last element of a list
+ \item[simplify]  Simplification of symbolic arithmetic expressions
+ \item[varInExp]  Nondeterministically returns a variable occuring
+                  in a symbolic arithmetic expression
+ \item[half]      Computes the half of a peano number using logic variables
+ \item[palindrom] Checks if a list is a palindrom
+%  \item[grep]
+%  \item[horseman]
+\end{description}
+\end{quote}
+
+Our benchmarks are executed on a Linux machine running Debian 5.0.7 with
+an Intel Core 2 Duo (3.0GHz) processor. Our compiler has been executed with
+the Glasgow Haskell Compiler (GHC 7.0.4, option -O2). For later comparison with
+other systems we considered PAKCS (Version 1.9.2, based on a SICStus-Prolog
+4.1.2 backend) and MCC (Version 0.9.10). The timings were performed with the
+time command measuring the execution time (in seconds) of a compiled
+executable for each benchmark as a mean of three runs.
+
+\begin{figure}
+\centering
 \begin{tabular}{|l|c|c|c|}
 \hline
-Expression: &\code{==} & \code{=:=} & \code{=:<=}\\
+Expression                              & \code{==} & \code{=:=} & \code{=:<=} \\
 \hline
-\code{last (take 100000 (repeat failed) ++ [1])} & \code{failed} & \code{failed} & 0.14\\ 
-\code{last (map (inc 0) [1..100000])} & stack overflow & 0.45 & 0.13\\
-\code{simplify}                       &10.34&10.33&7.27\\
-\code{varsInExp} & 0.49& 84.92 & 0.09\\
-\code{toPeano 5000 == halfS (toPeano 10000)} & 23.3 & 1.81 & no Fun Pat\\
+\code{last (map (inc 0) [1..100000])}   & 118.33    & 0.52       & 0.20        \\
+% \code{grep}                             & 0.94      & 0.13       &             \\
+\code{simplify}                         & 9.18      & 10.33      & 7.09        \\
+\code{varInExp}                         & 0.50      & 86.27      & 0.14        \\
+\code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
+\code{palindrom}                        & 31.12     & 13.45      & 17.55       \\
+% \code{horseman}                         & 3.88      & 4.93       &             \\
 \hline
 \end{tabular}
-
-The table above shows the evaluation of our approach. Primarily we
-used the same Benchmarks as in \cite{}
-
-Benchmarks for Last with ==, =:= and func. patterns
-
-also short comparison of =:= with PAKCS and MCC
+\caption{Benchmarks: comparing different types of equation}
+ \label{fig:search-strategies}
+\end{figure}
+
+\todo{Results from comparing, ==, =:=, =:<=}
+
+As expected, the creation and traversal of a large search space introduced
+by \code{==} in most cases is much slower than our presented approach for
+unification. In addition, the \code{last} example shows that the lazy 
+unification operator \code{=:<=} improves performance when unifying an
+expression which has to be evaluted only partially. Using strict unification,
+all elements of the list are (unnecessarily) evaluated. Interestingly, in the
+\code{varInExp} example the strict unification performs significantly worse
+than the other equality operators. \todo{Explain why}
+
+In contrast to the Curry implementations PAKCS and MCC, out approach is based
+upon an explicit representation of the search space and does not manipulate
+a global state for unifying logic variables.
+Nevertheless, our benchmarks show that it can in most cases
+compete or even outperform these implementations.
+Figure~\ref{fig:unification} shows a comparison of the curry systems where
+strict equality (\ccode{=:=}) is used for unification.\footnote{%
+\ccode{inc x n} is a naive addition that \code{n} times
+increases its argument \code{x} by 1.} We can deduce from the results
+that the unification of the MCC performs best. However, we can in most cases
+outperform the Prolog based PAKCS but for some examples where further
+research and optimization is necessary.
+
+\begin{figure}
+\centering
+\begin{tabular}{|l|c|c|c|}
+\hline
+Expression                              & KiCS2 & PAKCS & MCC  \\
+\hline
+\code{last (map (inc 0) [1..100000])}   & 0.52  & 4.27   & 0.15 \\
+% \code{grep}                             & 0.13  & 2.83   & 0.15 \\
+\code{simplify}                         & 10.33 & 0.15   & 0.01 \\
+\code{varInExp}                         & 86.27 & 0.35   & 0.03 \\
+\code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
+\code{palindrom}                        & 13.45 & 32.82  & 1.04 \\
+% \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
+\hline
+\end{tabular}
+\caption{Benchmarks: strict equational constraints}
+ \label{fig:unification}
+\end{figure}
+
+As the MCC does not support functional patterns, the performance of lazy
+unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
+Again, our compiler performs well against PAKCS
+and outperforms it in most cases. \todo{extend}
+
+\begin{figure}
+\centering
+\begin{tabular}{|l|c|c|}
+\hline
+Expression                              & KiCS2 & PAKCS    \\
+\hline
+\code{last (map (inc 0) [1..100000])}   & 0.20  & 3.93     \\
+\code{simplify}                         & 7.09  & 0.26     \\
+\code{varInExp}                         & 0.14  & 0.69     \\
+\code{fromPeano (half (toPeano 10000))} & 11.57 & $\infty$ \\
+\code{palindrom}                        & 17.55 & $\infty$ \\
+\hline
+\end{tabular}
+\caption{Benchmarks: functional patterns}
+ \label{fig:functional-patterns}
+\end{figure}
 

### 2011-07-04 papers/wlp2011/paper.tex

@@
 and better abstractions for application programming, but it also gave rise
-to new challenges of implementation.
+to new implementation challenges.
 
@@
 \end{enumerate}
-
+%
 The latter approach requires the implementation
-of non-deterministic evaluations but has the advantage
+of non-deterministic computations in a deterministic language
+but has the advantage
 that the explicit handling of non-determinism allows for
@@
 (note that, in contrast to Prolog, unification is performed
-modulo function evaluation in functional logic languages).
+modulo function evaluation).
 Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
@@
 a powerful feature to express arbitrary selections in term structures.
-For instance, functional patterns supports a straightforward
+For instance, functional patterns support a straightforward
 processing of XML data with incompletely specified
@@
 we must also extend the definition of operations performing
-pattern matching. For instance, consider the definition of lists
+pattern matching. For instance, consider the definition of polymorphic lists
 \begin{curry}
@@
 
-In Section \ref{sec:Unification} we stated that equational
-constraints could also be implemented using pattern matching, but that
-this implementation would increase the search space that has to be traversed 
-to extract the values of a computation. To show the benificial effects
-of our equality constraint implementations, we compare them to
-the results using the equality operator \code{==} that
-produces such a search space. The programs used for the benchmarks 
-are partially taken from \cite{AntoyHanus05LOPSTR}.
-%
-\begin{quote}
-\begin{description}
- \item[last]      Computes the last element of a list
- \item[simplify]  Simplification of symbolic arithmetic expressions
- \item[varInExp]  Nondeterministically returns a variable occuring
-                  in a symbolic arithmetic expression
- \item[half]      Computes the half of a peano number using logic variables
- \item[palindrom] Checks if a list is a palindrom
-%  \item[grep]
-%  \item[horseman]
-\end{description}
-\end{quote}
-
-Our benchmarks are executed on a Linux machine running Debian 5.0.7 with
-an Intel Core 2 Duo (3.0GHz) processor. Our compiler has been executed with
-the Glasgow Haskell Compiler (GHC 7.0.4, option -O2). For later comparison with
-other systems we considered PAKCS (Version 1.9.2, based on a SICStus-Prolog
-4.1.2 backend) and MCC (Version 0.9.10). The timings were performed with the
+In this section we evaluate our implementation by a set of benchmarks.
+They were executed on a Linux machine running Debian 5.0.7 with
+an Intel Core 2 Duo (3.0GHz) processor.
+KiCS2 has been executed with
+the Glasgow Haskell Compiler (GHC 7.0.4, option -O2).
+For a comparison with other mature implementations of Curry,
+we considered PAKCS \cite{Hanus10PAKCS}
+(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
+and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
 time command measuring the execution time (in seconds) of a compiled
 executable for each benchmark as a mean of three runs.
+The programs used for the benchmarks,
+partially taken from \cite{AntoyHanus05LOPSTR},
+are \code{last} (compute the last element of a list),
+\code{simplify} (simplify a symbolic arithmetic expression),
+\code{varInExp} (non-deterministically return a variable occuring
+                  in a symbolic arithmetic expression),
+\code{half} (compute the half of a peano number using logic variables),
+and
+\code{palindrom} (check whether a list is a palindrom).
+
+In Section \ref{sec:Unification} we discussed that equational
+constraints could also be solved by generators without variable bindings,
+but this technique might increase the search space due to the
+possibly superfluous generation of all values.
+To show the beneficial effects of our implementation
+of equational constraints with variable bindings,
+we compare the results of equational constraints to results
+where the Boolean equality operator \code{==} is used (which
+does not perform bindings but enumerate all values):
 
@@
 \centering
-\begin{tabular}{|l|c|c|c|}
+\begin{tabular}{|l|r|r|r|}
 \hline
-Expression                              & \code{==} & \code{=:=} & \code{=:<=} \\
+Expression                              & \code{==~}& \code{=:=} & \code{=:<=} \\
 \hline
@@
 
-\todo{Results from comparing, ==, =:=, =:<=}
-
+\noindent
 As expected, the creation and traversal of a large search space introduced
-by \code{==} in most cases is much slower than our presented approach for
-unification. In addition, the \code{last} example shows that the lazy 
+by \code{==} is in most cases much slower than our presented approach
+with variable bindings.
+In addition, the example \code{last} shows that the lazy 
 unification operator \code{=:<=} improves performance when unifying an
@@
 
-In contrast to the Curry implementations PAKCS and MCC, out approach is based
-upon an explicit representation of the search space and does not manipulate
-a global state for unifying logic variables.
+In contrast to the Curry implementations PAKCS and MCC, our approach is based
+on an explicit representation of the search space instead of manipulating
+a global state containing bindings for logic variables.
 Nevertheless, our benchmarks show that it can in most cases
-compete or even outperform these implementations.
-Figure~\ref{fig:unification} shows a comparison of the curry systems where
-strict equality (\ccode{=:=}) is used for unification.\footnote{%
+compete with or even outperform these implementations.
+Figure~\ref{fig:unification} shows a comparison of the Curry systems where
+equational constraints are used in the examples.\footnote{%
 \ccode{inc x n} is a naive addition that \code{n} times
 increases its argument \code{x} by 1.} We can deduce from the results
-that the unification of the MCC performs best. However, we can in most cases
-outperform the Prolog based PAKCS but for some examples where further
-research and optimization is necessary.
+that the unification of MCC performs best. However, in most cases
+our implementation outperforms the Prolog-based PAKCS implementation,
+except for some examples where further investigations
+and optimizations are necessary.
 
@@
 \centering
-\begin{tabular}{|l|c|c|c|}
+\begin{tabular}{|l|l|l|l|}
 \hline
@@
 
-As the MCC does not support functional patterns, the performance of lazy
+\noindent
+As MCC does not support functional patterns, the performance of lazy
 unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
@@
 
-Future work: adding other constraint systems
-
-\bibliographystyle{plain}
+We have presented an implementation of equational constraints
+and functional patterns in a purely functional implementation
+of Curry. Our implementation is based on adding binding constraints
+to computed values and processing them when values are extracted
+at the top level of a computation.
+Since we have added only additional constructors and rules
+for these constructors, our implementation does not cause
+additional work if these features are not used, i.e.,
+our implementation does not sacrifice the high efficiency
+of the kernel implementation as described in
+\cite{BrasselFischerHanusReck11}.
+However, if these features are used, they usually lead
+to a more efficient execution, as shown by our benchmarks.
+
+Other implementations of equational constraints in functional logic
+languages use side effects for their implementation.
+For instance, PAKCS \cite{Hanus10PAKCS}
+exploits the implementation of logic variables in Prolog,
+which are implemented on the primitive level by side effects.
+MCC \cite{Lux99FLOPS} compiles into C where a specific
+abstract machine implements the handling of logic variables.
+We have shown that our implementation is competitive to those.
+In contrast to those systems, our implementation supports
+a variety of search strategies like breadth-first or parallel search.
+
+\todo{Future work: adding other constraint systems}
+
+\bibliographystyle{mhnumber} %plain}
 \bibliography{mh}

### 2011-07-05 papers/wlp2011/paper.tex

@@
 \documentclass{llncs}
+
+\addtolength{\textheight}{0.2cm}
+\addtolength{\topmargin}{-0.1cm}
+\addtolength{\textwidth}{0.2cm}
+\addtolength{\oddsidemargin}{-0.1cm}
+\addtolength{\evensidemargin}{-0.1cm}
 
@@
 \usepackage{listings}
-\lstset{aboveskip=1.0ex,
-        belowskip=1.0ex,
+\lstset{aboveskip=0.5ex,
+        belowskip=0.5ex,
         showstringspaces=false, % no special string space
@@
 or computing with generators by rewriting
-are equivalent, i.e., compute the same results.
+are equivalent, i.e., compute the same values.
 Although this implementation technique is correct \cite{Brassel11Thesis},
@@
 where values of the arguments of an operation are determined before the
-operation is evaluated. Note that this does not necessarily require
-an eager evaluation of arguments.
-Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
-define lazy evaluation strategies for functional logic programs
-with call-time choice semantics where actual arguments passed to
-operations are shared. Hence, we can evaluate the expression above
-lazily, provided that all occurrences of \code{aBool}
+operation is evaluated. This can be enforced in a lazy strategy
+by sharing actual arguments.
+For instance, the expression above can be lazily evaluated
+provided that all occurrences of \code{aBool}
 are shared so that all of them reduce either to \code{True} or to \code{False}
-consistently. The requirements of this call-time choice semantics are the
-reason why it is not simply possible to use list comprehensions
-or non-determinism monads for a straightforward implementation
-of functional logic programs in Haskell \cite{FischerKiselyovShan09}.
+consistently.
 
@@
 
-In this section we sketch the translation of Curry programs
-into Haskell programs as performed by KiCS2,
-since this is necessary to understand the extensions
-described in the subsequent sections.
+To understand the extensions described in the subsequent sections,
+we sketch the translation of Curry programs
+into Haskell programs as performed by KiCS2.
 More details about this translation scheme can be found in
@@
 \end{haskell}
-The first argument of each \code{Choice} constructor of type \code{ID}
+The first argument of type \code{ID} of each \code{Choice} constructor
 is used to implement the call-time choice semantics
@@
 Now we can print all values contained in a choice structure
-in a depth-first manner by the following operation:\footnote{%
+in a depth-first manner by the following I/O operation:\footnote{%
 Note that this code has been simplified for readability
@@
 \end{haskell}
-This operation ignores failures and prints values that are no \code{Choice}.
+This operation ignores failures and prints values that are not a \code{Choice}.
 For a \code{Choice}, it checks whether a decision for
@@
 To avoid an unnecessary growth of the search space represented by
-\code{Choice} constructors, out compiler performs an optimization for
+\code{Choice} constructors, our compiler performs an optimization for
 deterministic operations. If an operation does not call, 
@@
   False  =:= False   =  Success
-
   []     =:= []      =  Success
@@
 Now consider the equational constraint \ccode{x =:= [True]}.
-If the logic variable \code{x} is replaced by \code{aBList},
+If the logic variable \code{x} is replaced by \code{aBoolList},
 the translated expression \ccode{aBoolList s =:= [True]}
@@
 \end{haskell}
-Thus, free variables are prefixed by \code{\us{}x}.
+Here, free variables are marked by the prefix \code{\us{}x}.
 
@@
 %
-
 The consistency of constraints is checked when values are extracted
@@
 The operation \code{add} checks the consistency of the constraint \code{c}
-with the decisions made so far and stores the decision made by the constraint
-if consistency is sustained. In this case, the constrained value is evaluated
-before the contstraint is removed to allow backtracking.
+with the decisions made so far and, in case of consistency,
+stores the decision made by the constraint.
+In this case, the constrained value is evaluated
+before the constraint is removed to allow backtracking.
 Furthermore, the operations \code{lookupDecision} and \code{setDecision}
@@
 a list.
-Using equational constraint, we can define it as follows:
+Using equational constraints, we can define it as follows:
 \begin{curry}
@@
 \code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
-which corresponds to the value \code{True} of the generator \code{aBool}.
+which corresponds to the value \code{True} of the generator \code{(aBool 3)}.
 Note that the variable with identifier \code{4} does not occur
@@
 
-In this section we evaluate our implementation by a set of benchmarks.
+In this section we evaluate our implementation by some benchmarks.
 They were executed on a Linux machine running Debian 5.0.7 with
@@
 partially taken from \cite{AntoyHanus05LOPSTR},
-are \code{last} (compute the last element of a list),
+are \code{last} (compute the last element of a list),\footnote{%
+\ccode{inc x n} is a naive addition that \code{n} times
+increases its argument \code{x} by 1.}
 \code{simplify} (simplify a symbolic arithmetic expression),
@@
 and
-\code{palindrom} (check whether a list is a palindrom).
-
-In Section \ref{sec:Unification} we discussed that equational
-constraints could also be solved by generators without variable bindings,
-but this technique might increase the search space due to the
-possibly superfluous generation of all values.
-To show the beneficial effects of our implementation
-of equational constraints with variable bindings,
-we compare the results of equational constraints to results
-where the Boolean equality operator \code{==} is used (which
-does not perform bindings but enumerate all values):
-
-\begin{figure}
+\code{palindrome} (check whether a list is a palindrome).
+
+\begin{figure}[t]
 \centering
@@
 \code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
-\code{palindrom}                        & 31.12     & 13.45      & 17.55       \\
+\code{palindrome}                       & 31.12     & 13.45      & 17.55       \\
 % \code{horseman}                         & 3.88      & 4.93       &             \\
@@
 \end{tabular}
-\caption{Benchmarks: comparing different types of equation}
+\caption{Benchmarks: comparing different equation solving methods}
  \label{fig:search-strategies}
@@
 
-\noindent
+In Section~\ref{sec:Unification} we mentioned that equational
+constraints could also be solved by generators without variable bindings,
+but this technique might increase the search space due to the
+possibly superfluous generation of all values.
+To show the beneficial effects of our implementation
+of equational constraints with variable bindings,
+we compare in Fig.~\ref{fig:search-strategies}
+the results of using equational constraints (\code{=:=})
+to the results
+where the Boolean equality operator (\code{==}) is used (which
+does not perform bindings but enumerate all values).
 As expected, the creation and traversal of a large search space introduced
@@
 In addition, the example \code{last} shows that the lazy 
-unification operator \code{=:<=} improves performance when unifying an
+unification operator (\code{=:<=}) improves performance when unifying an
 expression which has to be evaluted only partially. Using strict unification,
 all elements of the list are (unnecessarily) evaluated. Interestingly, in the
-\code{varInExp} example the strict unification performs significantly worse
+\code{varInExp} example equational constraints perform significantly worse
 than the other equality operators. \todo{Explain why}
 
-In contrast to the Curry implementations PAKCS and MCC, our approach is based
-on an explicit representation of the search space instead of manipulating
-a global state containing bindings for logic variables.
-Nevertheless, our benchmarks show that it can in most cases
-compete with or even outperform these implementations.
-Figure~\ref{fig:unification} shows a comparison of the Curry systems where
-equational constraints are used in the examples.\footnote{%
-\ccode{inc x n} is a naive addition that \code{n} times
-increases its argument \code{x} by 1.} We can deduce from the results
-that the unification of MCC performs best. However, in most cases
-our implementation outperforms the Prolog-based PAKCS implementation,
-except for some examples where further investigations
-and optimizations are necessary.
-
-\begin{figure}
+\begin{figure}[t]
 \centering
-\begin{tabular}{|l|l|l|l|}
+\begin{tabular}{|l|r|r|r|}
 \hline
@@
 \code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
-\code{palindrom}                        & 13.45 & 32.82  & 1.04 \\
+\code{palindrome}                       & 13.45 & 32.82  & 1.04 \\
 % \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
@@
 \end{tabular}
-\caption{Benchmarks: strict equational constraints}
+\caption{Benchmarks: equational constraints in different Curry implementations}
  \label{fig:unification}
@@
 
-\noindent
-As MCC does not support functional patterns, the performance of lazy
-unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
-Again, our compiler performs well against PAKCS
-and outperforms it in most cases. \todo{extend}
-
-\begin{figure}
+In contrast to the Curry implementations PAKCS and MCC, our approach is based
+on an explicit representation of the search space instead of
+backtracking and manipulating
+a global state containing bindings for logic variables.
+Nevertheless, the benchmarks in Fig.~\ref{fig:unification}
+show that it can compete with or even outperform these implementations.
+In these benchmarks only equational constraints are used.
+The results show that the implementation of unification
+of MCC performs best. However, in most cases
+our implementation outperforms the Prolog-based PAKCS implementation,
+except for some examples where further investigations
+and optimizations are necessary.
+
+\begin{figure}[t]
 \centering
-\begin{tabular}{|l|c|c|}
+\begin{tabular}{|l|r|r|}
 \hline
@@
 \code{fromPeano (half (toPeano 10000))} & 11.57 & $\infty$ \\
-\code{palindrom}                        & 17.55 & $\infty$ \\
+\code{palindrome}                       & 17.55 & $\infty$ \\
 \hline
 \end{tabular}
-\caption{Benchmarks: functional patterns}
+\caption{Benchmarks: comparing different implementations of functional patterns}
  \label{fig:functional-patterns}
 \end{figure}
+
+As MCC does not support functional patterns, the performance of lazy
+unification is compared with PAKCS only (Fig.~\ref{fig:functional-patterns}).
+Again, our compiler performs well against PAKCS
+and outperforms it in most cases (``$\infty$'' denotes a run time
+of more than 30 minutes). \todo{extend}
 
@@
 at the top level of a computation.
-Since we have added only additional constructors and rules
-for these constructors, our implementation does not cause
+Since we have added only new constructors and rules
+for them in our implementation, it does not introduce
 additional work if these features are not used, i.e.,
 our implementation does not sacrifice the high efficiency
-of the kernel implementation as described in
+of the kernel implementation as shown in
 \cite{BrasselFischerHanusReck11}.
 However, if these features are used, they usually lead
-to a more efficient execution, as shown by our benchmarks.
+to a more efficient execution, as demonstrated by our benchmarks.
 
@@
 In contrast to those systems, our implementation supports
-a variety of search strategies like breadth-first or parallel search.
-
-\todo{Future work: adding other constraint systems}
+a variety of search strategies, like breadth-first or parallel search,
+where the avoidance of side effects is important.
+
+For future work it might be interesting to add
+further constraint structures, like real arithmetic
+or finite domain constraints, to our implementation.
+This might be possible by extending the kinds of constraints
+of our implementation and solving them by
+functional approaches like \cite{SchrijversStuckeyWadler09}.
 

### 2011-07-06 lib/External_Prelude.hs
Renaming of ID constructors (ChoiceID, NarrowedID), requires bootstrapping!

@@ -147,14 +147,14 @@ instance Unifiable C_Int where
   bind i (C_CurryInt x2) = (i :=: ChooseN 0 1) : bind (leftID i) x2
   bind i (Choice_C_Int j l r) = [(ConstraintChoice j (bind i l) (bind i r))]
   bind i (Choices_C_Int j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  bind i (Choices_C_Int j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
+  bind i (Choices_C_Int j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
   bind _ Fail_C_Int = [Unsolvable]
   bind i (Guard_C_Int cs e) = cs ++ (bind i e)
   lazyBind i (C_Int      x2) = [i :=: ChooseN 0 1, leftID i :=: LazyBind (lazyBind (leftID i) (primint2curryint x2))]
   lazyBind i (C_CurryInt x2) = [i :=: ChooseN 0 1, leftID i :=: LazyBind (lazyBind (leftID i) x2)]
   lazyBind i (Choice_C_Int j l r) = [(ConstraintChoice j (lazyBind i l) (lazyBind i r))]
   lazyBind i (Choices_C_Int j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  lazyBind i (Choices_C_Int j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (lazyBind i) xs))]
+  lazyBind i (Choices_C_Int j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (lazyBind i) xs))]
   lazyBind _ Fail_C_Int = [Unsolvable]
   lazyBind i (Guard_C_Int cs e) = cs ++ [(i :=: (LazyBind (lazyBind i e)))]
 
@@ -273,12 +273,12 @@ instance Unifiable C_Float where
   (=.<=) _ _ = Fail_C_Success
   bind i (Choice_C_Float j l r) = [(ConstraintChoice j (bind i l) (bind i r))]
   bind i (Choices_C_Float j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  bind i (Choices_C_Float j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
[…]

### 2011-07-06 runtime/ID.hs

@@ -26,10 +26,12 @@ data Constraint
   -- |Unsolvable constraint
   | Unsolvable
   -- |Non-deterministic choice between two lists of constraints
-  | ConstraintChoice ID [Constraint] [Constraint]
+  | ConstraintChoice ID Constraints Constraints
   -- |Non-deterministic choice between a list of lists of constraints
-  | ConstraintChoices ID [[Constraint]]
+  | ConstraintChoices ID [Constraints]
     deriving (Eq, Show)
+
+type Constraints = [Constraint]
 
 -- ---------------------------------------------------------------------------
 -- Choice
@@ -55,7 +57,7 @@ data Choice
   | BoundTo ID Int
     -- |A free variable is lazily bound to an expression by a function
    --   pattern
-  | LazyBind [Constraint]
+  | LazyBind Constraints
     deriving Show
 
 -- TODO: Idea: Split ChooseN Int Int into two cases
@@ -92,17 +94,17 @@ isDefaultChoice _        = False
 -- Type to identify different Choice structures in a non-deterministic result
 data ID
     -- |Identifier for a choice introduced by the use of the (?) operator
-  = ID Ref
+  = ChoiceID Ref
     -- |Identifier for a choice for a free variable
   | FreeID [Int] IDSupply
     -- |Identifier for a choice for a narrowed variable (free before)
-  | Narrowed [Int] IDSupply
+  | NarrowedID [Int] IDSupply
     deriving Eq
 
 instance Show ID where
-  show (ID i)       = "?" ++ show i
-  show (FreeID _ i) = "_x" ++ show i
-  show (Narrowed _ i) = "Narrowed" ++ show i
+  show (ChoiceID     i) = "?" ++ show i
+  show (FreeID     _ i) = "_x" ++ show i
+  show (NarrowedID _ i) = "Narrowed" ++ show i
 
 -- |Construct an 'ID' for a free variable from an 'IDSupply'
 freeID :: [Int] -> IDSupply -> ID
@@ -110,19 +112,19 @@ freeID = FreeID
 
 -- |Construct an 'ID' for a binary choice from an 'IDSupply'
 thisID :: IDSupply -> ID
-thisID s = ID (thisRef s)
+thisID s = ChoiceID (thisRef s)
 
[…]

### 2011-07-06 runtime/PrimTypes.hs

@@ -87,7 +87,7 @@ instance Unifiable BinInt where
   bind i (Pos x2) = ((i :=: (ChooseN 2 1)):(concat [(bind (leftID i) x2)]))
   bind i (Choice_BinInt j l r) = [(ConstraintChoice j (bind i l) (bind i r))]
   bind i (Choices_BinInt j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  bind i (Choices_BinInt j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
+  bind i (Choices_BinInt j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
   bind _ Fail_BinInt = [Unsolvable]
   bind i (Guard_BinInt cs e) = cs ++ (bind i e)
   lazyBind i (Neg x2) = [(i :=: (ChooseN 0 1)),((leftID i) :=: (LazyBind (lazyBind (leftID i) x2)))]
@@ -95,7 +95,7 @@ instance Unifiable BinInt where
   lazyBind i (Pos x2) = [(i :=: (ChooseN 2 1)),((leftID i) :=: (LazyBind (lazyBind (leftID i) x2)))]
   lazyBind i (Choice_BinInt j l r) = [(ConstraintChoice j (lazyBind i l) (lazyBind i r))]
   lazyBind i (Choices_BinInt j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  lazyBind i (Choices_BinInt j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (lazyBind i) xs))]
+  lazyBind i (Choices_BinInt j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (lazyBind i) xs))]
   lazyBind _ Fail_BinInt = [Unsolvable]
   lazyBind i (Guard_BinInt cs e) = cs ++ [(i :=: (LazyBind (lazyBind i e)))]
 
@@ -180,7 +180,7 @@ instance Unifiable Nat where
   bind i (I x2) = ((i :=: (ChooseN 2 1)):(concat [(bind (leftID i) x2)]))
   bind i (Choice_Nat j l r) = [(ConstraintChoice j (bind i l) (bind i r))]
   bind i (Choices_Nat j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
[…]

### 2011-07-06 runtime/Search.hs

@@ -118,7 +118,7 @@ printValsDFS' fb cont (Frees i xs)   = lookupChoiceID i >>= choose
     choose (ChooseN c _, _) = printValsDFS fb cont (xs !! c)
     choose (NoChoice   , j) = cont $ choicesCons j xs
 
-printValsDFS' fb cont (Choices i@(Narrowed pns _) xs) = lookupChoiceID i >>= choose
+printValsDFS' fb cont (Choices i@(NarrowedID pns _) xs) = lookupChoiceID i >>= choose
   where
     choose (LazyBind cs, _) = processLazyBind fb cs i xs (printValsDFS fb cont)
     choose (ChooseN c _, _) = printValsDFS fb cont (xs !! c)
@@ -270,7 +270,7 @@ searchDFS' cont (Frees i xs) = lookupChoiceID i >>= choose
       reset <- setUnsetChoice i NoChoice
       searchDFS cont (guardCons cs $ choicesCons i xs) |< reset
 
-searchDFS' cont (Choices i@(Narrowed pns _) xs) = lookupChoice i >>= choose
+searchDFS' cont (Choices i@(NarrowedID pns _) xs) = lookupChoice i >>= choose
   where
     choose (LazyBind cs) = processLB cs
     choose (ChooseN c _) = searchDFS cont (xs !! c)
@@ -583,7 +583,7 @@ searchMPlus'' cont (Choice i x y) = lookupChoice' i >>= choose
     choose NoChoice    = (setChoice' i ChooseLeft  >> searchMPlus' cont x)
                          `mplus`
                          (setChoice' i ChooseRight >> searchMPlus' cont y)
-searchMPlus'' cont (Choices i@(Narrowed pns _)  branches) =
+searchMPlus'' cont (Choices i@(NarrowedID pns _)  branches) =
    lookupChoice' i >>= choose
   where
     choose (ChooseN c _) = searchMPlus' cont (branches !! c)

### 2011-07-06 runtime/Types.hs

@@ -65,13 +65,13 @@ instance Unifiable C_Success where
   bind i C_Success = ((i :=: (ChooseN 0 0)):(concat []))
   bind i (Choice_C_Success j l r) = [(ConstraintChoice j (bind i l) (bind i r))]
   bind i (Choices_C_Success j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  bind i (Choices_C_Success j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
+  bind i (Choices_C_Success j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
   bind _ Fail_C_Success = [Unsolvable]
   bind i (Guard_C_Success cs e) = cs ++ (bind i e)
   lazyBind i C_Success = [(i :=: (ChooseN 0 0))]
   lazyBind i (Choice_C_Success j l r) = [(ConstraintChoice j (lazyBind i l) (lazyBind i r))]
   lazyBind i (Choices_C_Success j@(FreeID _ _) xs) = [(i :=: (BindTo j))]
-  lazyBind i (Choices_C_Success j@(Narrowed _ _) xs) = [(ConstraintChoices j (map (lazyBind i) xs))]
+  lazyBind i (Choices_C_Success j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (lazyBind i) xs))]
   lazyBind _ Fail_C_Success = [Unsolvable]
   lazyBind i (Guard_C_Success cs e) = cs ++ [(i :=: (LazyBind (lazyBind i e)))]
 -- END GENERATED FROM PrimTypes.curry
@@ -123,13 +123,13 @@ data Try a
     deriving Show
 
 tryChoice :: ID -> a -> a -> Try a
-tryChoice i@(ID _) = Choice i
-tryChoice _        = error "Basics.tryChoice: no ID"
+tryChoice i@(ChoiceID _) = Choice i
+tryChoice _              = error "Basics.tryChoice: no ChoiceID"
 
 tryChoices :: ID -> [a] -> Try a
-tryChoices (ID _)         = error "Basics.tryChoices: ID"
-tryChoices i@(FreeID _ _) = Frees   i
-tryChoices i@(Narrowed _ _) = Choices i
+tryChoices (ChoiceID       _) = error "Basics.tryChoices: Choice"
+tryChoices i@(FreeID     _ _) = Frees   i
[…]

### 2011-07-06 runtime/Basics.hs
Renaming of Try constructors (Choices -> Narrowed, Frees -> Free)

@@ -23,24 +23,24 @@ nonAsciiChr i = chr# i
 d_dollar_bang :: (NonDet a, NonDet b) => (a -> b) -> a -> b
 d_dollar_bang f x = hnf (try x)
   where
-   hnf (Val v)        = f v -- inlined d_apply f v
-   hnf Fail           = failCons
-   hnf (Choice i a b) = choiceCons i (hnf (try a)) (hnf (try b))
-   hnf (Choices i xs) = choicesCons i (map (hnf . try) xs)
-   hnf (Frees i xs)   = f (choicesCons i xs)
-   hnf (Guard c e)    = guardCons c (hnf (try e))
+   hnf (Val v)         = f v -- inlined d_apply f v
+   hnf Fail            = failCons
+   hnf (Choice i a b)  = choiceCons i (hnf (try a)) (hnf (try b))
+   hnf (Narrowed i xs) = choicesCons i (map (hnf . try) xs)
+   hnf (Free i xs)     = f (choicesCons i xs)
+   hnf (Guard c e)     = guardCons c (hnf (try e))
 
 -- Apply a non-deterministic function to the head normal form
 nd_dollar_bang :: (NonDet a, NonDet b) => (Func a b) -> a -> IDSupply -> b
 nd_dollar_bang f x s = hnf (try x)
   where
-   hnf (Val v)        = nd_apply f v s
-   hnf Fail           = failCons
+   hnf (Val v)         = nd_apply f v s
+   hnf Fail            = failCons
    -- TODO Do we better use leftSupply and rightSupply?
-   hnf (Choice i a b) = choiceCons i (hnf (try a)) (hnf (try b))
-   hnf (Choices i xs) = choicesCons i (map (hnf . try) xs)
-   hnf (Frees i xs)   = nd_apply f (choicesCons i xs) s
-   hnf (Guard c e)    = guardCons c (hnf (try e))
+   hnf (Choice i a b)  = choiceCons i (hnf (try a)) (hnf (try b))
+   hnf (Narrowed i xs) = choicesCons i (map (hnf . try) xs)
+   hnf (Free i xs)     = nd_apply f (choicesCons i xs) s
+   hnf (Guard c e)     = guardCons c (hnf (try e))
 
 -- TODO: test implementation for $! replace if more efficient
 -- d_dollar_bang_test :: (NonDet a, NonDet b) => (a -> b) -> a -> b

### 2011-07-06 runtime/ID.hs

@@ -2,7 +2,7 @@
 -- ID module
 -- ---------------------------------------------------------------------------
 module ID
-  ( Constraint (..), Choice (..), defaultChoice, isDefaultChoice
+  ( Constraint (..), Constraints, Choice (..), defaultChoice, isDefaultChoice
   , ID (..), IDSupply
   , mkInt, leftID, rightID, narrowID
   , initSupply, leftSupply, rightSupply, thisID, freeID

### 2011-07-06 runtime/Search.hs

@@ -112,13 +112,13 @@ printValsDFS' fb cont (Choice i x y) = lookupChoice i >>= choose
 
 
 
-printValsDFS' fb cont (Frees i xs)   = lookupChoiceID i >>= choose
+printValsDFS' fb cont (Free i xs)   = lookupChoiceID i >>= choose
   where
     choose (LazyBind cs, _) = processLazyBind fb cs i xs (printValsDFS fb cont)
     choose (ChooseN c _, _) = printValsDFS fb cont (xs !! c)
     choose (NoChoice   , j) = cont $ choicesCons j xs
 
-printValsDFS' fb cont (Choices i@(NarrowedID pns _) xs) = lookupChoiceID i >>= choose
+printValsDFS' fb cont (Narrowed i@(NarrowedID pns _) xs) = lookupChoiceID i >>= choose
   where
     choose (LazyBind cs, _) = processLazyBind fb cs i xs (printValsDFS fb cont)
     choose (ChooseN c _, _) = printValsDFS fb cont (xs !! c)
@@ -259,7 +259,7 @@ searchDFS' cont (Choice i x1 x2) = lookupChoice i >>= choose
       reset <- setUnsetChoice i c
       searchDFS cont x |< reset
 
-searchDFS' cont (Frees i xs) = lookupChoiceID i >>= choose
+searchDFS' cont (Free i xs) = lookupChoiceID i >>= choose
   where
     choose (LazyBind cs, _) = processLB cs
     choose (ChooseN c _, _) = searchDFS cont (xs !! c)
@@ -270,7 +270,7 @@ searchDFS' cont (Frees i xs) = lookupChoiceID i >>= choose
       reset <- setUnsetChoice i NoChoice
       searchDFS cont (guardCons cs $ choicesCons i xs) |< reset
 
-searchDFS' cont (Choices i@(NarrowedID pns _) xs) = lookupChoice i >>= choose
+searchDFS' cont (Narrowed i@(NarrowedID pns _) xs) = lookupChoice i >>= choose
   where
     choose (LazyBind cs) = processLB cs
     choose (ChooseN c _) = searchDFS cont (xs !! c)
@@ -347,7 +347,7 @@ searchBFS x = bfs [] [] (return ()) (return ()) x
         newSet c = set   >> setChoice i c
         newReset = reset >> setChoice i NoChoice
 
-    bfs xs ys set reset (Choices i cs) = set   >> lookupChoice i >>= choose
[…]

### 2011-07-06 runtime/Types.hs

@@ -4,6 +4,268 @@ module Types where
 import ID
 
 -- ---------------------------------------------------------------------------
+-- Try structure
+-- ---------------------------------------------------------------------------
+
+-- TODO: Reason about pros and cons of reusing Choices for free, narrowed, (?)
+
+-- |Data type to wrap values in a generic structure
+data Try a
+  = Val a               -- ^Value in head normal form (HNF)
+  | Fail                -- ^Fail
+  | Choice ID a a       -- ^Binary choice, used for (?)
+  | Narrowed ID [a]     -- ^N-ary choice for narrowed variable
+  | Free ID [a]         -- ^N-ary choice for free variable
+  | Guard Constraints a -- ^Constrained value
+    deriving Show
+
+-- |Convert a binary choice of type a into one of type 'Try' a
+tryChoice :: ID -> a -> a -> Try a
+tryChoice i@(ChoiceID _) = Choice i
+tryChoice _              = error "Basics.tryChoice: no ChoiceID"
+
+-- |Convert a n-ary choice of type a into one of type 'Try' a
+tryChoices :: ID -> [a] -> Try a
+tryChoices (ChoiceID       _) = error "Basics.tryChoices: ChoiceID"
+tryChoices i@(FreeID     _ _) = Free     i
+tryChoices i@(NarrowedID _ _) = Narrowed i
+
+-- ---------------------------------------------------------------------------
+-- Non-determinism
+-- ---------------------------------------------------------------------------
+
+-- |Class for data that supports nondeterministic values
+class NonDet a where
+  -- |Constructor for a binary choice, used for (?)
+  choiceCons :: ID -> a -> a -> a
+  -- |Constructor for a n-ary choice, used for free variables and narrowing
+  choicesCons:: ID -> [a] -> a
+  -- |Constructor for a failed computation
+  failCons   :: a
+  -- |Constructor for a constrained value
+  guardCons  :: Constraints -> a -> a
+  -- |conversion of a value into its generic 'Try' structure
+  try        :: a -> Try a
+
+  -- Alternative approach: Replace generic Try structure with a matching
+  -- function to gain more efficiency.
+
+  -- |Matching with different function for the respective constructors
[…]

### 2011-07-11 papers/wlp2011/paper.tex
benchmark update

@@
   printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
-  printValsDFS v                = print v
    where
@@
                          setDecision i NoDecision
+  printValsDFS v                = print v
 \end{haskell}
@@
   Fail         & _     =  Fail
-  _            & Fail  =  Fail
   Choice i l r & c     =  Choice i (l & c) (r & c)
@@
 
+% The benchmark results are from unif@climens2011_7_5_20_38_47.bench
+
 In this section we evaluate our implementation by some benchmarks.
@@
 KiCS2 has been executed with
-the Glasgow Haskell Compiler (GHC 7.0.4, option -O2).
+the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) and uses
+an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
 For a comparison with other mature implementations of Curry,
@@
 \hline
-\code{last (map (inc 0) [1..100000])}   & 118.33    & 0.52       & 0.20        \\
-% \code{grep}                             & 0.94      & 0.13       &             \\
-\code{simplify}                         & 9.18      & 10.33      & 7.09        \\
-\code{varInExp}                         & 0.50      & 86.27      & 0.14        \\
-\code{fromPeano (half (toPeano 10000))} & 26.49     & 6.26       & 11.58       \\
-\code{palindrome}                       & 31.12     & 13.45      & 17.55       \\
-% \code{horseman}                         & 3.88      & 4.93       &             \\
+\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
+% \code{grep}                             & 1.06      & 0.19      &             \\
+\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
+\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
+% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
+% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
+\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
+\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
+% \code{horseman}                         & 3.58      & 3.13      &             \\
 \hline
@@
 expression which has to be evaluted only partially. Using strict unification,
-all elements of the list are (unnecessarily) evaluated. Interestingly, in the
-\code{varInExp} example equational constraints perform significantly worse
-than the other equality operators. \todo{Explain why}
+all elements of the list are (unnecessarily) evaluated.
 
@@
 \hline
-\code{last (map (inc 0) [1..100000])}   & 0.52  & 4.27   & 0.15 \\
-% \code{grep}                             & 0.13  & 2.83   & 0.15 \\
-\code{simplify}                         & 10.33 & 0.15   & 0.01 \\
-\code{varInExp}                         & 86.27 & 0.35   & 0.03 \\
-\code{fromPeano (half (toPeano 10000))} & 6.26  & 109.71 & 3.17 \\
-\code{palindrome}                       & 13.45 & 32.82  & 1.04 \\
-% \code{horseman}                         & 4.93  & 8.85   & 0.42 \\
+\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
+% \code{grep}                             & 0.19  & 2.88   & 0.14 \\
+\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
+\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
+% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
+% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
+\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
+\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
+% \code{horseman}                         & 3.13  & 8.70   & 0.42 \\
 \hline
@@
 \hline
-\code{last (map (inc 0) [1..100000])}   & 0.20  & 3.93     \\
-\code{simplify}                         & 7.09  & 0.26     \\
-\code{varInExp}                         & 0.14  & 0.69     \\
-\code{fromPeano (half (toPeano 10000))} & 11.57 & $\infty$ \\
-\code{palindrome}                       & 17.55 & $\infty$ \\
+\code{last (map (inc 0) [1..100000])}   & 0.26  & 3.97     \\
+\code{simplify}                         & 7.34  & 0.27     \\
+\code{varInExp}                         & 1.40  & 8.88     \\
+% \code{varInExp2}                         & 1.04  & 0.08    \\
+% \code{varInExp3}                         & 11.28 & 16.62   \\
+\code{fromPeano (half (toPeano 10000))} & 11.88 & $\infty$ \\
+\code{palindrome}                       & 23.21 & $\infty$ \\
 \hline
@@
 and outperforms it in most cases (``$\infty$'' denotes a run time
-of more than 30 minutes). \todo{extend}
+of more than 30 minutes).
 

### 2011-07-12 papers/wlp2011/paper.tex

@@
 
-In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
+In this paper we consider KiCS2 \cite{BrasselHanusPeemoellerReck11},
 a new system that compiles functional logic programs of the
@@
 into purely functional Haskell programs.
-We have shown in \cite{BrasselFischerHanusReck11}
+We have shown in \cite{BrasselHanusPeemoellerReck11}
 that this implementation can compete with or outperform
@@
 \begin{curry}
-  xor True  False = True
-  xor True  True  = False
-  xor False x     = x
+  xor True  x = not x             not True  = False
+  xor False x = x                 not False = True
 
@@
   xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
-                 $\to~$  xor True False      $\to~$  True
+                 $\to~$  xor True False      $\to~$  not False        $\to~$ True
 \end{haskell}
@@
 More details about this translation scheme can be found in
-\cite{BrasselFischer08IFL,BrasselFischerHanusReck11}.
+\cite{BrasselFischer08IFL,BrasselHanusPeemoellerReck11}.
 
@@
 which is avoided in KiCS2 by introducing an intermediate operation,
-see \cite{BrasselFischerHanusReck11}) and evaluate the main expression
+see \cite{BrasselHanusPeemoellerReck11}) and evaluate the main expression
 \code{(main 1)},
@@
   printValsDFS :: a -> IO ()
-  printValsDFS Fail             = return ()
+  printValsDFS Fail             = return ()$\listline$
   printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
@@
     follow NoDecision  = do newDecision ChooseLeft  x1
-                            newDecision ChooseRight x2
-
+                            newDecision ChooseRight x2$\listline$
     newDecision d x = do setDecision i d
                          printValsDFS x
-                         setDecision i NoDecision
+                         setDecision i NoDecision$\listline$
   printValsDFS v = print v
@@
 is nearly identical to a corresponding functional program.
-Actually, the benchmarks presented in \cite{BrasselFischerHanusReck11}
+Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
 show that this implementation outperforms all other Curry implementations
@@
 KiCS2 translates occurrences of logic variables
-into generators. For instance, consider the operation \code{not}
-defined by the rules:
-\begin{curry}
-  not True  = False
-  not False = True
-\end{curry}
-Then the expression \ccode{not x}, where \code{x} is a logic variable,
+into generators.
+For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
 is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
@@
                   in a symbolic arithmetic expression),
-\code{half} (compute the half of a peano number using logic variables),
+\code{half} (compute the half of a Peano number using logic variables),
+\code{palindrome} (check whether a list is a palindrome),
+\code{horseman} (solving an equation relating heads and feet of horses and men
+based on Peano numbers),
 and
-\code{palindrome} (check whether a list is a palindrome).
+\code{grep} (string matching based on a non-deterministic
+specification of regular expressions \cite{AntoyHanus10CACM}).
 
@@
 \hline
-Expression                              & \code{==~}& \code{=:=} & \code{=:<=} \\
+Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
 \hline
 \code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
-% \code{grep}                             & 1.06      & 0.19      &             \\
 \code{simplify}                         & 8.50      & 7.15      & 7.34        \\
@@
 \code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
-% \code{horseman}                         & 3.58      & 3.13      &             \\
+\code{horseman}                         & 3.58      & 3.13      & n/a         \\
+\code{grep}                             & 1.06      & 0.19      & n/a         \\
 \hline
@@
 \code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
-% \code{grep}                             & 0.19  & 2.88   & 0.14 \\
 \code{simplify}                         & 7.15  & 0.15   & 0.00 \\
@@
 \code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
-% \code{horseman}                         & 3.13  & 8.70   & 0.42 \\
+\code{horseman}                         & 3.13  & 8.70   & 0.42 \\
+\code{grep}                             & 0.19  & 2.88   & 0.14 \\
 \hline
@@
 of the kernel implementation as shown in
-\cite{BrasselFischerHanusReck11}.
+\cite{BrasselHanusPeemoellerReck11}.
 However, if these features are used, they usually lead

### 2011-07-12 papers/wlp2011/paper.tex
WLP paper: Refinements of section 1 to 3

@@
 
- \item compilation into logic languages (Prolog) and reusing
+ \item compilation into logic languages like Prolog and reusing
 the existing backtracking implementation for non-deterministic
@@
 that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
-or computing with generators by rewriting
+and computing with generators by rewriting
 are equivalent, i.e., compute the same values.
@@
 Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
-i.e., patterns containing evaluable operations providing
-for more powerful pattern matching than in logic or functional languages,
+i.e., patterns containing evaluable operations for more powerful pattern 
+matching than in logic or functional languages,
 can be implemented in this framework.
@@
 In the next section, we review the source language Curry
-and the features that are considered in this paper.
+and the features considered in this paper.
 Section~\ref{sec:Compilation} sketches the implementation scheme
@@
 logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
-\label{ctc-semantics}
+\label{ctc-semantics},
 where values of the arguments of an operation are determined before the
@@
 
-The condition $c$ in rule (\ref{rule}) is typically a
+The condition $c$ in rule (\ref{rule}) typically is a
 conjunction of \emph{equational constraints}
@@
 \cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
-defined operations. For instance, we can define the last element
-of a list also by:
+defined operations. For instance, we can also define the last element
+of a list by:
 %
@@
 of \code{(xs++[e])} is any list having \code{e} as its last element.
-Functional patterns are
-a powerful feature to express arbitrary selections in term structures.
-For instance, functional patterns support a straightforward
+Functional patterns are a powerful feature to express arbitrary 
+selections in term structures. For instance, they support a straightforward
 processing of XML data with incompletely specified
@@
 to each \code{Choice}. The difficulty is to get a unique identifier
-when needed, i.e., when some operation evaluates to a \code{Choice}.
+on demand, i.e., when some operation evaluates to a \code{Choice}.
 Since we want to compile into \emph{purely} functional programs
@@
 also called \emph{identifier supply},
-to each operation so that a \code{Choice} can pick an identifier from
+to each operation so that a \code{Choice} can pick its unique identifier from
 this set.
@@
 There are different implementations available \cite{AugustssonRittriSynek94}
-(see below for a simple implementation) and our implementation
+(see below for a simple implementation) and our system
 is parametric over concrete implementations of \code{IDSupply}.
@@
 \end{haskell}
-Thus, the result is non-deterministic and contains several choices,
-but all of them have the same identifier.
-This identifier is used when the values contained in this result
-are searched for: one has to make \emph{consistent} selections in choices with
-same identifiers. Thus, if we select the left branch as the value
+Thus, the result is non-deterministic and contains three choices,
+whereby all of them have the same identifier.
+To extract all values from such a \code{Choice} structure,
+we have to traverse it and  compute all possible choices,
+but consider the choice identifiers to make consistent (left/right)
+decisions.
+Thus, if we select the left branch as the value
 of the outermost \code{Choice}, we also have to select the left branch
-in the selected argument \code{(Choice 2 False True)} so that only
-the value \code{False} is possible here.
+in the selected argument \code{(Choice 2 False True)} so that \code{False}
+is the only value possible for this branch.
 Similarly, if we select the right branch as the value of the outermost
 \code{Choice}, we also have to select the right branch in
-its selected argument \code{(Choice 2 True False)} which yields the only
-value \code{False}.
-Thus, the unintended value \code{True} is not produced.
-
-In general, to extract all values from a \code{Choice} structure,
-we have to traverse it, compute all possible choices
-but consider the choice identifiers to make consistent (left/right)
-decisions.
-The latter requirement can be implemented by storing
+its selected argument \code{(Choice 2 True False)}, which again yields 
+\code{False} as the only possible value. In consequence, 
+the unintended value \code{True} is not produced.
+
+The requirement to make consistent decisions can be implemented by storing
 the decisions already made for some choices during the traversal.
@@
 \end{haskell}
-where \code{NoDecision} represents the fact that the value of a choice
-is yet undecided.
-Furthermore, we assume operations to lookup the current decision
-for a given identifier or change its decision (depending on the implementation
-of \code{IDSupply}, KiCS2 supports various implementations
+where \code{NoDecision} represents the fact that the value of a choice has
+not been decided yet. Furthermore, we assume operations to lookup the 
+current decision for a given identifier or change it
+(depending on the implementation
+of \code{IDSupply}, KiCS2 supports several implementations
 based on memory cells or finite maps):
@@
 \end{haskell}
-This operation ignores failures and prints values that are not a \code{Choice}.
-For a \code{Choice}, it checks whether a decision for
+This operation ignores failures and prints values that are not rooted by a
+\code{Choice} constructor. For a \code{Choice} constructor, 
+it checks whether a decision for
 this identifier has already been made (note that the initial value
@@
 Otherwise, the left alternative is used and this decision is stored.
-After printing
-all the values w.r.t.\ this decision,
+After printing all values w.r.t.\ this decision,
 the decision is undone (like in backtracking)
@@
 compute an initial identifier supply passed to this expression).
-The normal form computation is necessary for structured data like lists
+The normal form computation is necessary for structured data like lists,
 so that a failure or choice in some part of the data is moved to
-the root. Other search strategies, like
+the root.
+
+Other search strategies, like
 breadth-first search, iterative deepening, or parallel search,
 can be obtained by different implementations of this main operation
-to print all values. Instead of printing all values,
-one can also collect all values in a tree-like data structure
+to print all values.
+Furthermore, one can also collect all values in a tree-like data structure
 so that the programmer can implement his own search strategies
 (this corresponds to encapsulating search \cite{BrasselHanusHuch04JFLP}).
-Finally, one can also have other options rather than printing all values
-of the main expression.
-For instance, one can easily define operations
+Finally, instead of printing all values, one can easily define operations
 to print either the first solution only or one by one upon user request.
@@
 \code{Choice} constructors, our compiler performs an optimization for
-deterministic operations. If an operation does not call, 
-neither directly nor indirectly through
-other operations, the choice operation \ccode{?},\footnote{%
-Note that before applying this compilation scheme, overlapping rules in
-user-defined operations are transformed into calls of the choice operation
-\ccode{?}.}
-then it is not necessary to pass an identifier supply to the operation.
-In consequence, only the matching rules are extended so that the generated code
+deterministic operations. If an operation is defined by non-overlapping
+rules and does not call, neither directly nor indirectly through
+other operations, a function defined by overlapping rules, it does not
+introduce non-determinism (like \code{xor} or \code{not}).
+Thus, it is not necessary to pass an identifier supply to the operation.
+In consequence, only the matching rules are extended, so that the generated code
 is nearly identical to a corresponding functional program.
 Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
-show that this implementation outperforms all other Curry implementations
-for deterministic operations,
+show that for deterministic operations this implementation outperforms 
+all other Curry implementations,
 and, for non-deterministic operations, outperforms Prolog-based
@@
 
-As mentioned in the introduction,
-KiCS2 translates occurrences of logic variables
-into generators.
+As mentioned in the introduction, occurrences of logic variables are
+translated into generators.
 For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
-is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
-here). The latter expression is evaluated by reducing the
-argument \code{aBool} to a choice between \code{True} or \code{False}
+is translated into \ccode{not (aBool s)}, where s is an \code{IDSupply}
+provided by the context of the expression. The latter expression is 
+evaluated by reducing the
+argument \code{aBool s} to a choice between \code{True} or \code{False}
 followed by applying \code{not} to this choice.
@@
 non-deterministic operations, they are translated like any other operation
-so that one does not need additional run-time support for them.
-However, if equational constraints should be solved,
+and therefore do not require any additional run-time support.
+However, in the presence of equational constraints,
 there are methods which are more efficient than generating all values.

### 2011-07-13 papers/wlp2011/paper.tex
WLP paper: section 4..7 refined

@@
 Functional patterns are a powerful feature to express arbitrary 
-selections in term structures. For instance, they support a straightforward
+selections in term structures.
+For instance, they support a straightforward
 processing of XML data with incompletely specified
@@
 can be considered as a binary function defined by the
-following rules (here we show only the rules for the Boolean
+following rules (we only present the rules for the Boolean
 and list types, where \code{Success} denotes the only constructor
@@
 creates a search space when evaluating its first argument,
-although there is no search required since there is only one possible
-binding for \code{x}.
-Furthermore and even worse, there is an infinite search space
-when unifying two logic variables. For instance,
+although there is no search required since there is only one
+binding for \code{x} satisfying the constraint.
+Furthermore and even worse, unifying two logic variables
+introduces an infinite search space. For instance,
 the expression \ccode{xs =:= ys \& xs++ys =:= [True]}
@@
 occurring in an equational constraint,
-we \emph{bind} them to another variable or term.
+we \emph{bind} the variables to another variable or term.
 Since we compile into a purely functional language,
 the binding cannot be performed by some side effect.
-Instead, we add binding constraints to the computed results.
+Instead, we add binding constraints to the computed results to be 
+processed during search.
 
 To implement unification, we have to distinguish free variables
-from ``standard choices'' (introduced by the operation \ccode{?})
+from ``standard choices'' (introduced by overlapping rules)
 in the target code. For this purpose, we refine the definition
 of the type \code{ID} as follows:\footnote{For the sake
-of simplicity, we consider in the following the
-implementation of \code{IDSupply} as unbounded integers.}
+of simplicity, in the following, we consider the
+implementation of \code{IDSupply} to be unbounded integers.}
 \begin{haskell}
@@
 \end{haskell}
-The new constructor \code{FreeID} marks a choice corresponding to
+The new constructor \code{FreeID} identifies a choice corresponding to
 a free variable, e.g., the generator for Boolean variables is
@@
 \end{haskell}
-With this small change, the run-time system can distinguish
-free variables (choices with an identifier of the form \code{(FreeID $\ldots$)})
-from other values. This is quite useful at the top-level
-to print values containing logic variables in a specific form
-rather than enumerating all values, similarly to logic programming systems.
-For instance, KiCS2 evaluates the application of \code{head}
-to an unknown list as follows:
-\begin{haskell}
-  Prelude> head xs where xs free
-  {xs = (_x2:_x3)} _x2
-\end{haskell}
-Here, free variables are marked by the prefix \code{\us{}x}.
 
@@
 a binding constraint as a pair of a choice identifier
-and a decision for this choice:
+and a decision for this identifier:
 \begin{haskell}
@@
 Furthermore, we extend each data type by the possibility to
-add such a constraint:
-\begin{haskell}
-  data Bool   = $\ldots$ | Guard Constraint Bool
-  data List a = $\ldots$ | Guard Constraint (List a)
-\end{haskell}
-Thus, \code{(Guard $c$ $v$)} represents a \emph{constrained value},
-i.e., the value $v$ is only valid if the constraint $c$ is
+add constraints:
+\begin{haskell}
+  data Bool   = $\ldots$ | Guard [Constraint] Bool
+  data List a = $\ldots$ | Guard [Constraint] (List a)
+\end{haskell}
+Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
+i.e., the value $v$ is only valid if the constraints $cs$ are
 consistent with the decisions previously made during search.
-The binding constraints are created by the
+These binding constraints are created by the
 equational constraint operation \ccode{=:=}: if a free variable
 should be bound to a constructor, we make the same
-decision as it would be done in the successful branch
+decisions as it would be done in the successful branch
 of the generator. In case of Boolean values,
@@
 \begin{haskell}
-  Choice (FreeID i) _ _ =:= True   =  Guard (i :=: ChooseLeft ) Success
-  Choice (FreeID i) _ _ =:= False  =  Guard (i :=: ChooseRight) Success
+  Choice (FreeID i) _ _ =:= True   =  Guard [i :=: ChooseLeft ] Success
+  Choice (FreeID i) _ _ =:= False  =  Guard [i :=: ChooseRight] Success
 \end{haskell}
@@
   $\ldots$
-  printValsDFS (Guard c x) = do consistent <- add c
+  printValsDFS (Guard cs x) = do consistent <- add c
                                  if consistent then do printValsDFS x
@@
 
-Due to the extensions described in this section,
-we obtain an implementation of a unification procedure
-that integrates the demand-driven evaluation of operations
-occurring in the arguments.
-Benchmarks evaluating the efficiency of this approach are
-shown in Section~\ref{sec:Benchmarks}.
+Finally, with the ability to distinguish free variables 
+(choices with an identifier of the form \code{(FreeID $\ldots$)})
+from other values during search, values containing logic variables 
+can also be printed in a specific form rather than enumerating all values, 
+like in logic programming systems. For instance, KiCS2 evaluates the 
+application of \code{head} to an unknown list as follows:
+\begin{haskell}
+  Prelude> head xs where xs free
+  {xs = (_x2:_x3)} _x2
+\end{haskell}
+Here, free variables are marked by the prefix \code{\us{}x}.
+
+% Due to the extensions described in this section,
+% we obtain an implementation of a unification procedure
+% that integrates the demand-driven evaluation of operations
+% occurring in the arguments.
+% Benchmarks evaluating the efficiency of this approach are
+% shown in Section~\ref{sec:Benchmarks}.
 
@@
 requires that \code{ys++[e]} as well as \code{xs}
-must be reducible to identical values so that the input list \code{xs}
-is completely evaluated.
+must be reducible to unifiable terms so that in consequence
+the input list \code{xs} is completely evaluated.
 Hence, if \code{failed} denotes an operation whose evaluation fails,
@@
 the value \code{True}, i.e., the definition of \code{last'} is less strict.
-This is due to the fact that a functional pattern like
-\code{(xs++[e])} abbreviates all values to which it can be evaluated
-(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
-Conceptually, the rule defining \code{last'}
-abbreviates the following (infinite) set of rules:
-\begin{curry}
-  last' [e] = e
-  last' [x1,e] = e
-  last' [x1,x2,e] = e
-  $\ldots$
-\end{curry}
-As a further example,
-consider an operation that returns the first duplicated element in
+
+As another example for the advantage of the reduced strictness,
+consider an operation that returns the first duplicate element in
 a list.
@@
 %
-Functional patterns are quite useful in situations
-where one wants to select parts of some structure
-at arbitrary positions. For instance,
-functional patterns have been applied to support
-the high-level programming of querying or transforming
-XML data \cite{Hanus11ICLP}.
+Because of the reduced strictness, the logic variable \code{zs} (matching
+the tail list after the first duplicate) is never evaluated.
+This is due to the fact that a functional pattern like
+\code{(xs++[e])} abbreviates all values to which it can be evaluated
+(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
+Conceptually, the rule defining \code{last'}
+abbreviates the following (infinite) set of rules:
+\begin{curry}
+  last' [e] = e
+  last' [x1,e] = e
+  last' [x1,x2,e] = e
+  $\ldots$
+\end{curry}
 
@@
 \end{curry}
-The behavior of \ccode{=:<=} is similar to \ccode{=:=}
+The behavior of \ccode{=:<=} is similar to \ccode{=:=},
 except for the case that a variable in the left argument
@@
 
-The following discussion shows the requirements on an 
-implementation of functional patterns.
-The general structure
-is quite similar to equational constraints
+The general structure of the implementation of functional patterns in KiCS2
+is quite similar to that of equational constraints,
 with the exception that variables could be also bound
-to unevaluated expressions. Only if such variables are later
+to unevaluated expressions. 
+Only if such variables are later
 accessed, the expressions they are bound to are evaluated.
@@
 \begin{haskell}
-  data Decision = $\ldots$ | LazyBind Constraint
+  data Decision = $\ldots$ | LazyBind [Constraint]
 \end{haskell}
@@
   Choice (FreeID i) _ _ =:<= x
-    = Guard (i :=: LazyBind (lazyBind i x)) Success
+    = Guard [i :=: LazyBind (lazyBind i x)] Success
 \end{haskell}
@@
 \begin{haskell}
-  lazyBind :: ID -> a -> Constraint
-  lazyBind i True  = i :=: ChooseLeft
-  lazyBind i False = i :=: ChooseRight
+  lazyBind :: ID -> a -> [Constraint]
+  lazyBind i True  = i :=: [ChooseLeft]
+  lazyBind i False = i :=: [ChooseRight]
 \end{haskell}
@@
 \code{xs++[e] =:<= [failed,True]} that occurs when the expression
-\code{last' [failed,True]} is evaluated (here we omit failed branches
+\code{last' [failed,True]} is evaluated (we omit failed branches
 and some other details; note that logic variables are replaced
@@
 \begin{haskell}
-  aBoolList 2 ++ [aBool 3] =:<= [failed,True]
-  $\leadsto$ [aBool 4,aBool 3] =:<= [failed,True]
+     aBoolList 2 ++ [aBool 3] =:<= [failed, True]
+  $\leadsto$ [aBool 4, aBool 3] =:<= [failed, True]
   $\leadsto$ aBool 4 =:<= failed & aBool 3 =:<= True & [] =:<= []
-  $\leadsto$ Guard (4 :=: LazyBind (lazyBind 4 failed))
-           (Guard (3 :=: LazyBind (lazyBind 3 True)) Success)
-\end{haskell}
-If the binding of identifier \code{3} is later required,
-\code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
+  $\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed)
+           , 3 :=: LazyBind (lazyBind 3 True)] Success
+\end{haskell}
+If the value of the expression \code{last' [failed,True]} is later required,
+the value of the variable \code{e} (with the identifier \code{3}) is in turn
+required.
+Thus, \code{(lazyBind 3 True)} is evaluated to \code{[3 :=: ChooseLeft]}
 which corresponds to the value \code{True} of the generator \code{(aBool 3)}.
 Note that the variable with identifier \code{4} does not occur
-anywhere else so that the binding \code{(lazyBind 4 failed)}
+anywhere else, so that the binding \code{(lazyBind 4 failed)}
 will never be evaluated, as intended.
@@
 
-In this section we evaluate our implementation by some benchmarks.
-They were executed on a Linux machine running Debian 5.0.7 with
+In this section we evaluate our implementation of
+equational constraints and functional patterns by some benchmarks.
+The benchmarks were executed on a Linux machine running Debian 5.0.7 with
 an Intel Core 2 Duo (3.0GHz) processor.
-KiCS2 has been executed with
-the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) and uses
+KiCS2 has been used with
+the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) as its backend and
 an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
@@
 we considered PAKCS \cite{Hanus10PAKCS}
-(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
-and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
+(version 1.9.2, based on a SICStus-Prolog 4.1.2)
+and MCC \cite{Lux99FLOPS} (version 0.9.10). The timings were performed with the
 time command measuring the execution time (in seconds) of a compiled
@@
 \end{tabular}
-\caption{Benchmarks: comparing different equation solving methods}
+\caption{Benchmarks: comparing different representations for equations}
  \label{fig:search-strategies}
@@
 of equational constraints with variable bindings,
-we compare in Fig.~\ref{fig:search-strategies}
+in Fig.~\ref{fig:search-strategies} we compare
 the results of using equational constraints (\code{=:=})
@@
 does not perform bindings but enumerate all values).
-As expected, the creation and traversal of a large search space introduced
-by \code{==} is in most cases much slower than our presented approach
-with variable bindings.
+As expected, in most cases the creation and traversal of a large
+search space introduced by \code{==} is much slower than our
+presented approach with variable bindings.
 In addition, the example \code{last} shows that the lazy 
-unification operator (\code{=:<=}) improves performance when unifying an
+unification operator (\ccode{=:<=}) improves the performance when unifying an
 expression which has to be evaluted only partially. Using strict unification,
@@
 \end{tabular}
-\caption{Benchmarks: equational constraints in different Curry implementations}
+\caption{Benchmarks: strict unification in different Curry implementations}
  \label{fig:unification}
@@
 
-In contrast to the Curry implementations PAKCS and MCC, our approach is based
+In contrast to the Curry implementations PAKCS and MCC, our approach 
+of strict unification is based
 on an explicit representation of the search space instead of
@@
 a global state containing bindings for logic variables.
-Nevertheless, the benchmarks in Fig.~\ref{fig:unification}
-show that it can compete with or even outperform these implementations.
-In these benchmarks only equational constraints are used.
+Nevertheless, the benchmarks in Fig.~\ref{fig:unification},
+using equational constraints only,
+show that it can compete with or even outperform the other implementations.
 The results show that the implementation of unification
-of MCC performs best. However, in most cases
-our implementation outperforms the Prolog-based PAKCS implementation,
+of MCC performs best. However, in most cases our implementation 
+outperforms the Prolog-based PAKCS implementation, 
 except for some examples where further investigations
@@
 \end{tabular}
-\caption{Benchmarks: comparing different implementations of functional patterns}
+\caption{Benchmarks: functional patterns in different Curry implementations}
  \label{fig:functional-patterns}
@@
 We have presented an implementation of equational constraints
-and functional patterns in a purely functional implementation
+and functional patterns in KiCS2, a purely functional implementation
 of Curry. Our implementation is based on adding binding constraints
@@
 at the top level of a computation.
-Since we have added only new constructors and rules
-for them in our implementation, it does not introduce
-additional work if these features are not used, i.e.,
+Since we only have added new constructors and pattern matching rules 
+for them in our implementation, no overhead is introduced for programs
+without equational constraints, i.e.,
 our implementation does not sacrifice the high efficiency
-of the kernel implementation as shown in
+of the kernel implementation shown in
 \cite{BrasselHanusPeemoellerReck11}.
 However, if these features are used, they usually lead
-to a more efficient execution, as demonstrated by our benchmarks.
+to a comparably efficient execution, as demonstrated by our benchmarks.
 
@@
 For future work it might be interesting to add
-further constraint structures, like real arithmetic
-or finite domain constraints, to our implementation.
+further constraint structures to our implementation, like real arithmetic
+or finite domain constraints.
 This might be possible by extending the kinds of constraints

### 2011-07-13 papers/wlp2011/paper.tex
Small corrections to paper

@@
 For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
-is translated into \ccode{not (aBool s)}, where s is an \code{IDSupply}
+is translated into \ccode{not (aBool s)}, where \code{s} is an \code{IDSupply}
 provided by the context of the expression. The latter expression is 
@@
 non-deterministic operations, they are translated like any other operation
-and therefore do not require any additional run-time support.
+and, therefore, do not require any additional run-time support.
 However, in the presence of equational constraints,
@@
 Instead, we add binding constraints to the computed results to be 
-processed during search.
+processed by a search strategy that extract values from choice structures.
 
@@
   $\ldots$
-  printValsDFS (Guard cs x) = do consistent <- add c
+  printValsDFS (Guard cs x) = do consistent <- add cs
                                  if consistent then do printValsDFS x
-                                                       remove c
+                                                       remove cs
                                                else return ()
@@
 \end{haskell}
-The operation \code{add} checks the consistency of the constraint \code{c}
+The operation \code{add} checks the consistency of the constraint \code{cs}
 with the decisions made so far and, in case of consistency,
@@
 can also be printed in a specific form rather than enumerating all values, 
-like in logic programming systems. For instance, KiCS2 evaluates the 
+similarly to logic programming systems. For instance, KiCS2 evaluates the 
 application of \code{head} to an unknown list as follows:
@@
   lazyBind :: ID -> a -> [Constraint]
-  lazyBind i True  = i :=: [ChooseLeft]
-  lazyBind i False = i :=: [ChooseRight]
+  lazyBind i True  = [i :=: ChooseLeft]
+  lazyBind i False = [i :=: ChooseRight]
 \end{haskell}
@@
 
-In contrast to the Curry implementations PAKCS and MCC, our approach 
-of strict unification is based
+In contrast to the Curry implementations PAKCS and MCC,
+our implementation of strict unification is based
 on an explicit representation of the search space instead of

### 2011-07-13 papers/wlp2011/paper.tex
changed benchmarks to prdfs

--- 2011-07-12 papers/wlp2011/paper.tex
@@
 
- \item compilation into logic languages like Prolog and reusing
+ \item compilation into logic languages (Prolog) and reusing
 the existing backtracking implementation for non-deterministic
@@
 that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
-and computing with generators by rewriting
+or computing with generators by rewriting
 are equivalent, i.e., compute the same values.
@@
 Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
-i.e., patterns containing evaluable operations for more powerful pattern 
-matching than in logic or functional languages,
+i.e., patterns containing evaluable operations providing
+for more powerful pattern matching than in logic or functional languages,
 can be implemented in this framework.
@@
 In the next section, we review the source language Curry
-and the features considered in this paper.
+and the features that are considered in this paper.
 Section~\ref{sec:Compilation} sketches the implementation scheme
@@
 logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
-\label{ctc-semantics},
+\label{ctc-semantics}
 where values of the arguments of an operation are determined before the
@@
 
-The condition $c$ in rule (\ref{rule}) typically is a
+The condition $c$ in rule (\ref{rule}) is typically a
 conjunction of \emph{equational constraints}
@@
 \cite{AntoyHanus05LOPSTR}, i.e., $t_i$ might additionally contain calls to
-defined operations. For instance, we can also define the last element
-of a list by:
+defined operations. For instance, we can define the last element
+of a list also by:
 %
@@
 of \code{(xs++[e])} is any list having \code{e} as its last element.
-Functional patterns are a powerful feature to express arbitrary 
-selections in term structures. For instance, they support a straightforward
+Functional patterns are
+a powerful feature to express arbitrary selections in term structures.
+For instance, functional patterns support a straightforward
 processing of XML data with incompletely specified
@@
 to each \code{Choice}. The difficulty is to get a unique identifier
-on demand, i.e., when some operation evaluates to a \code{Choice}.
+when needed, i.e., when some operation evaluates to a \code{Choice}.
 Since we want to compile into \emph{purely} functional programs
@@
 also called \emph{identifier supply},
-to each operation so that a \code{Choice} can pick its unique identifier from
+to each operation so that a \code{Choice} can pick an identifier from
 this set.
@@
 There are different implementations available \cite{AugustssonRittriSynek94}
-(see below for a simple implementation) and our system
+(see below for a simple implementation) and our implementation
 is parametric over concrete implementations of \code{IDSupply}.
@@
 \end{haskell}
-Thus, the result is non-deterministic and contains three choices,
-whereby all of them have the same identifier.
-To extract all values from such a \code{Choice} structure,
-we have to traverse it and  compute all possible choices,
+Thus, the result is non-deterministic and contains several choices,
+but all of them have the same identifier.
+This identifier is used when the values contained in this result
+are searched for: one has to make \emph{consistent} selections in choices with
+same identifiers. Thus, if we select the left branch as the value
+of the outermost \code{Choice}, we also have to select the left branch
+in the selected argument \code{(Choice 2 False True)} so that only
+the value \code{False} is possible here.
+Similarly, if we select the right branch as the value of the outermost
+\code{Choice}, we also have to select the right branch in
+its selected argument \code{(Choice 2 True False)} which yields the only
+value \code{False}.
+Thus, the unintended value \code{True} is not produced.
+
+In general, to extract all values from a \code{Choice} structure,
+we have to traverse it, compute all possible choices
 but consider the choice identifiers to make consistent (left/right)
 decisions.
-Thus, if we select the left branch as the value
-of the outermost \code{Choice}, we also have to select the left branch
-in the selected argument \code{(Choice 2 False True)} so that \code{False}
-is the only value possible for this branch.
-Similarly, if we select the right branch as the value of the outermost
-\code{Choice}, we also have to select the right branch in
-its selected argument \code{(Choice 2 True False)}, which again yields 
-\code{False} as the only possible value. In consequence, 
-the unintended value \code{True} is not produced.
-
-The requirement to make consistent decisions can be implemented by storing
+The latter requirement can be implemented by storing
 the decisions already made for some choices during the traversal.
@@
 \end{haskell}
-where \code{NoDecision} represents the fact that the value of a choice has
-not been decided yet. Furthermore, we assume operations to lookup the 
-current decision for a given identifier or change it
-(depending on the implementation
-of \code{IDSupply}, KiCS2 supports several implementations
+where \code{NoDecision} represents the fact that the value of a choice
+is yet undecided.
+Furthermore, we assume operations to lookup the current decision
+for a given identifier or change its decision (depending on the implementation
+of \code{IDSupply}, KiCS2 supports various implementations
 based on memory cells or finite maps):
@@
 \end{haskell}
-This operation ignores failures and prints values that are not rooted by a
-\code{Choice} constructor. For a \code{Choice} constructor, 
-it checks whether a decision for
+This operation ignores failures and prints values that are not a \code{Choice}.
+For a \code{Choice}, it checks whether a decision for
 this identifier has already been made (note that the initial value
@@
 Otherwise, the left alternative is used and this decision is stored.
-After printing all values w.r.t.\ this decision,
+After printing
+all the values w.r.t.\ this decision,
 the decision is undone (like in backtracking)
@@
 compute an initial identifier supply passed to this expression).
-The normal form computation is necessary for structured data like lists,
+The normal form computation is necessary for structured data like lists
 so that a failure or choice in some part of the data is moved to
-the root.
-
-Other search strategies, like
+the root. Other search strategies, like
 breadth-first search, iterative deepening, or parallel search,
 can be obtained by different implementations of this main operation
-to print all values.
-Furthermore, one can also collect all values in a tree-like data structure
+to print all values. Instead of printing all values,
+one can also collect all values in a tree-like data structure
 so that the programmer can implement his own search strategies
 (this corresponds to encapsulating search \cite{BrasselHanusHuch04JFLP}).
-Finally, instead of printing all values, one can easily define operations
+Finally, one can also have other options rather than printing all values
+of the main expression.
+For instance, one can easily define operations
 to print either the first solution only or one by one upon user request.
@@
 \code{Choice} constructors, our compiler performs an optimization for
-deterministic operations. If an operation is defined by non-overlapping
-rules and does not call, neither directly nor indirectly through
-other operations, a function defined by overlapping rules, it does not
-introduce non-determinism (like \code{xor} or \code{not}).
-Thus, it is not necessary to pass an identifier supply to the operation.
-In consequence, only the matching rules are extended, so that the generated code
+deterministic operations. If an operation does not call, 
+neither directly nor indirectly through
+other operations, the choice operation \ccode{?},\footnote{%
+Note that before applying this compilation scheme, overlapping rules in
+user-defined operations are transformed into calls of the choice operation
+\ccode{?}.}
+then it is not necessary to pass an identifier supply to the operation.
+In consequence, only the matching rules are extended so that the generated code
 is nearly identical to a corresponding functional program.
 Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
-show that for deterministic operations this implementation outperforms 
-all other Curry implementations,
+show that this implementation outperforms all other Curry implementations
+for deterministic operations,
 and, for non-deterministic operations, outperforms Prolog-based
@@
 
-As mentioned in the introduction, occurrences of logic variables are
-translated into generators.
+As mentioned in the introduction,
+KiCS2 translates occurrences of logic variables
+into generators.
 For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
-is translated into \ccode{not (aBool s)}, where s is an \code{IDSupply}
-provided by the context of the expression. The latter expression is 
-evaluated by reducing the
-argument \code{aBool s} to a choice between \code{True} or \code{False}
+is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
+here). The latter expression is evaluated by reducing the
+argument \code{aBool} to a choice between \code{True} or \code{False}
 followed by applying \code{not} to this choice.
@@
 non-deterministic operations, they are translated like any other operation
-and therefore do not require any additional run-time support.
-However, in the presence of equational constraints,
+so that one does not need additional run-time support for them.
+However, if equational constraints should be solved,
 there are methods which are more efficient than generating all values.
@@
 \hline
-\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
-\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
-\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
-% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
-% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
-\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
-\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
-\code{horseman}                         & 3.58      & 3.13      & n/a         \\
-\code{grep}                             & 1.06      & 0.19      & n/a         \\
+\code{last (map (inc 0) [1..100000])}   &  2.91   &    0.05   &    0.01   \\
+\code{simplify}                         & 10.30      & 6.77      & 7.07        \\
+\code{varInExp}                         & 2.34    & 0.24      & 0.21        \\
+% \code{varInExp2}                        & 0.34   & 0.02     & 0.01        \\
+% \code{varInExp3}                        & 13.28    & 1.36     & 1.02       \\
+\code{fromPeano (half (toPeano 10000))} & 26.67     & 5.95      & 11.19       \\
+\code{palindrome}                       & 30.86     & 14.05     & 20.26       \\
+\code{horseman}                         & 3.24      & 3.31      & n/a         \\
+\code{grep}                             & 1.06      & 0.10      & n/a         \\
 \hline
@@
 \hline
-\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
-\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
-\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
-% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
-% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
-\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
-\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
-\code{horseman}                         & 3.13  & 8.70   & 0.42 \\
-\code{grep}                             & 0.19  & 2.88   & 0.14 \\
+\code{last (map (inc 0) [1..1000])}   & 0.05  & 0.40   &  0.01\\
+\code{simplify}                         & 6.77  & 0.15   & 0.00 \\
+\code{varInExp}                         & 0.24  & 0.89   & 0.07 \\
+% \code{varInExp2}                        & 0.02 & 0.06   & 0.00 \\
+% \code{varInExp3}                        & 1.36 & 2.07  & 0.14 \\
+\code{fromPeano (half (toPeano 10000))} & 5.95  & 108.88 & 3.22 \\
+\code{palindrome}                       & 14.05 & 32.56  & 1.07 \\
+\code{horseman}                         & 3.31  & 8.70   & 0.42 \\
+\code{grep}                             & 0.10  & 2.88   & 0.14 \\
 \hline
@@
 \hline
-\code{last (map (inc 0) [1..100000])}   & 0.26  & 3.97     \\
-\code{simplify}                         & 7.34  & 0.27     \\
-\code{varInExp}                         & 1.40  & 8.88     \\
-% \code{varInExp2}                         & 1.04  & 0.08    \\
-% \code{varInExp3}                         & 11.28 & 16.62   \\
-\code{fromPeano (half (toPeano 10000))} & 11.88 & $\infty$ \\
-\code{palindrome}                       & 23.21 & $\infty$ \\
+\code{last (map (inc 0) [1..10000])}   &  0.01 & 0.33     \\
+\code{simplify}                         & 7.07  & 0.27     \\
+\code{varInExp}                         &  0.21  &  1.87    \\
+% \code{varInExp2}                         & 0.01  & 0.02    \\
+% \code{varInExp3}                         & 1.02 & 2.16   \\
+\code{fromPeano (half (toPeano 10000))} & 11.19 & $\infty$ \\
+\code{palindrome}                       & 20.26 & $\infty$ \\
 \hline

### 2011-07-13 papers/wlp2011/paper.tex
added explanations for [Constraint] and performance of simplify

--- 2011-07-13 papers/wlp2011/paper.tex
@@
 Furthermore, we extend each data type by the possibility to
-add constraints:
+add constraints:\footnote{A \code{Constraint} only provides the decision for
+                          one constructor. To support constraints for
+                          structured data, we need a list of \code{Constraint}s.}
 \begin{haskell}
@@
   Choice (FreeID i) _ _ =:= Choice (FreeID j) _ _
-    =  Guard (i :=: BindTo j) Success
+    =  Guard [i :=: BindTo j] Success
 \end{haskell}
@@
 \end{haskell}
-The operation \code{add} checks the consistency of the constraint \code{c}
+The operation \code{add} checks the consistency of the constraints \code{cs}
 with the decisions made so far and, in case of consistency,
-stores the decision made by the constraint.
+stores the decisions made by the constraints.
 In this case, the constrained value is evaluated
-before the constraint is removed to allow backtracking.
+before the constraints are removed to allow backtracking.
 Furthermore, the operations \code{lookupDecision} and \code{setDecision}
@@
 outperforms the Prolog-based PAKCS implementation, 
-except for some examples where further investigations
-and optimizations are necessary.
+except for some examples. Especially \code{simplify} performs very badly
+due to expensive bindings of free variables to large arithmetic expressions
+in unsuccessful branches of the search. Further investigation
+and optimization will hopefully lead to a better performance in
+such cases.
 

### 2011-07-14 papers/wlp2011/paper.tex
corrected error in benchmark table

@@
 \hline
-\code{last (map (inc 0) [1..100000])}   &  2.91   &    0.05   &    0.01   \\
+\code{last (map (inc 0) [1..10000])}   &  2.91   &    0.05   &    0.01   \\
 \code{simplify}                         & 10.30      & 6.77      & 7.07        \\
@@
 \hline
-\code{last (map (inc 0) [1..1000])}   & 0.05  & 0.40   &  0.01\\
+\code{last (map (inc 0) [1..10000])}   & 0.05  & 0.40   &  0.01\\
 \code{simplify}                         & 6.77  & 0.15   & 0.00 \\

### 2011-07-14 papers/wlp2011/paper.tex

@@
 Furthermore, we extend each data type by the possibility to
-add constraints:\footnote{A \code{Constraint} only provides the decision for
-                          one constructor. To support constraints for
-                          structured data, we need a list of \code{Constraint}s.}
+add constraints:\footnote{A single \code{Constraint} provides the decision for
+one constructor. To support constraints for structured data,
+we add a list of \code{Constraint}s to the data.}
 \begin{haskell}
@@
 outperforms the Prolog-based PAKCS implementation, 
-except for some examples. Especially \code{simplify} performs very badly
+except for some examples. In particular, \code{simplify} does not performs well
 due to expensive bindings of free variables to large arithmetic expressions

### 2011-07-15 runtime/Basics.hs
Some compiler warnings removed

@@ -156,4 +156,6 @@ d_apply :: (a -> b) -> a -> b
 d_apply f a = f a
 
 nd_apply :: NonDet b => Func a b -> a -> IDSupply -> b
-nd_apply fun a s = (\(Func f) -> f a s) `d_dollar_bang` fun
+nd_apply fun a s = apply `d_dollar_bang` fun
+  where apply (Func f) = f a s
+        apply _        = error "Basics.nd_apply.apply: no ground term"

### 2011-07-15 runtime/ID.hs

@@ -2,10 +2,13 @@
 -- ID module
 -- ---------------------------------------------------------------------------
 module ID
-  ( Constraint (..), Constraints, Choice (..), defaultChoice, isDefaultChoice
-  , ID (..), IDSupply
-  , mkInt, leftID, rightID, narrowID
-  , initSupply, leftSupply, rightSupply, thisID, freeID
+  ( -- * Constraints
+    Constraint (..), Constraints
+    -- * Choices
+  , Choice (..), defaultChoice, isDefaultChoice
+    -- * IDs
+  , ID (..), mkInt, leftID, rightID, narrowID
+  , IDSupply, initSupply, leftSupply, rightSupply, thisID, freeID
     -- * Choice management
   , lookupChoice, lookupID, lookupChoiceID, setChoice, setUnsetChoice
   , nextNIDs
@@ -60,12 +63,6 @@ data Choice
   | LazyBind Constraints
     deriving Show
 
--- TODO: Idea: Split ChooseN Int Int into two cases
---   - ChooseFree Int Int     for free variables and
---   - ChooseNarrowed Int     for narrowed variables
--- to further separate these constructs and avoid the use of irregular values
--- like ChooseN 0 (-1)
-
 
 instance Eq Choice where
   NoChoice    == NoChoice    = True
@@ -91,9 +88,9 @@ isDefaultChoice _        = False
 -- ID
 -- ---------------------------------------------------------------------------
 
--- Type to identify different Choice structures in a non-deterministic result
+-- |Type to identify different Choice structures in a non-deterministic result
 data ID
-    -- |Identifier for a choice introduced by the use of the (?) operator
+    -- |Identifier for a choice introduced by using of the (?) operator
   = ChoiceID Ref
     -- |Identifier for a choice for a free variable
   | FreeID [Int] IDSupply

### 2011-07-15 runtime/Search.hs

@@ -582,7 +582,8 @@ searchMPlus'' cont (Free i branches) = lookupChoice' i >>= choose
 searchMPlus'' cont  (Guard cs e) =
   solves' cs >> searchMPlus' cont e
 
-
+processLazyBind' :: (NormalForm a, MonadPlus m) => ID
+  -> (a -> StateT SetOfChoices m b) -> Constraints -> [a] -> StateT SetOfChoices m b
 processLazyBind' i cont cs branches = do
   setChoice' i NoChoice
   searchMPlus' cont (guardCons cs (choicesCons i branches))

### 2011-07-15 papers/wlp2011/paper.tex
Typo removed
small changes according to Bernds remarks

@@
 i.e., the equation \code{$e_1$\,=:=\,$e_2$} is satisfied iff
-$e_1$ and $e_2$ are reducible to the same ground constructor term.
+$e_1$ and $e_2$ are reducible to unifiable ground constructor terms.
 In particular, expressions that do not have a value
@@
 Instead, we add binding constraints to the computed results to be 
-processed by a search strategy that extract values from choice structures.
+processed by a search strategy that extracts values from choice structures.
 
@@
 outperforms the Prolog-based PAKCS implementation, 
-except for some examples. In particular, \code{simplify} does not performs well
+except for some examples. In particular, \code{simplify} does not perform well
 due to expensive bindings of free variables to large arithmetic expressions

### 2011-07-15 papers/wlp2011/paper.tex
remark about missing occur check added

@@
 Furthermore, we add the rule that an equational constraint
-between two variables yields a binding for these variables:
+between two variables yields a binding for these variables:%
+\footnote{An occur check is not performed for efficiency reasons.}
 \begin{haskell}

### 2011-07-18 papers/wlp2011/paper.tex

@@
 Furthermore, we extend each data type by the possibility to
-add constraints:\footnote{A single \code{Constraint} provides the decision for
-one constructor. To support constraints for structured data,
-we add a list of \code{Constraint}s to the data.}
+add constraints:
 \begin{haskell}
@@
 \end{haskell}
+A single \code{Constraint} provides the decision for
+one constructor. In order to support constraints for structured data,
+a list of \code{Constraint}s provides the decision for the outermost
+constructor and the decisions for all its arguments.
 Thus, \code{(Guard $cs$ $v$)} represents a \emph{constrained value},
@@
 between two variables yields a binding for these variables:%
-\footnote{An occur check is not performed for efficiency reasons.}
 \begin{haskell}

### 2011-08-29 papers/wlp2011/paper.tex

--- 2011-07-15 papers/wlp2011/paper.tex
@@
 
-\addtolength{\textheight}{0.2cm}
-\addtolength{\topmargin}{-0.1cm}
-\addtolength{\textwidth}{0.2cm}
-\addtolength{\oddsidemargin}{-0.1cm}
-\addtolength{\evensidemargin}{-0.1cm}
-
-\usepackage{times}
+\usepackage{a4,a4wide}
 \usepackage{url}
@@
 \usepackage{listings}
-\lstset{aboveskip=0.5ex,
-        belowskip=0.5ex,
+\lstset{aboveskip=0.8ex,
+        belowskip=0.8ex,
         showstringspaces=false, % no special string space
@@
 \newcommand{\us}{\char95\xspace} % underscore
-\newcommand{\dollar}{\code{\char36}} % dollar in programs
-\newcommand{\Choice}[2]{#1 \mathop{?} #2}
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
 
@@
 Benchmarks demonstrating the usefulness of this scheme
-are presented in Section~\ref{sec:Benchmarks}
-before we conclude in Section~\ref{sec:Conclusions}.
+are presented in Sect.~\ref{sec:Benchmarks}
+before we conclude in Sect.~\ref{sec:Conclusions}.
 
@@
 \begin{curry}
-  xor True  x = not x             not True  = False
-  xor False x = x                 not False = True
+  not True  = False
+  not False = True
+
+  xor True  x = not x
+  xor False x = x
 
@@
 Here, the functional pattern \code{(xs++[e])}
-states that \code{last'} is reducible
-to \code{e} provided that the argument can be matched against
+states that \code{(last' t)} is reducible
+to \code{e} provided that the argument \code{t} can be matched against
 some value of \code{(xs++[e])} where \code{xs} and \code{e} are
@@
 is used to implement the call-time choice semantics
-discussed in Section~\ref{sec:Curry}.
+discussed in Sect.~\ref{sec:Curry}.
 Since the evaluation of \code{xorSelf aBool} duplicates
@@
 For instance, the operation \code{aBool}
-defined in Section~\ref{ex:aBool} is translated into:
+defined in Sect.~\ref{ex:aBool} is translated into:
 \begin{haskell}
@@
 we must also extend the definition of operations performing
-pattern matching. For instance, consider the definition of polymorphic lists
+pattern matching.\footnote{To obtain a simple compilation scheme,
+KiCS2 transforms source programs into uniform programs
+\cite{BrasselHanusPeemoellerReck11} where pattern matching
+is restricted to a single argument.
+This is always possible by introducing auxiliary operations.}
+For instance, consider the definition of polymorphic lists
 \begin{curry}
@@
 If we apply the same transformation to the rules defining \code{xor}
-(actually, \code{xor} performs pattern matching on two arguments
-which is avoided in KiCS2 by introducing an intermediate operation,
-see \cite{BrasselHanusPeemoellerReck11}) and evaluate the main expression
-\code{(main 1)},
+and evaluate the main expression \code{(main 1)},
 we obtain the result
@@
 To extract all values from such a \code{Choice} structure,
-we have to traverse it and  compute all possible choices,
+we have to traverse it and  compute all possible choices
 but consider the choice identifiers to make consistent (left/right)
@@
 \begin{haskell}
-  printValsDFS :: a -> IO ()
+  printValsDFS :: a -> IO ()$\listline$
   printValsDFS Fail             = return ()$\listline$
@@
 rules and does not call, neither directly nor indirectly through
-other operations, a function defined by overlapping rules, it does not
-introduce non-determinism (like \code{xor} or \code{not}).
+other operations, a function defined by overlapping rules,
+the evaluation of such an operation (like \code{xor} or \code{not})
+cannot introduce non-deterministic values.
 Thus, it is not necessary to pass an identifier supply to the operation.
-In consequence, only the matching rules are extended, so that the generated code
+In consequence, only the matching rules are extended by additional cases
+for handling \code{Choice} and \code{Fail} so that the generated code
 is nearly identical to a corresponding functional program.
@@
 that instantiates the variable \code{x} to \code{True} or \code{False}.
-% Actually, it has been shown in \cite{AntoyHanus06ICLP}
-% that narrowing expressions with logic variables
-% is equivalent to rewriting with generators.
 Since such generators are standard
@@
 of an equational constraint in the operation
-\code{last} in Section~\ref{ex:last}.
+\code{last} in Sect.~\ref{ex:last}.
 
@@
 i.e., the equation \code{$e_1$\,=:=\,$e_2$} is satisfied iff
-$e_1$ and $e_2$ are reducible to unifiable ground constructor terms.
+$e_1$ and $e_2$ are reducible to unifiable constructor terms.
 In particular, expressions that do not have a value
@@
 If we translate these operations into Haskell by the scheme
-shown in Section~\ref{sec:Compilation},
+presented in Sect.~\ref{sec:Compilation},
 the following rules are added to these rules
@@
 \end{haskell}
-
+%
 As mentioned above, the consideration of free variables
@@
 Furthermore, we add the rule that an equational constraint
-between two variables yields a binding for these variables:
+between two variables yields a binding for these variables:%
 \begin{haskell}
@@
 
-% Due to the extensions described in this section,
-% we obtain an implementation of a unification procedure
-% that integrates the demand-driven evaluation of operations
-% occurring in the arguments.
-% Benchmarks evaluating the efficiency of this approach are
-% shown in Section~\ref{sec:Benchmarks}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 For instance, the equational constraint in the condition
-defining \code{last} (see Section~\ref{ex:last})
+defining \code{last} (see Sect.~\ref{ex:last})
 requires that \code{ys++[e]} as well as \code{xs}
@@
 On the other hand, the evaluation of \code{last' [failed,True]} yields
-the value \code{True}, i.e., the definition of \code{last'} is less strict.
-
-As another example for the advantage of the reduced strictness,
+the value \code{True}, i.e., the definition of \code{last'} is less strict
+thanks to the use of functional patterns.
+
+As another example for the advantage of the reduced strictness
+implied by functional patterns,
 consider an operation that returns the first duplicate element in
@@
 into three sublists.
-The last constraint ensures that the first sublist
+The last equational constraint ensures that the first sublist
 \code{ys} does not contain duplicated elements
 (the library operation \code{nub} removes all duplicates from a list)
-and the second constraint ensures that the first element after \code{ys}
-occurs in \code{ys}.
+and the second equational constraint ensures that the first element
+after \code{ys} occurs in \code{ys}.
 Although this implementation is concise, it cannot be applied
@@
 \end{curry}
-
+%
 Obviously, one cannot implement functional patterns by
@@
 is almost identical to the strict unification operation \ccode{=:=}
-as shown in Section~\ref{sec:Unification}.
+as shown in Sect.~\ref{sec:Unification}.
 The only difference is in the rules where a free variable occurs
@@
 and some other details; note that logic variables are replaced
-by generators):
+by generators, i.e., we assume that \code{xs} is replaced by
+\code{aBoolList 2} and \code{e} is replaced by \code{aBool 3}):
 \begin{haskell}
@@
 \label{sec:Benchmarks}
-
-% The benchmark results are from unif@climens2011_7_5_20_38_47.bench
 
@@
 \code{varInExp}                         & 2.34    & 0.24      & 0.21        \\
-% \code{varInExp2}                        & 0.34   & 0.02     & 0.01        \\
-% \code{varInExp3}                        & 13.28    & 1.36     & 1.02       \\
 \code{fromPeano (half (toPeano 10000))} & 26.67     & 5.95      & 11.19       \\
@@
 
-In Section~\ref{sec:Unification} we mentioned that equational
+In Sect.~\ref{sec:Unification} we mentioned that equational
 constraints could also be solved by generators without variable bindings,
@@
 \code{varInExp}                         & 0.24  & 0.89   & 0.07 \\
-% \code{varInExp2}                        & 0.02 & 0.06   & 0.00 \\
-% \code{varInExp3}                        & 1.36 & 2.07  & 0.14 \\
 \code{fromPeano (half (toPeano 10000))} & 5.95  & 108.88 & 3.22 \\
@@
 \code{varInExp}                         &  0.21  &  1.87    \\
-% \code{varInExp2}                         & 0.01  & 0.02    \\
-% \code{varInExp3}                         & 1.02 & 2.16   \\
 \code{fromPeano (half (toPeano 10000))} & 11.19 & $\infty$ \\
@@
 
-\bibliographystyle{mhnumber} %plain}
-\bibliography{mh}
+\begin{thebibliography}{10}
+
+\bibitem{AntoyHanus00FROCOS}
+S.~Antoy and M.~Hanus.
+\newblock Compiling multi-paradigm declarative programs into {Prolog}.
+\newblock In {\em Proc. International Workshop on Frontiers of Combining
+  Systems (FroCoS'2000)}, pages 171--185. Springer LNCS 1794, 2000.
+
+\bibitem{AntoyHanus02FLOPS}
+S.~Antoy and M.~Hanus.
+\newblock Functional logic design patterns.
+\newblock In {\em Proc.\ of the 6th International Symposium on Functional and
+  Logic Programming (FLOPS 2002)}, pages 67--87. Springer LNCS 2441, 2002.
+
+\bibitem{AntoyHanus05LOPSTR}
+S.~Antoy and M.~Hanus.
+\newblock Declarative programming with function patterns.
+\newblock In {\em Proceedings of the International Symposium on Logic-based
+  Program Synthesis and Transformation (LOPSTR'05)}, pages 6--22. Springer LNCS
+  3901, 2005.
+
+\bibitem{AntoyHanus06ICLP}
+S.~Antoy and M.~Hanus.
+\newblock Overlapping rules and logic variables in functional logic programs.
+\newblock In {\em Proceedings of the 22nd International Conference on Logic
+  Programming (ICLP 2006)}, pages 87--101. Springer LNCS 4079, 2006.
+
+\bibitem{AntoyHanus10CACM}
+S.~Antoy and M.~Hanus.
+\newblock Functional logic programming.
+\newblock {\em Communications of the ACM}, 53(4):74--85, 2010.
+
+\bibitem{AntoyHanus11WFLP}
+S.~Antoy and M.~Hanus.
+\newblock New functional logic design patterns.
+\newblock In {\em Proc.\ of the 20th International Workshop on Functional and
+  (Constraint) Logic Programming (WFLP 2011)}, pages 19--34. Springer LNCS
+  6816, 2011.
+
+\bibitem{AntoyHanusLiuTolmach05}
+S.~Antoy, M.~Hanus, J.~Liu, and A.~Tolmach.
+\newblock A virtual machine for functional logic computations.
+\newblock In {\em Proc.\ of the 16th International Workshop on Implementation
+  and Application of Functional Languages (IFL 2004)}, pages 108--125. Springer
+  LNCS 3474, 2005.
+
+\bibitem{AugustssonRittriSynek94}
+L.~Augustsson, M.~Rittri, and D.~Synek.
+\newblock On generating unique names.
+\newblock {\em Journal of Functional Programming}, 4(1):117--123, 1994.
+
+\bibitem{Brassel11Thesis}
+B.~Bra{\ss}el.
+\newblock {\em Implementing Functional Logic Programs by Translation into
+  Purely Functional Programs}.
+\newblock PhD thesis, Christian-Albrechts-Universit{\"a}t zu Kiel, 2011.
+
+\bibitem{BrasselFischer08IFL}
+B.~Bra{\ss}el and S.~Fischer.
+\newblock From functional logic programs to purely functional programs
+  preserving laziness.
+\newblock In {\em Pre-Proceedings of the 20th Workshop on Implementation and
+  Application of Functional Languages (IFL 2008)}, 2008.
+
+\bibitem{BrasselHanusHuch04JFLP}
+B.~Bra{\ss}el, M.~Hanus, and F.~Huch.
+\newblock Encapsulating non-determinism in functional logic computations.
+\newblock {\em Journal of Functional and Logic Programming}, 2004(6), 2004.
+
+\bibitem{BrasselHanusPeemoellerReck11}
+B.~Bra{\ss}el, M.~Hanus, B.~Peem{\"o}ller, and F.~Reck.
+\newblock {KiCS2}: A new compiler from {Curry} to {Haskell}.
+\newblock In {\em Proc. of the 20th International Workshop on Functional and
+  (Constraint) Logic Programming (WFLP 2011)}, pages 1--18. Springer LNCS 6816,
+  2011.
+
+\bibitem{BrasselHuch07}
+B.~Bra{\ss}el and F.~Huch.
+\newblock On a tighter integration of functional and logic programming.
+\newblock In {\em Proc. APLAS 2007}, pages 122--138. Springer LNCS 4807, 2007.
+
+\bibitem{BrasselHuch09}
+B.~Bra{\ss}el and F.~Huch.
+\newblock The {Kiel} {Curry} {System} {KiCS}.
+\newblock In {\em Applications of Declarative Programming and Knowledge
+  Management}, pages 195--205. Springer LNAI 5437, 2009.
+
+\bibitem{GiovannettiLeviMoisoPalamidessi91}
+E.~Giovannetti, G.~Levi, C.~Moiso, and C.~Palamidessi.
+\newblock Kernel leaf: A logic plus functional language.
+\newblock {\em Journal of Computer and System Sciences}, 42(2):139--185, 1991.
+
+\bibitem{GonzalezEtAl99}
+J.C. Gonz{\'a}lez-Moreno, M.T. Hortal{\'a}-Gonz{\'a}lez, F.J.
+  L{\'o}pez-Fraguas, and M.~Rodr{\'\i}guez-Artalejo.
+\newblock An approach to declarative programming based on a rewriting logic.
+\newblock {\em Journal of Logic Programming}, 40:47--87, 1999.
+
+\bibitem{Hanus07ICLP}
+M.~Hanus.
+\newblock Multi-paradigm declarative languages.
+\newblock In {\em Proceedings of the International Conference on Logic
+  Programming (ICLP 2007)}, pages 45--75. Springer LNCS 4670, 2007.
+
+\bibitem{Hanus11ICLP}
+M.~Hanus.
+\newblock Declarative processing of semistructured web data.
+\newblock In {\em Technical Communications of the 27th International Conference
+  on Logic Programming}, volume~11, pages 198--208. Leibniz International
+  Proceedings in Informatics (LIPIcs), 2011.
+
+\bibitem{Hanus10PAKCS}
+M.~Hanus, S.~Antoy, B.~Bra{\ss}el, M.~Engelke, K.~H{\"o}ppner, J.~Koj,
+  P.~Niederau, R.~Sadre, and F.~Steiner.
+\newblock {PAKCS}: The {P}ortland {A}achen {K}iel {C}urry {S}ystem.
+\newblock Available at \url{http://www.informatik.uni-kiel.de/~pakcs/}, 2010.
+
+\bibitem{HanusSadre99JFLP}
+M.~Hanus and R.~Sadre.
+\newblock An abstract machine for curry and its concurrent implementation in
+  java.
+\newblock {\em Journal of Functional and Logic Programming}, 1999(6), 1999.
+
+\bibitem{Hanus06Curry}
+M.~Hanus~(ed.).
+\newblock Curry: An integrated functional logic language (vers.\ 0.8.2).
+\newblock Available at \url{http://www.curry-language.org}, 2006.
+
+\bibitem{Hussmann92}
+H.~Hussmann.
+\newblock Nondeterministic algebraic specifications and nonconfluent term
+  rewriting.
+\newblock {\em Journal of Logic Programming}, 12:237--255, 1992.
+
+\bibitem{Lopez-FraguasSanchez-Hernandez99}
+F.~L\'opez-Fraguas and J.~S\'anchez-Hern\'andez.
+\newblock {TOY}: A multiparadigm declarative system.
+\newblock In {\em Proc. of RTA'99}, pages 244--247. Springer LNCS 1631, 1999.
+
+\bibitem{Lux99FLOPS}
+W.~Lux.
+\newblock Implementing encapsulated search for a lazy functional logic
+  language.
+\newblock In {\em Proc. 4th Fuji International Symposium on Functional and
+  Logic Programming (FLOPS'99)}, pages 100--113. Springer LNCS 1722, 1999.
+
+\bibitem{ODonnell85}
+M.J. O'Donnell.
+\newblock {\em Equational Logic as a Programming Language}.
+\newblock MIT Press, 1985.
+
+\bibitem{PeytonJones03Haskell}
+S.~Peyton~Jones, editor.
+\newblock {\em Haskell 98 Language and Libraries---The Revised Report}.
+\newblock Cambridge University Press, 2003.
+
+\bibitem{Reddy85}
+U.S. Reddy.
+\newblock Narrowing as the operational semantics of functional languages.
+\newblock In {\em Proc. IEEE Internat. Symposium on Logic Programming}, pages
+  138--151, Boston, 1985.
+
+\bibitem{Robinson65}
+J.A. Robinson.
+\newblock A machine-oriented logic based on the resolution principle.
+\newblock {\em Journal of the ACM}, 12(1):23--41, 1965.
+
+\bibitem{SchrijversStuckeyWadler09}
+T.~Schrijvers, P.~Stuckey, and P.~Wadler.
+\newblock Monadic constraint programming.
+\newblock {\em Journal of Functional Programming}, 19(6):663--697, 2009.
+
+\bibitem{Slagle74}
+J.R. Slagle.
+\newblock Automated theorem-proving for theories with simplifiers,
+  commutativity, and associativity.
+\newblock {\em Journal of the ACM}, 21(4):622--642, 1974.
+
+\end{thebibliography}
 

### 2011-10-08 docs/src/manual.tex
First version of manual added

\documentclass[11pt,fleqn]{article}

\usepackage{latexsym}
\usepackage{makeidx}
\usepackage{url}
\usepackage{xspace}
\usepackage{graphicx}

%%% ------------------------------------------------------------------

\usepackage[colorlinks,linkcolor=blue]{hyperref}
\hypersetup{bookmarksopen=true}
\hypersetup{bookmarksopenlevel=0}
\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
\hypersetup{pdfauthor={Michael Hanus}}
%\hypersetup{pdfstartview=Title}
\hypersetup{pdfstartview=FitH}
\usepackage{thumbpdf}

%%% ------------------------------------------------------------------

\setlength{\textwidth}{16.5cm}
\setlength{\textheight}{23cm}
\renewcommand{\baselinestretch}{1.1}
\setlength{\topmargin}{-1cm}
\setlength{\oddsidemargin}{0cm}
\setlength{\evensidemargin}{0cm}
\setlength{\marginparwidth}{0.0cm}
\setlength{\marginparsep}{0.0cm}

\newlength{\figurewidth}
\setlength{\figurewidth}{\textwidth}
\addtolength{\figurewidth}{-0.4cm}

% font for program texts
\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
\newcommand{\codefont}{\tt}

% environment for typing program texts:
\makeatletter
\newenvironment{prog}{\par\vspace{0.7ex}
\setlength{\parindent}{1.0cm}
\setlength{\parskip}{-0.1ex}
\obeylines\@vobeyspaces\tt}{\vspace{0.7ex}\noindent
}
\makeatother
\newcommand{\startprog}{\begin{prog}}
\newcommand{\stopprog}{\end{prog}\noindent}

% program text in normal text
\newcommand{\code}[1]{\mbox{\codefont #1}}

% program text in normal text with apostrophs
\newcommand{\ccode}[1]{``\mbox{\codefont #1}''}

\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index

\newcommand{\labs}{\mbox{\tt\char92}}  % lambda abstraction in Curry
\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
\newcommand{\cb}{CurryBrowser\xspace}

% allow underscores in programs:
\catcode`\_=\active
\let_=\sb
\catcode`\_=12

% produce an index:
\makeindex

\begin{document}
\sloppy

\begin{titlepage}
\pdfbookmark[1]{Title}{Title}
\begin{center}
\fbox{
\begin{minipage}[t]{\figurewidth}
\begin{center}\vspace{10ex}
{\Huge\bf KiCS2 0.1}\\[4ex]
{\huge The Kiel Curry System (Version 2)}\\[7ex]
{\huge User Manual}\\[4ex]
Version of September 29, 2011\\[6ex]
\Large
Michael Hanus$^1$ [editor] \\[3ex]
{\large Additional Contributors:}\\[2ex]
Bernd Bra\ss{}el$^2$ \\
Bj\"orn Peem\"oller$^3$\\
Fabian Reck$^4$ \\[4ex]
\normalsize
(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
\end{center}
\end{minipage}}
\end{center}
\end{titlepage}

\pdfbookmark[1]{Contents}{Contents}
\tableofcontents

\newpage

\addcontentsline{toc}{section}{Preface}
\section*{Preface}

This document describes KiCS2 (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
Version \textbf{2}),
an implementation of the multi-paradigm language Curry
\cite{Hanus97POPL,Hanus06Curry} that is
based on compiling Curry programs into Haskell programs.
Curry is a universal programming language aiming at the amalgamation
of the most important declarative programming paradigms,
namely functional programming and logic programming.  
Curry combines in a seamless way features from functional programming
(nested expressions, lazy evaluation, higher-order functions),
logic programming (logical variables, partial data structures,
built-in search), and concurrent programming (concurrent evaluation
of constraints with synchronization on logical variables).
The current KiCS2 implementation does not support concurrent
constraints. Alternatively, one can write distributed applications
by the use of sockets that can be registered and accessed
with symbolic names.
Moreover, KiCS2 also supports
the high-level implementation of
graphical user interfaces and web services
(as described in more detail in
\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).

We assume familiarity with the ideas and features
of Curry as described in the Curry language definition \cite{Hanus06Curry}.
Therefore, this document only explains the use of the different
components of KiCS2
and the differences and restrictions of KiCS2
(see Section~\ref{sec-restrictions})
compared with the language Curry (Version 0.8.2).
The basic ideas of the implementation of KiCS2
can be found in
\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.

\bigskip

\subsection*{Acknowledgements}

This work has been supported in part by the
DFG grants Ha 2457/5-1 and Ha 2457/5-2.

\newpage

\section{Overview of KiCS2}

\subsection{General Use}
\label{sec-general}

This version of KiCS2 has been developed and
tested on Linux systems.
In principle, it should be also executable on other
platforms on which a Haskell implementation (Glasgow Haskell Compiler
and Cabal) exists, like in many Linux distributions, Sun Solaris,
or Mac OS X systems
(see the file \code{INSTALL.html} in the KiCS2 directory
for a description of the necessary software to install KiCS2).

All executables required to use the different components
of KiCS2 are stored in the directory \code{$kics2home$/bin}
(where $kics2home$ is the installation directory of the complete
KiCS2 installation). You should add this directory
to your path (e.g., by the \code{bash} command
\ccode{export PATH=$kics2home$/bin:\$PATH}).

The source code of the Curry program
must be stored in a file with the suffix \ccode{.curry},
e.g., \code{prog.curry}. 
Literate programs must be stored in files with the extension \ccode{.lcurry}.

Since the translation of Curry programs with KiCS2 creates
some auxiliary files (see Section~\ref{sec-auxfiles} for details),
you need write permission
in the directory where you have stored your Curry programs.
Moreover, the current implementation also recompiles
system libraries according to the setting of some options.
Therefore, the KiCS2 system should be locally installed
in your user account.
The auxiliary files for all Curry programs in the current
directory can be deleted by the command\pindex{cleancurry}
\startprog
cleancurry
\stopprog
(this is a shell script stored in the \code{bin} directory of the
KiCS2 installation, see above).
The command
\startprog
cleancurry -r
\stopprog
also deletes the auxiliary files in all subdirectories.

\subsection{Restrictions on Curry Programs}
\label{sec-restrictions}

There are a few minor restrictions on Curry programs
when they are processed with KiCS2:
\begin{itemize}
\item \index{singleton variables}\index{variables!singleton}
\emph{Singleton variables}, i.e., variables that occur only once
in a rule, should be denoted as an anonymous variable \ccode{_},
otherwise the parser will print a warning since this is a
typical source of programming errors.
\item
KiCS2 translates all \emph{local declarations} into global functions with
additional arguments (``lambda lifting'', see Appendix~D of the
Curry language report).
Thus, in the various run-time systems, the definition of
functions with local declarations look different from
their original definition (in order to see the result
of this transformation, you can use the \cb, see
Section~\ref{sec-currybrowser}).
\item \index{tabulator stops}
Tabulator stops instead of blank spaces in source files are
interpreted as stops at columns 9, 17, 25, 33, and so on.
%\item Threads created by a concurrent conjunction are not executed
%in a fair manner (usually, threads corresponding to leftmost constraints
%are executed with higher priority).
\item
Encapsulated search\index{encapsulated search}: In order
to allow the integration of non-deterministic computations
in programs performing I/O at the top-level, KiCS2 provides
the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
and \code{SearchTree} (Section~\ref{Library:SearchTree})
where one can defined search operators and various search strategies.
The general definition of encapsulated search of the Curry report
\cite{HanusSteiner98PLILP} is not supported.
\item
There is currently no general connection to external constraint solvers.
\end{itemize}

\subsection{Modules in KiCS2}
\label{sec-modules}

The current implementation of KiCS2 supports only flat module names,
i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
In order to allow the structuring of modules in different directories,
KiCS2 searches for imported modules in various directories.
By default, imported modules are searched in the directory
of the main program and the system module directories
\ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
This search path can be extended
by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
(which can be also set in a KiCS2 session by the option
\ccode{:set path}\pindex{path}\pindex{:set path},
see below)
to a list of directory names separated by colons (\ccode{:}).
In addition, a local standard search path
can be defined in the \ccode{.kics2rc} file
(see Section~\ref{sec-customization}).
Thus, modules to be loaded are searched in the following
directories (in this order, i.e., the first occurrence of a module file
in this search path is imported):
\begin{enumerate}
\item Current working directory (\ccode{.}) or directory prefix
of the main module (e.g., directory \ccode{/home/joe/curryprogs}
if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
\item The directories enumerated in the environment variable \code{CURRYPATH}.
\item The directories enumerated in the \ccode{.kics2rc} variable
      \ccode{libraries}.
\item The directories \ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
\end{enumerate}
Note that the standard prelude (\code{$kics2home$/lib/Prelude.curry})
will be always implicitly imported to all modules if a module
does not contain an explicit import declaration for the module
\code{Prelude}.

\newpage

\section{Using the Interactive Environment of KiCS2}
\label{sec-kics2}

This section describes the interactive environment
KiCS2\index{KiCS2}
that supports the development of applications written in Curry.
The implementation of KiCS2 contains also a separate compiler
which is automatically invoked by the interactive environment.

\subsection{Invoking KiCS2}
\label{sec:invoke-kics2}

To start KiCS2, execute the command
\ccode{kics2}\pindex{kics2}
(this is a shell script stored in
\code{$kics2home$/bin} where $kics2home$ is the installation directory
of KiCS2).
When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
the prelude (\code{$kics2home$/lib/Prelude.curry})
is already loaded, i.e., all definitions in the prelude are accessible.
Now you can type various commands (see next section)
or an expression to be evaluated.

One can also invoke KiCS2 with parameters.
These parameters are usual a sequence of commands
(see next section) that are executed before the user interaction
starts. For instance, the invocation
\startprog
kics2 :load Mod :add List
\stopprog
starts KiCS2, loads the main module \code{Mod}, and adds the additional
module \code{List}.
The invocation
\startprog
kics2 :load Mod :eval config
\stopprog
starts KiCS2, loads the main module \code{Mod}, and evaluates
the operation \code{config} before the user interaction starts.
As a final example, the invocation
\startprog
kics2 :load Mod :save :quit
\stopprog
starts KiCS2, loads the main module \code{Mod}, creates an executable,
and terminates KiCS2. This invocation could be useful in ``make'' files
for systems implemented in Curry.

\subsection{Command of KiCS2}
\label{sec:kics2-commands}

The {\bf most important commands} of KiCS2 are
(it is sufficient to type a unique prefix of a command if it is unique,
e.g., one can type \ccode{:r} instead of \ccode{:reload}):

\begin{description}
\item[\fbox{\code{:help}}]\pindex{:help}
Show a list of all available commands.

\item[\fbox{\code{:load $prog$}}]\pindex{:load}
Compile and load the program stored in \code{$prog$.curry}
together with all its imported modules.
% If this file does not exist, the system looks for a FlatCurry
% file \code{$prog$.fcy} and compiles from this intermediate representation.
% If the file \code{$prog$.fcy} does not exists, too, the system looks
% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
% XML representation (compare command \ccode{:xml}\pindex{:xml}),
% translates this into a FlatCurry file \code{$prog$.fcy}
% and compiles from this intermediate representation.

\item[\fbox{\code{:reload}}]\pindex{:reload}
Recompile all currently loaded modules.

\item[\fbox{\code{:add} $m$}]\pindex{:add}
Add module $m$ to the set of currently loaded modules
so that its exported entities are available in the top-level environment.

\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
and show the computed results.
In the default mode, all results of
non-deterministic computations are printed.
One can also print first one result and the next result
only if the user requests it. This behavior can be set by
the option \code{interactive}\index{interactive} (see below).

\textbf{Free variables in initial expressions} must be declared as
in Curry programs, i.e., in order to see the results of their bindings,
they must be introduced by a \ccode{where\ldots{}free} declaration.
For instance, one can write
\startprog
xs++ys\,=:=\,[1,2]  where xs,ys free
\stopprog
in order to obtain the following three possible bindings:
\startprog
\{xs = [], ys = [1,2]\} Success
\{xs = [1], ys = [2]\} Success
\{xs = [1,2], ys = []\} Success
\stopprog
Without these declarations, an error is reported in order to
avoid the unintended introduction of free variables in initial expressions
by typos.

If the free variables in the initial goal are of a polymorphic type,
as in the expression
\startprog
xs++ys=:=[z] where xs,ys,z free
\stopprog
they are specialized to the type \ccode{()}
(since the current implementation of KiCS2 cannot compute
with polymorphic logic variables).

\item[\fbox{:eval $expr$}]\pindex{:eval}
Same as $expr$. This command might be useful when putting
commands as arguments when invoking \code{kics2}.

% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
% Define the identifier $x$ as an abbreviation for the expression $expr$
% which can be used in subsequent expressions. The identifier $x$
% is visible until the next \code{load} or \code{reload} command.

\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
\end{description}
%
\bigskip
%
There are also a number of {\bf further commands} that are often
useful:
%
\begin{description}
\item[\fbox{\code{:type $expr$}}]\pindex{:type}
Show the type of the expression $expr$.

\item[\fbox{\code{:programs}}]\pindex{:programs}
Show the list of all Curry programs that are available in the load path.

\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
Change the current working directory to $dir$.

\item[\fbox{\code{:edit}}]\pindex{:edit}
Load the source code of the current main module into a text editor.
If the variable \code{editcommand} is set in the
configuration file \ccode{.kics2rc}
(see Section~\ref{sec-customization}),
its value is used as an editor command, otherwise
the environment variable \ccode{EDITOR} is used as the editor program.

\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
Load file $file$ into a text editor which is defined
as in the command \ccode{:edit}.

\item[\fbox{\code{:show}}]\pindex{:show}
Show the source text of the currently loaded Curry program.
If the variable \code{showcommand} is set in the
configuration file \ccode{.kics2rc}
(see Section~\ref{sec-customization}),
its value is used as a command to show the source text,
otherwise the \ccode{cat} is used.

\item[\fbox{\code{:show $m$}}]\pindex{:show}
Show the source text of module $m$ which must be accessible
via the current load path.

\item[\fbox{\code{:source $f$}}]\pindex{:source}
Show the source code of function $f$ (which must be visible
in the currently loaded module) in a separate window.

\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
Show the source code of function $f$ defined in module $m$
in a separate window.

\item[\fbox{\code{:browse}}]\pindex{:browse}
Start the CurryBrowser to analyze the currently loaded
module together with all its imported modules
(see Section~\ref{sec-currybrowser} for more details).

\item[\fbox{\code{:interface}}]\pindex{:interface}
Show the interface of the currently loaded
module, i.e., show the names of all imported modules,
the fixity declarations of all exported operators,
the exported datatypes declarations and the types
of all exported functions.

\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
Similar to \ccode{:interface}
but shows the interface of the module $m$
which must be in the load path of KiCS2.

\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
Show all calls to imported functions in the currently loaded module.
This might be useful to see which import declarations are really necessary.

\item[\fbox{\code{:set $option$}}]\pindex{:set}
Set or turn on/off a specific option
of the KiCS2 environment (see \ref{sec:options} for a description
of all options). Options are turned on by the prefix
\ccode{+} and off by the prefix \ccode{-}. Options that can only
be set (e.g., \code{path}) must not contain a prefix.

\item[\fbox{\code{:set}}]\pindex{:set}
Show a help text on the possible options
together with the current values of all options.

\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
program as an executable evaluating the main expression \ccode{main}.
The executable is stored in the file \code{Mod}
if \code{Mod} is the name of the currently loaded main module.

\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
but the expression $expr$ (typically: a call to the main
function) will be evaluated by the executable.

\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
The expression $expr$, which is typically of type \ccode{IO ()},
is evaluated in an independent process which runs in
parallel to the current KiCS2 process.
All output and error messages from this new process are suppressed.
This command is useful to test distributed Curry programs
where one can start
a new server process by this command. The new process
will be terminated when the evaluation of the expression $expr$
is finished.

\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.

\end{description}

\subsection{Option of KiCS2}
\label{sec:options}

The following options (which can be set by the command \ccode{:set})
are currently supported:

\begin{description}
\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
for loading modules to $path$.
Note that this search path is only used for loading modules
inside this invocation of KiCS2.

\item[\fbox{\code{prdfs }}]\pindex{prdfs}
Set the search mode to evaluate non-deterministic expressions
to primitive depth-first search (which is usually the fastest method
to print all non-deterministic values).

\item[\fbox{\code{dfs}}]\pindex{dfs}
Set the search mode to evaluate non-deterministic expressions
to depth-first search. Usually, all non-deterministic values
are printed with a depth-first strategy, but one can also
print only the first value or all values by interactively requesting them
(see below for these options).

\item[\fbox{\code{bfs}}]\pindex{bfs}
Similarly to \code{dfs} but use a breadth-first search strategy
to print the values.

\item[\fbox{\code{ids $n$}}]\pindex{ids}
Similarly to \code{dfs} but use an iterative-deepening strategy
with an initial depth bound of $n$. The depth-bound is
duplicated after each iteration.
If $n$ is missing, the default value $n=100$ is used.

\item[\fbox{\code{par $n$}}]\pindex{par}
Similarly to \code{dfs} but use a parallel search strategy
with $n$ threads.
If $n$ is missing, the system chooses an appropriate number
of threads according the current number of available processors.

\item[\fbox{\code{choices}}]\pindex{choices}
Show the internal choice structure
(according to the implementation described in
\cite{BrasselHanusPeemoellerReck11})
resulting from the complete evaluation of the main expression.
This mode is only useful for debugging purposes.

\item[\fbox{\code{supply $i$}}]\pindex{supply}
Use the implementation $i$ for the identifier supply for choice structures
(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
Currently, the values \code{ghc} (default), \code{integer},
or \code{ioref} are supported for $i$.

\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
Set the verbosity level to $n$. The following values are allowed
for $n$:
\begin{description}
\item[$n=0$:] Do not show any messages (except for errors).
\item[$n=1$:] Show only messages of the front-end, like loading
of modules.
\item[$n=2$:]
Show also messages of the back end, like compilation messages
from the Haskell compiler.
\item[$n=3$:]
Show also intermediate messages and commands of the compilation
process.
\item[$n=4$:]
Show also all intermediate results of the compilation process.
\end{description}

\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
Turn on/off the interactive mode.
In the interactive mode, the next non-deterministic value
is only computed when the user requests it.
In this mode, the computation is performed in a separate
window created by the command \code{interactivecommand}
whose value is defined in \ccode{.kics2rc}
(see Section~\ref{sec-customization}).

\item[\fbox{\code{+/-first}}]\pindex{first}
Turn on/off the first-only mode.
In the first-only mode, only the first value
of the main expression is printed (instead of all values).

\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
Turn on/off the optimization of the target program.

\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
Turn on/off the binding mode.
If the binding mode is on (default),
then the bindings of the free variables of the initial expression
are printed together with the result of the expression.

\item[\fbox{\code{+/-time}}]\pindex{time}
Turn on/off the time mode. If the time mode is on,
the cpu time and the elapsed time
of the computation is always printed together with the result
of an evaluation.

\item[\fbox{\code{+/-ghci}}]
Turn on/off the ghci mode.
In the ghci mode, the initial goal is send to the interactive version
of the Glasgow Haskell Compiler. This might result in a slower
execution but in a faster startup time since the linker
to create the main executable is not used.

\item[\fbox{\code{copts $o$}}]\pindex{copts}
Define additional options passed to the KiCS2 compiler.

\item[\fbox{\code{rts $o$}}]\pindex{rts}
Define additional run-time options passed to the Glasgow Haskell Compiler
(by \ccode{+RTS $o$ -RTS}).

\end{description}

\subsection{Source-File Options}

If the evaluation of operations in some main module loaded into
KiCS2 requires specific options, like an iterative-deepening
search strategy, one can also put these options into the source
code of this module in order to avoid setting these options
every time when this module is loaded.
Such
{\bf source-file options}\index{source-file option}\index{option!in source file}
must occur before the module header, i.e., before the first declaration
(module header, imports, fixity declaration, defining rules, etc)
occurring in the module.
Each source file option must be in a line of the form
\startprog
\{-\# KiCS2_OPTION $opt$ \#-\}
\stopprog
where $opt$ is an option that can occur in a \ccode{:set} command
(compare Section~\ref{sec:options}).
Such a line in the source code (which is a comment according to
the syntax of Curry)
has the effect that this option is set by the KiCS2 command
\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
as a main module. For instance, if a module starts with the
lines
\startprog
\{-\# KiCS2_OPTION ids \#-\}
\{-\# KiCS2_OPTION +ghci \#-\}
\{-\# KiCS2_OPTION v2 \#-\}
module M where
$\ldots$
\stopprog
then the load command \ccode{:load M} will also
set the options for iterative deepening, using \code{ghci}
and verbosity level 2.

\subsection{Command Line Editing}

In order to have support for line editing or history functionality
in the command line of KiCS2 (as often supported by the \code{readline}
library), you should have the Unix command \code{rlwrap} installed
on your local machine.
If \code{rlwrap} is installed, it is used by KiCS2 if called on a terminal.
If it should not be used (e.g., because it is executed
in an editor with \code{readline} functionality), one can
call KiCS2 with the parameter \ccode{--noreadline}
(which must occur as the first parameter).

\subsection{Customization}
\label{sec-customization}

In order to customize the behavior of KiCS2 to your own preferences,
there is a configuration file which is read by KiCS2 when it is invoked.
When you start KiCS2 for the first time, a standard version of
this configuration file is copied with the name
\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
into your home directory. The file contains definitions
of various settings, e.g., about showing warnings, using Curry extensions,
programs etc.
After you have started KiCS2 for the first time, look into this file
and adapt it to your own preferences.

\subsection{Emacs Interface}

Emacs is a powerful programmable editor suitable for program development.
It is freely available for many platforms
(see \url{http://www.emacs.org}).
The distribution of KiCS2 contains also a special
\emph{Curry mode}\index{Curry mode}\index{Emacs}
that supports the development of Curry programs in
the Emacs environment.
This mode includes support for syntax highlighting,
finding declarations in the current buffer, and
loading Curry programs into KiCS2
in an Emacs shell.

The Curry mode has been adapted from a similar mode for Haskell programs.
Its installation is described in the file \code{README}
in directory \ccode{$kics2home$/tools/emacs} which also contains
the sources of the Curry mode and a short description about
the use of this mode.

\newpage

\section{Extensions}
\label{sec-extensions}

KiCS2 supports some extensions in Curry programs that are not (yet)
part of the definition of Curry. These extensions are described below.

\subsection{Recursive Variable Bindings}

Local variable declarations (introduced by \code{let}\pindex{let}
or \code{where}\pindex{where}) can be (mutually) recursive in KiCS2.
For instance, the declaration
\startprog
ones5 = let ones = 1 : ones
         in take 5 ones
\stopprog
introduces the local variable \code{ones} which is bound
to a \emph{cyclic structure}\index{cyclic structure}
representing an infinite list of \code{1}'s.
Similarly, the definition
\startprog
onetwo n = take n one2
 where
   one2 = 1 : two1
   two1 = 2 : one2
\stopprog
introduces a local variables \code{one2} that represents
an infinite list of alternating \code{1}'s and \code{2}'s
so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.

\subsection{Functional Patterns}

Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
to code operations in a more readable way. Furthermore,
defining operations with functional patterns avoids problems
caused by strict equality (\ccode{=:=}) and leads to programs
that are potentially more efficient.

Consider the definition of an operation to compute the last element
of a list \code{xs} based on the prelude operation \ccode{++}
for list concatenation:
\startprog
last xs | ys++[y] =:= xs  = y   where y,ys free
\stopprog
Since the equality constraint \ccode{=:=} evaluates both sides
to a constructor term, all elements of the list \code{xs} are
fully evaluated in order to satisfy the constraint.

Functional patterns can help to improve this computational behavior.
A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
is a function call at a pattern position. With functional patterns,
we can define the operation \code{last} as follows:
\startprog
last (_++[y]) = y
\stopprog
This definition is not only more compact but also avoids the complete
evaluation of the list elements: since a functional pattern is considered
as an abbreviation for the set of constructor terms obtained by all
evaluations of the functional pattern to normal form (see
\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
definition is conceptually equivalent to the set of rules
\startprog
last [y] = y
last [_,y] = y
last [_,_,y] = y
\ldots
\stopprog
which shows that the evaluation of the list elements is not demanded
by the functional pattern.

In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
is interpreted as a functional pattern if $f$ is not a visible constructor
but a defined function that is visible in the scope of the pattern.

\subsection {Records}
\label{records}

A record is a data structure for bundling several data of various types.
It consists of typed data fields where each field is associated with
a unique label. These labels can be used to construct, select or update
fields in a record.

Unlike labeled data fields in Haskell, records are 
not syntactic sugar but a real extension of the
language\footnote{The current version allows to transform records
  into abstract data types. Future extensions may not have
  this facility.}.
The basic concept is described in \cite{Leijen05} but the current
version does not yet provide all features mentioned there. 
The restrictions are explained in Section~\ref{sec-restrinrecs}.

\subsubsection{Record Type Declaration}
\label{sec-recordtypedecl}

It is necessary to declare a record type before a record
can be constructed or used. The declaration has the following form:
\startprog
type $R$ $\alpha_1$ \ldots $\alpha_n$ = \{ $l_1$ :: $\tau_1$, \ldots, $l_m$ :: $\tau_m$ \}
\stopprog
It introduces a new $n$-ary record type $R$ which represents a
record consisting of $m$ fields. Each field has a unique label $l_i$ 
representing a value of the type $\tau_i$. Labels
are identifiers which refer to the corresponding
fields. The following examples define some record types:
\startprog
type Person = \{name :: String, age :: Int\}
type Address = \{person :: Person, street :: String, city :: String\}
type Branch a b = \{left :: a, right :: b\}
\stopprog
It is possible to summarize different labels which have the same
type. For instance, the record \code{Address} can also be declared as follows:
\startprog
type Address = \{person :: Person, street,city :: String\}
\stopprog
The fields can occur in an arbitrary order. The example above
can also be written as
\startprog
type Address = \{street,city :: String, person :: Person\}
\stopprog
The record type can be used in every type expression to represent
the corresponding record, e.g.
\startprog
data BiTree = Node (Branch BiTree BiTree) | Leaf Int
\stopprog
\startprog
getName :: Person -> String
getName \ldots
\stopprog

Labels can only be used in the context of
records. They do not share the name space with 
functions/constructors/variables or type constructors/type variables. 
For instance it is possible to use 
the same identifier for a label and a function at the same time. Label
identifiers cannot be shadowed by other identifiers.

Like in type synonym declarations, recursive or mutually 
dependent record declarations are not allowed. Records can only
be declared at the top level. Further restrictions are described in
section \ref{sec-restrinrecs}.

\subsubsection{Record Construction}
\label{sec-recordconstr}

The record construction generates a record with initial values for
each data field. It has the following form:
\startprog
\{ $l_1$ = $v_1$, \ldots, $l_m$ = $v_m$ \}
\stopprog
It generates a record where each label $l_i$ refers to the
value $v_i$. The type of the record results from the record type
declaration where the labels $l_i$ are defined.
A mix of labels from different
record types is not allowed. All labels must be specified with 
exactly one assignment. Examples for record constructions are
\startprog
\{name = "Johnson", age = 30\}     -- generates a record of type 'Person'
\{left = True, right = 20\}        -- generates a record of type 'Branch'
\stopprog
Assignments to labels can occur in an arbitrary order. For instance a
record of type \code{Person} can also be generated as follows:
\startprog
\{age = 30, name = "Johnson"\}     -- generates a record of type 'Person'
\stopprog
Unlike labeled fields in record type declarations, 
record constructions can be used in expressions without any restrictions
(as well as all kinds of record expressions). For instance the following
expression is valid:
\startprog
\{person = \{name = "Smith", age = 20\},   -- generates a record of
 street = "Main Street",                -- type 'Address'
 city   = "Springfield"\}
\stopprog

\subsubsection{Field Selection}
\label{sec-fieldsel}

The field selection is used to extract data from records. 
It has the following form:
\startprog
$r$ -> $l$
\stopprog
It returns the value to which the label $l$ refers to from the
record expression $r$. The label must occur in the declaration of
the record type of $r$.
An example for a field selection is:
\startprog
pers -> name
\stopprog
This returns the value of the label \code{name} from the record \code{pers}
(which has the type \code{Person}).
Sequential application of field selections are also possible:
\startprog
(addr -> person) -> age
\stopprog
The value of the label \code{age} is extracted from a record which itself
is the value of the label \code{person} in the record \code{addr}
(which has the type \code{Address}). When a field selection is used in
expressions, it has to be parenthesized.

\subsubsection{Field Update}
\label{sec-fieldupd}

Records can be updated by reassigning a new value to a label:
\startprog
\{$l_1$ := $v_1$, \ldots, $l_k$ := $v_k$ | $r$\}
\stopprog
The label $l_i$ is associated with the new value $v_i$ which
replaces the current value in the record $r$.
The labels must occur in the declaration 
of the record type of $r$. In contrast to record constructions,
it is not necessary to specify all labels of a record. 
Assignments can occur in an arbitrary order. It is not allowed to 
specify more than one assignment for a label in a record update.
Examples for record updates are:
\startprog
\{name := "Scott", age := 25 | pers\}
\{person := \{name := "Scott", age := 25 | pers\} | addr\}
\stopprog
In these examples \code{pers} is a record of type \code{Person} and \code{addr}
is a record of type \code{Address}.

\subsubsection{Records in Pattern Matching}
\label{sec-recsinpm}

It is possible to apply pattern matching to records (e.g., in functions,
let expressions or case branches). Two kinds of record patterns
are available:
\startprog
\{$l_1$ = $p_1$, \ldots, $l_n$ = $p_n$\}
\{$l_1$ = $p_1$, \ldots, $l_k$ = $p_k$ | _\}
\stopprog
In both cases each label $l_i$ is specified with a pattern $p_i$. 
All labels must occur only once in the record pattern.
The first case is used to match the whole record. Thus, all labels
of the record must occur in the pattern. 
The second case is used to match only a part of
the record. Here it is not necessary to specify all labels.
This case is represented by a vertical bar followed by the underscore
(anonymous variable). It is
not allowed to use a pattern term instead of the underscore.

When trying to match a record against a record pattern, the 
patterns of the specified labels are matched against 
the corresponding values in the record expression. On success, all pattern
variables occurring in the patterns are replaced by their actual expression.
If none of the patterns matches, the computation fails.

Here are some examples of pattern matching with records:
\startprog
isSmith30 :: Person -> Bool
isSmith30 \{name = "Smith", age = 30\} = True
\stopprog
\startprog
startsWith :: Char -> Person -> Bool
startsWith c \{name = (d:_) | _\} = c == d
\stopprog
\startprog
getPerson :: Address -> Person
getPerson \{person = p | _\} = p
\stopprog
As shown in the last example, a field selection can also be obtained
by pattern matching.

\subsubsection{Export of Records}
\label{sec-exprecs}

Exporting record types and labels is very similar to exporting
data types and constructors. There are three ways 
to specify an export:
\begin{itemize}
\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
  exports the record $R$ without any of its labels.
\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
  exports the record $R$ together with all its labels.
\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
\end{itemize}
%
Note that imported labels cannot be overwritten in record declarations
of the importing module. It is also not possible to import equal labels
from different modules.

\subsubsection{Restrictions in the Usage of Records}
\label{sec-restrinrecs}

In contrast to the basic concept in \cite{Leijen05}, KiCS2/Curry provides a
simpler version of records. Some of the features described there are
currently not supported or even restricted.

\begin{itemize}
\item Labels must be unique within the whole scope of the program.
  In particular, it is not allowed to define the same label within
  different records, not even when they are imported from other
  modules. However, it is possible to use equal identifiers for other
  entities without restrictions, since labels have an independent 
  name space.
\item The record type representation with labeled fields can only be
  used as the right-hand-side of a record type declaration. It is
  not allowed to use it in any other type annotation.
\item Records are not extensible or reducible. The structure of a
  record is specified in its record declaration and cannot be
  modified at the runtime of the program.
\item Empty records are not allowed.
\item It is not allowed  to use a pattern term
  at the right side of the vertical bar in a record pattern
  except for the underscore (anonymous pattern variable).
\item Labels cannot be sequentially associated with multiple values
  (record fields do not behave like stacks).
\end{itemize}

\newpage

\section{CurryDoc: A Documentation Generator for Curry Programs}
\label{sec-currydoc}

CurryDoc\index{CurryDoc}
\index{program!documentation}\index{documentation generator}
is a tool in the KiCS2 distribution that generates
the documentation for a Curry program (i.e., the main module
and all its imported modules) in HTML format.
The generated HTML pages contain information about
all data types and functions exported by a module as well
as links between the different entities.
Furthermore, some information about the definitional status
of functions (like external, complete, or
overlapping definitions) are provided and combined with
documentation comments provided by the programmer.

A \emph{documentation comment}
\index{documentation comment}\index{comment!documentation}
starts at the beginning of a line
with \ccode{--- }\pindex{---} (also in literate programs!).
All documentation comments immediately before a
definition of a datatype or (top-level) function are kept together.\footnote{%
The documentation tool recognizes this association from the first identifier
in a program line. If one wants to add a documentation comment
to the definition of a function which is an infix operator,
the first line of the operator definition should be a type definition,
otherwise the documentation comment is not recognized.}
The documentation comments for the complete module occur before
the first ``module'' or ``import'' line in the module.
The comments can also contain several special tags. These tags
must be the first thing on its line (in the documentation comment)
and continues until the next tag is encountered or until the
end of the comment. The following tags are recognized:
\begin{description}
\item[\code{@author}]\pindex{"@author} \emph{comment}\\
Specifies the author of a module (only reasonable in module comments).
\item[\code{@version}]\pindex{"@version} \emph{comment}\\
Specifies the version of a module (only reasonable in module comments).
\item[\code{@cons}]\pindex{"@cons} \emph{id} \emph{comment}\\
A comment for the constructor \emph{id} of a datatype
(only reasonable in datatype comments).
\item[\code{@param}]\pindex{"@param} \emph{id} \emph{comment}\\
A comment for function parameter \emph{id}
(only reasonable in function comments).
Due to pattern matching, this need not be the name of a parameter
given in the declaration of the function but all parameters
for this functions must be commented in left-to-right order
(if they are commented at all).
\item[\code{@return}]\pindex{"@return} \emph{comment}\\
A comment for the return value of a function
(only reasonable in function comments).
\end{description}
The comment of a documented entity can be any string
in HTML format, i.e., special characters like \ccode{<}
must be quoted (e.g., \ccode{\&lt;}).
It can also contain HTML tags.
However, header tags like \code{<h1>} should not be used
since the structuring is generated by CurryDoc.
For the same reason, preformatted text (tag \code{<pre>})
should not be used since the formatting is the task of CurryDoc.

The following example text shows a Curry program with some
documentation comments:
{\renewcommand{\baselinestretch}{1.0}
\begin{verbatim}
--- This is an
--- example module.
--- @author Michael Hanus
--- @version 0.1

module Example where

--- The function conc concatenates two lists.
--- @param xs - the first list
--- @param ys - the second list
--- @return a list containing all elements of xs and ys
conc []     ys = ys
conc (x:xs) ys = x : conc xs ys
-- this comment will not be included in the documentation

--- The function last computes the last element of a given list.
--- @param xs - the given input list
--- @return last element of the input list
last xs | conc ys [x] =:= xs  = x   where x,ys free

--- This datatype defines polymorphic trees.
--- @cons Leaf - a leaf of the tree
--- @cons Node - an inner node of the tree
data Tree a = Leaf a | Node [Tree a]
\end{verbatim}
}
\noindent
To generate the documentation, execute the command\pindex{currydoc}
\startprog
currydoc --html Example
\stopprog
(\code{currydoc} is a command usually stored in \code{$kics2home$/bin}
where $kics2home$ is the installation directory of KiCS2;
see Section~\ref{sec-general}).
This command creates the directory \code{DOC_Example} (if it does not exist)
and puts all HTML documentation files for the main program module
\code{Example}
and all its imported modules in this directory together with
a main index file \code{index.html}.
If one prefers another directory for the documentation files,
one can also execute the command
\startprog
currydoc --html docdir Example
\stopprog
where \code{docdir} is the directory for the documentation files.

In order to generate the common documentation for large collections
of Curry modules (e.g., the libraries contained in the KiCS2 distribution),
one can call \code{currydoc} with the following options:
\begin{description}
\item[\code{currydoc --noindexhtml docdir Mod}~:]
\pindex{noindex}
This command generates the documentation for module \code{Mod}
in the directory \code{docdir} without the index pages (i.e., main index page
and index pages for all functions and constructors defined in \code{Mod}
and its imported modules).
\item[\code{currydoc --onlyindexhtml docdir Mod1 Mod2 \ldots Mod$n$}~:]
\pindex{onlyindex}
This command generates only the index pages (i.e., a main index page
and index pages for all functions and constructors defined in the modules
\code{Mod1}, \code{M2},\ldots,\code{Mod$n$} and their imported modules)
in the directory \code{docdir}.
\end{description}

\newpage

\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
\label{sec-currybrowser}

\cb is a tool to browse through the modules and functions
of a Curry application, show them in various formats,
and analyze their properties.\footnote{Although \cb is
implemented in Curry, some functionalities of it require an
installed graph visualization tool (dot \url{http://www.graphviz.org/}),
otherwise they have no effect.}
Moreover, it is constructed in a way so that
new analyzers can be easily connected to \cb.
A detailed description of the ideas behind this tool can be
found in \cite{Hanus05WCFLP,Hanus06WLPE}.

\cb is part of the KiCS2 distribution and can be
started in two ways:
\begin{itemize}
\item
In the command shell via the command: \code{$kics2home$/bin/currybrowser mod}
\item
In the KiCS2 environment after loading the module
\code{mod} and typing the command \ccode{:browse}.
\end{itemize}
Here, \ccode{mod} is the name of the main module of a Curry application.
After the start, \cb loads the interfaces of the main
module and all imported modules before a GUI is created
for interactive browsing.

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.7]{currybrowser.jpg}
\end{center}
\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
\end{figure}
%
To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
shows a snapshot of its use on a particular application
(here: the implementation of \cb).
The upper list box in the left column shows the modules and their imports
in order to browse through the modules of an application.
Similarly to directory browsers, the list of imported modules of a module
can be opened or closed by clicking.
After selecting a module in the list of modules, its source code,
interface, or various other formats of the module can be shown
in the main (right) text area. For instance, one can show
pretty-printed versions of the intermediate flat programs (see below)
in order to see how local function definitions are translated by lambda lifting
\cite{Johnsson85}
or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
Since Curry is a language with parametric polymorphism and type inference,
programmers often omit the type signatures when defining functions.
Therefore, one can also view (and store) the selected module as source code where
missing type signatures are added.

Below the list box for selecting modules, there is a menu
(``Analyze selected module'') to analyze all functions
of the currently selected module at once. This is useful
to spot some functions of a module that could be problematic
in some application contexts, like functions that are impure (i.e., the result
depends on the evaluation time) or partially defined (i.e.,
not evaluable on all ground terms).
If such an analysis is selected,
the names of all functions are shown in the
lower list box of the left column (the ``function list'')
with prefixes indicating the properties of the individual functions.

The function list box can be also filled with functions
via the menu ``Select functions''. For instance, all functions
or only the exported functions defined in the currently selected
module can be shown there, or all functions from different modules
that are directly or indirectly called from
a currently selected function.
This list box is central to focus on a function in the
source code of some module or to analyze some function,
i.e., showing their properties. In order to focus on a function,
it is sufficient to check the ``focus on code'' button.
To analyze an individually selected function, one can
select an analysis from the list of available program analyses
(through the menu ``Select analysis'').
In this case, the analysis results are either shown
in the text box below the main text area
or visualized by separate tools, e.g., by a graph drawing tool for
visualizing call graphs.
Some analyses are local, i.e., they need only to consider the local definition
of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
``Pattern completeness''),
where other analyses are global, i.e.,
they consider the definitions of all functions directly or indirectly called
by this function (e.g., ``Depends on,'' ``Solution complete,''
``Set-valued'').
%
Finally, there are a few additional tools integrated into \cb,
for instance, to visualize the import relation between all modules
as a dependency graph. These tools are available through the ``Tools'' menu.

More details about the use of \cb and all built-in analyses
are available through the ``Help'' menu of \cb.

\newpage

\section{CurryTest: A Tool for Testing Curry Programs}
\label{sec-currytest}

CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
is a simple tool in the KiCS2 distribution to write
and run repeatable tests. CurryTest simplifies the task
of writing test cases for a module and executing them.
The tool is easy to use. Assume one has implemented a module \code{MyMod}
and wants to write some test cases to test its functionality,
making regression tests in future versions, etc.
For this purpose, there is a system library \code{Assertion}
(Section~ref{Library:Assertion}) which
contains the necessary definitions for writing tests.
In particular, it exports an abstract polymorphic type \ccode{Assertion a}
together with the following operations:
\startprog
assertTrue      :: String -> Bool -> Assertion ()
assertEqual     :: String -> a -> a -> Assertion a
assertValues    :: String -> a -> [a] -> Assertion a
assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
assertIO        :: String -> IO a -> a -> Assertion a
assertEqualIO   :: String -> IO a -> IO a -> Assertion a
\stopprog
The expression \ccode{assertTrue $s$ $b$}
is an assertion (named $s$) that the expression $b$ has the value \code{True}.
Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
asserts that the expressions $e_1$ and $e_2$
must be equal (i.e., \code{$e_1$==$e_2$} must hold),
the expression \ccode{assertValues $s$ $e$ $vs$} asserts
that $vs$ is the multiset of all values of $e$,
and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
that the constraint abstraction $c$ has the multiset of solutions $vs$.
Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
asserts that the I/O action $a$ yields the value $v$ whenever it is
executed, and
the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
asserts that the I/O actions $a_1$ and $a_2$ yields equal values.
The name $s$ provided as a first argument in each assertion
is used in the protocol produced by the test tool.

One can define a test program by importing the module
to be tested together with the module \code{Assertion} and defining
top-level functions of type \code{Assertion} in this module
(which must also be exported).
As an example, consider the following program
that can be used to test some list processing functions:
\startprog
\medskip
import List
import Assertion
\medskip
test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
\medskip
test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
\medskip
test3 = assertSolutions "prefix" (\labs{}x -> let y free in  x\,++\,y =:= [1,2])
                                 [[],[1],[1,2]]
\medskip
\stopprog
For instance, \code{test1} asserts that the result of evaluating the
expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.

We can execute a test suite by the command\pindex{currytest}
\startprog
currytest testList
\stopprog
(\code{currytest} is a program stored in \code{$kics2home$/bin}
where $kics2home$ is the installation directory of KiCS2;
see Section~\ref{sec-general}).
In our example, \ccode{testList.curry} is the program containing the
definition of all assertions. This has the effect
that all exported top-level functions
of type \code{Assertion} are tested (i.e., the corresponding
assertions are checked) and the results
(\ccode{OK} or failure) are reported together with the name of each assertion.
%If failures occur, the complete test results are also
%written into a file named \ccode{testList.testlog}.''
For our example above, we obtain the following successful protocol:
\startprog
============================================================
Testing module "testList"...
OK: ++
OK: all
OK: prefix
All tests successfully passed.
============================================================
\stopprog
There is also a graphical interface that summarizes the results
more nicely.
In order to start this interface, one has to add the parameter
\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
\startprog
currytest --window testList
\stopprog
or
\startprog
currytest -w testList
\stopprog
A snapshot of the interface is shown in Figure~\ref{fig-currytest}.

\begin{figure}%[t]
\begin{center}
\includegraphics[scale=0.7]{currytest.jpg}
\end{center}
\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
\end{figure}

\newpage

\section{Technical Problems}

One can implement distributed systems with KiCS2
by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
that supports a socket communication with symbolic names
rather than natural numbers.
For instance, this library is the basis of programming
dynamic web pages with the libraries
\code{HTML} (Section~\ref{Library:HTML})
or \code{WUI} (Section~\ref{Library:WUI}).
However, it might be possible that some technical problems
arise due to the use of named sockets.
Therefore, this section gives some information
about the technical requirements of KiCS2 and how to solve
problems due to these requirements.

There is one fixed port that is used by the implementation of KiCS2:
\begin{description}
\item[Port 8767:] This port is used by the
{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
named sockets in Curry.
If some other process uses this port on the machine,
the distribution facilities defined in the module \code{NamedSocket}
cannot be used.
\end{description}
If these features do not work, you can try to find out
whether this port is in use by the shell command
\ccode{netstat -a | fgrep 8767} (or similar).

The CPNS is implemented as a demon listening on its port 8767
in order to serve requests about registering a new symbolic
name for a named socket or asking the physical port number
of an registered named socket. The demon will be automatically started for
the first time on a machine when a user runs a program
using named sockets.
It can also be manually started and terminated by the
scripts \code{$kics2home$/cpns/start} and
\code{$kics2home$/cpns/stop}.
If the demon is already running, the command \code{$kics2home$/cpns/start}
does nothing (so it can be always executed
before invoking a Curry program using named sockets).

If you detect any further technical problem,
please write to
\begin{center}
\code{mh@informatik.uni-kiel.de}
\end{center}

\newpage

\addcontentsline{toc}{section}{Bibliography}
\bibliography{mh}
\bibliographystyle{plain}

\newpage
\appendix

\section{Libraries of the KiCS2 Distribution}
\label{sec:libraries}

{\setlength{\parindent}{0.0cm}

The KiCS2 distribution comes with an extensive collection
of libraries for application programming.
The libraries for meta-programming by representing
Curry programs as datatypes in Curry are described
in the following subsection in more detail.
The complete set of libraries with all exported types and functions
are described in the further subsections.
For a more detailed online documentation of all libraries of KiCS2,
see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.

\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
\label{sec-flatcurry}

\index{AbstractCurry}
\index{FlatCurry}
To support meta-programming, i.e., the manipulation of Curry programs
in Curry, there are system modules
\code{FlatCurry} (Section~\ref{Library:FlatCurry})
and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
stored in the directory \ccode{$kics2home$/lib/meta},
which define datatypes for the representation
of Curry programs.
\code{AbstractCurry} is a more direct representation of a Curry program,
whereas \code{FlatCurry} is a simplified representation
where local function definitions are replaced by global definitions
(i.e., lambda lifting has been performed) and pattern matching
is translated into explicit case/or expressions.
Thus, \code{FlatCurry} can be used for more back-end oriented
program manipulations (or, for writing new back ends for Curry),
whereas \code{AbstractCurry} is intended for manipulations of
programs that are more oriented towards the source program.

Both modules contain predefined I/O actions to read programs
in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
or \code{FlatCurry}
(\code{readFlatCurry}\pindex{readFlatCurry}) format.
These actions parse the corresponding source program and return
a data term representing this program (according to the definitions
in the modules \code{AbstractCurry} and \code{FlatCurry}).

Since all datatypes are explained in detail in these modules,
we refer to the online documentation\footnote{%
\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
of these modules.

As an example, consider a program file \ccode{test.curry}
containing the following two lines:
\startprog
rev []     = []
rev (x:xs) = (rev xs) ++ [x]
\stopprog
Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
following term:
\startprog
 (Prog "test"
  ["Prelude"]
  []
  [Func ("test","rev") 1 Public
        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
                  (TCons ("Prelude","[]") [(TVar 0)]))
        (Rule [0]
           (Case Flex (Var 1)
              [Branch (Pattern ("Prelude","[]") [])
                  (Comb ConsCall ("Prelude","[]") []),
               Branch (Pattern ("Prelude",":") [2,3])
                  (Comb FuncCall ("Prelude","++")
                        [Comb FuncCall ("test","rev") [Var 3],
                         Comb ConsCall ("Prelude",":")
                              [Var 2,Comb ConsCall ("Prelude","[]") []]
                        ])
              ]))]
  []
 )
\stopprog

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Definitions in order to LaTeX documents generated by "currydoc --tex"
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\currymodule}[2]{\subsubsection{Library #1}\label{Library:#1} #2}
\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
\newcommand{\currytypesstop}{}
\newcommand{\currytype}[3]{{\tt type #2}\pindex{#1} \begin{quote} #3 \end{quote}}
\newcommand{\currydata}[3]{{\tt data #1}\pindex{#1} \begin{quote}#2\end{quote}%
\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:} #3\end{itemize}}
\newcommand{\currycons}[3]{\item {\tt #1~::~#2}\pindex{#1}\par #3}
\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
\newcommand{\curryfuncstop}{}
\newcommand{\curryfunction}[3]{#2\pindex{#1}  \begin{quote}#3\end{quote}}
\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}

\subsection{General Libraries}

\input{lib/AllSolutions}
\input{lib/Assertion}
\input{lib/Char}
\input{lib/Combinatorial}
\input{lib/Constraint}
\input{lib/CSV}
\input{lib/Directory}
\input{lib/FileGoodies}
\input{lib/Float}
\input{lib/Global}
\input{lib/GUI}
\input{lib/Integer}
\input{lib/IO}
\input{lib/IOExts}
\input{lib/JavaScript}
\input{lib/KeyDatabaseSQLite}
\input{lib/List}
\input{lib/Maybe}
\input{lib/NamedSocket}
\input{lib/Parser}
\input{lib/Pretty}
\input{lib/Profile}
\input{lib/PropertyFile}
\input{lib/Read}
\input{lib/ReadNumeric}
\input{lib/ReadShowTerm}
\input{lib/SearchTree}
\input{lib/Socket}
\input{lib/System}
\input{lib/Time}
\input{lib/Unsafe}

\subsection{Data Structures and Algorithms}

\input{lib/Array}
\input{lib/Dequeue}
\input{lib/FiniteMap}
\input{lib/GraphInductive}
\input{lib/Random}
\input{lib/RedBlackTree}
\input{lib/SetRBT}
\input{lib/Sort}
\input{lib/TableRBT}
\input{lib/Traversal}

\subsection{Libraries for Web Applications}

\input{lib/CategorizedHtmlList}
\input{lib/HTML}
\input{lib/HtmlParser}
\input{lib/Mail}
\input{lib/WUI}
\input{lib/URL}
\input{lib/XML}
\input{lib/XmlConv}

\subsection{Libraries for Meta-Programming}

\input{lib/AbstractCurry}
\input{lib/AbstractCurryPrinter}
\input{lib/CompactFlatCurry}
\input{lib/CurryStringClassifier}
\input{lib/FlatCurry}
\input{lib/FlatCurryGoodies}
\input{lib/FlatCurryRead}
\input{lib/FlatCurryShow}
\input{lib/FlatCurryXML}
\input{lib/FlexRigid}
\input{lib/PrettyAbstract}

} % end setlength parindent

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Auxiliary Files}
\label{sec-auxfiles}

During the translation and execution of a Curry program with KiCS2,
various intermediate representations of the source program are created
and stored in different files which are shortly explained in this section.
In general, it is not necessary to know about
these auxiliary files because they are automatically generated
and updated. You should only remember the command for deleting
all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
to clean up your directories.

The various components of KiCS2 create the following auxiliary files.
\begin{description}
\item[\code{prog.fcy}:] This file contains the Curry program
in the so-called ``FlatCurry'' representation where all functions are global
(i.e., lambda lifting has been performed) and pattern matching
is translated into explicit case/or expressions
(compare Appendix~\ref{sec-flatcurry}).
This representation might be useful for other back ends and
compilers for Curry and is the basis doing meta-programming in Curry.
This file is implicitly
generated when a program is compiled with KiCS2.
The FlatCurry representation of a Curry program is usually
generated by the front-end after parsing, type checking and eliminating
local declarations.
If $dir$ is the directory where the Curry program is stored,
the corresponding FlatCurry program is stored in the directory
\ccode{$dir$/.curry}.

\item[\code{prog.fint}:] This file contains the interface
of the program in the so-called ``FlatCurry'' representation,
i.e., it is similar to \code{prog.fcy} but contains only exported
entities and the bodies of all functions omitted (i.e., ``external'').
This representation is useful for providing a fast access
to module interfaces.
This file is implicitly generated when a program is compiled with KiCS2
and stored in the same directory as \code{prog.fcy}.

\item[\code{Curry_prog.nda}:] This file contains some information about
the determinism behavior of operations that is used by the
KiCS2 compiler (see \cite{BrasselHanusPeemoellerReck11}
for more details about the use of this information).
If $dir$ is the directory where the Curry program is stored,
the corresponding Haskell program is stored in the directory
\ccode{$dir$/.curry/.kics2}.

\item[\code{Curry_prog.info}:] This file contains some
information about the top-level functions of module \code{prog}
that are used by the interactive environment,
like determinism behavior or IO status.
This file is stored in the same directory as \code{Curry_prog.nda}.

\item[\code{Curry_prog.hs}:] This file contains a Haskell program
as the result of translating the Curry program with the
KiCS2 compiler.
This file is stored in the same directory as \code{Curry_prog.nda}.

\item[\code{Curry_prog.o}:] This file contains the object code
of the Haskell program \code{Curry_prog.hs} when the latter program is
compiled in order to execute it.
This file is stored in the same directory as \code{Curry_prog.hs}.

\item[\code{Curry_prog.hi}:] This file contains the interface
of the Haskell program \code{Curry_prog.hs} when the latter program is
compiled in order to execute it.
This file is stored in the same directory as \code{Curry_prog.hs}.

\item[\code{prog}:] This file contains the executable
after compiling and saving a program with KiCS2
(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).

\end{description}

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{External Operations}
\label{sec:external-operations}

\index{operation!external}\index{external operation}
Currently, KiCS2 has no general interface to external operations,
i.e., operations whose semantics is not defined by program rules
in a Curry program but by some code written in another programming
language.
Therefore, if a new external operation should be added
to the system, this operation must be declared as \code{external}
in the Curry source code
and an implementation for this external operation
must be provided for the run-time system.
An external operation is defined as follows in the Curry source code:
\begin{enumerate}
\item
Add a type declaration for the external operation somewhere
in a module defining this operation (usually, the prelude
or some system module).
\item
For external operations it is not allowed to define any
rule since their semantics is determined by an external implementation.
Instead of the defining rules, you have to write
\startprog
f external
\stopprog
somewhere in the file containing the type declaration for 
the external operation \code{f}.
\end{enumerate}
Furthermore, an implementation of the external operation
must be provided in the target language of the KiCS2 compiler,
i.e., in Haskell, and inserted in the compiled code.
In order to simplify this task, KiCS2 follows some code
conventions that are described in the following.

Assume you want to implement your own concatenation
for strings in a module \code{String}.
The name and type of this string concatenation should be
\startprog
sconc :: String -> String -> String
\stopprog
Since the primitive Haskell implementation of this operation
does not now anything about the operational mechanism of Curry
(e.g., needed narrowing, non-deterministic rewriting),
the arguments need to be completely evaluated before
the primitive implementation is called.
This can be easily obtained by the prelude operation \code{(\$\#\#)}
that applies an operation to the \emph{normal form} of the given
argument, i.e., this operation evaluates the argument
to its normal form before applying the operation to it.\footnote{%
There is also a similar prelude operation \code{(\$\#)}
which evaluates the argument only to head-normal
which is a bit more efficient and can be used for unstructured
types like \code{Bool}.}
Thus, we define \code{sconc} by
\startprog
sconc :: String -> String -> String
sconc s1 s2 = (prim_sconc \$\#\# s1) \$\#\# s2
~
prim_sconc :: String -> String -> String
prim_sconc external
\stopprog
so that it is ensured that the external operation \code{prim_sconc}
is always called with complete evaluated arguments.

In order to define the Haskell code implementing \code{prim_sconc},
one has to satisfy the naming conventions of KiCS2.
The KiCS2 compiler generates the following code for the
external operation \code{prim_sconc} (note that the generated
Haskell code for the module \code{String} is stored in the file
\code{.curry/kics2/Curry_String.hs}):
\startprog
d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
d_C_prim_sconc x1 x2 = external_d_C_prim_sconc x1 x2
\stopprog
The type constructors \code{OP_List} and \code{C_Char}
of the prelude \code{Curry_Prelude}\footnote{Note that all
translated Curry modules are imported in the Haskell code fully qualified
in order to avoid name conflicts.}
correspond to the Curry type constructors for lists and characters.
The Haskell operation \code{external_d_C_prim_sconc}
is the external operation to be implemented in Haskell by the
programmer. If \code{String.curry} contains the code
of the Curry function \code{sconc} described above,
the Haskell code implementing the external operations
occurring in the module \code{String} must be in the
file \code{External_String.hs} which is located in the same
directory as the file \code{String.curry}.
The KiCS2 compiler appends the code contained in
\code{External_String.hs} to the generated code
stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
If the file \code{External_String.hs} contains also
some import declarations at the beginning, these import declarations
are put after the generated import declarations.}

In order to complete our example, we have to write into the
file \code{External_String.hs} a definition of the Haskell function
\code{external_d_C_prim_sconc}.
Thus, we start with the following definitions:
\startprog
import qualified Curry_Prelude as CP
~
external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
                        -> CP.OP_List CP.C_Char
\stopprog
First, we import the standard prelude with the name \code{CP}
in order to shorten the writing of type declarations.
In order to write the final code of this operation,
we have to convert the Curry-related types
(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
Note that the Curry-related types contain information about
non-deterministic or constrained values
(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
that are meaningless in Haskell.
To solve this conversion problem, the implementation of KiCS2
provides a family of operations to perform these conversions
for the predefined types occurring in the standard prelude.
For instance, \code{fromCurry} converts a Curry type into the
corresponding Haskell type, and \code{toCurry} converts
the Haskell type into the corresponding Curry type.
Thus, we complete our example with the definition
\startprog
external_d_C_prim_sconc s1 s2 =
  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
\stopprog
Here, we use Haskell's concatenation operation \ccode{++}
to concatenate the string arguments.
The type annotation \ccode{:: String} is necessary
because \ccode{++} is a polymorphic function
so that the type inference system of Haskell
has problems to determine the right instance of the conversion
function.

In order to use Haskell I/O actions as an implementation
for external Curry I/O actions in an easy manner, the KiCS2 implementation
also provides a family of operations \code{fromHaskellIO$n$},
where $n$ denotes the number of parameters of the I/O action,
which map a Haskell I/O action into a Curry I/O action and performing
all data conversions. For instance, if we want to implement
an external operation to print some string as an output line,
we start by declaring the external operations in the Curry module \code{String}:
\startprog
printString :: String -> IO ()
printString s = prim_printString \$\#\# s
~
prim_printString :: String -> IO ()
prim_printString external
\stopprog
Next we add the corresponding implementation in the file
\code{External_String.hs} (where \code{OP_Unit} is the name
of the Haskell representation of the Curry datatype \ccode{()}):
\startprog
external_d_C_prim_printString :: CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
external_d_C_prim_printString = fromHaskellIO1 putStrLn
\stopprog
When we compile the Curry module \code{String},
KiCS2 combines these definitions in the target program so that
we can immediately use the externally defined operation
\code{printString} in Curry programs.

As we have seen, KiCS2 transforms a name like
\code{primOP} of an external operation into the name
\code{external_d_C_primOP} for the Haskell operation
to be implemented, i.e., only a specific prefix is added.
However, this is only valid if no special characters occur
in the Curry names.
Otherwise (in order to generate a correct Haskell program),
special characters are translated into specific
names prefixed by \ccode{OP_}. For instance,
if we declare the external operation
\startprog
(<\&>) :: Int -> Int -> Int
(<\&>) external
\stopprog
the generated Haskell module contains the code
\startprog
d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
                     -> Curry_Prelude.C_Int
d_OP_lt_ampersand_gt x1 x2 = external_d_OP_lt_ampersand_gt x1 x2
\stopprog
so that one has to implement the operation
\code{external_d_OP_lt_ampersand_gt} in Haskell.
If in doubt, one should look into the generated Haskell code
about the names and types of the operations to be implemented.

Finally, note that this method to connect functions implemented in Haskell
to Curry programs provides the opportunity to connect
also operations written in other programming languages
to Curry via Haskell's foreign function interface.

\newpage
\addcontentsline{toc}{section}{Index}
\printindex

\end{document}

% LocalWords:  KiCS

### 2011-10-08 docs/src/manual.tex
More option explanations

@@
 Similarly to \code{dfs} but use a breadth-first search strategy
-to print the values.
+to compute and print the values of the given expression.
+
+\item[\fbox{\code{ids}}]\pindex{ids}
+Similarly to \code{dfs} but use an iterative-deepening strategy
+to compute and print the values of the initial expression.
+The initial depth bound is 100 and the depth-bound is
+doubled after each iteration.
 
 \item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-with an initial depth bound of $n$. The depth-bound is
-duplicated after each iteration.
-If $n$ is missing, the default value $n=100$ is used.
+Similarly to \code{ids} but use an initial depth bound of $n$.
+
+\item[\fbox{\code{par}}]\pindex{par}
+Similarly to \code{dfs} but use a parallel search strategy
+to compute and print the values of the initial expression.
+The system chooses an appropriate number
+of threads according the current number of available processors.
 
 \item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-with $n$ threads.
-If $n$ is missing, the system chooses an appropriate number
-of threads according the current number of available processors.
+Similarly to \code{par} but use $n$ parallel threads.
 
@@
 Define additional options passed to the KiCS2 compiler.
+For instance, setting the option
+\startprog
+:set copts -O 0
+\stopprog
+has the effect that all optimizations performed by the KiCS2 compiler
+are turned off.
 
 \item[\fbox{\code{rts $o$}}]\pindex{rts}
-Define additional run-time options passed to the Glasgow Haskell Compiler
-(by \ccode{+RTS $o$ -RTS}).
+Define additional run-time options passed to the executable
+generated by the Glasgow Haskell Compiler, i.e., the parameters
+\ccode{+RTS $o$ -RTS} are passed to the executable.
+For instance, setting the option
+\startprog
+:set rts -H512m
+\stopprog
+has the effect that the minimum heap size is set to 512 megabytes.
 

### 2011-10-10 runtime/Types.hs
NonDet instance for primitive functions implemented

@@ -172,12 +172,12 @@ x =:= y = match uniChoice uniNarrowed uniFree failCons uniGuard uniVal x
     uniNarrowed i xs  = checkFail (choicesCons i (map (\x' -> x' =:= y) xs)) y
     uniFree i _       = bindTo y
      where
-      bindTo = match bindChoice bindNarrowed bindFree failCons bindGuard bindVal 
+      bindTo = match bindChoice bindNarrowed bindFree failCons bindGuard bindVal
       bindChoice j y1 y2 = choiceCons j  (bindTo y1) (bindTo y2)
       bindNarrowed j ys  = choicesCons j (map bindTo ys)
       bindFree j _       = guardCons [i :=: BindTo j] C_Success
       bindGuard c e      = guardCons c (bindTo e)
-      bindVal v          = guardCons (bind i v) C_Success   
+      bindVal v          = guardCons (bind i v) C_Success
     uniGuard c e      = checkFail (guardCons c (e =:= y)) y
     uniVal v          = uniWith y
      where
@@ -207,7 +207,7 @@ x =:<= y = match uniChoice uniNarrowed uniFree failCons uniGuard uniVal x
   uniFree i _       = guardCons [i :=: LazyBind (lazyBind i y)] C_Success
   -- constructor-rooted term
   uniVal vx = unifyWith y
-   where 
+   where
     unifyWith = match uniyChoice uniyNarrowed uniyFree failCons uniyGuard uniyVal
     uniyChoice j y1 y2  = choiceCons j (unifyWith y1) (unifyWith y2)
     uniyNarrowed j ys   = choicesCons j (map unifyWith ys)
@@ -357,24 +357,24 @@ instance Show (a -> b) where
 instance Read (a -> b) where
   readsPrec = error "read for function is undefined"
 
-instance NonDet (a -> b) where
-  choiceCons  = error "choiceCons for function is undefined"
-  choicesCons = error "choicesCons for function is undefined"
-  failCons    = error "failCons for function is undefined"
-  guardCons   = error "guardCons for function is undefined"
-  try         = Val
+instance NonDet b => NonDet (a -> b) where
+  choiceCons    i f g = \ x -> choiceCons i (f x) (g x)
+  choicesCons    i fs = \ x -> choicesCons i (map ($x) fs)
+  failCons            = \ _ -> failCons
+  guardCons       c f = \ x -> guardCons c (f x)
+  try                 = Val
   match _ _ _ _ _ f x = f x
 
[…]

### 2011-10-12 docs/src/manual.tex
CurryDoc documentation improved
Date updated

@@
 {\huge User Manual}\\[4ex]
-Version of September 29, 2011\\[6ex]
+Version of October 12, 2011\\[6ex]
 \Large
@@
 
+\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
+\label{sec-erd2curry}
+
+ERD2Curry\index{ERD2Curry}\index{database programming}
+is a tool to generate Curry code to access and manipulate data
+persistently stored from
+entity relationship diagrams.\index{entity relationship diagrams}
+The idea of this tool is described in detail in
+\cite{BrasselHanusMueller08PADL}.
+Thus, we describe only the basic steps to use this tool
+in the following.
+
+If one creates an entity relationship diagram (ERD)
+with the Umbrello UML Modeller, one has to store its
+XML description in XMI format (as offered by Umbrello)
+in a file, e.g., \ccode{myerd.xmi}.
+This description can be compiled into a Curry program by the
+command\pindex{erd2curry}
+\startprog
+erd2curry myerd.xmi
+\stopprog
+(\code{erd2curry} is a program stored in \code{$kics2home$/bin}
+where $kics2home$ is the installation directory of KiCS2;
+see Section~\ref{sec-general}).
+If \code{MyData} is the name of the ERD, the Curry program file
+\ccode{MyData.curry} is generated containing all the necessary
+database access code as described in \cite{BrasselHanusMueller08PADL}.
+
+If one does not want to use the Umbrello UML Modeller,
+one can also create a textual description of the ERD
+as a Curry term of type \code{ERD}
+(w.r.t.\ the type definition given in module
+\code{$kics2home$/tools/erd2curry/ERD.curry})
+and store it in some file, e.g., \ccode{myerd.term}.
+This description can be compiled into a Curry program by the
+command\pindex{erd2curry}
+\startprog
+erd2curry -t myerd.term
+\stopprog
+%
+There is also the possibility to visualize an ERD term
+as a graph with the graph visualization program \code{dotty}
+(for this purpose, it might be necessary to adapt the definition
+of \code{dotviewcommand} in your \ccode{.kics2rc} file,
+see Section~\ref{sec-customization},
+according to your local environment).
+This can be done by the command
+\startprog
+erd2curry -v myerd.term
+\stopprog
+
+\paragraph{Inclusion in the Curry application:}
+To compile the generated database code, either
+include the directory \code{$kics2home$/tools/erd2curry}
+into your Curry load path
+(e.g., by setting  the environment variable
+\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
+or copy the file
+\code{$kics2home$/tools/erd2curry/ERDGeneric.curry}
+into the directory of the generated database code.
+
+\newpage
+
 \section{Technical Problems}
@@
 
-\newcommand{\currymodule}[2]{\subsubsection{Library #1}\label{Library:#1} #2}
+\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
 \newcommand{\currytypesstart}{\subsubsection*{Exported types:}}

### 2011-10-17 lib/External_System.hs
imports changed to comply with Haskell 2010 and GHC 7.*

@@ -1,9 +1,13 @@
 {-# LANGUAGE MagicHash, MultiParamTypeClasses #-}
-import System
-import System.Posix.Process(getProcessID)
-import System.CPUTime(getCPUTime)
-import Network.BSD(getHostName)
+
 import qualified Curry_Prelude as CP
+
+import Network.BSD (getHostName)
+import System.Cmd
+import System.CPUTime (getCPUTime)
+import System.Environment
+import System.Exit
+import System.Posix.Process (getProcessID)
 
 external_d_C_getCPUTime :: CP.C_IO CP.C_Int
 external_d_C_getCPUTime =

### 2011-10-17 docs/src/manual.tex

@@
 {\huge User Manual}\\[4ex]
-Version of October 12, 2011\\[6ex]
+Version of September 29, 2011\\[6ex]
 \Large

### 2011-10-28 docs/src/manual.tex
markdown support added to CurryDoc

--- 2011-10-12 docs/src/manual.tex
@@
 {\huge User Manual}\\[4ex]
-Version of October 12, 2011\\[6ex]
+Version of October 28, 2011\\[6ex]
 \Large
@@
 
+The documentation comment for the module (i.e., the documentation
+comment at the beginning of the module before the module header)
+can be formatted with
+\href{http://en.wikipedia.org/wiki/Markdown}{Markdown's syntax}\index{markdown}
+provided that the
+\href{http://johnmacfarlane.net/pandoc/}{pandoc}\index{pandoc}
+ tool is installed.
+If \code{pandoc} is not installed, one should run CurryDoc
+with the parameter \ccode{--nopandoc}.
+
 The following example text shows a Curry program with some

### 2011-10-31 docs/src/currydoc.tex
CurryDoc improvements

--- 2011-10-28 docs/src/manual.tex
@@
-\documentclass[11pt,fleqn]{article}
-
-\usepackage{latexsym}
-\usepackage{makeidx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{graphicx}
-
-%%% ------------------------------------------------------------------
-
-\usepackage[colorlinks,linkcolor=blue]{hyperref}
-\hypersetup{bookmarksopen=true}
-\hypersetup{bookmarksopenlevel=0}
-\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
-\hypersetup{pdfauthor={Michael Hanus}}
-%\hypersetup{pdfstartview=Title}
-\hypersetup{pdfstartview=FitH}
-\usepackage{thumbpdf}
-
-%%% ------------------------------------------------------------------
-
-\setlength{\textwidth}{16.5cm}
-\setlength{\textheight}{23cm}
-\renewcommand{\baselinestretch}{1.1}
-\setlength{\topmargin}{-1cm}
-\setlength{\oddsidemargin}{0cm}
-\setlength{\evensidemargin}{0cm}
-\setlength{\marginparwidth}{0.0cm}
-\setlength{\marginparsep}{0.0cm}
-
-\newlength{\figurewidth}
-\setlength{\figurewidth}{\textwidth}
-\addtolength{\figurewidth}{-0.4cm}
-
-% font for program texts
-\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\tt}
-
-% environment for typing program texts:
-\makeatletter
-\newenvironment{prog}{\par\vspace{0.7ex}
-\setlength{\parindent}{1.0cm}
-\setlength{\parskip}{-0.1ex}
-\obeylines\@vobeyspaces\tt}{\vspace{0.7ex}\noindent
-}
-\makeatother
-\newcommand{\startprog}{\begin{prog}}
-\newcommand{\stopprog}{\end{prog}\noindent}
-
-% program text in normal text
-\newcommand{\code}[1]{\mbox{\codefont #1}}
-
-% program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\mbox{\codefont #1}''}
-
-\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index
-
-\newcommand{\labs}{\mbox{\tt\char92}}  % lambda abstraction in Curry
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
-\newcommand{\cb}{CurryBrowser\xspace}
-
-% allow underscores in programs:
-\catcode`\_=\active
-\let_=\sb
-\catcode`\_=12
-
-% produce an index:
-\makeindex
-
-\begin{document}
-\sloppy
-
-\begin{titlepage}
-\pdfbookmark[1]{Title}{Title}
-\begin{center}
-\fbox{
-\begin{minipage}[t]{\figurewidth}
-\begin{center}\vspace{10ex}
-{\Huge\bf KiCS2 0.1}\\[4ex]
-{\huge The Kiel Curry System (Version 2)}\\[7ex]
-{\huge User Manual}\\[4ex]
-Version of October 28, 2011\\[6ex]
-\Large
-Michael Hanus$^1$ [editor] \\[3ex]
-{\large Additional Contributors:}\\[2ex]
-Bernd Bra\ss{}el$^2$ \\
-Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
-\normalsize
-(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
-(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
-(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
-\end{center}
-\end{minipage}}
-\end{center}
-\end{titlepage}
-
-\pdfbookmark[1]{Contents}{Contents}
-\tableofcontents
-
-\newpage
-
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes KiCS2 (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current KiCS2 implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, KiCS2 also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus06Curry}.
-Therefore, this document only explains the use of the different
-components of KiCS2
-and the differences and restrictions of KiCS2
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.2).
-The basic ideas of the implementation of KiCS2
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of KiCS2}
-
-\subsection{General Use}
-\label{sec-general}
-
-This version of KiCS2 has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems
-(see the file \code{INSTALL.html} in the KiCS2 directory
-for a description of the necessary software to install KiCS2).
-
-All executables required to use the different components
-of KiCS2 are stored in the directory \code{$kics2home$/bin}
-(where $kics2home$ is the installation directory of the complete
-KiCS2 installation). You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=$kics2home$/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with KiCS2 creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the KiCS2 system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\startprog
-cleancurry
-\stopprog
-(this is a shell script stored in the \code{bin} directory of the
-KiCS2 installation, see above).
-The command
-\startprog
-cleancurry -r
-\stopprog
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions on Curry Programs}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with KiCS2:
-\begin{itemize}
-\item \index{singleton variables}\index{variables!singleton}
-\emph{Singleton variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-KiCS2 translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-%\item Threads created by a concurrent conjunction are not executed
-%in a fair manner (usually, threads corresponding to leftmost constraints
-%are executed with higher priority).
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, KiCS2 provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in KiCS2}
-\label{sec-modules}
-
-The current implementation of KiCS2 supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-KiCS2 searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a KiCS2 session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{$kics2home$/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of KiCS2}
-\label{sec-kics2}
-
-This section describes the interactive environment
-KiCS2\index{KiCS2}
-that supports the development of applications written in Curry.
-The implementation of KiCS2 contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking KiCS2}
-\label{sec:invoke-kics2}
-
-To start KiCS2, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{$kics2home$/bin} where $kics2home$ is the installation directory
-of KiCS2).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{$kics2home$/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke KiCS2 with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\startprog
-kics2 :load Mod :add List
-\stopprog
-starts KiCS2, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\startprog
-kics2 :load Mod :eval config
-\stopprog
-starts KiCS2, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\startprog
-kics2 :load Mod :save :quit
-\stopprog
-starts KiCS2, loads the main module \code{Mod}, creates an executable,
-and terminates KiCS2. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of KiCS2}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of KiCS2 are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs, i.e., in order to see the results of their bindings,
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\startprog
-xs++ys\,=:=\,[1,2]  where xs,ys free
-\stopprog
-in order to obtain the following three possible bindings:
-\startprog
-\{xs = [], ys = [1,2]\} Success
-\{xs = [1], ys = [2]\} Success
-\{xs = [1,2], ys = []\} Success
-\stopprog
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\startprog
-xs++ys=:=[z] where xs,ys,z free
-\stopprog
-they are specialized to the type \ccode{()}
-(since the current implementation of KiCS2 cannot compute
-with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of KiCS2.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the KiCS2 environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current KiCS2 process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Option of KiCS2}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of KiCS2.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression.
-This mode is only useful for debugging purposes.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-Use the implementation $i$ for the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the values \code{ghc} (default), \code{integer},
-or \code{ioref} are supported for $i$.
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-In this mode, the computation is performed in a separate
-window created by the command \code{interactivecommand}
-whose value is defined in \ccode{.kics2rc}
-(see Section~\ref{sec-customization}).
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{copts $o$}}]\pindex{copts}
-Define additional options passed to the KiCS2 compiler.
-For instance, setting the option
-\startprog
-:set copts -O 0
-\stopprog
-has the effect that all optimizations performed by the KiCS2 compiler
-are turned off.
-
-\item[\fbox{\code{rts $o$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\startprog
-:set rts -H512m
-\stopprog
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-KiCS2 requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\startprog
-\{-\# KiCS2_OPTION $opt$ \#-\}
-\stopprog
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the KiCS2 command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\startprog
-\{-\# KiCS2_OPTION ids \#-\}
-\{-\# KiCS2_OPTION +ghci \#-\}
-\{-\# KiCS2_OPTION v2 \#-\}
-module M where
-$\ldots$
-\stopprog
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of KiCS2 (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by KiCS2 if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call KiCS2 with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of KiCS2 to your own preferences,
-there is a configuration file which is read by KiCS2 when it is invoked.
-When you start KiCS2 for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started KiCS2 for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of KiCS2 contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into KiCS2
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{$kics2home$/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
-\section{Extensions}
-\label{sec-extensions}
-
-KiCS2 supports some extensions in Curry programs that are not (yet)
-part of the definition of Curry. These extensions are described below.
-
-\subsection{Recursive Variable Bindings}
-
-Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in KiCS2.
-For instance, the declaration
-\startprog
-ones5 = let ones = 1 : ones
-         in take 5 ones
-\stopprog
-introduces the local variable \code{ones} which is bound
-to a \emph{cyclic structure}\index{cyclic structure}
-representing an infinite list of \code{1}'s.
-Similarly, the definition
-\startprog
-onetwo n = take n one2
- where
-   one2 = 1 : two1
-   two1 = 2 : one2
-\stopprog
-introduces a local variables \code{one2} that represents
-an infinite list of alternating \code{1}'s and \code{2}'s
-so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.
-
-\subsection{Functional Patterns}
-
-Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
-defining operations with functional patterns avoids problems
-caused by strict equality (\ccode{=:=}) and leads to programs
-that are potentially more efficient.
-
-Consider the definition of an operation to compute the last element
-of a list \code{xs} based on the prelude operation \ccode{++}
-for list concatenation:
-\startprog
-last xs | ys++[y] =:= xs  = y   where y,ys free
-\stopprog
-Since the equality constraint \ccode{=:=} evaluates both sides
-to a constructor term, all elements of the list \code{xs} are
-fully evaluated in order to satisfy the constraint.
-
-Functional patterns can help to improve this computational behavior.
-A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
-is a function call at a pattern position. With functional patterns,
-we can define the operation \code{last} as follows:
-\startprog
-last (_++[y]) = y
-\stopprog
-This definition is not only more compact but also avoids the complete
-evaluation of the list elements: since a functional pattern is considered
-as an abbreviation for the set of constructor terms obtained by all
-evaluations of the functional pattern to normal form (see
-\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
-definition is conceptually equivalent to the set of rules
-\startprog
-last [y] = y
-last [_,y] = y
-last [_,_,y] = y
-\ldots
-\stopprog
-which shows that the evaluation of the list elements is not demanded
-by the functional pattern.
-
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
-is interpreted as a functional pattern if $f$ is not a visible constructor
-but a defined function that is visible in the scope of the pattern.
-
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are 
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there. 
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\startprog
-type $R$ $\alpha_1$ \ldots $\alpha_n$ = \{ $l_1$ :: $\tau_1$, \ldots, $l_m$ :: $\tau_m$ \}
-\stopprog
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$ 
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\startprog
-type Person = \{name :: String, age :: Int\}
-type Address = \{person :: Person, street :: String, city :: String\}
-type Branch a b = \{left :: a, right :: b\}
-\stopprog
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\startprog
-type Address = \{person :: Person, street,city :: String\}
-\stopprog
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\startprog
-type Address = \{street,city :: String, person :: Person\}
-\stopprog
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\startprog
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\stopprog
-\startprog
-getName :: Person -> String
-getName \ldots
-\stopprog
-
-Labels can only be used in the context of
-records. They do not share the name space with 
-functions/constructors/variables or type constructors/type variables. 
-For instance it is possible to use 
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually 
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\startprog
-\{ $l_1$ = $v_1$, \ldots, $l_m$ = $v_m$ \}
-\stopprog
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with 
-exactly one assignment. Examples for record constructions are
-\startprog
-\{name = "Johnson", age = 30\}     -- generates a record of type 'Person'
-\{left = True, right = 20\}        -- generates a record of type 'Branch'
-\stopprog
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\startprog
-\{age = 30, name = "Johnson"\}     -- generates a record of type 'Person'
-\stopprog
-Unlike labeled fields in record type declarations, 
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\startprog
-\{person = \{name = "Smith", age = 20\},   -- generates a record of
- street = "Main Street",                -- type 'Address'
- city   = "Springfield"\}
-\stopprog
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records. 
-It has the following form:
-\startprog
-$r$ -> $l$
-\stopprog
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\startprog
-pers -> name
-\stopprog
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\startprog
-(addr -> person) -> age
-\stopprog
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}). When a field selection is used in
-expressions, it has to be parenthesized.
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\startprog
-\{$l_1$ := $v_1$, \ldots, $l_k$ := $v_k$ | $r$\}
-\stopprog
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration 
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record. 
-Assignments can occur in an arbitrary order. It is not allowed to 
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\startprog
-\{name := "Scott", age := 25 | pers\}
-\{person := \{name := "Scott", age := 25 | pers\} | addr\}
-\stopprog
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\startprog
-\{$l_1$ = $p_1$, \ldots, $l_n$ = $p_n$\}
-\{$l_1$ = $p_1$, \ldots, $l_k$ = $p_k$ | _\}
-\stopprog
-In both cases each label $l_i$ is specified with a pattern $p_i$. 
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern. 
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the 
-patterns of the specified labels are matched against 
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\startprog
-isSmith30 :: Person -> Bool
-isSmith30 \{name = "Smith", age = 30\} = True
-\stopprog
-\startprog
-startsWith :: Char -> Person -> Bool
-startsWith c \{name = (d:_) | _\} = c == d
-\stopprog
-\startprog
-getPerson :: Address -> Person
-getPerson \{person = p | _\} = p
-\stopprog
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways 
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, KiCS2/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent 
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
-
-\newpage
-
 \section{CurryDoc: A Documentation Generator for Curry Programs}
-\label{sec-currydoc}
 
@@
  tool is installed.
+Other documentations comments, like comments for operations
+or data types, can contain Markdown annotations for
+emphasizing elements (e.g., \verb!*emph*! or \verb!_verb_!),
+strong elements (e.g., \verb!**important**! or \verb!__important__!),
+or code elements (e.g., \verb!`3+4`!).
 If \code{pandoc} is not installed, one should run CurryDoc
-with the parameter \ccode{--nopandoc}.
+with the parameter \ccode{--nomarkdown}.
 
@@
 
---- This datatype defines polymorphic trees.
+--- This data type defines polymorphic trees.
 --- @cons Leaf - a leaf of the tree
@@
 \end{description}
-
-\newpage
-
-\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
-\label{sec-currybrowser}
-
-\cb is a tool to browse through the modules and functions
-of a Curry application, show them in various formats,
-and analyze their properties.\footnote{Although \cb is
-implemented in Curry, some functionalities of it require an
-installed graph visualization tool (dot \url{http://www.graphviz.org/}),
-otherwise they have no effect.}
-Moreover, it is constructed in a way so that
-new analyzers can be easily connected to \cb.
-A detailed description of the ideas behind this tool can be
-found in \cite{Hanus05WCFLP,Hanus06WLPE}.
-
-\cb is part of the KiCS2 distribution and can be
-started in two ways:
-\begin{itemize}
-\item
-In the command shell via the command: \code{$kics2home$/bin/currybrowser mod}
-\item
-In the KiCS2 environment after loading the module
-\code{mod} and typing the command \ccode{:browse}.
-\end{itemize}
-Here, \ccode{mod} is the name of the main module of a Curry application.
-After the start, \cb loads the interfaces of the main
-module and all imported modules before a GUI is created
-for interactive browsing.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.7]{currybrowser.jpg}
-\end{center}
-\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
-\end{figure}
-%
-To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
-shows a snapshot of its use on a particular application
-(here: the implementation of \cb).
-The upper list box in the left column shows the modules and their imports
-in order to browse through the modules of an application.
-Similarly to directory browsers, the list of imported modules of a module
-can be opened or closed by clicking.
-After selecting a module in the list of modules, its source code,
-interface, or various other formats of the module can be shown
-in the main (right) text area. For instance, one can show
-pretty-printed versions of the intermediate flat programs (see below)
-in order to see how local function definitions are translated by lambda lifting
-\cite{Johnsson85}
-or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
-Since Curry is a language with parametric polymorphism and type inference,
-programmers often omit the type signatures when defining functions.
-Therefore, one can also view (and store) the selected module as source code where
-missing type signatures are added.
-
-Below the list box for selecting modules, there is a menu
-(``Analyze selected module'') to analyze all functions
-of the currently selected module at once. This is useful
-to spot some functions of a module that could be problematic
-in some application contexts, like functions that are impure (i.e., the result
-depends on the evaluation time) or partially defined (i.e.,
-not evaluable on all ground terms).
-If such an analysis is selected,
-the names of all functions are shown in the
-lower list box of the left column (the ``function list'')
-with prefixes indicating the properties of the individual functions.
-
-The function list box can be also filled with functions
-via the menu ``Select functions''. For instance, all functions
-or only the exported functions defined in the currently selected
-module can be shown there, or all functions from different modules
-that are directly or indirectly called from
-a currently selected function.
-This list box is central to focus on a function in the
-source code of some module or to analyze some function,
-i.e., showing their properties. In order to focus on a function,
-it is sufficient to check the ``focus on code'' button.
-To analyze an individually selected function, one can
-select an analysis from the list of available program analyses
-(through the menu ``Select analysis'').
-In this case, the analysis results are either shown
-in the text box below the main text area
-or visualized by separate tools, e.g., by a graph drawing tool for
-visualizing call graphs.
-Some analyses are local, i.e., they need only to consider the local definition
-of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
-``Pattern completeness''),
-where other analyses are global, i.e.,
-they consider the definitions of all functions directly or indirectly called
-by this function (e.g., ``Depends on,'' ``Solution complete,''
-``Set-valued'').
-%
-Finally, there are a few additional tools integrated into \cb,
-for instance, to visualize the import relation between all modules
-as a dependency graph. These tools are available through the ``Tools'' menu.
-
-More details about the use of \cb and all built-in analyses
-are available through the ``Help'' menu of \cb.
-
-\newpage
-
-\section{CurryTest: A Tool for Testing Curry Programs}
-\label{sec-currytest}
-
-CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the KiCS2 distribution to write
-and run repeatable tests. CurryTest simplifies the task
-of writing test cases for a module and executing them.
-The tool is easy to use. Assume one has implemented a module \code{MyMod}
-and wants to write some test cases to test its functionality,
-making regression tests in future versions, etc.
-For this purpose, there is a system library \code{Assertion}
-(Section~ref{Library:Assertion}) which
-contains the necessary definitions for writing tests.
-In particular, it exports an abstract polymorphic type \ccode{Assertion a}
-together with the following operations:
-\startprog
-assertTrue      :: String -> Bool -> Assertion ()
-assertEqual     :: String -> a -> a -> Assertion a
-assertValues    :: String -> a -> [a] -> Assertion a
-assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
-assertIO        :: String -> IO a -> a -> Assertion a
-assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\stopprog
-The expression \ccode{assertTrue $s$ $b$}
-is an assertion (named $s$) that the expression $b$ has the value \code{True}.
-Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
-asserts that the expressions $e_1$ and $e_2$
-must be equal (i.e., \code{$e_1$==$e_2$} must hold),
-the expression \ccode{assertValues $s$ $e$ $vs$} asserts
-that $vs$ is the multiset of all values of $e$,
-and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
-that the constraint abstraction $c$ has the multiset of solutions $vs$.
-Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
-asserts that the I/O action $a$ yields the value $v$ whenever it is
-executed, and
-the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yields equal values.
-The name $s$ provided as a first argument in each assertion
-is used in the protocol produced by the test tool.
-
-One can define a test program by importing the module
-to be tested together with the module \code{Assertion} and defining
-top-level functions of type \code{Assertion} in this module
-(which must also be exported).
-As an example, consider the following program
-that can be used to test some list processing functions:
-\startprog
-\medskip
-import List
-import Assertion
-\medskip
-test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-\medskip
-test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-\medskip
-test3 = assertSolutions "prefix" (\labs{}x -> let y free in  x\,++\,y =:= [1,2])
-                                 [[],[1],[1,2]]
-\medskip
-\stopprog
-For instance, \code{test1} asserts that the result of evaluating the
-expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.
-
-We can execute a test suite by the command\pindex{currytest}
-\startprog
-currytest testList
-\stopprog
-(\code{currytest} is a program stored in \code{$kics2home$/bin}
-where $kics2home$ is the installation directory of KiCS2;
-see Section~\ref{sec-general}).
-In our example, \ccode{testList.curry} is the program containing the
-definition of all assertions. This has the effect
-that all exported top-level functions
-of type \code{Assertion} are tested (i.e., the corresponding
-assertions are checked) and the results
-(\ccode{OK} or failure) are reported together with the name of each assertion.
-%If failures occur, the complete test results are also
-%written into a file named \ccode{testList.testlog}.''
-For our example above, we obtain the following successful protocol:
-\startprog
-============================================================
-Testing module "testList"...
-OK: ++
-OK: all
-OK: prefix
-All tests successfully passed.
-============================================================
-\stopprog
-There is also a graphical interface that summarizes the results
-more nicely.
-In order to start this interface, one has to add the parameter
-\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\startprog
-currytest --window testList
-\stopprog
-or
-\startprog
-currytest -w testList
-\stopprog
-A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
-
-\begin{figure}%[t]
-\begin{center}
-\includegraphics[scale=0.7]{currytest.jpg}
-\end{center}
-\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
-\end{figure}
-
-\newpage
-
-\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
-\label{sec-erd2curry}
-
-ERD2Curry\index{ERD2Curry}\index{database programming}
-is a tool to generate Curry code to access and manipulate data
-persistently stored from
-entity relationship diagrams.\index{entity relationship diagrams}
-The idea of this tool is described in detail in
-\cite{BrasselHanusMueller08PADL}.
-Thus, we describe only the basic steps to use this tool
-in the following.
-
-If one creates an entity relationship diagram (ERD)
-with the Umbrello UML Modeller, one has to store its
-XML description in XMI format (as offered by Umbrello)
-in a file, e.g., \ccode{myerd.xmi}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\startprog
-erd2curry myerd.xmi
-\stopprog
-(\code{erd2curry} is a program stored in \code{$kics2home$/bin}
-where $kics2home$ is the installation directory of KiCS2;
-see Section~\ref{sec-general}).
-If \code{MyData} is the name of the ERD, the Curry program file
-\ccode{MyData.curry} is generated containing all the necessary
-database access code as described in \cite{BrasselHanusMueller08PADL}.
-
-If one does not want to use the Umbrello UML Modeller,
-one can also create a textual description of the ERD
-as a Curry term of type \code{ERD}
-(w.r.t.\ the type definition given in module
-\code{$kics2home$/tools/erd2curry/ERD.curry})
-and store it in some file, e.g., \ccode{myerd.term}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\startprog
-erd2curry -t myerd.term
-\stopprog
-%
-There is also the possibility to visualize an ERD term
-as a graph with the graph visualization program \code{dotty}
-(for this purpose, it might be necessary to adapt the definition
-of \code{dotviewcommand} in your \ccode{.kics2rc} file,
-see Section~\ref{sec-customization},
-according to your local environment).
-This can be done by the command
-\startprog
-erd2curry -v myerd.term
-\stopprog
-
-\paragraph{Inclusion in the Curry application:}
-To compile the generated database code, either
-include the directory \code{$kics2home$/tools/erd2curry}
-into your Curry load path
-(e.g., by setting  the environment variable
-\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
-or copy the file
-\code{$kics2home$/tools/erd2curry/ERDGeneric.curry}
-into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with KiCS2
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of KiCS2 and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of KiCS2:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{$kics2home$/cpns/start} and
-\code{$kics2home$/cpns/stop}.
-If the demon is already running, the command \code{$kics2home$/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
-\addcontentsline{toc}{section}{Bibliography}
-\bibliography{mh}
-\bibliographystyle{plain}
-
-\newpage
-\appendix
-
-\section{Libraries of the KiCS2 Distribution}
-\label{sec:libraries}
-
-{\setlength{\parindent}{0.0cm}
-
-The KiCS2 distribution comes with an extensive collection
-of libraries for application programming.
-The libraries for meta-programming by representing
-Curry programs as datatypes in Curry are described
-in the following subsection in more detail.
-The complete set of libraries with all exported types and functions
-are described in the further subsections.
-For a more detailed online documentation of all libraries of KiCS2,
-see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
-
-\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
-\label{sec-flatcurry}
-
-\index{AbstractCurry}
-\index{FlatCurry}
-To support meta-programming, i.e., the manipulation of Curry programs
-in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{$kics2home$/lib/meta},
-which define datatypes for the representation
-of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
-where local function definitions are replaced by global definitions
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
-program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
-programs that are more oriented towards the source program.
-
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
-These actions parse the corresponding source program and return
-a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
-
-Since all datatypes are explained in detail in these modules,
-we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
-of these modules.
-
-As an example, consider a program file \ccode{test.curry}
-containing the following two lines:
-\startprog
-rev []     = []
-rev (x:xs) = (rev xs) ++ [x]
-\stopprog
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
-following term:
-\startprog
- (Prog "test"
-  ["Prelude"]
-  []
-  [Func ("test","rev") 1 Public
-        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
-                  (TCons ("Prelude","[]") [(TVar 0)]))
-        (Rule [0]
-           (Case Flex (Var 1)
-              [Branch (Pattern ("Prelude","[]") [])
-                  (Comb ConsCall ("Prelude","[]") []),
-               Branch (Pattern ("Prelude",":") [2,3])
-                  (Comb FuncCall ("Prelude","++")
-                        [Comb FuncCall ("test","rev") [Var 3],
-                         Comb ConsCall ("Prelude",":")
-                              [Var 2,Comb ConsCall ("Prelude","[]") []]
-                        ])
-              ]))]
-  []
- )
-\stopprog
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Definitions in order to LaTeX documents generated by "currydoc --tex"
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
-\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
-\newcommand{\currytypesstop}{}
-\newcommand{\currytype}[3]{{\tt type #2}\pindex{#1} \begin{quote} #3 \end{quote}}
-\newcommand{\currydata}[3]{{\tt data #1}\pindex{#1} \begin{quote}#2\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:} #3\end{itemize}}
-\newcommand{\currycons}[3]{\item {\tt #1~::~#2}\pindex{#1}\par #3}
-\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
-\newcommand{\curryfuncstop}{}
-\newcommand{\curryfunction}[3]{#2\pindex{#1}  \begin{quote}#3\end{quote}}
-\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
-
-\subsection{General Libraries}
-
-\input{lib/AllSolutions}
-\input{lib/Assertion}
-\input{lib/Char}
-\input{lib/Combinatorial}
-\input{lib/Constraint}
-\input{lib/CSV}
-\input{lib/Directory}
-\input{lib/FileGoodies}
-\input{lib/Float}
-\input{lib/Global}
-\input{lib/GUI}
-\input{lib/Integer}
-\input{lib/IO}
-\input{lib/IOExts}
-\input{lib/JavaScript}
-\input{lib/KeyDatabaseSQLite}
-\input{lib/List}
-\input{lib/Maybe}
-\input{lib/NamedSocket}
-\input{lib/Parser}
-\input{lib/Pretty}
-\input{lib/Profile}
-\input{lib/PropertyFile}
-\input{lib/Read}
-\input{lib/ReadNumeric}
-\input{lib/ReadShowTerm}
-\input{lib/SearchTree}
-\input{lib/Socket}
-\input{lib/System}
-\input{lib/Time}
-\input{lib/Unsafe}
-
-\subsection{Data Structures and Algorithms}
-
-\input{lib/Array}
-\input{lib/Dequeue}
-\input{lib/FiniteMap}
-\input{lib/GraphInductive}
-\input{lib/Random}
-\input{lib/RedBlackTree}
-\input{lib/SetRBT}
-\input{lib/Sort}
-\input{lib/TableRBT}
-\input{lib/Traversal}
-
-\subsection{Libraries for Web Applications}
-
-\input{lib/CategorizedHtmlList}
-\input{lib/HTML}
-\input{lib/HtmlParser}
-\input{lib/Mail}
-\input{lib/WUI}
-\input{lib/URL}
-\input{lib/XML}
-\input{lib/XmlConv}
-
-\subsection{Libraries for Meta-Programming}
-
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
-\input{lib/CompactFlatCurry}
-\input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
-
-} % end setlength parindent
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with KiCS2,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of KiCS2 create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with KiCS2.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with KiCS2
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-KiCS2 compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-KiCS2 compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with KiCS2
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, KiCS2 has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Therefore, if a new external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\startprog
-f external
-\stopprog
-somewhere in the file containing the type declaration for 
-the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the KiCS2 compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, KiCS2 follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\startprog
-sconc :: String -> String -> String
-\stopprog
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal
-which is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\startprog
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc \$\#\# s1) \$\#\# s2
-~
-prim_sconc :: String -> String -> String
-prim_sconc external
-\stopprog
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of KiCS2.
-The KiCS2 compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\startprog
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 = external_d_C_prim_sconc x1 x2
-\stopprog
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the
-programmer. If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The KiCS2 compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\startprog
-import qualified Curry_Prelude as CP
-~
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> CP.OP_List CP.C_Char
-\stopprog
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of KiCS2
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-\startprog
-external_d_C_prim_sconc s1 s2 =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\stopprog
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-In order to use Haskell I/O actions as an implementation
-for external Curry I/O actions in an easy manner, the KiCS2 implementation
-also provides a family of operations \code{fromHaskellIO$n$},
-where $n$ denotes the number of parameters of the I/O action,
-which map a Haskell I/O action into a Curry I/O action and performing
-all data conversions. For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\startprog
-printString :: String -> IO ()
-printString s = prim_printString \$\#\# s
-~
-prim_printString :: String -> IO ()
-prim_printString external
-\stopprog
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{OP_Unit} is the name
-of the Haskell representation of the Curry datatype \ccode{()}):
-\startprog
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString = fromHaskellIO1 putStrLn
-\stopprog
-When we compile the Curry module \code{String},
-KiCS2 combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, KiCS2 transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\startprog
-(<\&>) :: Int -> Int -> Int
-(<\&>) external
-\stopprog
-the generated Haskell module contains the code
-\startprog
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 = external_d_OP_lt_ampersand_gt x1 x2
-\stopprog
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
-\addcontentsline{toc}{section}{Index}
-\printindex
-
-\end{document}
-
-% LocalWords:  KiCS

### 2011-10-31 docs/src/manual.tex
CurryDoc improvements

--- 2011-10-12 docs/src/manual.tex
@@
 
-\section{CurryDoc: A Documentation Generator for Curry Programs}
-\label{sec-currydoc}
-
-CurryDoc\index{CurryDoc}
-\index{program!documentation}\index{documentation generator}
-is a tool in the KiCS2 distribution that generates
-the documentation for a Curry program (i.e., the main module
-and all its imported modules) in HTML format.
-The generated HTML pages contain information about
-all data types and functions exported by a module as well
-as links between the different entities.
-Furthermore, some information about the definitional status
-of functions (like external, complete, or
-overlapping definitions) are provided and combined with
-documentation comments provided by the programmer.
-
-A \emph{documentation comment}
-\index{documentation comment}\index{comment!documentation}
-starts at the beginning of a line
-with \ccode{--- }\pindex{---} (also in literate programs!).
-All documentation comments immediately before a
-definition of a datatype or (top-level) function are kept together.\footnote{%
-The documentation tool recognizes this association from the first identifier
-in a program line. If one wants to add a documentation comment
-to the definition of a function which is an infix operator,
-the first line of the operator definition should be a type definition,
-otherwise the documentation comment is not recognized.}
-The documentation comments for the complete module occur before
-the first ``module'' or ``import'' line in the module.
-The comments can also contain several special tags. These tags
-must be the first thing on its line (in the documentation comment)
-and continues until the next tag is encountered or until the
-end of the comment. The following tags are recognized:
-\begin{description}
-\item[\code{@author}]\pindex{"@author} \emph{comment}\\
-Specifies the author of a module (only reasonable in module comments).
-\item[\code{@version}]\pindex{"@version} \emph{comment}\\
-Specifies the version of a module (only reasonable in module comments).
-\item[\code{@cons}]\pindex{"@cons} \emph{id} \emph{comment}\\
-A comment for the constructor \emph{id} of a datatype
-(only reasonable in datatype comments).
-\item[\code{@param}]\pindex{"@param} \emph{id} \emph{comment}\\
-A comment for function parameter \emph{id}
-(only reasonable in function comments).
-Due to pattern matching, this need not be the name of a parameter
-given in the declaration of the function but all parameters
-for this functions must be commented in left-to-right order
-(if they are commented at all).
-\item[\code{@return}]\pindex{"@return} \emph{comment}\\
-A comment for the return value of a function
-(only reasonable in function comments).
-\end{description}
-The comment of a documented entity can be any string
-in HTML format, i.e., special characters like \ccode{<}
-must be quoted (e.g., \ccode{\&lt;}).
-It can also contain HTML tags.
-However, header tags like \code{<h1>} should not be used
-since the structuring is generated by CurryDoc.
-For the same reason, preformatted text (tag \code{<pre>})
-should not be used since the formatting is the task of CurryDoc.
-
-The following example text shows a Curry program with some
-documentation comments:
-{\renewcommand{\baselinestretch}{1.0}
-\begin{verbatim}
---- This is an
---- example module.
---- @author Michael Hanus
---- @version 0.1
-
-module Example where
-
---- The function conc concatenates two lists.
---- @param xs - the first list
---- @param ys - the second list
---- @return a list containing all elements of xs and ys
-conc []     ys = ys
-conc (x:xs) ys = x : conc xs ys
--- this comment will not be included in the documentation
-
---- The function last computes the last element of a given list.
---- @param xs - the given input list
---- @return last element of the input list
-last xs | conc ys [x] =:= xs  = x   where x,ys free
-
---- This datatype defines polymorphic trees.
---- @cons Leaf - a leaf of the tree
---- @cons Node - an inner node of the tree
-data Tree a = Leaf a | Node [Tree a]
-\end{verbatim}
-}
-\noindent
-To generate the documentation, execute the command\pindex{currydoc}
-\startprog
-currydoc --html Example
-\stopprog
-(\code{currydoc} is a command usually stored in \code{$kics2home$/bin}
-where $kics2home$ is the installation directory of KiCS2;
-see Section~\ref{sec-general}).
-This command creates the directory \code{DOC_Example} (if it does not exist)
-and puts all HTML documentation files for the main program module
-\code{Example}
-and all its imported modules in this directory together with
-a main index file \code{index.html}.
-If one prefers another directory for the documentation files,
-one can also execute the command
-\startprog
-currydoc --html docdir Example
-\stopprog
-where \code{docdir} is the directory for the documentation files.
-
-In order to generate the common documentation for large collections
-of Curry modules (e.g., the libraries contained in the KiCS2 distribution),
-one can call \code{currydoc} with the following options:
-\begin{description}
-\item[\code{currydoc --noindexhtml docdir Mod}~:]
-\pindex{noindex}
-This command generates the documentation for module \code{Mod}
-in the directory \code{docdir} without the index pages (i.e., main index page
-and index pages for all functions and constructors defined in \code{Mod}
-and its imported modules).
-\item[\code{currydoc --onlyindexhtml docdir Mod1 Mod2 \ldots Mod$n$}~:]
-\pindex{onlyindex}
-This command generates only the index pages (i.e., a main index page
-and index pages for all functions and constructors defined in the modules
-\code{Mod1}, \code{M2},\ldots,\code{Mod$n$} and their imported modules)
-in the directory \code{docdir}.
-\end{description}
+\input{currydoc}
 

### 2011-11-02 docs/src/currydoc.tex
CurryDoc and comments improved

@@
 comment at the beginning of the module before the module header)
-can be formatted with
-\href{http://en.wikipedia.org/wiki/Markdown}{Markdown's syntax}\index{markdown}
-provided that the
-\href{http://johnmacfarlane.net/pandoc/}{pandoc}\index{pandoc}
- tool is installed.
-Other documentations comments, like comments for operations
-or data types, can contain Markdown annotations for
+can be formatted with a subset of
+\href{http://en.wikipedia.org/wiki/Markdown}{Markdown's syntax}\index{markdown}.
+For instance, it can contain Markdown annotations for
 emphasizing elements (e.g., \verb!*emph*! or \verb!_verb_!),
 strong elements (e.g., \verb!**important**! or \verb!__important__!),
-or code elements (e.g., \verb!`3+4`!).
-If \code{pandoc} is not installed, one should run CurryDoc
+code elements (e.g., \verb!`3+4`!), code blocks (lines prefixed by four blanks),
+unordered lists (lines prefixed by two blanks followed by ``\verb!* !''),
+ordered lists (lines prefixed by two blanks followed by a digit and a dot),
+quotations (lines prefixed by ``\verb!> !''),
+and web links of the form \ccode{<http://\ldots>}
+or \ccode{[link text](http://\ldots)}.
+If the Markdown syntax should not be used, one could run CurryDoc
 with the parameter \ccode{--nomarkdown}.
@@
 
---- The function conc concatenates two lists.
+--- The function `conc` concatenates two lists.
 --- @param xs - the first list
 --- @param ys - the second list
---- @return a list containing all elements of xs and ys
+--- @return a list containing all elements of `xs` and `ys`
 conc []     ys = ys
@@
 
---- The function last computes the last element of a given list.
+--- The function `last` computes the last element of a given list.
 --- @param xs - the given input list
@@
 
---- This data type defines polymorphic trees.
+--- This data type defines _polymorphic_ trees.
 --- @cons Leaf - a leaf of the tree

### 2011-11-04 docs/src/currydoc.tex
Markdown lib and doc added

@@
 can be formatted with a subset of
-\href{http://en.wikipedia.org/wiki/Markdown}{Markdown's syntax}\index{markdown}.
+\href{http://en.wikipedia.org/wiki/Markdown}{Markdown's syntax}\index{markdown}
+(which is described in detail in the appendix).
 For instance, it can contain Markdown annotations for
-emphasizing elements (e.g., \verb!*emph*! or \verb!_verb_!),
-strong elements (e.g., \verb!**important**! or \verb!__important__!),
+emphasizing elements (e.g., \verb!_verb_!),
+strong elements (e.g., \verb!**important**!),
 code elements (e.g., \verb!`3+4`!), code blocks (lines prefixed by four blanks),
-unordered lists (lines prefixed by two blanks followed by ``\verb!* !''),
-ordered lists (lines prefixed by two blanks followed by a digit and a dot),
+unordered lists (lines prefixed by  ``\verb! * !''),
+ordered lists (lines prefixed by blanks followed by a digit and a dot),
 quotations (lines prefixed by ``\verb!> !''),

### 2011-11-14 docs/src/currydoc.tex
CurryDoc improvements

@@
 \end{description}
-The comment of a documented entity can be any string
-in HTML format, i.e., special characters like \ccode{<}
-must be quoted (e.g., \ccode{\&lt;}).
-It can also contain HTML tags.
-However, header tags like \code{<h1>} should not be used
-since the structuring is generated by CurryDoc.
-For the same reason, preformatted text (tag \code{<pre>})
-should not be used since the formatting is the task of CurryDoc.
-
-The documentation comment for the module (i.e., the documentation
-comment at the beginning of the module before the module header)
-can be formatted with a subset of
+The comment of a documented entity can be any string in
 \href{http://en.wikipedia.org/wiki/Markdown}{Markdown's syntax}\index{markdown}
-(which is described in detail in the appendix).
+(the currently supported set of elements
+is described in detail in the appendix).
 For instance, it can contain Markdown annotations for
@@
 
+The comments can also contain markups in HTML format
+so that special characters like \ccode{<}
+must be quoted (e.g., \ccode{\&lt;}).
+However, header tags like \code{<h1>} should not be used
+since the structuring is generated by CurryDoc.
+In addition to Markdown or HTML markups,
+one can also mark \emph{references to names} of operations or data types
+in Curry programs which are translated into links inside
+the generated HTML documentation. Such references have to be
+enclosed in single quotes. For instance, the text
+\verb!'conc'! refers to the Curry operation \code{conc}
+inside the current module whereas the text
+\verb!'Prelude.reverse'! refers to the operation \code{reverse}
+of the module \code{Prelude}.
+If one wants to write single quotes without this specific
+meaning, one can escape them with a backslash:
+\begin{verbatim}
+--- This is a comment without a \'reference\'.
+\end{verbatim}
+To simplify the writing of documentation comments,
+such escaping is only necessary for single words,
+i.e., if the text inside quotes has not the syntax of
+an identifier, the escaping can be omitted, as in
+\begin{verbatim}
+--- This isn't a reference.
+\end{verbatim}
+%
 The following example text shows a Curry program with some
@@
 --- The function `last` computes the last element of a given list.
+--- It is based on the operation 'conc' to concatenate two lists.
 --- @param xs - the given input list
@@
 \startprog
-currydoc --html Example
+currydoc Example
 \stopprog
@@
 \startprog
-currydoc --html docdir Example
+currydoc docdir Example
 \stopprog
@@
 \end{description}
+
+%%% Local Variables: 
+%%% mode: pdflatex
+%%% TeX-master: "manual"
+%%% End:

### 2011-11-14 docs/src/manual.tex
CurryDoc improvements
Markdown lib and doc added

@@
 {\huge User Manual}\\[4ex]
-Version of October 28, 2011\\[6ex]
+Version of November 9, 2011\\[6ex]
 \Large
@@
 \newcommand{\currytypesstop}{}
-\newcommand{\currytype}[3]{{\tt type #2}\pindex{#1} \begin{quote} #3 \end{quote}}
-\newcommand{\currydata}[3]{{\tt data #1}\pindex{#1} \begin{quote}#2\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:} #3\end{itemize}}
-\newcommand{\currycons}[3]{\item {\tt #1~::~#2}\pindex{#1}\par #3}
+\newcommand{\currytypesynstart}[2]{{\tt type #2}\pindex{#1} \begin{quote}}
+\newcommand{\currytypesynstop}{\end{quote}}
+\newcommand{\currydatastart}[1]{{\tt data #1}\pindex{#1} \begin{quote}}
+\newcommand{\currydatacons}{\end{quote}%
+\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:}}
+\newcommand{\currydatastop}{\end{itemize}}
+\newcommand{\curryconsstart}[2]{\item {\tt #1~::~#2}\par}
 \newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
 \newcommand{\curryfuncstop}{}
-\newcommand{\curryfunction}[3]{#2\pindex{#1}  \begin{quote}#3\end{quote}}
+\newcommand{\curryfunctionstart}[2]{#2\pindex{#1}\begin{quote}}
+\newcommand{\curryfunctionstop}{\end{quote}}
 \newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
@@
 \input{lib/Mail}
+\input{lib/Markdown}
 \input{lib/WUI}
@@
 } % end setlength parindent
+
+\newpage
+
+\input{markdown_syntax}
 

### 2011-11-18 lib/External_Directory.hs
added a local and global constraint store which map identifier to values (only works with prdfs, ghc and integer ids)

@@ -3,49 +3,49 @@ import System.Directory
 import System.IO
 import qualified Curry_Prelude as CP
 
-external_d_C_prim_doesFileExist :: CP.C_String -> CP.C_IO CP.C_Bool
-external_d_C_prim_doesFileExist = fromHaskellIO1 doesFileExist
+external_d_C_prim_doesFileExist :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Bool
+external_d_C_prim_doesFileExist s _ = fromHaskellIO1 doesFileExist s
 
-external_d_C_prim_doesDirectoryExist :: CP.C_String -> CP.C_IO CP.C_Bool
-external_d_C_prim_doesDirectoryExist = fromHaskellIO1 doesDirectoryExist
+external_d_C_prim_doesDirectoryExist :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Bool
+external_d_C_prim_doesDirectoryExist s _ = fromHaskellIO1 doesDirectoryExist s
 
-external_d_C_prim_fileSize :: CP.C_String -> CP.C_IO CP.C_Int
-external_d_C_prim_fileSize = fromHaskellIO1
+external_d_C_prim_fileSize :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Int
+external_d_C_prim_fileSize s _ = fromHaskellIO1
   (\f -> do h <- openFile f ReadMode 
             i <- hFileSize h
             hClose h
             return i
-  )
+  ) s
 
-external_d_C_prim_getModificationTime :: CP.C_String
+external_d_C_prim_getModificationTime :: CP.C_String -> ConstStore
                                       -> CP.C_IO Curry_Time.C_ClockTime
-external_d_C_prim_getModificationTime = fromHaskellIO1 getModificationTime
+external_d_C_prim_getModificationTime s _ = fromHaskellIO1 getModificationTime s
 
-external_d_C_getCurrentDirectory :: CP.C_IO (CP.C_String)
-external_d_C_getCurrentDirectory =
+external_d_C_getCurrentDirectory :: ConstStore -> CP.C_IO (CP.C_String)
+external_d_C_getCurrentDirectory _ =
   fromHaskellIO0 System.Directory.getCurrentDirectory
 
-external_d_C_prim_setCurrentDirectory :: CP.C_String -> CP.C_IO CP.OP_Unit
-external_d_C_prim_setCurrentDirectory = fromHaskellIO1 setCurrentDirectory
[…]

### 2011-11-18 lib/External_Distribution.hs

@@ -1,23 +1,23 @@
 import Installation
 import qualified Curry_Prelude as CP
 
-external_d_C_curryCompiler :: CP.C_String
-external_d_C_curryCompiler = toCurry "kics2"
+external_d_C_curryCompiler ::  ConstStore -> CP.C_String
+external_d_C_curryCompiler _ = toCurry "kics2"
 
-external_d_C_curryCompilerMajorVersion :: CP.C_Int
-external_d_C_curryCompilerMajorVersion = toCurry (majorVersion :: Int)
+external_d_C_curryCompilerMajorVersion ::  ConstStore -> CP.C_Int
+external_d_C_curryCompilerMajorVersion _ = toCurry (majorVersion :: Int)
 
-external_d_C_curryCompilerMinorVersion :: CP.C_Int
-external_d_C_curryCompilerMinorVersion = toCurry (minorVersion ::Int)
+external_d_C_curryCompilerMinorVersion ::  ConstStore -> CP.C_Int
+external_d_C_curryCompilerMinorVersion _ = toCurry (minorVersion ::Int)
 
-external_d_C_curryRuntime :: CP.C_String
-external_d_C_curryRuntime = toCurry "ghc"
+external_d_C_curryRuntime ::  ConstStore -> CP.C_String
+external_d_C_curryRuntime _ = toCurry "ghc"
 
-external_d_C_curryRuntimeMajorVersion :: CP.C_Int
-external_d_C_curryRuntimeMajorVersion = toCurry (6::Int)
+external_d_C_curryRuntimeMajorVersion ::  ConstStore -> CP.C_Int
+external_d_C_curryRuntimeMajorVersion _ = toCurry (6::Int)
 
-external_d_C_curryRuntimeMinorVersion :: CP.C_Int
-external_d_C_curryRuntimeMinorVersion = toCurry (12::Int)
+external_d_C_curryRuntimeMinorVersion ::  ConstStore -> CP.C_Int
+external_d_C_curryRuntimeMinorVersion _ = toCurry (12::Int)
 
-external_d_C_installDir :: CP.C_String
-external_d_C_installDir = toCurry installDir
+external_d_C_installDir ::  ConstStore -> CP.C_String
+external_d_C_installDir _ = toCurry installDir

### 2011-11-18 lib/External_IOExts.hs

@@ -7,16 +7,16 @@ import System.IO
 import qualified Curry_Prelude as CP
 
 external_d_C_prim_execCmd :: CP.C_String
- -> CP.C_IO (CP.OP_Tuple3 Curry_IO.C_Handle Curry_IO.C_Handle Curry_IO.C_Handle)
-external_d_C_prim_execCmd = fromHaskellIO1
+ -> ConstStore -> CP.C_IO (CP.OP_Tuple3 Curry_IO.C_Handle Curry_IO.C_Handle Curry_IO.C_Handle)
+external_d_C_prim_execCmd str _ = fromHaskellIO1
   (\s -> do (h1,h2,h3,_) <- runInteractiveCommand s
-            return (OneHandle h1, OneHandle h2, OneHandle h3))
+            return (OneHandle h1, OneHandle h2, OneHandle h3)) str
 
-external_d_C_prim_connectToCmd :: CP.C_String -> CP.C_IO Curry_IO.C_Handle
-external_d_C_prim_connectToCmd = fromHaskellIO1
+external_d_C_prim_connectToCmd :: CP.C_String -> ConstStore -> CP.C_IO Curry_IO.C_Handle
+external_d_C_prim_connectToCmd str _ = fromHaskellIO1
   (\s -> do (hin,hout,herr,_) <- runInteractiveCommand s
             forkIO (forwardError herr)
-            return (InOutHandle hout hin))
+            return (InOutHandle hout hin)) str
 
 forwardError :: Handle -> IO ()
 forwardError h = do
@@ -34,15 +34,15 @@ type Assocs = [(String,String)]
 assocs :: IORef Assocs
 assocs = unsafePerformIO (newIORef [])
 
-external_d_C_prim_setAssoc :: CP.C_String -> CP.C_String -> CP.C_IO CP.OP_Unit
-external_d_C_prim_setAssoc = fromHaskellIO2
+external_d_C_prim_setAssoc :: CP.C_String -> CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
+external_d_C_prim_setAssoc str1 str2 _ = fromHaskellIO2
   (\key val -> do as <- readIORef assocs
-                  writeIORef assocs ((key,val):as))
+                  writeIORef assocs ((key,val):as)) str1 str2
 
-external_d_C_prim_getAssoc :: CP.C_String -> CP.C_IO (CP.C_Maybe (CP.C_String))
-external_d_C_prim_getAssoc = fromHaskellIO1
[…]

### 2011-11-18 lib/External_Prelude.hs

@@ -23,11 +23,11 @@ import PrimTypes
 class (Show a, Read a, NonDet a, Generable a, NormalForm a, Unifiable a)
       => Curry a where
   -- implementation of strict equalit (==) for a data type
-  (=?=) :: a -> a -> C_Bool
+  (=?=) :: a -> a -> ConstStore -> C_Bool
   (=?=) = error "(==) is undefined"
 
   -- implementation of less-or-equal (<=) for a data type
-  (<?=) :: a -> a -> C_Bool
+  (<?=) :: a -> a -> ConstStore -> C_Bool
   (<?=) = error "(<=) is undefined"
 
 instance Curry (PrimData a) where
@@ -36,24 +36,24 @@ instance Curry (PrimData a) where
 
 -- BEGIN GENERATED FROM PrimTypes.curry
 instance Curry_Prelude.Curry C_Success where
-  (=?=) (Choice_C_Success i x y) z = narrow i (x Curry_Prelude.=?= z) (y Curry_Prelude.=?= z)
-  (=?=) (Choices_C_Success i xs) y = narrows i (map ((Curry_Prelude.=?= y)) xs)
-  (=?=) (Guard_C_Success cs e) y = guardCons cs (e Curry_Prelude.=?= y)
-  (=?=) Fail_C_Success _ = failCons
-  (=?=) z (Choice_C_Success i x y) = narrow i (z Curry_Prelude.=?= x) (z Curry_Prelude.=?= y)
-  (=?=) y (Choices_C_Success i xs) = narrows i (map ((y Curry_Prelude.=?=)) xs)
-  (=?=) y (Guard_C_Success cs e) = guardCons cs (y Curry_Prelude.=?= e)
-  (=?=) _ Fail_C_Success = failCons
-  (=?=) C_Success C_Success = Curry_Prelude.C_True
-  (<?=) (Choice_C_Success i x y) z = narrow i (x Curry_Prelude.<?= z) (y Curry_Prelude.<?= z)
-  (<?=) (Choices_C_Success i xs) y = narrows i (map ((Curry_Prelude.<?= y)) xs)
-  (<?=) (Guard_C_Success cs e) y = guardCons cs (e Curry_Prelude.<?= y)
-  (<?=) Fail_C_Success _ = failCons
[…]

### 2011-11-18 lib/External_Profile.hs

@@ -3,17 +3,17 @@ import System.CPUTime
 import qualified Curry_Prelude as CP
 
 external_d_C_getProcessInfos ::
-            CP.C_IO (CP.OP_List (CP.OP_Tuple2 C_ProcessInfo CP.C_Int))
-external_d_C_getProcessInfos = fromIO $ do
+            ConstStore -> CP.C_IO (CP.OP_List (CP.OP_Tuple2 C_ProcessInfo CP.C_Int))
+external_d_C_getProcessInfos _ = fromIO $ do
   t <- getCPUTime
   return (CP.OP_Cons (CP.OP_Tuple2 C_RunTime (toCurry (t `div` (10^9))))
                      CP.OP_List)
 
-external_d_C_garbageCollectorOff :: CP.C_IO CP.OP_Unit
-external_d_C_garbageCollectorOff = fromHaskellIO0 (return ()) -- not supported
+external_d_C_garbageCollectorOff :: ConstStore -> CP.C_IO CP.OP_Unit
+external_d_C_garbageCollectorOff _ = fromHaskellIO0 (return ()) -- not supported
 
-external_d_C_garbageCollectorOn :: CP.C_IO CP.OP_Unit
-external_d_C_garbageCollectorOn = fromHaskellIO0 (return ()) -- not supported
+external_d_C_garbageCollectorOn :: ConstStore -> CP.C_IO CP.OP_Unit
+external_d_C_garbageCollectorOn _ = fromHaskellIO0 (return ()) -- not supported
 
-external_d_C_garbageCollect :: CP.C_IO CP.OP_Unit
-external_d_C_garbageCollect = fromHaskellIO0 performGC
+external_d_C_garbageCollect :: ConstStore -> CP.C_IO CP.OP_Unit
+external_d_C_garbageCollect _ = fromHaskellIO0 performGC

### 2011-11-18 lib/External_ReadShowTerm.hs

@@ -1,20 +1,20 @@
 import GHC.Exts (Char (C#))
 import qualified Curry_Prelude as CP
 
-external_d_C_prim_showTerm :: Show a => a -> CP.C_String
-external_d_C_prim_showTerm t = toCurry (show t)
+external_d_C_prim_showTerm :: Show a => a -> ConstStore -> CP.C_String
+external_d_C_prim_showTerm t _ = toCurry (show t)
 
-external_d_C_prim_showQTerm :: Show a => a -> CP.C_String
-external_d_C_prim_showQTerm t = toCurry (show t)
+external_d_C_prim_showQTerm :: Show a => a -> ConstStore -> CP.C_String
+external_d_C_prim_showQTerm t _= toCurry (show t)
 
 external_d_C_prim_readsUnqualifiedTerm ::
-  Read a => CP.OP_List CP.C_String -> CP.C_String
-         -> CP.OP_List (CP.OP_Tuple2 a CP.C_String)
+  Read a => CP.OP_List CP.C_String -> CP.C_String ->
+         ConstStore -> CP.OP_List (CP.OP_Tuple2 a CP.C_String)
 external_d_C_prim_readsUnqualifiedTerm _ = external_d_C_prim_readsQTerm
 
-external_d_C_prim_readsQTerm :: Read a => CP.C_String
-                             -> CP.OP_List (CP.OP_Tuple2 a CP.C_String)
-external_d_C_prim_readsQTerm s = toCurryPairs (reads (fromCurry s))
+external_d_C_prim_readsQTerm :: Read a => CP.C_String -> 
+                             ConstStore -> CP.OP_List (CP.OP_Tuple2 a CP.C_String)
+external_d_C_prim_readsQTerm s _ = toCurryPairs (reads (fromCurry s))
   where 
    toCurryPairs [] = CP.OP_List
    toCurryPairs ((v,s):xs) = CP.OP_Cons (CP.OP_Tuple2 v (toCurry s))

### 2011-11-18 lib/External_SearchTree.hs

@@ -14,6 +14,6 @@ instance MonadPlus C_SearchTree where
   mzero = C_Fail
   mplus = C_Or
 
-external_d_C_someSearchTree :: (NormalForm a,NonDet a) => a -> C_SearchTree a 
-external_d_C_someSearchTree = searchMPlus
+external_d_C_someSearchTree :: (NormalForm a,NonDet a) => a -> ConstStore -> C_SearchTree a 
+external_d_C_someSearchTree x _ = searchMPlus x
 

### 2011-11-18 lib/External_Socket.hs

@@ -10,11 +10,11 @@ instance ConvertCurryHaskell CP.C_Int PortID where
   toCurry (PortNumber i) = toCurry (toInteger i)
   fromCurry i = PortNumber (fromInteger (fromCurry i))
 
-external_d_C_prim_listenOn :: CP.C_Int -> CP.C_IO C_Socket
-external_d_C_prim_listenOn = fromHaskellIO1 listenOn
+external_d_C_prim_listenOn :: CP.C_Int -> ConstStore -> CP.C_IO C_Socket
+external_d_C_prim_listenOn i _ = fromHaskellIO1 listenOn i
 
-external_d_C_listenOnFresh :: CP.C_IO (CP.OP_Tuple2 CP.C_Int C_Socket)
-external_d_C_listenOnFresh = fromHaskellIO0 listenOnFreshPort
+external_d_C_listenOnFresh :: ConstStore -> CP.C_IO (CP.OP_Tuple2 CP.C_Int C_Socket)
+external_d_C_listenOnFresh _ = fromHaskellIO0 listenOnFreshPort
  where
    listenOnFreshPort :: IO (PortID,Socket)
    listenOnFreshPort = do
@@ -23,14 +23,14 @@ external_d_C_listenOnFresh = fromHaskellIO0 listenOnFreshPort
      return (p,s)
 
 external_d_C_prim_socketAccept :: C_Socket
-  -> CP.C_IO (CP.OP_Tuple2 CP.C_String Curry_IO.C_Handle)
-external_d_C_prim_socketAccept =
- fromHaskellIO1 (\s -> Network.accept s >>= \ (h,s,_) -> return (s,OneHandle h))
+  -> ConstStore -> CP.C_IO (CP.OP_Tuple2 CP.C_String Curry_IO.C_Handle)
+external_d_C_prim_socketAccept socket _ =
+ fromHaskellIO1 (\s -> Network.accept s >>= \ (h,s,_) -> return (s,OneHandle h)) socket
 
 
 external_d_C_prim_waitForSocketAccept :: C_Socket -> CP.C_Int
- -> CP.C_IO (CP.C_Maybe (CP.OP_Tuple2 (CP.OP_List CP.C_Char) Curry_IO.C_Handle))
-external_d_C_prim_waitForSocketAccept = fromHaskellIO2 wait
+ -> ConstStore -> CP.C_IO (CP.C_Maybe (CP.OP_Tuple2 (CP.OP_List CP.C_Char) Curry_IO.C_Handle))
[…]

### 2011-11-18 lib/External_System.hs

@@ -5,34 +5,34 @@ import System.CPUTime(getCPUTime)
 import Network.BSD(getHostName)
 import qualified Curry_Prelude as CP
 
-external_d_C_getCPUTime :: CP.C_IO CP.C_Int
-external_d_C_getCPUTime =
+external_d_C_getCPUTime :: ConstStore -> CP.C_IO CP.C_Int
+external_d_C_getCPUTime _ =
   fromIO (getCPUTime >>= return . toCurry . (`div` 1000000000))
 
-external_d_C_getElapsedTime :: CP.C_IO CP.C_Int
-external_d_C_getElapsedTime =
+external_d_C_getElapsedTime :: ConstStore -> CP.C_IO CP.C_Int
+external_d_C_getElapsedTime _ =
   fromHaskellIO0 (return (0 :: Int))
 
-external_d_C_getArgs :: CP.C_IO (CP.OP_List CP.C_String)
-external_d_C_getArgs = fromHaskellIO0 getArgs
+external_d_C_getArgs :: ConstStore -> CP.C_IO (CP.OP_List CP.C_String)
+external_d_C_getArgs _ = fromHaskellIO0 getArgs
 
-external_d_C_prim_getEnviron :: CP.C_String -> CP.C_IO CP.C_String
-external_d_C_prim_getEnviron =
-  fromHaskellIO1 (\var -> getEnv var `catch` (\_ -> return ""))
+external_d_C_prim_getEnviron :: CP.C_String -> ConstStore -> CP.C_IO CP.C_String
+external_d_C_prim_getEnviron str _ =
+  fromHaskellIO1 (\var -> getEnv var `catch` (\_ -> return "")) str
 
-external_d_C_getHostname :: CP.C_IO CP.C_String
-external_d_C_getHostname = fromHaskellIO0 getHostName
+external_d_C_getHostname :: ConstStore -> CP.C_IO CP.C_String
+external_d_C_getHostname _ = fromHaskellIO0 getHostName
 
-external_d_C_getPID :: CP.C_IO CP.C_Int
-external_d_C_getPID = fromIO (do pid <- getProcessID
-                                 return (toCurry (fromIntegral pid :: Int))
-                             )
+external_d_C_getPID :: ConstStore -> CP.C_IO CP.C_Int
+external_d_C_getPID _ = fromIO (do pid <- getProcessID
+                                   return (toCurry (fromIntegral pid :: Int))
[…]

### 2011-11-18 lib/External_Time.hs

@@ -27,17 +27,17 @@ instance ConvertCurryHaskell C_CalendarTime T.CalendarTime where
                          (toCurry s)
                          (toCurry tz)
 
-external_d_C_getClockTime :: CP.C_IO C_ClockTime
-external_d_C_getClockTime =
+external_d_C_getClockTime ::  ConstStore -> CP.C_IO C_ClockTime
+external_d_C_getClockTime _ =
   fromIO (T.getClockTime >>= return . toCurry)
 
-external_d_C_prim_toCalendarTime :: C_ClockTime -> CP.C_IO C_CalendarTime
-external_d_C_prim_toCalendarTime ct =
+external_d_C_prim_toCalendarTime :: C_ClockTime -> ConstStore -> CP.C_IO C_CalendarTime
+external_d_C_prim_toCalendarTime ct _ =
   fromIO (T.toCalendarTime (fromCurry ct) >>= return . toCurry)
 
-external_d_C_prim_toUTCTime :: C_ClockTime -> C_CalendarTime
-external_d_C_prim_toUTCTime ct = toCurry (T.toUTCTime (fromCurry ct))
+external_d_C_prim_toUTCTime :: C_ClockTime ->  ConstStore -> C_CalendarTime
+external_d_C_prim_toUTCTime ct _ = toCurry (T.toUTCTime (fromCurry ct))
 
-external_d_C_prim_toClockTime :: C_CalendarTime -> C_ClockTime
-external_d_C_prim_toClockTime ct =
+external_d_C_prim_toClockTime :: C_CalendarTime -> ConstStore -> C_ClockTime
+external_d_C_prim_toClockTime ct _ =
   toCurry (T.toClockTime (fromCurry ct))

### 2011-11-18 lib/External_Unsafe.hs

@@ -1,7 +1,7 @@
 import System.IO.Unsafe(unsafePerformIO)
 import qualified Curry_Prelude as CP
 
-external_d_C_unsafePerformIO :: CP.C_IO a -> a
-external_d_C_unsafePerformIO = unsafePerformIO . toIO
+external_d_C_unsafePerformIO :: CP.C_IO a -> ConstStore -> a
+external_d_C_unsafePerformIO io cs = unsafePerformIO  (toIO io cs)
 
 -----------------------------------------------------------------------

### 2011-11-18 runtime/Basics.hs

@@ -20,24 +20,24 @@ nonAsciiChr :: Int# -> Char#
 nonAsciiChr i = chr# i
 
 -- Apply a function to the head normal form
-d_dollar_bang :: (NonDet a, NonDet b) => (a -> b) -> a -> b
-d_dollar_bang f x = match hnfChoice hnfNarrowed hnfFree failCons hnfGuard f x
+d_dollar_bang :: (NonDet a, NonDet b) => (a -> ConstStore -> b) -> a -> ConstStore -> b
+d_dollar_bang f x cs = match hnfChoice hnfNarrowed hnfFree failCons hnfGuard (flip f cs) x
   where
-   hnfChoice i a b  = choiceCons i (d_dollar_bang f a) (d_dollar_bang f b)
-   hnfNarrowed i xs = choicesCons i (map (d_dollar_bang f) xs)
-   hnfFree i xs     = f (choicesCons i xs)
-   hnfGuard c e     = guardCons c (d_dollar_bang f e)
+   hnfChoice i a b  = choiceCons i (d_dollar_bang f a cs) (d_dollar_bang f b cs)
+   hnfNarrowed i xs = choicesCons i (map (\y -> d_dollar_bang f y cs) xs)
+   hnfFree i xs     = f (choicesCons i xs) cs
+   hnfGuard c e     = guardCons c (d_dollar_bang f e $! combConstr c cs)
 
 -- Apply a non-deterministic function to the head normal form
-nd_dollar_bang :: (NonDet a, NonDet b) => (Func a b) -> a -> IDSupply -> b
-nd_dollar_bang f x s = match hnfChoice hnfNarrowed hnfFree failCons hnfGuard hnfVal x
+nd_dollar_bang :: (NonDet a, NonDet b) => (Func a b) -> a -> IDSupply -> ConstStore -> b
+nd_dollar_bang f x s cs = match hnfChoice hnfNarrowed hnfFree failCons hnfGuard hnfVal x
   where
-   hnfVal v         = nd_apply f v s
+   hnfVal v         = nd_apply f v s cs
    -- TODO Do we better use leftSupply and rightSupply?
-   hnfChoice i a b  = choiceCons i (nd_dollar_bang f a s) (nd_dollar_bang f b s)
-   hnfNarrowed i xs = choicesCons i (map (\x -> nd_dollar_bang f x s) xs)
-   hnfFree i xs     = nd_apply f (choicesCons i xs) s
-   hnfGuard c e     = guardCons c (nd_dollar_bang f e s)
[…]

### 2011-11-18 runtime/ID.hs

@@ -1,9 +1,11 @@
+{-# LANGUAGE ExistentialQuantification #-}
+
 -- ---------------------------------------------------------------------------
 -- ID module
 -- ---------------------------------------------------------------------------
 module ID
   ( -- * Constraints
-    Constraint (..), Constraints
+    Constraint (..), Constraints(..), getConstrList
     -- * Choices
   , Choice (..), defaultChoice, isDefaultChoice
     -- * IDs
@@ -29,12 +31,27 @@ data Constraint
   -- |Unsolvable constraint
   | Unsolvable
   -- |Non-deterministic choice between two lists of constraints
-  | ConstraintChoice ID Constraints Constraints
+  | ConstraintChoice ID [Constraint] [Constraint]
   -- |Non-deterministic choice between a list of lists of constraints
-  | ConstraintChoices ID [Constraints]
-    deriving (Eq, Show)
+  | ConstraintChoices ID [[Constraint]]
+ deriving (Show,Eq)
 
-type Constraints = [Constraint]
+-- A Value Constraint is used to bind a Value to an id it also contains the
+-- structural constraint information that describes the choice to be taken
+-- for a given id, a Struct Constraint has only the structural information
+data Constraints = forall a . ValConstr ID a [Constraint] | StructConstr [Constraint]
+
+-- a selector to get the strucural constraint information from a constraint
+getConstrList :: Constraints -> [Constraint]
+getConstrList (ValConstr _ _ c) = c
+getConstrList (StructConstr c) = c
+
+instance Show Constraints where
+  showsPrec _ (ValConstr _ _ c) = ("(ValC " ++) .  shows c . (')':)
+  showsPrec _ (StructConstr c) = ("(StructC " ++) . shows c . (')':)
+
+instance Eq Constraints where 
+ c1 == c2 = getConstrList c1 == getConstrList c2
 
 -- ---------------------------------------------------------------------------
 -- Choice
@@ -60,7 +77,7 @@ data Choice
   | BoundTo ID Int
     -- |A free variable is lazily bound to an expression by a function
    --   pattern
-  | LazyBind Constraints
+  | LazyBind [Constraint]
     deriving Show
 
 

### 2011-11-18 runtime/PrimTypes.hs

@@ -15,7 +15,7 @@ data BinInt
      | Choice_BinInt ID BinInt BinInt
      | Choices_BinInt ID ([BinInt])
      | Fail_BinInt
-     | Guard_BinInt ([Constraint]) BinInt
+     | Guard_BinInt (Constraints) BinInt
 
 instance Show BinInt where
   showsPrec d (Choice_BinInt i x y) = showsChoice d i x y
@@ -55,20 +55,20 @@ instance Generable BinInt where
 
 
 instance NormalForm BinInt where
-  ($!!) cont (Neg x1) = (\y1 -> cont (Neg y1)) $!! x1
-  ($!!) cont Zero = cont Zero
-  ($!!) cont (Pos x1) = (\y1 -> cont (Pos y1)) $!! x1
-  ($!!) cont (Choice_BinInt i x y) = nfChoice cont i x y
-  ($!!) cont (Choices_BinInt i xs) = nfChoices cont i xs
-  ($!!) cont (Guard_BinInt c x) = guardCons c (cont $!! x)
-  ($!!) _ Fail_BinInt = failCons
-  ($##) cont (Neg x1) = (\y1 -> cont (Neg y1)) $## x1
-  ($##) cont Zero = cont Zero
-  ($##) cont (Pos x1) = (\y1 -> cont (Pos y1)) $## x1
-  ($##) cont (Choice_BinInt i x y) = gnfChoice cont i x y
-  ($##) cont (Choices_BinInt i xs) = gnfChoices cont i xs
-  ($##) cont (Guard_BinInt c x) = guardCons c (cont $## x)
-  ($##) _ Fail_BinInt = failCons
+  ($!!) cont (Neg x1) cs = ((\y1 cs -> cont (Neg y1) cs) $!! x1) cs
+  ($!!) cont Zero cs = cont Zero cs
+  ($!!) cont (Pos x1) cs = ((\y1 cs -> cont (Pos y1) cs) $!! x1) cs
+  ($!!) cont (Choice_BinInt i x y) cs = nfChoice cont i x y cs
+  ($!!) cont (Choices_BinInt i xs) cs = nfChoices cont i xs cs
+  ($!!) cont (Guard_BinInt c x) cs = guardCons c ((cont $!! x) $! combConstr c cs)
+  ($!!) _ Fail_BinInt _ = failCons
+  ($##) cont (Neg x1) cs = ((\y1 cs -> cont (Neg y1) cs) $## x1) cs
[…]

### 2011-11-18 runtime/Search.hs

@@ -4,7 +4,7 @@ import Control.Monad
 import Control.Monad.State.Strict
 import Control.Parallel.TreeSearch
 import qualified Data.Map
-
+import System.IO.Unsafe (unsafePerformIO)
 import Debug
 import ID
 import MonadList
@@ -12,27 +12,47 @@ import PrimTypes
 import Solver (SolutionTree (..), solves)
 import Types
 
--- TODO what to do whith choices and failures
-toIO :: C_IO a -> IO a
-toIO (C_IO io) = io
-toIO (Choice_C_IO _ _ _) = error "toIO: Non-determinism in IO occured"
-toIO (Guard_C_IO constraints e) = do
-  st <- solves constraints
+toIO :: C_IO a -> ConstStore -> IO a
+toIO (C_IO io) _ = io
+toIO (Choice_C_IO _ _ _) _ = error "toIO: Non-determinism in IO occured"
+toIO (Guard_C_IO constraints e) cs = do
+  st <- solves (getConstrList constraints)
   case st of
-    SuccessST _ -> toIO e
+    SuccessST _ -> do
+      -- adds the constraint to the global constraint store
+      -- to make it available to subsequent operations
+      -- this is safe because no backtracking is done in IO
+      addToGlobalCs constraints
+      toIO e $! (combConstr constraints cs)
     FailST           -> error "toIO (Guard): failed"
     ChoiceST  _ _ _  -> error "toIO (Guard): Non-determinism in IO occured"
     ChoicesST _ _    -> error "toIO (Guard): Non-determinism in IO occured"
-toIO Fail_C_IO = error "toIO: failed"
-toIO (Choices_C_IO i choices) = do
+toIO Fail_C_IO _ = error "toIO: failed"
+-- TODO: lookup value in constraint map and global map?
+toIO (Choices_C_IO i@(NarrowedID _ _) choices) cs = followToIO i choices cs
+
+toIO (Choices_C_IO i@(FreeID _ _) choices) cs = do
+  -- bindings of free variables are looked up first in the local
+  -- constraint store and then in the global constraint store
+  lookupCs cs i (flip toIO cs) tryGlobal
+ where
+  tryGlobal = do
+    csg <- lookupGlobalCs
+    lookupCs csg i (flip toIO cs) (followToIO i choices cs)
+
+followToIO :: ID -> [C_IO a] -> ConstStore -> IO a
+followToIO i choices cs =  do
   c <- lookupChoice i
   case c of
[…]

### 2011-11-18 runtime/Types.hs

@@ -1,8 +1,59 @@
-{-# LANGUAGE MultiParamTypeClasses, Rank2Types #-}
+{-# LANGUAGE MultiParamTypeClasses, Rank2Types, ExistentialQuantification #-}
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 module Types where
 
 import ID
+import qualified Data.Map as Map
+import Data.Map (Map)
+import Data.IORef
+import Unsafe.Coerce(unsafeCoerce)
+import System.IO.Unsafe (unsafePerformIO)
+
+-- ---------------------------------------------------------------------------
+-- Constraint Store
+-- ---------------------------------------------------------------------------
+
+type ConstStore = Map Integer Value
+
+data Value = forall a . V a
+
+unV :: Value -> a
+unV (V x) = unsafeCoerce x
+
+emptyCs :: ConstStore
+emptyCs = Map.empty
+
+-- combines a constraint and a constraint sote
+combConstr :: Constraints -> ConstStore -> ConstStore
+combConstr (StructConstr _)  cs = cs
+combConstr (ValConstr i v _) cs = Map.insert (mkInt i) (V v) cs
+
+-- combines two constraint stores
+combConstStores :: ConstStore -> ConstStore -> ConstStore
+combConstStores = Map.union
+
+-- lookupCs looks for a velue bound in a constraint store
+-- if the value is found the given function is applied,
+-- if not, the default value is returned
+lookupCs :: ConstStore -> ID -> (a -> b) -> b -> b
+lookupCs cs i f def = maybe def (f . unV) (Map.lookup (mkInt i) cs)
+
+-- ---------------------------------------------------------------------------
+-- Global Constraint Store
+------------------------------------------------------------------------------
+
+globalCs = unsafePerformIO $ newIORef emptyCs
+
+-- adds a Constraint to the global constraint store
+addToGlobalCs :: Constraints -> IO ()
+addToGlobalCs (StructConstr _) = return ()
+addToGlobalCs cs@(ValConstr i x _ ) = do
+  gcs <- readIORef globalCs
+  let gcs' = combConstr cs gcs
+  writeIORef globalCs $! gcs'
+
+lookupGlobalCs :: IO ConstStore
[…]

### 2011-11-18 docs/src/manual.tex
Curry tests improved

@@
 For this purpose, there is a system library \code{Assertion}
-(Section~ref{Library:Assertion}) which
+(Section~\ref{Library:Assertion}) which
 contains the necessary definitions for writing tests.

### 2011-11-18 runtime/Types.hs
Deactivated global constraint store

@@ -47,6 +47,7 @@ globalCs = unsafePerformIO $ newIORef emptyCs
 -- adds a Constraint to the global constraint store
 addToGlobalCs :: Constraints -> IO ()
 addToGlobalCs (StructConstr _) = return ()
+addToGlobalCs _ = return ()
 addToGlobalCs cs@(ValConstr i x _ ) = do
   gcs <- readIORef globalCs
   let gcs' = combConstr cs gcs

### 2011-11-29 runtime/Types.hs
added conditional compilation to disable the constraint store and the use of normal forms for value constraints
the c preprocessor is now activated, non-normalform value constraints are the standard

@@ -25,37 +25,56 @@ emptyCs = Map.empty
 
 -- combines a constraint and a constraint sote
 combConstr :: Constraints -> ConstStore -> ConstStore
+#ifdef DISABLE_CS
+combConstr = const
+#else
 combConstr (StructConstr _)  cs = cs
 combConstr (ValConstr i v _) cs = Map.insert (getKey i) (V v) cs
-
+#endif
 -- combines two constraint stores
 combConstStores :: ConstStore -> ConstStore -> ConstStore
+#ifdef DISABLE_CS
+combConstStores = const
+#else
 combConstStores = Map.union
+#endif
 
 -- lookupCs looks for a velue bound in a constraint store
 -- if the value is found the given function is applied,
 -- if not, the default value is returned
 lookupCs :: ConstStore -> ID -> (a -> b) -> b -> b
+#ifdef DISABLE_CS
+lookupCs _ _ _ def = def
+#else
 lookupCs cs i f def = maybe def (f . unV) (Map.lookup (getKey i) cs)
+#endif
 
 -- ---------------------------------------------------------------------------
 -- Global Constraint Store
 ------------------------------------------------------------------------------
-
+#ifndef DISABLE_CS
 globalCs :: IORef ConstStore
 globalCs = unsafePerformIO $ newIORef emptyCs
+#endif
 
 -- adds a Constraint to the global constraint store
 addToGlobalCs :: Constraints -> IO ()
+#ifdef DISABLE_CS
+addToGlobalCs = return ()
+#else
 addToGlobalCs (StructConstr _) = return ()
 addToGlobalCs cs@(ValConstr i x _ ) = do
   gcs <- readIORef globalCs
   let gcs' = combConstr cs gcs
   writeIORef globalCs $! gcs'
+#endif
 
 lookupGlobalCs :: IO ConstStore
+#ifdef DISABLE_CS
+lookupGlobalCs = return emptyCs
+#else
 lookupGlobalCs = readIORef globalCs
-
+#endif
 -- ---------------------------------------------------------------------------
 -- Try structure
 -- ---------------------------------------------------------------------------
@@ -238,7 +257,7 @@ class NormalForm a => Unifiable a where
   where
   uniChoice i x1 x2 = checkFail (choiceCons i ((x1 =:= y) cs) ((x2 =:= y) cs)) y
[…]

### 2011-11-29 docs/src/currydoc.tex

@@
 meaning, one can escape them with a backslash:
-\begin{verbatim}
+\begin{curry}
 --- This is a comment without a \'reference\'.
-\end{verbatim}
+\end{curry}
 To simplify the writing of documentation comments,
@@
 an identifier, the escaping can be omitted, as in
-\begin{verbatim}
+\begin{curry}
 --- This isn't a reference.
-\end{verbatim}
+\end{curry}
 %
@@
 documentation comments:
-{\renewcommand{\baselinestretch}{1.0}
-\begin{verbatim}
+\begin{curry}
 --- This is an
@@
 data Tree a = Leaf a | Node [Tree a]
-\end{verbatim}
-}
+\end{curry}
 \noindent
 To generate the documentation, execute the command\pindex{currydoc}
-\startprog
+\begin{curry}
 currydoc Example
-\stopprog
+\end{curry}
 (\code{currydoc} is a command usually stored in \code{$kics2home$/bin}
@@
 one can also execute the command
-\startprog
+\begin{curry}
 currydoc docdir Example
-\stopprog
+\end{curry}
 where \code{docdir} is the directory for the documentation files.

### 2011-11-29 docs/src/manual.tex

@@
 \renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\tt}
-
-% environment for typing program texts:
-\makeatletter
-\newenvironment{prog}{\par\vspace{0.7ex}
-\setlength{\parindent}{1.0cm}
-\setlength{\parskip}{-0.1ex}
-\obeylines\@vobeyspaces\tt}{\vspace{0.7ex}\noindent
+\newcommand{\codefont}{\small\tt}
+
+\usepackage{listings}
+\lstset{aboveskip=1.5ex,
+        belowskip=1.5ex,
+        showstringspaces=false, % no special string space
+        mathescape=true,
+        flexiblecolumns=false,
+        xleftmargin=2ex,
+        basewidth=0.52em,
+        basicstyle=\small\ttfamily}
+
+\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
        }
-\makeatother
-\newcommand{\startprog}{\begin{prog}}
-\newcommand{\stopprog}{\end{prog}\noindent}
+
+\lstnewenvironment{curry}{}{}
+\newcommand{\listline}{\vrule width0pt depth1.75ex}
 
@@
 % program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\mbox{\codefont #1}''}
+\newcommand{\ccode}[1]{``\code{#1}''}
 
@@
 {\huge User Manual}\\[4ex]
-Version of November 9, 2011\\[6ex]
+Version of November 29, 2011\\[6ex]
 \Large
@@
 directory can be deleted by the command\pindex{cleancurry}
-\startprog
+\begin{curry}
 cleancurry
-\stopprog
+\end{curry}
 (this is a shell script stored in the \code{bin} directory of the
@@
 The command
-\startprog
+\begin{curry}
 cleancurry -r
-\stopprog
+\end{curry}
 also deletes the auxiliary files in all subdirectories.
 
-\subsection{Restrictions on Curry Programs}
+\subsection{Restrictions}
 \label{sec-restrictions}
@@
 interpreted as stops at columns 9, 17, 25, 33, and so on.
-%\item Threads created by a concurrent conjunction are not executed
-%in a fair manner (usually, threads corresponding to leftmost constraints
-%are executed with higher priority).
+In general, tabulator stops should be avoided in source programs.
 \item
@@
 \cite{HanusSteiner98PLILP} is not supported.
+\item Concurrent computations\index{concurrency}
+based on the suspension of expressions
+containing free variables are not yet supported.
+KiCS2 supports \emph{value generators} for free variables
+so that a free variable is instantiated when its value is demanded.
+For instance, the initial expression
+\begin{curry}
+x == True where x free
+\end{curry}
+is non-deterministically evaluated to \code{False} and \code{True}
+by instantiating \code{x} to \code{False} and \code{True}, respectively.
+Thus, a computation is never suspended due to free variables.
+This behavior also applies to free variables of primitive types
+like integers. For instance, the initial expression
+\begin{curry}
+x*y=:=1 where x,y free
+\end{curry}
+is non-deterministically evaluated to the two solutions
+\begin{curry}
+{x = -1, y = -1} Success
+{x = 1, y = 1} Success
+\end{curry}
 \item
@@
 starts. For instance, the invocation
-\startprog
+\begin{curry}
 kics2 :load Mod :add List
-\stopprog
+\end{curry}
 starts KiCS2, loads the main module \code{Mod}, and adds the additional
@@
 The invocation
-\startprog
+\begin{curry}
 kics2 :load Mod :eval config
-\stopprog
+\end{curry}
 starts KiCS2, loads the main module \code{Mod}, and evaluates
@@
 As a final example, the invocation
-\startprog
+\begin{curry}
 kics2 :load Mod :save :quit
-\stopprog
+\end{curry}
 starts KiCS2, loads the main module \code{Mod}, creates an executable,
@@
 \textbf{Free variables in initial expressions} must be declared as
-in Curry programs, i.e., in order to see the results of their bindings,
+in Curry programs.
+In order to see the results of their bindings,\footnote{Currently,
+bindings are only printed if the initial expression is not an I/O action
+(i.e., not of type \ccode{IO\ldots})
+and there are not more than ten free variables
+in the initial expression.}
 they must be introduced by a \ccode{where\ldots{}free} declaration.
 For instance, one can write
-\startprog
-xs++ys\,=:=\,[1,2]  where xs,ys free
-\stopprog
+\begin{curry}
+xs++ys =:= [1,2]  where xs,ys free
+\end{curry}
 in order to obtain the following three possible bindings:
-\startprog
-\{xs = [], ys = [1,2]\} Success
-\{xs = [1], ys = [2]\} Success
-\{xs = [1,2], ys = []\} Success
-\stopprog
+\begin{curry}
+{xs = [], ys = [1,2]} Success
+{xs = [1], ys = [2]} Success
+{xs = [1,2], ys = []} Success
+\end{curry}
 Without these declarations, an error is reported in order to
@@
 as in the expression
-\startprog
+\begin{curry}
 xs++ys=:=[z] where xs,ys,z free
-\stopprog
+\end{curry}
 they are specialized to the type \ccode{()}
-(since the current implementation of KiCS2 cannot compute
-with polymorphic logic variables).
+(since the current implementation of KiCS2 does not support
+computations with polymorphic logic variables).
 
@@
 For instance, setting the option
-\startprog
+\begin{curry}
 :set copts -O 0
-\stopprog
+\end{curry}
 has the effect that all optimizations performed by the KiCS2 compiler
@@
 For instance, setting the option
-\startprog
+\begin{curry}
 :set rts -H512m
-\stopprog
+\end{curry}
 has the effect that the minimum heap size is set to 512 megabytes.
@@
 Each source file option must be in a line of the form
-\startprog
-\{-\# KiCS2_OPTION $opt$ \#-\}
-\stopprog
+\begin{curry}
+{-# KiCS2_OPTION $opt$ #-}
+\end{curry}
 where $opt$ is an option that can occur in a \ccode{:set} command
@@
 lines
-\startprog
-\{-\# KiCS2_OPTION ids \#-\}
-\{-\# KiCS2_OPTION +ghci \#-\}
-\{-\# KiCS2_OPTION v2 \#-\}
+\begin{curry}
+{-# KiCS2_OPTION ids #-}
+{-# KiCS2_OPTION +ghci #-}
+{-# KiCS2_OPTION v2 #-}
 module M where
 $\ldots$
-\stopprog
+\end{curry}
 then the load command \ccode{:load M} will also
@@
 For instance, the declaration
-\startprog
+\begin{curry}
 ones5 = let ones = 1 : ones
          in take 5 ones
-\stopprog
+\end{curry}
 introduces the local variable \code{ones} which is bound
@@
 Similarly, the definition
-\startprog
+\begin{curry}
 onetwo n = take n one2
@@
    two1 = 2 : one2
-\stopprog
+\end{curry}
 introduces a local variables \code{one2} that represents
@@
 for list concatenation:
-\startprog
+\begin{curry}
 last xs | ys++[y] =:= xs  = y   where y,ys free
-\stopprog
+\end{curry}
 Since the equality constraint \ccode{=:=} evaluates both sides
@@
 we can define the operation \code{last} as follows:
-\startprog
+\begin{curry}
 last (_++[y]) = y
-\stopprog
+\end{curry}
 This definition is not only more compact but also avoids the complete
@@
 definition is conceptually equivalent to the set of rules
-\startprog
+\begin{curry}
 last [y] = y
@@
 last [_,_,y] = y
-\ldots
-\stopprog
+$\ldots$
+\end{curry}
 which shows that the evaluation of the list elements is not demanded
@@
 can be constructed or used. The declaration has the following form:
-\startprog
-type $R$ $\alpha_1$ \ldots $\alpha_n$ = \{ $l_1$ :: $\tau_1$, \ldots, $l_m$ :: $\tau_m$ \}
-\stopprog
+\begin{curry}
+type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
+\end{curry}
 It introduces a new $n$-ary record type $R$ which represents a
@@
 fields. The following examples define some record types:
-\startprog
-type Person = \{name :: String, age :: Int\}
-type Address = \{person :: Person, street :: String, city :: String\}
-type Branch a b = \{left :: a, right :: b\}
-\stopprog
+\begin{curry}
+type Person = {name :: String, age :: Int}
+type Address = {person :: Person, street :: String, city :: String}
+type Branch a b = {left :: a, right :: b}
+\end{curry}
 It is possible to summarize different labels which have the same
 type. For instance, the record \code{Address} can also be declared as follows:
-\startprog
-type Address = \{person :: Person, street,city :: String\}
-\stopprog
+\begin{curry}
+type Address = {person :: Person, street,city :: String}
+\end{curry}
 The fields can occur in an arbitrary order. The example above
 can also be written as
-\startprog
-type Address = \{street,city :: String, person :: Person\}
-\stopprog
+\begin{curry}
+type Address = {street,city :: String, person :: Person}
+\end{curry}
 The record type can be used in every type expression to represent
 the corresponding record, e.g.
-\startprog
+\begin{curry}
 data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\stopprog
-\startprog
+\end{curry}
+\begin{curry}
 getName :: Person -> String
-getName \ldots
-\stopprog
+getName $\ldots$
+\end{curry}
 
@@
 each data field. It has the following form:
-\startprog
-\{ $l_1$ = $v_1$, \ldots, $l_m$ = $v_m$ \}
-\stopprog
+\begin{curry}
+{ $l_1$ = $v_1$,$\ldots$, $l_m$ = $v_m$ }
+\end{curry}
 It generates a record where each label $l_i$ refers to the
@@
 exactly one assignment. Examples for record constructions are
-\startprog
-\{name = "Johnson", age = 30\}     -- generates a record of type 'Person'
-\{left = True, right = 20\}        -- generates a record of type 'Branch'
-\stopprog
+\begin{curry}
+{name = "Johnson", age = 30}     -- generates a record of type 'Person'
+{left = True, right = 20}        -- generates a record of type 'Branch'
+\end{curry}
 Assignments to labels can occur in an arbitrary order. For instance a
 record of type \code{Person} can also be generated as follows:
-\startprog
-\{age = 30, name = "Johnson"\}     -- generates a record of type 'Person'
-\stopprog
+\begin{curry}
+{age = 30, name = "Johnson"}     -- generates a record of type 'Person'
+\end{curry}
 Unlike labeled fields in record type declarations, 
@@
 expression is valid:
-\startprog
-\{person = \{name = "Smith", age = 20\},   -- generates a record of
+\begin{curry}
+{person = {name = "Smith", age = 20},   -- generates a record of
  street = "Main Street",                -- type 'Address'
- city   = "Springfield"\}
-\stopprog
+ city   = "Springfield"}
+\end{curry}
 
@@
 It has the following form:
-\startprog
+\begin{curry}
 $r$ -> $l$
-\stopprog
+\end{curry}
 It returns the value to which the label $l$ refers to from the
@@
 An example for a field selection is:
-\startprog
+\begin{curry}
 pers -> name
-\stopprog
+\end{curry}
 This returns the value of the label \code{name} from the record \code{pers}
@@
 Sequential application of field selections are also possible:
-\startprog
+\begin{curry}
 (addr -> person) -> age
-\stopprog
+\end{curry}
 The value of the label \code{age} is extracted from a record which itself
@@
 Records can be updated by reassigning a new value to a label:
-\startprog
-\{$l_1$ := $v_1$, \ldots, $l_k$ := $v_k$ | $r$\}
-\stopprog
+\begin{curry}
+{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
+\end{curry}
 The label $l_i$ is associated with the new value $v_i$ which
@@
 Examples for record updates are:
-\startprog
-\{name := "Scott", age := 25 | pers\}
-\{person := \{name := "Scott", age := 25 | pers\} | addr\}
-\stopprog
+\begin{curry}
+{name := "Scott", age := 25 | pers}
+{person := {name := "Scott", age := 25 | pers} | addr}
+\end{curry}
 In these examples \code{pers} is a record of type \code{Person} and \code{addr}
@@
 are available:
-\startprog
-\{$l_1$ = $p_1$, \ldots, $l_n$ = $p_n$\}
-\{$l_1$ = $p_1$, \ldots, $l_k$ = $p_k$ | _\}
-\stopprog
+\begin{curry}
+{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
+{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
+\end{curry}
 In both cases each label $l_i$ is specified with a pattern $p_i$. 
@@
 Here are some examples of pattern matching with records:
-\startprog
+\begin{curry}
 isSmith30 :: Person -> Bool
-isSmith30 \{name = "Smith", age = 30\} = True
-\stopprog
-\startprog
+isSmith30 {name = "Smith", age = 30} = True
+\end{curry}
+\begin{curry}
 startsWith :: Char -> Person -> Bool
-startsWith c \{name = (d:_) | _\} = c == d
-\stopprog
-\startprog
+startsWith c {name = (d:_) | _} = c == d
+\end{curry}
+\begin{curry}
 getPerson :: Address -> Person
-getPerson \{person = p | _\} = p
-\stopprog
+getPerson {person = p | _} = p
+\end{curry}
 As shown in the last example, a field selection can also be obtained
@@
 together with the following operations:
-\startprog
+\begin{curry}
 assertTrue      :: String -> Bool -> Assertion ()
@@
 assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\stopprog
+\end{curry}
 The expression \ccode{assertTrue $s$ $b$}
@@
 that can be used to test some list processing functions:
-\startprog
-\medskip
+\begin{curry}
 import List
 import Assertion
-\medskip
+
 test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-\medskip
+
 test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-\medskip
-test3 = assertSolutions "prefix" (\labs{}x -> let y free in  x\,++\,y =:= [1,2])
+
+test3 = assertSolutions "prefix" (\x -> let y free in  x++y =:= [1,2])
                                  [[],[1],[1,2]]
-\medskip
-\stopprog
+
+\end{curry}
 For instance, \code{test1} asserts that the result of evaluating the
@@
 We can execute a test suite by the command\pindex{currytest}
-\startprog
+\begin{curry}
 currytest testList
-\stopprog
+\end{curry}
 (\code{currytest} is a program stored in \code{$kics2home$/bin}
@@
 For our example above, we obtain the following successful protocol:
-\startprog
+\begin{curry}
 ============================================================
@@
 ============================================================
-\stopprog
+\end{curry}
 There is also a graphical interface that summarizes the results
@@
 \ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\startprog
+\begin{curry}
 currytest --window testList
-\stopprog
+\end{curry}
 or
-\startprog
+\begin{curry}
 currytest -w testList
-\stopprog
+\end{curry}
 A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
@@
 command\pindex{erd2curry}
-\startprog
+\begin{curry}
 erd2curry myerd.xmi
-\stopprog
+\end{curry}
 (\code{erd2curry} is a program stored in \code{$kics2home$/bin}
@@
 command\pindex{erd2curry}
-\startprog
+\begin{curry}
 erd2curry -t myerd.term
-\stopprog
+\end{curry}
 %
@@
 This can be done by the command
-\startprog
+\begin{curry}
 erd2curry -v myerd.term
-\stopprog
+\end{curry}
 
@@
 containing the following two lines:
-\startprog
+\begin{curry}
 rev []     = []
 rev (x:xs) = (rev xs) ++ [x]
-\stopprog
+\end{curry}
 Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
 following term:
-\startprog
+\begin{curry}
  (Prog "test"
@@
  )
-\stopprog
+\end{curry}
 
@@
 Instead of the defining rules, you have to write
-\startprog
+\begin{curry}
 f external
-\stopprog
+\end{curry}
 somewhere in the file containing the type declaration for 
@@
 The name and type of this string concatenation should be
-\startprog
+\begin{curry}
 sconc :: String -> String -> String
-\stopprog
+\end{curry}
 Since the primitive Haskell implementation of this operation
@@
 Thus, we define \code{sconc} by
-\startprog
+\begin{curry}
 sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc \$\#\# s1) \$\#\# s2
-~
+sconc s1 s2 = (prim_sconc $\$$## s1) $\$$## s2
+
 prim_sconc :: String -> String -> String
 prim_sconc external
-\stopprog
+\end{curry}
 so that it is ensured that the external operation \code{prim_sconc}
@@
 \code{.curry/kics2/Curry_String.hs}):
-\startprog
+\begin{curry}
 d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
@@
 d_C_prim_sconc x1 x2 = external_d_C_prim_sconc x1 x2
-\stopprog
+\end{curry}
 The type constructors \code{OP_List} and \code{C_Char}
@@
 Thus, we start with the following definitions:
-\startprog
+\begin{curry}
 import qualified Curry_Prelude as CP
-~
+
 external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
                         -> CP.OP_List CP.C_Char
-\stopprog
+\end{curry}
 First, we import the standard prelude with the name \code{CP}
@@
 Thus, we complete our example with the definition
-\startprog
+\begin{curry}
 external_d_C_prim_sconc s1 s2 =
   toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\stopprog
+\end{curry}
 Here, we use Haskell's concatenation operation \ccode{++}
@@
 we start by declaring the external operations in the Curry module \code{String}:
-\startprog
+\begin{curry}
 printString :: String -> IO ()
-printString s = prim_printString \$\#\# s
-~
+printString s = prim_printString $\$$## s
+
 prim_printString :: String -> IO ()
 prim_printString external
-\stopprog
+\end{curry}
 Next we add the corresponding implementation in the file
@@
 of the Haskell representation of the Curry datatype \ccode{()}):
-\startprog
+\begin{curry}
 external_d_C_prim_printString :: CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
 external_d_C_prim_printString = fromHaskellIO1 putStrLn
-\stopprog
+\end{curry}
 When we compile the Curry module \code{String},
@@
 if we declare the external operation
-\startprog
-(<\&>) :: Int -> Int -> Int
-(<\&>) external
-\stopprog
+\begin{curry}
+(<&>) :: Int -> Int -> Int
+(<&>) external
+\end{curry}
 the generated Haskell module contains the code
-\startprog
+\begin{curry}
 d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
@@
 d_OP_lt_ampersand_gt x1 x2 = external_d_OP_lt_ampersand_gt x1 x2
-\stopprog
+\end{curry}
 so that one has to implement the operation

### 2011-12-02 docs/src/manual.tex

@@
 
-\item[\fbox{\code{+/-ghci}}]
+\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
 Turn on/off the ghci mode.
@@
 
-\item[\fbox{\code{copts $o$}}]\pindex{copts}
+\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
 Define additional options passed to the KiCS2 compiler.
@@
 \begin{curry}
-:set copts -O 0
+:set cmp -O 0
 \end{curry}
@@
 
-\item[\fbox{\code{rts $o$}}]\pindex{rts}
+\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
+Define additional options passed to the Glasgow Haskell Compiler
+when the generated Haskell programs are compiled.
+Usually, such options are useful only for experimental purposes.
+For instance, setting the option
+\begin{curry}
+:set ghc -DDISABLE_CS
+\end{curry}
+has the effect that the constraint store used to enable
+an efficient access to complex bindings is disabled.
+Similarly,
+\begin{curry}
+:set ghc -DSTRICT_VAL_BIND
+\end{curry}
+has the effect that expressions in a unification constraint
+(\code{=:=}) are always fully evaluated
+(instead of the evaluation to a head normal form only)
+before unifying both sides.
+Since these options influence the compilation of the run-time system,
+one should also enforce the recompilation of Haskell programs
+by the GHC option \ccode{-fforce-recomp}, e.g., one should set
+\begin{curry}
+:set ghc -DDISABLE_CS -fforce-recomp
+\end{curry}
+
+\item[\fbox{\code{rts $opts$}}]\pindex{rts}
 Define additional run-time options passed to the executable
@@
 has the effect that the minimum heap size is set to 512 megabytes.
+
+\item[\fbox{\code{args $arguments$}}]\pindex{args}
+Define run-time arguments passed to the executable
+generated by the Glasgow Haskell Compiler.
+For instance, setting the option
+\begin{curry}
+:set args first second
+\end{curry}
+has the effect that the I/O operation \code{getArgs}
+(see library \code{System} (Section~\ref{Library:System})
+returns the value \code{["first","second"]}.
 

### 2011-12-13 docs/src/manual.tex
Makefile improved for git install, descrs of ext. ops. up-to-date

@@
 {\huge User Manual}\\[4ex]
-Version of November 29, 2011\\[6ex]
+Version of December 13, 2011\\[6ex]
 \Large
@@
 language.
-Therefore, if a new external operation should be added
+Thus, if an external operation should be added
 to the system, this operation must be declared as \code{external}
@@
 \end{curry}
-somewhere in the file containing the type declaration for 
-the external operation \code{f}.
+below the type declaration for the external operation \code{f}.
 \end{enumerate}
@@
 There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal
-which is a bit more efficient and can be used for unstructured
+which evaluates the argument only to head-normal form.
+This is a bit more efficient and can be used for unstructured
 types like \code{Bool}.}
@@
                -> Curry_Prelude.OP_List Curry_Prelude.C_Char
+               -> ConstStore
                -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 = external_d_C_prim_sconc x1 x2
+d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
 \end{curry}
@@
 The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the
-programmer. If \code{String.curry} contains the code
+is the external operation to be implemented in Haskell by the programmer.
+The additional argument of type \code{ConstStore}
+represents the current set of constraints when this
+operation is called. This argument is intended to provide
+a more efficient access to binding constraints and can be
+ignored in standard operations.
+
+If \code{String.curry} contains the code
 of the Curry function \code{sconc} described above,
@@
 external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> CP.OP_List CP.C_Char
+                        -> ConstStore -> CP.OP_List CP.C_Char
 \end{curry}
@@
 Thus, we complete our example with the definition
-\begin{curry}
-external_d_C_prim_sconc s1 s2 =
+(note that we simply ignore the final argument representing the
+constraint store)
+\begin{curry}
+external_d_C_prim_sconc s1 s2 _ =
   toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
@@
 also provides a family of operations \code{fromHaskellIO$n$},
-where $n$ denotes the number of parameters of the I/O action,
-which map a Haskell I/O action into a Curry I/O action and performing
-all data conversions. For instance, if we want to implement
+where $n$ denotes the number of parameters of the I/O action.
+Each operation \code{fromHaskellIO$n$}
+maps a Haskell I/O action into a Curry I/O action and performs
+all data conversions for the arguments and result.
+For instance, if we want to implement
 an external operation to print some string as an output line,
@@
 \begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString = fromHaskellIO1 putStrLn
+external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
+                              -> CP.C_IO CP.OP_Unit
+external_d_C_prim_printString s _ = fromHaskellIO1 putStrLn s
 \end{curry}
@@
 d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 = external_d_OP_lt_ampersand_gt x1 x2
+                     -> ConstStore -> Curry_Prelude.C_Int
+d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
 \end{curry}

### 2011-12-16 lib/External_Directory.hs
Removed functions fromHaskellIO<n>, replace by toCurry

@@ -1,17 +1,18 @@
-import System.Time
 import System.Directory
 import System.IO
+import System.Time
+
 import qualified Curry_Prelude as CP
 
 external_d_C_prim_doesFileExist :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Bool
-external_d_C_prim_doesFileExist s _ = fromHaskellIO1 doesFileExist s
+external_d_C_prim_doesFileExist s _ = toCurry doesFileExist s
 
 external_d_C_prim_doesDirectoryExist :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Bool
-external_d_C_prim_doesDirectoryExist s _ = fromHaskellIO1 doesDirectoryExist s
+external_d_C_prim_doesDirectoryExist s _ = toCurry doesDirectoryExist s
 
 external_d_C_prim_fileSize :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Int
-external_d_C_prim_fileSize s _ = fromHaskellIO1
-  (\f -> do h <- openFile f ReadMode 
+external_d_C_prim_fileSize s _ = toCurry
+  (\f -> do h <- openFile f ReadMode
             i <- hFileSize h
             hClose h
             return i
@@ -19,33 +20,31 @@ external_d_C_prim_fileSize s _ = fromHaskellIO1
 
 external_d_C_prim_getModificationTime :: CP.C_String -> ConstStore
                                       -> CP.C_IO Curry_Time.C_ClockTime
-external_d_C_prim_getModificationTime s _ = fromHaskellIO1 getModificationTime s
+external_d_C_prim_getModificationTime s _ = toCurry getModificationTime s
 
 external_d_C_getCurrentDirectory :: ConstStore -> CP.C_IO (CP.C_String)
-external_d_C_getCurrentDirectory _ =
-  fromHaskellIO0 System.Directory.getCurrentDirectory
+external_d_C_getCurrentDirectory _ = toCurry getCurrentDirectory
 
 external_d_C_prim_setCurrentDirectory :: CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_setCurrentDirectory s _ = fromHaskellIO1 setCurrentDirectory s
+external_d_C_prim_setCurrentDirectory s _ = toCurry setCurrentDirectory s
 
 external_d_C_prim_getDirectoryContents :: CP.C_String -> ConstStore
[…]

### 2011-12-16 lib/External_IOExts.hs

@@ -8,12 +8,12 @@ import qualified Curry_Prelude as CP
 
 external_d_C_prim_execCmd :: CP.C_String
  -> ConstStore -> CP.C_IO (CP.OP_Tuple3 Curry_IO.C_Handle Curry_IO.C_Handle Curry_IO.C_Handle)
-external_d_C_prim_execCmd str _ = fromHaskellIO1
+external_d_C_prim_execCmd str _ = toCurry
   (\s -> do (h1,h2,h3,_) <- runInteractiveCommand s
             return (OneHandle h1, OneHandle h2, OneHandle h3)) str
 
 external_d_C_prim_connectToCmd :: CP.C_String -> ConstStore -> CP.C_IO Curry_IO.C_Handle
-external_d_C_prim_connectToCmd str _ = fromHaskellIO1
+external_d_C_prim_connectToCmd str _ = toCurry
   (\s -> do (hin,hout,herr,_) <- runInteractiveCommand s
             forkIO (forwardError herr)
             return (InOutHandle hout hin)) str
@@ -35,12 +35,12 @@ assocs :: IORef Assocs
 assocs = unsafePerformIO (newIORef [])
 
 external_d_C_prim_setAssoc :: CP.C_String -> CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_setAssoc str1 str2 _ = fromHaskellIO2
+external_d_C_prim_setAssoc str1 str2 _ = toCurry
   (\key val -> do as <- readIORef assocs
                   writeIORef assocs ((key,val):as)) str1 str2
 
 external_d_C_prim_getAssoc :: CP.C_String -> ConstStore -> CP.C_IO (CP.C_Maybe (CP.C_String))
-external_d_C_prim_getAssoc str _ = fromHaskellIO1
+external_d_C_prim_getAssoc str _ = toCurry
   (\key -> do as <- readIORef assocs
               return (lookup key as)) str
 
@@ -48,11 +48,11 @@ external_d_C_prim_getAssoc str _ = fromHaskellIO1
 -- Implementation of IORefs in Curry. Note that we store Curry values
 -- (and not the corresponding Haskell values) in the Haskell IORefs
 data C_IORef a
-     = Choice_C_IORef ID (C_IORef a) (C_IORef a)
-     | Choices_C_IORef ID ([C_IORef a])
-     | Fail_C_IORef
-     | Guard_C_IORef Constraints (C_IORef a)
[…]

### 2011-12-16 lib/External_Socket.hs

@@ -11,26 +11,26 @@ instance ConvertCurryHaskell CP.C_Int PortID where
   fromCurry i = PortNumber (fromInteger (fromCurry i))
 
 external_d_C_prim_listenOn :: CP.C_Int -> ConstStore -> CP.C_IO C_Socket
-external_d_C_prim_listenOn i _ = fromHaskellIO1 listenOn i
+external_d_C_prim_listenOn i _ = toCurry listenOn i
 
 external_d_C_listenOnFresh :: ConstStore -> CP.C_IO (CP.OP_Tuple2 CP.C_Int C_Socket)
-external_d_C_listenOnFresh _ = fromHaskellIO0 listenOnFreshPort
- where
-   listenOnFreshPort :: IO (PortID,Socket)
-   listenOnFreshPort = do
-     s <- listenOn (PortNumber aNY_PORT)
-     p <- Network.socketPort s
-     return (p,s)
+external_d_C_listenOnFresh _ = toCurry listenOnFreshPort
+  where
+  listenOnFreshPort :: IO (PortID,Socket)
+  listenOnFreshPort = do
+    s <- listenOn (PortNumber aNY_PORT)
+    p <- Network.socketPort s
+    return (p,s)
 
 external_d_C_prim_socketAccept :: C_Socket
   -> ConstStore -> CP.C_IO (CP.OP_Tuple2 CP.C_String Curry_IO.C_Handle)
 external_d_C_prim_socketAccept socket _ =
- fromHaskellIO1 (\s -> Network.accept s >>= \ (h,s,_) -> return (s,OneHandle h)) socket
+ toCurry (\s -> Network.accept s >>= \ (h,s,_) -> return (s,OneHandle h)) socket
 
 
 external_d_C_prim_waitForSocketAccept :: C_Socket -> CP.C_Int
  -> ConstStore -> CP.C_IO (CP.C_Maybe (CP.OP_Tuple2 (CP.OP_List CP.C_Char) Curry_IO.C_Handle))
-external_d_C_prim_waitForSocketAccept s i _ = fromHaskellIO2 wait s i
+external_d_C_prim_waitForSocketAccept s i _ = toCurry wait s i
 
 wait :: Socket -> Int -> IO (Maybe (String,CurryHandle))
 wait s t = do
@@ -43,9 +43,9 @@ wait s t = do
   return res
 
 external_d_C_prim_sClose :: C_Socket -> ConstStore -> CP.C_IO CP.OP_Unit
[…]

### 2011-12-16 lib/External_Time.hs

@@ -7,37 +7,34 @@ instance ConvertCurryHaskell C_ClockTime T.ClockTime where
   toCurry   (T.TOD i _) = C_CTime (toCurry i)
 
 instance ConvertCurryHaskell C_CalendarTime T.CalendarTime where
-  fromCurry (C_CalendarTime y m d h min s tz ) =
-           T.CalendarTime (fromCurry y)
-                          (toEnum (fromCurry m - 1))
-                          (fromCurry d)
-                          (fromCurry h)
-                          (fromCurry min)
-                          (fromCurry s)
-                          0 undefined undefined undefined
-                          (fromCurry tz)
-                          undefined
+  fromCurry (C_CalendarTime y m d h min s tz) =
+    T.CalendarTime  (fromCurry y)
+                    (toEnum (fromCurry m - 1))
+                    (fromCurry d)
+                    (fromCurry h)
+                    (fromCurry min)
+                    (fromCurry s)
+                    0 undefined undefined undefined
+                    (fromCurry tz)
+                    undefined
 
   toCurry (T.CalendarTime y m d h min s _ _ _ _ tz _) =
-          C_CalendarTime (toCurry y)
-                         (toCurry (fromEnum m + 1))
-                         (toCurry d)
-                         (toCurry h)
-                         (toCurry min)
-                         (toCurry s)
-                         (toCurry tz)
+    C_CalendarTime  (toCurry y)
+                    (toCurry (fromEnum m + 1))
+                    (toCurry d)
+                    (toCurry h)
+                    (toCurry min)
+                    (toCurry s)
+                    (toCurry tz)
 
-external_d_C_getClockTime ::  ConstStore -> CP.C_IO C_ClockTime
-external_d_C_getClockTime _ =
-  fromIO (T.getClockTime >>= return . toCurry)
+external_d_C_getClockTime :: ConstStore -> CP.C_IO C_ClockTime
+external_d_C_getClockTime _ = toCurry T.getClockTime
 
 external_d_C_prim_toCalendarTime :: C_ClockTime -> ConstStore -> CP.C_IO C_CalendarTime
-external_d_C_prim_toCalendarTime ct _ =
-  fromIO (T.toCalendarTime (fromCurry ct) >>= return . toCurry)
[…]

### 2011-12-16 runtime/Basics.hs

@@ -2,8 +2,6 @@
 
 module Basics
   ( module Basics
-  , module ConstStore
-  , module ID
   , module PrimTypes
   , module Search
   , module Types
@@ -12,8 +10,6 @@ module Basics
 import Data.Char(ord)
 import GHC.Exts (Int#, Char#, chr#)
 
-import ConstStore
-import ID
 import PrimTypes
 import Search
 import Types
@@ -132,27 +128,3 @@ maySwitch y (Choices_C_Success i xs) = Choices_C_Success (narrowID i) (map (& y)
 maySwitch y (Guard_C_Success cs e)   = Guard_C_Success cs (e & y)
 maySwitch y x                        = error $ "maySwitch: " ++ show y ++ " " ++ show x
 -}
-
- -- Use a Haskell IO action to implement a Curry IO action:
-fromHaskellIO0 :: ConvertCurryHaskell ca ha => IO ha -> C_IO ca
-fromHaskellIO0 = toCurry
--- fromHaskellIO0 hact = fromIO (hact >>= return . toCurry)
-
-fromHaskellIO1 :: (ConvertCurryHaskell ca ha, ConvertCurryHaskell cb hb)
-               => (ha -> IO hb) -> ca -> C_IO cb
-fromHaskellIO1 = toCurry
--- fromHaskellIO1 hact ca = fromIO (hact (fromCurry ca) >>= return . toCurry)
-
-fromHaskellIO2 :: (ConvertCurryHaskell ca ha, ConvertCurryHaskell cb hb,
-                   ConvertCurryHaskell cc hc)
-               => (ha -> hb -> IO hc) -> ca -> cb -> C_IO cc
-fromHaskellIO2 = toCurry
--- fromHaskellIO2 hact ca cb =
---   fromIO (hact (fromCurry ca) (fromCurry cb) >>= return . toCurry)
-
-fromHaskellIO3 :: (ConvertCurryHaskell ca ha, ConvertCurryHaskell cb hb,
-                   ConvertCurryHaskell cc hc, ConvertCurryHaskell cd hd)
-               => (ha -> hb -> hc -> IO hd) -> ca -> cb -> cc -> C_IO cd
-fromHaskellIO3 = toCurry
--- fromHaskellIO3 hact ca cb cc =
---  fromIO (hact (fromCurry ca) (fromCurry cb) (fromCurry cc) >>= return . toCurry)

### 2011-12-16 lib/External_IOExts.hs
Fixed some typing errors

@@ -125,7 +125,7 @@ external_d_C_newIORef :: CP.Curry a => a -> ConstStore -> CP.C_IO (C_IORef a)
 external_d_C_newIORef cv _ = toCurry (newIORef cv)
 
 external_d_C_prim_readIORef :: CP.Curry a => C_IORef a -> ConstStore -> CP.C_IO a
-external_d_C_prim_readIORef ref _ = toCurry (readIORef ref)
+external_d_C_prim_readIORef ref _ = fromIO (readIORef (fromCurry ref))
 
 external_d_C_prim_writeIORef :: CP.Curry a => C_IORef a -> a
                                            -> ConstStore -> CP.C_IO CP.OP_Unit

### 2011-12-16 lib/External_Prelude.hs

@@ -219,7 +219,6 @@ curryint2primint (Pos n) = currynat2primint n
 curryint2primint (Neg n) = negateInt# (currynat2primint n)
 curryint2primint _ = error "KiCS2 error: Prelude.curryint2primint: no ground term"
 
-
 -- ---------------------------------------------------------------------------
 -- Float
 -- ---------------------------------------------------------------------------
@@ -468,9 +467,9 @@ instance ConvertCurryHaskell C_Int Integer where
    where
     int2C_Int (I# c) = C_Int c
 
-  fromCurry (C_Int i) = toInteger (I# i)
+  fromCurry (C_Int      i) = toInteger (I# i)
   fromCurry (C_CurryInt i) = toInteger (I# (curryint2primint i))
-  fromCurry _         = error "KiCS2 error: Int data with no ground term"
+  fromCurry _              = error "KiCS2 error: Int data with no ground term"
 
 instance ConvertCurryHaskell C_Float Float where
   toCurry (F# f) = C_Float f
@@ -481,9 +480,9 @@ instance ConvertCurryHaskell C_Float Float where
 instance ConvertCurryHaskell C_Char Char where
   toCurry (C# c) = C_Char c
 
-  fromCurry (C_Char c) = C# c
+  fromCurry (C_Char    c) = C# c
   fromCurry (CurryChar c) = C# (curryChar2primChar c)
-  fromCurry _          = error "KiCS2 error: Char data with no ground term"
+  fromCurry _             = error "KiCS2 error: Char data with no ground term"
 
 instance (ConvertCurryHaskell ct ht) =>
          ConvertCurryHaskell (OP_List ct) [ht] where

### 2011-12-16 runtime/PrimTypes.hs

@@ -350,7 +350,7 @@ instance Unifiable t0 => Unifiable (C_IO t0) where
 -- END GENERATED FROM PrimTypes.curry
 
 instance ConvertCurryHaskell ca ha => ConvertCurryHaskell (C_IO ca) (IO ha)
-  where -- needs FlexibleInstances
+  where
   toCurry io  = C_IO (io >>= return . toCurry)
   fromCurry _ = error "C_IO.fromCurry: Use top-level search instead."
 

### 2011-12-16 docs/src/manual.tex
Bug fix in External_Profile.hs, manual updated

@@
 
-In order to use Haskell I/O actions as an implementation
-for external Curry I/O actions in an easy manner, the KiCS2 implementation
-also provides a family of operations \code{fromHaskellIO$n$},
-where $n$ denotes the number of parameters of the I/O action.
-Each operation \code{fromHaskellIO$n$}
-maps a Haskell I/O action into a Curry I/O action and performs
-all data conversions for the arguments and result.
+The conversion between Curry types and Haskell types,
+i.e., the family of conversion operation \code{fromCurry}
+and \code{toCurry}, is defined in the KiCS2 implementation
+for all standard data types.
+In particular, it is also defined on function types so that one can
+easily implement external Curry I/O actions by using
+Haskell I/O actions.
 For instance, if we want to implement
@@
 Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{OP_Unit} is the name
-of the Haskell representation of the Curry datatype \ccode{()}):
+\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
+of the Haskell representation of the Curry type constructor \code{IO}
+and the Curry data type \ccode{()}, respectively):
 \begin{curry}
@@
                               -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = fromHaskellIO1 putStrLn s
+external_d_C_prim_printString s _ = toCurry putStrLn s
+\end{curry}
+Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
+is transformed into a Curry I/O action \ccode{toCurry putStrLn}
+which has the type
+\begin{curry}
+CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
 \end{curry}

### 2011-12-16 lib/External_Profile.hs

@@ -10,10 +10,10 @@ external_d_C_getProcessInfos _ = fromIO $ do
                      CP.OP_List)
 
 external_d_C_garbageCollectorOff :: ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_garbageCollectorOff _ = CP.done -- not supported
+external_d_C_garbageCollectorOff _ = fromIO (return CP.OP_Unit) -- not supported
 
 external_d_C_garbageCollectorOn :: ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_garbageCollectorOn _ = CP.done -- not supported
+external_d_C_garbageCollectorOn _ = fromIO (return CP.OP_Unit) -- not supported
 
 external_d_C_garbageCollect :: ConstStore -> CP.C_IO CP.OP_Unit
 external_d_C_garbageCollect _ = toCurry performGC

### 2011-12-16 docs/src/manual.tex

@@
 {\huge User Manual}\\[4ex]
-Version of December 13, 2011\\[6ex]
+Version of December 16, 2011\\[6ex]
 \Large
@@
 \cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression.
-This mode is only useful for debugging purposes.
+resulting from the complete evaluation of the main expression
+in a tree-like structure.
+This mode is only useful for debugging or
+understanding the implementation of non-deterministic
+evaluations used in KiCS2.
 
 \item[\fbox{\code{supply $i$}}]\pindex{supply}
-Use the implementation $i$ for the identifier supply for choice structures
+Use implementation $i$ as the identifier supply for choice structures
 (see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the values \code{ghc} (default), \code{integer},
-or \code{ioref} are supported for $i$.
+Currently, the following values for $i$ are supported:
+\begin{description}
+\item[\code{integer}:]\pindex{integer}
+Use unbounded integers as choice identifiers.
+This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
+\item[\code{ghc} (default):]\pindex{ghc}
+Use a more sophisticated implementation of choice
+identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
+provided by the Glasgow Haskell Compiler.
+\item[\code{pureio}:]\pindex{pureio}
+Use IO references (i.e., memory cells) for choice identifiers.
+This is the most efficient implementation for top-level depth-first search
+but cannot be used for more sophisticated search methods
+like encapsulated search.
+\item[\code{ioref}:]\pindex{ioref}
+Use a mixture of \code{ghc} and \code{pureio}.
+IO references are used for top-level depth-first search
+and \code{ghc} identifiers are used for encapsulated search methods.
+\end{description}
 

### 2011-12-16 lib/External_Profile.hs
Fixed bug in Profile module

@@ -10,10 +10,10 @@ external_d_C_getProcessInfos _ = fromIO $ do
                      CP.OP_List)
 
 external_d_C_garbageCollectorOff :: ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_garbageCollectorOff _ = CP.done -- not supported
+external_d_C_garbageCollectorOff _ = toCurry (return () :: IO ()) -- not supported
 
 external_d_C_garbageCollectorOn :: ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_garbageCollectorOn _ = CP.done -- not supported
+external_d_C_garbageCollectorOn _ = toCurry (return () :: IO ()) -- not supported
 
 external_d_C_garbageCollect :: ConstStore -> CP.C_IO CP.OP_Unit
 external_d_C_garbageCollect _ = toCurry performGC

### 2011-12-16 docs/src/manual.tex

@@
 
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the KiCS2 implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
+In order to use Haskell I/O actions as an implementation
+for external Curry I/O actions in an easy manner, the KiCS2 implementation
+also provides a family of operations \code{fromHaskellIO$n$},
+where $n$ denotes the number of parameters of the I/O action.
+Each operation \code{fromHaskellIO$n$}
+maps a Haskell I/O action into a Curry I/O action and performs
+all data conversions for the arguments and result.
 For instance, if we want to implement
@@
 Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
+\code{External_String.hs} (where \code{OP_Unit} is the name
+of the Haskell representation of the Curry datatype \ccode{()}):
 \begin{curry}
@@
                               -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
+external_d_C_prim_printString s _ = fromHaskellIO1 putStrLn s
 \end{curry}

### 2012-01-12 papers/atps2012/paper.tex

\documentclass[english]{lni}

%\IfFileExists{latin1.sty}{\usepackage{latin1}}{\usepackage{isolatin1}}

\usepackage{graphicx}

\author{
Michael Hanus
\quad
Bj{\"o}rn Peem{\"o}ller
\quad
Fabian Reck \\
\\
Institut f\"ur Informatik, CAU Kiel, D-24098 Kiel, Germany \\
\texttt{\{mh|bjp|fre\}@informatik.uni-kiel.de}
}
\title{Search Strategies for Functional Logic Programming}
\begin{document}
\maketitle

\begin{abstract}
In this paper we discuss our practical experiences
with the use of different search strategies
in functional logic programs.
In particular, we show that complete strategies,
like breadth-first search or iterative deepening search,
are a reasonable alternative to incomplete strategies, like depth-first
search, that have been favored in the past for logic programming.
\end{abstract}

\section{Introduction}

Functional logic languages combine the most important
features of functional and logic programming in a single language
(see \cite{AntoyHanus10CACM,Hanus07ICLP} for recent surveys).
In particular, they provide higher-order functions and demand-driven
evaluation from functional programming together with logic programming features
like non-deterministic search and computing with partial information
(logic variables).
This combination
led to new design patterns \cite{AntoyHanus02FLOPS,AntoyHanus11WFLP}
and better abstractions for application programming,
e.g., as shown for programming with databases
\cite{BrasselHanusMueller08PADL,Fischer05},
GUI programming \cite{Hanus00PADL},
web programming \cite{Hanus01PADL,Hanus06PPDP,HanusKoschnicke10PADL},
or string parsing \cite{CaballeroLopez99}.
Moreover, it is also a good basis
to teach the ideas of functional and logic programming,
or declarative programming in general,
with a single computation model and programming language
\cite{Hanus97DPLE}.

An important feature of logic programming languages
is non-deterministic search.
In Prolog, which is still the standard language for logic programming,
non-deterministic search is implemented via backtracking,
which corresponds to a depth-first search traversal
through the SLD proof tree \cite{Lloyd87}.

classical view: LP = lvars, unif + Backtracking

-> FP view: add backtracking to obtain LP

beginners: backtracking considered harmful

better: complete strategies

here: complete strategies reasonable, thus, should be the default

\section{Curry}

A short standard intro to Curry.

Mention that search strategy is not fixed
 (discuss top-level vs. encaps. search)

\section{Search Strategies}

Show enacpsulated search, i.e., introduce search tree
and implementation of search strategies in search trees
(maybe BFS and IDS in the appendix).

\section{Benchmarks}

Compare DFS, BFS, IDS (top-level and encapsulated, all solutions and first sol)

Hopefully, they show what we want...

\section{Conclusions}

If efficiency important and DFS safe (e.g., finite search space): use DFS

Future work: approximate safeness of DFS

\bibliography{mh}
%\bibliography{paper}

\end{document}

### 2012-01-12 papers/atps2012/paper.tex
ATPS update (Sect. 1 finished)

@@
 \usepackage{graphicx}
+\usepackage{url}
+\usepackage{listings}
+\lstset{aboveskip=0.8ex,
+        belowskip=0.8ex,
+        xleftmargin=2ex,
+        showstringspaces=false, % no special string space
+        mathescape=true,
+        flexiblecolumns=false,
+        basewidth=0.52em,
+        basicstyle=\small\ttfamily}
+\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
+       }
+\lstnewenvironment{curry}{}{}
+\newcommand{\listline}{\vrule width0pt depth1.75ex}
 
@@
 through the SLD proof tree \cite{Lloyd87}.
+Due to this feature of Prolog,
+logic programming is often considered as unification and backtracking
+as shown by approaches to add logic programming features
+to existing functional languages
+(e.g., \cite{ClaessenLjungloef00,Hinze01}).
+This limited ``backtracking'' view of logic programming
+is also harmful to beginners when newbies
+define their family relationships using a Prolog rule like
+\begin{lstlisting}
+sibling(X,Y) :- sibling(Y,X).
+\end{lstlisting}
+In such cases, one has to explain from the beginning
+the pitfalls of backtracking.
+From a declarative point of view,
+a logic program defines a set of rules
+and a logic programming system tries to find
+a solution to a query w.r.t.\ the set of rules.
+In order to abstract from operational details,
+the search strategy has to be complete.
+Due to these considerations,
+the functional logic language Curry \cite{Hanus06Curry}
+does not fix a particular search strategy
+so that different Curry implementations can support
+different (or also several) search strategies.
+Moreover, Curry implementations also
+support encapsulated search
+where non-deterministic computations are represented
+in a data structure so that different search strategies
+can be implemented as tree traversals
+\cite{BrasselHanusHuch04JFLP,HanusSteiner98PLILP,Lux99FLOPS}.
 
-classical view: LP = lvars, unif + Backtracking
+In this paper, we present our practical results
+with different search strategies implemented
+in a new implementation of Curry, called
+KiCS2 \cite{BrasselHanusPeemoellerReck11}.
+KiCS2 compiles Curry programs into Haskell programs
+where non-deterministic computations are implemented
+as tree structures so that flexible search strategies
+are supported.
+Although the incomplete depth-first search strategy
+is the most efficient one (provided that it is able
+to find a result value),
+we show that complete strategies,
+like breadth-first search or iterative deepening search,
+are a reasonable alternative that does not force
+the programmer to consider the applied search strategy
+in his program.
 
--> FP view: add backtracking to obtain LP
-
-beginners: backtracking considered harmful
-
-better: complete strategies
-
-here: complete strategies reasonable, thus, should be the default
+In the next section, we introduce the basics of Curry
+that are necessary to understand the remaining part of the paper.
+Different search strategies and their implementation
+are discussed in Section~\ref{sec:strategies}.
+These strategies are evaluated with a number of benchmarks
+in Section~\ref{sec:benchmarks}
+before we conclude in Section~\ref{sec:conclusions}
 
@@
 \section{Search Strategies}
+\label{sec:strategies}
 
@@
 \section{Benchmarks}
+\label{sec:benchmarks}
 
@@
 \section{Conclusions}
+\label{sec:conclusions}
+
+Complete strategies could be the default!
 

### 2012-01-13 papers/atps2012/paper.tex
Section on Curry started

@@
 \newcommand{\listline}{\vrule width0pt depth1.75ex}
+\newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\ccode}[1]{``\code{#1}''}
 
@@
 
-In the next section, we introduce the basics of Curry
+In the next section, we briefly recall some principles
+of functional logic programming and the programming language Curry
 that are necessary to understand the remaining part of the paper.
@@
 
-\section{Curry}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Functional Logic Programming and Curry}
 
-A short standard intro to Curry.
+Curry \cite{Hanus06Curry}
+is a declarative multi-paradigm language combining features 
+from functional programming (demand-driven evaluation, parametric
+polymorphism, higher-order functions) and logic programming
+(computing with partial information, unification, constraints).
+Recent surveys are available in \cite{AntoyHanus10CACM,Hanus07ICLP}.
+The syntax of Curry is close to Haskell \cite{PeytonJones03Haskell}.
+Thus, variables and function names usually
+start with lowercase letters and the names of type and data constructors
+start with an uppercase letter. The application of $f$
+to $e$ is denoted by juxtaposition (``$f~e$'').
+In addition, Curry allows free (logic) 
+variables in conditions and right-hand sides of defining rules.
+The operational semantics is based on an optimal evaluation strategy
+\cite{AntoyEchahedHanus00JACM} which is a conservative extension
+of lazy functional programming and (concurrent) logic programming.
 
+A Curry program consists of the definition of data types and
+operations on these types. 
+Note that in a functional logic language
+operations might yield more than one result on the same input due to 
+the logic programming features.
+For instance, Curry contains a \emph{choice} operation defined by:
+%
+\begin{curry}
+  x ? _ = x
+  _ ? y = y
+\end{curry}
+%
+Thus, the expression \ccode{0$~$?$~$1} has two values: \code{0} and \code{1}.
+If expressions have more than one value, one wants to select
+intended values according to some constraints,
+typically in conditions of program rules.
+A \emph{rule} has the form \ccode{$f~t_1\ldots{}t_n$ | $c$ = $e$}
+where the (optional) condition $c$ is a \emph{constraint},
+i.e., an expression of the built-in type
+\code{Success}. For instance, the trivial constraint
+\code{success} is a value of type \code{Success} that
+denotes the always satisfiable constraint.
+Thus, we say that a constraint $c$ is \emph{satisfied} if it
+can be evaluated to \code{success}.
+An \emph{equational constraint} $e_1 \,\code{=:=}\, e_2$ is satisfiable
+if both sides $e_1$ and $e_2$ are reducible to unifiable values.
+Furthermore, if $c_1$ and $c_2$ are constraints,
+\code{$c_1\,$\&$\,\,c_2$} denotes their concurrent conjunction
+(i.e., both argument constraints are concurrently evaluated).
+
+As a simple example, consider the following Curry program
+which defines a polymorphic 
+data type for lists and operations to compute the 
+concatenation of lists and the last element of a list:\footnote{Note
+that lists are a built-in data type with a more convenient syntax, 
+e.g., one can write \code{[x,y,z]} instead of \code{x:y:z:[]}
+and \code{[a]} instead of the list type \ccode{List a}.}
+%
+\begin{curry}
+  data List a = [] | a : List a    --[a] denotes "List a"
+
+  -- "++" is a right-associative infix operator
+  (++) :: [a] -> [a] -> [a]
+  []     ++ ys = ys
+  (x:xs) ++ ys = x : (xs ++ ys)
+
+  last :: [a] -> a
+  last xs | (ys ++ [z]) =:= xs
+          = z                    where ys,z free
+\end{curry}
+%
+Logic programming is supported by admitting function calls with free
+variables (e.g., \code{(ys++[z])} in the rule defining \code{last})
+and constraints in the condition of a defining rule. In contrast to
+Prolog, free variables need to be declared explicitly to make their
+scopes clear (e.g., \ccode{where ys,z free} in the example).
+A conditional rule is applicable if its condition  is satisfiable.
+Thus, the rule defining \code{last} states in its condition
+that \code{z} is the last element of a given list
+\code{xs} if there exists a list \code{ys}
+such that the concatenation of \code{ys} and the one-element list
+\code{[z]} is equal to the given list \code{xs}.
+
+{\it
 Mention that search strategy is not fixed
- (discuss top-level vs. encaps. search)
+ (discuss top-level vs. encaps. search)}
 

### 2012-01-15 papers/atps2012/paper.tex
ATPS paper: Sect. 2 completed

@@
 
-Curry \cite{Hanus06Curry}
-is a declarative multi-paradigm language combining features 
+Integrated functional logic programming languages
+combine features 
 from functional programming (demand-driven evaluation, parametric
@@
 Recent surveys are available in \cite{AntoyHanus10CACM,Hanus07ICLP}.
-The syntax of Curry is close to Haskell \cite{PeytonJones03Haskell}.
-Thus, variables and function names usually
+Curry \cite{Hanus06Curry} is a functional logic language
+which extends lazy functional programming as to be found in
+Haskell \cite{PeytonJones03Haskell} and supports these features.
+
+A Curry program consists of the definition of data types and
+operations on these types. 
+Since the syntax of Curry is close to Haskell,
+variables and function names usually
 start with lowercase letters and the names of type and data constructors
@@
 variables in conditions and right-hand sides of defining rules.
-The operational semantics is based on an optimal evaluation strategy
-\cite{AntoyEchahedHanus00JACM} which is a conservative extension
-of lazy functional programming and (concurrent) logic programming.
-
-A Curry program consists of the definition of data types and
-operations on these types. 
 Note that in a functional logic language
@@
 typically in conditions of program rules.
-A \emph{rule} has the form \ccode{$f~t_1\ldots{}t_n$ | $c$ = $e$}
+A \emph{rule} has the form
+\begin{curry}
+$f~t_1\ldots{}t_n$ | $c$ = $e$
+\end{curry}
 where the (optional) condition $c$ is a \emph{constraint},
@@
 if both sides $e_1$ and $e_2$ are reducible to unifiable values.
-Furthermore, if $c_1$ and $c_2$ are constraints,
-\code{$c_1\,$\&$\,\,c_2$} denotes their concurrent conjunction
-(i.e., both argument constraints are concurrently evaluated).
+%Furthermore, if $c_1$ and $c_2$ are constraints,
+%\code{$c_1\,$\&$\,\,c_2$} denotes their concurrent conjunction
+%(i.e., both argument constraints are concurrently evaluated).
 
@@
 
-{\it
-Mention that search strategy is not fixed
- (discuss top-level vs. encaps. search)}
+Curry also offers standard features of
+functional languages, like modules or monadic I/O
+which is identical to Haskell's I/O concept \cite{Wadler97}.
+Thus, \ccode{IO $\alpha$} denotes the type of an I/O action that returns values
+of type $\alpha$.
 
+The operational semantics of Curry is based on an optimal evaluation strategy
+\cite{AntoyEchahedHanus00JACM} which is a conservative extension
+of lazy functional programming and (concurrent) logic programming.
+A big-step and a small-step operational semantics of Curry
+can be found in \cite{AlbertHanusHuchOliverVidal05}.
+Curry's semantics is sound in the sense of logic programming,
+i.e., each computed result is correct and for each
+correct result there is a more general computed one
+\cite{AntoyEchahedHanus00JACM}.
+In order to achieve completeness,
+one has to take all possible non-deterministic derivation paths
+into account. In contrast to Prolog, which fixes
+a (potentially incomplete) depth-first search strategy to find solutions,
+Curry does not fix a particular search strategy.
+Thus, different Curry implementations
+can support various search strategies.
+For instance, the Curry implementation PAKCS \cite{Hanus10PAKCS},
+which compiles Curry programs into Prolog programs,
+supports only a depth-first search strategy.
+MCC \cite{Lux99FLOPS} compiles Curry programs into C programs
+and uses also a depth-first search strategy to find all
+solutions of a given top-level goal.
+In addition, MCC offers the encapsulation of search
+(see below) so that other search strategies,
+like a complete breadth-first search strategy, can be used
+inside a Curry program.
+The Curry implementation KiCS \cite{BrasselHuch07,BrasselHuch09},
+which compiles Curry programs into Haskell programs,
+offers depth-first and breadth-first search strategies
+for top-level goals as well the encapsulation of search
+with user-definable strategies.
+In this paper we consider the Curry implementation
+KiCS2 \cite{BrasselHanusPeemoellerReck11},
+which is based on similar ideas than KiCS but uses
+a different compilation model avoiding side effects
+to enable better optimizations for target programs.
+KiCS2 also offers different search strategies for top-level goals
+and the encapsulation of search with user-definable strategies,
+which is described next.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Search Strategies}
 \label{sec:strategies}
+
+As mentioned above, Curry does not enforce a particular search
+strategy so that Curry implementation
+can provide different search strategies to find solutions
+or values for a given constraint or expression.
+The most advanced system in this respect is
+KiCS2 \cite{BrasselHanusPeemoellerReck11},
+which supports the evaluation of top-level expressions
+with depth-first search, breadth-first search,
+iterative deepening search, or parallel search strategies.
+This flexibility is possible since all operations with side effects
+are collected in monadic I/O operations \cite{Wadler97}.
+As a consequence of this computation model,
+all non-deterministic computations between I/O operations
+must be encapsulated since one cannot apply two alternative
+I/O operations to some world and proceed with two non-deterministic worlds
+(``one cannot duplicate the world'').
+Therefore, Curry offers the encapsulation of search
+by representing non-deterministic computations
+in a data structure so that the computation of solutions
+(one solution or all solutions) can conceptually be implemented
+as traversals on this data structure.
+
+Early: try \cite{HanusSteiner98PLILP,Lux99FLOPS}
+
+Later: \cite{BrasselHanusHuch04JFLP}
 
@@
 
-\bibliography{mh}
-%\bibliography{paper}
+%\bibliography{mh}
+\bibliography{paper}
 

### 2012-01-16 papers/atps2012/paper.tex

@@
 \newcommand{\ccode}[1]{``\code{#1}''}
+\newcommand{\bs}{\char92} % backslash
+\newcommand{\us}{\char95} % underscore
 
@@
 Haskell \cite{PeytonJones03Haskell} and supports these features.
+Another functional logic language based on similar principles is
+${\cal TOY}$ \cite{Lopez-FraguasSanchez-Hernandez99}.
+However, ${\cal TOY}$ does not offer flexible search strategies
+by a concept of encapsulating search
+(although it provides a concept of nested computation spaces
+in order to deal with failures in functional logic
+programming \cite{LopezSanchez04,SanchezHernandez06}).
+Therefore, we use Curry throughout this paper, although
+the concepts presented here could be also integrated in
+other functional logic languages.
 
@@
 %
-Thus, the expression \ccode{0$~$?$~$1} has two values: \code{0} and \code{1}.
+The choice operation can be used to define other non-deterministic operations
+like
+%
+\begin{curry}
+coin = 0 ? 1  
+\end{curry}
+Thus, the expression \ccode{coin} has two values: \code{0} and \code{1}.
 If expressions have more than one value, one wants to select
@@
 As mentioned above, Curry does not enforce a particular search
-strategy so that Curry implementation
+strategy. A Curry implementation
 can provide different search strategies to find solutions
@@
 iterative deepening search, or parallel search strategies.
-This flexibility is possible since all operations with side effects
+Curry supports flexibility since all operations with side effects
 are collected in monadic I/O operations \cite{Wadler97}.
@@
 all non-deterministic computations between I/O operations
-must be encapsulated since one cannot apply two alternative
-I/O operations to some world and proceed with two non-deterministic worlds
-(``one cannot duplicate the world'').
+must be encapsulated since one can not apply two alternative
+I/O operations to an existing ``world''
+and non-deterministically proceed with two alternative worlds
+(``one can not duplicate the world'').
 Therefore, Curry offers the encapsulation of search
 by representing non-deterministic computations
-in a data structure so that the computation of solutions
+in a data structure so that the computation of different solutions
 (one solution or all solutions) can conceptually be implemented
@@
 
-Early: try \cite{HanusSteiner98PLILP,Lux99FLOPS}
+An early approach to encapsulating search
+\cite{HanusSteiner98PLILP,Lux99FLOPS}
+is based on a primitive search operator
+\begin{curry}
+try :: (a->Success) -> [a->Success]
+\end{curry}
+that takes a constraint abstraction, e.g., \code{(\bs{}x->x\,=:=\,coin)},
+as input, evaluates it until the first non-deterministic step occurs,
+and returns the result: the empty list in case of failure, a list with a single
+element in case of success, or a list with at least two elements representing
+a nondeterministic choice. For instance,
+\code{try\,(\bs{}x->x\,=:=\,coin)} evaluates to
+\code{[\bs{}x->x\,=:=\,0,\,\bs{}x->x\,=:=\,1]}.
+Based on this primitive, one can define various search strategies
+to explore the search space and return its solutions.
+\cite{Lux99FLOPS} shows an implementation of this primitive.
 
-Later: \cite{BrasselHanusHuch04JFLP}
+Although typical search operators of Prolog,
+like \code{findall}, \code{once}, or negation-as-failure,
+can be implemented using the primitive \code{try},
+it became also clear that the combination with demand-driven evaluation
+and sharing causes further complications.
+For instance, in an expression like
+\begin{curry}
+let y=coin in try (\x->x=:=y)
+\end{curry}
+it is not obvious whether the evaluation of \code{coin} (introduced outside
+but demanded inside the search operator) should be encapsulated or not.
+Hence, the result of this expression might depend on the evaluation order.
+For instance,
+if \code{coin} is evaluated before the \code{try} expression,
+it results in two computations where \code{y} is bound to \code{0}
+in one computation and to \code{1} in the other computation.
+Hence, \code{try} does not encapsulate the nondeterminism of \code{coin}
+(this is the semantics of \code{try} implemented in \cite{Lux99FLOPS}).
+However, if \code{coin} is evaluated inside the capsule of \code{try}
+(because it is not demanded before), then the nondeterminism of
+\code{coin} is encapsulated.
+These and more peculiarities are discussed in \cite{BrasselHanusHuch04JFLP}.
+Furthermore, the order of the solutions might depend on the
+textual order of program rules or the evaluation time
+(e.g., in parallel implementations).
+Therefore, \cite{BrasselHanusHuch04JFLP} contains a proposal
+for another primitive search operator:
+\begin{curry}
+getSearchTree :: a -> IO (SearchTree a)
+\end{curry}
+Since \code{getSearchTree} is an I/O action, its result (in particular,
+the order of solutions) depends on the current environment, e.g.,
+time of evaluation.
+It takes an expression and delivers a search tree representing
+the search space when evaluating the input:
+\begin{curry}
+data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
+\end{curry}
+To avoid the complications w.r.t.\ shared variables,
+\code{getSearchTree} implements a \emph{strong encapsulation view},
+i.e., conceptually, the argument of \code{getSearchTree}
+is cloned before the evaluation starts in order to cut any sharing
+with the environment.
+Furthermore, the structure of the search tree is computed lazily
+so that an expression with infinitely many values does not cause
+the nontermination of the search operator if one is interested in only
+one solution.
+
+This primitive has been implemented for the first time
+in KiCS \cite{BrasselHuch07,BrasselHuch09}
+and it is also provided in KiCS2 \cite{BrasselHanusPeemoellerReck11}
+considered in this paper.
+With this primitive,
+the programmer can define its own search strategy as
+\code{SearchTree} traversals in order to
+collect all non-deterministic values into a list structure.
 
@@
 
-%\bibliography{mh}
-\bibliography{paper}
+\bibliography{mh}
+%\bibliography{paper}
 

### 2012-01-16 runtime/Search.hs
Altered Benchmarks

@@ -437,7 +437,7 @@ searchBFS act goal = do
 -- A function to increase the depth for the iterative deepening strategy
 -- (here: double the depth after reaching the depth bound)
 incrDepth4IDFS :: Int -> Int
-incrDepth4IDFS n = n + 1 -- * 2
+incrDepth4IDFS n = n + 1 -- * 2 
 
 -- Print all values of an expression with iterative deepening where
 -- the first argument is the initial depth size which will be increased

### 2012-01-16 runtime/Strategies.hs

@@ -26,7 +26,7 @@ dfsSearch (Choice x y) = dfsSearch x ++ dfsSearch y
 
 -- |Iterative deepening
 idsSearch :: Int -> SearchTree a -> [a]
-idsSearch initDepth = startIDS initDepth (2*)
+idsSearch initDepth = startIDS initDepth  (+ 1) -- (2*)
 
 idsDefaultDepth :: Int
 idsDefaultDepth = 100

### 2012-01-16 papers/atps2012/paper.tex
ATPS paper: Sect. 3 completed

@@
 \end{curry}
+\code{(Value v)} and \code{Fail} represents a single value
+or a failure (i.e., no value), respectively,
+and \code{(Or t1 t2)} represents a choice (i.e., a non-deterministic value)
+between two search trees \code{t1} and \code{t2}.
 To avoid the complications w.r.t.\ shared variables,
@@
 collect all non-deterministic values into a list structure.
+For instance, a depth-first search strategy can be easily defined
+as follows:
+\begin{curry}
+allValuesDFS :: SearchTree a -> [a]
+allValuesDFS Fail      = []
+allValuesDFS (Value x) = [x]
+allValuesDFS (Or x y)  = allValuesDFS x ++ allValuesDFS y
+\end{curry}
+Note that the lazy evaluation of traversal operations like
+\code{allValuesDFS} has the advantage that the search strategy
+is decoupled from the control. For instance, a single value
+of a non-deterministic expression $exp$ can be printed by
+\begin{curry}
+getSearchTree $exp$ >>= print . head . allValuesBFS
+\end{curry}
+even if $exp$ has infinitely many values.
+This is in contrast to Prolog's constructs for controlling
+search where different operators are necessary to compute
+one or all solutions.
 
-Show enacpsulated search, i.e., introduce search tree
-and implementation of search strategies in search trees
-(maybe BFS and IDS in the appendix).
+As a further example, we show an implementation
+of a breadth-first search strategy:
+\begin{curry}
+allValuesBFS :: SearchTree a -> [a]
+allValuesBFS t = collect [t]
 
+collect []     = []
+collect (t:ts) = values (t:ts) ++ collect (children (t:ts))
+
+values []           = []
+values (Fail:ts)    = values ts
+values (Value x:ts) = x : values ts
+values (Or _ _:ts)  = values ts
+
+children []           = []
+children (Fail:ts)    = children ts
+children (Value _:ts) = children ts
+children (Or x y:ts)  = x:y:children ts
+\end{curry}
+The operation \code{values} extracts the values
+in each level of the tree and the operation \code{children}
+extracts all direct ancestors of a level in order to recursively
+collect their values.
+
+It is well known that depth-first search lacks completeness,
+i.e., it is not able to compute all existing values.
+For instance, consider the following operation
+that non-deterministically returns all increasing numbers from a given value:
+\begin{curry}
+f n = f (n+1) ? n
+\end{curry}
+Obviously, \code{0} is a value of \code{(f 0)}, which can be printed
+by the expression
+\begin{curry}
+getSearchTree (f 0) >>= print . head . allValuesBFS
+\end{curry}
+However, if we replace \code{allValuesBFS} by \code{allValuesDFS},
+we run into an infinite loop since \ccode{?} returns
+its values in the given left-to-right order.
+In order to abstract from the details of the evaluation order,
+it would be preferable to use complete search strategies.
+Complete strategies are often neglected due to performance reasons.
+Thus, we evaluate and compare in the next section the
+performance behavior of different search strategies in KiCS2.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Benchmarks}

### 2012-01-17 papers/atps2012/paper.tex

@@
 
+In principle, the operator to encapsulate search can be
+also used to implement an interactive top-level
+to print all values of an expression as requested by the user.
+For instance, the following operation prints all elements
+of a given list as requested by the user:
+\begin{curry}
+printResults :: [_] -> IO ()
+printResults [] = putStrLn "No more values"
+printResults (x:xs) =
+  do print x
+     putStr "More values? "
+     inp <- getLine
+     if inp=="yes" then printResults xs
+                   else done
+\end{curry}
+Hence, a Prolog-like top-level interactive environment
+to show all values of an expression $exp$ in depth-first order
+can be obtained by
+\begin{curry}
+getSearchTree $exp$ >>= printResults . allValuesDFS
+\end{curry}
+In addition, we can print the results in breadth-first order
+by using \code{allValuesBFS} instead of \code{allValuesDFS}.
+Based these ideas, KiCS2 provides an interactive top-level
+where the user can select various search strategies
+(depth-first, breadth-first, iterative deepening, parallel).
+However, the top-level search in KiCS2 is not implemented
+via the primitive encapsulation operator but in more direct
+monadic style (see also \cite{BrasselHanusPeemoellerReck11})
+in order to avoid the explicit construction
+of the \code{SearchTree} structure.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-01-17 papers/atps2012/paper.tex
ATPS paper revised and IDS added

--- 2012-01-16 papers/atps2012/paper.tex
@@
 \usepackage{listings}
-\lstset{aboveskip=0.8ex,
-        belowskip=0.8ex,
+\lstset{aboveskip=1.2ex,
+        belowskip=1.2ex,
         xleftmargin=2ex,
@@
 \newcommand{\listline}{\vrule width0pt depth1.75ex}
-\newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\code}[1]{\texttt{\small{}#1}}
 \newcommand{\ccode}[1]{``\code{#1}''}
@@
 are a reasonable alternative to incomplete strategies, like depth-first
-search, that have been favored in the past for logic programming.
+search, that have been favored in the past for logic programming languages.
 \end{abstract}
@@
 Due to this feature of Prolog,
-logic programming is often considered as unification and backtracking
+the idea of logic programming is often reduced to
+the combination of unification and backtracking,
 as shown by approaches to add logic programming features
@@
 This limited ``backtracking'' view of logic programming
-is also harmful to beginners when newbies
+is also harmful to beginners, e.g., when newbies
 define their family relationships using a Prolog rule like
@@
 In such cases, one has to explain from the beginning
-the pitfalls of backtracking.
+the pitfalls of backtracking which harms the understanding
+of declarative programming.
 From a declarative point of view,
@@
 with different search strategies implemented
-in a new implementation of Curry, called
+in a new Curry system called
 KiCS2 \cite{BrasselHanusPeemoellerReck11}.
 KiCS2 compiles Curry programs into Haskell programs
-where non-deterministic computations are implemented
+where non-deterministic values or computations are represented
 as tree structures so that flexible search strategies
-are supported.
+can be supported.
 Although the incomplete depth-first search strategy
@@
 that are necessary to understand the remaining part of the paper.
-Different search strategies and their implementation
-are discussed in Section~\ref{sec:strategies}.
+The encapsulation of search and the implementation of different
+search strategies are discussed in Section~\ref{sec:strategies}.
 These strategies are evaluated with a number of benchmarks
@@
 Integrated functional logic programming languages
-combine features 
-from functional programming (demand-driven evaluation, parametric
-polymorphism, higher-order functions) and logic programming
-(computing with partial information, unification, constraints).
+combine features from functional programming and logic programming.
 Recent surveys are available in \cite{AntoyHanus10CACM,Hanus07ICLP}.
@@
 which extends lazy functional programming as to be found in
-Haskell \cite{PeytonJones03Haskell} and supports these features.
+Haskell \cite{PeytonJones03Haskell} and additionally supports
+logic programming features.
 Another functional logic language based on similar principles is
@@
 \begin{curry}
-data List a = [] | a : List a    --[a] denotes "List a"
+data List a = [] | a : List a    -- [a] denotes "List a"
 
@@
 In order to achieve completeness,
-one has to take all possible non-deterministic derivation paths
+one has to take \emph{all} possible non-deterministic derivation paths
 into account. In contrast to Prolog, which fixes
@@
 MCC \cite{Lux99FLOPS} compiles Curry programs into C programs
-and uses also a depth-first search strategy to find all
+and uses also a depth-first search strategy to find the
 solutions of a given top-level goal.
@@
 strategy. A Curry implementation
-can provide different search strategies to find solutions
+can provide various search strategies to find solutions
 or values for a given constraint or expression.
@@
 iterative deepening search, or parallel search strategies.
-Curry supports flexibility since all operations with side effects
+Curry supports this flexibility since all operations with side effects
 are collected in monadic I/O operations \cite{Wadler97}.
@@
 
-An early approach to encapsulating search
+An early approach
 \cite{HanusSteiner98PLILP,Lux99FLOPS}
+to encapsulating search in Curry
 is based on a primitive search operator
@@
 like \code{findall}, \code{once}, or negation-as-failure,
-can be implemented using the primitive \code{try},
-it became also clear that the combination with demand-driven evaluation
-and sharing causes further complications.
+can be implemented using \code{try},
+it became also clear that the combination of
+encapsulated search and demand-driven evaluation
+and sharing causes further complications \cite{BrasselHanusHuch04JFLP}.
 For instance, in an expression like
@@
 \end{curry}
-it is not obvious whether the evaluation of \code{coin} (introduced outside
+it is not obvious whether the non-determinism caused by the evaluation
+of \code{coin} (introduced outside
 but demanded inside the search operator) should be encapsulated or not.
@@
 \end{curry}
-Since \code{getSearchTree} is an I/O action, its result (in particular,
-the order of solutions) depends on the current environment, e.g.,
-time of evaluation.
 It takes an expression and delivers a search tree representing
@@
 between two search trees \code{t1} and \code{t2}.
+Since \code{getSearchTree} is an I/O action, its result (in particular,
+the order of solutions) depends on the current environment, e.g.,
+time of evaluation.
 To avoid the complications w.r.t.\ shared variables,
@@
 \code{allValuesDFS} has the advantage that the search strategy
-is decoupled from the control. For instance, a single value
-of a non-deterministic expression $exp$ can be printed by
+is decoupled from the control. For instance, we can define
+the following operation to print a single value
+of a non-deterministic expression:
 \begin{curry}
-getSearchTree $exp$ >>= print . head . allValuesBFS
+printFirstValueDFS x = getSearchTree x >>= print . head . allValuesDFS
 \end{curry}
-even if $exp$ has infinitely many values.
+Thus, \code{(printFirstValueDFS $exp$)} can print some value
+even if the non-deterministic expression $exp$ has infinitely many values.
 This is in contrast to Prolog's constructs for controlling
@@
 
-As a further example, we show an implementation
-of a breadth-first search strategy:
+It is well known that depth-first search lacks completeness,
+i.e., it might not be able to compute all existing values.
+For instance, consider the following operation
+that non-deterministically returns all increasing numbers
+from a given number \code{n}:
+\begin{curry}
+f n = f (n+1) ? n
+\end{curry}
+Although \code{0} is a value of \code{(f 0)},
+the evaluation of \code{(printFirstValueDFS (f 0))} does not terminate
+(provided that the primitive \code{getSearchTree}
+explores the non-determinism of \ccode{?} in left-to-right order).
+This problem can be avoided with complete search strategies,
+like breadth-first search strategy,
+which can be defined on \code{SearchTree} structures as follows:
 \begin{curry}
@@
 
-It is well known that depth-first search lacks completeness,
-i.e., it is not able to compute all existing values.
-For instance, consider the following operation
-that non-deterministically returns all increasing numbers from a given value:
-\begin{curry}
-f n = f (n+1) ? n
-\end{curry}
-Obviously, \code{0} is a value of \code{(f 0)}, which can be printed
-by the expression
+Using \code{allValuesBFS}, we can print a value
+of the expression \code{(f 0)} by
 \begin{curry}
@@
 \end{curry}
-However, if we replace \code{allValuesBFS} by \code{allValuesDFS},
-we run into an infinite loop since \ccode{?} returns
-its values in the given left-to-right order.
+%
 In order to abstract from the details of the evaluation order,
@@
 performance behavior of different search strategies in KiCS2.
+
+Another complete search strategy is iterative-deepening search.
+Basically, it is depth-first search strategy with a depth-bound
+which is incremented in each iteration.
+Thus, we compute in each iteration a list of values
+together with some information whether we have aborted
+(due to the depth-bound) the computation of further possible values.
+For this purpose, we define list structures that can also end
+with an \code{Abort}:
+\begin{curry}
+data AbortList a = Nil | Cons a (AbortList a) | Abort
+\end{curry}
+and define the concatenation on such lists:
+\begin{curry}
+concA :: AbortList a -> AbortList a -> AbortList a
+concA Abort       Abort       = Abort
+concA Abort       Nil         = Abort
+concA Abort       (Cons x xs) = Cons x (concA Abort xs)
+concA Nil         ys          = ys
+concA (Cons x xs) ys          = Cons x (concA xs ys)
+\end{curry}
+Now we define an operation to collect values in a search tree
+within some level bounds (to avoid the repeated collection
+of values found in each iteration):
+\begin{curry}
+collectInBounds :: Int -> Int -> SearchTree a -> AbortList a
+collectInBounds oldbound newbound st = collectLevel newbound st
+ where
+  collectLevel _ Fail      = Nil
+  collectLevel d (Value x) = if d<=newbound-oldbound then Cons x Nil
+                                                     else Nil
+  collectLevel d (Or x y)  =
+    if d>0 then concA (collectLevel (d-1) x) (collectLevel (d-1) y)
+           else Abort
+\end{curry}
+Now, the entire search strategy consists of
+repeated calls to \code{collectInBounds} as long as the result list
+is aborted. In order to experiment with different parameters,
+the initial depth bound and the method to increase the depth bound
+in each iteration are passed as parameters to the main operation:
+\begin{curry}
+allValuesIDS :: Int -> (Int -> Int) -> SearchTree a -> [a]
+allValuesIDS initdepth incrdepth st =
+  iterIDS initdepth (collectInBounds 0 initdepth st)
+ where
+  iterIDS _ Nil = []
+  iterIDS n (Cons x xs) = x : iterIDS n xs
+  iterIDS n Abort = let newdepth = incrdepth n
+                     in iterIDS newdepth (collectInBounds n newdepth st)
+\end{curry}
+%
+All operators to encapsulate search can also
+be used to implement an interactive top-level
+to print all values of an expression as requested by the user.
+For instance, the following I/O operation prints all elements
+of a given list as requested by the user:
+\begin{curry}
+printResults :: [a] -> IO ()
+printResults []     = putStrLn "No more values"
+printResults (x:xs) = do print x
+                         putStr "More values? "
+                         inp <- getLine
+                         if inp=="yes" then printResults xs
+                                       else done
+\end{curry}
+Hence, an interactive Prolog-like top-level behavior
+to show the values of an expression $exp$ in depth-first order
+can be obtained by
+\begin{curry}
+getSearchTree $exp$ >>= printResults . allValuesDFS
+\end{curry}
+In addition, we can print the results in breadth-first order
+by using \code{allValuesBFS} instead of \code{allValuesDFS}.
+Based on these ideas, KiCS2 provides an interactive top-level
+where the user can select various search strategies
+(e.g., depth-first, breadth-first, iterative deepening, parallel).
+However, the top-level search in KiCS2 is not implemented
+via the primitive encapsulation operator but in
+a monadic style (see also \cite{BrasselHanusPeemoellerReck11})
+in order to avoid the explicit construction
+of the \code{SearchTree} structure.
+Thus, we compare in the next section also top-level search
+with encapsulated search in order to evaluate the potential overhead
+caused by abstracting and programming with search structures.
 

### 2012-01-17 papers/atps2012/paper.tex
ATPS: Sect. 4 started

@@
 \section{Functional Logic Programming and Curry}
+\label{sec:curry}
 
@@
 %
-All operators to encapsulate search can also
+Although this definition describes the strategy of
+iterative deepening search, it misses one important motivation.
+An advantage of depth-first search
+in comparison to breadth-first search is its memory behavior:
+whereas breadth-first search has to store all nodes
+in a level of the search tree in parallel, depth-first search
+need only to store the nodes in a branch from the root
+to the current node under investigation.
+Since iterative deepening uses depth-first search in each iteration,
+it should have a memory behavior similarly to depth-first search
+and, in case of wide search trees,
+better than breadth-first search.
+The price for this behavior is the recomputation of the initial
+goal in each iteration. Since we defined iteration deepening
+on a search tree, recomputation is not required. Instead, the
+already evaluated part of the search tree is kept in memory.
+Therefore, we also implemented the iterative deepening strategy
+for top-level goals without an explicit search tree
+but with a recomputation of the initial goal in each iteration
+(see Section~\ref{sec:benchmarks}).
+
+The various operators to encapsulate search can also
 be used to implement an interactive top-level
@@
 Based on these ideas, KiCS2 provides an interactive top-level
-where the user can select various search strategies
-(e.g., depth-first, breadth-first, iterative deepening, parallel).
+where the user can select various search strategies,
+e.g., depth-first, breadth-first, iterative deepening, or
+an experimental implementation of parallel search.
 However, the top-level search in KiCS2 is not implemented
-via the primitive encapsulation operator but in
+via the primitive encapsulation operators but in
 a monadic style (see also \cite{BrasselHanusPeemoellerReck11})
@@
 
-Compare DFS, BFS, IDS (top-level and encapsulated, all solutions and first sol)
+In this section we evaluate the practical bevavior
+of the various search strategies discussed so far.
+Since they are only supported by the Curry implementation KiCS2,
+we use only this system for our evaluation.
+A general comparison of KiCS2 and other Curry implementations
+can be found in \cite{BrasselHanusPeemoellerReck11}.
+Since KiCS2 compiles Curry programs into Haskell programs,
+we used the Glasgow Haskell Compiler
+(GHC 7.0.4, option -O2) to compile and execute
+the generated target programs.
+All benchmarks were executed on a Linux machine
+running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor.
+The timings were performed with the time command measuring the
+execution time (in seconds) of a compiled executable for each benchmark.
+Due to the lack of precise measurements of the space behavior,
+we measured only the execution times where
+``oom'' denotes a memory overflow in a computation.
 
-Hopefully, they show what we want...
+\begin{figure}[ht]
+\centering
+\begin{tabular}{|@{~~}l@{~~}|*{7}{@{~~}r@{~~}|}}
+\hline
+Program &   DFS  & IDS(+1) & IDS(*2) & eDFS & eBFS & eIDS(+1) & eIDS(*2) \\\hline
+PermSort  &     \\
+Last      &     \\
+Half      &     \\
+\hline
+\end{tabular}
+\caption{Benchmarks: computing all values}
+\label{fig:bench-all-solutions}
+\end{figure}
 
+Figure~\ref{fig:bench-all-solutions} shows the benchmark results
+when all values of a given expression are computed.
+All benchmark programs are non-deterministic programs.
+For instance, ``PermSort'' sorts
+a list containing 15 elements by enumerating all permutations
+and selecting the sorted ones,
+``Last'' compute the last element \code{x}
+of a list \code{xs} containing 100,000 elements
+by solving the equation \ccode{ys++[x] =:= xs}
+(see Section~\ref{sec:curry}),
+``Half'' computes the half \code{y} of a natural number \code{x}
+(in Peano representation) by solving the equation \code{y+y=:=x},
+\ldots
+
+The columns in Figure~\ref{fig:bench-all-solutions}
+are the various search strategies considered in this paper.
+``DFS'' denotes the top-level depth-first search strategy
+which is similar to \code{allValuesDFS} but implements
+this strategy in a monadic style without the explicit
+construction of a search tree.
+The encapsulated search strategies are prefixed by ``e'',
+i.e., ``eDFS'' and ``eBFS'' correspond to \code{allValuesDFS}
+and \code{allValuesBFS}, respectively.
+``eIDS(+1)'' corresponds to \code{allValuesIDS}
+with an initial depth bound of \code{10} and the depth increment
+operation \code{(+1)}, whereas ``eIDS(*2)'' uses the same initial depth bound
+but the depth increment operation \code{(*2)}, i.e., to depth bound
+is doubled in each iteration.
+Finally, ``IDS(+1)'' and ``IDS(*2)'' are iterative deepening
+strategies with similar parameters but recompute the initial
+expression in each iteration in order to trade space for run time,
+as discussed in Section~\ref{sec:strategies}.
+
+{\sc should we also show relative run times?}
+
+As one can see, depth-first search is the most efficient
+search strategy.
+However, complete search strategies are viable alternatives.
+Their overhead is acceptable taking into account the fact
+that one need not to reason about the details of exploring the
+search space. The behavior of iterative deepening is largely
+influenced by the parameters.
+A constant increment of the depth bound causes a big overhead
+compared to doubling the depth bound in each iteration.
+Breadth-first search seems also a good strategy
+taking into account the size of memory provided by modern computers.
+However, there are extreme cases when the search tree is wide
+where iterative deepening with recomputation is the single
+strategy which is able to compute a result.
+This is shown in Figure~\ref{fig:bench-first-solution}
+that contains benchmarks to compute only a first value
+of an expression.
+
+\begin{figure}[ht]
+\centering
+\begin{tabular}{|@{~~}l@{~~}|*{7}{@{~~}r@{~~}|}}
+\hline
+Program &   DFS  & IDS(+1) & IDS(*2) & eDFS & eBFS & eIDS(+1) & eIDS(*2) \\\hline
+PermSort  &     \\
+Last      &     \\
+Half      &     \\
+\hline
+\end{tabular}
+\caption{Benchmarks: computing a single value}
+\label{fig:bench-first-solution}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}

### 2012-01-17 papers/atps2012/paper.tex

@@
 like breadth-first search or iterative deepening search,
-are a reasonable alternative to incomplete strategies, like depth-first
+are a viable alternative to incomplete strategies, like depth-first
 search, that have been favored in the past for logic programming languages.
@@
 All benchmark programs are non-deterministic programs.
-For instance, ``PermSort'' sorts
-a list containing 15 elements by enumerating all permutations
-and selecting the sorted ones,
+``PermSort'' sorts a list containing 15 elements by enumerating
+all permutations and selecting the sorted ones,
 ``Last'' compute the last element \code{x}
@@
 (in Peano representation) by solving the equation \code{y+y=:=x},
+``Embed'' searches for an embedding of a message
+which is ``hidden'' in another message
+(see \cite[Sect.~4.1]{Antoy10JSC}),
 \ldots
@@
 
-Complete strategies could be the default!
+We discussed the use of different search strategies
+in functional logic programs.
+In order to enable user-programmable search strategies,
+modern functional logic languages like Curry
+provide primitives to represent non-deterministic computations
+or values as data structures that can be traversed like
+any term structure.
+The Curry implementation KiCS2, considered in this paper,
+provides a primitive \code{getSearchTree}
+that returns a tree representation of a non-deterministic computation.
+This representation can be used to define various search strategies,
+like depth-first search, breadth-first search, or iterative deepening
+search.
 
-If efficiency important and DFS safe (e.g., finite search space): use DFS
-
-Future work: approximate safeness of DFS
+We have practically evaluated and compared
+the efficiency of these strategies.
+The benchmarks indicated that complete strategies
+are a viable alternative to incomplete strategies,
+that have been favored in the past due to limited memory requirements.
+Thus, it could be reasonable to make
+complete strategies the default in functional logic languages.
+This would have a good impact on teaching declarative programming
+to beginners.
+Moreover, decoupling non-deterministic programs
+from their search strategy could lead to a more declarative
+programming style (instead of the use of predicates with
+side effects as in Prolog) and enable more potential for
+optimization, e.g., parallel search strategies.
+Furthermore, if efficiency and memory limitations are important,
+one can still use an efficient strategy like depth-first search
+provided that it is able to find all solutions
+(e.g., in case of a finite search space).
+It is an interesting topic for future work
+to statically approximate situations where
+the use of incomplete strategies is sufficient.
 

### 2012-01-18 papers/atps2012/paper.tex
added part of the benchmark results

@@
 Program     &   DFS  & IDS(+1) & IDS(*2) & eDFS  & eBFS  & eIDS(+1) & eIDS(*2) \\\hline
-PermSort  &     \\
+PermSort    &  13.12 & 841.58  & 32.77   & 38.48 & 66.82 &    72.64 &    46.79 \\
 Last        &     \\
 Half        &     \\
+Graph       &   1.76 &   52.53 &    2.40 & 3.51  & 4.22  &     4.26 &     3.83 \\
+HorseMan    &   2.64 &  639.64 &    6.63 & 3.23  & 3.29  &     3.20 &     3.17 \\
+MAC         &   6.80 &  544.05 &    7.84 & 21.19 & 18.34 &    18.59 &    21.03 \\
+Queens      &  24.74 &  374.64 &   33.62 & 42.71 & 43.98 &    44.01 &    42.65
 \hline

### 2012-01-18 papers/atps2012/paper.tex
ATPS paper: Typos fixed

--- 2012-01-17 papers/atps2012/paper.tex
@@
 in Section~\ref{sec:benchmarks}
-before we conclude in Section~\ref{sec:conclusions}
+before we conclude in Section~\ref{sec:conclusions}.
 
@@
 in a level of the search tree in parallel, depth-first search
-need only to store the nodes in a branch from the root
+only needs to store the nodes in the branch from the root
 to the current node under investigation.
@@
 The price for this behavior is the recomputation of the initial
-goal in each iteration. Since we defined iteration deepening
-on a search tree, recomputation is not required. Instead, the
+goal in each iteration. 
+However, since we defined iterative deepening on a search tree,
+recomputation is not required. Instead, the
 already evaluated part of the search tree is kept in memory.
@@
 The various operators to encapsulate search can also
-be used to implement an interactive top-level
+be used to implement an interactive top-level search
 to print all values of an expression as requested by the user.
-For instance, the following I/O operation prints all elements
-of a given list as requested by the user:
+For instance, the following I/O operation interactively 
+prints all elements of a given list:
 \begin{curry}
@@
 by using \code{allValuesBFS} instead of \code{allValuesDFS}.
-Based on these ideas, KiCS2 provides an interactive top-level
+Based on these ideas, KiCS2 provides an interactive top-level search
 where the user can select various search strategies,
@@
 of the \code{SearchTree} structure.
-Thus, we compare in the next section also top-level search
-with encapsulated search in order to evaluate the potential overhead
+Thus, in the next section we also compare the top-level search
+with the encapsulated search in order to evaluate the potential overhead
 caused by abstracting and programming with search structures.
@@
 
-In this section we evaluate the practical bevavior
+In this section we evaluate the practical behavior
 of the various search strategies discussed so far.
@@
 the generated target programs.
-All benchmarks were executed on a Linux machine
-running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor.
+All benchmarks were executed on a 32bit Linux machine
+running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor and 4 GiB RAM.
 The timings were performed with the time command measuring the
@@
 all permutations and selecting the sorted ones,
-``Last'' compute the last element \code{x}
+``Last'' computes the last element \code{x}
 of a list \code{xs} containing 100,000 elements

### 2012-01-18 papers/atps2012/paper.tex
added new benchmark results to atps paper

--- 2012-01-18 papers/atps2012/paper.tex
@@
 Program     &   DFS  & IDS(+1) & IDS(*2) & eDFS  & eBFS  & eIDS(+1) & eIDS(*2) \\\hline
-PermSort  &     \\
-Last      &     \\
-Half      &     \\
+PermSort    &  13.12 &  841.58 &   32.77 & 38.48 & 66.82 &    72.64 &    46.79 \\
+Last        &   0.10 & >10min  &    0.26 &  0.34 &  0.18 &     6.58 &     0.38 \\
+Half        &   0.67 &  561.45 &    1.62 &  1.43 &  1.22 &     1.86 &     1.71 \\
+Graph       &   1.76 &   52.53 &    2.40 & 3.51  & 4.22  &     4.26 &     3.83 \\
+HorseMan    &   2.64 &  639.64 &    6.63 & 3.23  & 3.29  &     3.20 &     3.17 \\
+MAC         &   6.80 &  544.05 &    7.84 & 21.19 & 18.34 &    18.59 &    21.03 \\
+Queens      &  24.74 &  374.64 &   33.62 & 42.71 & 43.98 &    44.01 &    42.65
 \hline

### 2012-01-18 papers/atps2012/paper.tex
ATPS: Improved crossover from bfs to ids

--- 2012-01-18 papers/atps2012/paper.tex
@@
 which can be defined on \code{SearchTree} structures as follows:
+%
 \begin{curry}
@@
 \end{curry}
+%
 The operation \code{values} extracts the values
@@
 of the expression \code{(f 0)} by
+%
 \begin{curry}
@@
 it would be preferable to use complete search strategies.
-Complete strategies are often neglected due to performance reasons.
-Thus, we evaluate and compare in the next section the
-performance behavior of different search strategies in KiCS2.
+However, complete strategies are often neglected due to performance reasons.
+For example, the breadth-first search strategy stores all child nodes 
+of a tree level in a list to be explored later, which, as the search space
+potentially doubles on each level, may lead to an exponentially
+growing memory usage.
 
-Another complete search strategy is iterative-deepening search.
+Another complete search strategy with a superior memory behavior
+is iterative-deepening search.
 Basically, it is depth-first search strategy with a depth-bound
@@
 with an \code{Abort}:
+%
 \begin{curry}
@@
 \end{curry}
+%
 and define the concatenation on such lists:
+%
 \begin{curry}
@@
 \end{curry}
+%
 Now we define an operation to collect values in a search tree
@@
 of values found in each iteration):
+%
 \begin{curry}
@@
 \end{curry}
+%
 Now, the entire search strategy consists of
@@
 in each iteration are passed as parameters to the main operation:
+%
 \begin{curry}
@@
 %
-Although this definition describes the strategy of
-iterative deepening search, it misses one important motivation.
-An advantage of depth-first search
+The key advantage of depth-first search
 in comparison to breadth-first search is its memory behavior:
@@
 prints all elements of a given list:
+%
 \begin{curry}
@@
 \end{curry}
+%
 Hence, an interactive Prolog-like top-level behavior
@@
 of the \code{SearchTree} structure.
-Thus, in the next section we also compare the top-level search
-with the encapsulated search in order to evaluate the potential overhead
+Thus, in the next section we both compare the different search strategies
+introduced above as well as the top-level search
+with the encapsulated search, in order to evaluate the potential overhead
 caused by abstracting and programming with search structures.
@@
 \hline
-Program &   DFS  & IDS(+1) & IDS(*2) & eDFS & eBFS & eIDS(+1) & eIDS(*2) \\\hline
-PermSort  &     \\
-Last      &     \\
-Half      &     \\
+Program  &   DFS  & IDS(+1)& IDS(*2)&  eDFS  &  eBFS  &eIDS(+1)&eIDS(*2)\\
+\hline
+PermSort &  13.12 & 841.58 &  32.77 &  38.48 &  66.82 &  72.64 &  46.79 \\
+Last     &        &        &        &        &        &        &        \\
+Half     &        &        &        &        &        &        &        \\
+Graph    &   1.76 &  52.53 &   2.40 &   3.51 &   4.22 &   4.26 &   3.83 \\
+HorseMan &   2.64 & 639.64 &   6.63 &   3.23 &   3.29 &   3.20 &   3.17 \\
+MAC      &   6.80 & 544.05 &   7.84 &  21.19 &  18.34 &  18.59 &  21.03 \\
+Queens   &  24.74 & 374.64 &  33.62 &  42.71 &  43.98 &  44.01 &  42.65 \\
 \hline
@@
 \hline
-Program &   DFS  & IDS(+1) & IDS(*2) & eDFS & eBFS & eIDS(+1) & eIDS(*2) \\\hline
-PermSort  &     \\
-Last      &     \\
-Half      &     \\
+Program  &   DFS  & IDS(+1)& IDS(*2)&  eDFS  &  eBFS  &eIDS(+1)&eIDS(*2)\\
+\hline
+PermSort &        &        &        &        &        &        &        \\
+Last     &        &        &        &        &        &        &        \\
+Half     &        &        &        &        &        &        &        \\
 \hline

### 2012-01-18 papers/atps2012/paper.tex

@@
 PermSort &  13.12 & 841.58 &  32.77 &  38.48 &  66.82 &  72.64 &  46.79 \\
-Last     &        &        &        &        &        &        &        \\
-Half     &        &        &        &        &        &        &        \\
+Last     &   0.10 & >10min &   0.26 &   0.34 &   0.18 &   6.58 &   0.38 \\
+Half     &   0.67 & 561.45 &   1.62 &   1.43 &   1.22 &   1.86 &   1.71 \\
 Graph    &   1.76 &  52.53 &   2.40 &   3.51 &   4.22 &   4.26 &   3.83 \\

### 2012-01-18 papers/atps2012/paper.tex
atps update

--- 2012-01-18 papers/atps2012/paper.tex
@@
 PermSort &  13.12 & 841.58 &  32.77 &  38.48 &  66.82 &  72.64 &  46.79 \\
-Last     &        &        &        &        &        &        &        \\
-Half     &        &        &        &        &        &        &        \\
+Last     &   0.10 & $>$10min &   0.26 &   0.34 &   0.18 &   6.58 &   0.38 \\
+Half     &   0.67 & 561.45 &   1.62 &   1.43 &   1.22 &   1.86 &   1.71 \\
 Graph    &   1.76 &  52.53 &   2.40 &   3.51 &   4.22 &   4.26 &   3.83 \\
@@
 Figure~\ref{fig:bench-all-solutions} shows the benchmark results
-when all values of a given expression are computed.
+when all values of a given expression are computed
+with various search strategies.
 All benchmark programs are non-deterministic programs.
@@
 ``Last'' computes the last element \code{x}
-of a list \code{xs} containing 100,000 elements
+of a list \code{xs} containing 10,000 elements
 by solving the equation \ccode{ys++[x] =:= xs}
@@
 (in Peano representation) by solving the equation \code{y+y=:=x},
-``Embed'' searches for an embedding of a message
-which is ``hidden'' in another message
-(see \cite[Sect.~4.1]{Antoy10JSC}),
-\ldots
+%``Embed'' searches for an embedding of a message
+%which is ``hidden'' in another message
+%(see \cite[Sect.~4.1]{Antoy10JSC}),
+``Graph'' computes some path in a graph where the edges
+are represented by a non-deterministic ``successor'' operation,
+``HorseMan'' computes the numbers of horses and men
+from given numbers of heads and feet by searching for
+appropriate numbers in Peano representation,
+``MAC'' solves the ``Missionaries and Cannibals'' puzzle
+(see \cite[Sect.~3.1]{AntoyHanus02FLOPS})
+for reasonable numbers of missionaries and cannibals
+to obtain measurable run times,
+and ``Queens'' computes safe placements of queens on a chess board.
 
@@
 
-{\sc should we also show relative run times?}
-
-As one can see, depth-first search is the most efficient
+The table entries in Figure~\ref{fig:bench-all-solutions}
+contain the run times in seconds to compute all values of the initial
+expression.
+As one can see, the top-level depth-first search is the most efficient
 search strategy.
-However, complete search strategies are viable alternatives.
-Their overhead is acceptable taking into account the fact
+The encapsulated version of depth-first search (eDFS)
+with the explicit construction of the search tree
+causes some overhead, since the search strategy
+is encoded in the source program rather than in the run-time system
+as in the top-level search (DFS).
+The benchmarks also show that
+complete search strategies, like ``eBFS'' and ``eIDS(*2)''
+are viable alternatives to the incomplete depth-first search strategy.
+Their overhead (sometimes they are even faster than ``eDFS'',
+but they are always slower than top-level search)
+is acceptable taking into account the fact
 that one need not to reason about the details of exploring the
@@
 A constant increment of the depth bound causes a big overhead
-compared to doubling the depth bound in each iteration.
-Breadth-first search seems also a good strategy
+compared to doubling the depth bound in each iteration,
+in particular, when the search tree is slim as in ``Last''.
+This is even worse for top-level iterative deepening
+which recomputes the initial expression in each iteration,
+see ``IDS(+1)''.
+
+Breadth-first search seems a good strategy
 taking into account the size of memory provided by modern computers.

### 2012-01-18 papers/atps2012/paper.tex
ATPS: reference added
Added Benchmark results for first Solution

@@
 Curry does not fix a particular search strategy.
+Actually, descriptions of the model-theoretic \cite{GonzalezEtAl99}
+or operational \cite{AlbertHanusHuchOliverVidal05}
+semantics of functional logic languages do not take
+a search strategy into account.
 Thus, different Curry implementations
@@
 \hline
-PermSort &        &        &        &        &        &        &        \\
-Last     &        &        &        &        &        &        &        \\
-Half     &        &        &        &        &        &        &        \\
+PermSort &  12.57 & 804.82 &  32.09 &  34.75 &  67.02 &  70.47 &  46.66 \\
+Last     &   0.10 & 661.92 &   0.24 &   0.35 &   0.19 &   6.57 &   0.39 \\
+Half     &   0.34 &  72.64 &   0.58 &   0.69 &   0.59 &   0.78 &   0.78 \\
+Graph    &   0.00 &   0.10 &   0.04 &   0.00 &   0.04 &   0.04 &   0.03 \\
+HorseMan &   2.06 & 636.62 &   6.64 &   2.52 &   2.60 &   2.61 &   2.57 \\
+MAC      &   0.14 &  38.75 &   0.94 &   0.94 &   2.81 &   2.75 &   2.79 \\
+Queens   &   1.40 & 370.26 &  33.65 &   2.44 &   8.95 &   8.86 &   3.04 \\
+NDNums   &   oom  &  47.89 &   0.14 &    oom &   oom  &    oom &   0.47 \\
 \hline

### 2012-01-18 papers/atps2012/paper.tex
updated benchmarks for Last with all Solutions

@@
 PermSort &  13.12 & 841.58 &  32.77 &  38.48 &  66.82 &  72.64 &  46.79 \\
-Last     &   0.10 & $>$10min &   0.26 &   0.34 &   0.18 &   6.58 &   0.38 \\
+Last     &   0.10 & 660.08 &   0.26 &   0.34 &   0.19 &   6.59 &   0.39 \\
 Half     &   0.67 & 561.45 &   1.62 &   1.43 &   1.22 &   1.86 &   1.71 \\

### 2012-01-18 papers/atps2012/paper.tex
ATPS: Benchmark comments improved

@@
 \begin{document}
+\pagestyle{plain} % remove it in the camera-ready version
 \maketitle
@@
 Due to the lack of precise measurements of the space behavior,
-we measured only the execution times where
-``oom'' denotes a memory overflow in a computation.
+we measured only the execution times.
 
@@
 search space. The behavior of iterative deepening is largely
-influenced by the parameters.
+influenced by its parameterization.
 A constant increment of the depth bound causes a big overhead
@@
 
-Breadth-first search seems a good strategy
-taking into account the size of memory provided by modern computers.
-However, there are extreme cases when the search tree is wide
-where iterative deepening with recomputation is the single
-strategy which is able to compute a result.
-This is shown in Figure~\ref{fig:bench-first-solution}
-that contains benchmarks to compute only a first value
-of an expression.
-
-\begin{figure}[ht]
+\begin{figure}[t]
 \centering
@@
 \end{figure}
+
+The benchmarks indicate that breadth-first search seems to be a good strategy
+taking into account the size of memory provided by modern computers.
+However, if the search tree is wide (i.e., the nodes in each level
+increases with the depth of the tree),
+then iterative deepening is superior to breadth-first search.
+To examine such a case, we added a benchmark ``NDNums''
+which defines a non-deterministic operation
+with a high branching factor that returns all inreasing numbers
+\begin{curry}
+g n = g (n+1) ? n ? g (n+1)
+\end{curry}
+and solve the equation \ccode{g 0 =:= 29}.
+Obviously, the computation of all values will never terminate.
+Therefore, we executed the benchmarks to compute
+only a first value of an expression via different search strategies.
+Figure~\ref{fig:bench-first-solution} contains the
+corresponding results where
+``oom'' denotes a memory overflow in a computation.
+As one can see, iterative deepening is the only search strategy
+that is able to find a solution in all examples.
 

### 2012-01-18 papers/atps2012/paper.tex
ATPS: System parameters of adamo
fixed typo

--- 2012-01-18 papers/atps2012/paper.tex
@@
 offers depth-first and breadth-first search strategies
-for top-level goals as well the encapsulation of search
+for top-level goals as well as the encapsulation of search
 with user-definable strategies.

### 2012-01-18 papers/atps2012/paper.tex
Final changes to ATPS version

--- 2012-01-18 papers/atps2012/paper.tex
@@
         flexiblecolumns=false,
-        basewidth=0.52em,
+        basewidth=0.6em,
         basicstyle=\small\ttfamily}
@@
 \begin{curry}
-data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
+data SearchTree a = Value a
+                  | Fail
+                  | Or (SearchTree a) (SearchTree a)
 \end{curry}
@@
 \begin{curry}
-printFirstValueDFS x = getSearchTree x >>= print . head . allValuesDFS
+printFirstValueDFS x =
+  getSearchTree x >>= print . head . allValuesDFS
 \end{curry}
@@
   collectLevel _ Fail      = Nil
-  collectLevel d (Value x) = if d<=newbound-oldbound then Cons x Nil
+  collectLevel d (Value x) = if d <= newbound - oldbound
+                             then Cons x Nil
                              else Nil
   collectLevel d (Or x y)  =
-    if d>0 then concA (collectLevel (d-1) x) (collectLevel (d-1) y)
+    if d>0
+    then concA (collectLevel (d-1) x) (collectLevel (d-1) y)
     else Abort
@@
   iterIDS n (Cons x xs) = x : iterIDS n xs
-  iterIDS n Abort = let newdepth = incrdepth n
+  iterIDS n Abort =
+    let newdepth = incrdepth n
      in iterIDS newdepth (collectInBounds n newdepth st)
@@
 All benchmarks were executed on a 32bit Linux machine
-running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor and 4 GiB RAM.
+running Ubuntu 11.10 with an Intel Core 2 Duo (2.13~GHz) processor and 4~GiB RAM.
 The timings were performed with the time command measuring the

### 2012-01-25 talks/ag2011/talk1.tex
Last years AG talks moved

--- 2011-06-16 talks/badhonnef2011/talk.tex
@@
 
-\date{May 04, 2011}
+\date{June 08, 2011}
 
@@
 \begin{itemize}
-\item Lazy functional logic programing language.
-\item Haskell-like syntax.
-\item Extended by non-determinism, free variables, constraints, unification.
+\item Lazy functional logic programming language
+\item Haskell-like syntax
+\item Extended by non-determinism, free variables, constraints, unification
 \end{itemize}
@@
 \begin{itemize}
-\item Compile Curry programs into Haskell
+\item Compile Curry programs into Haskell programs
 \item Reuse features of Haskell (lazy evaluation, higher-order functions)
@@
 \begin{itemize}
-  \item No built-in non-determinism in Haskell.
-  \item Idea: Explicit representation in data types as constructors.
+  \item No built-in non-determinism in Haskell
+  \item Idea: Explicit representation in data types as constructors
 \end{itemize}
@@
 \begin{itemize}
-  \item Operations have to deal with non-deterministic values.
-  \item Non-determinism in arguments is propagated.
+  \item Operations have to deal with non-deterministic values
+  \item Non-determinism in arguments is propagated
 %   \item Purely functional Curry programs are Haskell programs (no runtime overhead) (first-order)
@@
   \item Yields two results \verb!True! and \verb!False!
-  \item Only \verb!False! is correct w.r.t. the semantics of Curry.
+  \item Only \verb!False! is correct w.r.t. the semantics of Curry
 \end{itemize}
@@
 \begin{itemize}
-\item Non-deterministic functions need an additional parameter
-\item An identifier is passed to functional argument when applied
+\item Non-deterministic functions need an additional parameter for fresh \verb!ID!s
+\item The identifier is passed to the functional argument when applied
 \end{itemize}
@@
   \end{itemize}
+\item Deterministic variant is used inside other deterministic functions
 \item Benchmarks show enormously improved performance
@@
 \begin{center}
-\includegraphics[width=9cm]{reverse}
+\includegraphics[width=9cm]{gfx/reverse}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=9cm]{reverseho}
+\includegraphics[width=9cm]{gfx/reverseho}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=9cm]{permsort}
+\includegraphics[width=9cm]{gfx/permsort}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=9cm]{half}
+\includegraphics[width=9cm]{gfx/half}
 \begin{program}
@@
 \begin{itemize}
-\item Transformation of a functional logic program into a functional program.
-\item Resulting program can be optimized by the Haskell compiler.
-\item Our Approach can compete with or outperform existing implementations.
-\item A formal description of our approach can be found in our paper (WFLP 2011).
+\item Transformation of a functional logic program into a functional program
+\item Resulting program can be optimized by the Haskell compiler
+\item Our approach can compete with or outperform existing implementations
+\item A formal description of our approach can be found in our paper
 \end{itemize}

### 2012-02-06 papers/wlp2011/paper.tex
Unused environment removed; authors formatted using \and

@@
 
-\usepackage{a4,a4wide}
 \usepackage{url}
@@
 \usepackage{listings}
-\lstset{aboveskip=0.8ex,
-        belowskip=0.8ex,
+\lstset{aboveskip=1.0ex,
+        belowskip=1.0ex,
         showstringspaces=false, % no special string space
@@
 
-% Redefine description env as equation list to gain key alignment
-\usepackage{eqlist}
-\let\description=\eqlist
-\let\enddescription=\endeqlist
-\let\eqlistlabel\descriptionlabel
-
 \begin{document}
-%\pagestyle{plain}
+\pagestyle{plain} % remove for camera ready version
 \sloppy
@@
 Bernd Bra{\ss}el
-\kern1em
+\and
 Michael Hanus
-\kern1em
+\and
 Bj{\"o}rn Peem{\"o}ller
-\kern1em
+\and
 Fabian Reck
@@
 \begin{haskell}
-  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
+  data List a = Nil | Cons a (List a)
+              | Choice ID (List a) (List a) | Fail
 \end{haskell}

### 2012-02-14 papers/atps2012/paper.tex
Minor improvements and additions to ATPS'12 paper

@@
 \documentclass[english]{lni}
-
-%\IfFileExists{latin1.sty}{\usepackage{latin1}}{\usepackage{isolatin1}}
 
@@
 \begin{document}
-\pagestyle{plain} % remove it in the camera-ready version
 \maketitle
@@
 or string parsing \cite{CaballeroLopez99}.
-Moreover, it is also a good basis
+Moreover, it is a good basis
 to teach the ideas of functional and logic programming,
@@
 \cite{Hanus97DPLE}.
+The operational principles of functional logic languages
+have also been used for other computation tasks,
+like inverse computations \cite{AbramovGlueckKlimov06},
+partial evaluation \cite{AlpuenteFalaschiVidal98TOPLAS}, or
+generation of test cases \cite{FischerKuchen07,RuncimanNaylorLindblad08}.
 
@@
 KiCS2 compiles Curry programs into Haskell programs
-where non-deterministic values or computations are represented
+where non-deterministic values and computations are represented
 as tree structures so that flexible search strategies
@@
 if both sides $e_1$ and $e_2$ are reducible to unifiable values.
-%Furthermore, if $c_1$ and $c_2$ are constraints,
-%\code{$c_1\,$\&$\,\,c_2$} denotes their concurrent conjunction
-%(i.e., both argument constraints are concurrently evaluated).
 
@@
 element in case of success, or a list with at least two elements representing
-a nondeterministic choice. For instance,
+a non-deterministic choice. For instance,
 \code{try\,(\bs{}x->x\,=:=\,coin)} evaluates to
@@
 in one computation and to \code{1} in the other computation.
-Hence, \code{try} does not encapsulate the nondeterminism of \code{coin}
+Hence, \code{try} does not encapsulate the non-determinism of \code{coin}
 (this is the semantics of \code{try} implemented in \cite{Lux99FLOPS}).
 However, if \code{coin} is evaluated inside the capsule of \code{try}
-(because it is not demanded before), then the nondeterminism of
+(because it is not demanded before), then the non-determinism of
 \code{coin} is encapsulated.
@@
 (e.g., in parallel implementations).
-Therefore, \cite{BrasselHanusHuch04JFLP} contains a proposal
+Hence, it is difficult to define a search operator as a pure function.
+
+Due to these considerations,
+\cite{BrasselHanusHuch04JFLP} contains a proposal
 for another primitive search operator:
@@
 \end{curry}
-\code{(Value v)} and \code{Fail} represents a single value
+\code{(Value v)} and \code{Fail} represent a single value
 or a failure (i.e., no value), respectively,
@@
 Since \code{getSearchTree} is an I/O action, its result (in particular,
-the order of solutions) depends on the current environment, e.g.,
+the order of subtrees) depends on the current environment, e.g.,
 time of evaluation.
@@
 values []             = []
-values (Fail:ts)    = values ts
-values (Value x:ts) = x : values ts
-values (Or _ _:ts)  = values ts
+values (Fail    : ts) = values ts
+values (Value x : ts) = x : values ts
+values (Or _ _  : ts) = values ts
 
 children []             = []
-children (Fail:ts)    = children ts
-children (Value _:ts) = children ts
-children (Or x y:ts)  = x:y:children ts
+children (Fail    : ts) = children ts
+children (Value _ : ts) = children ts
+children (Or x y  : ts) = x : y : children ts
 \end{curry}
@@
 is iterative-deepening search.
-Basically, it is depth-first search strategy with a depth-bound
+Basically, it is a depth-first search strategy with a depth-bound
 which is incremented in each iteration.
@@
 Since they are only supported by the Curry implementation KiCS2,
-we use only this system for our evaluation.
+we use this system for our evaluation.
 A general comparison of KiCS2 and other Curry implementations
@@
 (in Peano representation) by solving the equation \code{y+y=:=x},
-%``Embed'' searches for an embedding of a message
-%which is ``hidden'' in another message
-%(see \cite[Sect.~4.1]{Antoy10JSC}),
 ``Graph'' computes some path in a graph where the edges
@@
 that have been favored in the past due to limited memory requirements.
-Thus, it could be reasonable to make
-complete strategies the default in functional logic languages.
-This would have a good impact on teaching declarative programming
-to beginners.
-Moreover, decoupling non-deterministic programs
+Decoupling non-deterministic programs
 from their search strategy could lead to a more declarative
@@
 optimization, e.g., parallel search strategies.
+The explicit definition of search strategies
+has also been advocated for combining logic programs
+with different constraint solvers
+\cite{FrankHofstedtPepperReckmann06,Schulte97CP}.
+
+Our results indicate that it could be reasonable to make
+complete strategies the default in functional logic languages.
+This would have a good impact on teaching declarative programming
+to beginners.
 Furthermore, if efficiency and memory limitations are important,
-one can still use an efficient strategy like depth-first search
+one can still use an efficient strategy, like depth-first search,
 provided that it is able to find all solutions
@@
 to statically approximate situations where
-the use of incomplete strategies is sufficient.
+the use of theoretically incomplete strategies is sufficient.
 

### 2012-02-15 papers/atps2012/paper.tex
ATPS 2012: Final paper version

@@
 and a logic programming system tries to find
-a solution to a query w.r.t.\ the set of rules.
+a solution to a query w.r.t.\ this set of rules.
 In order to abstract from operational details,
@@
 last :: [a] -> a
-last xs | (ys ++ [z]) =:= xs
-        = z                    where ys,z free
+last xs | (ys ++ [z]) =:= xs = z
+  where ys, z free
 \end{curry}
@@
 \begin{curry}
-let y=coin in try (\x->x=:=y)
+let y = coin in try (\x->x =:= y)
 \end{curry}
@@
 in each level of the tree and the operation \code{children}
-extracts all direct ancestors of a level in order to recursively
+extracts all direct successors of a level in order to recursively
 collect their values.
@@
   collectLevel d (Or  x y) =
-    if d>0
+    if d > 0
     then concA (collectLevel (d-1) x) (collectLevel (d-1) y)
@@
 recomputation is not required. Instead, the
-already evaluated part of the search tree is kept in memory.
+already evaluated part of the search tree is kept in memory,
+sacrificing the better memory behaviour.
 Therefore, we also implemented the iterative deepening strategy
@@
                          inp <- getLine
-                         if inp=="yes" then printResults xs
+                         if inp == "yes" then printResults xs
                                          else done

### 2012-02-17 papers/wlp2011/paper.tex
WLP 2011: Extended paper, final version

@@
 
+\usepackage[T1]{fontenc} % To suppress font warnings
 \usepackage{url}
@@
 
-\begin{document}
 \pagestyle{plain} % remove for camera ready version
@@
 
-\title{Implementing Equational Constraints\\ in a Functional Language}
+% Metadata
+
+% texorpdfstring to avoid warning of hyperref package
+\title{\texorpdfstring
+  {Implementing Equational Constraints \\ in a Functional Language}
+  {Implementing Equational Constraints in a Functional Language}
+}
 
@@
 }
+
 \institute{
@@
 }
+
+\begin{document}
 
@@
 programming with non-strict and non-deterministic operations.  This
-logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
+logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}%
 \label{ctc-semantics},
@@
 \end{haskell}
+%
 The second rule transforms a non-deterministic argument

### 2012-04-02 docs/Set_Functions_and_Failures.tex
documentation for Set Functions added

\documentclass{article}

\usepackage{listings}

\title{On the combination of Set Functions and Failures}

\begin{document}

\maketitle

\section{Towards a sound semantics}

The example that led to the need to reconsider the semantics of failure w.r.t.
Set Functions was a more complex instance of the following:

\begin{lstlisting}
data Size = Small | Big

sizes :: Size
sizes = Small ? Big

getSmallest | isEmpty (set1 isSmallest aSize) = aSize
  where aSize = sizes

isSmallest size | size < sizes = size
\end{lstlisting}

This program is a straightforward implementation of the intuitive definition
that a value from a non-deterministic expression is the smallest if there
is no smaller one. And with our implementation of Set Function it works
just fine, there is only the one result \lstinline{Small}.

No let's change the definition of \lstinline{getSmallest}:

\begin{lstlisting}
getSmallest 
 | isEmpty (set1 isSmallest (aSize =:= sizes &> aSize))
 = aSize
 where aSize = size
\end{lstlisting}

This new definition is more complex and somewhat contrived but
it still should yield the same results. The single difference is
that there is a constraint that states that there should be a
size in the results of \lstinline{sizes} that can be unified
with \lstinline{aSize}. This, of course, is trivially satisfied.

Still, with our current implementation in addition to the expected
result \lstinline{Small} there are two other ones, a second
\lstinline{Small} and \lstinline{Big}. This is obviously not intended
but can easily be explained. The new constraint introduces two
failing computations for the unification of \lstinline{Small} with
\lstinline{Big} and vice versa to the argument of 
\lstinline{isSmallest}. Since our implementation does not
capsule the non-determinism introduced by the argument and regards
a failing computation as the absence of any results, the result
of the capsuled computation is non-deterministically four sets:
$\{\texttt{Small}\},\{\},\{\}$ and $\{\}$. The first set contains
\lstinline{Small} as the result for a smaller value than \lstinline{Big},
the second one is empty because there is no smaller value than \lstinline{Small}.
The other two empty sets stem from the failing computations introduced by
the constraint.

In order to avoid the incorrect behavior we need to mark the failing computations
that are introduced by the argument to a set-valued function. In the example
above a failing computation in the argument should yield a failing computation
in the result of the set-valued function rather than an empty set of results.
This marking can be done with the same mechanism that is used to cover the
non-determinism of the argument.

The next example will show that failing computations in the argument should
not always result in a failure of the set-valued function. Consider the
following simple program:

\begin{lstlisting}
trueOrSomething x = True ? x
main = set1 trueOrSomething failed
\end{lstlisting}

Now the non-determinism is introduced not by the argument to the set-valued function
but by the function itself. Therefore, it is capsuled. That means the resulting set
should contain both \lstinline{True} and a failing computation. Since failing
computations are not considered values, there are two simple options:
\begin{enumerate}
\item A set that contains a failing computation is a failing computation
\item A set that contains failing computation is the set without the
      failing computations.
\end{enumerate}

Our first example motivates the first option. However, this option is ruled out
by the semantics of Set Functions as stated in the PPDP-Paper by Sergio Antoy
and Michael Hanus. There is specified that every result that can be obtained
from the application of the original function to some arguments has to belong
to the result of the set-valued version of the function applied to the same
arguments. In our example \lstinline{True} is a result of 
\lstinline{trueOrSomething failed}, therefore \lstinline{True} has to be an
element of a result of \lstinline{set1 trueOrSomething failed}.

The second option is a valid choice for the second example, but, as we have
already seen, it will lead to unintended results is the modified version
of the first example.

A way out of this dilemma is to combine the two options:

\textit{A set that contains only failing computations that stem from an argument
        is a failing computation it is the set without any failing computations
        otherwise}

This implies that the following program has to return the empty set rather than
a failing computation.

\begin{lstlisting}
failOrSomething x = failed ? x
main = set1 failOrSomething failed
\end{lstlisting}

The non-determinism is encapsulated and the resulting set contains both a failing
computation that stem from the argument and one that is introduced by the set-valued
function.

Note that the prototypical implementation of Set Functions in PAKCS would yield no 
result for the latter two examples since the argument is strictly evaluated and
only deterministic values are passed to the set-valued function.

Following this semantics we can make an interesting observation if we consider
the following call to \lstinline{trueOrSomething}:

\begin{lstlisting}
main = set1 trueOrSomething (failed ? failed)
\end{lstlisting}

This call will yield the set $\{\texttt{True}\}$ twice. And for additional
non-deterministic occurrences of failing computations in the argument there
would be an additional result.

So, failing computations in arguments of Set Functions can affect the
multiplicity of the results.

So far we had to distinguish failing computations in the arguments to
set-valued functions and those that were introduced by such functions.
Now we want to investigate failing computations that stem from different
levels of set-valued functions.

We contrive an adequate example:

\begin{lstlisting}
f x = failed ? x
g x = set1 f (failed ? x)
main = set1 g failed
\end{lstlisting}

Now, what is the intended result of a call to \lstinline{main}? The non-determinism
introduced by \lstinline{f} is capsuled by the call of the set-valued version of the
function \lstinline{f} in \lstinline{g}. So the result of this call to \lstinline{f}$_{S}$
is $\{\}?\{\}$ since \lstinline{f} introduces an failure itself. The remaining non-determinism
is the one introduced by \lstinline{g}. Since \lstinline{main} calls the set-valued version
of \lstinline{g} the result of this call is $\{\{\},\{\}\}$.

Now let's change the program:

\begin{lstlisting}
f x = x
g x = set1 f (failed ? x)
main = set1 g failed
\end{lstlisting}

Here \lstinline{f} is just an other name for \lstinline{id}. This time the
call to \lstinline{f}$_S$ by g will non-deterministically yield two
failing computations, since the non-determinism is covered an all
failing computations stem from an argument to \lstinline{f}$_S$.

In the call to \lstinline{g}$_S$ by \lstinline{main} the non-determinism is
capsuled and the resulting set contains the two failing computations. One of
these failures was introduced by \lstinline{g}, so, the result is the
empty set.

The last example regards the non-deterministic combination of failing
computations. Consider the following program:

\begin{lstlisting}
g x = set2 (?) x failed
main = set1 g failed
\end{lstlisting}

This time the non-determinism introduced by \lstinline{(?)} is capsuled and
both failures stem from arguments to \lstinline{(?)}. Therefore, the
result of \lstinline{(?)}$_S$ is a failing computation. Now we need to
determine the result of the call to \lstinline{g}$_S$ by \lstinline{main}.
The failure of the call to \lstinline{(?)}$_S$ is a result of the failure
introduced by \lstinline{g}, therefore, this call to \lstinline{g}$_S$
should yield the empty set. To be able to implement this behavior, a set
valued function that fails because of failing computations that stem
from the arguments needs to determine which of the failures stems from the
deepest nested call and has to associate this information with the resulting
failure.

\end{document}
% LocalWords:  failOrSomething trueOrSomething PPDP isSmallest aSize LocalWords
% LocalWords:  getSmallest isEmpty

### 2012-04-03 docs/Set_Functions_and_Failures.tex
Minor corrections to the Set Functions semantics paper

@@
 \usepackage{listings}
+\lstset{basicstyle=\small\ttfamily}
 
 \title{On the combination of Set Functions and Failures}
+\author{Fabian Reck}
 
@@
 
-sizes :: Size
-sizes = Small ? Big
+aSize :: Size
+aSize = Small ? Big
 
-getSmallest | isEmpty (set1 isSmallest aSize) = aSize
-  where aSize = sizes
+getSmallest | isEmpty (set1 smallerThan size) = size
+  where size = aSize
 
-isSmallest size | size < sizes = size
+smallerThan size | anotherSize < size = anotherSize
+  where anotherSize = aSize
 \end{lstlisting}
@@
 that a value from a non-deterministic expression is the smallest if there
-is no smaller one. And with our implementation of Set Function it works
+is no smaller one. And with our implementation of Set Functions it works
 just fine, there is only the one result \lstinline{Small}.
 
-No let's change the definition of \lstinline{getSmallest}:
+Now let's change the definition of \lstinline{getSmallest}:
 
@@
 getSmallest 
- | isEmpty (set1 isSmallest (aSize =:= sizes &> aSize))
- = aSize
- where aSize = size
+ | isEmpty (set1 smallerThan (size =:= aSize &> size))
+ = size
+ where size = aSize
 \end{lstlisting}
@@
 
-\textit{A set that contains only failing computations that stem from an argument
-        is a failing computation it is the set without any failing computations
-        otherwise}
+% TODO: Definition?
+\begin{quote}
+A set that contains only failing computations that stem from an argument
+is a failing computation.
+Otherwise, it is the set without any failing computations.
+\end{quote}
 
-This implies that the following program has to return the empty set rather than
-a failing computation.
+This implies that the following program has to return the empty set
+rather than a failing computation.
 

### 2012-04-24 talks/wlp2011/slides.tex
Slides of WLP2011 added

\documentclass[10pt]{beamer}
%\documentclass[draft,10pt]{beamer} % for faster formatting
%\documentclass[handout,10pt]{beamer} % to produce a handout

\input{mhbeamer}

\usepackage{listings}
\lstset{mathescape=true,
        aboveskip=0.5ex,
        belowskip=0.5ex,
        showstringspaces=false, % no special string space
        basewidth=0.55em,
        basicstyle=\ttfamily,
      literate={=}{{=}}1
               {\\}{{$\lambda$}}1
               {\\\\}{{\char`\\\char`\\}}1
               {->}{{$\rightarrow{}\!\!\!$}}3
               {<-}{{$\leftarrow{}$}}2
               {\ .}{{$\circ$}}2 
               {\ .\ }{{$\circ$}}2
               {>>}{{>>}}2 
               {>>=}{{>>=}}4
               {|}{{$\mid$}}1
      }
\lstnewenvironment{curry}{}{}
\newcommand{\listline}{\vrule width0pt depth1.75ex}

\newcommand{\suc}{\mathit{success}}  % success
\newcommand{\Suc}{\mathtt{Success}}  % success type
\newcommand{\pre}[1]{#1^{pre}}  % precondition for a function
\newcommand{\post}[1]{#1^{post}} % postcondition for a function
\newcommand{\funset}{\ensuremath{_{\cal S}}}

\date{WLP 2011}
\author[Michael Hanus]{Michael Hanus\\[2ex]
{\small Joint work with Bernd Bra\ss{}el, Bj\"orn Peem\"oller and Fabian Reck}}
\institute[CAU Kiel]{Christian-Albrechts-University of Kiel\\
  Programming Languages and Compiler Construction}

\begin{document}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}

\title[Implementing Equational Constraints in a Functional Language]
      {Implementing Equational Constraints\\
       in a Functional Language}
\frame{\titlepage}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Functional Logic Languages}

\begin{block}{Goal: combine best of declarative paradigms in a single model}
\begin{itemize}%\expandlist{1ex}
\item {\blue efficient execution} principles of functional languages\\
(determinism, laziness)
\item {\blue flexibility} of logic languages\\
(computation with partial information, built-in search)
\item {\blue application-domains} of constraint languages\\
(constraint solvers for specific domains)
\item avoid non-declarative features of Prolog\\
(arithmetic, cut, I/O, side-effects)
\end{itemize}
\end{block}
\vfill
\pause

\begin{block}{Curry {\lgrey [POPL'97,\ldots]} $\leadsto$ {\tt http://www.curry-language.org/}}
\begin{itemize}%\expandlist{1ex}
\item declarative multi-paradigm language\\
 (higher-order concurrent functional logic language)
\item extension of Haskell (non-strict functional language)
\item SWE advantage: better (high-level) APIs for various application domains\\
(GUI programming, web programming, database programming,\ldots)
\end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Curry Programs}

\begin{exampleblock}{Datatypes (values): enumerate all constructors}
\begin{prog}
 data Bool       = {\blue True}    | {\blue{}False}
 data List a     = {\blue []}      | a {\blue :} List a   {\red-- [a]}
\end{prog}
\end{exampleblock}
\vfill
\pause

\begin{block}{Program rules:~ $f~t_1\ldots t_n~|~c~=~r$}
\begin{tabular}{r@{~~:~~}l@{\qquad}r@{~~:~~}l}
$f$ & function name & $t_1\ldots t_n$ & data terms\\
$c$ & condition of type \code{Success} (optional) & $r$ & expression
\end{tabular}
\end{block}
\vfill

\begin{exampleblock}{Example \hspace{4cm}\only<3>{(with \alert{equational constraint}~ \texttt{=:=})}\only<4>{(with \alert{functional pattern})}}
\begin{curry}
(++) :: [a] -> [a] -> [a]    $\only<3-4>{\code{last :: [a] -> a}}$
[]     ++ ys = ys              $\only<3>{\code{last xs | \alert{ys ++ [x] =:= xs}}}$$\only<4>{\code{last \alert{(ys ++ [x])} = x}}$
(x:xs) ++ ys = x : xs ++ ys             $\only<3>{\code{= x  where ys,x free}}$
\end{curry}
\end{exampleblock}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Non-deterministic Operations}

\begin{exampleblock}{Choice operation}
\vspace{-1ex}
\begin{curry}
 x ? _  =  x               aBool = True ? False
 _ ? y  =  y$\listline\pause$
not True  = False          xor True  x = not x
not False = True           xor False x = x$\listline$
xorSelf x = xor x x
\end{curry}
\end{exampleblock}
\pause

\begin{exampleblock}{Challenge: combining laziness and non-determinism}
\begin{curry}
 xorSelf aBool$\pause$  $\to~~$xor aBool aBool$\pause$    $\to~$  xor True aBool$\pause$
                $\to~~$xor True False$\pause$     $\to~$  not False$\pause$       $\to~$ True
\end{curry}
\end{exampleblock}
\pause

\begin{block}{Avoid unintended results: ``call-time choice''}
\begin{itemize}
\item semantics: arguments $\approx$ values (not non-deterministic operations!)
\item execution: keep laziness via \emph{sharing}
\item \alert{share/identify choices coming from same source}
\end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Implementing Curry}

\begin{block}{Abstract machines}
\begin{itemize}
\item C (MCC, M\"unster Curry Compiler)
\item Java (FLVM, Portland)
\end{itemize}
\end{block}
\pause

\begin{block}{Compiling to Prolog}
\begin{itemize}
\item PAKCS (Portland Aachen Kiel Curry System), TOY (Madrid)
\item reuse non-determinism (backtracking), logic variables, constraint solvers
\item implement laziness, higher-order
\end{itemize}
\end{block}
\pause

\begin{block}{Compiling to Haskell: {\bf KiCS2} {\lgrey[WFLP'11]}}
\begin{itemize}
\item reuse laziness, higher-order
\item implement non-determinism, logic variables
\item advantages: \alert{search strategy not fixed}\\
      (backtracking, breadth-first, iterative deepening, parallel,$\ldots$)
\end{itemize}
\end{block}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Compiling Curry to Haskell: KiCS2}

\begin{block}{Principles of KiCS2}
\begin{itemize}\expandlist{2ex}
\item represent search space as data ($\approx$ search tree)
\item defined operation: generate \emph{all} its values
\item search strategy $\approx$ search tree traversal
\item Haskell's laziness: no problem with infinite trees
\end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Representing Non-deterministic Values}

\begin{block}{Every operation could have non-deterministic results!}
$\leadsto$ represent a choice between values as data:
\begin{curry}
data Bool = False | True $\pause$| Choice $\only<5->{\texttt{\alert{ID}}}$ Bool Bool | $~$Fail$\listline\pause$
aBool = True ? False $\leadsto$  aBool $\only<5->{\texttt{\alert{s}}}$ =$~$Choice $\only<5->{\texttt{\alert{(thisID s)}}}$ True False
\end{curry}
\pause
Missing: \alert{identify choices}\pause~~$\leadsto$~~add unique(!) identifier to each choice
\end{block}
\pause
\vfill
\begin{block}{Creation of choice identifiers}
\begin{itemize}
\item avoid side effects (to support optimization, parallel evaluation)
\item supply each operation with \alert{set of choice ids} (\code{IDSupply})
\item different operations get distinct subsets of choice ids
\end{itemize}
\end{block}
\vfill\pause

\begin{exampleblock}{Interface of {\tt IDSupply}}
\vspace{-1ex}
\begin{curry}
initSupply  :: IO IDSupply            $\only<7>{\alert{\mbox{\tt{-- create}}}}\only<8>{\alert{\mbox{\tt{initSupply = return 1}}}}$
thisID      :: IDSupply -> ID        $\only<7>{\alert{\mbox{\tt{-- get one id}}}}\only<8>{\alert{\mbox{\tt{thisID~~~~~ n = n}}}}$
leftSupply  :: IDSupply -> IDSupply  $\only<7>{\alert{\mbox{\tt{-- split into}}}}\only<8>{\alert{\mbox{\tt{leftSupply~ n = 2*n}}}}$
rightSupply :: IDSupply -> IDSupply  $\only<7>{\alert{\mbox{\tt{-- disjoint subsets}}}}\only<8>{\alert{\mbox{\tt{rightSupply n = 2*n+1}}}}$
\end{curry}
\end{exampleblock}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Compilation Scheme}

\begin{block}{Translate Curry programs into functional programs}
\begin{itemize}
\item every operation: additional \code{IDSupply} argument
\item supply each \code{Choice} with unique identifier
\item pattern matching: extended with cases for \code{Choice} and \code{Fail}
\end{itemize}
\end{block}
\pause
\vfill

\begin{exampleblock}{}
\begin{curry}
aBool :: Bool         $\leadsto$   aBool :: IDSupply -> Bool
aBool = True ? False  $\leadsto$   aBool s =$~$Choice (thisID s) True False
$\pause$
not :: Bool $\only<4->{\alert{\to~\mbox{\tt{IDSupply}}}}$ -> Bool
not True  $\only<4->{\alert{\mbox{\tt{s}}}}$ =$~$False
not False $\only<4->{\alert{\mbox{\tt{s}}}}$ =$~$True$\pause$
$\alert{\mbox{\tt{not (Choice i x1 x2) s = Choice i (not x1) (not x2)}}}$
$\alert{\mbox{\tt{not Fail s = Fail}}}\pause$

main :: $\only<6->{\alert{\mbox{\tt{IDSupply}}~\to}}$ Bool
main $\only<6->{\alert{\mbox{\tt{s}}}}$ =$~$xorSelf $\only<6->{\alert{\mbox{\tt{(}}}}$aBool $\only<6->{\alert{\mbox{\tt{(leftSupply s)) (rightSupply s)}}}}$
\end{curry}
\end{exampleblock}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Choices with Identifiers}

\begin{block}{Principle: each choice gets its own id}
\begin{curry}
xorSelf aBool $\pause\leadsto$
Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{curry}
Interpretation: essentially one choice $\leadsto$ two solutions (\code{False}, \code{False})
\pause
\bigskip
\begin{curry}
xor aBool aBool $\pause\leadsto$
Choice 1 (Choice 2 False True) (Choice 2 True False)
\end{curry}
\medskip
Two different choices $\leadsto$ four solutions (\code{False}, \code{True}, \code{True}, \code{False})
\end{block}
\pause
\vfill

\begin{block}{Overall computation strategy}
\begin{itemize}\expandlist{0.5ex}
\item operations compute all non-deterministic results (as \code{Choice} structure)
\item top-level selects different values and prints them
\item top-level responsible for search strategy
\end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Top Level: Print Results}

\begin{block}{Extract values from computed choice structure}
\begin{itemize}\expandlist{0.5ex}
\item make all possible selections (left, right for each \code{Choice})
\item make consistent selections (same choice for identified \code{Choice})
\end{itemize}
\pause
\begin{curry}
   xorSelf aBool
   $\leadsto$ Choice 1 (Choice 1 False True) (Choice 1 True False)
   $\pause\leadsto$   False   $\vee$   False
\end{curry}
\pause
\begin{itemize}\expandlist{0.5ex}
\item manage global structure (e.g., finite map) to store current selections
\item search strategy $\approx$ strategy to visit tree of choices
\end{itemize}
\end{block}
\vfill\pause

\begin{block}{Management of choice selections}
\begin{curry}
  data Decision = NoDecision | ChooseLeft | ChooseRight$\listline$
  lookupDecision :: ID -> IO Decision          $\alert{\mbox{\tt{-- retrieve}}}$
  setDecision    :: ID -> Decision -> IO ()   $\alert{\mbox{\tt{-- modify}}}$
\end{curry}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Depth-first Search}
\vspace{-1ex}
\begin{exampleblock}{Print all values of a choice structure in depth-first order}
\vspace{-1ex}
\begin{curry}
  printValsDFS :: a -> IO ()$\listline$
  printValsDFS Fail             = return ()$\listline$
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision    = do newDecision ChooseLeft  x1
                              newDecision ChooseRight x2

    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision$\listline$
  printValsDFS v = print v
\end{curry}
\vfill\pause

Main call: \code{initSupply >>= \bs{}s -> printValsDFS (main s)}\\
Result:~~~~ \code{False}\\
\hspace{1.62cm}\code{False}
\end{exampleblock}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Benchmarks {\lgrey [WFLP'11]}}

\begin{block}{Deterministic Programs}
\begin{center}
\begin{tabular}{|@{~~}l@{~~}|l@{~~}|*{4}{@{~~}r@{~~}|}}
\hline
System  & Target   & ReverseUser & Reverse &  Tak  & TakPeano \\\hline
KiCS2   & Haskell  &      0.12   &   0.12  &  0.21 &    0.79  \\[0.5ex]
PAKCS   & SICStus-Prolog    &      2.05   &   1.88  & 39.80 &   62.43  \\[0.5ex]
MCC     & VM/C     &      0.43   &   0.47  &  1.21 &    5.49  \\[0.5ex]
\hline
\end{tabular}
\end{center}
\end{block}
\vfill

\begin{block}{Non-deterministic Programs}
\begin{center}
\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
\hline
System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
KiCS2   &    2.83  &        3.68   &  0.14 &   0.49 \\[0.5ex]
PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\[0.5ex]
MCC     &    1.46  &        5.74   &  0.09 &   0.57 \\[0.5ex]
\hline
\end{tabular}
\end{center}
\end{block}
\vfill
\begin{block}{}
  \begin{itemize}
  \item faster than Prolog-based implementations (PAKCS)
  \item competitive with VM/C-based implementation (MCC)
  \end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{frame}[fragile]
\frametitle{Logic Variables}
\vspace{-1ex}
\begin{block}{Kernel language supports choices but no logic variables}
\begin{itemize}
\item logic variables are replaced by non-deterministic generators
\item generator: evaluates to all possible values
\end{itemize}
\end{block}
\vfill
\pause

\begin{exampleblock}{}\vspace{-1ex}
\begin{curry}
-- Generator for Booleans
aBool :: Bool
aBool = True ? False$\listline$
-- Generator for list of Booleans
aBoolList :: [Bool]
aBoolList = [] ? (aBool : aBoolList)$\listline$
main = not x where x free   $\leadsto~~$  main =$~$not aBool
\end{curry}
\end{exampleblock}
\vfill
\pause

\begin{block}{}
\citebib{Antoy/H. ICLP'06}:\\[1ex]
narrowing (resolution) on logic variables \\
$~~~~~~~~~~~~~~~~~~~~~~~\Updownarrow$\\
non-deterministic rewriting with generators
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Logic Variables and Unification}

\begin{block}{Equational Constraints}
\begin{itemize}
\item generators instead of logic variables sufficient in principle
\item could be inefficient with constraints on partially known values
\end{itemize}
\end{block}
\vfill
\pause

\begin{exampleblock}{}
\code{xs},\code{ys} logic variables of type \code{[Bool]}:\\
\pause
\code{xs =:= [True]} $~\leadsto$ search space (instead of deterministic binding)\\
\pause
\code{xs =:= ys} \hspace{5.7ex}$\leadsto$ infinitely many solutions!
\end{exampleblock}
\vfill
\pause

\begin{block}{Unification}
\begin{itemize}
\item better: bind logic variables instead of enumerating identical values
\item necessary extensions:
\begin{itemize}
\item mark unevaluated generators ($\approx$ unbound variables)
\item add binding constraints to values
\end{itemize}
\end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Implementing Equational Constraints}

\begin{block}{Representation of logic variables}
\begin{curry}
data ID = ChoiceID Integer | $\alert{\mbox{\tt{FreeID}}}$ Integer$\listline$
aBool s = Choice ($\alert{\mbox{\tt{FreeID}}}$ (thisID s)) True False
\end{curry}
\end{block}
\vfill
\pause

\begin{block}{Add binding constraints to values}
\begin{curry}
data Constraint = ID :=: Decision$\listline$
data Bool   = $\ldots$ |$~$Guard [Constraint] Bool
data List a = $\ldots$ |$~$Guard [Constraint] (List a)
\end{curry}
\pause
\medskip
Implementation of equational constraints:
\begin{curry}
Choice (FreeID i) _ _ =:= True  = Guard$\,$[i:=:ChooseLeft ]$\,$Success
Choice (FreeID i) _ _ =:= False = Guard$\,$[i:=:ChooseRight]$\,$Success
$\ldots$
\end{curry}
\medskip
Binding constraints are processed/checked by top-level printing
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Functional Patterns {\lgrey\small[Antoy/H. LOPSTR'05]}}

\begin{block}{Useful extension: allow defined functions in patterns}
\begin{curry}
last (xs++[x]) = x
\end{curry}
\end{block}
\vfill
\pause

\begin{block}{Semantics: abbreviates all values of functional pattern}
\begin{curry}
last [x] = x              -- xs$\mapsto$[]
last [x1,x] = x           -- xs$\mapsto$[x1]
last [x1,x2,x] = x        -- xs$\mapsto$[x1,x2]
$\ldots$
\end{curry}
\end{block}
\vfill
\pause

\begin{block}{Implementation: lazy unification ``{\tt{=:<=}}''}
\begin{curry}
last ys | (xs++[x]) =:<= ys  = x   where xs, x free
\end{curry}
\medskip
Difference to ``{\tt{=:=}}'': bind variable to \emph{unevaluated} expression\\
\code{last [failed,True] $\leadsto$ True}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}[fragile]
\frametitle{Implementing Lazy Unification}

\begin{block}{Add lazy binding constraints as decisions}
\begin{curry}
data Decision = $\ldots$ |$~$LazyBind [Constraint]
\end{curry}
\end{block}
\vfill
\pause

\begin{block}{Lazy unification with variables: add lazy binding constraints}
\begin{curry}
Choice (FreeID i) _ _ =:<= x
    = Guard [i :=: LazyBind (lazyBind i x)] Success$\listline$
lazyBind :: ID -> a -> [Constraint]
lazyBind i True  = [i :=: ChooseLeft]
lazyBind i False = [i :=: ChooseRight]
\end{curry}
\medskip
$\leadsto$ these constraints are evaluated when value is demanded
\end{block}
\vfill
\pause

\begin{exampleblock}{}
\begin{curry}
xs++[x] =:<= [failed,True]
$\leadsto$ Guard [ 4 :=: LazyBind (lazyBind 4 failed),
           3 :=: LazyBind (lazyBind 3 True)] Success
\end{curry}
\end{exampleblock}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Benchmarks}

\begin{block}{Equational constraints}
\begin{center}
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS  & MCC  \\
\hline
\code{last (map (inc 0) [1..10000])}   & 0.05  & 0.40   &  0.01\\
\code{simplify}                         & 6.77  & 0.15   & 0.00 \\
\code{varInExp}                         & 0.24  & 0.89   & 0.07 \\
\code{fromPeano (half (toPeano 10000))} & 5.95  & 108.88 & 3.22 \\
\code{palindrome}                       & 14.05 & 32.56  & 1.07 \\
\code{horseman}                         & 3.31  & 8.70   & 0.42 \\
\code{grep}                             & 0.10  & 2.88   & 0.14 \\
\hline
\end{tabular}
\end{center}
\end{block}
\vfill
\begin{block}{}
  \begin{itemize}
  \item faster than Prolog-based implementations (PAKCS)
  \item primitive VM/C-based implementation (MCC) performs best
  \item with some exceptions ($\leadsto$ further investigations required)
  \end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Benchmarks}

\begin{block}{Functional patterns}
\begin{center}
\begin{tabular}{|l|r|r|}
\hline
Expression                              & KiCS2 & PAKCS    \\
\hline
\code{last (map (inc 0) [1..10000])}   &  0.01 & 0.33     \\
\code{simplify}                         & 7.07  & 0.27     \\
\code{varInExp}                         &  0.21  &  1.87    \\
\code{fromPeano (half (toPeano 10000))} & 11.19 & $\infty$ \\
\code{palindrome}                       & 20.26 & $\infty$ \\
\hline
\end{tabular}
\end{center}
\end{block}
\vfill
\begin{block}{}
  \begin{itemize}
  \item currently only supported by PAKCS and KiCS2
  \item in many cases bettern than Prolog-based implementation
  \end{itemize}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Benchmarks}

\begin{block}{Comparing different representations for equations in KiCS2}
\begin{center}
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..10000])}   &  2.91   &    0.05   &    0.01   \\
\code{simplify}                         & 10.30      & 6.77      & 7.07        \\
\code{varInExp}                         & 2.34    & 0.24      & 0.21        \\
\code{fromPeano (half (toPeano 10000))} & 26.67     & 5.95      & 11.19       \\
\code{palindrome}                       & 30.86     & 14.05     & 20.26       \\
\code{horseman}                         & 3.24      & 3.31      & n/a         \\
\code{grep}                             & 1.06      & 0.10      & n/a         \\
\hline
\end{tabular}
\end{center}
\end{block}
\vfill

\begin{block}{}
\begin{description}
\item[\code{==}~:] Boolean equality without variable bindings
\item[\code{=:=}~:] equational constraint with variable bindings
\item[\code{=:<=}~:] lazy unification with unevaluated variable bindings
\end{description}
\end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{frame}
\frametitle{Conclusions}
\vspace{-1ex}
\begin{block}{}
\begin{itemize}
\item novel implementation techniques for equational constraints
\item add bindings constraints to computed values
\item process them when extracting values from search space
\item only new constructors added, no overhead for existing implementation
\item more efficient execution if features are used
\end{itemize}
\end{block}
\vfill
\begin{block}{Advantages of KiCS2}
\begin{itemize}
\item highly efficient for deterministic parts, competitive for non-det. programs
\item flexible search strategies
\item many opportunities for optimizations (determinism analysis)
\item exploit future improvements of Haskell implementations
\end{itemize}
\end{block}
\vfill
\begin{block}{Future Work}
\begin{itemize}
\item Add further constraint structures
\end{itemize}
\end{block}
\end{frame}

\end{document}

% LocalWords:  non-deterministic

### 2012-04-27 docs/src/manual.tex
SetFunctions documentation improved
Version update

--- 2011-12-16 docs/src/manual.tex
@@
 {\huge User Manual}\\[4ex]
-Version of December 16, 2011\\[6ex]
+Version of April 25, 2012\\[6ex]
 \Large
@@
 \input{lib/ReadShowTerm}
+\input{lib/SetFunctions}
 \input{lib/SearchTree}

### 2012-05-05 docs/src/manual.tex
Bug fix in Makefile

@@
 \begin{center}\vspace{10ex}
-{\Huge\bf KiCS2 0.1}\\[4ex]
+{\Huge\bf KiCS2 0.2}\\[4ex]
 {\huge The Kiel Curry System (Version 2)}\\[7ex]

### 2012-05-16 runtime/ConstStore.hs
force-recomp removed from runtime libs, REPL controls this flag if ghc options changed

@@ -1,5 +1,4 @@
 {-# LANGUAGE CPP, ExistentialQuantification #-}
-{-# OPTIONS_GHC -fforce-recomp #-}
 
 module ConstStore where
 

### 2012-05-16 runtime/Debug.hs

@@ -1,5 +1,4 @@
 {-# LANGUAGE CPP #-}
-{-# OPTIONS_GHC -fforce-recomp #-}
 module Debug  where
 
 trace :: String -> IO ()

### 2012-05-16 runtime/Types.hs

@@ -4,7 +4,7 @@
 -- ---------------------------------------------------------------------------
 {-# LANGUAGE CPP #-}
 {-# LANGUAGE MultiParamTypeClasses, Rank2Types #-}
-{-# OPTIONS_GHC -fno-warn-orphans -fforce-recomp #-}
+{-# OPTIONS_GHC -fno-warn-orphans #-}
 
 module Types
   ( module ConstStore

### 2012-05-16 docs/src/manual.tex
Distinction between local and global installation (default) added

@@
 {\huge User Manual}\\[4ex]
-Version of April 25, 2012\\[6ex]
+Version of May 16, 2012\\[6ex]
 \Large
@@
 
-\subsection{General Use}
-\label{sec-general}
+\subsection{Installation}
+\label{sec-install}
 
@@
 and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems
-(see the file \code{INSTALL.html} in the KiCS2 directory
-for a description of the necessary software to install KiCS2).
+or Mac OS X systems.
+
+Installation instructions for KiCS2 can be found in
+the file \code{INSTALL.txt} stored in the KiCS2 installation directory.
+Note that there are two possibilities to install KiCS2:
+\begin{description}
+\item[Global installation:]\index{global installation}\index{installation!global}
+KiCS2 is installed in some global system directory where users
+have no write permission. In this case, some options for experimenting
+with KiCS2 (like \code{supply} or \code{ghc}, see below)
+are not available (since they require the recompilation of parts of
+the installed system).
+\item[Local installation:]\index{local installation}\index{installation!local}
+KiCS2 is installed in some local user directory where the user
+has write permission. The requires the setting of the option
+\begin{lstlisting}
+GLOBALINSTALL=no
+\end{lstlisting}
+in the \code{Makefile} of the KiCS2 installation.
+In this case, all options of KiCS2 are available.
+\end{description}
+In the following, $kics2home$ denotes the installation directory
+of the KiCS2 installation.
+
+\subsection{General Use}
+\label{sec-general}
 
 All executables required to use the different components
-of KiCS2 are stored in the directory \code{$kics2home$/bin}
-(where $kics2home$ is the installation directory of the complete
-KiCS2 installation). You should add this directory
+of KiCS2 are stored in the directory \code{$kics2home$/bin}.
+You should add this directory
 to your path (e.g., by the \code{bash} command
@@
 
-\item[\fbox{\code{choices}}]\pindex{choices}
+\item[\fbox{\code{choices $n$}}]\pindex{choices}
 Show the internal choice structure
@@
 evaluations used in KiCS2.
+If the optional argument $n$ is provided,
+the tree is shown up to depth $n$.
 
 \item[\fbox{\code{supply $i$}}]\pindex{supply}
+(not available in global installations, see Section~\ref{sec-install})
 Use implementation $i$ as the identifier supply for choice structures
@@
 \item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
+(not available in global installations, see Section~\ref{sec-install})
 Define additional options passed to the Glasgow Haskell Compiler
@@
 where $kics2home$ is the installation directory of KiCS2;
-see Section~\ref{sec-general}).
+see Section~\ref{sec-install}).
 In our example, \ccode{testList.curry} is the program containing the
@@
 where $kics2home$ is the installation directory of KiCS2;
-see Section~\ref{sec-general}).
+see Section~\ref{sec-install}).
 If \code{MyData} is the name of the ERD, the Curry program file

### 2012-05-30 docs/src/manual.tex
Some improvement for "make"

@@
 {\huge User Manual}\\[4ex]
-Version of May 16, 2012\\[6ex]
+Version of May 30, 2012\\[6ex]
 \Large
@@
 KiCS2 is installed in some local user directory where the user
-has write permission. The requires the setting of the option
+has write permission and the option \code{GLOBALINSTALL}
+in the \code{Makefile} of the KiCS2 installation is set as follows:
 \begin{lstlisting}
@@
 \end{lstlisting}
-in the \code{Makefile} of the KiCS2 installation.
 In this case, all options of KiCS2 are available.

### 2012-05-31 docs/src/manual.tex
Set idsupply ioref as default in REPL

@@
 This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc} (default):]\pindex{ghc}
+\item[\code{ghc}:]\pindex{ghc}
 Use a more sophisticated implementation of choice
@@
 like encapsulated search.
-\item[\code{ioref}:]\pindex{ioref}
+\item[\code{ioref} (default):]\pindex{ioref}
 Use a mixture of \code{ghc} and \code{pureio}.

### 2012-06-01 src/makecurrycgi.sh
Shell scripts are generated with appropriate KICS2HOME values

@@ -1,6 +1,10 @@
 #!/bin/sh
 # Compile a Curry program (using the HTML library) into a cgi script
 
+# The installation directory of KiCS2
+KICS2HOME=`echo KICS2HOME must be defined here!`
+export KICS2HOME
+
 # Standard suffix that will be added to the main script:
 CGISUFFIX="_CGIMAIN_$$"
 # Name of the main function in the main script (should not be in conflict
@@ -89,11 +93,6 @@ if [ $# != 1 -a $# != 3 ] ; then
   echo "             specifications) when generating JavaScript support code"
   exit 1
 fi
-
-# Compute the main installation directory of KiCS2:
-PATHNAME=`(cd \`dirname $0\` > /dev/null ; pwd)`
-KICS2HOME=`expr $PATHNAME : '\(.*\)/bin'`
-export KICS2HOME
 
 # Definitions for WUI/JavaScript generation:
 WUIJS_PREPROCESSOR=$KICS2HOME/tools/curry2js/Curry2JS

### 2012-06-01 docs/src/manual.tex
Version update

@@
 {\huge User Manual}\\[4ex]
-Version of May 30, 2012\\[6ex]
+Version of June 1, 2012\\[6ex]
 \Large

### 2012-06-01 runtime/Basics.hs
Extended FailInfo in runtime

@@ -2,6 +2,7 @@
 
 module Basics
   ( module Basics
+  , module FailInfo
   , module PrimTypes
   , module Search
   , module Types
@@ -11,6 +12,7 @@ import Data.Char(ord)
 import GHC.Exts (Int#, Char#, chr#)
 
 import Debug (internalError)
+import FailInfo
 import PrimTypes
 import Search
 import Types

### 2012-06-01 runtime/FailInfo.hs

@@ -1,0 +1,32 @@
+-- ---------------------------------------------------------------------------
+-- This module contains the types and functions to collect information
+-- about the origin of failures
+-- ---------------------------------------------------------------------------
+module FailInfo
+  ( FailInfo (..)
+  , defFailInfo
+  , consFail
+  , traceFail
+  ) where
+
+import Data.List
+
+data FailInfo = FailInfo
+  { failTrace :: [String] }
+  deriving (Eq, Show)
+
+defFailInfo :: FailInfo
+defFailInfo = FailInfo []
+
+addTrace :: String -> FailInfo -> FailInfo
+addTrace t fi = fi { failTrace = t : failTrace fi }
+
+consFail :: String -> [String] -> FailInfo
+consFail fun args = addTrace (fmtCall fun args) defFailInfo
+
+traceFail :: String -> [String] -> FailInfo -> FailInfo
+traceFail fun args fi = addTrace (fmtCall fun args) fi
+
+fmtCall ::  String -> [String] -> String
+fmtCall fun args = intercalate " " (fun : map addParens args)
+  where addParens s = '(' : s ++ ")"

### 2012-06-01 runtime/ID.hs

@@ -3,10 +3,8 @@
 -- ID module
 -- ---------------------------------------------------------------------------
 module ID
-  ( -- * FailInfo
-    FailInfo, defFailInfo
-    -- * Cover
-  , Cover, incCover, decCover, defCover, isCovered
+  ( -- * Cover
+    Cover, incCover, decCover, defCover, isCovered
     -- * Constraints
   , Constraint (..), Constraints(..), getConstrList
     -- * Decisions
@@ -21,35 +19,33 @@ module ID
   ) where
 
 import Control.Monad (liftM, when, zipWithM_)
-import Data.List (partition)
 
 import Debug
+import FailInfo (FailInfo)
 import IDSupply hiding (getDecisionRaw, setDecisionRaw, unsetDecisionRaw)
 import qualified IDSupply
 
 -- ---------------------------------------------------------------------------
--- Fail Info
---  - will eventually collect information about the origin of failures
--- ---------------------------------------------------------------------------
-
-type FailInfo = ()
-
-defFailInfo :: FailInfo
-defFailInfo = ()
-
--- ---------------------------------------------------------------------------
 -- Cover
--- - used to store information about the covering depth of choices,
---   guards and failures
+-- ---------------------------------------------------------------------------
 
+-- |Type used to store information about the covering depth of choices,
+-- guards and failures
 type Cover = Int
+
+-- |Increase covering depth
 incCover :: Cover -> Cover
 incCover = (+ 1)
+
+-- |Decrease covering depth
 decCover :: Cover -> Cover
 decCover = flip (-) 1
+
+-- |Default covering depth (corresponds to no covering)
 defCover :: Cover
 defCover = 0
 
+-- |Is the non-determinism covered?
 isCovered :: Cover -> Bool
 isCovered x = x > 0
 
@@ -68,7 +64,6 @@ data Constraint
   -- |Non-deterministic choice between a list of lists of constraints
   | ConstraintChoices Cover ID [[Constraint]]
  deriving (Show,Eq)
-
 
 -- A Value Constraint is used to bind a Value to an id it also contains the
 -- structural constraint information that describes the choice to be taken

### 2012-06-01 runtime/PrimTypes.hs

@@ -4,6 +4,7 @@ module PrimTypes where
 import System.IO (Handle)
 
 import Debug
+import FailInfo
 import Types
 
 -- BinInt

### 2012-06-01 runtime/Search.hs

@@ -231,14 +231,14 @@ printValsDFS backTrack cont goal = do
       -- [NoDecision, ChooseLeft, ChooseRight], therefore the reset action may
       -- be ignored in between
       where decide bt c a = setDecision i c >> printValsDFS bt cont a
-    follow c           = error $ "Search.prChoice: " ++ show c
+    follow c           = internalError $ "Search.prChoice: " ++ show c
 
   prFree _ i xs   = lookupDecisionID i >>= follow
     where
     follow (LazyBind cs, _) = processLB backTrack cs i xs
     follow (ChooseN c _, _) = printValsDFS backTrack cont (xs !! c)
     follow (NoDecision , j) = cont $ choicesCons defCover j xs
-    follow c                = error $ "Search.prFree: " ++ show c
+    follow c                = internalError $ "Search.prFree: " ++ show c
 
   prNarrowed _ i@(NarrowedID pns _) xs = lookupDecision i >>= follow
     where
@@ -251,8 +251,8 @@ printValsDFS backTrack cont goal = do
       | otherwise        = foldr1 (>>) $
         zipWithButLast3 (decide True) (decide False) [0 ..] xs pns
       where decide bt n a pn = setDecision i (ChooseN n pn) >> printValsDFS bt cont a
-    follow c           = error $ "Search.prNarrowed: Bad choice " ++ show c
-  prNarrowed _ i _ = error $ "Search.prNarrowed: Bad narrowed ID " ++ show i
+    follow c           = internalError $ "Search.prNarrowed: Bad choice " ++ show c
+  prNarrowed _ i _ = internalError $ "Search.prNarrowed: Bad narrowed ID " ++ show i
 
   prGuard _ cs e = solve cs e >>= \mbSltn -> case mbSltn of
     Nothing                      -> return ()
@@ -320,14 +320,14 @@ searchDFS act goal = do
       follow ChooseLeft  = dfs cont x1
       follow ChooseRight = dfs cont x2
       follow NoDecision  = decide i ChooseLeft x1 +++ decide i ChooseRight x2
-      follow c           = error $ "Search.dfsChoice: Bad choice " ++ show c
+      follow c           = internalError $ "Search.dfsChoice: Bad choice " ++ show c
 
     dfsFree cd i xs = lookupDecisionID i >>= follow
       where
       follow (LazyBind cs, _) = processLB i cs xs
[…]

### 2012-06-01 runtime/Types.hs

@@ -14,6 +14,7 @@ module Types
 
 import ConstStore
 import Debug
+import FailInfo
 import ID
 
 -- ---------------------------------------------------------------------------

### 2012-06-01 runtime/ID.hs
Fixed bug that caused covered non-determinism to be encapsulated if the non-deterministic value was bound to a variable

@@ -64,9 +64,9 @@ data Constraint
   -- |Unsolvable constraint
   | Unsolvable FailInfo
   -- |Non-deterministic choice between two lists of constraints
-  | ConstraintChoice ID [Constraint] [Constraint]
+  | ConstraintChoice Cover ID [Constraint] [Constraint]
   -- |Non-deterministic choice between a list of lists of constraints
-  | ConstraintChoices ID [[Constraint]]
+  | ConstraintChoices Cover ID [[Constraint]]
  deriving (Show,Eq)
 
 

### 2012-06-01 runtime/Solver.hs

@@ -35,21 +35,21 @@ solve cnstrs val = solve' (getConstrList cnstrs) val
 
 solveOne :: (Store m, NonDet a) => Constraint -> a -> Solution m a
 solveOne (Unsolvable _) _ = noSolution
-solveOne (ConstraintChoice i lcs rcs) e = lookupDecision i >>= follow
+solveOne (ConstraintChoice cd i lcs rcs) e = lookupDecision i >>= follow
   where
-  follow ChooseLeft  = mkSolution $ guardCons  defCover (StructConstr lcs) e
-  follow ChooseRight = mkSolution $ guardCons  defCover (StructConstr rcs) e
-  follow NoDecision  = mkSolution $ choiceCons defCover i
-                                    (guardCons defCover (StructConstr lcs) e)
-                                    (guardCons defCover (StructConstr rcs) e)
+  follow ChooseLeft  = mkSolution $ guardCons  cd (StructConstr lcs) e
+  follow ChooseRight = mkSolution $ guardCons  cd (StructConstr rcs) e
+  follow NoDecision  = mkSolution $ choiceCons cd i
+                                    (guardCons cd (StructConstr lcs) e)
+                                    (guardCons cd (StructConstr rcs) e)
   follow c           = error $ "Solver.solve.choose: CC:" ++ show c
 
-solveOne cc@(ConstraintChoices i css) e = lookupDecision i >>= follow
+solveOne cc@(ConstraintChoices cd i css) e = lookupDecision i >>= follow
   where
-  follow (ChooseN c _) = mkSolution $ guardCons defCover (StructConstr (css !! c)) e
-  follow NoDecision    = mkSolution $ choicesCons defCover i
-                                    $ map (\cs -> guardCons defCover (StructConstr cs) e) css
-  follow (LazyBind cs) = mkSolution $ guardCons defCover (StructConstr (cs ++ [cc])) e
+  follow (ChooseN c _) = mkSolution $ guardCons cd (StructConstr (css !! c)) e
+  follow NoDecision    = mkSolution $ choicesCons cd i
+                                    $ map (\cs -> guardCons cd (StructConstr cs) e) css
+  follow (LazyBind cs) = mkSolution $ guardCons cd (StructConstr (cs ++ [cc])) e
   follow c             = error $ "Solver.solve.choose: CCs:" ++ show c
 
 solveOne (i :=: cc) e = lookupDecision i >>= follow cc

### 2012-06-01 runtime/Types.hs

@@ -117,6 +117,15 @@ narrows cs cd i@(FreeID        p s) f xs
 narrows _  cd i@(NarrowedID      _ _) f xs = choicesCons cd i (map f xs)
 narrows _  _    (ChoiceID          _) _ _  = internalError "Types.narrows: ChoiceID"
 
+
+bindOrNarrow :: Unifiable a => ID -> Cover -> ID -> [a] -> [Constraint]
+bindOrNarrow i cd j@(FreeID p s) xs | isCovered cd = [ i:=: BindTo j ]
+                                    | otherwise    = [ConstraintChoices cd (NarrowedID p s) (map (bind i) xs)]
+
+lazyBindOrNarrow :: Unifiable a => ID -> Cover -> ID -> [a] -> [Constraint]
+lazyBindOrNarrow i cd j@(FreeID p s) xs | isCovered cd = [ i:=: BindTo j ]
+                                        | otherwise    = [ConstraintChoices cd (NarrowedID p s) (map (lazyBind i) xs)]
+
 -- ---------------------------------------------------------------------------
 -- Computation of normal forms
 -- ---------------------------------------------------------------------------
@@ -456,16 +465,16 @@ instance Unifiable C_Success where
   (=.<=) C_Success C_Success _ = C_Success
   (=.<=) _ _ _ = Fail_C_Success defCover defFailInfo
   bind i C_Success = ((i :=: (ChooseN 0 0)):(concat []))
-  bind i (Choice_C_Success cd j l r) = [(ConstraintChoice j (bind i l) (bind i r))]
-  bind i (Choices_C_Success cd j@(FreeID _ _) _) = [(i :=: (BindTo j))]
-  bind i (Choices_C_Success cd j@(NarrowedID _ _) xs) = [(ConstraintChoices j (map (bind i) xs))]
+  bind i (Choice_C_Success cd j l r) = [(ConstraintChoice cd j (bind i l) (bind i r))]
+  bind i (Choices_C_Success cd j@(FreeID _ _) xs) = bindOrNarrow i cd j xs 
+  bind i (Choices_C_Success cd j@(NarrowedID _ _) xs) = [(ConstraintChoices cd j (map (bind i) xs))]
[…]

### 2012-06-07 runtime/FailInfo.hs
added customFail function

@@ -5,6 +5,7 @@
 module FailInfo
   ( FailInfo (..)
   , defFailInfo
+  , customFail
   , consFail
   , traceFail
   ) where
@@ -20,6 +21,9 @@ defFailInfo = FailInfo []
 
 addTrace :: String -> FailInfo -> FailInfo
 addTrace t fi = fi { failTrace = t : failTrace fi }
+
+customFail :: String -> FailInfo
+customFail msg = addTrace msg defFailInfo
 
 consFail :: String -> [String] -> FailInfo
 consFail fun args = addTrace (fmtCall fun args) defFailInfo

### 2012-06-07 runtime/Search.hs
Minimal display of fail information (just to start with)

@@ -171,7 +171,7 @@ showChoiceTree n goal = showsTree n [] "" (try goal) []
     | d <= 0    = indent l k . showChar '_' . nl
     | otherwise = indent l k . case ndVal of
       Val v           -> showString "Val " . shows v . nl
-      Fail _ _        -> showChar '!' . nl
+      Fail _ info     -> showChar '!' . nl . shows info . nl
       Choice  _ i x y -> shows i  . nl . showsChildren d l [("L", try x), ("R", try y)]
       Narrowed _ i xs -> shows i  . nl . showsChildren d l (zip (map show [(0 :: Int) ..]) (map try xs))
       Free     _ i xs -> shows i  . nl . showsChildren d l (zip (map show [(0 :: Int) ..]) (map try xs))

### 2012-06-08 talks/badhonnef2012/set-functions/talk.tex

\documentclass[utf8]{beamer}

\usepackage{graphicx}
\usepackage{color}
\usepackage{listings}
\usetheme{Boadilla}

\lstset{basicstyle=\small}

\setbeamertemplate{footline}%{infolines theme}
{
\leavevmode%
\hbox{%
\begin{beamercolorbox}[wd=.333333\paperwidth,ht=2.25ex,dp=1ex,center]{author in head/foot}%
\usebeamerfont{author in head/foot}\insertshortauthor%~~(\insertshortinstitute)
\end{beamercolorbox}%
\begin{beamercolorbox}[wd=.333333\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
\usebeamerfont{title in head/foot}\insertshorttitle
\end{beamercolorbox}%
\begin{beamercolorbox}[wd=.333333\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
\usebeamerfont{date in head/foot}\insertshortdate{}\hspace*{2em}
\insertframenumber{} / \inserttotalframenumber\hspace*{2ex}
\end{beamercolorbox}}%
\vskip0pt%
}

\begin{document}

\title{Negation as Failure with Set Functions}

\date{May 04, 2012}

\author[M. Hanus, B. Peemöller, \underline{F. Reck}]
 { Michael Hanus \and Björn Peemöller\\
            \underline{Fabian Reck} \\
  \texttt{\{mh, bjp, fre\}@informatik.uni-kiel.de}}

\institute{Kiel University}

\begin{frame}
\titlepage
\end{frame}

\section{Introduction}

\begin{frame}[fragile]
\frametitle{Curry}
\begin{columns}[t]
\column{.45\textwidth}
\begin{itemize}
\item Lazy functional logic programing language.
\item Haskell-like syntax.
\item Extended by non-determinism, free variables, constraints, unification,
      failure.
\end{itemize}
\begin{example}<2>
\begin{lstlisting}
> ensure aBool
True
\end{lstlisting}
\end{example}
\column{.45\textwidth}
\begin{block}{Running Code Example}
\begin{lstlisting}
data Bool = True | False

aBool :: Bool
aBool = True ? False

not True  = False
not False = True

ensure True = True
\end{lstlisting}
\end{block}
\end{columns}
\end{frame}

\begin{frame}[fragile]
\frametitle{Example}
\begin{block}{Flight data as non-deterministic operation}
\begin{lstlisting}[mathescape]
flight = (LH469, Portland, Frankfurt, 10:.15)
       ? (NWA92, Portland, Amsterdam, 10:.00)
       ? (LH10,  Frankfurt,Hamburg,    1:.00)
       ? (KL1783,Amsterdam,Hamburg,    1:.52)

itinerary orig dest 
   | flight =:= (num,orig,dest,len)
   = [num]                         where num,len free

itinerary orig dest 
   | flight =:= (num1,orig,stop,len1)
   & flight =:= (num2,stop,dest,len2)
   = [num1,num2]      where num1,len1,num2,len2,stop free
\end{lstlisting}
\end{block}
\pause
Problem: itinerary with shortest air time? $\leadsto$ reason about \emph{all} values
\end{frame}

\begin{frame}[fragile]
\frametitle{Encapsulated Search}
\begin{itemize}
\item Allows to access the non-deterministic results of an expression
      within the program
\end{itemize}
\begin{block}{Example}
\begin{lstlisting}
> allValues aBool
[True,False]
\end{lstlisting}
\end{block}
\pause
\begin{block}{Problem}
\begin{lstlisting}
> allValues $! aBool
[True]
[False]
\end{lstlisting}%$
\end{block}
\begin{itemize}
\item Result depends on the evaluation order
\item Declarative languages are order independent 
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Set Functions}
\begin{itemize}
\item Novel approach to encapsulated search
\item For every function $f$ there is a function $f_S$
\item $f_S$ takes the same arguments as $f$
\item The result of $f_S$: Set of non-deterministic results of $f$
\item Non-determinism in arguments to $f_S$ is not capsuled
\item Every result of $f \overline{t}$ is element of a
      result of $f_S \overline{t}$
\end{itemize}
\begin{block}{Example}
\begin{lstlisting}[mathescape]
> aBool$_S$ 
[True,False]

> not$_S$ aBool
[False]
[True]
\end{lstlisting}
\end{block} 
\end{frame}

\begin{frame}[fragile]
\frametitle{Implementation Approaches}
\begin{itemize}
\item Evaluate the arguments strictly \emph{before}
      passing them to the function
\pause
\begin{block}{Problem}
\begin{lstlisting}[mathescape]
answer x = 42

> answer (ensure False)
42

> answer$_S$ (ensure False)
no more solutions
\end{lstlisting}
\end{block}
\item Identify non-determinism that stems from an argument
\end{itemize}
\end{frame}

\section{Compilation scheme}

\begin{frame}[fragile]
\frametitle{Representation of Non-Determinism and Failure in KiCS2}
\begin{itemize}
  \item No built-in non-determinism in Haskell.
  \item Idea: Explicit representation in data types as constructors.
\end{itemize}
\pause
\begin{block}{Curry}
\begin{lstlisting}
data Bool = True | False
aBool = True ? False
ensure True = True
\end{lstlisting}
\end{block}

\begin{block}{Haskell}
\begin{lstlisting}
data Bool = True | False | Choice Bool Bool | Fail
aBool = Choice True False
ensure True = True
ensure (Choice l r) = Choice (ensure l) (ensure r)
ensure _    = Fail
\end{lstlisting}
\end{block}
\end{frame}

\begin{frame}[fragile]
\frametitle{Implementation of Set Functions}
\begin{itemize}
\item Mark choices that stem from arguments
\begin{block}{Curry}
\begin{lstlisting}[mathescape]
answer$_S$ x = search (answer (cover x))

cover True         = True
cover False        = False
cover (Choice l r) = Choice$^C$ (cover l) (cover r)
cover Fail         = Fail

search (Choice l r)  = search l ++ search r
search (Choice$^C$ l r) = Choice (search l) (search r)
search Fail          = []
search x             = [x]
\end{lstlisting}
\end{block}
\item Marked choices are retained 
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Negation as Failure}
\begin{itemize}
\item Just check if the result of a set function is the empty set
\begin{block}{Example}
\begin{lstlisting}[mathescape]
data Size = Small | Big
aSize = Small ? Big

getSmallest | isEmpty (smallerThan$_S$ size) = size
  where size = aSize

smallerThan size | anotherSize < size = anotherSize
  where anotherSize = aSize

> getSmallest
Small
\end{lstlisting}
\end{block}
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Negation as Failure and Failures}
\begin{block}{Problem}
\begin{lstlisting}[mathescape]
getSmallest | isEmpty (smallerThan$_S$ 
                         (size =:= aSize &> size)) 
            = size
  where size = aSize

> getSmallest
Small
Small
Big
\end{lstlisting}
\end{block}
\begin{itemize}
 \item Failures in arguments lead to empty results of set functions
 \item Set functions should behave as if arguments were
       evaluated outside
 \item Implementation: Also mark failures
\end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Be Lazy}
\begin{block}{Problem}
\begin{lstlisting}[mathescape]
trueOrSomething x = True ? x
> trueOrSomething failed
True
> trueOrSomething$_S$ failed
no more solutions
\end{lstlisting}
\end{block}
\begin{itemize}
\item Set functions should behave as if arguments were
      evaluated ouside as far as all their non-deterministic
      branches require
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Conclusion and future work}
\begin{itemize}
\item Set functions are an order independent approach to encapsulated
      search
\item Check for empty result set allows programming with negation as failure
\item Difficult to handle failing arguments
\item Need for theoretic justification of the behaviour
\item Need for formal semantics
\end{itemize}
\end{frame}

\end{document}

### 2012-06-29 lib/External_Directory.hs
moved libs to a separate repository and imported them as submodule

@@ -1,50 +1,0 @@
-import System.Directory
-import System.IO
-import System.Time
-
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_doesFileExist :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Bool
-external_d_C_prim_doesFileExist s _ = toCurry doesFileExist s
-
-external_d_C_prim_doesDirectoryExist :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Bool
-external_d_C_prim_doesDirectoryExist s _ = toCurry doesDirectoryExist s
-
-external_d_C_prim_fileSize :: CP.C_String -> ConstStore -> CP.C_IO CP.C_Int
-external_d_C_prim_fileSize s _ = toCurry
-  (\f -> do h <- openFile f ReadMode
-            i <- hFileSize h
-            hClose h
-            return i
-  ) s
-
-external_d_C_prim_getModificationTime :: CP.C_String -> ConstStore
-                                      -> CP.C_IO Curry_Time.C_ClockTime
-external_d_C_prim_getModificationTime s _ = toCurry getModificationTime s
-
-external_d_C_getCurrentDirectory :: ConstStore -> CP.C_IO (CP.C_String)
-external_d_C_getCurrentDirectory _ = toCurry getCurrentDirectory
-
-external_d_C_prim_setCurrentDirectory :: CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_setCurrentDirectory s _ = toCurry setCurrentDirectory s
-
-external_d_C_prim_getDirectoryContents :: CP.C_String -> ConstStore
-                                       -> CP.C_IO (CP.OP_List (CP.C_String))
-external_d_C_prim_getDirectoryContents s _ = toCurry getDirectoryContents s
-
-external_d_C_prim_createDirectory :: CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_createDirectory s _ = toCurry createDirectory s
-
-external_d_C_prim_removeFile :: CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_removeFile s _ = toCurry removeFile s
-
-external_d_C_prim_removeDirectory :: CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_removeDirectory s _ = toCurry removeDirectory s
-
[…]

### 2012-06-29 lib/External_IO.hs

@@ -1,110 +1,0 @@
-{-# LANGUAGE MultiParamTypeClasses #-}
-import System.IO
-import Control.Concurrent
-import qualified Curry_Prelude as CP
-
-type C_Handle = PrimData CurryHandle
-
-instance ConvertCurryHaskell C_IOMode IOMode where
-  toCurry ReadMode   = C_ReadMode
-  toCurry WriteMode  = C_WriteMode
-  toCurry AppendMode = C_AppendMode
-
-  fromCurry C_ReadMode   = ReadMode
-  fromCurry C_WriteMode  = WriteMode
-  fromCurry C_AppendMode = AppendMode
-  fromCurry _            = error "IOMode data with no ground term occurred"
-
-instance ConvertCurryHaskell C_SeekMode SeekMode where
-  toCurry AbsoluteSeek = C_AbsoluteSeek
-  toCurry RelativeSeek = C_RelativeSeek
-  toCurry SeekFromEnd  = C_SeekFromEnd
-
-  fromCurry C_AbsoluteSeek = AbsoluteSeek
-  fromCurry C_RelativeSeek = RelativeSeek
-  fromCurry C_SeekFromEnd  = SeekFromEnd
-  fromCurry _            = error "SeekMode data with no ground term occurred"
-
-
-external_d_C_stdin :: ConstStore -> C_Handle
-external_d_C_stdin _ = PrimData (OneHandle stdin)
-
-external_d_C_stdout :: ConstStore -> C_Handle
-external_d_C_stdout _ = PrimData (OneHandle stdout)
-
-external_d_C_stderr :: ConstStore -> C_Handle
-external_d_C_stderr _ = PrimData (OneHandle stderr)
-
-external_d_C_prim_openFile :: CP.OP_List CP.C_Char -> C_IOMode
-                           -> ConstStore -> CP.C_IO C_Handle
-external_d_C_prim_openFile fn  mode _ =
-  toCurry (\s m -> openFile s m >>= return . OneHandle) fn mode
-
-external_d_C_prim_hClose :: C_Handle -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_hClose handle _ = toCurry
-  (\ch -> case ch of OneHandle h       -> hClose h
-                     InOutHandle h1 h2 -> hClose h1 >> hClose h2) handle
-
[…]

### 2012-06-29 lib/External_IOExts.hs

@@ -1,144 +1,0 @@
-{-# LANGUAGE MultiParamTypeClasses #-}
-import Data.IORef
-import System.IO.Unsafe   (unsafePerformIO) -- for global associations
-import System.Process     (readProcessWithExitCode, runInteractiveCommand)
-import Control.Concurrent (forkIO)
-import System.IO
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_execCmd :: CP.C_String
- -> ConstStore -> CP.C_IO (CP.OP_Tuple3 Curry_IO.C_Handle Curry_IO.C_Handle Curry_IO.C_Handle)
-external_d_C_prim_execCmd str _ = toCurry
-  (\s -> do (h1,h2,h3,_) <- runInteractiveCommand s
-            return (OneHandle h1, OneHandle h2, OneHandle h3)) str
-
-external_d_C_prim_evalCmd :: CP.C_String -> CP.OP_List CP.C_String -> CP.C_String
-  -> ConstStore -> CP.C_IO (CP.OP_Tuple3 CP.C_Int CP.C_String CP.C_String)
-external_d_C_prim_evalCmd cmd args input _
-  = toCurry readProcessWithExitCode cmd args input
-
-external_d_C_prim_connectToCmd :: CP.C_String -> ConstStore -> CP.C_IO Curry_IO.C_Handle
-external_d_C_prim_connectToCmd str _ = toCurry
-  (\s -> do (hin,hout,herr,_) <- runInteractiveCommand s
-            forkIO (forwardError herr)
-            return (InOutHandle hout hin)) str
-
-forwardError :: Handle -> IO ()
-forwardError h = do
-   eof <- hIsEOF h
-   if eof then return ()
-          else hGetLine h >>= hPutStrLn stderr >> forwardError h
-
-
------------------------------------------------------------------------
--- Implementation of global associations as simple association lists
--- (could be later improved by a more efficient implementation, e.g., maps)
-
-type Assocs = [(String,String)]
-
-assocs :: IORef Assocs
-assocs = unsafePerformIO (newIORef [])
-
-external_d_C_prim_setAssoc :: CP.C_String -> CP.C_String -> ConstStore -> CP.C_IO CP.OP_Unit
-external_d_C_prim_setAssoc str1 str2 _ = toCurry
[…]

### 2012-06-29 lib/External_SetFunctions.hs

@@ -1,4 +1,0 @@
-
-
-external_d_C_cover :: Coverable a => a -> ConstStore -> a
-external_d_C_cover x _ = cover x

### 2012-06-29 lib/External_Unsafe.hs

@@ -1,7 +1,0 @@
-import System.IO.Unsafe(unsafePerformIO)
-import qualified Curry_Prelude as CP
-
-external_d_C_unsafePerformIO :: CP.C_IO a -> ConstStore -> a
-external_d_C_unsafePerformIO io cs = unsafePerformIO  (toIO io cs)
-
------------------------------------------------------------------------

### 2012-09-11 docs/src/manual.tex
Manual updated w.r.t. new Curry version

@@
 \usepackage{graphicx}
+
+\input{version}
 
@@
 \begin{center}\vspace{10ex}
-{\Huge\bf KiCS2 0.2}\\[4ex]
+{\Huge\bf KiCS2}\\[4ex]
 {\huge The Kiel Curry System (Version 2)}\\[7ex]
 {\huge User Manual}\\[4ex]
-Version of June 1, 2012\\[6ex]
+\kicsversiondate\\[6ex]
 \Large
@@
 We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus06Curry}.
+of Curry as described in the Curry language definition \cite{Hanus12Curry}.
 Therefore, this document only explains the use of the different
@@
 (see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.2).
+compared with the language Curry (Version 0.8.3).
 The basic ideas of the implementation of KiCS2
@@
 \begin{itemize}
-\item \index{singleton variables}\index{variables!singleton}
-\emph{Singleton variables}, i.e., variables that occur only once
+\item
+\code{fcase} expressions are not yet supported.
+\item
+The left-hand sides of program rules must be linear,
+i.e., multiple occurrences of a variable in a pattern
+are not allowed.
+\item
+\index{singleton variables}\index{variables!singleton}
+\emph{Singleton pattern variables}, i.e., variables that occur only once
 in a rule, should be denoted as an anonymous variable \ccode{_},

### 2012-09-17 papers/wlp2011/paper.tex
Paper updated

@@
 \usepackage{color}
+\usepackage{times}
 
@@
 and functional patterns can be added without sacrificing
-the efficiency of the kernel implementation.
+the efficiency of the kernel implementation.\\[1ex]
+\textbf{Keywords:} functional logic programming,
+implementation,
+unification,
+functional patterns
 \end{abstract}
@@
 
-Previous implementations of functional logic languages
-can be classified into three categories:
-\begin{enumerate}
- \item designing new abstract machines appropriately supporting
-these operational features and implementing them in some (typically,
+In order to implement a functional logic language,
+one can develop a suitable abstract machine and
+implement it in some (typically,
 imperative) language, like C \cite{Lux99FLOPS}
-or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP},
-
- \item compilation into logic languages like Prolog and reusing
-the existing backtracking implementation for non-deterministic
+or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
+One could also use compile into logic languages like Prolog
+and reuse existing backtracking implementations for non-deterministic
 search as well as logic variables and unification for computing with partial
-information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}, or
-
- \item compilation into non-strict functional languages like Haskell
-and reusing the implementation
-of lazy evaluation and higher-order functions
-\cite{BrasselHuch07,BrasselHuch09}.
-\end{enumerate}
-%
-The latter approach requires the implementation
-of non-deterministic computations in a deterministic language
-but has the advantage
+information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}.
+There are also more recent approaches
+\cite{BrasselHuch07,BrasselHuch09}
+where functional logic programs are compiled
+into non-strict functional programs
+to reuse the implementation
+of lazy evaluation and higher-order functions.
+Although this requires the implementation
+of non-deterministic computations in a deterministic language,
+it has the advantage
 that the explicit handling of non-determinism allows for
@@
 
-In this paper we consider KiCS2 \cite{BrasselHanusPeemoellerReck11},
+This paper is related to the latter implementation approach.
+In particular, we consider KiCS2 \cite{BrasselHanusPeemoellerReck11},
 a new system that compiles functional logic programs of the
-source language Curry \cite{Hanus06Curry}
+source language Curry \cite{Hanus12Curry}
 into purely functional Haskell programs.
-We have shown in \cite{BrasselHanusPeemoellerReck11}
-that this implementation can compete with or outperform
-other existing implementations of Curry.
 KiCS2 is based on the idea to represent
@@
 implementing various strategies.
-Furthermore, logic variables are replaced by generators,
+In KICS2, logic variables are replaced by generators,
 i.e., operations that non-deterministically evaluate to
 all possible ground values of the type of the logic variable.
-It has been shown \cite{AntoyHanus06ICLP}
-that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
+This is justified by the fact that
+computing with logic variables by narrowing \cite{Reddy85,Slagle74}
 and computing with generators by rewriting
-are equivalent, i.e., compute the same values.
-Although this implementation technique is correct \cite{Brassel11Thesis},
+is equivalent, i.e., yields the same values \cite{AntoyHanus06ICLP}.
+Although this implementation technique
+outperforms other existing implementations of Curry
+on deterministic programs and can compete with them on
+non-deterministic programs (see \cite{BrasselHanusPeemoellerReck11}
+for benchmarks),
 the generation of all values for logic variables
@@
 to \code{Y} without enumerating any values for \code{X} or \code{Y}.
-Therefore, we propose in this paper new techniques
-to implement equational constraints in the framework of KiCS2
-(note that, in contrast to Prolog, unification is performed
-modulo function evaluation).
-Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
+In order to obtain a similar behavior in KiCS2,
+we propose in this paper new techniques
+to implement equational constraints
+(in contrast to Prolog, Curry performs unification
+modulo function evaluation)
+in a purely functional target language.
+A purely functional, i.e., side-effect free, implementation
+is reasonable in order to support different, in particular,
+parallel search strategies.
+Beyond equational constraints,
+we also show how functional patterns \cite{AntoyHanus05LOPSTR},
 i.e., patterns containing evaluable operations for more powerful pattern 
@@
 
-The syntax of the functional logic language Curry \cite{Hanus06Curry}
+The syntax of the functional logic language Curry \cite{Hanus12Curry}
 is close to Haskell \cite{PeytonJones03Haskell},
@@
 where values of the arguments of an operation are determined before the
-operation is evaluated. This can be enforced in a lazy strategy
-by sharing actual arguments.
+operation is evaluated.
+In a lazy strategy, this can be enforced by sharing actual arguments.
 For instance, the expression above can be lazily evaluated
@@
 To understand the extensions described in the subsequent sections,
-we sketch the translation of Curry programs
+we review the translation of Curry programs
 into Haskell programs as performed by KiCS2.
@@
 This is obtained by assigning a unique identifier (of type \code{ID})
-to each \code{Choice}. The difficulty is to get a unique identifier
+to each \code{Choice} constructor. The difficulty is to get a unique identifier
 on demand, i.e., when some operation evaluates to a \code{Choice}.
@@
 so that the set \code{s} is split into a set \code{(leftSupply s)}
-containing identifiers for the evaluation of \code{aBool}
+containing identifiers for the evaluation of the argument \code{aBool}
 and a set \code{(rightSupply s)} containing identifiers
@@
 \end{curry}
-The type definition is then extended as follows:
-\begin{haskell}
-  data List a = Nil | Cons a (List a)
-              | Choice ID (List a) (List a) | Fail
+The type definition is then extended as described above:
+\begin{haskell}
+  data List a = Nil | Cons a (List a) | Choice ID (List a) (List a) | Fail
 \end{haskell}
@@
 compute an initial identifier supply passed to this expression).
-The normal form computation is necessary for structured data like lists,
+The normal form computation is necessary for structured data, like lists,
 so that a failure or choice in some part of the data is moved to
@@
 to print all values.
-Furthermore, one can also collect all values in a tree-like data structure
+Instead of printing them, one can also collect
+the values in a tree-like data structure
 so that the programmer can implement his own search strategies
-(this corresponds to encapsulating search \cite{BrasselHanusHuch04JFLP}).
-Finally, instead of printing all values, one can easily define operations
-to print either the first solution only or one by one upon user request.
-Due to the lazy evaluation strategy of Haskell,
-such operations can also be applied to infinite choice structures.
+as tree traversals. Thus, KiCS2 supports a primitive
+\begin{curry}
+  getSearchTree :: a -> IO (SearchTree a)
+\end{curry}
+that returns the search tree corresponding to the evaluation
+of its argument, where the search tree is some computed value,
+a failure, or a choice between two trees:
+\begin{curry}
+  data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
+\end{curry}
+This primitive is useful to encapsulate non-deterministic
+operations and select some result value, e.g., the
+``first'' or the ``best'' one according to some ordering.
+Since the search tree is created in a demand-driven manner,
+the primitive is also applicable to infinite search spaces
+(in contrast to Prolog's \code{findall} primitive \cite{Naish85SLP}).
+Based on this representation, a Curry programmer
+can define his own search strategies
+as tree traversals (see \cite{HanusPeemoellerReck12ATPS}
+for detailed examples).
+Note that these kinds of applications demand for a side-effect
+free implementation of non-deterministic computations
+(in contrast to traditional Prolog implementations
+\cite{Ait-Kaci91})---which is the challenge addressed in this paper.
 
 To avoid an unnecessary growth of the search space represented by
-\code{Choice} constructors, our compiler performs an optimization for
+\code{Choice} constructors, KiCS2 performs an optimization for
 deterministic operations. If an operation is defined by non-overlapping
@@
 
-As mentioned in the introduction, occurrences of logic variables are
-translated into generators.
+As mentioned in the introduction, KiCS2 translates
+occurrences of logic variables into generators.
 For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
@@
 
-The general structure of the implementation of functional patterns in KiCS2
-is quite similar to that of equational constraints,
+Our proposed implementation of functional patterns in KiCS2
+has a structure that is quite similar to that of equational constraints,
 with the exception that variables could be also bound
@@
 and functional patterns in KiCS2, a purely functional implementation
-of Curry. Our implementation is based on adding binding constraints
-to computed values and processing them when values are extracted
+of Curry.
+In addition to the kernel implementation described in
+\cite{BrasselHanusPeemoellerReck11},
+we add binding constraints
+to computed values which are processed when values are extracted
 at the top level of a computation.
-Since we only have added new constructors and pattern matching rules 
-for them in our implementation, no overhead is introduced for programs
+Since only new constructors and pattern matching rules 
+for them are added in our implementation,
+no overhead is introduced for programs
 without equational constraints, i.e.,
@@
 In contrast to those systems, our implementation supports
-a variety of search strategies, like breadth-first or parallel search,
+a variety of ``top-level'' search strategies,
+like iterative deepening, breadth-first or parallel search,
+as well as user-programmable search strategies,
 where the avoidance of side effects is important.
@@
 
-\begin{thebibliography}{10}
-
-\bibitem{AntoyHanus00FROCOS}
-S.~Antoy and M.~Hanus.
-\newblock Compiling multi-paradigm declarative programs into {Prolog}.
-\newblock In {\em Proc. International Workshop on Frontiers of Combining
-  Systems (FroCoS'2000)}, pages 171--185. Springer LNCS 1794, 2000.
-
-\bibitem{AntoyHanus02FLOPS}
-S.~Antoy and M.~Hanus.
-\newblock Functional logic design patterns.
-\newblock In {\em Proc.\ of the 6th International Symposium on Functional and
-  Logic Programming (FLOPS 2002)}, pages 67--87. Springer LNCS 2441, 2002.
-
-\bibitem{AntoyHanus05LOPSTR}
-S.~Antoy and M.~Hanus.
-\newblock Declarative programming with function patterns.
-\newblock In {\em Proceedings of the International Symposium on Logic-based
-  Program Synthesis and Transformation (LOPSTR'05)}, pages 6--22. Springer LNCS
-  3901, 2005.
-
-\bibitem{AntoyHanus06ICLP}
-S.~Antoy and M.~Hanus.
-\newblock Overlapping rules and logic variables in functional logic programs.
-\newblock In {\em Proceedings of the 22nd International Conference on Logic
-  Programming (ICLP 2006)}, pages 87--101. Springer LNCS 4079, 2006.
-
-\bibitem{AntoyHanus10CACM}
-S.~Antoy and M.~Hanus.
-\newblock Functional logic programming.
-\newblock {\em Communications of the ACM}, 53(4):74--85, 2010.
-
-\bibitem{AntoyHanus11WFLP}
-S.~Antoy and M.~Hanus.
-\newblock New functional logic design patterns.
-\newblock In {\em Proc.\ of the 20th International Workshop on Functional and
-  (Constraint) Logic Programming (WFLP 2011)}, pages 19--34. Springer LNCS
-  6816, 2011.
-
-\bibitem{AntoyHanusLiuTolmach05}
-S.~Antoy, M.~Hanus, J.~Liu, and A.~Tolmach.
-\newblock A virtual machine for functional logic computations.
-\newblock In {\em Proc.\ of the 16th International Workshop on Implementation
-  and Application of Functional Languages (IFL 2004)}, pages 108--125. Springer
-  LNCS 3474, 2005.
-
-\bibitem{AugustssonRittriSynek94}
-L.~Augustsson, M.~Rittri, and D.~Synek.
-\newblock On generating unique names.
-\newblock {\em Journal of Functional Programming}, 4(1):117--123, 1994.
-
-\bibitem{Brassel11Thesis}
-B.~Bra{\ss}el.
-\newblock {\em Implementing Functional Logic Programs by Translation into
-  Purely Functional Programs}.
-\newblock PhD thesis, Christian-Albrechts-Universit{\"a}t zu Kiel, 2011.
-
-\bibitem{BrasselFischer08IFL}
-B.~Bra{\ss}el and S.~Fischer.
-\newblock From functional logic programs to purely functional programs
-  preserving laziness.
-\newblock In {\em Pre-Proceedings of the 20th Workshop on Implementation and
-  Application of Functional Languages (IFL 2008)}, 2008.
-
-\bibitem{BrasselHanusHuch04JFLP}
-B.~Bra{\ss}el, M.~Hanus, and F.~Huch.
-\newblock Encapsulating non-determinism in functional logic computations.
-\newblock {\em Journal of Functional and Logic Programming}, 2004(6), 2004.
-
-\bibitem{BrasselHanusPeemoellerReck11}
-B.~Bra{\ss}el, M.~Hanus, B.~Peem{\"o}ller, and F.~Reck.
-\newblock {KiCS2}: A new compiler from {Curry} to {Haskell}.
-\newblock In {\em Proc. of the 20th International Workshop on Functional and
-  (Constraint) Logic Programming (WFLP 2011)}, pages 1--18. Springer LNCS 6816,
-  2011.
-
-\bibitem{BrasselHuch07}
-B.~Bra{\ss}el and F.~Huch.
-\newblock On a tighter integration of functional and logic programming.
-\newblock In {\em Proc. APLAS 2007}, pages 122--138. Springer LNCS 4807, 2007.
-
-\bibitem{BrasselHuch09}
-B.~Bra{\ss}el and F.~Huch.
-\newblock The {Kiel} {Curry} {System} {KiCS}.
-\newblock In {\em Applications of Declarative Programming and Knowledge
-  Management}, pages 195--205. Springer LNAI 5437, 2009.
-
-\bibitem{GiovannettiLeviMoisoPalamidessi91}
-E.~Giovannetti, G.~Levi, C.~Moiso, and C.~Palamidessi.
-\newblock Kernel leaf: A logic plus functional language.
-\newblock {\em Journal of Computer and System Sciences}, 42(2):139--185, 1991.
-
-\bibitem{GonzalezEtAl99}
-J.C. Gonz{\'a}lez-Moreno, M.T. Hortal{\'a}-Gonz{\'a}lez, F.J.
-  L{\'o}pez-Fraguas, and M.~Rodr{\'\i}guez-Artalejo.
-\newblock An approach to declarative programming based on a rewriting logic.
-\newblock {\em Journal of Logic Programming}, 40:47--87, 1999.
-
-\bibitem{Hanus07ICLP}
-M.~Hanus.
-\newblock Multi-paradigm declarative languages.
-\newblock In {\em Proceedings of the International Conference on Logic
-  Programming (ICLP 2007)}, pages 45--75. Springer LNCS 4670, 2007.
-
-\bibitem{Hanus11ICLP}
-M.~Hanus.
-\newblock Declarative processing of semistructured web data.
-\newblock In {\em Technical Communications of the 27th International Conference
-  on Logic Programming}, volume~11, pages 198--208. Leibniz International
-  Proceedings in Informatics (LIPIcs), 2011.
-
-\bibitem{Hanus10PAKCS}
-M.~Hanus, S.~Antoy, B.~Bra{\ss}el, M.~Engelke, K.~H{\"o}ppner, J.~Koj,
-  P.~Niederau, R.~Sadre, and F.~Steiner.
-\newblock {PAKCS}: The {P}ortland {A}achen {K}iel {C}urry {S}ystem.
-\newblock Available at \url{http://www.informatik.uni-kiel.de/~pakcs/}, 2010.
-
-\bibitem{HanusSadre99JFLP}
-M.~Hanus and R.~Sadre.
-\newblock An abstract machine for curry and its concurrent implementation in
-  java.
-\newblock {\em Journal of Functional and Logic Programming}, 1999(6), 1999.
-
-\bibitem{Hanus06Curry}
-M.~Hanus~(ed.).
-\newblock Curry: An integrated functional logic language (vers.\ 0.8.2).
-\newblock Available at \url{http://www.curry-language.org}, 2006.
-
-\bibitem{Hussmann92}
-H.~Hussmann.
-\newblock Nondeterministic algebraic specifications and nonconfluent term
-  rewriting.
-\newblock {\em Journal of Logic Programming}, 12:237--255, 1992.
-
-\bibitem{Lopez-FraguasSanchez-Hernandez99}
-F.~L\'opez-Fraguas and J.~S\'anchez-Hern\'andez.
-\newblock {TOY}: A multiparadigm declarative system.
-\newblock In {\em Proc. of RTA'99}, pages 244--247. Springer LNCS 1631, 1999.
-
-\bibitem{Lux99FLOPS}
-W.~Lux.
-\newblock Implementing encapsulated search for a lazy functional logic
-  language.
-\newblock In {\em Proc. 4th Fuji International Symposium on Functional and
-  Logic Programming (FLOPS'99)}, pages 100--113. Springer LNCS 1722, 1999.
-
-\bibitem{ODonnell85}
-M.J. O'Donnell.
-\newblock {\em Equational Logic as a Programming Language}.
-\newblock MIT Press, 1985.
-
-\bibitem{PeytonJones03Haskell}
-S.~Peyton~Jones, editor.
-\newblock {\em Haskell 98 Language and Libraries---The Revised Report}.
-\newblock Cambridge University Press, 2003.
-
-\bibitem{Reddy85}
-U.S. Reddy.
-\newblock Narrowing as the operational semantics of functional languages.
-\newblock In {\em Proc. IEEE Internat. Symposium on Logic Programming}, pages
-  138--151, Boston, 1985.
-
-\bibitem{Robinson65}
-J.A. Robinson.
-\newblock A machine-oriented logic based on the resolution principle.
-\newblock {\em Journal of the ACM}, 12(1):23--41, 1965.
-
-\bibitem{SchrijversStuckeyWadler09}
-T.~Schrijvers, P.~Stuckey, and P.~Wadler.
-\newblock Monadic constraint programming.
-\newblock {\em Journal of Functional Programming}, 19(6):663--697, 2009.
-
-\bibitem{Slagle74}
-J.R. Slagle.
-\newblock Automated theorem-proving for theories with simplifiers,
-  commutativity, and associativity.
-\newblock {\em Journal of the ACM}, 21(4):622--642, 1974.
-
-\end{thebibliography}
+\bibliographystyle{mhnumber}
+\bibliography{mh}
 

### 2012-09-21 docs/src/manual.tex

@@
 is only computed when the user requests it.
-In this mode, the computation is performed in a separate
-window created by the command \code{interactivecommand}
-whose value is defined in \ccode{.kics2rc}
-(see Section~\ref{sec-customization}).
+Thus, one has also the possibility to terminate the
+enumeration of all values after having seen some values.
 

### 2012-09-24 papers/wlp2011/paper.tex
added an explanation for the cases where =:<= is slower than =:=
PADL 2013 paper updated

@@
 or Java \cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP}.
-One could also use compile into logic languages like Prolog
+One could also compile into logic languages like Prolog
 and reuse existing backtracking implementations for non-deterministic
@@
 information \cite{AntoyHanus00FROCOS,Lopez-FraguasSanchez-Hernandez99}.
-There are also more recent approaches
+More recent approaches
 \cite{BrasselHuch07,BrasselHuch09}
-where functional logic programs are compiled
+compile functional logic programs
 into non-strict functional programs
@@
 \begin{haskell}
-  printValsDFS :: a -> IO ()$\listline$
-  printValsDFS Fail             = return ()$\listline$
+  printValsDFS :: a -> IO ()
+  printValsDFS Fail             = return ()
   printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
@@
                            printValsDFS x
-                           setDecision i NoDecision$\listline$
+                           setDecision i NoDecision
   printValsDFS v = print v
@@
 can define his own search strategies
-as tree traversals (see \cite{HanusPeemoellerReck12ATPS}
+as tree traversals in his source program without any modification
+of the Curry compiler (see \cite{HanusPeemoellerReck12ATPS}
 for detailed examples).
@@
 The last equational constraint ensures that the first sublist
-\code{ys} does not contain duplicated elements
+\code{ys} does not contain duplicate elements
 (the library operation \code{nub} removes all duplicates from a list)
@@
 expression which has to be evaluted only partially. Using strict unification,
-all elements of the list are (unnecessarily) evaluated.
+all elements of the list are (unnecessarily) evaluated. On the other
+hand, lazy unification causes some unnecessary overhead when all elements are
+evaluated anyway, which is shown by the \code{fromPeano} and 
+\code{palindrome} examples.
 

### 2012-09-24 papers/wlp2011/paper.tex
Further updates, git to 16 pages

@@
 \usepackage{listings}
-\lstset{aboveskip=1.0ex,
-        belowskip=1.0ex,
+\lstset{aboveskip=0.9ex,
+        belowskip=0.9ex,
         showstringspaces=false, % no special string space
@@
 This has the advantage that one can apply various,
-and in particular, complete search strategies to compute solutions.
+in particular, complete search strategies
+or even user-defined strategies to compute solutions.
 However, the generation of all values for logic variables
@@
 More recent approaches
-\cite{BrasselHuch07,BrasselHuch09}
+\cite{BrasselFischer08IFL,BrasselHuch07,BrasselHuch09}
 compile functional logic programs
@@
 that the explicit handling of non-determinism allows for
-various search strategies like depth-first, breadth-first, parallel,
-or iterative deepening instead of committing to a fixed (incomplete)
+various search strategies, like depth-first, breadth-first, parallel,
+or iterative deepening, instead of committing to a fixed (incomplete)
 strategy like backtracking \cite{BrasselHuch07}.
@@
 implementing various strategies.
-In KICS2, logic variables are replaced by generators,
+Logic variables are replaced by generators,
 i.e., operations that non-deterministically evaluate to
@@
 and computing with generators by rewriting
-is equivalent, i.e., yields the same values \cite{AntoyHanus06ICLP}.
+are equivalent, i.e., yield the same values \cite{AntoyHanus06ICLP}.
 Although this implementation technique
-outperforms other existing implementations of Curry
+outperforms other implementations of Curry
 on deterministic programs and can compete with them on
@@
 might be inefficient for applications that
-exploit constraints on partial values.
-For instance, in Prolog the equality constraint \ccode{X=c(a)}
-is solved by instantiating the variable \code{X} to \code{c(a)},
+exploit constraints on partially known values.
+For instance, the equality constraint \ccode{X=c(a)}
+is solved in Prolog by instantiating the variable \code{X} to \code{c(a)},
 but the equality constraint \ccode{X=Y} is solved by binding \code{X}
@@
 is reasonable in order to support different, in particular,
-parallel search strategies.
+parallel or user-defined search strategies.
 Beyond equational constraints,
@@
 \begin{curry}
-  not True  = False
-  not False = True
-
-  xor True  x = not x
-  xor False x = x
+  not True  = False            xor True  x = not x
+  not False = True             xor False x = x
 
@@
 and exclude such unintended results,
-Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
-the rewriting logic CRWL as a logical
+the rewriting logic CRWL is proposed in \cite{GonzalezEtAl99} as a logical
 (execution- and strategy-independent) foundation for declarative
@@
 \end{haskell}
-The first argument of type \code{ID} of each \code{Choice} constructor
+where \code{Fail} represents a failure and \code{(Choice i t1 t2)}
+a non-deterministic value, i.e.,
+a selection of two values \code{t1} and \code{t2}
+that can be chosen by some search strategy.
+The first argument \code{i} of type \code{ID} of a \code{Choice} constructor
 is used to implement the call-time choice semantics
@@
 There are different implementations available \cite{AugustssonRittriSynek94}
-(see below for a simple implementation) and our system
+(see below for a simple one) and our system
 is parametric over concrete implementations of \code{IDSupply}.
@@
 \code{False} as the only possible value. In consequence, 
-the unintended value \code{True} is not produced.
+the unintended value \code{True} is not extracted as a result.
 
@@
 \begin{haskell}
-  printValsDFS :: a -> IO ()
-  printValsDFS Fail             = return ()
+  printValsDFS :: a -> IO ()$\listline$
+  printValsDFS Fail             = return ()$\listline$
   printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
@@
                            printValsDFS x
-                           setDecision i NoDecision
+                           setDecision i NoDecision$\listline$
   printValsDFS v = print v
@@
 Our proposed implementation of functional patterns in KiCS2
-has a structure that is quite similar to that of equational constraints,
+has a structure that is quite similar to that of equational constraints
 with the exception that variables could be also bound
@@
 \code{last' [failed,True]} is evaluated (we omit failed branches
-and some other details; note that logic variables are replaced
+and some other details). Note that logic variables are replaced
 by generators, i.e., we assume that \code{xs} is replaced by
-\code{aBoolList 2} and \code{e} is replaced by \code{aBool 3}):
+\code{aBoolList 2} and \code{e} is replaced by \code{aBool 3}:
 \begin{haskell}
@@
 Note that the variable with identifier \code{4} does not occur
-anywhere else, so that the binding \code{(lazyBind 4 failed)}
+anywhere else so that the binding \code{(lazyBind 4 failed)}
 will never be evaluated, as intended.
@@
 expression which has to be evaluted only partially. Using strict unification,
-all elements of the list are (unnecessarily) evaluated. On the other
-hand, lazy unification causes some unnecessary overhead when all elements are
-evaluated anyway, which is shown by the \code{fromPeano} and 
-\code{palindrome} examples.
+all elements of the list are (unnecessarily) evaluated.
+On the other hand, lazy unification causes some overhead
+when the expressions are fully evaluated,
+which is shown by the \code{fromPeano} and \code{palindrome} examples.
+Thus, it is reasonable to use it only if its improved computational
+power is really required, as intended by its design.
 
@@
 \hline
-\code{last (map (inc 0) [1..10000])}   & 0.05  & 0.40   &  0.01\\
+\code{last (map (inc 0) [1..10000])}    & 0.05  & 0.40   & 0.01 \\
 \code{simplify}                         & 6.77  & 0.15   & 0.00 \\

### 2012-09-24 papers/wlp2011/paper.tex
Sentence about practical usage added

@@
 to a comparably efficient execution, as demonstrated by our benchmarks.
+Although the benchmarks were small in order to evaluate
+our unification implementation,
+it should be noted that KiCS2 is used in larger applications,
+like the curricula and module information system of our department%
+\footnote{\url{http://www-ps.informatik.uni-kiel.de/~mh/studiengaenge/}}.
+In these and similar applications, where large parts are
+purely functional computations,
+KiCS2 is 15-20 times faster than PAKCS \cite{Hanus10PAKCS}.
 
 Other implementations of equational constraints in functional logic
-languages use side effects for their implementation.
+languages are based on side effects.
 For instance, PAKCS \cite{Hanus10PAKCS}

### 2012-10-09 docs/src/manual.tex
Manual updated w.r.t. record syntax

@@
 \begin{itemize}
-\item
-\code{fcase} expressions are not yet supported.
-\item
-The left-hand sides of program rules must be linear,
-i.e., multiple occurrences of a variable in a pattern
-are not allowed.
 \item
@@
 \begin{curry}
-{ $l_1$ = $v_1$,$\ldots$, $l_m$ = $v_m$ }
+{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
 \end{curry}
@@
 \begin{curry}
-{name = "Johnson", age = 30}     -- generates a record of type 'Person'
-{left = True, right = 20}        -- generates a record of type 'Branch'
+{name := "Johnson", age := 30}     -- generates a record of type 'Person'
+{left := True, right := 20}        -- generates a record of type 'Branch'
 \end{curry}
@@
 \begin{curry}
-{age = 30, name = "Johnson"}     -- generates a record of type 'Person'
+{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
 \end{curry}
@@
 \begin{curry}
-{person = {name = "Smith", age = 20},   -- generates a record of
- street = "Main Street",                -- type 'Address'
- city   = "Springfield"}
+{person := {name := "Smith", age := 20},   -- generates a record of
+ street := "Main Street",                  -- type 'Address'
+ city   := "Springfield"}
 \end{curry}
@@
 \begin{curry}
-$r$ -> $l$
+$r$ :> $l$
 \end{curry}
@@
 \begin{curry}
-pers -> name
+pers :> name
 \end{curry}
@@
 \begin{curry}
-(addr -> person) -> age
+(addr :> person) :> age
 \end{curry}

### 2012-11-08 papers/wlp2011/paper.tex
Modifications to the PADL paper

@@
 and the features considered in this paper.
-Section~\ref{sec:Compilation} sketches the implementation scheme
-of KiCS2.
+Section~\ref{sec:Compilation} recapitulates the implementation scheme
+of KiCS2 originaly presented in \cite{BrasselHanusPeemoellerReck11}.
 Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
-discuss the extensions to implement unification modulo
+discuss our new extensions to implement unification modulo
 functional evaluation and functional patterns, respectively.
-Benchmarks demonstrating the usefulness of this scheme
+Benchmarks demonstrating the usefulness of these extensions
 are presented in Sect.~\ref{sec:Benchmarks}
@@
 on demand, i.e., when some operation evaluates to a \code{Choice}.
+We cannot thread an identifier supply, e.g. a counter,
+through the search tree without fixing an evaluation order.
 Since we want to compile into \emph{purely} functional programs
 (in order to enable powerful program optimizations),
-we cannot use unsafe features with side effects to generate
+we can neither use unsafe features with side effects to generate
 such identifiers. 
-Hence, we pass a (conceptually infinite) set of identifiers,
+Hence, we follow the idea presented in \cite{AugustssonRittriSynek94} and
+pass a (conceptually infinite) set of identifiers,
 also called \emph{identifier supply},
@@
 obtained by \code{thisID}.
-There are different implementations available \cite{AugustssonRittriSynek94}
+There are different implementations available
 (see below for a simple one) and our system
@@
 is already a failed computation (failure propagation).
+Since \code{head} is a deterministic operation, it does
+not use the identifier supply \code{s}.
 
@@
 of \code{IDSupply}, KiCS2 supports several implementations
-based on memory cells or finite maps):
+based on memory cells or finite maps).
+For a top level operation that prints all values contained in a choice structure
+in a depth-first manner, these operations would be of the following types:
 \begin{haskell}
@@
 %
-Now we can print all values contained in a choice structure
-in a depth-first manner by the following I/O operation:\footnote{%
+Now the search operation can be defined by the I/O operation below:\footnote{%
 Note that this code has been simplified for readability
@@
 the values in a tree-like data structure
-so that the programmer can implement his own search strategies
-as tree traversals. Thus, KiCS2 supports a primitive
+for further processing. Thus, KiCS2 supports a primitive
 \begin{curry}
@@
 other operations, a function defined by overlapping rules,
-the evaluation of such an operation (like \code{xor} or \code{not})
+the evaluation of such an operation (like \code{head})
 cannot introduce non-deterministic values.
@@
 
-Due to this constructive definition, \ccode{=:=}
+Due to this definition, \ccode{=:=}
 can be considered as a binary function defined by the
@@
 \end{haskell}
-Furthermore, we use this operation in narrowing steps,
+We use this operation in narrowing steps,
 i.e., in all rules operating on \code{Choice} constructors.
@@
 \end{haskell}
-%
+to ensure that the resulting choice is not considered a free variable.
+
 As mentioned above, the consideration of free variables
@@
 target language Haskell).
-However, it is evaluated by \code{lazyBind} when its binding
+However, it is evaluated by \code{lazyBind} 
+as far as necessary when its binding
 is required by another part of the computation.

### 2012-11-09 papers/wlp2011/paper.tex
Typo + more explanation of LazyBind evaluation

@@
 Section~\ref{sec:Compilation} recapitulates the implementation scheme
-of KiCS2 originaly presented in \cite{BrasselHanusPeemoellerReck11}.
+of KiCS2 originally presented in \cite{BrasselHanusPeemoellerReck11}.
 Sections~\ref{sec:Unification} and~\ref{sec:FuncPatterns}
@@
 target language Haskell).
-However, it is evaluated by \code{lazyBind} 
-as far as necessary when its binding
-is required by another part of the computation.
+However, it is evaluated by \code{lazyBind} to head normal form
+when its binding is required by another part of the computation,
+whereas the binding constraints for any sub-expression
+are in turn lazily computed using `LazyBind`.
+
 Similarly to equational constraints,

### 2012-11-09 papers/wlp2011/paper.tex
typo corrected

@@
 whereas the binding constraints for any sub-expression
-are in turn lazily computed using `LazyBind`.
+are in turn lazily computed using \code{lazyBind}.
 

### 2012-11-09 papers/wlp2011/paper.tex
changes to padl paper according to meeting

@@
 
-\pagestyle{plain} % remove for camera ready version
+% \pagestyle{plain} % remove for camera ready version
 \sloppy
@@
 is already a failed computation (failure propagation).
-Since \code{head} is a deterministic operation, it does
-not use the identifier supply \code{s}.
+Since deterministic operations do not introduce new
+\code{Choice} constructors, \code{head} does not use the
+ identifier supply \code{s}.
 
@@
 Now the search operation can be defined by the I/O operation below:\footnote{%
-Note that this code has been simplified for readability
-since the type system of Haskell does not
+Note that this code has been simplified and slightly renamed 
+compared to \cite{BrasselHanusPeemoellerReck11}
+for readability. The type system of Haskell does not
 allow this direct definition.}
@@
 
-Due to this definition, \ccode{=:=}
+According to this definition, \ccode{=:=}
 can be considered as a binary function defined by the
@@
 
-As another example for the advantage of the reduced strictness
-implied by functional patterns,
-consider an operation that returns the first duplicate element in
-a list.
-Using equational constraints, we can define it as follows:
-\begin{curry}
-  fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
-            = e    where ys, zs, e free
-\end{curry}
-The first equational constraint is used to split the input list \code{xs}
-into three sublists.
-The last equational constraint ensures that the first sublist
-\code{ys} does not contain duplicate elements
-(the library operation \code{nub} removes all duplicates from a list)
-and the second equational constraint ensures that the first element
-after \code{ys} occurs in \code{ys}.
-Although this implementation is concise, it cannot be applied
-to infinite lists due to the strict interpretation of \ccode{=:=}.
-This is not the case if we define this operation by a functional pattern:
-\begin{curry}
-  fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
-                        = e
-\end{curry}
-%
-Because of the reduced strictness, the logic variable \code{zs} (matching
-the tail list after the first duplicate) is never evaluated.
-This is due to the fact that a functional pattern like
+TODO fuer mh: Weitere Beispiele \cite{AntoyHanus05LOPSTR,Hanus11ICLP}
+
+% As another example for the advantage of the reduced strictness
+% implied by functional patterns,
+% consider an operation that returns the first duplicate element in
+% a list.
+% Using equational constraints, we can define it as follows:
+% \begin{curry}
+%   fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
+%             = e    where ys, zs, e free
+% \end{curry}
+% The first equational constraint is used to split the input list \code{xs}
+% into three sublists.
+% The last equational constraint ensures that the first sublist
+% \code{ys} does not contain duplicate elements
+% (the library operation \code{nub} removes all duplicates from a list)
+% and the second equational constraint ensures that the first element
+% after \code{ys} occurs in \code{ys}.
+% Although this implementation is concise, it cannot be applied
+% to infinite lists due to the strict interpretation of \ccode{=:=}.
+% This is not the case if we define this operation by a functional pattern:
+% \begin{curry}
+%   fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
+%                         = e
+% \end{curry}
+% %
+% Because of the reduced strictness, the logic variable \code{zs} (matching
+% the tail list after the first duplicate) is never evaluated.
+
+Conceptually, a functional pattern like
 \code{(xs++[e])} abbreviates all values to which it can be evaluated
-(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]} etc.
-Conceptually, the rule defining \code{last'}
+(by narrowing), like \code{[e]}, \code{[x1,e]}, \code{[x1,x2,e]}, and so on.
+In consequence, the rule defining \code{last'}
 abbreviates the following (infinite) set of rules:
@@
 In addition, the example \code{last} shows that the lazy 
-unification operator (\ccode{=:<=}) improves the performance when unifying an
+unification operator (\code{=:<=}) improves the performance when unifying an
 expression which has to be evaluted only partially. Using strict unification,

### 2012-11-14 papers/wlp2011/paper.tex
Comment added, bbl included

@@
 on demand, i.e., when some operation evaluates to a \code{Choice}.
-We cannot thread an identifier supply, e.g. a counter,
+We cannot thread an identifier supply, e.g., a counter,
 through the search tree without fixing an evaluation order.
@@
 thanks to the use of functional patterns.
-
-TODO fuer mh: Weitere Beispiele \cite{AntoyHanus05LOPSTR,Hanus11ICLP}
-
-% As another example for the advantage of the reduced strictness
-% implied by functional patterns,
-% consider an operation that returns the first duplicate element in
-% a list.
-% Using equational constraints, we can define it as follows:
-% \begin{curry}
-%   fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
-%             = e    where ys, zs, e free
-% \end{curry}
-% The first equational constraint is used to split the input list \code{xs}
-% into three sublists.
-% The last equational constraint ensures that the first sublist
-% \code{ys} does not contain duplicate elements
-% (the library operation \code{nub} removes all duplicates from a list)
-% and the second equational constraint ensures that the first element
-% after \code{ys} occurs in \code{ys}.
-% Although this implementation is concise, it cannot be applied
-% to infinite lists due to the strict interpretation of \ccode{=:=}.
-% This is not the case if we define this operation by a functional pattern:
-% \begin{curry}
-%   fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
-%                         = e
-% \end{curry}
-% %
-% Because of the reduced strictness, the logic variable \code{zs} (matching
-% the tail list after the first duplicate) is never evaluated.
+%
+Beyond this improved operational behavior,
+functional patterns can lead to more expressive programs
+(e.g., matching and unification on infinite structures,
+pattern matching at arbitrary depth in recursive data structures)
+and more elegant program patterns
+(see \cite{AntoyHanus05LOPSTR,AntoyHanus11WFLP,Hanus11ICLP} for examples).
 
@@
 
-\bibliographystyle{mhnumber}
-\bibliography{mh}
+\begin{thebibliography}{10}
+
+\bibitem{Ait-Kaci91}
+H.~A{\"\i}t-Kaci.
+\newblock {\em Warren's Abstract Machine}.
+\newblock MIT Press, 1991.
+
+\bibitem{AntoyHanus00FROCOS}
+S.~Antoy and M.~Hanus.
+\newblock Compiling Multi-Paradigm Declarative Programs into {Prolog}.
+\newblock In {\em Proc. International Workshop on Frontiers of Combining
+  Systems (FroCoS'2000)}, pp. 171--185. Springer LNCS 1794, 2000.
+
+\bibitem{AntoyHanus02FLOPS}
+S.~Antoy and M.~Hanus.
+\newblock Functional Logic Design Patterns.
+\newblock In {\em Proc.\ of the 6th International Symposium on Functional and
+  Logic Programming (FLOPS 2002)}, pp. 67--87. Springer LNCS 2441, 2002.
+
+\bibitem{AntoyHanus05LOPSTR}
+S.~Antoy and M.~Hanus.
+\newblock Declarative Programming with Function Patterns.
+\newblock In {\em Proceedings of the International Symposium on Logic-based
+  Program Synthesis and Transformation (LOPSTR'05)}, pp. 6--22. Springer LNCS
+  3901, 2005.
+
+\bibitem{AntoyHanus06ICLP}
+S.~Antoy and M.~Hanus.
+\newblock Overlapping Rules and Logic Variables in Functional Logic Programs.
+\newblock In {\em Proceedings of the 22nd International Conference on Logic
+  Programming (ICLP 2006)}, pp. 87--101. Springer LNCS 4079, 2006.
+
+\bibitem{AntoyHanus10CACM}
+S.~Antoy and M.~Hanus.
+\newblock Functional Logic Programming.
+\newblock {\em Communications of the ACM}, Vol.~53, No.~4, pp. 74--85, 2010.
+
+\bibitem{AntoyHanus11WFLP}
+S.~Antoy and M.~Hanus.
+\newblock New Functional Logic Design Patterns.
+\newblock In {\em Proc.\ of the 20th International Workshop on Functional and
+  (Constraint) Logic Programming (WFLP 2011)}, pp. 19--34. Springer LNCS 6816,
+  2011.
+
+\bibitem{AntoyHanusLiuTolmach05}
+S.~Antoy, M.~Hanus, J.~Liu, and A.~Tolmach.
+\newblock A Virtual Machine for Functional Logic Computations.
+\newblock In {\em Proc.\ of the 16th International Workshop on Implementation
+  and Application of Functional Languages (IFL 2004)}, pp. 108--125. Springer
+  LNCS 3474, 2005.
+
+\bibitem{AugustssonRittriSynek94}
+L.~Augustsson, M.~Rittri, and D.~Synek.
+\newblock On generating unique names.
+\newblock {\em Journal of Functional Programming}, Vol.~4, No.~1, pp. 117--123,
+  1994.
+
+\bibitem{BrasselFischer08IFL}
+B.~Bra{\ss}el and S.~Fischer.
+\newblock From Functional Logic Programs to Purely Functional Programs
+  Preserving Laziness.
+\newblock In {\em Proceedings of the 20th International Symposium on
+  Implementation and Application of Functional Languages (IFL 2008)}, pp.
+  25--42. Springer LNCS 5836, 2008.
+
+\bibitem{BrasselHanusPeemoellerReck11}
+B.~Bra{\ss}el, M.~Hanus, B.~Peem{\"o}ller, and F.~Reck.
+\newblock {KiCS2}: A New Compiler from {Curry} to {Haskell}.
+\newblock In {\em Proc. of the 20th International Workshop on Functional and
+  (Constraint) Logic Programming (WFLP 2011)}, pp. 1--18. Springer LNCS 6816,
+  2011.
+
+\bibitem{BrasselHuch07}
+B.~Bra{\ss}el and F.~Huch.
+\newblock On a Tighter Integration of Functional and Logic Programming.
+\newblock In {\em Proc. APLAS 2007}, pp. 122--138. Springer LNCS 4807, 2007.
+
+\bibitem{BrasselHuch09}
+B.~Bra{\ss}el and F.~Huch.
+\newblock The {Kiel} {Curry} {System} {KiCS}.
+\newblock In {\em Applications of Declarative Programming and Knowledge
+  Management}, pp. 195--205. Springer LNAI 5437, 2009.
+
+\bibitem{GiovannettiLeviMoisoPalamidessi91}
+E.~Giovannetti, G.~Levi, C.~Moiso, and C.~Palamidessi.
+\newblock Kernel LEAF: A Logic plus Functional Language.
+\newblock {\em Journal of Computer and System Sciences}, Vol.~42, No.~2, pp.
+  139--185, 1991.
+
+\bibitem{GonzalezEtAl99}
+J.C. Gonz{\'a}lez-Moreno, M.T. Hortal{\'a}-Gonz{\'a}lez, F.J.
+  L{\'o}pez-Fraguas, and M.~Rodr{\'\i}guez-Artalejo.
+\newblock An approach to declarative programming based on a rewriting logic.
+\newblock {\em Journal of Logic Programming}, Vol.~40, pp. 47--87, 1999.
+
+\bibitem{Hanus07ICLP}
+M.~Hanus.
+\newblock Multi-paradigm Declarative Languages.
+\newblock In {\em Proceedings of the International Conference on Logic
+  Programming (ICLP 2007)}, pp. 45--75. Springer LNCS 4670, 2007.
+
+\bibitem{Hanus11ICLP}
+M.~Hanus.
+\newblock Declarative Processing of Semistructured Web Data.
+\newblock In {\em Technical Communications of the 27th International Conference
+  on Logic Programming}, volume~11, pp. 198--208. Leibniz International
+  Proceedings in Informatics (LIPIcs), 2011.
+
+\bibitem{Hanus10PAKCS}
+M.~Hanus, S.~Antoy, B.~Bra{\ss}el, M.~Engelke, K.~H{\"o}ppner, J.~Koj,
+  P.~Niederau, R.~Sadre, and F.~Steiner.
+\newblock {PAKCS}: The {P}ortland {A}achen {K}iel {C}urry {S}ystem.
+\newblock Available at \url{http://www.informatik.uni-kiel.de/~pakcs/}, 2010.
+
+\bibitem{HanusPeemoellerReck12ATPS}
+M.~Hanus, B.~Peem{\"o}ller, and F.~Reck.
+\newblock Search Strategies for Functional Logic Programming.
+\newblock In {\em Proc. of the 5th Working Conference on Programming Languages
+  (ATPS'12)}, pp. 61--74. Springer LNI 199, 2012.
+
+\bibitem{HanusSadre99JFLP}
+M.~Hanus and R.~Sadre.
+\newblock An Abstract Machine for {Curry} and its Concurrent Implementation in
+  {Java}.
+\newblock {\em Journal of Functional and Logic Programming}, Vol.~1999, No.~6,
+  1999.
+
+\bibitem{Hanus12Curry}
+M.~Hanus~(ed.).
+\newblock Curry: An Integrated Functional Logic Language (Vers.\ 0.8.3).
+\newblock Available at \url{http://www.curry-language.org}, 2012.
+
+\bibitem{Hussmann92}
+H.~Hussmann.
+\newblock Nondeterministic Algebraic Specifications and Nonconfluent Term
+  Rewriting.
+\newblock {\em Journal of Logic Programming}, Vol.~12, pp. 237--255, 1992.
+
+\bibitem{Lopez-FraguasSanchez-Hernandez99}
+F.~L\'opez-Fraguas and J.~S\'anchez-Hern\'andez.
+\newblock {TOY}: A Multiparadigm Declarative System.
+\newblock In {\em Proc. of RTA'99}, pp. 244--247. Springer LNCS 1631, 1999.
+
+\bibitem{Lux99FLOPS}
+W.~Lux.
+\newblock Implementing Encapsulated Search for a Lazy Functional Logic
+  Language.
+\newblock In {\em Proc. 4th Fuji International Symposium on Functional and
+  Logic Programming (FLOPS'99)}, pp. 100--113. Springer LNCS 1722, 1999.
+
+\bibitem{Naish85SLP}
+L.~Naish.
+\newblock All Solutions Predicates in Prolog.
+\newblock In {\em Proc. IEEE Internat. Symposium on Logic Programming}, pp.
+  73--77, Boston, 1985.
+
+\bibitem{ODonnell85}
+M.J. O'Donnell.
+\newblock {\em Equational Logic as a Programming Language}.
+\newblock MIT Press, 1985.
+
+\bibitem{PeytonJones03Haskell}
+S.~Peyton~Jones, editor.
+\newblock {\em Haskell 98 Language and Libraries---The Revised Report}.
+\newblock Cambridge University Press, 2003.
+
+\bibitem{Reddy85}
+U.S. Reddy.
+\newblock Narrowing as the Operational Semantics of Functional Languages.
+\newblock In {\em Proc. IEEE Internat. Symposium on Logic Programming}, pp.
+  138--151, Boston, 1985.
+
+\bibitem{Robinson65}
+J.A. Robinson.
+\newblock A Machine-Oriented Logic Based on the Resolution Principle.
+\newblock {\em Journal of the ACM}, Vol.~12, No.~1, pp. 23--41, 1965.
+
+\bibitem{SchrijversStuckeyWadler09}
+T.~Schrijvers, P.~Stuckey, and P.~Wadler.
+\newblock Monadic Constraint Programming.
+\newblock {\em Journal of Functional Programming}, Vol.~19, No.~6, pp.
+  663--697, 2009.
+
+\bibitem{Slagle74}
+J.R. Slagle.
+\newblock Automated Theorem-Proving for Theories with Simplifiers,
+  Commutativity, and Associativity.
+\newblock {\em Journal of the ACM}, Vol.~21, No.~4, pp. 622--642, 1974.
+
+\end{thebibliography}
 

### 2013-01-11 docs/src/manual.tex
Manual: Removed parenthesis requirement for record expressions

@@
 \item
+Unification is performed without an occur check.
+\item
 There is currently no general connection to external constraint solvers.
@@
 \item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-(not available in global installations, see Section~\ref{sec-install})
-Define additional options passed to the Glasgow Haskell Compiler
+Define additional options passed to the Glasgow Haskell Compiler (GHC)
 when the generated Haskell programs are compiled.
-Usually, such options are useful only for experimental purposes.
+One has to be careful when using such options.
+For instance, in a global installation of KiCS2
+(see Section~\ref{sec-install}), libraries are pre-compiled
+so that inconsistencies might occur if compilation options
+might be changed.
+
+It is safe to pass specific GHC linking options.
+For instance, to enforce the static linking of libraries
+in order to generate an executable (see command \ccode{:save})
+that can be executed in another environment, one could set the
+options
+\begin{curry}
+:set ghc -static -optl-static -optl-pthread
+\end{curry}
+
+Other options are useful for experimental purposes,
+but those should be used only in local installations
+(see Section~\ref{sec-install}) to avoid inconsistent
+target codes for different libraries.
 For instance, setting the option
@@
 \begin{curry}
-(addr :> person) :> age
+addr :> person :> age
 \end{curry}
@@
 is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}). When a field selection is used in
-expressions, it has to be parenthesized.
+(which has the type \code{Address}).
 

### 2013-01-11 docs/src/manual.tex

@@
 \lstnewenvironment{curry}{}{}
+\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
+
 \newcommand{\listline}{\vrule width0pt depth1.75ex}
@@
 Thus, we define \code{sconc} by
-\begin{curry}
+\begin{currynomath}
 sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $\$$## s1) $\$$## s2
+sconc s1 s2 = (prim_sconc $## s1) $## s2
 
@@
 prim_sconc external
-\end{curry}
+\end{currynomath}
 so that it is ensured that the external operation \code{prim_sconc}
@@
 we start by declaring the external operations in the Curry module \code{String}:
-\begin{curry}
+\begin{currynomath}
 printString :: String -> IO ()
-printString s = prim_printString $\$$## s
+printString s = prim_printString $## s
 
@@
 prim_printString external
-\end{curry}
+\end{currynomath} % $
 Next we add the corresponding implementation in the file

### 2013-02-11 docs/src/manual.tex

@@
 \begin{curry}
-last xs | ys++[y] =:= xs  = y   where y,ys free
+last xs | _++[y] =:= xs  = y   where y free
 \end{curry}
@@
 but a defined function that is visible in the scope of the pattern.
+
+\subsection{Order of Pattern Matching}
+
+Curry allows multiple occurrences of pattern variables
+in standard patterns. These are an abbreviation of equational constraints
+between pattern variables.
+Functional patterns might also contain multiple occurrences of
+pattern variables.
+For instance, the operation
+\begin{curry}
+f (_++[x]++_++[x]++_) = x
+\end{curry}
+returns all elements with at least two occurrences in a list.
+
+If functional patterns as well as multiple occurrences of
+pattern variables occur in a pattern defining an operation,
+there are various orders to match an expression against such
+an operation. In the current implementation, the order
+is as follows:
+\begin{enumerate}
+\item Standard pattern matching: First, it is checked whether
+the constructor patterns match. Thus, functional patterns
+and multiple occurrences of pattern variables are ignored.
+\item Functional pattern matching: In the next phase,
+functional patterns are matched but occurrences of standard
+pattern variables in the functional patterns are ignored.
+\item Non-linear patterns: If standard and functional pattern matching
+is successful, the equational constraints which correspond
+to multiple occurrences pattern variables are solved.
+\item Guards: Finally, the guards supplied by the programmer
+are checked.
+\end{enumerate}
+The order of pattern matching should not influence the computed
+result. However, it might have some influence on the termination
+behavior of programs, i.e., a program might not terminate
+instead of finitely failing.
+In such cases, it could be necessary to consider the influence
+of the order of pattern matching. Note that other orders of pattern matching
+can be obtained using auxiliary operations.
 

### 2013-02-13 docs/src/manual.tex
Release notes and manual updated

@@
 
+It is also possible to combine functional patterns with
+as-patterns.\index{as-pattern}\pindex{"@}
+Similarly to the meaning of as-patterns
+in standard constructor patterns,
+as-patterns in functional patterns are interpreted
+as a sequence of pattern matching where the variable of the as-pattern
+is matched before the given pattern is matched.
+This process can be described by introducing an auxiliary operation
+for this two-level pattern matching process.
+For instance, the definition
+\begin{curry}
+f (_ ++ x@[(42,_)] ++ _) = x
+\end{curry}
+is considered as syntactic sugar for the expanded definition
+\begin{curry}
+f (_ ++ x ++ _) = f' x
+ where
+  f' [(42,_)] = x
+\end{curry}
+However, as-patterns are usually implemented
+in a more efficient way without introducing auxiliary operations.
+
 \subsection{Order of Pattern Matching}
@@
                                  [[],[1],[1,2]]
-
 \end{curry}

### 2013-02-14 docs/src/manual.tex
Manual updated

@@
 \lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
-
 \newcommand{\listline}{\vrule width0pt depth1.75ex}
@@
 
-\newcommand{\labs}{\mbox{\tt\char92}}  % lambda abstraction in Curry
 \newcommand{\todo}[1]{\fbox{\sc To do: #1}}
+\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
+\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
 \newcommand{\cb}{CurryBrowser\xspace}
@@
 
-This document describes KiCS2 (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
+This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
 Version \textbf{2}),
@@
 of constraints with synchronization on logical variables).
-The current KiCS2 implementation does not support concurrent
+The current \CYS implementation does not support concurrent
 constraints. Alternatively, one can write distributed applications
@@
 with symbolic names.
-Moreover, KiCS2 also supports
+Moreover, \CYS also supports
 the high-level implementation of
@@
 Therefore, this document only explains the use of the different
-components of KiCS2
-and the differences and restrictions of KiCS2
+components of \CYS
+and the differences and restrictions of \CYS
 (see Section~\ref{sec-restrictions})
 compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of KiCS2
+The basic ideas of the implementation of \CYS
 can be found in
@@
 
-\section{Overview of KiCS2}
+\section{Overview of \CYS}
 
@@
 
-This version of KiCS2 has been developed and
+This version of \CYS has been developed and
 tested on Linux systems.
@@
 
-Installation instructions for KiCS2 can be found in
-the file \code{INSTALL.txt} stored in the KiCS2 installation directory.
-Note that there are two possibilities to install KiCS2:
+Installation instructions for \CYS can be found in
+the file \code{INSTALL.txt} stored in the \CYS installation directory.
+Note that there are two possibilities to install \CYS:
 \begin{description}
 \item[Global installation:]\index{global installation}\index{installation!global}
-KiCS2 is installed in some global system directory where users
+\CYS is installed in some global system directory where users
 have no write permission. In this case, some options for experimenting
-with KiCS2 (like \code{supply} or \code{ghc}, see below)
+with \CYS (like \code{supply} or \code{ghc}, see below)
 are not available (since they require the recompilation of parts of
@@
 \item[Local installation:]\index{local installation}\index{installation!local}
-KiCS2 is installed in some local user directory where the user
+\CYS is installed in some local user directory where the user
 has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the KiCS2 installation is set as follows:
+in the \code{Makefile} of the \CYS installation is set as follows:
 \begin{lstlisting}
@@
 \end{lstlisting}
-In this case, all options of KiCS2 are available.
+In this case, all options of \CYS are available.
 \end{description}
-In the following, $kics2home$ denotes the installation directory
-of the KiCS2 installation.
+In the following, \cyshome denotes the installation directory
+of the \CYS installation.
 
@@
 All executables required to use the different components
-of KiCS2 are stored in the directory \code{$kics2home$/bin}.
+of \CYS are stored in the directory \code{\cyshome/bin}.
 You should add this directory
 to your path (e.g., by the \code{bash} command
-\ccode{export PATH=$kics2home$/bin:\$PATH}).
+\ccode{export PATH=\cyshome/bin:\$PATH}).
 
@@
 
-Since the translation of Curry programs with KiCS2 creates
+Since the translation of Curry programs with \CYS creates
 some auxiliary files (see Section~\ref{sec-auxfiles} for details),
@@
 system libraries according to the setting of some options.
-Therefore, the KiCS2 system should be locally installed
+Therefore, the \CYS system should be locally installed
 in your user account.
@@
 (this is a shell script stored in the \code{bin} directory of the
-KiCS2 installation, see above).
+\CYS installation, see above).
 The command
@@
 There are a few minor restrictions on Curry programs
-when they are processed with KiCS2:
+when they are processed with \CYS:
 \begin{itemize}
@@
 \item
-KiCS2 translates all \emph{local declarations} into global functions with
+\CYS translates all \emph{local declarations} into global functions with
 additional arguments (``lambda lifting'', see Appendix~D of the
@@
 to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, KiCS2 provides
+in programs performing I/O at the top-level, \CYS provides
 the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
@@
 containing free variables are not yet supported.
-KiCS2 supports \emph{value generators} for free variables
+\CYS supports \emph{value generators} for free variables
 so that a free variable is instantiated when its value is demanded.
@@
 
-\subsection{Modules in KiCS2}
+\subsection{Modules in \CYS}
 \label{sec-modules}
 
-The current implementation of KiCS2 supports only flat module names,
+The current implementation of \CYS supports only flat module names,
 i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
 In order to allow the structuring of modules in different directories,
-KiCS2 searches for imported modules in various directories.
+\CYS searches for imported modules in various directories.
 By default, imported modules are searched in the directory
 of the main program and the system module directories
-\ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
+\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
 This search path can be extended
 by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a KiCS2 session by the option
+(which can be also set in a \CYS session by the option
 \ccode{:set path}\pindex{path}\pindex{:set path},
@@
       \ccode{libraries}.
-\item The directories \ccode{$kics2home$/lib} and \ccode{$kics2home$/lib/meta}.
+\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
 \end{enumerate}
-Note that the standard prelude (\code{$kics2home$/lib/Prelude.curry})
+Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
 will be always implicitly imported to all modules if a module
@@
 
-\section{Using the Interactive Environment of KiCS2}
+\section{Using the Interactive Environment of \CYS}
 \label{sec-kics2}
@@
 This section describes the interactive environment
-KiCS2\index{KiCS2}
+\CYS\index{\CYS}
 that supports the development of applications written in Curry.
-The implementation of KiCS2 contains also a separate compiler
+The implementation of \CYS contains also a separate compiler
 which is automatically invoked by the interactive environment.
 
-\subsection{Invoking KiCS2}
+\subsection{Invoking \CYS}
 \label{sec:invoke-kics2}
 
-To start KiCS2, execute the command
+To start \CYS, execute the command
 \ccode{kics2}\pindex{kics2}
 (this is a shell script stored in
-\code{$kics2home$/bin} where $kics2home$ is the installation directory
-of KiCS2).
+\code{\cyshome/bin} where \cyshome is the installation directory
+of \CYS).
 When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{$kics2home$/lib/Prelude.curry})
+the prelude (\code{\cyshome/lib/Prelude.curry})
 is already loaded, i.e., all definitions in the prelude are accessible.
@@
 
-One can also invoke KiCS2 with parameters.
+One can also invoke \CYS with parameters.
 These parameters are usual a sequence of commands
@@
 \end{curry}
-starts KiCS2, loads the main module \code{Mod}, and adds the additional
+starts \CYS, loads the main module \code{Mod}, and adds the additional
 module \code{List}.
@@
 \end{curry}
-starts KiCS2, loads the main module \code{Mod}, and evaluates
+starts \CYS, loads the main module \code{Mod}, and evaluates
 the operation \code{config} before the user interaction starts.
@@
 \end{curry}
-starts KiCS2, loads the main module \code{Mod}, creates an executable,
-and terminates KiCS2. This invocation could be useful in ``make'' files
+starts \CYS, loads the main module \code{Mod}, creates an executable,
+and terminates \CYS. This invocation could be useful in ``make'' files
 for systems implemented in Curry.
 
-\subsection{Command of KiCS2}
+\subsection{Command of \CYS}
 \label{sec:kics2-commands}
 
-The {\bf most important commands} of KiCS2 are
+The {\bf most important commands} of \CYS are
 (it is sufficient to type a unique prefix of a command if it is unique,
@@
 they are specialized to the type \ccode{()}
-(since the current implementation of KiCS2 does not support
+(since the current implementation of \CYS does not support
 computations with polymorphic logic variables).
@@
 but shows the interface of the module $m$
-which must be in the load path of KiCS2.
+which must be in the load path of \CYS.
 
@@
 Set or turn on/off a specific option
-of the KiCS2 environment (see \ref{sec:options} for a description
+of the \CYS environment (see \ref{sec:options} for a description
 of all options). Options are turned on by the prefix
@@
 is evaluated in an independent process which runs in
-parallel to the current KiCS2 process.
+parallel to the current \CYS process.
 All output and error messages from this new process are suppressed.
@@
 
-\subsection{Option of KiCS2}
+\subsection{Options of \CYS}
 \label{sec:options}
@@
 Note that this search path is only used for loading modules
-inside this invocation of KiCS2.
+inside this invocation of \CYS.
 
@@
 understanding the implementation of non-deterministic
-evaluations used in KiCS2.
+evaluations used in \CYS.
 If the optional argument $n$ is provided,
@@
 \item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the KiCS2 compiler.
+Define additional options passed to the \CYS compiler.
 For instance, setting the option
@@
 \end{curry}
-has the effect that all optimizations performed by the KiCS2 compiler
+has the effect that all optimizations performed by the \CYS compiler
 are turned off.
@@
 One has to be careful when using such options.
-For instance, in a global installation of KiCS2
+For instance, in a global installation of \CYS
 (see Section~\ref{sec-install}), libraries are pre-compiled
@@
 If the evaluation of operations in some main module loaded into
-KiCS2 requires specific options, like an iterative-deepening
+\CYS requires specific options, like an iterative-deepening
 search strategy, one can also put these options into the source
@@
 the syntax of Curry)
-has the effect that this option is set by the KiCS2 command
+has the effect that this option is set by the \CYS command
 \ccode{:set $opt$} whenever this module is loaded (not reloaded!)
@@
 In order to have support for line editing or history functionality
-in the command line of KiCS2 (as often supported by the \code{readline}
+in the command line of \CYS (as often supported by the \code{readline}
 library), you should have the Unix command \code{rlwrap} installed
 on your local machine.
-If \code{rlwrap} is installed, it is used by KiCS2 if called on a terminal.
+If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
 If it should not be used (e.g., because it is executed
 in an editor with \code{readline} functionality), one can
-call KiCS2 with the parameter \ccode{--noreadline}
+call \CYS with the parameter \ccode{--noreadline}
 (which must occur as the first parameter).
@@
 
-In order to customize the behavior of KiCS2 to your own preferences,
-there is a configuration file which is read by KiCS2 when it is invoked.
-When you start KiCS2 for the first time, a standard version of
+In order to customize the behavior of \CYS to your own preferences,
+there is a configuration file which is read by \CYS when it is invoked.
+When you start \CYS for the first time, a standard version of
 this configuration file is copied with the name
@@
 programs etc.
-After you have started KiCS2 for the first time, look into this file
+After you have started \CYS for the first time, look into this file
 and adapt it to your own preferences.
@@
 (see \url{http://www.emacs.org}).
-The distribution of KiCS2 contains also a special
+The distribution of \CYS contains also a special
 \emph{Curry mode}\index{Curry mode}\index{Emacs}
@@
 finding declarations in the current buffer, and
-loading Curry programs into KiCS2
+loading Curry programs into \CYS
 in an Emacs shell.
@@
 Its installation is described in the file \code{README}
-in directory \ccode{$kics2home$/tools/emacs} which also contains
+in directory \ccode{\cyshome/tools/emacs} which also contains
 the sources of the Curry mode and a short description about
@@
 
-KiCS2 supports some extensions in Curry programs that are not (yet)
+\CYS supports some extensions in Curry programs that are not (yet)
 part of the definition of Curry. These extensions are described below.
@@
 Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in KiCS2.
+or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
 For instance, the declaration
@@
 
-In contrast to the basic concept in \cite{Leijen05}, KiCS2/Curry provides a
+In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
 simpler version of records. Some of the features described there are
@@
 
-\cb is part of the KiCS2 distribution and can be
+\cb is part of the \CYS distribution and can be
 started in two ways:
@@
 \item
-In the command shell via the command: \code{$kics2home$/bin/currybrowser mod}
+In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
 \item
-In the KiCS2 environment after loading the module
+In the \CYS environment after loading the module
 \code{mod} and typing the command \ccode{:browse}.
@@
 CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the KiCS2 distribution to write
+is a simple tool in the \CYS distribution to write
 and run repeatable tests. CurryTest simplifies the task
@@
 \end{curry}
-(\code{currytest} is a program stored in \code{$kics2home$/bin}
-where $kics2home$ is the installation directory of KiCS2;
+(\code{currytest} is a program stored in \code{\cyshome/bin}
+where \cyshome is the installation directory of \CYS;
 see Section~\ref{sec-install}).
@@
 \end{curry}
-(\code{erd2curry} is a program stored in \code{$kics2home$/bin}
-where $kics2home$ is the installation directory of KiCS2;
+(\code{erd2curry} is a program stored in \code{\cyshome/bin}
+where \cyshome is the installation directory of \CYS;
 see Section~\ref{sec-install}).
@@
 (w.r.t.\ the type definition given in module
-\code{$kics2home$/tools/erd2curry/ERD.curry})
+\code{\cyshome/tools/erd2curry/ERD.curry})
 and store it in some file, e.g., \ccode{myerd.term}.
@@
 To compile the generated database code, either
-include the directory \code{$kics2home$/tools/erd2curry}
+include the directory \code{\cyshome/tools/erd2curry}
 into your Curry load path
@@
 or copy the file
-\code{$kics2home$/tools/erd2curry/ERDGeneric.curry}
+\code{\cyshome/tools/erd2curry/ERDGeneric.curry}
 into the directory of the generated database code.
@@
 
-One can implement distributed systems with KiCS2
+One can implement distributed systems with \CYS
 by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
@@
 Therefore, this section gives some information
-about the technical requirements of KiCS2 and how to solve
+about the technical requirements of \CYS and how to solve
 problems due to these requirements.
 
-There is one fixed port that is used by the implementation of KiCS2:
+There is one fixed port that is used by the implementation of \CYS:
 \begin{description}
@@
 It can also be manually started and terminated by the
-scripts \code{$kics2home$/cpns/start} and
-\code{$kics2home$/cpns/stop}.
-If the demon is already running, the command \code{$kics2home$/cpns/start}
+scripts \code{\cyshome/cpns/start} and
+\code{\cyshome/cpns/stop}.
+If the demon is already running, the command \code{\cyshome/cpns/start}
 does nothing (so it can be always executed
@@
 
-\section{Libraries of the KiCS2 Distribution}
+\section{Libraries of the \CYS Distribution}
 \label{sec:libraries}
@@
 
-The KiCS2 distribution comes with an extensive collection
+The \CYS distribution comes with an extensive collection
 of libraries for application programming.
@@
 are described in the further subsections.
-For a more detailed online documentation of all libraries of KiCS2,
+For a more detailed online documentation of all libraries of \CYS,
 see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
@@
 and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{$kics2home$/lib/meta},
+stored in the directory \ccode{\cyshome/lib/meta},
 which define datatypes for the representation
@@
 
-During the translation and execution of a Curry program with KiCS2,
+During the translation and execution of a Curry program with \CYS,
 various intermediate representations of the source program are created
@@
 
-The various components of KiCS2 create the following auxiliary files.
+The various components of \CYS create the following auxiliary files.
 \begin{description}
@@
 This file is implicitly
-generated when a program is compiled with KiCS2.
+generated when a program is compiled with \CYS.
 The FlatCurry representation of a Curry program is usually
@@
 to module interfaces.
-This file is implicitly generated when a program is compiled with KiCS2
+This file is implicitly generated when a program is compiled with \CYS
 and stored in the same directory as \code{prog.fcy}.
@@
 the determinism behavior of operations that is used by the
-KiCS2 compiler (see \cite{BrasselHanusPeemoellerReck11}
+\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
 for more details about the use of this information).
@@
 as the result of translating the Curry program with the
-KiCS2 compiler.
+\CYS compiler.
 This file is stored in the same directory as \code{Curry_prog.nda}.
@@
 \item[\code{prog}:] This file contains the executable
-after compiling and saving a program with KiCS2
+after compiling and saving a program with \CYS
 (see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
@@
 \index{operation!external}\index{external operation}
-Currently, KiCS2 has no general interface to external operations,
+Currently, \CYS has no general interface to external operations,
 i.e., operations whose semantics is not defined by program rules
@@
 Furthermore, an implementation of the external operation
-must be provided in the target language of the KiCS2 compiler,
+must be provided in the target language of the \CYS compiler,
 i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, KiCS2 follows some code
+In order to simplify this task, \CYS follows some code
 conventions that are described in the following.
@@
 In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of KiCS2.
-The KiCS2 compiler generates the following code for the
+one has to satisfy the naming conventions of \CYS.
+The \CYS compiler generates the following code for the
 external operation \code{prim_sconc} (note that the generated
@@
 directory as the file \code{String.curry}.
-The KiCS2 compiler appends the code contained in
+The \CYS compiler appends the code contained in
 \code{External_String.hs} to the generated code
@@
 that are meaningless in Haskell.
-To solve this conversion problem, the implementation of KiCS2
+To solve this conversion problem, the implementation of \CYS
 provides a family of operations to perform these conversions
@@
 i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the KiCS2 implementation
+and \code{toCurry}, is defined in the \CYS implementation
 for all standard data types.
@@
 When we compile the Curry module \code{String},
-KiCS2 combines these definitions in the target program so that
+\CYS combines these definitions in the target program so that
 we can immediately use the externally defined operation
@@
 
-As we have seen, KiCS2 transforms a name like
+As we have seen, \CYS transforms a name like
 \code{primOP} of an external operation into the name

### 2013-03-07 runtime/Search.hs
test run with new cover implementation

@@ -573,7 +573,7 @@ searchMSearch cd x = evalStateT (searchMSearch' cd return x) (Map.empty :: Decis
 searchMSearch' :: (NormalForm a, MonadSearch m, Store m) => Cover -> (a -> m b) -> a -> m b
 searchMSearch' cd cont = match smpChoice smpChoices smpChoices smpFail smpGuard smpVal
   where
-  smpFail d info = if isCovered d then szero d info else mzero
+  smpFail d info  = szero d info
   smpVal v        = searchNF (searchMSearch' cd) cont v
 
   smpChoice d i x y = lookupDecision i >>= follow
@@ -599,7 +599,7 @@ searchMSearch' cd cont = match smpChoice smpChoices smpChoices smpFail smpGuard 
 
   smpGuard d cs e 
    | isCovered d = constrainMSearch d cs (searchMSearch' cd cont e)
-   | otherwise = solve cd cs e >>= maybe mzero (searchMSearch' cd cont . snd)
+   | otherwise = solve cd cs e >>= maybe (szero d defFailInfo) (searchMSearch' cd cont . snd)
 
   processLB d i cs xs = decide i NoDecision
                         $ guardCons d (StructConstr cs) (choicesCons d i xs)

### 2013-03-07 docs/src/manual.tex

@@
 the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yields equal values.
+asserts that the I/O actions $a_1$ and $a_2$ yield equal values.
 The name $s$ provided as a first argument in each assertion
@@
 
-test3 = assertSolutions "prefix" (\x -> let y free in  x++y =:= [1,2])
+test3 = assertSolutions "prefix" (\x -> x++_ =:= [1,2])
                                  [[],[1],[1,2]]
@@
 \begin{curry}
-currytest testList
+currytest TestList
 \end{curry}
@@
 see Section~\ref{sec-install}).
-In our example, \ccode{testList.curry} is the program containing the
+In our example, \ccode{TestList.curry} is the program containing the
 definition of all assertions. This has the effect
@@
 %If failures occur, the complete test results are also
-%written into a file named \ccode{testList.testlog}.''
+%written into a file named \ccode{TestList.testlog}.''
 For our example above, we obtain the following successful protocol:
@@
 ============================================================
-Testing module "testList"...
+Testing module "TestList"...
 OK: ++
@@
 \begin{curry}
-currytest --window testList
+currytest --window TestList
 \end{curry}
@@
 \begin{curry}
-currytest -w testList
+currytest -w TestList
 \end{curry}

### 2013-04-15 docs/src/manual.tex
Manual update (:set prompt added)

@@
 
+\item[\fbox{\code{prompt $p$}}]\pindex{prompt}
+Sets the user prompt which is shown when \CYS is waiting for input.
+If the parameter $p$ starts with a letter or a percent sign,
+the prompt is printed as the given parameter,
+where the sequence \ccode{\%s} is expanded to the list of
+currently loaded modules
+and \ccode{\%\%} is expanded to a percent sign.
+If the prompt starts with a double quote, it is read as a string and,
+therefore, also supports the normal escape sequences that can occur
+in Curry programs. The default setting is
+\begin{curry}
+:set prompt "%s> "
+\end{curry}
+
 \item[\fbox{\code{+/-interactive}}]\pindex{interactive}

### 2013-05-21 docs/src/manual.tex

@@
 \item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
+Encapsulated search\index{encapsulated search}:
 The general definition of encapsulated search of the Curry report
 \cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
+Thus, the corresponding prelude operations like
+\code{try}\pindex{try},
+\code{solveAll}\pindex{solveAll},
+\code{once}\pindex{once},
+\code{findall}\pindex{findall}, or
+\code{best}\pindex{best}
+are not defined in the \CYS prelude.
+However, \CYS supports appropriate alternatives
+to encapsulate non-deterministic computations:
+\begin{description}
+\item[Strong encapsulation:]
+This means that all potential
+non-determinism is encapsulated. Since this might
+result in dependencies on the evaluation strategy
+(see \cite{BrasselHanusHuch04JFLP} for a detailed discussion),
+this kind of encapsulation is only available as I/O operations.
+For instance, the library \code{AllSolutions}\pindex{AllSolutions}
+(Section~\ref{Library:AllSolutions})
+defines the operation
+\begin{curry}
+getAllValues :: a -> IO [a]
+\end{curry}
+to compute all values of a given argument expression.
+There is also the library \code{SearchTree}\pindex{SearchTree}
+(Section~\ref{Library:SearchTree})
+which supports user-programmable search strategies
+and contains some predefined strategies
+like depth-first, breadth-first, iterative deepening search.
+\item[Weak encapsulation:]
+This means that only the non-determinism
+defined inside an encapsulation operator is encapsulated.
+Conceptually, these operators are offered as
+\emph{set functions}\index{set functions} \cite{AntoyHanus09}
+which compute the set of all results but do not encapsulate
+non-determinism in the actual arguments.
+See the library \code{SetFunctions}
+(Section~\ref{Library:SetFunctions}) for more details.
+\end{description}
+\item
+Concurrent computations\index{concurrency}
 based on the suspension of expressions

### 2013-07-12 docs/src/currybrowser.tex
Split manual into separate tex files

--- 2013-03-07 docs/src/manual.tex
@@
-\documentclass[11pt,fleqn]{article}
-
-\usepackage{latexsym}
-\usepackage{makeidx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{graphicx}
-
-\input{version}
-
-%%% ------------------------------------------------------------------
-
-\usepackage[colorlinks,linkcolor=blue]{hyperref}
-\hypersetup{bookmarksopen=true}
-\hypersetup{bookmarksopenlevel=0}
-\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
-\hypersetup{pdfauthor={Michael Hanus}}
-%\hypersetup{pdfstartview=Title}
-\hypersetup{pdfstartview=FitH}
-\usepackage{thumbpdf}
-
-%%% ------------------------------------------------------------------
-
-\setlength{\textwidth}{16.5cm}
-\setlength{\textheight}{23cm}
-\renewcommand{\baselinestretch}{1.1}
-\setlength{\topmargin}{-1cm}
-\setlength{\oddsidemargin}{0cm}
-\setlength{\evensidemargin}{0cm}
-\setlength{\marginparwidth}{0.0cm}
-\setlength{\marginparsep}{0.0cm}
-
-\newlength{\figurewidth}
-\setlength{\figurewidth}{\textwidth}
-\addtolength{\figurewidth}{-0.4cm}
-
-% font for program texts
-\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\small\tt}
-
-\usepackage{listings}
-\lstset{aboveskip=1.5ex,
-        belowskip=1.5ex,
-        showstringspaces=false, % no special string space
-        mathescape=true,
-        flexiblecolumns=false,
-        xleftmargin=2ex,
-        basewidth=0.52em,
-        basicstyle=\small\ttfamily}
-
-\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
-       }
-
-\lstnewenvironment{curry}{}{}
-\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
-\newcommand{\listline}{\vrule width0pt depth1.75ex}
-
-% program text in normal text
-\newcommand{\code}[1]{\mbox{\codefont #1}}
-
-% program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\code{#1}''}
-
-\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index
-
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
-\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
-\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
-\newcommand{\cb}{CurryBrowser\xspace}
-
-% allow underscores in programs:
-\catcode`\_=\active
-\let_=\sb
-\catcode`\_=12
-
-% produce an index:
-\makeindex
-
-\begin{document}
-\sloppy
-
-\begin{titlepage}
-\pdfbookmark[1]{Title}{Title}
-\begin{center}
-\fbox{
-\begin{minipage}[t]{\figurewidth}
-\begin{center}\vspace{10ex}
-{\Huge\bf KiCS2}\\[4ex]
-{\huge The Kiel Curry System (Version 2)}\\[7ex]
-{\huge User Manual}\\[4ex]
-\kicsversiondate\\[6ex]
-\Large
-Michael Hanus$^1$ [editor] \\[3ex]
-{\large Additional Contributors:}\\[2ex]
-Bernd Bra\ss{}el$^2$ \\
-Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
-\normalsize
-(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
-(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
-(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
-\end{center}
-\end{minipage}}
-\end{center}
-\end{titlepage}
-
-\pdfbookmark[1]{Contents}{Contents}
-\tableofcontents
-
-\newpage
-
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current \CYS implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, \CYS also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
-Therefore, this document only explains the use of the different
-components of \CYS
-and the differences and restrictions of \CYS
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of \CYS
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of \CYS}
-
-\subsection{Installation}
-\label{sec-install}
-
-This version of \CYS has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems.
-
-Installation instructions for \CYS can be found in
-the file \code{INSTALL.txt} stored in the \CYS installation directory.
-Note that there are two possibilities to install \CYS:
-\begin{description}
-\item[Global installation:]\index{global installation}\index{installation!global}
-\CYS is installed in some global system directory where users
-have no write permission. In this case, some options for experimenting
-with \CYS (like \code{supply} or \code{ghc}, see below)
-are not available (since they require the recompilation of parts of
-the installed system).
-\item[Local installation:]\index{local installation}\index{installation!local}
-\CYS is installed in some local user directory where the user
-has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the \CYS installation is set as follows:
-\begin{lstlisting}
-GLOBALINSTALL=no
-\end{lstlisting}
-In this case, all options of \CYS are available.
-\end{description}
-In the following, \cyshome denotes the installation directory
-of the \CYS installation.
-
-\subsection{General Use}
-\label{sec-general}
-
-All executables required to use the different components
-of \CYS are stored in the directory \code{\cyshome/bin}.
-You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=\cyshome/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with \CYS creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\begin{curry}
-cleancurry
-\end{curry}
-(this is a shell script stored in the \code{bin} directory of the
-\CYS installation, see above).
-The command
-\begin{curry}
-cleancurry -r
-\end{curry}
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with \CYS:
-\begin{itemize}
-\item
-\index{singleton variables}\index{variables!singleton}
-\emph{Singleton pattern variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-\CYS translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-In general, tabulator stops should be avoided in source programs.
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
-based on the suspension of expressions
-containing free variables are not yet supported.
-\CYS supports \emph{value generators} for free variables
-so that a free variable is instantiated when its value is demanded.
-For instance, the initial expression
-\begin{curry}
-x == True where x free
-\end{curry}
-is non-deterministically evaluated to \code{False} and \code{True}
-by instantiating \code{x} to \code{False} and \code{True}, respectively.
-Thus, a computation is never suspended due to free variables.
-This behavior also applies to free variables of primitive types
-like integers. For instance, the initial expression
-\begin{curry}
-x*y=:=1 where x,y free
-\end{curry}
-is non-deterministically evaluated to the two solutions
-\begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
-\end{curry}
-\item
-Unification is performed without an occur check.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in \CYS}
-\label{sec-modules}
-
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-\CYS searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a \CYS session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of \CYS}
-\label{sec-kics2}
-
-This section describes the interactive environment
-\CYS\index{\CYS}
-that supports the development of applications written in Curry.
-The implementation of \CYS contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking \CYS}
-\label{sec:invoke-kics2}
-
-To start \CYS, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{\cyshome/bin} where \cyshome is the installation directory
-of \CYS).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{\cyshome/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke \CYS with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\begin{curry}
-kics2 :load Mod :add List
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\begin{curry}
-kics2 :load Mod :eval config
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\begin{curry}
-kics2 :load Mod :save :quit
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, creates an executable,
-and terminates \CYS. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of \CYS}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of \CYS are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs.
-In order to see the results of their bindings,\footnote{Currently,
-bindings are only printed if the initial expression is not an I/O action
-(i.e., not of type \ccode{IO\ldots})
-and there are not more than ten free variables
-in the initial expression.}
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
-\end{curry}
-in order to obtain the following three possible bindings:
-\begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
-\end{curry}
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\begin{curry}
-xs++ys=:=[z] where xs,ys,z free
-\end{curry}
-they are specialized to the type \ccode{()}
-(since the current implementation of \CYS does not support
-computations with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of \CYS.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the \CYS environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current \CYS process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Options of \CYS}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of \CYS.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices $n$}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression
-in a tree-like structure.
-This mode is only useful for debugging or
-understanding the implementation of non-deterministic
-evaluations used in \CYS.
-If the optional argument $n$ is provided,
-the tree is shown up to depth $n$.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-(not available in global installations, see Section~\ref{sec-install})
-Use implementation $i$ as the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the following values for $i$ are supported:
-\begin{description}
-\item[\code{integer}:]\pindex{integer}
-Use unbounded integers as choice identifiers.
-This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc}:]\pindex{ghc}
-Use a more sophisticated implementation of choice
-identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
-provided by the Glasgow Haskell Compiler.
-\item[\code{pureio}:]\pindex{pureio}
-Use IO references (i.e., memory cells) for choice identifiers.
-This is the most efficient implementation for top-level depth-first search
-but cannot be used for more sophisticated search methods
-like encapsulated search.
-\item[\code{ioref} (default):]\pindex{ioref}
-Use a mixture of \code{ghc} and \code{pureio}.
-IO references are used for top-level depth-first search
-and \code{ghc} identifiers are used for encapsulated search methods.
-\end{description}
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-Thus, one has also the possibility to terminate the
-enumeration of all values after having seen some values.
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the \CYS compiler.
-For instance, setting the option
-\begin{curry}
-:set cmp -O 0
-\end{curry}
-has the effect that all optimizations performed by the \CYS compiler
-are turned off.
-
-\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-Define additional options passed to the Glasgow Haskell Compiler (GHC)
-when the generated Haskell programs are compiled.
-One has to be careful when using such options.
-For instance, in a global installation of \CYS
-(see Section~\ref{sec-install}), libraries are pre-compiled
-so that inconsistencies might occur if compilation options
-might be changed.
-
-It is safe to pass specific GHC linking options.
-For instance, to enforce the static linking of libraries
-in order to generate an executable (see command \ccode{:save})
-that can be executed in another environment, one could set the
-options
-\begin{curry}
-:set ghc -static -optl-static -optl-pthread
-\end{curry}
-
-Other options are useful for experimental purposes,
-but those should be used only in local installations
-(see Section~\ref{sec-install}) to avoid inconsistent
-target codes for different libraries.
-For instance, setting the option
-\begin{curry}
-:set ghc -DDISABLE_CS
-\end{curry}
-has the effect that the constraint store used to enable
-an efficient access to complex bindings is disabled.
-Similarly,
-\begin{curry}
-:set ghc -DSTRICT_VAL_BIND
-\end{curry}
-has the effect that expressions in a unification constraint
-(\code{=:=}) are always fully evaluated
-(instead of the evaluation to a head normal form only)
-before unifying both sides.
-Since these options influence the compilation of the run-time system,
-one should also enforce the recompilation of Haskell programs
-by the GHC option \ccode{-fforce-recomp}, e.g., one should set
-\begin{curry}
-:set ghc -DDISABLE_CS -fforce-recomp
-\end{curry}
-
-\item[\fbox{\code{rts $opts$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\begin{curry}
-:set rts -H512m
-\end{curry}
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\item[\fbox{\code{args $arguments$}}]\pindex{args}
-Define run-time arguments passed to the executable
-generated by the Glasgow Haskell Compiler.
-For instance, setting the option
-\begin{curry}
-:set args first second
-\end{curry}
-has the effect that the I/O operation \code{getArgs}
-(see library \code{System} (Section~\ref{Library:System})
-returns the value \code{["first","second"]}.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-\CYS requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\begin{curry}
-{-# KiCS2_OPTION $opt$ #-}
-\end{curry}
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the \CYS command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\begin{curry}
-{-# KiCS2_OPTION ids #-}
-{-# KiCS2_OPTION +ghci #-}
-{-# KiCS2_OPTION v2 #-}
-module M where
-$\ldots$
-\end{curry}
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of \CYS (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call \CYS with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of \CYS to your own preferences,
-there is a configuration file which is read by \CYS when it is invoked.
-When you start \CYS for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started \CYS for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of \CYS contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into \CYS
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{\cyshome/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
-\section{Extensions}
-\label{sec-extensions}
-
-\CYS supports some extensions in Curry programs that are not (yet)
-part of the definition of Curry. These extensions are described below.
-
-\subsection{Recursive Variable Bindings}
-
-Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
-For instance, the declaration
-\begin{curry}
-ones5 = let ones = 1 : ones
-         in take 5 ones
-\end{curry}
-introduces the local variable \code{ones} which is bound
-to a \emph{cyclic structure}\index{cyclic structure}
-representing an infinite list of \code{1}'s.
-Similarly, the definition
-\begin{curry}
-onetwo n = take n one2
- where
-   one2 = 1 : two1
-   two1 = 2 : one2
-\end{curry}
-introduces a local variables \code{one2} that represents
-an infinite list of alternating \code{1}'s and \code{2}'s
-so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.
-
-\subsection{Functional Patterns}
-
-Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
-defining operations with functional patterns avoids problems
-caused by strict equality (\ccode{=:=}) and leads to programs
-that are potentially more efficient.
-
-Consider the definition of an operation to compute the last element
-of a list \code{xs} based on the prelude operation \ccode{++}
-for list concatenation:
-\begin{curry}
-last xs | _++[y] =:= xs  = y   where y free
-\end{curry}
-Since the equality constraint \ccode{=:=} evaluates both sides
-to a constructor term, all elements of the list \code{xs} are
-fully evaluated in order to satisfy the constraint.
-
-Functional patterns can help to improve this computational behavior.
-A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
-is a function call at a pattern position. With functional patterns,
-we can define the operation \code{last} as follows:
-\begin{curry}
-last (_++[y]) = y
-\end{curry}
-This definition is not only more compact but also avoids the complete
-evaluation of the list elements: since a functional pattern is considered
-as an abbreviation for the set of constructor terms obtained by all
-evaluations of the functional pattern to normal form (see
-\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
-definition is conceptually equivalent to the set of rules
-\begin{curry}
-last [y] = y
-last [_,y] = y
-last [_,_,y] = y
-$\ldots$
-\end{curry}
-which shows that the evaluation of the list elements is not demanded
-by the functional pattern.
-
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
-is interpreted as a functional pattern if $f$ is not a visible constructor
-but a defined function that is visible in the scope of the pattern.
-
-It is also possible to combine functional patterns with
-as-patterns.\index{as-pattern}\pindex{"@}
-Similarly to the meaning of as-patterns
-in standard constructor patterns,
-as-patterns in functional patterns are interpreted
-as a sequence of pattern matching where the variable of the as-pattern
-is matched before the given pattern is matched.
-This process can be described by introducing an auxiliary operation
-for this two-level pattern matching process.
-For instance, the definition
-\begin{curry}
-f (_ ++ x@[(42,_)] ++ _) = x
-\end{curry}
-is considered as syntactic sugar for the expanded definition
-\begin{curry}
-f (_ ++ x ++ _) = f' x
- where
-  f' [(42,_)] = x
-\end{curry}
-However, as-patterns are usually implemented
-in a more efficient way without introducing auxiliary operations.
-
-\subsection{Order of Pattern Matching}
-
-Curry allows multiple occurrences of pattern variables
-in standard patterns. These are an abbreviation of equational constraints
-between pattern variables.
-Functional patterns might also contain multiple occurrences of
-pattern variables.
-For instance, the operation
-\begin{curry}
-f (_++[x]++_++[x]++_) = x
-\end{curry}
-returns all elements with at least two occurrences in a list.
-
-If functional patterns as well as multiple occurrences of
-pattern variables occur in a pattern defining an operation,
-there are various orders to match an expression against such
-an operation. In the current implementation, the order
-is as follows:
-\begin{enumerate}
-\item Standard pattern matching: First, it is checked whether
-the constructor patterns match. Thus, functional patterns
-and multiple occurrences of pattern variables are ignored.
-\item Functional pattern matching: In the next phase,
-functional patterns are matched but occurrences of standard
-pattern variables in the functional patterns are ignored.
-\item Non-linear patterns: If standard and functional pattern matching
-is successful, the equational constraints which correspond
-to multiple occurrences pattern variables are solved.
-\item Guards: Finally, the guards supplied by the programmer
-are checked.
-\end{enumerate}
-The order of pattern matching should not influence the computed
-result. However, it might have some influence on the termination
-behavior of programs, i.e., a program might not terminate
-instead of finitely failing.
-In such cases, it could be necessary to consider the influence
-of the order of pattern matching. Note that other orders of pattern matching
-can be obtained using auxiliary operations.
-
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are 
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there. 
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\begin{curry}
-type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
-\end{curry}
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$ 
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\begin{curry}
-type Person = {name :: String, age :: Int}
-type Address = {person :: Person, street :: String, city :: String}
-type Branch a b = {left :: a, right :: b}
-\end{curry}
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\begin{curry}
-type Address = {person :: Person, street,city :: String}
-\end{curry}
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\begin{curry}
-type Address = {street,city :: String, person :: Person}
-\end{curry}
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\begin{curry}
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\end{curry}
-\begin{curry}
-getName :: Person -> String
-getName $\ldots$
-\end{curry}
-
-Labels can only be used in the context of
-records. They do not share the name space with 
-functions/constructors/variables or type constructors/type variables. 
-For instance it is possible to use 
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually 
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\begin{curry}
-{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
-\end{curry}
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with 
-exactly one assignment. Examples for record constructions are
-\begin{curry}
-{name := "Johnson", age := 30}     -- generates a record of type 'Person'
-{left := True, right := 20}        -- generates a record of type 'Branch'
-\end{curry}
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\begin{curry}
-{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
-\end{curry}
-Unlike labeled fields in record type declarations, 
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\begin{curry}
-{person := {name := "Smith", age := 20},   -- generates a record of
- street := "Main Street",                  -- type 'Address'
- city   := "Springfield"}
-\end{curry}
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records. 
-It has the following form:
-\begin{curry}
-$r$ :> $l$
-\end{curry}
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\begin{curry}
-pers :> name
-\end{curry}
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\begin{curry}
-addr :> person :> age
-\end{curry}
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}).
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\begin{curry}
-{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
-\end{curry}
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration 
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record. 
-Assignments can occur in an arbitrary order. It is not allowed to 
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\begin{curry}
-{name := "Scott", age := 25 | pers}
-{person := {name := "Scott", age := 25 | pers} | addr}
-\end{curry}
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\begin{curry}
-{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
-{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
-\end{curry}
-In both cases each label $l_i$ is specified with a pattern $p_i$. 
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern. 
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the 
-patterns of the specified labels are matched against 
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\begin{curry}
-isSmith30 :: Person -> Bool
-isSmith30 {name = "Smith", age = 30} = True
-\end{curry}
-\begin{curry}
-startsWith :: Char -> Person -> Bool
-startsWith c {name = (d:_) | _} = c == d
-\end{curry}
-\begin{curry}
-getPerson :: Address -> Person
-getPerson {person = p | _} = p
-\end{curry}
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways 
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent 
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
-
-\newpage
-
-\input{currydoc}
-
-\newpage
-
 \section{\cb: A Tool for Analyzing and Browsing Curry Programs}
@@
 are available through the ``Help'' menu of \cb.
-
-\newpage
-
-\section{CurryTest: A Tool for Testing Curry Programs}
-\label{sec-currytest}
-
-CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the \CYS distribution to write
-and run repeatable tests. CurryTest simplifies the task
-of writing test cases for a module and executing them.
-The tool is easy to use. Assume one has implemented a module \code{MyMod}
-and wants to write some test cases to test its functionality,
-making regression tests in future versions, etc.
-For this purpose, there is a system library \code{Assertion}
-(Section~\ref{Library:Assertion}) which
-contains the necessary definitions for writing tests.
-In particular, it exports an abstract polymorphic type \ccode{Assertion a}
-together with the following operations:
-\begin{curry}
-assertTrue      :: String -> Bool -> Assertion ()
-assertEqual     :: String -> a -> a -> Assertion a
-assertValues    :: String -> a -> [a] -> Assertion a
-assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
-assertIO        :: String -> IO a -> a -> Assertion a
-assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\end{curry}
-The expression \ccode{assertTrue $s$ $b$}
-is an assertion (named $s$) that the expression $b$ has the value \code{True}.
-Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
-asserts that the expressions $e_1$ and $e_2$
-must be equal (i.e., \code{$e_1$==$e_2$} must hold),
-the expression \ccode{assertValues $s$ $e$ $vs$} asserts
-that $vs$ is the multiset of all values of $e$,
-and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
-that the constraint abstraction $c$ has the multiset of solutions $vs$.
-Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
-asserts that the I/O action $a$ yields the value $v$ whenever it is
-executed, and
-the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yield equal values.
-The name $s$ provided as a first argument in each assertion
-is used in the protocol produced by the test tool.
-
-One can define a test program by importing the module
-to be tested together with the module \code{Assertion} and defining
-top-level functions of type \code{Assertion} in this module
-(which must also be exported).
-As an example, consider the following program
-that can be used to test some list processing functions:
-\begin{curry}
-import List
-import Assertion
-
-test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-
-test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-
-test3 = assertSolutions "prefix" (\x -> x++_ =:= [1,2])
-                                 [[],[1],[1,2]]
-\end{curry}
-For instance, \code{test1} asserts that the result of evaluating the
-expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.
-
-We can execute a test suite by the command\pindex{currytest}
-\begin{curry}
-currytest TestList
-\end{curry}
-(\code{currytest} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-In our example, \ccode{TestList.curry} is the program containing the
-definition of all assertions. This has the effect
-that all exported top-level functions
-of type \code{Assertion} are tested (i.e., the corresponding
-assertions are checked) and the results
-(\ccode{OK} or failure) are reported together with the name of each assertion.
-%If failures occur, the complete test results are also
-%written into a file named \ccode{TestList.testlog}.''
-For our example above, we obtain the following successful protocol:
-\begin{curry}
-============================================================
-Testing module "TestList"...
-OK: ++
-OK: all
-OK: prefix
-All tests successfully passed.
-============================================================
-\end{curry}
-There is also a graphical interface that summarizes the results
-more nicely.
-In order to start this interface, one has to add the parameter
-\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\begin{curry}
-currytest --window TestList
-\end{curry}
-or
-\begin{curry}
-currytest -w TestList
-\end{curry}
-A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
-
-\begin{figure}%[t]
-\begin{center}
-\includegraphics[scale=0.7]{currytest.jpg}
-\end{center}
-\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
-\end{figure}
-
-\newpage
-
-\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
-\label{sec-erd2curry}
-
-ERD2Curry\index{ERD2Curry}\index{database programming}
-is a tool to generate Curry code to access and manipulate data
-persistently stored from
-entity relationship diagrams.\index{entity relationship diagrams}
-The idea of this tool is described in detail in
-\cite{BrasselHanusMueller08PADL}.
-Thus, we describe only the basic steps to use this tool
-in the following.
-
-If one creates an entity relationship diagram (ERD)
-with the Umbrello UML Modeller, one has to store its
-XML description in XMI format (as offered by Umbrello)
-in a file, e.g., \ccode{myerd.xmi}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry myerd.xmi
-\end{curry}
-(\code{erd2curry} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-If \code{MyData} is the name of the ERD, the Curry program file
-\ccode{MyData.curry} is generated containing all the necessary
-database access code as described in \cite{BrasselHanusMueller08PADL}.
-
-If one does not want to use the Umbrello UML Modeller,
-one can also create a textual description of the ERD
-as a Curry term of type \code{ERD}
-(w.r.t.\ the type definition given in module
-\code{\cyshome/tools/erd2curry/ERD.curry})
-and store it in some file, e.g., \ccode{myerd.term}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry -t myerd.term
-\end{curry}
-%
-There is also the possibility to visualize an ERD term
-as a graph with the graph visualization program \code{dotty}
-(for this purpose, it might be necessary to adapt the definition
-of \code{dotviewcommand} in your \ccode{.kics2rc} file,
-see Section~\ref{sec-customization},
-according to your local environment).
-This can be done by the command
-\begin{curry}
-erd2curry -v myerd.term
-\end{curry}
-
-\paragraph{Inclusion in the Curry application:}
-To compile the generated database code, either
-include the directory \code{\cyshome/tools/erd2curry}
-into your Curry load path
-(e.g., by setting  the environment variable
-\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
-or copy the file
-\code{\cyshome/tools/erd2curry/ERDGeneric.curry}
-into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with \CYS
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of \CYS and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of \CYS:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{\cyshome/cpns/start} and
-\code{\cyshome/cpns/stop}.
-If the demon is already running, the command \code{\cyshome/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
-\addcontentsline{toc}{section}{Bibliography}
-\bibliography{mh}
-\bibliographystyle{plain}
-
-\newpage
-\appendix
-
-\section{Libraries of the \CYS Distribution}
-\label{sec:libraries}
-
-{\setlength{\parindent}{0.0cm}
-
-The \CYS distribution comes with an extensive collection
-of libraries for application programming.
-The libraries for meta-programming by representing
-Curry programs as datatypes in Curry are described
-in the following subsection in more detail.
-The complete set of libraries with all exported types and functions
-are described in the further subsections.
-For a more detailed online documentation of all libraries of \CYS,
-see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
-
-\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
-\label{sec-flatcurry}
-
-\index{AbstractCurry}
-\index{FlatCurry}
-To support meta-programming, i.e., the manipulation of Curry programs
-in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{\cyshome/lib/meta},
-which define datatypes for the representation
-of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
-where local function definitions are replaced by global definitions
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
-program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
-programs that are more oriented towards the source program.
-
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
-These actions parse the corresponding source program and return
-a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
-
-Since all datatypes are explained in detail in these modules,
-we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
-of these modules.
-
-As an example, consider a program file \ccode{test.curry}
-containing the following two lines:
-\begin{curry}
-rev []     = []
-rev (x:xs) = (rev xs) ++ [x]
-\end{curry}
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
-following term:
-\begin{curry}
- (Prog "test"
-  ["Prelude"]
-  []
-  [Func ("test","rev") 1 Public
-        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
-                  (TCons ("Prelude","[]") [(TVar 0)]))
-        (Rule [0]
-           (Case Flex (Var 1)
-              [Branch (Pattern ("Prelude","[]") [])
-                  (Comb ConsCall ("Prelude","[]") []),
-               Branch (Pattern ("Prelude",":") [2,3])
-                  (Comb FuncCall ("Prelude","++")
-                        [Comb FuncCall ("test","rev") [Var 3],
-                         Comb ConsCall ("Prelude",":")
-                              [Var 2,Comb ConsCall ("Prelude","[]") []]
-                        ])
-              ]))]
-  []
- )
-\end{curry}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Definitions in order to LaTeX documents generated by "currydoc --tex"
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
-\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
-\newcommand{\currytypesstop}{}
-\newcommand{\currytypesynstart}[2]{{\tt type #2}\pindex{#1} \begin{quote}}
-\newcommand{\currytypesynstop}{\end{quote}}
-\newcommand{\currydatastart}[1]{{\tt data #1}\pindex{#1} \begin{quote}}
-\newcommand{\currydatacons}{\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:}}
-\newcommand{\currydatastop}{\end{itemize}}
-\newcommand{\curryconsstart}[2]{\item {\tt #1~::~#2}\par}
-\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
-\newcommand{\curryfuncstop}{}
-\newcommand{\curryfunctionstart}[2]{#2\pindex{#1}\begin{quote}}
-\newcommand{\curryfunctionstop}{\end{quote}}
-\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
-
-\subsection{General Libraries}
-
-\input{lib/AllSolutions}
-\input{lib/Assertion}
-\input{lib/Char}
-\input{lib/Combinatorial}
-\input{lib/Constraint}
-\input{lib/CSV}
-\input{lib/Directory}
-\input{lib/FileGoodies}
-\input{lib/Float}
-\input{lib/Global}
-\input{lib/GUI}
-\input{lib/Integer}
-\input{lib/IO}
-\input{lib/IOExts}
-\input{lib/JavaScript}
-\input{lib/KeyDatabaseSQLite}
-\input{lib/List}
-\input{lib/Maybe}
-\input{lib/NamedSocket}
-\input{lib/Parser}
-\input{lib/Pretty}
-\input{lib/Profile}
-\input{lib/PropertyFile}
-\input{lib/Read}
-\input{lib/ReadNumeric}
-\input{lib/ReadShowTerm}
-\input{lib/SetFunctions}
-\input{lib/SearchTree}
-\input{lib/Socket}
-\input{lib/System}
-\input{lib/Time}
-\input{lib/Unsafe}
-
-\subsection{Data Structures and Algorithms}
-
-\input{lib/Array}
-\input{lib/Dequeue}
-\input{lib/FiniteMap}
-\input{lib/GraphInductive}
-\input{lib/Random}
-\input{lib/RedBlackTree}
-\input{lib/SetRBT}
-\input{lib/Sort}
-\input{lib/TableRBT}
-\input{lib/Traversal}
-
-\subsection{Libraries for Web Applications}
-
-\input{lib/CategorizedHtmlList}
-\input{lib/HTML}
-\input{lib/HtmlParser}
-\input{lib/Mail}
-\input{lib/Markdown}
-\input{lib/WUI}
-\input{lib/URL}
-\input{lib/XML}
-\input{lib/XmlConv}
-
-\subsection{Libraries for Meta-Programming}
-
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
-\input{lib/CompactFlatCurry}
-\input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
-
-} % end setlength parindent
-
-\newpage
-
-\input{markdown_syntax}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with \CYS,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of \CYS create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with \CYS.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with \CYS
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-\CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with \CYS
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, \CYS has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Thus, if an external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\begin{curry}
-f external
-\end{curry}
-below the type declaration for the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the \CYS compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, \CYS follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\begin{curry}
-sconc :: String -> String -> String
-\end{curry}
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal form.
-This is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\begin{currynomath}
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $## s1) $## s2
-
-prim_sconc :: String -> String -> String
-prim_sconc external
-\end{currynomath}
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of \CYS.
-The \CYS compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\begin{curry}
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> ConstStore
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
-\end{curry}
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the programmer.
-The additional argument of type \code{ConstStore}
-represents the current set of constraints when this
-operation is called. This argument is intended to provide
-a more efficient access to binding constraints and can be
-ignored in standard operations.
-
-If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The \CYS compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\begin{curry}
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> ConstStore -> CP.OP_List CP.C_Char
-\end{curry}
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of \CYS
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-(note that we simply ignore the final argument representing the
-constraint store)
-\begin{curry}
-external_d_C_prim_sconc s1 s2 _ =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\end{curry}
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the \CYS implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
-For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\begin{currynomath}
-printString :: String -> IO ()
-printString s = prim_printString $## s
-
-prim_printString :: String -> IO ()
-prim_printString external
-\end{currynomath} % $
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
-\begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
-                              -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-\end{curry}
-When we compile the Curry module \code{String},
-\CYS combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, \CYS transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\begin{curry}
-(<&>) :: Int -> Int -> Int
-(<&>) external
-\end{curry}
-the generated Haskell module contains the code
-\begin{curry}
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> ConstStore -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
-\end{curry}
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
-\addcontentsline{toc}{section}{Index}
-\printindex
-
-\end{document}
-
-% LocalWords:  KiCS

### 2013-07-12 docs/src/currytest.tex
Split manual into separate tex files

--- 2013-03-07 docs/src/manual.tex
@@
-\documentclass[11pt,fleqn]{article}
-
-\usepackage{latexsym}
-\usepackage{makeidx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{graphicx}
-
-\input{version}
-
-%%% ------------------------------------------------------------------
-
-\usepackage[colorlinks,linkcolor=blue]{hyperref}
-\hypersetup{bookmarksopen=true}
-\hypersetup{bookmarksopenlevel=0}
-\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
-\hypersetup{pdfauthor={Michael Hanus}}
-%\hypersetup{pdfstartview=Title}
-\hypersetup{pdfstartview=FitH}
-\usepackage{thumbpdf}
-
-%%% ------------------------------------------------------------------
-
-\setlength{\textwidth}{16.5cm}
-\setlength{\textheight}{23cm}
-\renewcommand{\baselinestretch}{1.1}
-\setlength{\topmargin}{-1cm}
-\setlength{\oddsidemargin}{0cm}
-\setlength{\evensidemargin}{0cm}
-\setlength{\marginparwidth}{0.0cm}
-\setlength{\marginparsep}{0.0cm}
-
-\newlength{\figurewidth}
-\setlength{\figurewidth}{\textwidth}
-\addtolength{\figurewidth}{-0.4cm}
-
-% font for program texts
-\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\small\tt}
-
-\usepackage{listings}
-\lstset{aboveskip=1.5ex,
-        belowskip=1.5ex,
-        showstringspaces=false, % no special string space
-        mathescape=true,
-        flexiblecolumns=false,
-        xleftmargin=2ex,
-        basewidth=0.52em,
-        basicstyle=\small\ttfamily}
-
-\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
-       }
-
-\lstnewenvironment{curry}{}{}
-\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
-\newcommand{\listline}{\vrule width0pt depth1.75ex}
-
-% program text in normal text
-\newcommand{\code}[1]{\mbox{\codefont #1}}
-
-% program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\code{#1}''}
-
-\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index
-
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
-\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
-\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
-\newcommand{\cb}{CurryBrowser\xspace}
-
-% allow underscores in programs:
-\catcode`\_=\active
-\let_=\sb
-\catcode`\_=12
-
-% produce an index:
-\makeindex
-
-\begin{document}
-\sloppy
-
-\begin{titlepage}
-\pdfbookmark[1]{Title}{Title}
-\begin{center}
-\fbox{
-\begin{minipage}[t]{\figurewidth}
-\begin{center}\vspace{10ex}
-{\Huge\bf KiCS2}\\[4ex]
-{\huge The Kiel Curry System (Version 2)}\\[7ex]
-{\huge User Manual}\\[4ex]
-\kicsversiondate\\[6ex]
-\Large
-Michael Hanus$^1$ [editor] \\[3ex]
-{\large Additional Contributors:}\\[2ex]
-Bernd Bra\ss{}el$^2$ \\
-Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
-\normalsize
-(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
-(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
-(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
-\end{center}
-\end{minipage}}
-\end{center}
-\end{titlepage}
-
-\pdfbookmark[1]{Contents}{Contents}
-\tableofcontents
-
-\newpage
-
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current \CYS implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, \CYS also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
-Therefore, this document only explains the use of the different
-components of \CYS
-and the differences and restrictions of \CYS
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of \CYS
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of \CYS}
-
-\subsection{Installation}
-\label{sec-install}
-
-This version of \CYS has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems.
-
-Installation instructions for \CYS can be found in
-the file \code{INSTALL.txt} stored in the \CYS installation directory.
-Note that there are two possibilities to install \CYS:
-\begin{description}
-\item[Global installation:]\index{global installation}\index{installation!global}
-\CYS is installed in some global system directory where users
-have no write permission. In this case, some options for experimenting
-with \CYS (like \code{supply} or \code{ghc}, see below)
-are not available (since they require the recompilation of parts of
-the installed system).
-\item[Local installation:]\index{local installation}\index{installation!local}
-\CYS is installed in some local user directory where the user
-has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the \CYS installation is set as follows:
-\begin{lstlisting}
-GLOBALINSTALL=no
-\end{lstlisting}
-In this case, all options of \CYS are available.
-\end{description}
-In the following, \cyshome denotes the installation directory
-of the \CYS installation.
-
-\subsection{General Use}
-\label{sec-general}
-
-All executables required to use the different components
-of \CYS are stored in the directory \code{\cyshome/bin}.
-You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=\cyshome/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with \CYS creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\begin{curry}
-cleancurry
-\end{curry}
-(this is a shell script stored in the \code{bin} directory of the
-\CYS installation, see above).
-The command
-\begin{curry}
-cleancurry -r
-\end{curry}
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with \CYS:
-\begin{itemize}
-\item
-\index{singleton variables}\index{variables!singleton}
-\emph{Singleton pattern variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-\CYS translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-In general, tabulator stops should be avoided in source programs.
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
-based on the suspension of expressions
-containing free variables are not yet supported.
-\CYS supports \emph{value generators} for free variables
-so that a free variable is instantiated when its value is demanded.
-For instance, the initial expression
-\begin{curry}
-x == True where x free
-\end{curry}
-is non-deterministically evaluated to \code{False} and \code{True}
-by instantiating \code{x} to \code{False} and \code{True}, respectively.
-Thus, a computation is never suspended due to free variables.
-This behavior also applies to free variables of primitive types
-like integers. For instance, the initial expression
-\begin{curry}
-x*y=:=1 where x,y free
-\end{curry}
-is non-deterministically evaluated to the two solutions
-\begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
-\end{curry}
-\item
-Unification is performed without an occur check.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in \CYS}
-\label{sec-modules}
-
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-\CYS searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a \CYS session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of \CYS}
-\label{sec-kics2}
-
-This section describes the interactive environment
-\CYS\index{\CYS}
-that supports the development of applications written in Curry.
-The implementation of \CYS contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking \CYS}
-\label{sec:invoke-kics2}
-
-To start \CYS, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{\cyshome/bin} where \cyshome is the installation directory
-of \CYS).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{\cyshome/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke \CYS with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\begin{curry}
-kics2 :load Mod :add List
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\begin{curry}
-kics2 :load Mod :eval config
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\begin{curry}
-kics2 :load Mod :save :quit
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, creates an executable,
-and terminates \CYS. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of \CYS}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of \CYS are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs.
-In order to see the results of their bindings,\footnote{Currently,
-bindings are only printed if the initial expression is not an I/O action
-(i.e., not of type \ccode{IO\ldots})
-and there are not more than ten free variables
-in the initial expression.}
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
-\end{curry}
-in order to obtain the following three possible bindings:
-\begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
-\end{curry}
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\begin{curry}
-xs++ys=:=[z] where xs,ys,z free
-\end{curry}
-they are specialized to the type \ccode{()}
-(since the current implementation of \CYS does not support
-computations with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of \CYS.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the \CYS environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current \CYS process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Options of \CYS}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of \CYS.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices $n$}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression
-in a tree-like structure.
-This mode is only useful for debugging or
-understanding the implementation of non-deterministic
-evaluations used in \CYS.
-If the optional argument $n$ is provided,
-the tree is shown up to depth $n$.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-(not available in global installations, see Section~\ref{sec-install})
-Use implementation $i$ as the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the following values for $i$ are supported:
-\begin{description}
-\item[\code{integer}:]\pindex{integer}
-Use unbounded integers as choice identifiers.
-This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc}:]\pindex{ghc}
-Use a more sophisticated implementation of choice
-identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
-provided by the Glasgow Haskell Compiler.
-\item[\code{pureio}:]\pindex{pureio}
-Use IO references (i.e., memory cells) for choice identifiers.
-This is the most efficient implementation for top-level depth-first search
-but cannot be used for more sophisticated search methods
-like encapsulated search.
-\item[\code{ioref} (default):]\pindex{ioref}
-Use a mixture of \code{ghc} and \code{pureio}.
-IO references are used for top-level depth-first search
-and \code{ghc} identifiers are used for encapsulated search methods.
-\end{description}
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-Thus, one has also the possibility to terminate the
-enumeration of all values after having seen some values.
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the \CYS compiler.
-For instance, setting the option
-\begin{curry}
-:set cmp -O 0
-\end{curry}
-has the effect that all optimizations performed by the \CYS compiler
-are turned off.
-
-\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-Define additional options passed to the Glasgow Haskell Compiler (GHC)
-when the generated Haskell programs are compiled.
-One has to be careful when using such options.
-For instance, in a global installation of \CYS
-(see Section~\ref{sec-install}), libraries are pre-compiled
-so that inconsistencies might occur if compilation options
-might be changed.
-
-It is safe to pass specific GHC linking options.
-For instance, to enforce the static linking of libraries
-in order to generate an executable (see command \ccode{:save})
-that can be executed in another environment, one could set the
-options
-\begin{curry}
-:set ghc -static -optl-static -optl-pthread
-\end{curry}
-
-Other options are useful for experimental purposes,
-but those should be used only in local installations
-(see Section~\ref{sec-install}) to avoid inconsistent
-target codes for different libraries.
-For instance, setting the option
-\begin{curry}
-:set ghc -DDISABLE_CS
-\end{curry}
-has the effect that the constraint store used to enable
-an efficient access to complex bindings is disabled.
-Similarly,
-\begin{curry}
-:set ghc -DSTRICT_VAL_BIND
-\end{curry}
-has the effect that expressions in a unification constraint
-(\code{=:=}) are always fully evaluated
-(instead of the evaluation to a head normal form only)
-before unifying both sides.
-Since these options influence the compilation of the run-time system,
-one should also enforce the recompilation of Haskell programs
-by the GHC option \ccode{-fforce-recomp}, e.g., one should set
-\begin{curry}
-:set ghc -DDISABLE_CS -fforce-recomp
-\end{curry}
-
-\item[\fbox{\code{rts $opts$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\begin{curry}
-:set rts -H512m
-\end{curry}
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\item[\fbox{\code{args $arguments$}}]\pindex{args}
-Define run-time arguments passed to the executable
-generated by the Glasgow Haskell Compiler.
-For instance, setting the option
-\begin{curry}
-:set args first second
-\end{curry}
-has the effect that the I/O operation \code{getArgs}
-(see library \code{System} (Section~\ref{Library:System})
-returns the value \code{["first","second"]}.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-\CYS requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\begin{curry}
-{-# KiCS2_OPTION $opt$ #-}
-\end{curry}
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the \CYS command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\begin{curry}
-{-# KiCS2_OPTION ids #-}
-{-# KiCS2_OPTION +ghci #-}
-{-# KiCS2_OPTION v2 #-}
-module M where
-$\ldots$
-\end{curry}
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of \CYS (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call \CYS with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of \CYS to your own preferences,
-there is a configuration file which is read by \CYS when it is invoked.
-When you start \CYS for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started \CYS for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of \CYS contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into \CYS
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{\cyshome/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
-\section{Extensions}
-\label{sec-extensions}
-
-\CYS supports some extensions in Curry programs that are not (yet)
-part of the definition of Curry. These extensions are described below.
-
-\subsection{Recursive Variable Bindings}
-
-Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
-For instance, the declaration
-\begin{curry}
-ones5 = let ones = 1 : ones
-         in take 5 ones
-\end{curry}
-introduces the local variable \code{ones} which is bound
-to a \emph{cyclic structure}\index{cyclic structure}
-representing an infinite list of \code{1}'s.
-Similarly, the definition
-\begin{curry}
-onetwo n = take n one2
- where
-   one2 = 1 : two1
-   two1 = 2 : one2
-\end{curry}
-introduces a local variables \code{one2} that represents
-an infinite list of alternating \code{1}'s and \code{2}'s
-so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.
-
-\subsection{Functional Patterns}
-
-Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
-defining operations with functional patterns avoids problems
-caused by strict equality (\ccode{=:=}) and leads to programs
-that are potentially more efficient.
-
-Consider the definition of an operation to compute the last element
-of a list \code{xs} based on the prelude operation \ccode{++}
-for list concatenation:
-\begin{curry}
-last xs | _++[y] =:= xs  = y   where y free
-\end{curry}
-Since the equality constraint \ccode{=:=} evaluates both sides
-to a constructor term, all elements of the list \code{xs} are
-fully evaluated in order to satisfy the constraint.
-
-Functional patterns can help to improve this computational behavior.
-A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
-is a function call at a pattern position. With functional patterns,
-we can define the operation \code{last} as follows:
-\begin{curry}
-last (_++[y]) = y
-\end{curry}
-This definition is not only more compact but also avoids the complete
-evaluation of the list elements: since a functional pattern is considered
-as an abbreviation for the set of constructor terms obtained by all
-evaluations of the functional pattern to normal form (see
-\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
-definition is conceptually equivalent to the set of rules
-\begin{curry}
-last [y] = y
-last [_,y] = y
-last [_,_,y] = y
-$\ldots$
-\end{curry}
-which shows that the evaluation of the list elements is not demanded
-by the functional pattern.
-
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
-is interpreted as a functional pattern if $f$ is not a visible constructor
-but a defined function that is visible in the scope of the pattern.
-
-It is also possible to combine functional patterns with
-as-patterns.\index{as-pattern}\pindex{"@}
-Similarly to the meaning of as-patterns
-in standard constructor patterns,
-as-patterns in functional patterns are interpreted
-as a sequence of pattern matching where the variable of the as-pattern
-is matched before the given pattern is matched.
-This process can be described by introducing an auxiliary operation
-for this two-level pattern matching process.
-For instance, the definition
-\begin{curry}
-f (_ ++ x@[(42,_)] ++ _) = x
-\end{curry}
-is considered as syntactic sugar for the expanded definition
-\begin{curry}
-f (_ ++ x ++ _) = f' x
- where
-  f' [(42,_)] = x
-\end{curry}
-However, as-patterns are usually implemented
-in a more efficient way without introducing auxiliary operations.
-
-\subsection{Order of Pattern Matching}
-
-Curry allows multiple occurrences of pattern variables
-in standard patterns. These are an abbreviation of equational constraints
-between pattern variables.
-Functional patterns might also contain multiple occurrences of
-pattern variables.
-For instance, the operation
-\begin{curry}
-f (_++[x]++_++[x]++_) = x
-\end{curry}
-returns all elements with at least two occurrences in a list.
-
-If functional patterns as well as multiple occurrences of
-pattern variables occur in a pattern defining an operation,
-there are various orders to match an expression against such
-an operation. In the current implementation, the order
-is as follows:
-\begin{enumerate}
-\item Standard pattern matching: First, it is checked whether
-the constructor patterns match. Thus, functional patterns
-and multiple occurrences of pattern variables are ignored.
-\item Functional pattern matching: In the next phase,
-functional patterns are matched but occurrences of standard
-pattern variables in the functional patterns are ignored.
-\item Non-linear patterns: If standard and functional pattern matching
-is successful, the equational constraints which correspond
-to multiple occurrences pattern variables are solved.
-\item Guards: Finally, the guards supplied by the programmer
-are checked.
-\end{enumerate}
-The order of pattern matching should not influence the computed
-result. However, it might have some influence on the termination
-behavior of programs, i.e., a program might not terminate
-instead of finitely failing.
-In such cases, it could be necessary to consider the influence
-of the order of pattern matching. Note that other orders of pattern matching
-can be obtained using auxiliary operations.
-
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are 
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there. 
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\begin{curry}
-type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
-\end{curry}
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$ 
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\begin{curry}
-type Person = {name :: String, age :: Int}
-type Address = {person :: Person, street :: String, city :: String}
-type Branch a b = {left :: a, right :: b}
-\end{curry}
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\begin{curry}
-type Address = {person :: Person, street,city :: String}
-\end{curry}
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\begin{curry}
-type Address = {street,city :: String, person :: Person}
-\end{curry}
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\begin{curry}
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\end{curry}
-\begin{curry}
-getName :: Person -> String
-getName $\ldots$
-\end{curry}
-
-Labels can only be used in the context of
-records. They do not share the name space with 
-functions/constructors/variables or type constructors/type variables. 
-For instance it is possible to use 
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually 
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\begin{curry}
-{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
-\end{curry}
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with 
-exactly one assignment. Examples for record constructions are
-\begin{curry}
-{name := "Johnson", age := 30}     -- generates a record of type 'Person'
-{left := True, right := 20}        -- generates a record of type 'Branch'
-\end{curry}
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\begin{curry}
-{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
-\end{curry}
-Unlike labeled fields in record type declarations, 
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\begin{curry}
-{person := {name := "Smith", age := 20},   -- generates a record of
- street := "Main Street",                  -- type 'Address'
- city   := "Springfield"}
-\end{curry}
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records. 
-It has the following form:
-\begin{curry}
-$r$ :> $l$
-\end{curry}
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\begin{curry}
-pers :> name
-\end{curry}
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\begin{curry}
-addr :> person :> age
-\end{curry}
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}).
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\begin{curry}
-{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
-\end{curry}
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration 
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record. 
-Assignments can occur in an arbitrary order. It is not allowed to 
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\begin{curry}
-{name := "Scott", age := 25 | pers}
-{person := {name := "Scott", age := 25 | pers} | addr}
-\end{curry}
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\begin{curry}
-{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
-{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
-\end{curry}
-In both cases each label $l_i$ is specified with a pattern $p_i$. 
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern. 
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the 
-patterns of the specified labels are matched against 
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\begin{curry}
-isSmith30 :: Person -> Bool
-isSmith30 {name = "Smith", age = 30} = True
-\end{curry}
-\begin{curry}
-startsWith :: Char -> Person -> Bool
-startsWith c {name = (d:_) | _} = c == d
-\end{curry}
-\begin{curry}
-getPerson :: Address -> Person
-getPerson {person = p | _} = p
-\end{curry}
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways 
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent 
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
-
-\newpage
-
-\input{currydoc}
-
-\newpage
-
-\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
-\label{sec-currybrowser}
-
-\cb is a tool to browse through the modules and functions
-of a Curry application, show them in various formats,
-and analyze their properties.\footnote{Although \cb is
-implemented in Curry, some functionalities of it require an
-installed graph visualization tool (dot \url{http://www.graphviz.org/}),
-otherwise they have no effect.}
-Moreover, it is constructed in a way so that
-new analyzers can be easily connected to \cb.
-A detailed description of the ideas behind this tool can be
-found in \cite{Hanus05WCFLP,Hanus06WLPE}.
-
-\cb is part of the \CYS distribution and can be
-started in two ways:
-\begin{itemize}
-\item
-In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
-\item
-In the \CYS environment after loading the module
-\code{mod} and typing the command \ccode{:browse}.
-\end{itemize}
-Here, \ccode{mod} is the name of the main module of a Curry application.
-After the start, \cb loads the interfaces of the main
-module and all imported modules before a GUI is created
-for interactive browsing.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.7]{currybrowser.jpg}
-\end{center}
-\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
-\end{figure}
-%
-To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
-shows a snapshot of its use on a particular application
-(here: the implementation of \cb).
-The upper list box in the left column shows the modules and their imports
-in order to browse through the modules of an application.
-Similarly to directory browsers, the list of imported modules of a module
-can be opened or closed by clicking.
-After selecting a module in the list of modules, its source code,
-interface, or various other formats of the module can be shown
-in the main (right) text area. For instance, one can show
-pretty-printed versions of the intermediate flat programs (see below)
-in order to see how local function definitions are translated by lambda lifting
-\cite{Johnsson85}
-or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
-Since Curry is a language with parametric polymorphism and type inference,
-programmers often omit the type signatures when defining functions.
-Therefore, one can also view (and store) the selected module as source code where
-missing type signatures are added.
-
-Below the list box for selecting modules, there is a menu
-(``Analyze selected module'') to analyze all functions
-of the currently selected module at once. This is useful
-to spot some functions of a module that could be problematic
-in some application contexts, like functions that are impure (i.e., the result
-depends on the evaluation time) or partially defined (i.e.,
-not evaluable on all ground terms).
-If such an analysis is selected,
-the names of all functions are shown in the
-lower list box of the left column (the ``function list'')
-with prefixes indicating the properties of the individual functions.
-
-The function list box can be also filled with functions
-via the menu ``Select functions''. For instance, all functions
-or only the exported functions defined in the currently selected
-module can be shown there, or all functions from different modules
-that are directly or indirectly called from
-a currently selected function.
-This list box is central to focus on a function in the
-source code of some module or to analyze some function,
-i.e., showing their properties. In order to focus on a function,
-it is sufficient to check the ``focus on code'' button.
-To analyze an individually selected function, one can
-select an analysis from the list of available program analyses
-(through the menu ``Select analysis'').
-In this case, the analysis results are either shown
-in the text box below the main text area
-or visualized by separate tools, e.g., by a graph drawing tool for
-visualizing call graphs.
-Some analyses are local, i.e., they need only to consider the local definition
-of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
-``Pattern completeness''),
-where other analyses are global, i.e.,
-they consider the definitions of all functions directly or indirectly called
-by this function (e.g., ``Depends on,'' ``Solution complete,''
-``Set-valued'').
-%
-Finally, there are a few additional tools integrated into \cb,
-for instance, to visualize the import relation between all modules
-as a dependency graph. These tools are available through the ``Tools'' menu.
-
-More details about the use of \cb and all built-in analyses
-are available through the ``Help'' menu of \cb.
-
-\newpage
-
 \section{CurryTest: A Tool for Testing Curry Programs}
@@
 \end{figure}
-
-\newpage
-
-\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
-\label{sec-erd2curry}
-
-ERD2Curry\index{ERD2Curry}\index{database programming}
-is a tool to generate Curry code to access and manipulate data
-persistently stored from
-entity relationship diagrams.\index{entity relationship diagrams}
-The idea of this tool is described in detail in
-\cite{BrasselHanusMueller08PADL}.
-Thus, we describe only the basic steps to use this tool
-in the following.
-
-If one creates an entity relationship diagram (ERD)
-with the Umbrello UML Modeller, one has to store its
-XML description in XMI format (as offered by Umbrello)
-in a file, e.g., \ccode{myerd.xmi}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry myerd.xmi
-\end{curry}
-(\code{erd2curry} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-If \code{MyData} is the name of the ERD, the Curry program file
-\ccode{MyData.curry} is generated containing all the necessary
-database access code as described in \cite{BrasselHanusMueller08PADL}.
-
-If one does not want to use the Umbrello UML Modeller,
-one can also create a textual description of the ERD
-as a Curry term of type \code{ERD}
-(w.r.t.\ the type definition given in module
-\code{\cyshome/tools/erd2curry/ERD.curry})
-and store it in some file, e.g., \ccode{myerd.term}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry -t myerd.term
-\end{curry}
-%
-There is also the possibility to visualize an ERD term
-as a graph with the graph visualization program \code{dotty}
-(for this purpose, it might be necessary to adapt the definition
-of \code{dotviewcommand} in your \ccode{.kics2rc} file,
-see Section~\ref{sec-customization},
-according to your local environment).
-This can be done by the command
-\begin{curry}
-erd2curry -v myerd.term
-\end{curry}
-
-\paragraph{Inclusion in the Curry application:}
-To compile the generated database code, either
-include the directory \code{\cyshome/tools/erd2curry}
-into your Curry load path
-(e.g., by setting  the environment variable
-\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
-or copy the file
-\code{\cyshome/tools/erd2curry/ERDGeneric.curry}
-into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with \CYS
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of \CYS and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of \CYS:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{\cyshome/cpns/start} and
-\code{\cyshome/cpns/stop}.
-If the demon is already running, the command \code{\cyshome/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
-\addcontentsline{toc}{section}{Bibliography}
-\bibliography{mh}
-\bibliographystyle{plain}
-
-\newpage
-\appendix
-
-\section{Libraries of the \CYS Distribution}
-\label{sec:libraries}
-
-{\setlength{\parindent}{0.0cm}
-
-The \CYS distribution comes with an extensive collection
-of libraries for application programming.
-The libraries for meta-programming by representing
-Curry programs as datatypes in Curry are described
-in the following subsection in more detail.
-The complete set of libraries with all exported types and functions
-are described in the further subsections.
-For a more detailed online documentation of all libraries of \CYS,
-see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
-
-\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
-\label{sec-flatcurry}
-
-\index{AbstractCurry}
-\index{FlatCurry}
-To support meta-programming, i.e., the manipulation of Curry programs
-in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{\cyshome/lib/meta},
-which define datatypes for the representation
-of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
-where local function definitions are replaced by global definitions
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
-program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
-programs that are more oriented towards the source program.
-
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
-These actions parse the corresponding source program and return
-a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
-
-Since all datatypes are explained in detail in these modules,
-we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
-of these modules.
-
-As an example, consider a program file \ccode{test.curry}
-containing the following two lines:
-\begin{curry}
-rev []     = []
-rev (x:xs) = (rev xs) ++ [x]
-\end{curry}
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
-following term:
-\begin{curry}
- (Prog "test"
-  ["Prelude"]
-  []
-  [Func ("test","rev") 1 Public
-        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
-                  (TCons ("Prelude","[]") [(TVar 0)]))
-        (Rule [0]
-           (Case Flex (Var 1)
-              [Branch (Pattern ("Prelude","[]") [])
-                  (Comb ConsCall ("Prelude","[]") []),
-               Branch (Pattern ("Prelude",":") [2,3])
-                  (Comb FuncCall ("Prelude","++")
-                        [Comb FuncCall ("test","rev") [Var 3],
-                         Comb ConsCall ("Prelude",":")
-                              [Var 2,Comb ConsCall ("Prelude","[]") []]
-                        ])
-              ]))]
-  []
- )
-\end{curry}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Definitions in order to LaTeX documents generated by "currydoc --tex"
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
-\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
-\newcommand{\currytypesstop}{}
-\newcommand{\currytypesynstart}[2]{{\tt type #2}\pindex{#1} \begin{quote}}
-\newcommand{\currytypesynstop}{\end{quote}}
-\newcommand{\currydatastart}[1]{{\tt data #1}\pindex{#1} \begin{quote}}
-\newcommand{\currydatacons}{\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:}}
-\newcommand{\currydatastop}{\end{itemize}}
-\newcommand{\curryconsstart}[2]{\item {\tt #1~::~#2}\par}
-\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
-\newcommand{\curryfuncstop}{}
-\newcommand{\curryfunctionstart}[2]{#2\pindex{#1}\begin{quote}}
-\newcommand{\curryfunctionstop}{\end{quote}}
-\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
-
-\subsection{General Libraries}
-
-\input{lib/AllSolutions}
-\input{lib/Assertion}
-\input{lib/Char}
-\input{lib/Combinatorial}
-\input{lib/Constraint}
-\input{lib/CSV}
-\input{lib/Directory}
-\input{lib/FileGoodies}
-\input{lib/Float}
-\input{lib/Global}
-\input{lib/GUI}
-\input{lib/Integer}
-\input{lib/IO}
-\input{lib/IOExts}
-\input{lib/JavaScript}
-\input{lib/KeyDatabaseSQLite}
-\input{lib/List}
-\input{lib/Maybe}
-\input{lib/NamedSocket}
-\input{lib/Parser}
-\input{lib/Pretty}
-\input{lib/Profile}
-\input{lib/PropertyFile}
-\input{lib/Read}
-\input{lib/ReadNumeric}
-\input{lib/ReadShowTerm}
-\input{lib/SetFunctions}
-\input{lib/SearchTree}
-\input{lib/Socket}
-\input{lib/System}
-\input{lib/Time}
-\input{lib/Unsafe}
-
-\subsection{Data Structures and Algorithms}
-
-\input{lib/Array}
-\input{lib/Dequeue}
-\input{lib/FiniteMap}
-\input{lib/GraphInductive}
-\input{lib/Random}
-\input{lib/RedBlackTree}
-\input{lib/SetRBT}
-\input{lib/Sort}
-\input{lib/TableRBT}
-\input{lib/Traversal}
-
-\subsection{Libraries for Web Applications}
-
-\input{lib/CategorizedHtmlList}
-\input{lib/HTML}
-\input{lib/HtmlParser}
-\input{lib/Mail}
-\input{lib/Markdown}
-\input{lib/WUI}
-\input{lib/URL}
-\input{lib/XML}
-\input{lib/XmlConv}
-
-\subsection{Libraries for Meta-Programming}
-
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
-\input{lib/CompactFlatCurry}
-\input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
-
-} % end setlength parindent
-
-\newpage
-
-\input{markdown_syntax}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with \CYS,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of \CYS create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with \CYS.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with \CYS
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-\CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with \CYS
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, \CYS has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Thus, if an external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\begin{curry}
-f external
-\end{curry}
-below the type declaration for the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the \CYS compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, \CYS follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\begin{curry}
-sconc :: String -> String -> String
-\end{curry}
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal form.
-This is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\begin{currynomath}
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $## s1) $## s2
-
-prim_sconc :: String -> String -> String
-prim_sconc external
-\end{currynomath}
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of \CYS.
-The \CYS compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\begin{curry}
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> ConstStore
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
-\end{curry}
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the programmer.
-The additional argument of type \code{ConstStore}
-represents the current set of constraints when this
-operation is called. This argument is intended to provide
-a more efficient access to binding constraints and can be
-ignored in standard operations.
-
-If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The \CYS compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\begin{curry}
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> ConstStore -> CP.OP_List CP.C_Char
-\end{curry}
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of \CYS
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-(note that we simply ignore the final argument representing the
-constraint store)
-\begin{curry}
-external_d_C_prim_sconc s1 s2 _ =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\end{curry}
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the \CYS implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
-For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\begin{currynomath}
-printString :: String -> IO ()
-printString s = prim_printString $## s
-
-prim_printString :: String -> IO ()
-prim_printString external
-\end{currynomath} % $
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
-\begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
-                              -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-\end{curry}
-When we compile the Curry module \code{String},
-\CYS combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, \CYS transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\begin{curry}
-(<&>) :: Int -> Int -> Int
-(<&>) external
-\end{curry}
-the generated Haskell module contains the code
-\begin{curry}
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> ConstStore -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
-\end{curry}
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
-\addcontentsline{toc}{section}{Index}
-\printindex
-
-\end{document}
-
-% LocalWords:  KiCS

### 2013-07-12 docs/src/erd2curry.tex
Split manual into separate tex files

--- 2013-03-07 docs/src/manual.tex
@@
-\documentclass[11pt,fleqn]{article}
-
-\usepackage{latexsym}
-\usepackage{makeidx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{graphicx}
-
-\input{version}
-
-%%% ------------------------------------------------------------------
-
-\usepackage[colorlinks,linkcolor=blue]{hyperref}
-\hypersetup{bookmarksopen=true}
-\hypersetup{bookmarksopenlevel=0}
-\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
-\hypersetup{pdfauthor={Michael Hanus}}
-%\hypersetup{pdfstartview=Title}
-\hypersetup{pdfstartview=FitH}
-\usepackage{thumbpdf}
-
-%%% ------------------------------------------------------------------
-
-\setlength{\textwidth}{16.5cm}
-\setlength{\textheight}{23cm}
-\renewcommand{\baselinestretch}{1.1}
-\setlength{\topmargin}{-1cm}
-\setlength{\oddsidemargin}{0cm}
-\setlength{\evensidemargin}{0cm}
-\setlength{\marginparwidth}{0.0cm}
-\setlength{\marginparsep}{0.0cm}
-
-\newlength{\figurewidth}
-\setlength{\figurewidth}{\textwidth}
-\addtolength{\figurewidth}{-0.4cm}
-
-% font for program texts
-\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\small\tt}
-
-\usepackage{listings}
-\lstset{aboveskip=1.5ex,
-        belowskip=1.5ex,
-        showstringspaces=false, % no special string space
-        mathescape=true,
-        flexiblecolumns=false,
-        xleftmargin=2ex,
-        basewidth=0.52em,
-        basicstyle=\small\ttfamily}
-
-\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
-       }
-
-\lstnewenvironment{curry}{}{}
-\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
-\newcommand{\listline}{\vrule width0pt depth1.75ex}
-
-% program text in normal text
-\newcommand{\code}[1]{\mbox{\codefont #1}}
-
-% program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\code{#1}''}
-
-\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index
-
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
-\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
-\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
-\newcommand{\cb}{CurryBrowser\xspace}
-
-% allow underscores in programs:
-\catcode`\_=\active
-\let_=\sb
-\catcode`\_=12
-
-% produce an index:
-\makeindex
-
-\begin{document}
-\sloppy
-
-\begin{titlepage}
-\pdfbookmark[1]{Title}{Title}
-\begin{center}
-\fbox{
-\begin{minipage}[t]{\figurewidth}
-\begin{center}\vspace{10ex}
-{\Huge\bf KiCS2}\\[4ex]
-{\huge The Kiel Curry System (Version 2)}\\[7ex]
-{\huge User Manual}\\[4ex]
-\kicsversiondate\\[6ex]
-\Large
-Michael Hanus$^1$ [editor] \\[3ex]
-{\large Additional Contributors:}\\[2ex]
-Bernd Bra\ss{}el$^2$ \\
-Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
-\normalsize
-(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
-(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
-(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
-\end{center}
-\end{minipage}}
-\end{center}
-\end{titlepage}
-
-\pdfbookmark[1]{Contents}{Contents}
-\tableofcontents
-
-\newpage
-
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current \CYS implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, \CYS also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
-Therefore, this document only explains the use of the different
-components of \CYS
-and the differences and restrictions of \CYS
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of \CYS
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of \CYS}
-
-\subsection{Installation}
-\label{sec-install}
-
-This version of \CYS has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems.
-
-Installation instructions for \CYS can be found in
-the file \code{INSTALL.txt} stored in the \CYS installation directory.
-Note that there are two possibilities to install \CYS:
-\begin{description}
-\item[Global installation:]\index{global installation}\index{installation!global}
-\CYS is installed in some global system directory where users
-have no write permission. In this case, some options for experimenting
-with \CYS (like \code{supply} or \code{ghc}, see below)
-are not available (since they require the recompilation of parts of
-the installed system).
-\item[Local installation:]\index{local installation}\index{installation!local}
-\CYS is installed in some local user directory where the user
-has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the \CYS installation is set as follows:
-\begin{lstlisting}
-GLOBALINSTALL=no
-\end{lstlisting}
-In this case, all options of \CYS are available.
-\end{description}
-In the following, \cyshome denotes the installation directory
-of the \CYS installation.
-
-\subsection{General Use}
-\label{sec-general}
-
-All executables required to use the different components
-of \CYS are stored in the directory \code{\cyshome/bin}.
-You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=\cyshome/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with \CYS creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\begin{curry}
-cleancurry
-\end{curry}
-(this is a shell script stored in the \code{bin} directory of the
-\CYS installation, see above).
-The command
-\begin{curry}
-cleancurry -r
-\end{curry}
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with \CYS:
-\begin{itemize}
-\item
-\index{singleton variables}\index{variables!singleton}
-\emph{Singleton pattern variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-\CYS translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-In general, tabulator stops should be avoided in source programs.
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
-based on the suspension of expressions
-containing free variables are not yet supported.
-\CYS supports \emph{value generators} for free variables
-so that a free variable is instantiated when its value is demanded.
-For instance, the initial expression
-\begin{curry}
-x == True where x free
-\end{curry}
-is non-deterministically evaluated to \code{False} and \code{True}
-by instantiating \code{x} to \code{False} and \code{True}, respectively.
-Thus, a computation is never suspended due to free variables.
-This behavior also applies to free variables of primitive types
-like integers. For instance, the initial expression
-\begin{curry}
-x*y=:=1 where x,y free
-\end{curry}
-is non-deterministically evaluated to the two solutions
-\begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
-\end{curry}
-\item
-Unification is performed without an occur check.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in \CYS}
-\label{sec-modules}
-
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-\CYS searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a \CYS session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of \CYS}
-\label{sec-kics2}
-
-This section describes the interactive environment
-\CYS\index{\CYS}
-that supports the development of applications written in Curry.
-The implementation of \CYS contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking \CYS}
-\label{sec:invoke-kics2}
-
-To start \CYS, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{\cyshome/bin} where \cyshome is the installation directory
-of \CYS).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{\cyshome/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke \CYS with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\begin{curry}
-kics2 :load Mod :add List
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\begin{curry}
-kics2 :load Mod :eval config
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\begin{curry}
-kics2 :load Mod :save :quit
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, creates an executable,
-and terminates \CYS. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of \CYS}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of \CYS are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs.
-In order to see the results of their bindings,\footnote{Currently,
-bindings are only printed if the initial expression is not an I/O action
-(i.e., not of type \ccode{IO\ldots})
-and there are not more than ten free variables
-in the initial expression.}
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
-\end{curry}
-in order to obtain the following three possible bindings:
-\begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
-\end{curry}
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\begin{curry}
-xs++ys=:=[z] where xs,ys,z free
-\end{curry}
-they are specialized to the type \ccode{()}
-(since the current implementation of \CYS does not support
-computations with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of \CYS.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the \CYS environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current \CYS process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Options of \CYS}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of \CYS.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices $n$}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression
-in a tree-like structure.
-This mode is only useful for debugging or
-understanding the implementation of non-deterministic
-evaluations used in \CYS.
-If the optional argument $n$ is provided,
-the tree is shown up to depth $n$.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-(not available in global installations, see Section~\ref{sec-install})
-Use implementation $i$ as the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the following values for $i$ are supported:
-\begin{description}
-\item[\code{integer}:]\pindex{integer}
-Use unbounded integers as choice identifiers.
-This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc}:]\pindex{ghc}
-Use a more sophisticated implementation of choice
-identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
-provided by the Glasgow Haskell Compiler.
-\item[\code{pureio}:]\pindex{pureio}
-Use IO references (i.e., memory cells) for choice identifiers.
-This is the most efficient implementation for top-level depth-first search
-but cannot be used for more sophisticated search methods
-like encapsulated search.
-\item[\code{ioref} (default):]\pindex{ioref}
-Use a mixture of \code{ghc} and \code{pureio}.
-IO references are used for top-level depth-first search
-and \code{ghc} identifiers are used for encapsulated search methods.
-\end{description}
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-Thus, one has also the possibility to terminate the
-enumeration of all values after having seen some values.
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the \CYS compiler.
-For instance, setting the option
-\begin{curry}
-:set cmp -O 0
-\end{curry}
-has the effect that all optimizations performed by the \CYS compiler
-are turned off.
-
-\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-Define additional options passed to the Glasgow Haskell Compiler (GHC)
-when the generated Haskell programs are compiled.
-One has to be careful when using such options.
-For instance, in a global installation of \CYS
-(see Section~\ref{sec-install}), libraries are pre-compiled
-so that inconsistencies might occur if compilation options
-might be changed.
-
-It is safe to pass specific GHC linking options.
-For instance, to enforce the static linking of libraries
-in order to generate an executable (see command \ccode{:save})
-that can be executed in another environment, one could set the
-options
-\begin{curry}
-:set ghc -static -optl-static -optl-pthread
-\end{curry}
-
-Other options are useful for experimental purposes,
-but those should be used only in local installations
-(see Section~\ref{sec-install}) to avoid inconsistent
-target codes for different libraries.
-For instance, setting the option
-\begin{curry}
-:set ghc -DDISABLE_CS
-\end{curry}
-has the effect that the constraint store used to enable
-an efficient access to complex bindings is disabled.
-Similarly,
-\begin{curry}
-:set ghc -DSTRICT_VAL_BIND
-\end{curry}
-has the effect that expressions in a unification constraint
-(\code{=:=}) are always fully evaluated
-(instead of the evaluation to a head normal form only)
-before unifying both sides.
-Since these options influence the compilation of the run-time system,
-one should also enforce the recompilation of Haskell programs
-by the GHC option \ccode{-fforce-recomp}, e.g., one should set
-\begin{curry}
-:set ghc -DDISABLE_CS -fforce-recomp
-\end{curry}
-
-\item[\fbox{\code{rts $opts$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\begin{curry}
-:set rts -H512m
-\end{curry}
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\item[\fbox{\code{args $arguments$}}]\pindex{args}
-Define run-time arguments passed to the executable
-generated by the Glasgow Haskell Compiler.
-For instance, setting the option
-\begin{curry}
-:set args first second
-\end{curry}
-has the effect that the I/O operation \code{getArgs}
-(see library \code{System} (Section~\ref{Library:System})
-returns the value \code{["first","second"]}.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-\CYS requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\begin{curry}
-{-# KiCS2_OPTION $opt$ #-}
-\end{curry}
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the \CYS command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\begin{curry}
-{-# KiCS2_OPTION ids #-}
-{-# KiCS2_OPTION +ghci #-}
-{-# KiCS2_OPTION v2 #-}
-module M where
-$\ldots$
-\end{curry}
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of \CYS (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call \CYS with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of \CYS to your own preferences,
-there is a configuration file which is read by \CYS when it is invoked.
-When you start \CYS for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started \CYS for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of \CYS contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into \CYS
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{\cyshome/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
-\section{Extensions}
-\label{sec-extensions}
-
-\CYS supports some extensions in Curry programs that are not (yet)
-part of the definition of Curry. These extensions are described below.
-
-\subsection{Recursive Variable Bindings}
-
-Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
-For instance, the declaration
-\begin{curry}
-ones5 = let ones = 1 : ones
-         in take 5 ones
-\end{curry}
-introduces the local variable \code{ones} which is bound
-to a \emph{cyclic structure}\index{cyclic structure}
-representing an infinite list of \code{1}'s.
-Similarly, the definition
-\begin{curry}
-onetwo n = take n one2
- where
-   one2 = 1 : two1
-   two1 = 2 : one2
-\end{curry}
-introduces a local variables \code{one2} that represents
-an infinite list of alternating \code{1}'s and \code{2}'s
-so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.
-
-\subsection{Functional Patterns}
-
-Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
-defining operations with functional patterns avoids problems
-caused by strict equality (\ccode{=:=}) and leads to programs
-that are potentially more efficient.
-
-Consider the definition of an operation to compute the last element
-of a list \code{xs} based on the prelude operation \ccode{++}
-for list concatenation:
-\begin{curry}
-last xs | _++[y] =:= xs  = y   where y free
-\end{curry}
-Since the equality constraint \ccode{=:=} evaluates both sides
-to a constructor term, all elements of the list \code{xs} are
-fully evaluated in order to satisfy the constraint.
-
-Functional patterns can help to improve this computational behavior.
-A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
-is a function call at a pattern position. With functional patterns,
-we can define the operation \code{last} as follows:
-\begin{curry}
-last (_++[y]) = y
-\end{curry}
-This definition is not only more compact but also avoids the complete
-evaluation of the list elements: since a functional pattern is considered
-as an abbreviation for the set of constructor terms obtained by all
-evaluations of the functional pattern to normal form (see
-\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
-definition is conceptually equivalent to the set of rules
-\begin{curry}
-last [y] = y
-last [_,y] = y
-last [_,_,y] = y
-$\ldots$
-\end{curry}
-which shows that the evaluation of the list elements is not demanded
-by the functional pattern.
-
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
-is interpreted as a functional pattern if $f$ is not a visible constructor
-but a defined function that is visible in the scope of the pattern.
-
-It is also possible to combine functional patterns with
-as-patterns.\index{as-pattern}\pindex{"@}
-Similarly to the meaning of as-patterns
-in standard constructor patterns,
-as-patterns in functional patterns are interpreted
-as a sequence of pattern matching where the variable of the as-pattern
-is matched before the given pattern is matched.
-This process can be described by introducing an auxiliary operation
-for this two-level pattern matching process.
-For instance, the definition
-\begin{curry}
-f (_ ++ x@[(42,_)] ++ _) = x
-\end{curry}
-is considered as syntactic sugar for the expanded definition
-\begin{curry}
-f (_ ++ x ++ _) = f' x
- where
-  f' [(42,_)] = x
-\end{curry}
-However, as-patterns are usually implemented
-in a more efficient way without introducing auxiliary operations.
-
-\subsection{Order of Pattern Matching}
-
-Curry allows multiple occurrences of pattern variables
-in standard patterns. These are an abbreviation of equational constraints
-between pattern variables.
-Functional patterns might also contain multiple occurrences of
-pattern variables.
-For instance, the operation
-\begin{curry}
-f (_++[x]++_++[x]++_) = x
-\end{curry}
-returns all elements with at least two occurrences in a list.
-
-If functional patterns as well as multiple occurrences of
-pattern variables occur in a pattern defining an operation,
-there are various orders to match an expression against such
-an operation. In the current implementation, the order
-is as follows:
-\begin{enumerate}
-\item Standard pattern matching: First, it is checked whether
-the constructor patterns match. Thus, functional patterns
-and multiple occurrences of pattern variables are ignored.
-\item Functional pattern matching: In the next phase,
-functional patterns are matched but occurrences of standard
-pattern variables in the functional patterns are ignored.
-\item Non-linear patterns: If standard and functional pattern matching
-is successful, the equational constraints which correspond
-to multiple occurrences pattern variables are solved.
-\item Guards: Finally, the guards supplied by the programmer
-are checked.
-\end{enumerate}
-The order of pattern matching should not influence the computed
-result. However, it might have some influence on the termination
-behavior of programs, i.e., a program might not terminate
-instead of finitely failing.
-In such cases, it could be necessary to consider the influence
-of the order of pattern matching. Note that other orders of pattern matching
-can be obtained using auxiliary operations.
-
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are 
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there. 
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\begin{curry}
-type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
-\end{curry}
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$ 
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\begin{curry}
-type Person = {name :: String, age :: Int}
-type Address = {person :: Person, street :: String, city :: String}
-type Branch a b = {left :: a, right :: b}
-\end{curry}
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\begin{curry}
-type Address = {person :: Person, street,city :: String}
-\end{curry}
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\begin{curry}
-type Address = {street,city :: String, person :: Person}
-\end{curry}
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\begin{curry}
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\end{curry}
-\begin{curry}
-getName :: Person -> String
-getName $\ldots$
-\end{curry}
-
-Labels can only be used in the context of
-records. They do not share the name space with 
-functions/constructors/variables or type constructors/type variables. 
-For instance it is possible to use 
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually 
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\begin{curry}
-{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
-\end{curry}
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with 
-exactly one assignment. Examples for record constructions are
-\begin{curry}
-{name := "Johnson", age := 30}     -- generates a record of type 'Person'
-{left := True, right := 20}        -- generates a record of type 'Branch'
-\end{curry}
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\begin{curry}
-{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
-\end{curry}
-Unlike labeled fields in record type declarations, 
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\begin{curry}
-{person := {name := "Smith", age := 20},   -- generates a record of
- street := "Main Street",                  -- type 'Address'
- city   := "Springfield"}
-\end{curry}
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records. 
-It has the following form:
-\begin{curry}
-$r$ :> $l$
-\end{curry}
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\begin{curry}
-pers :> name
-\end{curry}
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\begin{curry}
-addr :> person :> age
-\end{curry}
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}).
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\begin{curry}
-{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
-\end{curry}
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration 
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record. 
-Assignments can occur in an arbitrary order. It is not allowed to 
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\begin{curry}
-{name := "Scott", age := 25 | pers}
-{person := {name := "Scott", age := 25 | pers} | addr}
-\end{curry}
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\begin{curry}
-{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
-{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
-\end{curry}
-In both cases each label $l_i$ is specified with a pattern $p_i$. 
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern. 
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the 
-patterns of the specified labels are matched against 
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\begin{curry}
-isSmith30 :: Person -> Bool
-isSmith30 {name = "Smith", age = 30} = True
-\end{curry}
-\begin{curry}
-startsWith :: Char -> Person -> Bool
-startsWith c {name = (d:_) | _} = c == d
-\end{curry}
-\begin{curry}
-getPerson :: Address -> Person
-getPerson {person = p | _} = p
-\end{curry}
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways 
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent 
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
-
-\newpage
-
-\input{currydoc}
-
-\newpage
-
-\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
-\label{sec-currybrowser}
-
-\cb is a tool to browse through the modules and functions
-of a Curry application, show them in various formats,
-and analyze their properties.\footnote{Although \cb is
-implemented in Curry, some functionalities of it require an
-installed graph visualization tool (dot \url{http://www.graphviz.org/}),
-otherwise they have no effect.}
-Moreover, it is constructed in a way so that
-new analyzers can be easily connected to \cb.
-A detailed description of the ideas behind this tool can be
-found in \cite{Hanus05WCFLP,Hanus06WLPE}.
-
-\cb is part of the \CYS distribution and can be
-started in two ways:
-\begin{itemize}
-\item
-In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
-\item
-In the \CYS environment after loading the module
-\code{mod} and typing the command \ccode{:browse}.
-\end{itemize}
-Here, \ccode{mod} is the name of the main module of a Curry application.
-After the start, \cb loads the interfaces of the main
-module and all imported modules before a GUI is created
-for interactive browsing.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.7]{currybrowser.jpg}
-\end{center}
-\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
-\end{figure}
-%
-To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
-shows a snapshot of its use on a particular application
-(here: the implementation of \cb).
-The upper list box in the left column shows the modules and their imports
-in order to browse through the modules of an application.
-Similarly to directory browsers, the list of imported modules of a module
-can be opened or closed by clicking.
-After selecting a module in the list of modules, its source code,
-interface, or various other formats of the module can be shown
-in the main (right) text area. For instance, one can show
-pretty-printed versions of the intermediate flat programs (see below)
-in order to see how local function definitions are translated by lambda lifting
-\cite{Johnsson85}
-or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
-Since Curry is a language with parametric polymorphism and type inference,
-programmers often omit the type signatures when defining functions.
-Therefore, one can also view (and store) the selected module as source code where
-missing type signatures are added.
-
-Below the list box for selecting modules, there is a menu
-(``Analyze selected module'') to analyze all functions
-of the currently selected module at once. This is useful
-to spot some functions of a module that could be problematic
-in some application contexts, like functions that are impure (i.e., the result
-depends on the evaluation time) or partially defined (i.e.,
-not evaluable on all ground terms).
-If such an analysis is selected,
-the names of all functions are shown in the
-lower list box of the left column (the ``function list'')
-with prefixes indicating the properties of the individual functions.
-
-The function list box can be also filled with functions
-via the menu ``Select functions''. For instance, all functions
-or only the exported functions defined in the currently selected
-module can be shown there, or all functions from different modules
-that are directly or indirectly called from
-a currently selected function.
-This list box is central to focus on a function in the
-source code of some module or to analyze some function,
-i.e., showing their properties. In order to focus on a function,
-it is sufficient to check the ``focus on code'' button.
-To analyze an individually selected function, one can
-select an analysis from the list of available program analyses
-(through the menu ``Select analysis'').
-In this case, the analysis results are either shown
-in the text box below the main text area
-or visualized by separate tools, e.g., by a graph drawing tool for
-visualizing call graphs.
-Some analyses are local, i.e., they need only to consider the local definition
-of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
-``Pattern completeness''),
-where other analyses are global, i.e.,
-they consider the definitions of all functions directly or indirectly called
-by this function (e.g., ``Depends on,'' ``Solution complete,''
-``Set-valued'').
-%
-Finally, there are a few additional tools integrated into \cb,
-for instance, to visualize the import relation between all modules
-as a dependency graph. These tools are available through the ``Tools'' menu.
-
-More details about the use of \cb and all built-in analyses
-are available through the ``Help'' menu of \cb.
-
-\newpage
-
-\section{CurryTest: A Tool for Testing Curry Programs}
-\label{sec-currytest}
-
-CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the \CYS distribution to write
-and run repeatable tests. CurryTest simplifies the task
-of writing test cases for a module and executing them.
-The tool is easy to use. Assume one has implemented a module \code{MyMod}
-and wants to write some test cases to test its functionality,
-making regression tests in future versions, etc.
-For this purpose, there is a system library \code{Assertion}
-(Section~\ref{Library:Assertion}) which
-contains the necessary definitions for writing tests.
-In particular, it exports an abstract polymorphic type \ccode{Assertion a}
-together with the following operations:
-\begin{curry}
-assertTrue      :: String -> Bool -> Assertion ()
-assertEqual     :: String -> a -> a -> Assertion a
-assertValues    :: String -> a -> [a] -> Assertion a
-assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
-assertIO        :: String -> IO a -> a -> Assertion a
-assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\end{curry}
-The expression \ccode{assertTrue $s$ $b$}
-is an assertion (named $s$) that the expression $b$ has the value \code{True}.
-Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
-asserts that the expressions $e_1$ and $e_2$
-must be equal (i.e., \code{$e_1$==$e_2$} must hold),
-the expression \ccode{assertValues $s$ $e$ $vs$} asserts
-that $vs$ is the multiset of all values of $e$,
-and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
-that the constraint abstraction $c$ has the multiset of solutions $vs$.
-Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
-asserts that the I/O action $a$ yields the value $v$ whenever it is
-executed, and
-the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yield equal values.
-The name $s$ provided as a first argument in each assertion
-is used in the protocol produced by the test tool.
-
-One can define a test program by importing the module
-to be tested together with the module \code{Assertion} and defining
-top-level functions of type \code{Assertion} in this module
-(which must also be exported).
-As an example, consider the following program
-that can be used to test some list processing functions:
-\begin{curry}
-import List
-import Assertion
-
-test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-
-test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-
-test3 = assertSolutions "prefix" (\x -> x++_ =:= [1,2])
-                                 [[],[1],[1,2]]
-\end{curry}
-For instance, \code{test1} asserts that the result of evaluating the
-expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.
-
-We can execute a test suite by the command\pindex{currytest}
-\begin{curry}
-currytest TestList
-\end{curry}
-(\code{currytest} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-In our example, \ccode{TestList.curry} is the program containing the
-definition of all assertions. This has the effect
-that all exported top-level functions
-of type \code{Assertion} are tested (i.e., the corresponding
-assertions are checked) and the results
-(\ccode{OK} or failure) are reported together with the name of each assertion.
-%If failures occur, the complete test results are also
-%written into a file named \ccode{TestList.testlog}.''
-For our example above, we obtain the following successful protocol:
-\begin{curry}
-============================================================
-Testing module "TestList"...
-OK: ++
-OK: all
-OK: prefix
-All tests successfully passed.
-============================================================
-\end{curry}
-There is also a graphical interface that summarizes the results
-more nicely.
-In order to start this interface, one has to add the parameter
-\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\begin{curry}
-currytest --window TestList
-\end{curry}
-or
-\begin{curry}
-currytest -w TestList
-\end{curry}
-A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
-
-\begin{figure}%[t]
-\begin{center}
-\includegraphics[scale=0.7]{currytest.jpg}
-\end{center}
-\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
-\end{figure}
-
-\newpage
-
 \section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
@@
 into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with \CYS
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of \CYS and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of \CYS:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{\cyshome/cpns/start} and
-\code{\cyshome/cpns/stop}.
-If the demon is already running, the command \code{\cyshome/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
-\addcontentsline{toc}{section}{Bibliography}
-\bibliography{mh}
-\bibliographystyle{plain}
-
-\newpage
-\appendix
-
-\section{Libraries of the \CYS Distribution}
-\label{sec:libraries}
-
-{\setlength{\parindent}{0.0cm}
-
-The \CYS distribution comes with an extensive collection
-of libraries for application programming.
-The libraries for meta-programming by representing
-Curry programs as datatypes in Curry are described
-in the following subsection in more detail.
-The complete set of libraries with all exported types and functions
-are described in the further subsections.
-For a more detailed online documentation of all libraries of \CYS,
-see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
-
-\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
-\label{sec-flatcurry}
-
-\index{AbstractCurry}
-\index{FlatCurry}
-To support meta-programming, i.e., the manipulation of Curry programs
-in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{\cyshome/lib/meta},
-which define datatypes for the representation
-of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
-where local function definitions are replaced by global definitions
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
-program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
-programs that are more oriented towards the source program.
-
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
-These actions parse the corresponding source program and return
-a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
-
-Since all datatypes are explained in detail in these modules,
-we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
-of these modules.
-
-As an example, consider a program file \ccode{test.curry}
-containing the following two lines:
-\begin{curry}
-rev []     = []
-rev (x:xs) = (rev xs) ++ [x]
-\end{curry}
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
-following term:
-\begin{curry}
- (Prog "test"
-  ["Prelude"]
-  []
-  [Func ("test","rev") 1 Public
-        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
-                  (TCons ("Prelude","[]") [(TVar 0)]))
-        (Rule [0]
-           (Case Flex (Var 1)
-              [Branch (Pattern ("Prelude","[]") [])
-                  (Comb ConsCall ("Prelude","[]") []),
-               Branch (Pattern ("Prelude",":") [2,3])
-                  (Comb FuncCall ("Prelude","++")
-                        [Comb FuncCall ("test","rev") [Var 3],
-                         Comb ConsCall ("Prelude",":")
-                              [Var 2,Comb ConsCall ("Prelude","[]") []]
-                        ])
-              ]))]
-  []
- )
-\end{curry}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Definitions in order to LaTeX documents generated by "currydoc --tex"
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
-\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
-\newcommand{\currytypesstop}{}
-\newcommand{\currytypesynstart}[2]{{\tt type #2}\pindex{#1} \begin{quote}}
-\newcommand{\currytypesynstop}{\end{quote}}
-\newcommand{\currydatastart}[1]{{\tt data #1}\pindex{#1} \begin{quote}}
-\newcommand{\currydatacons}{\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:}}
-\newcommand{\currydatastop}{\end{itemize}}
-\newcommand{\curryconsstart}[2]{\item {\tt #1~::~#2}\par}
-\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
-\newcommand{\curryfuncstop}{}
-\newcommand{\curryfunctionstart}[2]{#2\pindex{#1}\begin{quote}}
-\newcommand{\curryfunctionstop}{\end{quote}}
-\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
-
-\subsection{General Libraries}
-
-\input{lib/AllSolutions}
-\input{lib/Assertion}
-\input{lib/Char}
-\input{lib/Combinatorial}
-\input{lib/Constraint}
-\input{lib/CSV}
-\input{lib/Directory}
-\input{lib/FileGoodies}
-\input{lib/Float}
-\input{lib/Global}
-\input{lib/GUI}
-\input{lib/Integer}
-\input{lib/IO}
-\input{lib/IOExts}
-\input{lib/JavaScript}
-\input{lib/KeyDatabaseSQLite}
-\input{lib/List}
-\input{lib/Maybe}
-\input{lib/NamedSocket}
-\input{lib/Parser}
-\input{lib/Pretty}
-\input{lib/Profile}
-\input{lib/PropertyFile}
-\input{lib/Read}
-\input{lib/ReadNumeric}
-\input{lib/ReadShowTerm}
-\input{lib/SetFunctions}
-\input{lib/SearchTree}
-\input{lib/Socket}
-\input{lib/System}
-\input{lib/Time}
-\input{lib/Unsafe}
-
-\subsection{Data Structures and Algorithms}
-
-\input{lib/Array}
-\input{lib/Dequeue}
-\input{lib/FiniteMap}
-\input{lib/GraphInductive}
-\input{lib/Random}
-\input{lib/RedBlackTree}
-\input{lib/SetRBT}
-\input{lib/Sort}
-\input{lib/TableRBT}
-\input{lib/Traversal}
-
-\subsection{Libraries for Web Applications}
-
-\input{lib/CategorizedHtmlList}
-\input{lib/HTML}
-\input{lib/HtmlParser}
-\input{lib/Mail}
-\input{lib/Markdown}
-\input{lib/WUI}
-\input{lib/URL}
-\input{lib/XML}
-\input{lib/XmlConv}
-
-\subsection{Libraries for Meta-Programming}
-
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
-\input{lib/CompactFlatCurry}
-\input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
-
-} % end setlength parindent
-
-\newpage
-
-\input{markdown_syntax}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with \CYS,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of \CYS create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with \CYS.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with \CYS
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-\CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with \CYS
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, \CYS has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Thus, if an external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\begin{curry}
-f external
-\end{curry}
-below the type declaration for the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the \CYS compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, \CYS follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\begin{curry}
-sconc :: String -> String -> String
-\end{curry}
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal form.
-This is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\begin{currynomath}
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $## s1) $## s2
-
-prim_sconc :: String -> String -> String
-prim_sconc external
-\end{currynomath}
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of \CYS.
-The \CYS compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\begin{curry}
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> ConstStore
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
-\end{curry}
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the programmer.
-The additional argument of type \code{ConstStore}
-represents the current set of constraints when this
-operation is called. This argument is intended to provide
-a more efficient access to binding constraints and can be
-ignored in standard operations.
-
-If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The \CYS compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\begin{curry}
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> ConstStore -> CP.OP_List CP.C_Char
-\end{curry}
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of \CYS
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-(note that we simply ignore the final argument representing the
-constraint store)
-\begin{curry}
-external_d_C_prim_sconc s1 s2 _ =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\end{curry}
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the \CYS implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
-For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\begin{currynomath}
-printString :: String -> IO ()
-printString s = prim_printString $## s
-
-prim_printString :: String -> IO ()
-prim_printString external
-\end{currynomath} % $
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
-\begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
-                              -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-\end{curry}
-When we compile the Curry module \code{String},
-\CYS combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, \CYS transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\begin{curry}
-(<&>) :: Int -> Int -> Int
-(<&>) external
-\end{curry}
-the generated Haskell module contains the code
-\begin{curry}
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> ConstStore -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
-\end{curry}
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
-\addcontentsline{toc}{section}{Index}
-\printindex
-
-\end{document}
-
-% LocalWords:  KiCS

### 2013-07-12 docs/src/extensions.tex
Split manual into separate tex files

--- 2013-03-07 docs/src/manual.tex
@@
-\documentclass[11pt,fleqn]{article}
-
-\usepackage{latexsym}
-\usepackage{makeidx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{graphicx}
-
-\input{version}
-
-%%% ------------------------------------------------------------------
-
-\usepackage[colorlinks,linkcolor=blue]{hyperref}
-\hypersetup{bookmarksopen=true}
-\hypersetup{bookmarksopenlevel=0}
-\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
-\hypersetup{pdfauthor={Michael Hanus}}
-%\hypersetup{pdfstartview=Title}
-\hypersetup{pdfstartview=FitH}
-\usepackage{thumbpdf}
-
-%%% ------------------------------------------------------------------
-
-\setlength{\textwidth}{16.5cm}
-\setlength{\textheight}{23cm}
-\renewcommand{\baselinestretch}{1.1}
-\setlength{\topmargin}{-1cm}
-\setlength{\oddsidemargin}{0cm}
-\setlength{\evensidemargin}{0cm}
-\setlength{\marginparwidth}{0.0cm}
-\setlength{\marginparsep}{0.0cm}
-
-\newlength{\figurewidth}
-\setlength{\figurewidth}{\textwidth}
-\addtolength{\figurewidth}{-0.4cm}
-
-% font for program texts
-\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\small\tt}
-
-\usepackage{listings}
-\lstset{aboveskip=1.5ex,
-        belowskip=1.5ex,
-        showstringspaces=false, % no special string space
-        mathescape=true,
-        flexiblecolumns=false,
-        xleftmargin=2ex,
-        basewidth=0.52em,
-        basicstyle=\small\ttfamily}
-
-\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
-       }
-
-\lstnewenvironment{curry}{}{}
-\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
-\newcommand{\listline}{\vrule width0pt depth1.75ex}
-
-% program text in normal text
-\newcommand{\code}[1]{\mbox{\codefont #1}}
-
-% program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\code{#1}''}
-
-\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index
-
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
-\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
-\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
-\newcommand{\cb}{CurryBrowser\xspace}
-
-% allow underscores in programs:
-\catcode`\_=\active
-\let_=\sb
-\catcode`\_=12
-
-% produce an index:
-\makeindex
-
-\begin{document}
-\sloppy
-
-\begin{titlepage}
-\pdfbookmark[1]{Title}{Title}
-\begin{center}
-\fbox{
-\begin{minipage}[t]{\figurewidth}
-\begin{center}\vspace{10ex}
-{\Huge\bf KiCS2}\\[4ex]
-{\huge The Kiel Curry System (Version 2)}\\[7ex]
-{\huge User Manual}\\[4ex]
-\kicsversiondate\\[6ex]
-\Large
-Michael Hanus$^1$ [editor] \\[3ex]
-{\large Additional Contributors:}\\[2ex]
-Bernd Bra\ss{}el$^2$ \\
-Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
-\normalsize
-(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
-(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
-(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
-\end{center}
-\end{minipage}}
-\end{center}
-\end{titlepage}
-
-\pdfbookmark[1]{Contents}{Contents}
-\tableofcontents
-
-\newpage
-
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current \CYS implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, \CYS also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
-Therefore, this document only explains the use of the different
-components of \CYS
-and the differences and restrictions of \CYS
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of \CYS
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of \CYS}
-
-\subsection{Installation}
-\label{sec-install}
-
-This version of \CYS has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems.
-
-Installation instructions for \CYS can be found in
-the file \code{INSTALL.txt} stored in the \CYS installation directory.
-Note that there are two possibilities to install \CYS:
-\begin{description}
-\item[Global installation:]\index{global installation}\index{installation!global}
-\CYS is installed in some global system directory where users
-have no write permission. In this case, some options for experimenting
-with \CYS (like \code{supply} or \code{ghc}, see below)
-are not available (since they require the recompilation of parts of
-the installed system).
-\item[Local installation:]\index{local installation}\index{installation!local}
-\CYS is installed in some local user directory where the user
-has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the \CYS installation is set as follows:
-\begin{lstlisting}
-GLOBALINSTALL=no
-\end{lstlisting}
-In this case, all options of \CYS are available.
-\end{description}
-In the following, \cyshome denotes the installation directory
-of the \CYS installation.
-
-\subsection{General Use}
-\label{sec-general}
-
-All executables required to use the different components
-of \CYS are stored in the directory \code{\cyshome/bin}.
-You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=\cyshome/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with \CYS creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\begin{curry}
-cleancurry
-\end{curry}
-(this is a shell script stored in the \code{bin} directory of the
-\CYS installation, see above).
-The command
-\begin{curry}
-cleancurry -r
-\end{curry}
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with \CYS:
-\begin{itemize}
-\item
-\index{singleton variables}\index{variables!singleton}
-\emph{Singleton pattern variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-\CYS translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-In general, tabulator stops should be avoided in source programs.
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
-based on the suspension of expressions
-containing free variables are not yet supported.
-\CYS supports \emph{value generators} for free variables
-so that a free variable is instantiated when its value is demanded.
-For instance, the initial expression
-\begin{curry}
-x == True where x free
-\end{curry}
-is non-deterministically evaluated to \code{False} and \code{True}
-by instantiating \code{x} to \code{False} and \code{True}, respectively.
-Thus, a computation is never suspended due to free variables.
-This behavior also applies to free variables of primitive types
-like integers. For instance, the initial expression
-\begin{curry}
-x*y=:=1 where x,y free
-\end{curry}
-is non-deterministically evaluated to the two solutions
-\begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
-\end{curry}
-\item
-Unification is performed without an occur check.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in \CYS}
-\label{sec-modules}
-
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-\CYS searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a \CYS session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of \CYS}
-\label{sec-kics2}
-
-This section describes the interactive environment
-\CYS\index{\CYS}
-that supports the development of applications written in Curry.
-The implementation of \CYS contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking \CYS}
-\label{sec:invoke-kics2}
-
-To start \CYS, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{\cyshome/bin} where \cyshome is the installation directory
-of \CYS).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{\cyshome/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke \CYS with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\begin{curry}
-kics2 :load Mod :add List
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\begin{curry}
-kics2 :load Mod :eval config
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\begin{curry}
-kics2 :load Mod :save :quit
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, creates an executable,
-and terminates \CYS. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of \CYS}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of \CYS are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs.
-In order to see the results of their bindings,\footnote{Currently,
-bindings are only printed if the initial expression is not an I/O action
-(i.e., not of type \ccode{IO\ldots})
-and there are not more than ten free variables
-in the initial expression.}
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
-\end{curry}
-in order to obtain the following three possible bindings:
-\begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
-\end{curry}
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\begin{curry}
-xs++ys=:=[z] where xs,ys,z free
-\end{curry}
-they are specialized to the type \ccode{()}
-(since the current implementation of \CYS does not support
-computations with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of \CYS.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the \CYS environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current \CYS process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Options of \CYS}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of \CYS.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices $n$}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression
-in a tree-like structure.
-This mode is only useful for debugging or
-understanding the implementation of non-deterministic
-evaluations used in \CYS.
-If the optional argument $n$ is provided,
-the tree is shown up to depth $n$.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-(not available in global installations, see Section~\ref{sec-install})
-Use implementation $i$ as the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the following values for $i$ are supported:
-\begin{description}
-\item[\code{integer}:]\pindex{integer}
-Use unbounded integers as choice identifiers.
-This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc}:]\pindex{ghc}
-Use a more sophisticated implementation of choice
-identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
-provided by the Glasgow Haskell Compiler.
-\item[\code{pureio}:]\pindex{pureio}
-Use IO references (i.e., memory cells) for choice identifiers.
-This is the most efficient implementation for top-level depth-first search
-but cannot be used for more sophisticated search methods
-like encapsulated search.
-\item[\code{ioref} (default):]\pindex{ioref}
-Use a mixture of \code{ghc} and \code{pureio}.
-IO references are used for top-level depth-first search
-and \code{ghc} identifiers are used for encapsulated search methods.
-\end{description}
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-Thus, one has also the possibility to terminate the
-enumeration of all values after having seen some values.
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the \CYS compiler.
-For instance, setting the option
-\begin{curry}
-:set cmp -O 0
-\end{curry}
-has the effect that all optimizations performed by the \CYS compiler
-are turned off.
-
-\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-Define additional options passed to the Glasgow Haskell Compiler (GHC)
-when the generated Haskell programs are compiled.
-One has to be careful when using such options.
-For instance, in a global installation of \CYS
-(see Section~\ref{sec-install}), libraries are pre-compiled
-so that inconsistencies might occur if compilation options
-might be changed.
-
-It is safe to pass specific GHC linking options.
-For instance, to enforce the static linking of libraries
-in order to generate an executable (see command \ccode{:save})
-that can be executed in another environment, one could set the
-options
-\begin{curry}
-:set ghc -static -optl-static -optl-pthread
-\end{curry}
-
-Other options are useful for experimental purposes,
-but those should be used only in local installations
-(see Section~\ref{sec-install}) to avoid inconsistent
-target codes for different libraries.
-For instance, setting the option
-\begin{curry}
-:set ghc -DDISABLE_CS
-\end{curry}
-has the effect that the constraint store used to enable
-an efficient access to complex bindings is disabled.
-Similarly,
-\begin{curry}
-:set ghc -DSTRICT_VAL_BIND
-\end{curry}
-has the effect that expressions in a unification constraint
-(\code{=:=}) are always fully evaluated
-(instead of the evaluation to a head normal form only)
-before unifying both sides.
-Since these options influence the compilation of the run-time system,
-one should also enforce the recompilation of Haskell programs
-by the GHC option \ccode{-fforce-recomp}, e.g., one should set
-\begin{curry}
-:set ghc -DDISABLE_CS -fforce-recomp
-\end{curry}
-
-\item[\fbox{\code{rts $opts$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\begin{curry}
-:set rts -H512m
-\end{curry}
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\item[\fbox{\code{args $arguments$}}]\pindex{args}
-Define run-time arguments passed to the executable
-generated by the Glasgow Haskell Compiler.
-For instance, setting the option
-\begin{curry}
-:set args first second
-\end{curry}
-has the effect that the I/O operation \code{getArgs}
-(see library \code{System} (Section~\ref{Library:System})
-returns the value \code{["first","second"]}.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-\CYS requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\begin{curry}
-{-# KiCS2_OPTION $opt$ #-}
-\end{curry}
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the \CYS command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\begin{curry}
-{-# KiCS2_OPTION ids #-}
-{-# KiCS2_OPTION +ghci #-}
-{-# KiCS2_OPTION v2 #-}
-module M where
-$\ldots$
-\end{curry}
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of \CYS (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call \CYS with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of \CYS to your own preferences,
-there is a configuration file which is read by \CYS when it is invoked.
-When you start \CYS for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started \CYS for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of \CYS contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into \CYS
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{\cyshome/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
 \section{Extensions}
@@
 \end{itemize}
-
-\newpage
-
-\input{currydoc}
-
-\newpage
-
-\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
-\label{sec-currybrowser}
-
-\cb is a tool to browse through the modules and functions
-of a Curry application, show them in various formats,
-and analyze their properties.\footnote{Although \cb is
-implemented in Curry, some functionalities of it require an
-installed graph visualization tool (dot \url{http://www.graphviz.org/}),
-otherwise they have no effect.}
-Moreover, it is constructed in a way so that
-new analyzers can be easily connected to \cb.
-A detailed description of the ideas behind this tool can be
-found in \cite{Hanus05WCFLP,Hanus06WLPE}.
-
-\cb is part of the \CYS distribution and can be
-started in two ways:
-\begin{itemize}
-\item
-In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
-\item
-In the \CYS environment after loading the module
-\code{mod} and typing the command \ccode{:browse}.
-\end{itemize}
-Here, \ccode{mod} is the name of the main module of a Curry application.
-After the start, \cb loads the interfaces of the main
-module and all imported modules before a GUI is created
-for interactive browsing.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.7]{currybrowser.jpg}
-\end{center}
-\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
-\end{figure}
-%
-To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
-shows a snapshot of its use on a particular application
-(here: the implementation of \cb).
-The upper list box in the left column shows the modules and their imports
-in order to browse through the modules of an application.
-Similarly to directory browsers, the list of imported modules of a module
-can be opened or closed by clicking.
-After selecting a module in the list of modules, its source code,
-interface, or various other formats of the module can be shown
-in the main (right) text area. For instance, one can show
-pretty-printed versions of the intermediate flat programs (see below)
-in order to see how local function definitions are translated by lambda lifting
-\cite{Johnsson85}
-or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
-Since Curry is a language with parametric polymorphism and type inference,
-programmers often omit the type signatures when defining functions.
-Therefore, one can also view (and store) the selected module as source code where
-missing type signatures are added.
-
-Below the list box for selecting modules, there is a menu
-(``Analyze selected module'') to analyze all functions
-of the currently selected module at once. This is useful
-to spot some functions of a module that could be problematic
-in some application contexts, like functions that are impure (i.e., the result
-depends on the evaluation time) or partially defined (i.e.,
-not evaluable on all ground terms).
-If such an analysis is selected,
-the names of all functions are shown in the
-lower list box of the left column (the ``function list'')
-with prefixes indicating the properties of the individual functions.
-
-The function list box can be also filled with functions
-via the menu ``Select functions''. For instance, all functions
-or only the exported functions defined in the currently selected
-module can be shown there, or all functions from different modules
-that are directly or indirectly called from
-a currently selected function.
-This list box is central to focus on a function in the
-source code of some module or to analyze some function,
-i.e., showing their properties. In order to focus on a function,
-it is sufficient to check the ``focus on code'' button.
-To analyze an individually selected function, one can
-select an analysis from the list of available program analyses
-(through the menu ``Select analysis'').
-In this case, the analysis results are either shown
-in the text box below the main text area
-or visualized by separate tools, e.g., by a graph drawing tool for
-visualizing call graphs.
-Some analyses are local, i.e., they need only to consider the local definition
-of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
-``Pattern completeness''),
-where other analyses are global, i.e.,
-they consider the definitions of all functions directly or indirectly called
-by this function (e.g., ``Depends on,'' ``Solution complete,''
-``Set-valued'').
-%
-Finally, there are a few additional tools integrated into \cb,
-for instance, to visualize the import relation between all modules
-as a dependency graph. These tools are available through the ``Tools'' menu.
-
-More details about the use of \cb and all built-in analyses
-are available through the ``Help'' menu of \cb.
-
-\newpage
-
-\section{CurryTest: A Tool for Testing Curry Programs}
-\label{sec-currytest}
-
-CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the \CYS distribution to write
-and run repeatable tests. CurryTest simplifies the task
-of writing test cases for a module and executing them.
-The tool is easy to use. Assume one has implemented a module \code{MyMod}
-and wants to write some test cases to test its functionality,
-making regression tests in future versions, etc.
-For this purpose, there is a system library \code{Assertion}
-(Section~\ref{Library:Assertion}) which
-contains the necessary definitions for writing tests.
-In particular, it exports an abstract polymorphic type \ccode{Assertion a}
-together with the following operations:
-\begin{curry}
-assertTrue      :: String -> Bool -> Assertion ()
-assertEqual     :: String -> a -> a -> Assertion a
-assertValues    :: String -> a -> [a] -> Assertion a
-assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
-assertIO        :: String -> IO a -> a -> Assertion a
-assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\end{curry}
-The expression \ccode{assertTrue $s$ $b$}
-is an assertion (named $s$) that the expression $b$ has the value \code{True}.
-Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
-asserts that the expressions $e_1$ and $e_2$
-must be equal (i.e., \code{$e_1$==$e_2$} must hold),
-the expression \ccode{assertValues $s$ $e$ $vs$} asserts
-that $vs$ is the multiset of all values of $e$,
-and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
-that the constraint abstraction $c$ has the multiset of solutions $vs$.
-Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
-asserts that the I/O action $a$ yields the value $v$ whenever it is
-executed, and
-the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yield equal values.
-The name $s$ provided as a first argument in each assertion
-is used in the protocol produced by the test tool.
-
-One can define a test program by importing the module
-to be tested together with the module \code{Assertion} and defining
-top-level functions of type \code{Assertion} in this module
-(which must also be exported).
-As an example, consider the following program
-that can be used to test some list processing functions:
-\begin{curry}
-import List
-import Assertion
-
-test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-
-test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-
-test3 = assertSolutions "prefix" (\x -> x++_ =:= [1,2])
-                                 [[],[1],[1,2]]
-\end{curry}
-For instance, \code{test1} asserts that the result of evaluating the
-expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.
-
-We can execute a test suite by the command\pindex{currytest}
-\begin{curry}
-currytest TestList
-\end{curry}
-(\code{currytest} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-In our example, \ccode{TestList.curry} is the program containing the
-definition of all assertions. This has the effect
-that all exported top-level functions
-of type \code{Assertion} are tested (i.e., the corresponding
-assertions are checked) and the results
-(\ccode{OK} or failure) are reported together with the name of each assertion.
-%If failures occur, the complete test results are also
-%written into a file named \ccode{TestList.testlog}.''
-For our example above, we obtain the following successful protocol:
-\begin{curry}
-============================================================
-Testing module "TestList"...
-OK: ++
-OK: all
-OK: prefix
-All tests successfully passed.
-============================================================
-\end{curry}
-There is also a graphical interface that summarizes the results
-more nicely.
-In order to start this interface, one has to add the parameter
-\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\begin{curry}
-currytest --window TestList
-\end{curry}
-or
-\begin{curry}
-currytest -w TestList
-\end{curry}
-A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
-
-\begin{figure}%[t]
-\begin{center}
-\includegraphics[scale=0.7]{currytest.jpg}
-\end{center}
-\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
-\end{figure}
-
-\newpage
-
-\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
-\label{sec-erd2curry}
-
-ERD2Curry\index{ERD2Curry}\index{database programming}
-is a tool to generate Curry code to access and manipulate data
-persistently stored from
-entity relationship diagrams.\index{entity relationship diagrams}
-The idea of this tool is described in detail in
-\cite{BrasselHanusMueller08PADL}.
-Thus, we describe only the basic steps to use this tool
-in the following.
-
-If one creates an entity relationship diagram (ERD)
-with the Umbrello UML Modeller, one has to store its
-XML description in XMI format (as offered by Umbrello)
-in a file, e.g., \ccode{myerd.xmi}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry myerd.xmi
-\end{curry}
-(\code{erd2curry} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-If \code{MyData} is the name of the ERD, the Curry program file
-\ccode{MyData.curry} is generated containing all the necessary
-database access code as described in \cite{BrasselHanusMueller08PADL}.
-
-If one does not want to use the Umbrello UML Modeller,
-one can also create a textual description of the ERD
-as a Curry term of type \code{ERD}
-(w.r.t.\ the type definition given in module
-\code{\cyshome/tools/erd2curry/ERD.curry})
-and store it in some file, e.g., \ccode{myerd.term}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry -t myerd.term
-\end{curry}
-%
-There is also the possibility to visualize an ERD term
-as a graph with the graph visualization program \code{dotty}
-(for this purpose, it might be necessary to adapt the definition
-of \code{dotviewcommand} in your \ccode{.kics2rc} file,
-see Section~\ref{sec-customization},
-according to your local environment).
-This can be done by the command
-\begin{curry}
-erd2curry -v myerd.term
-\end{curry}
-
-\paragraph{Inclusion in the Curry application:}
-To compile the generated database code, either
-include the directory \code{\cyshome/tools/erd2curry}
-into your Curry load path
-(e.g., by setting  the environment variable
-\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
-or copy the file
-\code{\cyshome/tools/erd2curry/ERDGeneric.curry}
-into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with \CYS
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of \CYS and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of \CYS:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{\cyshome/cpns/start} and
-\code{\cyshome/cpns/stop}.
-If the demon is already running, the command \code{\cyshome/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
-\addcontentsline{toc}{section}{Bibliography}
-\bibliography{mh}
-\bibliographystyle{plain}
-
-\newpage
-\appendix
-
-\section{Libraries of the \CYS Distribution}
-\label{sec:libraries}
-
-{\setlength{\parindent}{0.0cm}
-
-The \CYS distribution comes with an extensive collection
-of libraries for application programming.
-The libraries for meta-programming by representing
-Curry programs as datatypes in Curry are described
-in the following subsection in more detail.
-The complete set of libraries with all exported types and functions
-are described in the further subsections.
-For a more detailed online documentation of all libraries of \CYS,
-see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
-
-\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
-\label{sec-flatcurry}
-
-\index{AbstractCurry}
-\index{FlatCurry}
-To support meta-programming, i.e., the manipulation of Curry programs
-in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{\cyshome/lib/meta},
-which define datatypes for the representation
-of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
-where local function definitions are replaced by global definitions
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
-program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
-programs that are more oriented towards the source program.
-
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
-These actions parse the corresponding source program and return
-a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
-
-Since all datatypes are explained in detail in these modules,
-we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
-of these modules.
-
-As an example, consider a program file \ccode{test.curry}
-containing the following two lines:
-\begin{curry}
-rev []     = []
-rev (x:xs) = (rev xs) ++ [x]
-\end{curry}
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
-following term:
-\begin{curry}
- (Prog "test"
-  ["Prelude"]
-  []
-  [Func ("test","rev") 1 Public
-        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
-                  (TCons ("Prelude","[]") [(TVar 0)]))
-        (Rule [0]
-           (Case Flex (Var 1)
-              [Branch (Pattern ("Prelude","[]") [])
-                  (Comb ConsCall ("Prelude","[]") []),
-               Branch (Pattern ("Prelude",":") [2,3])
-                  (Comb FuncCall ("Prelude","++")
-                        [Comb FuncCall ("test","rev") [Var 3],
-                         Comb ConsCall ("Prelude",":")
-                              [Var 2,Comb ConsCall ("Prelude","[]") []]
-                        ])
-              ]))]
-  []
- )
-\end{curry}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Definitions in order to LaTeX documents generated by "currydoc --tex"
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
-\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
-\newcommand{\currytypesstop}{}
-\newcommand{\currytypesynstart}[2]{{\tt type #2}\pindex{#1} \begin{quote}}
-\newcommand{\currytypesynstop}{\end{quote}}
-\newcommand{\currydatastart}[1]{{\tt data #1}\pindex{#1} \begin{quote}}
-\newcommand{\currydatacons}{\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:}}
-\newcommand{\currydatastop}{\end{itemize}}
-\newcommand{\curryconsstart}[2]{\item {\tt #1~::~#2}\par}
-\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
-\newcommand{\curryfuncstop}{}
-\newcommand{\curryfunctionstart}[2]{#2\pindex{#1}\begin{quote}}
-\newcommand{\curryfunctionstop}{\end{quote}}
-\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
-
-\subsection{General Libraries}
-
-\input{lib/AllSolutions}
-\input{lib/Assertion}
-\input{lib/Char}
-\input{lib/Combinatorial}
-\input{lib/Constraint}
-\input{lib/CSV}
-\input{lib/Directory}
-\input{lib/FileGoodies}
-\input{lib/Float}
-\input{lib/Global}
-\input{lib/GUI}
-\input{lib/Integer}
-\input{lib/IO}
-\input{lib/IOExts}
-\input{lib/JavaScript}
-\input{lib/KeyDatabaseSQLite}
-\input{lib/List}
-\input{lib/Maybe}
-\input{lib/NamedSocket}
-\input{lib/Parser}
-\input{lib/Pretty}
-\input{lib/Profile}
-\input{lib/PropertyFile}
-\input{lib/Read}
-\input{lib/ReadNumeric}
-\input{lib/ReadShowTerm}
-\input{lib/SetFunctions}
-\input{lib/SearchTree}
-\input{lib/Socket}
-\input{lib/System}
-\input{lib/Time}
-\input{lib/Unsafe}
-
-\subsection{Data Structures and Algorithms}
-
-\input{lib/Array}
-\input{lib/Dequeue}
-\input{lib/FiniteMap}
-\input{lib/GraphInductive}
-\input{lib/Random}
-\input{lib/RedBlackTree}
-\input{lib/SetRBT}
-\input{lib/Sort}
-\input{lib/TableRBT}
-\input{lib/Traversal}
-
-\subsection{Libraries for Web Applications}
-
-\input{lib/CategorizedHtmlList}
-\input{lib/HTML}
-\input{lib/HtmlParser}
-\input{lib/Mail}
-\input{lib/Markdown}
-\input{lib/WUI}
-\input{lib/URL}
-\input{lib/XML}
-\input{lib/XmlConv}
-
-\subsection{Libraries for Meta-Programming}
-
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
-\input{lib/CompactFlatCurry}
-\input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
-
-} % end setlength parindent
-
-\newpage
-
-\input{markdown_syntax}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with \CYS,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of \CYS create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with \CYS.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with \CYS
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-\CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with \CYS
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, \CYS has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Thus, if an external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\begin{curry}
-f external
-\end{curry}
-below the type declaration for the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the \CYS compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, \CYS follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\begin{curry}
-sconc :: String -> String -> String
-\end{curry}
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal form.
-This is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\begin{currynomath}
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $## s1) $## s2
-
-prim_sconc :: String -> String -> String
-prim_sconc external
-\end{currynomath}
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of \CYS.
-The \CYS compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\begin{curry}
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> ConstStore
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
-\end{curry}
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the programmer.
-The additional argument of type \code{ConstStore}
-represents the current set of constraints when this
-operation is called. This argument is intended to provide
-a more efficient access to binding constraints and can be
-ignored in standard operations.
-
-If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The \CYS compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\begin{curry}
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> ConstStore -> CP.OP_List CP.C_Char
-\end{curry}
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of \CYS
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-(note that we simply ignore the final argument representing the
-constraint store)
-\begin{curry}
-external_d_C_prim_sconc s1 s2 _ =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\end{curry}
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the \CYS implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
-For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\begin{currynomath}
-printString :: String -> IO ()
-printString s = prim_printString $## s
-
-prim_printString :: String -> IO ()
-prim_printString external
-\end{currynomath} % $
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
-\begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
-                              -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-\end{curry}
-When we compile the Curry module \code{String},
-\CYS combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, \CYS transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\begin{curry}
-(<&>) :: Int -> Int -> Int
-(<&>) external
-\end{curry}
-the generated Haskell module contains the code
-\begin{curry}
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> ConstStore -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
-\end{curry}
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
-\addcontentsline{toc}{section}{Index}
-\printindex
-
-\end{document}
-
-% LocalWords:  KiCS

### 2013-07-12 docs/src/libraries.tex
Split manual into separate tex files

--- 2013-03-07 docs/src/manual.tex
@@
-\documentclass[11pt,fleqn]{article}
-
-\usepackage{latexsym}
-\usepackage{makeidx}
-\usepackage{url}
-\usepackage{xspace}
-\usepackage{graphicx}
-
-\input{version}
-
-%%% ------------------------------------------------------------------
-
-\usepackage[colorlinks,linkcolor=blue]{hyperref}
-\hypersetup{bookmarksopen=true}
-\hypersetup{bookmarksopenlevel=0}
-\hypersetup{pdftitle={KiCS2: The Kiel Curry System Version 2}}
-\hypersetup{pdfauthor={Michael Hanus}}
-%\hypersetup{pdfstartview=Title}
-\hypersetup{pdfstartview=FitH}
-\usepackage{thumbpdf}
-
-%%% ------------------------------------------------------------------
-
-\setlength{\textwidth}{16.5cm}
-\setlength{\textheight}{23cm}
-\renewcommand{\baselinestretch}{1.1}
-\setlength{\topmargin}{-1cm}
-\setlength{\oddsidemargin}{0cm}
-\setlength{\evensidemargin}{0cm}
-\setlength{\marginparwidth}{0.0cm}
-\setlength{\marginparsep}{0.0cm}
-
-\newlength{\figurewidth}
-\setlength{\figurewidth}{\textwidth}
-\addtolength{\figurewidth}{-0.4cm}
-
-% font for program texts
-\renewcommand{\tt}{\usefont{OT1}{cmtt}{m}{n}\selectfont}
-\newcommand{\codefont}{\small\tt}
-
-\usepackage{listings}
-\lstset{aboveskip=1.5ex,
-        belowskip=1.5ex,
-        showstringspaces=false, % no special string space
-        mathescape=true,
-        flexiblecolumns=false,
-        xleftmargin=2ex,
-        basewidth=0.52em,
-        basicstyle=\small\ttfamily}
-
-\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
-       }
-
-\lstnewenvironment{curry}{}{}
-\lstnewenvironment{currynomath}{\lstset{mathescape=false}}{} % Curry w/o math
-\newcommand{\listline}{\vrule width0pt depth1.75ex}
-
-% program text in normal text
-\newcommand{\code}[1]{\mbox{\codefont #1}}
-
-% program text in normal text with apostrophs
-\newcommand{\ccode}[1]{``\code{#1}''}
-
-\newcommand{\pindex}[1]{\index{#1@{\tt #1}}}  % program elements in index
-
-\newcommand{\todo}[1]{\fbox{\sc To do: #1}}
-\newcommand{\CYS}{KiCS2\xspace} % name of Curry system described here
-\newcommand{\cyshome}{\mbox{\textit{kics2home}}\xspace} % symbolic installation directory
-\newcommand{\cb}{CurryBrowser\xspace}
-
-% allow underscores in programs:
-\catcode`\_=\active
-\let_=\sb
-\catcode`\_=12
-
-% produce an index:
-\makeindex
-
-\begin{document}
-\sloppy
-
-\begin{titlepage}
-\pdfbookmark[1]{Title}{Title}
-\begin{center}
-\fbox{
-\begin{minipage}[t]{\figurewidth}
-\begin{center}\vspace{10ex}
-{\Huge\bf KiCS2}\\[4ex]
-{\huge The Kiel Curry System (Version 2)}\\[7ex]
-{\huge User Manual}\\[4ex]
-\kicsversiondate\\[6ex]
-\Large
-Michael Hanus$^1$ [editor] \\[3ex]
-{\large Additional Contributors:}\\[2ex]
-Bernd Bra\ss{}el$^2$ \\
-Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
-\normalsize
-(1) University of Kiel, Germany, {\tt mh@informatik.uni-kiel.de} \\
-(2) University of Kiel, Germany, {\tt bbr@informatik.uni-kiel.de} \\
-(3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
-\end{center}
-\end{minipage}}
-\end{center}
-\end{titlepage}
-
-\pdfbookmark[1]{Contents}{Contents}
-\tableofcontents
-
-\newpage
-
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current \CYS implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, \CYS also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
-Therefore, this document only explains the use of the different
-components of \CYS
-and the differences and restrictions of \CYS
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of \CYS
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of \CYS}
-
-\subsection{Installation}
-\label{sec-install}
-
-This version of \CYS has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems.
-
-Installation instructions for \CYS can be found in
-the file \code{INSTALL.txt} stored in the \CYS installation directory.
-Note that there are two possibilities to install \CYS:
-\begin{description}
-\item[Global installation:]\index{global installation}\index{installation!global}
-\CYS is installed in some global system directory where users
-have no write permission. In this case, some options for experimenting
-with \CYS (like \code{supply} or \code{ghc}, see below)
-are not available (since they require the recompilation of parts of
-the installed system).
-\item[Local installation:]\index{local installation}\index{installation!local}
-\CYS is installed in some local user directory where the user
-has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the \CYS installation is set as follows:
-\begin{lstlisting}
-GLOBALINSTALL=no
-\end{lstlisting}
-In this case, all options of \CYS are available.
-\end{description}
-In the following, \cyshome denotes the installation directory
-of the \CYS installation.
-
-\subsection{General Use}
-\label{sec-general}
-
-All executables required to use the different components
-of \CYS are stored in the directory \code{\cyshome/bin}.
-You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=\cyshome/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with \CYS creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\begin{curry}
-cleancurry
-\end{curry}
-(this is a shell script stored in the \code{bin} directory of the
-\CYS installation, see above).
-The command
-\begin{curry}
-cleancurry -r
-\end{curry}
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with \CYS:
-\begin{itemize}
-\item
-\index{singleton variables}\index{variables!singleton}
-\emph{Singleton pattern variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-\CYS translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-In general, tabulator stops should be avoided in source programs.
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
-based on the suspension of expressions
-containing free variables are not yet supported.
-\CYS supports \emph{value generators} for free variables
-so that a free variable is instantiated when its value is demanded.
-For instance, the initial expression
-\begin{curry}
-x == True where x free
-\end{curry}
-is non-deterministically evaluated to \code{False} and \code{True}
-by instantiating \code{x} to \code{False} and \code{True}, respectively.
-Thus, a computation is never suspended due to free variables.
-This behavior also applies to free variables of primitive types
-like integers. For instance, the initial expression
-\begin{curry}
-x*y=:=1 where x,y free
-\end{curry}
-is non-deterministically evaluated to the two solutions
-\begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
-\end{curry}
-\item
-Unification is performed without an occur check.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in \CYS}
-\label{sec-modules}
-
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-\CYS searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a \CYS session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of \CYS}
-\label{sec-kics2}
-
-This section describes the interactive environment
-\CYS\index{\CYS}
-that supports the development of applications written in Curry.
-The implementation of \CYS contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking \CYS}
-\label{sec:invoke-kics2}
-
-To start \CYS, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{\cyshome/bin} where \cyshome is the installation directory
-of \CYS).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{\cyshome/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke \CYS with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\begin{curry}
-kics2 :load Mod :add List
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\begin{curry}
-kics2 :load Mod :eval config
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\begin{curry}
-kics2 :load Mod :save :quit
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, creates an executable,
-and terminates \CYS. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of \CYS}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of \CYS are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs.
-In order to see the results of their bindings,\footnote{Currently,
-bindings are only printed if the initial expression is not an I/O action
-(i.e., not of type \ccode{IO\ldots})
-and there are not more than ten free variables
-in the initial expression.}
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
-\end{curry}
-in order to obtain the following three possible bindings:
-\begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
-\end{curry}
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\begin{curry}
-xs++ys=:=[z] where xs,ys,z free
-\end{curry}
-they are specialized to the type \ccode{()}
-(since the current implementation of \CYS does not support
-computations with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of \CYS.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the \CYS environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current \CYS process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Options of \CYS}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of \CYS.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices $n$}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression
-in a tree-like structure.
-This mode is only useful for debugging or
-understanding the implementation of non-deterministic
-evaluations used in \CYS.
-If the optional argument $n$ is provided,
-the tree is shown up to depth $n$.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-(not available in global installations, see Section~\ref{sec-install})
-Use implementation $i$ as the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the following values for $i$ are supported:
-\begin{description}
-\item[\code{integer}:]\pindex{integer}
-Use unbounded integers as choice identifiers.
-This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc}:]\pindex{ghc}
-Use a more sophisticated implementation of choice
-identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
-provided by the Glasgow Haskell Compiler.
-\item[\code{pureio}:]\pindex{pureio}
-Use IO references (i.e., memory cells) for choice identifiers.
-This is the most efficient implementation for top-level depth-first search
-but cannot be used for more sophisticated search methods
-like encapsulated search.
-\item[\code{ioref} (default):]\pindex{ioref}
-Use a mixture of \code{ghc} and \code{pureio}.
-IO references are used for top-level depth-first search
-and \code{ghc} identifiers are used for encapsulated search methods.
-\end{description}
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-Thus, one has also the possibility to terminate the
-enumeration of all values after having seen some values.
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the \CYS compiler.
-For instance, setting the option
-\begin{curry}
-:set cmp -O 0
-\end{curry}
-has the effect that all optimizations performed by the \CYS compiler
-are turned off.
-
-\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-Define additional options passed to the Glasgow Haskell Compiler (GHC)
-when the generated Haskell programs are compiled.
-One has to be careful when using such options.
-For instance, in a global installation of \CYS
-(see Section~\ref{sec-install}), libraries are pre-compiled
-so that inconsistencies might occur if compilation options
-might be changed.
-
-It is safe to pass specific GHC linking options.
-For instance, to enforce the static linking of libraries
-in order to generate an executable (see command \ccode{:save})
-that can be executed in another environment, one could set the
-options
-\begin{curry}
-:set ghc -static -optl-static -optl-pthread
-\end{curry}
-
-Other options are useful for experimental purposes,
-but those should be used only in local installations
-(see Section~\ref{sec-install}) to avoid inconsistent
-target codes for different libraries.
-For instance, setting the option
-\begin{curry}
-:set ghc -DDISABLE_CS
-\end{curry}
-has the effect that the constraint store used to enable
-an efficient access to complex bindings is disabled.
-Similarly,
-\begin{curry}
-:set ghc -DSTRICT_VAL_BIND
-\end{curry}
-has the effect that expressions in a unification constraint
-(\code{=:=}) are always fully evaluated
-(instead of the evaluation to a head normal form only)
-before unifying both sides.
-Since these options influence the compilation of the run-time system,
-one should also enforce the recompilation of Haskell programs
-by the GHC option \ccode{-fforce-recomp}, e.g., one should set
-\begin{curry}
-:set ghc -DDISABLE_CS -fforce-recomp
-\end{curry}
-
-\item[\fbox{\code{rts $opts$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\begin{curry}
-:set rts -H512m
-\end{curry}
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\item[\fbox{\code{args $arguments$}}]\pindex{args}
-Define run-time arguments passed to the executable
-generated by the Glasgow Haskell Compiler.
-For instance, setting the option
-\begin{curry}
-:set args first second
-\end{curry}
-has the effect that the I/O operation \code{getArgs}
-(see library \code{System} (Section~\ref{Library:System})
-returns the value \code{["first","second"]}.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-\CYS requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\begin{curry}
-{-# KiCS2_OPTION $opt$ #-}
-\end{curry}
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the \CYS command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\begin{curry}
-{-# KiCS2_OPTION ids #-}
-{-# KiCS2_OPTION +ghci #-}
-{-# KiCS2_OPTION v2 #-}
-module M where
-$\ldots$
-\end{curry}
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of \CYS (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call \CYS with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of \CYS to your own preferences,
-there is a configuration file which is read by \CYS when it is invoked.
-When you start \CYS for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started \CYS for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of \CYS contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into \CYS
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{\cyshome/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
-\section{Extensions}
-\label{sec-extensions}
-
-\CYS supports some extensions in Curry programs that are not (yet)
-part of the definition of Curry. These extensions are described below.
-
-\subsection{Recursive Variable Bindings}
-
-Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
-For instance, the declaration
-\begin{curry}
-ones5 = let ones = 1 : ones
-         in take 5 ones
-\end{curry}
-introduces the local variable \code{ones} which is bound
-to a \emph{cyclic structure}\index{cyclic structure}
-representing an infinite list of \code{1}'s.
-Similarly, the definition
-\begin{curry}
-onetwo n = take n one2
- where
-   one2 = 1 : two1
-   two1 = 2 : one2
-\end{curry}
-introduces a local variables \code{one2} that represents
-an infinite list of alternating \code{1}'s and \code{2}'s
-so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.
-
-\subsection{Functional Patterns}
-
-Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
-defining operations with functional patterns avoids problems
-caused by strict equality (\ccode{=:=}) and leads to programs
-that are potentially more efficient.
-
-Consider the definition of an operation to compute the last element
-of a list \code{xs} based on the prelude operation \ccode{++}
-for list concatenation:
-\begin{curry}
-last xs | _++[y] =:= xs  = y   where y free
-\end{curry}
-Since the equality constraint \ccode{=:=} evaluates both sides
-to a constructor term, all elements of the list \code{xs} are
-fully evaluated in order to satisfy the constraint.
-
-Functional patterns can help to improve this computational behavior.
-A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
-is a function call at a pattern position. With functional patterns,
-we can define the operation \code{last} as follows:
-\begin{curry}
-last (_++[y]) = y
-\end{curry}
-This definition is not only more compact but also avoids the complete
-evaluation of the list elements: since a functional pattern is considered
-as an abbreviation for the set of constructor terms obtained by all
-evaluations of the functional pattern to normal form (see
-\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
-definition is conceptually equivalent to the set of rules
-\begin{curry}
-last [y] = y
-last [_,y] = y
-last [_,_,y] = y
-$\ldots$
-\end{curry}
-which shows that the evaluation of the list elements is not demanded
-by the functional pattern.
-
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
-is interpreted as a functional pattern if $f$ is not a visible constructor
-but a defined function that is visible in the scope of the pattern.
-
-It is also possible to combine functional patterns with
-as-patterns.\index{as-pattern}\pindex{"@}
-Similarly to the meaning of as-patterns
-in standard constructor patterns,
-as-patterns in functional patterns are interpreted
-as a sequence of pattern matching where the variable of the as-pattern
-is matched before the given pattern is matched.
-This process can be described by introducing an auxiliary operation
-for this two-level pattern matching process.
-For instance, the definition
-\begin{curry}
-f (_ ++ x@[(42,_)] ++ _) = x
-\end{curry}
-is considered as syntactic sugar for the expanded definition
-\begin{curry}
-f (_ ++ x ++ _) = f' x
- where
-  f' [(42,_)] = x
-\end{curry}
-However, as-patterns are usually implemented
-in a more efficient way without introducing auxiliary operations.
-
-\subsection{Order of Pattern Matching}
-
-Curry allows multiple occurrences of pattern variables
-in standard patterns. These are an abbreviation of equational constraints
-between pattern variables.
-Functional patterns might also contain multiple occurrences of
-pattern variables.
-For instance, the operation
-\begin{curry}
-f (_++[x]++_++[x]++_) = x
-\end{curry}
-returns all elements with at least two occurrences in a list.
-
-If functional patterns as well as multiple occurrences of
-pattern variables occur in a pattern defining an operation,
-there are various orders to match an expression against such
-an operation. In the current implementation, the order
-is as follows:
-\begin{enumerate}
-\item Standard pattern matching: First, it is checked whether
-the constructor patterns match. Thus, functional patterns
-and multiple occurrences of pattern variables are ignored.
-\item Functional pattern matching: In the next phase,
-functional patterns are matched but occurrences of standard
-pattern variables in the functional patterns are ignored.
-\item Non-linear patterns: If standard and functional pattern matching
-is successful, the equational constraints which correspond
-to multiple occurrences pattern variables are solved.
-\item Guards: Finally, the guards supplied by the programmer
-are checked.
-\end{enumerate}
-The order of pattern matching should not influence the computed
-result. However, it might have some influence on the termination
-behavior of programs, i.e., a program might not terminate
-instead of finitely failing.
-In such cases, it could be necessary to consider the influence
-of the order of pattern matching. Note that other orders of pattern matching
-can be obtained using auxiliary operations.
-
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are 
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there. 
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\begin{curry}
-type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
-\end{curry}
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$ 
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\begin{curry}
-type Person = {name :: String, age :: Int}
-type Address = {person :: Person, street :: String, city :: String}
-type Branch a b = {left :: a, right :: b}
-\end{curry}
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\begin{curry}
-type Address = {person :: Person, street,city :: String}
-\end{curry}
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\begin{curry}
-type Address = {street,city :: String, person :: Person}
-\end{curry}
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\begin{curry}
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\end{curry}
-\begin{curry}
-getName :: Person -> String
-getName $\ldots$
-\end{curry}
-
-Labels can only be used in the context of
-records. They do not share the name space with 
-functions/constructors/variables or type constructors/type variables. 
-For instance it is possible to use 
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually 
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\begin{curry}
-{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
-\end{curry}
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with 
-exactly one assignment. Examples for record constructions are
-\begin{curry}
-{name := "Johnson", age := 30}     -- generates a record of type 'Person'
-{left := True, right := 20}        -- generates a record of type 'Branch'
-\end{curry}
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\begin{curry}
-{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
-\end{curry}
-Unlike labeled fields in record type declarations, 
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\begin{curry}
-{person := {name := "Smith", age := 20},   -- generates a record of
- street := "Main Street",                  -- type 'Address'
- city   := "Springfield"}
-\end{curry}
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records. 
-It has the following form:
-\begin{curry}
-$r$ :> $l$
-\end{curry}
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\begin{curry}
-pers :> name
-\end{curry}
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\begin{curry}
-addr :> person :> age
-\end{curry}
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}).
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\begin{curry}
-{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
-\end{curry}
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration 
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record. 
-Assignments can occur in an arbitrary order. It is not allowed to 
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\begin{curry}
-{name := "Scott", age := 25 | pers}
-{person := {name := "Scott", age := 25 | pers} | addr}
-\end{curry}
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\begin{curry}
-{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
-{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
-\end{curry}
-In both cases each label $l_i$ is specified with a pattern $p_i$. 
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern. 
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the 
-patterns of the specified labels are matched against 
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\begin{curry}
-isSmith30 :: Person -> Bool
-isSmith30 {name = "Smith", age = 30} = True
-\end{curry}
-\begin{curry}
-startsWith :: Char -> Person -> Bool
-startsWith c {name = (d:_) | _} = c == d
-\end{curry}
-\begin{curry}
-getPerson :: Address -> Person
-getPerson {person = p | _} = p
-\end{curry}
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways 
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent 
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
-
-\newpage
-
-\input{currydoc}
-
-\newpage
-
-\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
-\label{sec-currybrowser}
-
-\cb is a tool to browse through the modules and functions
-of a Curry application, show them in various formats,
-and analyze their properties.\footnote{Although \cb is
-implemented in Curry, some functionalities of it require an
-installed graph visualization tool (dot \url{http://www.graphviz.org/}),
-otherwise they have no effect.}
-Moreover, it is constructed in a way so that
-new analyzers can be easily connected to \cb.
-A detailed description of the ideas behind this tool can be
-found in \cite{Hanus05WCFLP,Hanus06WLPE}.
-
-\cb is part of the \CYS distribution and can be
-started in two ways:
-\begin{itemize}
-\item
-In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
-\item
-In the \CYS environment after loading the module
-\code{mod} and typing the command \ccode{:browse}.
-\end{itemize}
-Here, \ccode{mod} is the name of the main module of a Curry application.
-After the start, \cb loads the interfaces of the main
-module and all imported modules before a GUI is created
-for interactive browsing.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.7]{currybrowser.jpg}
-\end{center}
-\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
-\end{figure}
-%
-To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
-shows a snapshot of its use on a particular application
-(here: the implementation of \cb).
-The upper list box in the left column shows the modules and their imports
-in order to browse through the modules of an application.
-Similarly to directory browsers, the list of imported modules of a module
-can be opened or closed by clicking.
-After selecting a module in the list of modules, its source code,
-interface, or various other formats of the module can be shown
-in the main (right) text area. For instance, one can show
-pretty-printed versions of the intermediate flat programs (see below)
-in order to see how local function definitions are translated by lambda lifting
-\cite{Johnsson85}
-or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
-Since Curry is a language with parametric polymorphism and type inference,
-programmers often omit the type signatures when defining functions.
-Therefore, one can also view (and store) the selected module as source code where
-missing type signatures are added.
-
-Below the list box for selecting modules, there is a menu
-(``Analyze selected module'') to analyze all functions
-of the currently selected module at once. This is useful
-to spot some functions of a module that could be problematic
-in some application contexts, like functions that are impure (i.e., the result
-depends on the evaluation time) or partially defined (i.e.,
-not evaluable on all ground terms).
-If such an analysis is selected,
-the names of all functions are shown in the
-lower list box of the left column (the ``function list'')
-with prefixes indicating the properties of the individual functions.
-
-The function list box can be also filled with functions
-via the menu ``Select functions''. For instance, all functions
-or only the exported functions defined in the currently selected
-module can be shown there, or all functions from different modules
-that are directly or indirectly called from
-a currently selected function.
-This list box is central to focus on a function in the
-source code of some module or to analyze some function,
-i.e., showing their properties. In order to focus on a function,
-it is sufficient to check the ``focus on code'' button.
-To analyze an individually selected function, one can
-select an analysis from the list of available program analyses
-(through the menu ``Select analysis'').
-In this case, the analysis results are either shown
-in the text box below the main text area
-or visualized by separate tools, e.g., by a graph drawing tool for
-visualizing call graphs.
-Some analyses are local, i.e., they need only to consider the local definition
-of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
-``Pattern completeness''),
-where other analyses are global, i.e.,
-they consider the definitions of all functions directly or indirectly called
-by this function (e.g., ``Depends on,'' ``Solution complete,''
-``Set-valued'').
-%
-Finally, there are a few additional tools integrated into \cb,
-for instance, to visualize the import relation between all modules
-as a dependency graph. These tools are available through the ``Tools'' menu.
-
-More details about the use of \cb and all built-in analyses
-are available through the ``Help'' menu of \cb.
-
-\newpage
-
-\section{CurryTest: A Tool for Testing Curry Programs}
-\label{sec-currytest}
-
-CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the \CYS distribution to write
-and run repeatable tests. CurryTest simplifies the task
-of writing test cases for a module and executing them.
-The tool is easy to use. Assume one has implemented a module \code{MyMod}
-and wants to write some test cases to test its functionality,
-making regression tests in future versions, etc.
-For this purpose, there is a system library \code{Assertion}
-(Section~\ref{Library:Assertion}) which
-contains the necessary definitions for writing tests.
-In particular, it exports an abstract polymorphic type \ccode{Assertion a}
-together with the following operations:
-\begin{curry}
-assertTrue      :: String -> Bool -> Assertion ()
-assertEqual     :: String -> a -> a -> Assertion a
-assertValues    :: String -> a -> [a] -> Assertion a
-assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
-assertIO        :: String -> IO a -> a -> Assertion a
-assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\end{curry}
-The expression \ccode{assertTrue $s$ $b$}
-is an assertion (named $s$) that the expression $b$ has the value \code{True}.
-Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
-asserts that the expressions $e_1$ and $e_2$
-must be equal (i.e., \code{$e_1$==$e_2$} must hold),
-the expression \ccode{assertValues $s$ $e$ $vs$} asserts
-that $vs$ is the multiset of all values of $e$,
-and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
-that the constraint abstraction $c$ has the multiset of solutions $vs$.
-Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
-asserts that the I/O action $a$ yields the value $v$ whenever it is
-executed, and
-the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yield equal values.
-The name $s$ provided as a first argument in each assertion
-is used in the protocol produced by the test tool.
-
-One can define a test program by importing the module
-to be tested together with the module \code{Assertion} and defining
-top-level functions of type \code{Assertion} in this module
-(which must also be exported).
-As an example, consider the following program
-that can be used to test some list processing functions:
-\begin{curry}
-import List
-import Assertion
-
-test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-
-test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-
-test3 = assertSolutions "prefix" (\x -> x++_ =:= [1,2])
-                                 [[],[1],[1,2]]
-\end{curry}
-For instance, \code{test1} asserts that the result of evaluating the
-expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.
-
-We can execute a test suite by the command\pindex{currytest}
-\begin{curry}
-currytest TestList
-\end{curry}
-(\code{currytest} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-In our example, \ccode{TestList.curry} is the program containing the
-definition of all assertions. This has the effect
-that all exported top-level functions
-of type \code{Assertion} are tested (i.e., the corresponding
-assertions are checked) and the results
-(\ccode{OK} or failure) are reported together with the name of each assertion.
-%If failures occur, the complete test results are also
-%written into a file named \ccode{TestList.testlog}.''
-For our example above, we obtain the following successful protocol:
-\begin{curry}
-============================================================
-Testing module "TestList"...
-OK: ++
-OK: all
-OK: prefix
-All tests successfully passed.
-============================================================
-\end{curry}
-There is also a graphical interface that summarizes the results
-more nicely.
-In order to start this interface, one has to add the parameter
-\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\begin{curry}
-currytest --window TestList
-\end{curry}
-or
-\begin{curry}
-currytest -w TestList
-\end{curry}
-A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
-
-\begin{figure}%[t]
-\begin{center}
-\includegraphics[scale=0.7]{currytest.jpg}
-\end{center}
-\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
-\end{figure}
-
-\newpage
-
-\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
-\label{sec-erd2curry}
-
-ERD2Curry\index{ERD2Curry}\index{database programming}
-is a tool to generate Curry code to access and manipulate data
-persistently stored from
-entity relationship diagrams.\index{entity relationship diagrams}
-The idea of this tool is described in detail in
-\cite{BrasselHanusMueller08PADL}.
-Thus, we describe only the basic steps to use this tool
-in the following.
-
-If one creates an entity relationship diagram (ERD)
-with the Umbrello UML Modeller, one has to store its
-XML description in XMI format (as offered by Umbrello)
-in a file, e.g., \ccode{myerd.xmi}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry myerd.xmi
-\end{curry}
-(\code{erd2curry} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-If \code{MyData} is the name of the ERD, the Curry program file
-\ccode{MyData.curry} is generated containing all the necessary
-database access code as described in \cite{BrasselHanusMueller08PADL}.
-
-If one does not want to use the Umbrello UML Modeller,
-one can also create a textual description of the ERD
-as a Curry term of type \code{ERD}
-(w.r.t.\ the type definition given in module
-\code{\cyshome/tools/erd2curry/ERD.curry})
-and store it in some file, e.g., \ccode{myerd.term}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry -t myerd.term
-\end{curry}
-%
-There is also the possibility to visualize an ERD term
-as a graph with the graph visualization program \code{dotty}
-(for this purpose, it might be necessary to adapt the definition
-of \code{dotviewcommand} in your \ccode{.kics2rc} file,
-see Section~\ref{sec-customization},
-according to your local environment).
-This can be done by the command
-\begin{curry}
-erd2curry -v myerd.term
-\end{curry}
-
-\paragraph{Inclusion in the Curry application:}
-To compile the generated database code, either
-include the directory \code{\cyshome/tools/erd2curry}
-into your Curry load path
-(e.g., by setting  the environment variable
-\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
-or copy the file
-\code{\cyshome/tools/erd2curry/ERDGeneric.curry}
-into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with \CYS
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of \CYS and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of \CYS:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{\cyshome/cpns/start} and
-\code{\cyshome/cpns/stop}.
-If the demon is already running, the command \code{\cyshome/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
-\addcontentsline{toc}{section}{Bibliography}
-\bibliography{mh}
-\bibliographystyle{plain}
-
-\newpage
-\appendix
-
 \section{Libraries of the \CYS Distribution}
@@
 } % end setlength parindent
-
-\newpage
-
-\input{markdown_syntax}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with \CYS,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of \CYS create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with \CYS.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with \CYS
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-\CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with \CYS
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, \CYS has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Thus, if an external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\begin{curry}
-f external
-\end{curry}
-below the type declaration for the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the \CYS compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, \CYS follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\begin{curry}
-sconc :: String -> String -> String
-\end{curry}
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal form.
-This is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\begin{currynomath}
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $## s1) $## s2
-
-prim_sconc :: String -> String -> String
-prim_sconc external
-\end{currynomath}
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of \CYS.
-The \CYS compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\begin{curry}
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> ConstStore
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
-\end{curry}
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the programmer.
-The additional argument of type \code{ConstStore}
-represents the current set of constraints when this
-operation is called. This argument is intended to provide
-a more efficient access to binding constraints and can be
-ignored in standard operations.
-
-If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The \CYS compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\begin{curry}
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> ConstStore -> CP.OP_List CP.C_Char
-\end{curry}
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of \CYS
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-(note that we simply ignore the final argument representing the
-constraint store)
-\begin{curry}
-external_d_C_prim_sconc s1 s2 _ =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\end{curry}
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the \CYS implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
-For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\begin{currynomath}
-printString :: String -> IO ()
-printString s = prim_printString $## s
-
-prim_printString :: String -> IO ()
-prim_printString external
-\end{currynomath} % $
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
-\begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
-                              -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-\end{curry}
-When we compile the Curry module \code{String},
-\CYS combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, \CYS transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\begin{curry}
-(<&>) :: Int -> Int -> Int
-(<&>) external
-\end{curry}
-the generated Haskell module contains the code
-\begin{curry}
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> ConstStore -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
-\end{curry}
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
-\addcontentsline{toc}{section}{Index}
-\printindex
-
-\end{document}
-
-% LocalWords:  KiCS

### 2013-07-12 docs/src/manual.tex
Split manual into separate tex files

--- 2013-03-07 docs/src/manual.tex
@@
 
-\newpage
+\include{preface}
+\include{overview}
+\include{usage}
+\include{extensions}
+\include{currydoc}
+\include{currybrowser}
+\include{currytest}
+\include{erd2curry}
+\include{problems}
 
-\addcontentsline{toc}{section}{Preface}
-\section*{Preface}
-
-This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
-Version \textbf{2}),
-an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
-based on compiling Curry programs into Haskell programs.
-Curry is a universal programming language aiming at the amalgamation
-of the most important declarative programming paradigms,
-namely functional programming and logic programming.  
-Curry combines in a seamless way features from functional programming
-(nested expressions, lazy evaluation, higher-order functions),
-logic programming (logical variables, partial data structures,
-built-in search), and concurrent programming (concurrent evaluation
-of constraints with synchronization on logical variables).
-The current \CYS implementation does not support concurrent
-constraints. Alternatively, one can write distributed applications
-by the use of sockets that can be registered and accessed
-with symbolic names.
-Moreover, \CYS also supports
-the high-level implementation of
-graphical user interfaces and web services
-(as described in more detail in
-\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).
-
-We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
-Therefore, this document only explains the use of the different
-components of \CYS
-and the differences and restrictions of \CYS
-(see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
-The basic ideas of the implementation of \CYS
-can be found in
-\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.
-
-\bigskip
-
-\subsection*{Acknowledgements}
-
-This work has been supported in part by the
-DFG grants Ha 2457/5-1 and Ha 2457/5-2.
-
-\newpage
-
-\section{Overview of \CYS}
-
-\subsection{Installation}
-\label{sec-install}
-
-This version of \CYS has been developed and
-tested on Linux systems.
-In principle, it should be also executable on other
-platforms on which a Haskell implementation (Glasgow Haskell Compiler
-and Cabal) exists, like in many Linux distributions, Sun Solaris,
-or Mac OS X systems.
-
-Installation instructions for \CYS can be found in
-the file \code{INSTALL.txt} stored in the \CYS installation directory.
-Note that there are two possibilities to install \CYS:
-\begin{description}
-\item[Global installation:]\index{global installation}\index{installation!global}
-\CYS is installed in some global system directory where users
-have no write permission. In this case, some options for experimenting
-with \CYS (like \code{supply} or \code{ghc}, see below)
-are not available (since they require the recompilation of parts of
-the installed system).
-\item[Local installation:]\index{local installation}\index{installation!local}
-\CYS is installed in some local user directory where the user
-has write permission and the option \code{GLOBALINSTALL}
-in the \code{Makefile} of the \CYS installation is set as follows:
-\begin{lstlisting}
-GLOBALINSTALL=no
-\end{lstlisting}
-In this case, all options of \CYS are available.
-\end{description}
-In the following, \cyshome denotes the installation directory
-of the \CYS installation.
-
-\subsection{General Use}
-\label{sec-general}
-
-All executables required to use the different components
-of \CYS are stored in the directory \code{\cyshome/bin}.
-You should add this directory
-to your path (e.g., by the \code{bash} command
-\ccode{export PATH=\cyshome/bin:\$PATH}).
-
-The source code of the Curry program
-must be stored in a file with the suffix \ccode{.curry},
-e.g., \code{prog.curry}. 
-Literate programs must be stored in files with the extension \ccode{.lcurry}.
-
-Since the translation of Curry programs with \CYS creates
-some auxiliary files (see Section~\ref{sec-auxfiles} for details),
-you need write permission
-in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
-The auxiliary files for all Curry programs in the current
-directory can be deleted by the command\pindex{cleancurry}
-\begin{curry}
-cleancurry
-\end{curry}
-(this is a shell script stored in the \code{bin} directory of the
-\CYS installation, see above).
-The command
-\begin{curry}
-cleancurry -r
-\end{curry}
-also deletes the auxiliary files in all subdirectories.
-
-\subsection{Restrictions}
-\label{sec-restrictions}
-
-There are a few minor restrictions on Curry programs
-when they are processed with \CYS:
-\begin{itemize}
-\item
-\index{singleton variables}\index{variables!singleton}
-\emph{Singleton pattern variables}, i.e., variables that occur only once
-in a rule, should be denoted as an anonymous variable \ccode{_},
-otherwise the parser will print a warning since this is a
-typical source of programming errors.
-\item
-\CYS translates all \emph{local declarations} into global functions with
-additional arguments (``lambda lifting'', see Appendix~D of the
-Curry language report).
-Thus, in the various run-time systems, the definition of
-functions with local declarations look different from
-their original definition (in order to see the result
-of this transformation, you can use the \cb, see
-Section~\ref{sec-currybrowser}).
-\item \index{tabulator stops}
-Tabulator stops instead of blank spaces in source files are
-interpreted as stops at columns 9, 17, 25, 33, and so on.
-In general, tabulator stops should be avoided in source programs.
-\item
-Encapsulated search\index{encapsulated search}: In order
-to allow the integration of non-deterministic computations
-in programs performing I/O at the top-level, \CYS provides
-the libraries \code{AllSolutions} (Section~\ref{Library:AllSolutions})
-and \code{SearchTree} (Section~\ref{Library:SearchTree})
-where one can defined search operators and various search strategies.
-The general definition of encapsulated search of the Curry report
-\cite{HanusSteiner98PLILP} is not supported.
-\item Concurrent computations\index{concurrency}
-based on the suspension of expressions
-containing free variables are not yet supported.
-\CYS supports \emph{value generators} for free variables
-so that a free variable is instantiated when its value is demanded.
-For instance, the initial expression
-\begin{curry}
-x == True where x free
-\end{curry}
-is non-deterministically evaluated to \code{False} and \code{True}
-by instantiating \code{x} to \code{False} and \code{True}, respectively.
-Thus, a computation is never suspended due to free variables.
-This behavior also applies to free variables of primitive types
-like integers. For instance, the initial expression
-\begin{curry}
-x*y=:=1 where x,y free
-\end{curry}
-is non-deterministically evaluated to the two solutions
-\begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
-\end{curry}
-\item
-Unification is performed without an occur check.
-\item
-There is currently no general connection to external constraint solvers.
-\end{itemize}
-
-\subsection{Modules in \CYS}
-\label{sec-modules}
-
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
-\CYS searches for imported modules in various directories.
-By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-This search path can be extended
-by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
-(which can be also set in a \CYS session by the option
-\ccode{:set path}\pindex{path}\pindex{:set path},
-see below)
-to a list of directory names separated by colons (\ccode{:}).
-In addition, a local standard search path
-can be defined in the \ccode{.kics2rc} file
-(see Section~\ref{sec-customization}).
-Thus, modules to be loaded are searched in the following
-directories (in this order, i.e., the first occurrence of a module file
-in this search path is imported):
-\begin{enumerate}
-\item Current working directory (\ccode{.}) or directory prefix
-of the main module (e.g., directory \ccode{/home/joe/curryprogs}
-if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
-\item The directories enumerated in the environment variable \code{CURRYPATH}.
-\item The directories enumerated in the \ccode{.kics2rc} variable
-      \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
-\end{enumerate}
-Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
-will be always implicitly imported to all modules if a module
-does not contain an explicit import declaration for the module
-\code{Prelude}.
-
-\newpage
-
-\section{Using the Interactive Environment of \CYS}
-\label{sec-kics2}
-
-This section describes the interactive environment
-\CYS\index{\CYS}
-that supports the development of applications written in Curry.
-The implementation of \CYS contains also a separate compiler
-which is automatically invoked by the interactive environment.
-
-\subsection{Invoking \CYS}
-\label{sec:invoke-kics2}
-
-To start \CYS, execute the command
-\ccode{kics2}\pindex{kics2}
-(this is a shell script stored in
-\code{\cyshome/bin} where \cyshome is the installation directory
-of \CYS).
-When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
-the prelude (\code{\cyshome/lib/Prelude.curry})
-is already loaded, i.e., all definitions in the prelude are accessible.
-Now you can type various commands (see next section)
-or an expression to be evaluated.
-
-One can also invoke \CYS with parameters.
-These parameters are usual a sequence of commands
-(see next section) that are executed before the user interaction
-starts. For instance, the invocation
-\begin{curry}
-kics2 :load Mod :add List
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and adds the additional
-module \code{List}.
-The invocation
-\begin{curry}
-kics2 :load Mod :eval config
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, and evaluates
-the operation \code{config} before the user interaction starts.
-As a final example, the invocation
-\begin{curry}
-kics2 :load Mod :save :quit
-\end{curry}
-starts \CYS, loads the main module \code{Mod}, creates an executable,
-and terminates \CYS. This invocation could be useful in ``make'' files
-for systems implemented in Curry.
-
-\subsection{Command of \CYS}
-\label{sec:kics2-commands}
-
-The {\bf most important commands} of \CYS are
-(it is sufficient to type a unique prefix of a command if it is unique,
-e.g., one can type \ccode{:r} instead of \ccode{:reload}):
-
-\begin{description}
-\item[\fbox{\code{:help}}]\pindex{:help}
-Show a list of all available commands.
-
-\item[\fbox{\code{:load $prog$}}]\pindex{:load}
-Compile and load the program stored in \code{$prog$.curry}
-together with all its imported modules.
-% If this file does not exist, the system looks for a FlatCurry
-% file \code{$prog$.fcy} and compiles from this intermediate representation.
-% If the file \code{$prog$.fcy} does not exists, too, the system looks
-% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
-% XML representation (compare command \ccode{:xml}\pindex{:xml}),
-% translates this into a FlatCurry file \code{$prog$.fcy}
-% and compiles from this intermediate representation.
-
-\item[\fbox{\code{:reload}}]\pindex{:reload}
-Recompile all currently loaded modules.
-
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
-
-\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
-and show the computed results.
-In the default mode, all results of
-non-deterministic computations are printed.
-One can also print first one result and the next result
-only if the user requests it. This behavior can be set by
-the option \code{interactive}\index{interactive} (see below).
-
-\textbf{Free variables in initial expressions} must be declared as
-in Curry programs.
-In order to see the results of their bindings,\footnote{Currently,
-bindings are only printed if the initial expression is not an I/O action
-(i.e., not of type \ccode{IO\ldots})
-and there are not more than ten free variables
-in the initial expression.}
-they must be introduced by a \ccode{where\ldots{}free} declaration.
-For instance, one can write
-\begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
-\end{curry}
-in order to obtain the following three possible bindings:
-\begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
-\end{curry}
-Without these declarations, an error is reported in order to
-avoid the unintended introduction of free variables in initial expressions
-by typos.
-
-If the free variables in the initial goal are of a polymorphic type,
-as in the expression
-\begin{curry}
-xs++ys=:=[z] where xs,ys,z free
-\end{curry}
-they are specialized to the type \ccode{()}
-(since the current implementation of \CYS does not support
-computations with polymorphic logic variables).
-
-\item[\fbox{:eval $expr$}]\pindex{:eval}
-Same as $expr$. This command might be useful when putting
-commands as arguments when invoking \code{kics2}.
-
-% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
-% Define the identifier $x$ as an abbreviation for the expression $expr$
-% which can be used in subsequent expressions. The identifier $x$
-% is visible until the next \code{load} or \code{reload} command.
-
-\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
-\end{description}
-%
-\bigskip
-%
-There are also a number of {\bf further commands} that are often
-useful:
-%
-\begin{description}
-\item[\fbox{\code{:type $expr$}}]\pindex{:type}
-Show the type of the expression $expr$.
-
-\item[\fbox{\code{:programs}}]\pindex{:programs}
-Show the list of all Curry programs that are available in the load path.
-
-\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
-Change the current working directory to $dir$.
-
-\item[\fbox{\code{:edit}}]\pindex{:edit}
-Load the source code of the current main module into a text editor.
-If the variable \code{editcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as an editor command, otherwise
-the environment variable \ccode{EDITOR} is used as the editor program.
-
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
-as in the command \ccode{:edit}.
-
-\item[\fbox{\code{:show}}]\pindex{:show}
-Show the source text of the currently loaded Curry program.
-If the variable \code{showcommand} is set in the
-configuration file \ccode{.kics2rc}
-(see Section~\ref{sec-customization}),
-its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
-
-\item[\fbox{\code{:show $m$}}]\pindex{:show}
-Show the source text of module $m$ which must be accessible
-via the current load path.
-
-\item[\fbox{\code{:source $f$}}]\pindex{:source}
-Show the source code of function $f$ (which must be visible
-in the currently loaded module) in a separate window.
-
-\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
-Show the source code of function $f$ defined in module $m$
-in a separate window.
-
-\item[\fbox{\code{:browse}}]\pindex{:browse}
-Start the CurryBrowser to analyze the currently loaded
-module together with all its imported modules
-(see Section~\ref{sec-currybrowser} for more details).
-
-\item[\fbox{\code{:interface}}]\pindex{:interface}
-Show the interface of the currently loaded
-module, i.e., show the names of all imported modules,
-the fixity declarations of all exported operators,
-the exported datatypes declarations and the types
-of all exported functions.
-
-\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
-Similar to \ccode{:interface}
-but shows the interface of the module $m$
-which must be in the load path of \CYS.
-
-\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
-Show all calls to imported functions in the currently loaded module.
-This might be useful to see which import declarations are really necessary.
-
-\item[\fbox{\code{:set $option$}}]\pindex{:set}
-Set or turn on/off a specific option
-of the \CYS environment (see \ref{sec:options} for a description
-of all options). Options are turned on by the prefix
-\ccode{+} and off by the prefix \ccode{-}. Options that can only
-be set (e.g., \code{path}) must not contain a prefix.
-
-\item[\fbox{\code{:set}}]\pindex{:set}
-Show a help text on the possible options
-together with the current values of all options.
-
-\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
-program as an executable evaluating the main expression \ccode{main}.
-The executable is stored in the file \code{Mod}
-if \code{Mod} is the name of the currently loaded main module.
-
-\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
-but the expression $expr$ (typically: a call to the main
-function) will be evaluated by the executable.
-
-\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
-The expression $expr$, which is typically of type \ccode{IO ()},
-is evaluated in an independent process which runs in
-parallel to the current \CYS process.
-All output and error messages from this new process are suppressed.
-This command is useful to test distributed Curry programs
-where one can start
-a new server process by this command. The new process
-will be terminated when the evaluation of the expression $expr$
-is finished.
-
-\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.
-
-\end{description}
-
-\subsection{Options of \CYS}
-\label{sec:options}
-
-The following options (which can be set by the command \ccode{:set})
-are currently supported:
-
-\begin{description}
-\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
-for loading modules to $path$.
-Note that this search path is only used for loading modules
-inside this invocation of \CYS.
-
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
-Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
-print only the first value or all values by interactively requesting them
-(see below for these options).
-
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
-
-\item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
-to compute and print the values of the initial expression.
-The initial depth bound is 100 and the depth-bound is
-doubled after each iteration.
-
-\item[\fbox{\code{ids $n$}}]\pindex{ids}
-Similarly to \code{ids} but use an initial depth bound of $n$.
-
-\item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
-to compute and print the values of the initial expression.
-The system chooses an appropriate number
-of threads according the current number of available processors.
-
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
-
-\item[\fbox{\code{choices $n$}}]\pindex{choices}
-Show the internal choice structure
-(according to the implementation described in
-\cite{BrasselHanusPeemoellerReck11})
-resulting from the complete evaluation of the main expression
-in a tree-like structure.
-This mode is only useful for debugging or
-understanding the implementation of non-deterministic
-evaluations used in \CYS.
-If the optional argument $n$ is provided,
-the tree is shown up to depth $n$.
-
-\item[\fbox{\code{supply $i$}}]\pindex{supply}
-(not available in global installations, see Section~\ref{sec-install})
-Use implementation $i$ as the identifier supply for choice structures
-(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
-Currently, the following values for $i$ are supported:
-\begin{description}
-\item[\code{integer}:]\pindex{integer}
-Use unbounded integers as choice identifiers.
-This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
-\item[\code{ghc}:]\pindex{ghc}
-Use a more sophisticated implementation of choice
-identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
-provided by the Glasgow Haskell Compiler.
-\item[\code{pureio}:]\pindex{pureio}
-Use IO references (i.e., memory cells) for choice identifiers.
-This is the most efficient implementation for top-level depth-first search
-but cannot be used for more sophisticated search methods
-like encapsulated search.
-\item[\code{ioref} (default):]\pindex{ioref}
-Use a mixture of \code{ghc} and \code{pureio}.
-IO references are used for top-level depth-first search
-and \code{ghc} identifiers are used for encapsulated search methods.
-\end{description}
-
-\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
-Set the verbosity level to $n$. The following values are allowed
-for $n$:
-\begin{description}
-\item[$n=0$:] Do not show any messages (except for errors).
-\item[$n=1$:] Show only messages of the front-end, like loading
-of modules.
-\item[$n=2$:]
-Show also messages of the back end, like compilation messages
-from the Haskell compiler.
-\item[$n=3$:]
-Show also intermediate messages and commands of the compilation
-process.
-\item[$n=4$:]
-Show also all intermediate results of the compilation process.
-\end{description}
-
-\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
-Turn on/off the interactive mode.
-In the interactive mode, the next non-deterministic value
-is only computed when the user requests it.
-Thus, one has also the possibility to terminate the
-enumeration of all values after having seen some values.
-
-\item[\fbox{\code{+/-first}}]\pindex{first}
-Turn on/off the first-only mode.
-In the first-only mode, only the first value
-of the main expression is printed (instead of all values).
-
-\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
-Turn on/off the optimization of the target program.
-
-\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
-Turn on/off the binding mode.
-If the binding mode is on (default),
-then the bindings of the free variables of the initial expression
-are printed together with the result of the expression.
-
-\item[\fbox{\code{+/-time}}]\pindex{time}
-Turn on/off the time mode. If the time mode is on,
-the cpu time and the elapsed time
-of the computation is always printed together with the result
-of an evaluation.
-
-\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
-Turn on/off the ghci mode.
-In the ghci mode, the initial goal is send to the interactive version
-of the Glasgow Haskell Compiler. This might result in a slower
-execution but in a faster startup time since the linker
-to create the main executable is not used.
-
-\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
-Define additional options passed to the \CYS compiler.
-For instance, setting the option
-\begin{curry}
-:set cmp -O 0
-\end{curry}
-has the effect that all optimizations performed by the \CYS compiler
-are turned off.
-
-\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
-Define additional options passed to the Glasgow Haskell Compiler (GHC)
-when the generated Haskell programs are compiled.
-One has to be careful when using such options.
-For instance, in a global installation of \CYS
-(see Section~\ref{sec-install}), libraries are pre-compiled
-so that inconsistencies might occur if compilation options
-might be changed.
-
-It is safe to pass specific GHC linking options.
-For instance, to enforce the static linking of libraries
-in order to generate an executable (see command \ccode{:save})
-that can be executed in another environment, one could set the
-options
-\begin{curry}
-:set ghc -static -optl-static -optl-pthread
-\end{curry}
-
-Other options are useful for experimental purposes,
-but those should be used only in local installations
-(see Section~\ref{sec-install}) to avoid inconsistent
-target codes for different libraries.
-For instance, setting the option
-\begin{curry}
-:set ghc -DDISABLE_CS
-\end{curry}
-has the effect that the constraint store used to enable
-an efficient access to complex bindings is disabled.
-Similarly,
-\begin{curry}
-:set ghc -DSTRICT_VAL_BIND
-\end{curry}
-has the effect that expressions in a unification constraint
-(\code{=:=}) are always fully evaluated
-(instead of the evaluation to a head normal form only)
-before unifying both sides.
-Since these options influence the compilation of the run-time system,
-one should also enforce the recompilation of Haskell programs
-by the GHC option \ccode{-fforce-recomp}, e.g., one should set
-\begin{curry}
-:set ghc -DDISABLE_CS -fforce-recomp
-\end{curry}
-
-\item[\fbox{\code{rts $opts$}}]\pindex{rts}
-Define additional run-time options passed to the executable
-generated by the Glasgow Haskell Compiler, i.e., the parameters
-\ccode{+RTS $o$ -RTS} are passed to the executable.
-For instance, setting the option
-\begin{curry}
-:set rts -H512m
-\end{curry}
-has the effect that the minimum heap size is set to 512 megabytes.
-
-\item[\fbox{\code{args $arguments$}}]\pindex{args}
-Define run-time arguments passed to the executable
-generated by the Glasgow Haskell Compiler.
-For instance, setting the option
-\begin{curry}
-:set args first second
-\end{curry}
-has the effect that the I/O operation \code{getArgs}
-(see library \code{System} (Section~\ref{Library:System})
-returns the value \code{["first","second"]}.
-
-\end{description}
-
-\subsection{Source-File Options}
-
-If the evaluation of operations in some main module loaded into
-\CYS requires specific options, like an iterative-deepening
-search strategy, one can also put these options into the source
-code of this module in order to avoid setting these options
-every time when this module is loaded.
-Such
-{\bf source-file options}\index{source-file option}\index{option!in source file}
-must occur before the module header, i.e., before the first declaration
-(module header, imports, fixity declaration, defining rules, etc)
-occurring in the module.
-Each source file option must be in a line of the form
-\begin{curry}
-{-# KiCS2_OPTION $opt$ #-}
-\end{curry}
-where $opt$ is an option that can occur in a \ccode{:set} command
-(compare Section~\ref{sec:options}).
-Such a line in the source code (which is a comment according to
-the syntax of Curry)
-has the effect that this option is set by the \CYS command
-\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
-as a main module. For instance, if a module starts with the
-lines
-\begin{curry}
-{-# KiCS2_OPTION ids #-}
-{-# KiCS2_OPTION +ghci #-}
-{-# KiCS2_OPTION v2 #-}
-module M where
-$\ldots$
-\end{curry}
-then the load command \ccode{:load M} will also
-set the options for iterative deepening, using \code{ghci}
-and verbosity level 2.
-
-\subsection{Command Line Editing}
-
-In order to have support for line editing or history functionality
-in the command line of \CYS (as often supported by the \code{readline}
-library), you should have the Unix command \code{rlwrap} installed
-on your local machine.
-If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
-If it should not be used (e.g., because it is executed
-in an editor with \code{readline} functionality), one can
-call \CYS with the parameter \ccode{--noreadline}
-(which must occur as the first parameter).
-
-\subsection{Customization}
-\label{sec-customization}
-
-In order to customize the behavior of \CYS to your own preferences,
-there is a configuration file which is read by \CYS when it is invoked.
-When you start \CYS for the first time, a standard version of
-this configuration file is copied with the name
-\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
-into your home directory. The file contains definitions
-of various settings, e.g., about showing warnings, using Curry extensions,
-programs etc.
-After you have started \CYS for the first time, look into this file
-and adapt it to your own preferences.
-
-\subsection{Emacs Interface}
-
-Emacs is a powerful programmable editor suitable for program development.
-It is freely available for many platforms
-(see \url{http://www.emacs.org}).
-The distribution of \CYS contains also a special
-\emph{Curry mode}\index{Curry mode}\index{Emacs}
-that supports the development of Curry programs in
-the Emacs environment.
-This mode includes support for syntax highlighting,
-finding declarations in the current buffer, and
-loading Curry programs into \CYS
-in an Emacs shell.
-
-The Curry mode has been adapted from a similar mode for Haskell programs.
-Its installation is described in the file \code{README}
-in directory \ccode{\cyshome/tools/emacs} which also contains
-the sources of the Curry mode and a short description about
-the use of this mode.
-
-\newpage
-
-\section{Extensions}
-\label{sec-extensions}
-
-\CYS supports some extensions in Curry programs that are not (yet)
-part of the definition of Curry. These extensions are described below.
-
-\subsection{Recursive Variable Bindings}
-
-Local variable declarations (introduced by \code{let}\pindex{let}
-or \code{where}\pindex{where}) can be (mutually) recursive in \CYS.
-For instance, the declaration
-\begin{curry}
-ones5 = let ones = 1 : ones
-         in take 5 ones
-\end{curry}
-introduces the local variable \code{ones} which is bound
-to a \emph{cyclic structure}\index{cyclic structure}
-representing an infinite list of \code{1}'s.
-Similarly, the definition
-\begin{curry}
-onetwo n = take n one2
- where
-   one2 = 1 : two1
-   two1 = 2 : one2
-\end{curry}
-introduces a local variables \code{one2} that represents
-an infinite list of alternating \code{1}'s and \code{2}'s
-so that the expression \code{(onetwo 6)} evaluates to \code{[1,2,1,2,1,2]}.
-
-\subsection{Functional Patterns}
-
-Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
-defining operations with functional patterns avoids problems
-caused by strict equality (\ccode{=:=}) and leads to programs
-that are potentially more efficient.
-
-Consider the definition of an operation to compute the last element
-of a list \code{xs} based on the prelude operation \ccode{++}
-for list concatenation:
-\begin{curry}
-last xs | _++[y] =:= xs  = y   where y free
-\end{curry}
-Since the equality constraint \ccode{=:=} evaluates both sides
-to a constructor term, all elements of the list \code{xs} are
-fully evaluated in order to satisfy the constraint.
-
-Functional patterns can help to improve this computational behavior.
-A \emph{functional pattern}\index{functional pattern}\index{pattern!functional}
-is a function call at a pattern position. With functional patterns,
-we can define the operation \code{last} as follows:
-\begin{curry}
-last (_++[y]) = y
-\end{curry}
-This definition is not only more compact but also avoids the complete
-evaluation of the list elements: since a functional pattern is considered
-as an abbreviation for the set of constructor terms obtained by all
-evaluations of the functional pattern to normal form (see
-\cite{AntoyHanus05LOPSTR} for an exact definition), the previous
-definition is conceptually equivalent to the set of rules
-\begin{curry}
-last [y] = y
-last [_,y] = y
-last [_,_,y] = y
-$\ldots$
-\end{curry}
-which shows that the evaluation of the list elements is not demanded
-by the functional pattern.
-
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
-is interpreted as a functional pattern if $f$ is not a visible constructor
-but a defined function that is visible in the scope of the pattern.
-
-It is also possible to combine functional patterns with
-as-patterns.\index{as-pattern}\pindex{"@}
-Similarly to the meaning of as-patterns
-in standard constructor patterns,
-as-patterns in functional patterns are interpreted
-as a sequence of pattern matching where the variable of the as-pattern
-is matched before the given pattern is matched.
-This process can be described by introducing an auxiliary operation
-for this two-level pattern matching process.
-For instance, the definition
-\begin{curry}
-f (_ ++ x@[(42,_)] ++ _) = x
-\end{curry}
-is considered as syntactic sugar for the expanded definition
-\begin{curry}
-f (_ ++ x ++ _) = f' x
- where
-  f' [(42,_)] = x
-\end{curry}
-However, as-patterns are usually implemented
-in a more efficient way without introducing auxiliary operations.
-
-\subsection{Order of Pattern Matching}
-
-Curry allows multiple occurrences of pattern variables
-in standard patterns. These are an abbreviation of equational constraints
-between pattern variables.
-Functional patterns might also contain multiple occurrences of
-pattern variables.
-For instance, the operation
-\begin{curry}
-f (_++[x]++_++[x]++_) = x
-\end{curry}
-returns all elements with at least two occurrences in a list.
-
-If functional patterns as well as multiple occurrences of
-pattern variables occur in a pattern defining an operation,
-there are various orders to match an expression against such
-an operation. In the current implementation, the order
-is as follows:
-\begin{enumerate}
-\item Standard pattern matching: First, it is checked whether
-the constructor patterns match. Thus, functional patterns
-and multiple occurrences of pattern variables are ignored.
-\item Functional pattern matching: In the next phase,
-functional patterns are matched but occurrences of standard
-pattern variables in the functional patterns are ignored.
-\item Non-linear patterns: If standard and functional pattern matching
-is successful, the equational constraints which correspond
-to multiple occurrences pattern variables are solved.
-\item Guards: Finally, the guards supplied by the programmer
-are checked.
-\end{enumerate}
-The order of pattern matching should not influence the computed
-result. However, it might have some influence on the termination
-behavior of programs, i.e., a program might not terminate
-instead of finitely failing.
-In such cases, it could be necessary to consider the influence
-of the order of pattern matching. Note that other orders of pattern matching
-can be obtained using auxiliary operations.
-
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are 
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there. 
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\begin{curry}
-type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
-\end{curry}
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$ 
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\begin{curry}
-type Person = {name :: String, age :: Int}
-type Address = {person :: Person, street :: String, city :: String}
-type Branch a b = {left :: a, right :: b}
-\end{curry}
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\begin{curry}
-type Address = {person :: Person, street,city :: String}
-\end{curry}
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\begin{curry}
-type Address = {street,city :: String, person :: Person}
-\end{curry}
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\begin{curry}
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\end{curry}
-\begin{curry}
-getName :: Person -> String
-getName $\ldots$
-\end{curry}
-
-Labels can only be used in the context of
-records. They do not share the name space with 
-functions/constructors/variables or type constructors/type variables. 
-For instance it is possible to use 
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually 
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\begin{curry}
-{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
-\end{curry}
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with 
-exactly one assignment. Examples for record constructions are
-\begin{curry}
-{name := "Johnson", age := 30}     -- generates a record of type 'Person'
-{left := True, right := 20}        -- generates a record of type 'Branch'
-\end{curry}
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\begin{curry}
-{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
-\end{curry}
-Unlike labeled fields in record type declarations, 
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\begin{curry}
-{person := {name := "Smith", age := 20},   -- generates a record of
- street := "Main Street",                  -- type 'Address'
- city   := "Springfield"}
-\end{curry}
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records. 
-It has the following form:
-\begin{curry}
-$r$ :> $l$
-\end{curry}
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\begin{curry}
-pers :> name
-\end{curry}
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\begin{curry}
-addr :> person :> age
-\end{curry}
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}).
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\begin{curry}
-{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
-\end{curry}
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration 
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record. 
-Assignments can occur in an arbitrary order. It is not allowed to 
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\begin{curry}
-{name := "Scott", age := 25 | pers}
-{person := {name := "Scott", age := 25 | pers} | addr}
-\end{curry}
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\begin{curry}
-{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
-{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
-\end{curry}
-In both cases each label $l_i$ is specified with a pattern $p_i$. 
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern. 
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the 
-patterns of the specified labels are matched against 
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\begin{curry}
-isSmith30 :: Person -> Bool
-isSmith30 {name = "Smith", age = 30} = True
-\end{curry}
-\begin{curry}
-startsWith :: Char -> Person -> Bool
-startsWith c {name = (d:_) | _} = c == d
-\end{curry}
-\begin{curry}
-getPerson :: Address -> Person
-getPerson {person = p | _} = p
-\end{curry}
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways 
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent 
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
-
-\newpage
-
-\input{currydoc}
-
-\newpage
-
-\section{\cb: A Tool for Analyzing and Browsing Curry Programs}
-\label{sec-currybrowser}
-
-\cb is a tool to browse through the modules and functions
-of a Curry application, show them in various formats,
-and analyze their properties.\footnote{Although \cb is
-implemented in Curry, some functionalities of it require an
-installed graph visualization tool (dot \url{http://www.graphviz.org/}),
-otherwise they have no effect.}
-Moreover, it is constructed in a way so that
-new analyzers can be easily connected to \cb.
-A detailed description of the ideas behind this tool can be
-found in \cite{Hanus05WCFLP,Hanus06WLPE}.
-
-\cb is part of the \CYS distribution and can be
-started in two ways:
-\begin{itemize}
-\item
-In the command shell via the command: \code{\cyshome/bin/currybrowser mod}
-\item
-In the \CYS environment after loading the module
-\code{mod} and typing the command \ccode{:browse}.
-\end{itemize}
-Here, \ccode{mod} is the name of the main module of a Curry application.
-After the start, \cb loads the interfaces of the main
-module and all imported modules before a GUI is created
-for interactive browsing.
-
-\begin{figure}[t]
-\begin{center}
-\includegraphics[scale=0.7]{currybrowser.jpg}
-\end{center}
-\caption{Snapshot of the main window of CurryBrowser\label{fig-currybrowser}}
-\end{figure}
-%
-To get an impression of the use of \cb, Figure~\ref{fig-currybrowser}
-shows a snapshot of its use on a particular application
-(here: the implementation of \cb).
-The upper list box in the left column shows the modules and their imports
-in order to browse through the modules of an application.
-Similarly to directory browsers, the list of imported modules of a module
-can be opened or closed by clicking.
-After selecting a module in the list of modules, its source code,
-interface, or various other formats of the module can be shown
-in the main (right) text area. For instance, one can show
-pretty-printed versions of the intermediate flat programs (see below)
-in order to see how local function definitions are translated by lambda lifting
-\cite{Johnsson85}
-or pattern matching is translated into case expressions \cite{Hanus97POPL,Wadler87}.
-Since Curry is a language with parametric polymorphism and type inference,
-programmers often omit the type signatures when defining functions.
-Therefore, one can also view (and store) the selected module as source code where
-missing type signatures are added.
-
-Below the list box for selecting modules, there is a menu
-(``Analyze selected module'') to analyze all functions
-of the currently selected module at once. This is useful
-to spot some functions of a module that could be problematic
-in some application contexts, like functions that are impure (i.e., the result
-depends on the evaluation time) or partially defined (i.e.,
-not evaluable on all ground terms).
-If such an analysis is selected,
-the names of all functions are shown in the
-lower list box of the left column (the ``function list'')
-with prefixes indicating the properties of the individual functions.
-
-The function list box can be also filled with functions
-via the menu ``Select functions''. For instance, all functions
-or only the exported functions defined in the currently selected
-module can be shown there, or all functions from different modules
-that are directly or indirectly called from
-a currently selected function.
-This list box is central to focus on a function in the
-source code of some module or to analyze some function,
-i.e., showing their properties. In order to focus on a function,
-it is sufficient to check the ``focus on code'' button.
-To analyze an individually selected function, one can
-select an analysis from the list of available program analyses
-(through the menu ``Select analysis'').
-In this case, the analysis results are either shown
-in the text box below the main text area
-or visualized by separate tools, e.g., by a graph drawing tool for
-visualizing call graphs.
-Some analyses are local, i.e., they need only to consider the local definition
-of this function (e.g., ``Calls directly,'' ``Overlapping rules,''
-``Pattern completeness''),
-where other analyses are global, i.e.,
-they consider the definitions of all functions directly or indirectly called
-by this function (e.g., ``Depends on,'' ``Solution complete,''
-``Set-valued'').
-%
-Finally, there are a few additional tools integrated into \cb,
-for instance, to visualize the import relation between all modules
-as a dependency graph. These tools are available through the ``Tools'' menu.
-
-More details about the use of \cb and all built-in analyses
-are available through the ``Help'' menu of \cb.
-
-\newpage
-
-\section{CurryTest: A Tool for Testing Curry Programs}
-\label{sec-currytest}
-
-CurryTest\index{CurryTest}\index{testing programs}\index{program!testing}
-is a simple tool in the \CYS distribution to write
-and run repeatable tests. CurryTest simplifies the task
-of writing test cases for a module and executing them.
-The tool is easy to use. Assume one has implemented a module \code{MyMod}
-and wants to write some test cases to test its functionality,
-making regression tests in future versions, etc.
-For this purpose, there is a system library \code{Assertion}
-(Section~\ref{Library:Assertion}) which
-contains the necessary definitions for writing tests.
-In particular, it exports an abstract polymorphic type \ccode{Assertion a}
-together with the following operations:
-\begin{curry}
-assertTrue      :: String -> Bool -> Assertion ()
-assertEqual     :: String -> a -> a -> Assertion a
-assertValues    :: String -> a -> [a] -> Assertion a
-assertSolutions :: String -> (a->Success) -> [a] -> Assertion a
-assertIO        :: String -> IO a -> a -> Assertion a
-assertEqualIO   :: String -> IO a -> IO a -> Assertion a
-\end{curry}
-The expression \ccode{assertTrue $s$ $b$}
-is an assertion (named $s$) that the expression $b$ has the value \code{True}.
-Similarly, the expression \ccode{assertEqual $s$ $e_1$ $e_2$}
-asserts that the expressions $e_1$ and $e_2$
-must be equal (i.e., \code{$e_1$==$e_2$} must hold),
-the expression \ccode{assertValues $s$ $e$ $vs$} asserts
-that $vs$ is the multiset of all values of $e$,
-and the expression \ccode{assertSolutions $s$ $c$ $vs$} asserts
-that the constraint abstraction $c$ has the multiset of solutions $vs$.
-Furthermore, the expression \ccode{assertIO $s$ $a$ $v$}
-asserts that the I/O action $a$ yields the value $v$ whenever it is
-executed, and
-the expression \ccode{assertEqualIO $s$ $a_1$ $a_2$}
-asserts that the I/O actions $a_1$ and $a_2$ yield equal values.
-The name $s$ provided as a first argument in each assertion
-is used in the protocol produced by the test tool.
-
-One can define a test program by importing the module
-to be tested together with the module \code{Assertion} and defining
-top-level functions of type \code{Assertion} in this module
-(which must also be exported).
-As an example, consider the following program
-that can be used to test some list processing functions:
-\begin{curry}
-import List
-import Assertion
-
-test1 = assertEqual     "++"     ([1,2]++[3,4]) [1,2,3,4]
-
-test2 = assertTrue      "all"    (all (<5) [1,2,3,4])
-
-test3 = assertSolutions "prefix" (\x -> x++_ =:= [1,2])
-                                 [[],[1],[1,2]]
-\end{curry}
-For instance, \code{test1} asserts that the result of evaluating the
-expression \code{([1,2]++[3,4])} is equal to \code{[1,2,3,4]}.
-
-We can execute a test suite by the command\pindex{currytest}
-\begin{curry}
-currytest TestList
-\end{curry}
-(\code{currytest} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-In our example, \ccode{TestList.curry} is the program containing the
-definition of all assertions. This has the effect
-that all exported top-level functions
-of type \code{Assertion} are tested (i.e., the corresponding
-assertions are checked) and the results
-(\ccode{OK} or failure) are reported together with the name of each assertion.
-%If failures occur, the complete test results are also
-%written into a file named \ccode{TestList.testlog}.''
-For our example above, we obtain the following successful protocol:
-\begin{curry}
-============================================================
-Testing module "TestList"...
-OK: ++
-OK: all
-OK: prefix
-All tests successfully passed.
-============================================================
-\end{curry}
-There is also a graphical interface that summarizes the results
-more nicely.
-In order to start this interface, one has to add the parameter
-\ccode{--window} (or \ccode{-w}), e.g., executing a test suite by
-\begin{curry}
-currytest --window TestList
-\end{curry}
-or
-\begin{curry}
-currytest -w TestList
-\end{curry}
-A snapshot of the interface is shown in Figure~\ref{fig-currytest}.
-
-\begin{figure}%[t]
-\begin{center}
-\includegraphics[scale=0.7]{currytest.jpg}
-\end{center}
-\caption{Snapshot of CurryTest's graphical interface\label{fig-currytest}}
-\end{figure}
-
-\newpage
-
-\section{ERD2Curry: A Tool to Generate Programs from ER Specifications}
-\label{sec-erd2curry}
-
-ERD2Curry\index{ERD2Curry}\index{database programming}
-is a tool to generate Curry code to access and manipulate data
-persistently stored from
-entity relationship diagrams.\index{entity relationship diagrams}
-The idea of this tool is described in detail in
-\cite{BrasselHanusMueller08PADL}.
-Thus, we describe only the basic steps to use this tool
-in the following.
-
-If one creates an entity relationship diagram (ERD)
-with the Umbrello UML Modeller, one has to store its
-XML description in XMI format (as offered by Umbrello)
-in a file, e.g., \ccode{myerd.xmi}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry myerd.xmi
-\end{curry}
-(\code{erd2curry} is a program stored in \code{\cyshome/bin}
-where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
-If \code{MyData} is the name of the ERD, the Curry program file
-\ccode{MyData.curry} is generated containing all the necessary
-database access code as described in \cite{BrasselHanusMueller08PADL}.
-
-If one does not want to use the Umbrello UML Modeller,
-one can also create a textual description of the ERD
-as a Curry term of type \code{ERD}
-(w.r.t.\ the type definition given in module
-\code{\cyshome/tools/erd2curry/ERD.curry})
-and store it in some file, e.g., \ccode{myerd.term}.
-This description can be compiled into a Curry program by the
-command\pindex{erd2curry}
-\begin{curry}
-erd2curry -t myerd.term
-\end{curry}
-%
-There is also the possibility to visualize an ERD term
-as a graph with the graph visualization program \code{dotty}
-(for this purpose, it might be necessary to adapt the definition
-of \code{dotviewcommand} in your \ccode{.kics2rc} file,
-see Section~\ref{sec-customization},
-according to your local environment).
-This can be done by the command
-\begin{curry}
-erd2curry -v myerd.term
-\end{curry}
-
-\paragraph{Inclusion in the Curry application:}
-To compile the generated database code, either
-include the directory \code{\cyshome/tools/erd2curry}
-into your Curry load path
-(e.g., by setting  the environment variable
-\ccode{CURRYPATH}\pindex{CURRYPATH}, see also Section~\ref{sec-modules})
-or copy the file
-\code{\cyshome/tools/erd2curry/ERDGeneric.curry}
-into the directory of the generated database code.
-
-\newpage
-
-\section{Technical Problems}
-
-One can implement distributed systems with \CYS
-by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
-that supports a socket communication with symbolic names
-rather than natural numbers.
-For instance, this library is the basis of programming
-dynamic web pages with the libraries
-\code{HTML} (Section~\ref{Library:HTML})
-or \code{WUI} (Section~\ref{Library:WUI}).
-However, it might be possible that some technical problems
-arise due to the use of named sockets.
-Therefore, this section gives some information
-about the technical requirements of \CYS and how to solve
-problems due to these requirements.
-
-There is one fixed port that is used by the implementation of \CYS:
-\begin{description}
-\item[Port 8767:] This port is used by the
-{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
-named sockets in Curry.
-If some other process uses this port on the machine,
-the distribution facilities defined in the module \code{NamedSocket}
-cannot be used.
-\end{description}
-If these features do not work, you can try to find out
-whether this port is in use by the shell command
-\ccode{netstat -a | fgrep 8767} (or similar).
-
-The CPNS is implemented as a demon listening on its port 8767
-in order to serve requests about registering a new symbolic
-name for a named socket or asking the physical port number
-of an registered named socket. The demon will be automatically started for
-the first time on a machine when a user runs a program
-using named sockets.
-It can also be manually started and terminated by the
-scripts \code{\cyshome/cpns/start} and
-\code{\cyshome/cpns/stop}.
-If the demon is already running, the command \code{\cyshome/cpns/start}
-does nothing (so it can be always executed
-before invoking a Curry program using named sockets).
-
-If you detect any further technical problem,
-please write to
-\begin{center}
-\code{mh@informatik.uni-kiel.de}
-\end{center}
-
-\newpage
-
+% Bibliography
 \addcontentsline{toc}{section}{Bibliography}
@@
 
-\newpage
 \appendix
 
-\section{Libraries of the \CYS Distribution}
-\label{sec:libraries}
+\include{libraries}
+\include{markdown_syntax}
+\include{auxiliary_files}
+\include{external_ops}
 
-{\setlength{\parindent}{0.0cm}
-
-The \CYS distribution comes with an extensive collection
-of libraries for application programming.
-The libraries for meta-programming by representing
-Curry programs as datatypes in Curry are described
-in the following subsection in more detail.
-The complete set of libraries with all exported types and functions
-are described in the further subsections.
-For a more detailed online documentation of all libraries of \CYS,
-see \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/index.html}.
-
-\subsection{AbstractCurry and FlatCurry: Meta-Programming in Curry}
-\label{sec-flatcurry}
-
-\index{AbstractCurry}
-\index{FlatCurry}
-To support meta-programming, i.e., the manipulation of Curry programs
-in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{\cyshome/lib/meta},
-which define datatypes for the representation
-of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
-where local function definitions are replaced by global definitions
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
-program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
-programs that are more oriented towards the source program.
-
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
-These actions parse the corresponding source program and return
-a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
-
-Since all datatypes are explained in detail in these modules,
-we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
-of these modules.
-
-As an example, consider a program file \ccode{test.curry}
-containing the following two lines:
-\begin{curry}
-rev []     = []
-rev (x:xs) = (rev xs) ++ [x]
-\end{curry}
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
-following term:
-\begin{curry}
- (Prog "test"
-  ["Prelude"]
-  []
-  [Func ("test","rev") 1 Public
-        (FuncType (TCons ("Prelude","[]") [(TVar 0)])
-                  (TCons ("Prelude","[]") [(TVar 0)]))
-        (Rule [0]
-           (Case Flex (Var 1)
-              [Branch (Pattern ("Prelude","[]") [])
-                  (Comb ConsCall ("Prelude","[]") []),
-               Branch (Pattern ("Prelude",":") [2,3])
-                  (Comb FuncCall ("Prelude","++")
-                        [Comb FuncCall ("test","rev") [Var 3],
-                         Comb ConsCall ("Prelude",":")
-                              [Var 2,Comb ConsCall ("Prelude","[]") []]
-                        ])
-              ]))]
-  []
- )
-\end{curry}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Definitions in order to LaTeX documents generated by "currydoc --tex"
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\currymodule}[1]{\subsubsection{Library #1}\label{Library:#1}}
-\newcommand{\currytypesstart}{\subsubsection*{Exported types:}}
-\newcommand{\currytypesstop}{}
-\newcommand{\currytypesynstart}[2]{{\tt type #2}\pindex{#1} \begin{quote}}
-\newcommand{\currytypesynstop}{\end{quote}}
-\newcommand{\currydatastart}[1]{{\tt data #1}\pindex{#1} \begin{quote}}
-\newcommand{\currydatacons}{\end{quote}%
-\begin{itemize}\item[] \hspace{-4ex}\emph{Exported constructors:}}
-\newcommand{\currydatastop}{\end{itemize}}
-\newcommand{\curryconsstart}[2]{\item {\tt #1~::~#2}\par}
-\newcommand{\curryfuncstart}{\subsubsection*{Exported functions:}}
-\newcommand{\curryfuncstop}{}
-\newcommand{\curryfunctionstart}[2]{#2\pindex{#1}\begin{quote}}
-\newcommand{\curryfunctionstop}{\end{quote}}
-\newcommand{\curryfuncsig}[2]{{\tt #1~::~#2}}
-
-\subsection{General Libraries}
-
-\input{lib/AllSolutions}
-\input{lib/Assertion}
-\input{lib/Char}
-\input{lib/Combinatorial}
-\input{lib/Constraint}
-\input{lib/CSV}
-\input{lib/Directory}
-\input{lib/FileGoodies}
-\input{lib/Float}
-\input{lib/Global}
-\input{lib/GUI}
-\input{lib/Integer}
-\input{lib/IO}
-\input{lib/IOExts}
-\input{lib/JavaScript}
-\input{lib/KeyDatabaseSQLite}
-\input{lib/List}
-\input{lib/Maybe}
-\input{lib/NamedSocket}
-\input{lib/Parser}
-\input{lib/Pretty}
-\input{lib/Profile}
-\input{lib/PropertyFile}
-\input{lib/Read}
-\input{lib/ReadNumeric}
-\input{lib/ReadShowTerm}
-\input{lib/SetFunctions}
-\input{lib/SearchTree}
-\input{lib/Socket}
-\input{lib/System}
-\input{lib/Time}
-\input{lib/Unsafe}
-
-\subsection{Data Structures and Algorithms}
-
-\input{lib/Array}
-\input{lib/Dequeue}
-\input{lib/FiniteMap}
-\input{lib/GraphInductive}
-\input{lib/Random}
-\input{lib/RedBlackTree}
-\input{lib/SetRBT}
-\input{lib/Sort}
-\input{lib/TableRBT}
-\input{lib/Traversal}
-
-\subsection{Libraries for Web Applications}
-
-\input{lib/CategorizedHtmlList}
-\input{lib/HTML}
-\input{lib/HtmlParser}
-\input{lib/Mail}
-\input{lib/Markdown}
-\input{lib/WUI}
-\input{lib/URL}
-\input{lib/XML}
-\input{lib/XmlConv}
-
-\subsection{Libraries for Meta-Programming}
-
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
-\input{lib/CompactFlatCurry}
-\input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
-
-} % end setlength parindent
-
-\newpage
-
-\input{markdown_syntax}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Auxiliary Files}
-\label{sec-auxfiles}
-
-During the translation and execution of a Curry program with \CYS,
-various intermediate representations of the source program are created
-and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
-these auxiliary files because they are automatically generated
-and updated. You should only remember the command for deleting
-all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
-to clean up your directories.
-
-The various components of \CYS create the following auxiliary files.
-\begin{description}
-\item[\code{prog.fcy}:] This file contains the Curry program
-in the so-called ``FlatCurry'' representation where all functions are global
-(i.e., lambda lifting has been performed) and pattern matching
-is translated into explicit case/or expressions
-(compare Appendix~\ref{sec-flatcurry}).
-This representation might be useful for other back ends and
-compilers for Curry and is the basis doing meta-programming in Curry.
-This file is implicitly
-generated when a program is compiled with \CYS.
-The FlatCurry representation of a Curry program is usually
-generated by the front-end after parsing, type checking and eliminating
-local declarations.
-If $dir$ is the directory where the Curry program is stored,
-the corresponding FlatCurry program is stored in the directory
-\ccode{$dir$/.curry}.
-
-\item[\code{prog.fint}:] This file contains the interface
-of the program in the so-called ``FlatCurry'' representation,
-i.e., it is similar to \code{prog.fcy} but contains only exported
-entities and the bodies of all functions omitted (i.e., ``external'').
-This representation is useful for providing a fast access
-to module interfaces.
-This file is implicitly generated when a program is compiled with \CYS
-and stored in the same directory as \code{prog.fcy}.
-
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.hs}:] This file contains a Haskell program
-as the result of translating the Curry program with the
-\CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
-\item[\code{Curry_prog.o}:] This file contains the object code
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
-This file is stored in the same directory as \code{Curry_prog.hs}.
-
-\item[\code{prog}:] This file contains the executable
-after compiling and saving a program with \CYS
-(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).
-
-\end{description}
-
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{External Operations}
-\label{sec:external-operations}
-
-\index{operation!external}\index{external operation}
-Currently, \CYS has no general interface to external operations,
-i.e., operations whose semantics is not defined by program rules
-in a Curry program but by some code written in another programming
-language.
-Thus, if an external operation should be added
-to the system, this operation must be declared as \code{external}
-in the Curry source code
-and an implementation for this external operation
-must be provided for the run-time system.
-An external operation is defined as follows in the Curry source code:
-\begin{enumerate}
-\item
-Add a type declaration for the external operation somewhere
-in a module defining this operation (usually, the prelude
-or some system module).
-\item
-For external operations it is not allowed to define any
-rule since their semantics is determined by an external implementation.
-Instead of the defining rules, you have to write
-\begin{curry}
-f external
-\end{curry}
-below the type declaration for the external operation \code{f}.
-\end{enumerate}
-Furthermore, an implementation of the external operation
-must be provided in the target language of the \CYS compiler,
-i.e., in Haskell, and inserted in the compiled code.
-In order to simplify this task, \CYS follows some code
-conventions that are described in the following.
-
-Assume you want to implement your own concatenation
-for strings in a module \code{String}.
-The name and type of this string concatenation should be
-\begin{curry}
-sconc :: String -> String -> String
-\end{curry}
-Since the primitive Haskell implementation of this operation
-does not now anything about the operational mechanism of Curry
-(e.g., needed narrowing, non-deterministic rewriting),
-the arguments need to be completely evaluated before
-the primitive implementation is called.
-This can be easily obtained by the prelude operation \code{(\$\#\#)}
-that applies an operation to the \emph{normal form} of the given
-argument, i.e., this operation evaluates the argument
-to its normal form before applying the operation to it.\footnote{%
-There is also a similar prelude operation \code{(\$\#)}
-which evaluates the argument only to head-normal form.
-This is a bit more efficient and can be used for unstructured
-types like \code{Bool}.}
-Thus, we define \code{sconc} by
-\begin{currynomath}
-sconc :: String -> String -> String
-sconc s1 s2 = (prim_sconc $## s1) $## s2
-
-prim_sconc :: String -> String -> String
-prim_sconc external
-\end{currynomath}
-so that it is ensured that the external operation \code{prim_sconc}
-is always called with complete evaluated arguments.
-
-In order to define the Haskell code implementing \code{prim_sconc},
-one has to satisfy the naming conventions of \CYS.
-The \CYS compiler generates the following code for the
-external operation \code{prim_sconc} (note that the generated
-Haskell code for the module \code{String} is stored in the file
-\code{.curry/kics2/Curry_String.hs}):
-\begin{curry}
-d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-               -> ConstStore
-               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
-d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
-\end{curry}
-The type constructors \code{OP_List} and \code{C_Char}
-of the prelude \code{Curry_Prelude}\footnote{Note that all
-translated Curry modules are imported in the Haskell code fully qualified
-in order to avoid name conflicts.}
-correspond to the Curry type constructors for lists and characters.
-The Haskell operation \code{external_d_C_prim_sconc}
-is the external operation to be implemented in Haskell by the programmer.
-The additional argument of type \code{ConstStore}
-represents the current set of constraints when this
-operation is called. This argument is intended to provide
-a more efficient access to binding constraints and can be
-ignored in standard operations.
-
-If \code{String.curry} contains the code
-of the Curry function \code{sconc} described above,
-the Haskell code implementing the external operations
-occurring in the module \code{String} must be in the
-file \code{External_String.hs} which is located in the same
-directory as the file \code{String.curry}.
-The \CYS compiler appends the code contained in
-\code{External_String.hs} to the generated code
-stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
-If the file \code{External_String.hs} contains also
-some import declarations at the beginning, these import declarations
-are put after the generated import declarations.}
-
-In order to complete our example, we have to write into the
-file \code{External_String.hs} a definition of the Haskell function
-\code{external_d_C_prim_sconc}.
-Thus, we start with the following definitions:
-\begin{curry}
-import qualified Curry_Prelude as CP
-
-external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
-                        -> ConstStore -> CP.OP_List CP.C_Char
-\end{curry}
-First, we import the standard prelude with the name \code{CP}
-in order to shorten the writing of type declarations.
-In order to write the final code of this operation,
-we have to convert the Curry-related types
-(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
-Note that the Curry-related types contain information about
-non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
-that are meaningless in Haskell.
-To solve this conversion problem, the implementation of \CYS
-provides a family of operations to perform these conversions
-for the predefined types occurring in the standard prelude.
-For instance, \code{fromCurry} converts a Curry type into the
-corresponding Haskell type, and \code{toCurry} converts
-the Haskell type into the corresponding Curry type.
-Thus, we complete our example with the definition
-(note that we simply ignore the final argument representing the
-constraint store)
-\begin{curry}
-external_d_C_prim_sconc s1 s2 _ =
-  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
-\end{curry}
-Here, we use Haskell's concatenation operation \ccode{++}
-to concatenate the string arguments.
-The type annotation \ccode{:: String} is necessary
-because \ccode{++} is a polymorphic function
-so that the type inference system of Haskell
-has problems to determine the right instance of the conversion
-function.
-
-The conversion between Curry types and Haskell types,
-i.e., the family of conversion operation \code{fromCurry}
-and \code{toCurry}, is defined in the \CYS implementation
-for all standard data types.
-In particular, it is also defined on function types so that one can
-easily implement external Curry I/O actions by using
-Haskell I/O actions.
-For instance, if we want to implement
-an external operation to print some string as an output line,
-we start by declaring the external operations in the Curry module \code{String}:
-\begin{currynomath}
-printString :: String -> IO ()
-printString s = prim_printString $## s
-
-prim_printString :: String -> IO ()
-prim_printString external
-\end{currynomath} % $
-Next we add the corresponding implementation in the file
-\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
-of the Haskell representation of the Curry type constructor \code{IO}
-and the Curry data type \ccode{()}, respectively):
-\begin{curry}
-external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
-                              -> CP.C_IO CP.OP_Unit
-external_d_C_prim_printString s _ = toCurry putStrLn s
-\end{curry}
-Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
-is transformed into a Curry I/O action \ccode{toCurry putStrLn}
-which has the type
-\begin{curry}
-CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
-\end{curry}
-When we compile the Curry module \code{String},
-\CYS combines these definitions in the target program so that
-we can immediately use the externally defined operation
-\code{printString} in Curry programs.
-
-As we have seen, \CYS transforms a name like
-\code{primOP} of an external operation into the name
-\code{external_d_C_primOP} for the Haskell operation
-to be implemented, i.e., only a specific prefix is added.
-However, this is only valid if no special characters occur
-in the Curry names.
-Otherwise (in order to generate a correct Haskell program),
-special characters are translated into specific
-names prefixed by \ccode{OP_}. For instance,
-if we declare the external operation
-\begin{curry}
-(<&>) :: Int -> Int -> Int
-(<&>) external
-\end{curry}
-the generated Haskell module contains the code
-\begin{curry}
-d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
-                     -> ConstStore -> Curry_Prelude.C_Int
-d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
-\end{curry}
-so that one has to implement the operation
-\code{external_d_OP_lt_ampersand_gt} in Haskell.
-If in doubt, one should look into the generated Haskell code
-about the names and types of the operations to be implemented.
-
-Finally, note that this method to connect functions implemented in Haskell
-to Curry programs provides the opportunity to connect
-also operations written in other programming languages
-to Curry via Haskell's foreign function interface.
-
-\newpage
+% Index
 \addcontentsline{toc}{section}{Index}

### 2013-07-12 docs/src/usage.tex
Split manual into separate tex files

\section{Using the Interactive Environment of \CYS}
\label{sec-kics2}

This section describes the interactive environment
\CYS\index{\CYS}
that supports the development of applications written in Curry.
The implementation of \CYS contains also a separate compiler
which is automatically invoked by the interactive environment.

\subsection{Invoking \CYS}
\label{sec:invoke-kics2}

To start \CYS, execute the command
\ccode{kics2}\pindex{kics2}
(this is a shell script stored in
\code{\cyshome/bin} where \cyshome is the installation directory
of \CYS).
When the system is ready (i.e., when the prompt \ccode{Prelude>} occurs),
the prelude (\code{\cyshome/lib/Prelude.curry})
is already loaded, i.e., all definitions in the prelude are accessible.
Now you can type various commands (see next section)
or an expression to be evaluated.

One can also invoke \CYS with parameters.
These parameters are usual a sequence of commands
(see next section) that are executed before the user interaction
starts. For instance, the invocation
\begin{curry}
kics2 :load Mod :add List
\end{curry}
starts \CYS, loads the main module \code{Mod}, and adds the additional
module \code{List}.
The invocation
\begin{curry}
kics2 :load Mod :eval config
\end{curry}
starts \CYS, loads the main module \code{Mod}, and evaluates
the operation \code{config} before the user interaction starts.
As a final example, the invocation
\begin{curry}
kics2 :load Mod :save :quit
\end{curry}
starts \CYS, loads the main module \code{Mod}, creates an executable,
and terminates \CYS. This invocation could be useful in ``make'' files
for systems implemented in Curry.

\subsection{Command of \CYS}
\label{sec:kics2-commands}

The {\bf most important commands} of \CYS are
(it is sufficient to type a unique prefix of a command if it is unique,
e.g., one can type \ccode{:r} instead of \ccode{:reload}):

\begin{description}
\item[\fbox{\code{:help}}]\pindex{:help}
Show a list of all available commands.

\item[\fbox{\code{:load $prog$}}]\pindex{:load}
Compile and load the program stored in \code{$prog$.curry}
together with all its imported modules.
% If this file does not exist, the system looks for a FlatCurry
% file \code{$prog$.fcy} and compiles from this intermediate representation.
% If the file \code{$prog$.fcy} does not exists, too, the system looks
% for a file \code{$prog$_flat.xml} containing a FlatCurry program in
% XML representation (compare command \ccode{:xml}\pindex{:xml}),
% translates this into a FlatCurry file \code{$prog$.fcy}
% and compiles from this intermediate representation.

\item[\fbox{\code{:reload}}]\pindex{:reload}
Recompile all currently loaded modules.

\item[\fbox{\code{:add} $m$}]\pindex{:add}
Add module $m$ to the set of currently loaded modules
so that its exported entities are available in the top-level environment.

\item[\fbox{$expr$}] Evaluate the expression $expr$ to normal form
and show the computed results.
In the default mode, all results of
non-deterministic computations are printed.
One can also print first one result and the next result
only if the user requests it. This behavior can be set by
the option \code{interactive}\index{interactive} (see below).

\textbf{Free variables in initial expressions} must be declared as
in Curry programs.
In order to see the results of their bindings,\footnote{Currently,
bindings are only printed if the initial expression is not an I/O action
(i.e., not of type \ccode{IO\ldots})
and there are not more than ten free variables
in the initial expression.}
they must be introduced by a \ccode{where\ldots{}free} declaration.
For instance, one can write
\begin{curry}
xs++ys =:= [1,2]  where xs,ys free
\end{curry}
in order to obtain the following three possible bindings:
\begin{curry}
{xs = [], ys = [1,2]} Success
{xs = [1], ys = [2]} Success
{xs = [1,2], ys = []} Success
\end{curry}
Without these declarations, an error is reported in order to
avoid the unintended introduction of free variables in initial expressions
by typos.

If the free variables in the initial goal are of a polymorphic type,
as in the expression
\begin{curry}
xs++ys=:=[z] where xs,ys,z free
\end{curry}
they are specialized to the type \ccode{()}
(since the current implementation of \CYS does not support
computations with polymorphic logic variables).

\item[\fbox{:eval $expr$}]\pindex{:eval}
Same as $expr$. This command might be useful when putting
commands as arguments when invoking \code{kics2}.

% \item[\fbox{\code{let} $x$ \code{=} $expr$}]
% Define the identifier $x$ as an abbreviation for the expression $expr$
% which can be used in subsequent expressions. The identifier $x$
% is visible until the next \code{load} or \code{reload} command.

\item[\fbox{\code{:quit}}]\pindex{:quit} Exit the system.
\end{description}
%
\bigskip
%
There are also a number of {\bf further commands} that are often
useful:
%
\begin{description}
\item[\fbox{\code{:type $expr$}}]\pindex{:type}
Show the type of the expression $expr$.

\item[\fbox{\code{:programs}}]\pindex{:programs}
Show the list of all Curry programs that are available in the load path.

\item[\fbox{\code{:cd $dir$}}]\pindex{:cd}
Change the current working directory to $dir$.

\item[\fbox{\code{:edit}}]\pindex{:edit}
Load the source code of the current main module into a text editor.
If the variable \code{editcommand} is set in the
configuration file \ccode{.kics2rc}
(see Section~\ref{sec-customization}),
its value is used as an editor command, otherwise
the environment variable \ccode{EDITOR} is used as the editor program.

\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
Load file $file$ into a text editor which is defined
as in the command \ccode{:edit}.

\item[\fbox{\code{:show}}]\pindex{:show}
Show the source text of the currently loaded Curry program.
If the variable \code{showcommand} is set in the
configuration file \ccode{.kics2rc}
(see Section~\ref{sec-customization}),
its value is used as a command to show the source text,
otherwise the \ccode{cat} is used.

\item[\fbox{\code{:show $m$}}]\pindex{:show}
Show the source text of module $m$ which must be accessible
via the current load path.

\item[\fbox{\code{:source $f$}}]\pindex{:source}
Show the source code of function $f$ (which must be visible
in the currently loaded module) in a separate window.

\item[\fbox{\code{:source $m$.$f$}}]\pindex{:source}
Show the source code of function $f$ defined in module $m$
in a separate window.

\item[\fbox{\code{:browse}}]\pindex{:browse}
Start the CurryBrowser to analyze the currently loaded
module together with all its imported modules
(see Section~\ref{sec-currybrowser} for more details).

\item[\fbox{\code{:interface}}]\pindex{:interface}
Show the interface of the currently loaded
module, i.e., show the names of all imported modules,
the fixity declarations of all exported operators,
the exported datatypes declarations and the types
of all exported functions.

\item[\fbox{\code{:interface $m$}}]\pindex{:interface}
Similar to \ccode{:interface}
but shows the interface of the module $m$
which must be in the load path of \CYS.

\item[\fbox{\code{:usedimports}}]\pindex{:usedimports}
Show all calls to imported functions in the currently loaded module.
This might be useful to see which import declarations are really necessary.

\item[\fbox{\code{:set $option$}}]\pindex{:set}
Set or turn on/off a specific option
of the \CYS environment (see \ref{sec:options} for a description
of all options). Options are turned on by the prefix
\ccode{+} and off by the prefix \ccode{-}. Options that can only
be set (e.g., \code{path}) must not contain a prefix.

\item[\fbox{\code{:set}}]\pindex{:set}
Show a help text on the possible options
together with the current values of all options.

\item[\fbox{\code{:save}}]\pindex{:save} Save the currently loaded
program as an executable evaluating the main expression \ccode{main}.
The executable is stored in the file \code{Mod}
if \code{Mod} is the name of the currently loaded main module.

\item[\fbox{\code{:save $expr$}}]\pindex{:save} Similar as \ccode{:save}
but the expression $expr$ (typically: a call to the main
function) will be evaluated by the executable.

\item[\fbox{\code{:fork $expr$}}]\pindex{:fork}
The expression $expr$, which is typically of type \ccode{IO ()},
is evaluated in an independent process which runs in
parallel to the current \CYS process.
All output and error messages from this new process are suppressed.
This command is useful to test distributed Curry programs
where one can start
a new server process by this command. The new process
will be terminated when the evaluation of the expression $expr$
is finished.

\item[\fbox{\code{:!$cmd$}}]\pindex{:"!} Shell escape: execute $cmd$ in a Unix shell.

\end{description}

\subsection{Options of \CYS}
\label{sec:options}

The following options (which can be set by the command \ccode{:set})
are currently supported:

\begin{description}
\item[\fbox{\code{path $path$}}]\pindex{path} Set the additional search path
for loading modules to $path$.
Note that this search path is only used for loading modules
inside this invocation of \CYS.

\item[\fbox{\code{prdfs }}]\pindex{prdfs}
Set the search mode to evaluate non-deterministic expressions
to primitive depth-first search (which is usually the fastest method
to print all non-deterministic values).

\item[\fbox{\code{dfs}}]\pindex{dfs}
Set the search mode to evaluate non-deterministic expressions
to depth-first search. Usually, all non-deterministic values
are printed with a depth-first strategy, but one can also
print only the first value or all values by interactively requesting them
(see below for these options).

\item[\fbox{\code{bfs}}]\pindex{bfs}
Similarly to \code{dfs} but use a breadth-first search strategy
to compute and print the values of the given expression.

\item[\fbox{\code{ids}}]\pindex{ids}
Similarly to \code{dfs} but use an iterative-deepening strategy
to compute and print the values of the initial expression.
The initial depth bound is 100 and the depth-bound is
doubled after each iteration.

\item[\fbox{\code{ids $n$}}]\pindex{ids}
Similarly to \code{ids} but use an initial depth bound of $n$.

\item[\fbox{\code{par}}]\pindex{par}
Similarly to \code{dfs} but use a parallel search strategy
to compute and print the values of the initial expression.
The system chooses an appropriate number
of threads according the current number of available processors.

\item[\fbox{\code{par $n$}}]\pindex{par}
Similarly to \code{par} but use $n$ parallel threads.

\item[\fbox{\code{choices $n$}}]\pindex{choices}
Show the internal choice structure
(according to the implementation described in
\cite{BrasselHanusPeemoellerReck11})
resulting from the complete evaluation of the main expression
in a tree-like structure.
This mode is only useful for debugging or
understanding the implementation of non-deterministic
evaluations used in \CYS.
If the optional argument $n$ is provided,
the tree is shown up to depth $n$.

\item[\fbox{\code{supply $i$}}]\pindex{supply}
(not available in global installations, see Section~\ref{sec-install})
Use implementation $i$ as the identifier supply for choice structures
(see \cite{BrasselHanusPeemoellerReck11} for a detailed explanation).
Currently, the following values for $i$ are supported:
\begin{description}
\item[\code{integer}:]\pindex{integer}
Use unbounded integers as choice identifiers.
This implementation is described in \cite{BrasselHanusPeemoellerReck11}.
\item[\code{ghc}:]\pindex{ghc}
Use a more sophisticated implementation of choice
identifiers (based on the ideas described in \cite{AugustssonRittriSynek94})
provided by the Glasgow Haskell Compiler.
\item[\code{pureio}:]\pindex{pureio}
Use IO references (i.e., memory cells) for choice identifiers.
This is the most efficient implementation for top-level depth-first search
but cannot be used for more sophisticated search methods
like encapsulated search.
\item[\code{ioref} (default):]\pindex{ioref}
Use a mixture of \code{ghc} and \code{pureio}.
IO references are used for top-level depth-first search
and \code{ghc} identifiers are used for encapsulated search methods.
\end{description}

\item[\fbox{\code{v$n$}}]\pindex{v}\index{verbosity}
Set the verbosity level to $n$. The following values are allowed
for $n$:
\begin{description}
\item[$n=0$:] Do not show any messages (except for errors).
\item[$n=1$:] Show only messages of the front-end, like loading
of modules.
\item[$n=2$:]
Show also messages of the back end, like compilation messages
from the Haskell compiler.
\item[$n=3$:]
Show also intermediate messages and commands of the compilation
process.
\item[$n=4$:]
Show also all intermediate results of the compilation process.
\end{description}

\item[\fbox{\code{prompt $p$}}]\pindex{prompt}
Sets the user prompt which is shown when \CYS is waiting for input.
If the parameter $p$ starts with a letter or a percent sign,
the prompt is printed as the given parameter,
where the sequence \ccode{\%s} is expanded to the list of
currently loaded modules
and \ccode{\%\%} is expanded to a percent sign.
If the prompt starts with a double quote, it is read as a string and,
therefore, also supports the normal escape sequences that can occur
in Curry programs. The default setting is
\begin{curry}
:set prompt "%s> "
\end{curry}

\item[\fbox{\code{+/-interactive}}]\pindex{interactive}
Turn on/off the interactive mode.
In the interactive mode, the next non-deterministic value
is only computed when the user requests it.
Thus, one has also the possibility to terminate the
enumeration of all values after having seen some values.

\item[\fbox{\code{+/-first}}]\pindex{first}
Turn on/off the first-only mode.
In the first-only mode, only the first value
of the main expression is printed (instead of all values).

\item[\fbox{\code{+/-optimize}}]\pindex{optimize}
Turn on/off the optimization of the target program.

\item[\fbox{\code{+/-bindings}}]\pindex{bindings}
Turn on/off the binding mode.
If the binding mode is on (default),
then the bindings of the free variables of the initial expression
are printed together with the result of the expression.

\item[\fbox{\code{+/-time}}]\pindex{time}
Turn on/off the time mode. If the time mode is on,
the cpu time and the elapsed time
of the computation is always printed together with the result
of an evaluation.

\item[\fbox{\code{+/-ghci}}]\pindex{ghci}
Turn on/off the ghci mode.
In the ghci mode, the initial goal is send to the interactive version
of the Glasgow Haskell Compiler. This might result in a slower
execution but in a faster startup time since the linker
to create the main executable is not used.

\item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
Define additional options passed to the \CYS compiler.
For instance, setting the option
\begin{curry}
:set cmp -O 0
\end{curry}
has the effect that all optimizations performed by the \CYS compiler
are turned off.

\item[\fbox{\code{ghc $opts$}}]\pindex{ghc}
Define additional options passed to the Glasgow Haskell Compiler (GHC)
when the generated Haskell programs are compiled.
One has to be careful when using such options.
For instance, in a global installation of \CYS
(see Section~\ref{sec-install}), libraries are pre-compiled
so that inconsistencies might occur if compilation options
might be changed.

It is safe to pass specific GHC linking options.
For instance, to enforce the static linking of libraries
in order to generate an executable (see command \ccode{:save})
that can be executed in another environment, one could set the
options
\begin{curry}
:set ghc -static -optl-static -optl-pthread
\end{curry}

Other options are useful for experimental purposes,
but those should be used only in local installations
(see Section~\ref{sec-install}) to avoid inconsistent
target codes for different libraries.
For instance, setting the option
\begin{curry}
:set ghc -DDISABLE_CS
\end{curry}
has the effect that the constraint store used to enable
an efficient access to complex bindings is disabled.
Similarly,
\begin{curry}
:set ghc -DSTRICT_VAL_BIND
\end{curry}
has the effect that expressions in a unification constraint
(\code{=:=}) are always fully evaluated
(instead of the evaluation to a head normal form only)
before unifying both sides.
Since these options influence the compilation of the run-time system,
one should also enforce the recompilation of Haskell programs
by the GHC option \ccode{-fforce-recomp}, e.g., one should set
\begin{curry}
:set ghc -DDISABLE_CS -fforce-recomp
\end{curry}

\item[\fbox{\code{rts $opts$}}]\pindex{rts}
Define additional run-time options passed to the executable
generated by the Glasgow Haskell Compiler, i.e., the parameters
\ccode{+RTS $o$ -RTS} are passed to the executable.
For instance, setting the option
\begin{curry}
:set rts -H512m
\end{curry}
has the effect that the minimum heap size is set to 512 megabytes.

\item[\fbox{\code{args $arguments$}}]\pindex{args}
Define run-time arguments passed to the executable
generated by the Glasgow Haskell Compiler.
For instance, setting the option
\begin{curry}
:set args first second
\end{curry}
has the effect that the I/O operation \code{getArgs}
(see library \code{System} (Section~\ref{Library:System})
returns the value \code{["first","second"]}.

\end{description}

\subsection{Source-File Options}

If the evaluation of operations in some main module loaded into
\CYS requires specific options, like an iterative-deepening
search strategy, one can also put these options into the source
code of this module in order to avoid setting these options
every time when this module is loaded.
Such
{\bf source-file options}\index{source-file option}\index{option!in source file}
must occur before the module header, i.e., before the first declaration
(module header, imports, fixity declaration, defining rules, etc)
occurring in the module.
Each source file option must be in a line of the form
\begin{curry}
{-# KiCS2_OPTION $opt$ #-}
\end{curry}
where $opt$ is an option that can occur in a \ccode{:set} command
(compare Section~\ref{sec:options}).
Such a line in the source code (which is a comment according to
the syntax of Curry)
has the effect that this option is set by the \CYS command
\ccode{:set $opt$} whenever this module is loaded (not reloaded!)
as a main module. For instance, if a module starts with the
lines
\begin{curry}
{-# KiCS2_OPTION ids #-}
{-# KiCS2_OPTION +ghci #-}
{-# KiCS2_OPTION v2 #-}
module M where
$\ldots$
\end{curry}
then the load command \ccode{:load M} will also
set the options for iterative deepening, using \code{ghci}
and verbosity level 2.

\subsection{Command Line Editing}

In order to have support for line editing or history functionality
in the command line of \CYS (as often supported by the \code{readline}
library), you should have the Unix command \code{rlwrap} installed
on your local machine.
If \code{rlwrap} is installed, it is used by \CYS if called on a terminal.
If it should not be used (e.g., because it is executed
in an editor with \code{readline} functionality), one can
call \CYS with the parameter \ccode{--noreadline}
(which must occur as the first parameter).

\subsection{Customization}
\label{sec-customization}

In order to customize the behavior of \CYS to your own preferences,
there is a configuration file which is read by \CYS when it is invoked.
When you start \CYS for the first time, a standard version of
this configuration file is copied with the name
\ccode{.kics2rc}\pindex{kics2rc}\pindex{.kics2rc}
into your home directory. The file contains definitions
of various settings, e.g., about showing warnings, using Curry extensions,
programs etc.
After you have started \CYS for the first time, look into this file
and adapt it to your own preferences.

\subsection{Emacs Interface}

Emacs is a powerful programmable editor suitable for program development.
It is freely available for many platforms
(see \url{http://www.emacs.org}).
The distribution of \CYS contains also a special
\emph{Curry mode}\index{Curry mode}\index{Emacs}
that supports the development of Curry programs in
the Emacs environment.
This mode includes support for syntax highlighting,
finding declarations in the current buffer, and
loading Curry programs into \CYS
in an Emacs shell.

The Curry mode has been adapted from a similar mode for Haskell programs.
Its installation is described in the file \code{README}
in directory \ccode{\cyshome/tools/emacs} which also contains
the sources of the Curry mode and a short description about
the use of this mode.

### 2013-07-25 docs/src/usage.tex
Default search strategy changed to breadth-first search.

@@
 
-\item[\fbox{\code{prdfs }}]\pindex{prdfs}
+\item[\fbox{\code{bfs}}]\pindex{bfs}
 Set the search mode to evaluate non-deterministic expressions
-to primitive depth-first search (which is usually the fastest method
-to print all non-deterministic values).
-
-\item[\fbox{\code{dfs}}]\pindex{dfs}
-Set the search mode to evaluate non-deterministic expressions
-to depth-first search. Usually, all non-deterministic values
-are printed with a depth-first strategy, but one can also
+to breadth-first search.
+This is the default search strategy.
+Usually, all non-deterministic values
+are enumerated and printed with a breadth-first strategy, but one can also
 print only the first value or all values by interactively requesting them
@@
 
-\item[\fbox{\code{bfs}}]\pindex{bfs}
-Similarly to \code{dfs} but use a breadth-first search strategy
-to compute and print the values of the given expression.
+\item[\fbox{\code{dfs}}]\pindex{dfs}
+Similarly to \code{bfs} but use a depth-first search strategy
+to compute and print the values of the initial expression.
 
 \item[\fbox{\code{ids}}]\pindex{ids}
-Similarly to \code{dfs} but use an iterative-deepening strategy
+Similarly to \code{bfs} but use an iterative-deepening strategy
 to compute and print the values of the initial expression.
@@
 \item[\fbox{\code{par}}]\pindex{par}
-Similarly to \code{dfs} but use a parallel search strategy
+Similarly to \code{bfs} but use a parallel search strategy
 to compute and print the values of the initial expression.
@@
 Similarly to \code{par} but use $n$ parallel threads.
+
+\item[\fbox{\code{prdfs }}]\pindex{prdfs}
+Set the search mode to evaluate non-deterministic expressions
+to primitive depth-first search. This is usually the fastest method
+to print \emph{all} non-deterministic values.
+However, it does not support the evaluation of values
+by interactively requesting them.
 

### 2013-10-30 docs/src/usage.tex
Change manual to mention the option as 'parallel'

@@
 
-\item[\fbox{\code{par}}]\pindex{par}
+\item[\fbox{\code{parallel}}]\pindex{parallel}
 Similarly to \code{bfs} but use a parallel search strategy
@@
 
-\item[\fbox{\code{par $n$}}]\pindex{par}
-Similarly to \code{par} but use $n$ parallel threads.
+\item[\fbox{\code{parallel $n$}}]\pindex{parallel}
+Similarly to \code{parallel} but use $n$ parallel threads.
 

### 2014-01-31 docs/src/usage.tex
safe mode improved

@@
 
+\item[\fbox{\code{+/-safe}}]\pindex{safe}
+Turn on/off the safe execution mode.
+In the safe execution mode, the initial goal is
+not allowed to be of type \code{IO} and the program should not
+import the module \code{Unsafe}.
+Furthermore, the allowed commands are
+\code{eval}, \code{load}, \code{quit}, and \code{reload}.
+This mode is useful to use \CYS in uncontrolled environments,
+like a computation service in a web page.
+
 \item[\fbox{\code{cmp $opts$}}]\pindex{cmp}
@@
 when the generated Haskell programs are compiled.
-One has to be careful when using such options.
+Many options necessary to compile Curry programs
+are already set (you can see them by setting the verbosity
+level to 2 or greater).
+One has to be careful when providing additional options.
 For instance, in a global installation of \CYS

### 2014-04-08 docs/src/auxiliary_files.tex

\section{Auxiliary Files}
\label{sec-auxfiles}

During the translation and execution of a Curry program with \CYS,
various intermediate representations of the source program are created
and stored in different files which are shortly explained in this section.
In general, it is not necessary to know about
these auxiliary files because they are automatically generated
and updated. You should only remember the command for deleting
all auxiliary files (\ccode{cleancurry}, see Section~\ref{sec-general})
to clean up your directories.

The various components of \CYS create the following auxiliary files.
\begin{description}
\item[\code{prog.fcy}:] This file contains the Curry program
in the so-called ``FlatCurry'' representation where all functions are global
(i.e., lambda lifting has been performed) and pattern matching
is translated into explicit case/or expressions
(compare Appendix~\ref{sec-flatcurry}).
This representation might be useful for other back ends and
compilers for Curry and is the basis doing meta-programming in Curry.
This file is implicitly
generated when a program is compiled with \CYS.
The FlatCurry representation of a Curry program is usually
generated by the front-end after parsing, type checking and eliminating
local declarations.
If $dir$ is the directory where the Curry program is stored,
the corresponding FlatCurry program is stored in the directory
\ccode{$dir$/.curry}.

\item[\code{prog.fint}:] This file contains the interface
of the program in the so-called ``FlatCurry'' representation,
i.e., it is similar to \code{prog.fcy} but contains only exported
entities and the bodies of all functions omitted (i.e., ``external'').
This representation is useful for providing a fast access
to module interfaces.
This file is implicitly generated when a program is compiled with \CYS
and stored in the same directory as \code{prog.fcy}.

\item[\code{Curry_prog.nda}:] This file contains some information about
the determinism behavior of operations that is used by the
\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
for more details about the use of this information).
If $dir$ is the directory where the Curry program is stored,
the corresponding Haskell program is stored in the directory
\ccode{$dir$/.curry/.kics2}.

\item[\code{Curry_prog.info}:] This file contains some
information about the top-level functions of module \code{prog}
that are used by the interactive environment,
like determinism behavior or IO status.
This file is stored in the same directory as \code{Curry_prog.nda}.

\item[\code{Curry_prog.hs}:] This file contains a Haskell program
as the result of translating the Curry program with the
\CYS compiler.
This file is stored in the same directory as \code{Curry_prog.nda}.

\item[\code{Curry_prog.o}:] This file contains the object code
of the Haskell program \code{Curry_prog.hs} when the latter program is
compiled in order to execute it.
This file is stored in the same directory as \code{Curry_prog.hs}.

\item[\code{Curry_prog.hi}:] This file contains the interface
of the Haskell program \code{Curry_prog.hs} when the latter program is
compiled in order to execute it.
This file is stored in the same directory as \code{Curry_prog.hs}.

\item[\code{prog}:] This file contains the executable
after compiling and saving a program with \CYS
(see command \ccode{:save} in Section~\ref{sec:kics2-commands}).

\end{description}

### 2014-04-08 docs/src/currytest.tex

@@
 where \cyshome is the installation directory of \CYS;
-see Section~\ref{sec-install}).
+see Section~\ref{sec-general}).
 In our example, \ccode{TestList.curry} is the program containing the

### 2014-04-08 docs/src/external_ops.tex

\section{External Operations}
\label{sec:external-operations}

\index{operation!external}\index{external operation}
Currently, \CYS has no general interface to external operations,
i.e., operations whose semantics is not defined by program rules
in a Curry program but by some code written in another programming
language.
Thus, if an external operation should be added
to the system, this operation must be declared as \code{external}
in the Curry source code
and an implementation for this external operation
must be provided for the run-time system.
An external operation is defined as follows in the Curry source code:
\begin{enumerate}
\item
Add a type declaration for the external operation somewhere
in a module defining this operation (usually, the prelude
or some system module).
\item
For external operations it is not allowed to define any
rule since their semantics is determined by an external implementation.
Instead of the defining rules, you have to write
\begin{curry}
f external
\end{curry}
below the type declaration for the external operation \code{f}.
\end{enumerate}
Furthermore, an implementation of the external operation
must be provided in the target language of the \CYS compiler,
i.e., in Haskell, and inserted in the compiled code.
In order to simplify this task, \CYS follows some code
conventions that are described in the following.

Assume you want to implement your own concatenation
for strings in a module \code{String}.
The name and type of this string concatenation should be
\begin{curry}
sconc :: String -> String -> String
\end{curry}
Since the primitive Haskell implementation of this operation
does not now anything about the operational mechanism of Curry
(e.g., needed narrowing, non-deterministic rewriting),
the arguments need to be completely evaluated before
the primitive implementation is called.
This can be easily obtained by the prelude operation \code{(\$\#\#)}
that applies an operation to the \emph{normal form} of the given
argument, i.e., this operation evaluates the argument
to its normal form before applying the operation to it.\footnote{%
There is also a similar prelude operation \code{(\$\#)}
which evaluates the argument only to head-normal form.
This is a bit more efficient and can be used for unstructured
types like \code{Bool}.}
Thus, we define \code{sconc} by
\begin{currynomath}
sconc :: String -> String -> String
sconc s1 s2 = (prim_sconc $## s1) $## s2

prim_sconc :: String -> String -> String
prim_sconc external
\end{currynomath}
so that it is ensured that the external operation \code{prim_sconc}
is always called with complete evaluated arguments.

In order to define the Haskell code implementing \code{prim_sconc},
one has to satisfy the naming conventions of \CYS.
The \CYS compiler generates the following code for the
external operation \code{prim_sconc} (note that the generated
Haskell code for the module \code{String} is stored in the file
\code{.curry/kics2/Curry_String.hs}):
\begin{curry}
d_C_prim_sconc :: Curry_Prelude.OP_List Curry_Prelude.C_Char
               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
               -> ConstStore
               -> Curry_Prelude.OP_List Curry_Prelude.C_Char
d_C_prim_sconc x1 x2 x3500 = external_d_C_prim_sconc x1 x2 x3500
\end{curry}
The type constructors \code{OP_List} and \code{C_Char}
of the prelude \code{Curry_Prelude}\footnote{Note that all
translated Curry modules are imported in the Haskell code fully qualified
in order to avoid name conflicts.}
correspond to the Curry type constructors for lists and characters.
The Haskell operation \code{external_d_C_prim_sconc}
is the external operation to be implemented in Haskell by the programmer.
The additional argument of type \code{ConstStore}
represents the current set of constraints when this
operation is called. This argument is intended to provide
a more efficient access to binding constraints and can be
ignored in standard operations.

If \code{String.curry} contains the code
of the Curry function \code{sconc} described above,
the Haskell code implementing the external operations
occurring in the module \code{String} must be in the
file \code{External_String.hs} which is located in the same
directory as the file \code{String.curry}.
The \CYS compiler appends the code contained in
\code{External_String.hs} to the generated code
stored in the file \code{.curry/kics2/Curry_String.hs}.\footnote{%
If the file \code{External_String.hs} contains also
some import declarations at the beginning, these import declarations
are put after the generated import declarations.}

In order to complete our example, we have to write into the
file \code{External_String.hs} a definition of the Haskell function
\code{external_d_C_prim_sconc}.
Thus, we start with the following definitions:
\begin{curry}
import qualified Curry_Prelude as CP

external_d_C_prim_sconc :: CP.OP_List CP.C_Char -> CP.OP_List CP.C_Char
                        -> ConstStore -> CP.OP_List CP.C_Char
\end{curry}
First, we import the standard prelude with the name \code{CP}
in order to shorten the writing of type declarations.
In order to write the final code of this operation,
we have to convert the Curry-related types
(like \code{C_Char}) into the corresponding Haskell types (like \code{Char}).
Note that the Curry-related types contain information about
non-deterministic or constrained values
(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
that are meaningless in Haskell.
To solve this conversion problem, the implementation of \CYS
provides a family of operations to perform these conversions
for the predefined types occurring in the standard prelude.
For instance, \code{fromCurry} converts a Curry type into the
corresponding Haskell type, and \code{toCurry} converts
the Haskell type into the corresponding Curry type.
Thus, we complete our example with the definition
(note that we simply ignore the final argument representing the
constraint store)
\begin{curry}
external_d_C_prim_sconc s1 s2 _ =
  toCurry ((fromCurry s1 ++ fromCurry s2) :: String)
\end{curry}
Here, we use Haskell's concatenation operation \ccode{++}
to concatenate the string arguments.
The type annotation \ccode{:: String} is necessary
because \ccode{++} is a polymorphic function
so that the type inference system of Haskell
has problems to determine the right instance of the conversion
function.

The conversion between Curry types and Haskell types,
i.e., the family of conversion operation \code{fromCurry}
and \code{toCurry}, is defined in the \CYS implementation
for all standard data types.
In particular, it is also defined on function types so that one can
easily implement external Curry I/O actions by using
Haskell I/O actions.
For instance, if we want to implement
an external operation to print some string as an output line,
we start by declaring the external operations in the Curry module \code{String}:
\begin{currynomath}
printString :: String -> IO ()
printString s = prim_printString $## s

prim_printString :: String -> IO ()
prim_printString external
\end{currynomath} % $
Next we add the corresponding implementation in the file
\code{External_String.hs} (where \code{C_IO} and \code{OP_Unit} are the names
of the Haskell representation of the Curry type constructor \code{IO}
and the Curry data type \ccode{()}, respectively):
\begin{curry}
external_d_C_prim_printString :: CP.OP_List CP.C_Char -> ConstStore
                              -> CP.C_IO CP.OP_Unit
external_d_C_prim_printString s _ = toCurry putStrLn s
\end{curry}
Here, Haskell's I/O action \code{putStrLn} of type \ccode{String -> IO ()}
is transformed into a Curry I/O action \ccode{toCurry putStrLn}
which has the type
\begin{curry}
CP.OP_List CP.C_Char -> CP.C_IO CP.OP_Unit
\end{curry}
When we compile the Curry module \code{String},
\CYS combines these definitions in the target program so that
we can immediately use the externally defined operation
\code{printString} in Curry programs.

As we have seen, \CYS transforms a name like
\code{primOP} of an external operation into the name
\code{external_d_C_primOP} for the Haskell operation
to be implemented, i.e., only a specific prefix is added.
However, this is only valid if no special characters occur
in the Curry names.
Otherwise (in order to generate a correct Haskell program),
special characters are translated into specific
names prefixed by \ccode{OP_}. For instance,
if we declare the external operation
\begin{curry}
(<&>) :: Int -> Int -> Int
(<&>) external
\end{curry}
the generated Haskell module contains the code
\begin{curry}
d_OP_lt_ampersand_gt :: Curry_Prelude.C_Int -> Curry_Prelude.C_Int
                     -> ConstStore -> Curry_Prelude.C_Int
d_OP_lt_ampersand_gt x1 x2 x3500 = external_d_OP_lt_ampersand_gt x1 x2 x3500
\end{curry}
so that one has to implement the operation
\code{external_d_OP_lt_ampersand_gt} in Haskell.
If in doubt, one should look into the generated Haskell code
about the names and types of the operations to be implemented.

Finally, note that this method to connect functions implemented in Haskell
to Curry programs provides the opportunity to connect
also operations written in other programming languages
to Curry via Haskell's foreign function interface.

### 2014-04-08 docs/src/overview.tex

\section{Overview of \CYS}

\subsection{Installation}
\label{sec-install}

This version of \CYS has been developed and
tested on Linux systems.
In principle, it should be also executable on other
platforms on which a Haskell implementation (Glasgow Haskell Compiler
and Cabal) exists, like in many Linux distributions, Sun Solaris,
or Mac OS X systems.

Installation instructions for \CYS can be found in
the file \code{INSTALL.txt} stored in the \CYS installation directory.
Note that there are two possibilities to install \CYS:
\begin{description}
\item[Global installation:]\index{global installation}\index{installation!global}
\CYS is installed in some global system directory where users
have no write permission. In this case, some options for experimenting
with \CYS (like \code{supply} or \code{ghc}, see below)
are not available (since they require the recompilation of parts of
the installed system).
\item[Local installation:]\index{local installation}\index{installation!local}
\CYS is installed in some local user directory where the user
has write permission and the option \code{GLOBALINSTALL}
in the \code{Makefile} of the \CYS installation is set as follows:
\begin{lstlisting}
GLOBALINSTALL=no
\end{lstlisting}
In this case, all options of \CYS are available.
\end{description}
In the following, \cyshome denotes the installation directory
of the \CYS installation.

\subsection{General Use}
\label{sec-general}

All executables required to use the different components
of \CYS are stored in the directory \code{\cyshome/bin}.
You should add this directory
to your path (e.g., by the \code{bash} command
\ccode{export PATH=\cyshome/bin:\$PATH}).

The source code of the Curry program
must be stored in a file with the suffix \ccode{.curry},
e.g., \code{prog.curry}.
Literate programs must be stored in files with the extension \ccode{.lcurry}.

Since the translation of Curry programs with \CYS creates
some auxiliary files (see Section~\ref{sec-auxfiles} for details),
you need write permission
in the directory where you have stored your Curry programs.
Moreover, the current implementation also recompiles
system libraries according to the setting of some options.
Therefore, the \CYS system should be locally installed
in your user account.
The auxiliary files for all Curry programs in the current
directory can be deleted by the command\pindex{cleancurry}
\begin{curry}
cleancurry
\end{curry}
(this is a shell script stored in the \code{bin} directory of the
\CYS installation, see above).
The command
\begin{curry}
cleancurry -r
\end{curry}
also deletes the auxiliary files in all subdirectories.

\subsection{Restrictions}
\label{sec-restrictions}

There are a few minor restrictions on Curry programs
when they are processed with \CYS:
\begin{itemize}
\item
\index{singleton variables}\index{variables!singleton}
\emph{Singleton pattern variables}, i.e., variables that occur only once
in a rule, should be denoted as an anonymous variable \ccode{_},
otherwise the parser will print a warning since this is a
typical source of programming errors.
\item
\CYS translates all \emph{local declarations} into global functions with
additional arguments (``lambda lifting'', see Appendix~D of the
Curry language report).
Thus, in the various run-time systems, the definition of
functions with local declarations look different from
their original definition (in order to see the result
of this transformation, you can use the \cb, see
Section~\ref{sec-currybrowser}).
\item \index{tabulator stops}
Tabulator stops instead of blank spaces in source files are
interpreted as stops at columns 9, 17, 25, 33, and so on.
In general, tabulator stops should be avoided in source programs.
\item
Encapsulated search\index{encapsulated search}:
The general definition of encapsulated search of the Curry report
\cite{HanusSteiner98PLILP} is not supported.
Thus, the corresponding prelude operations like
\code{try}\pindex{try},
\code{solveAll}\pindex{solveAll},
\code{once}\pindex{once},
\code{findall}\pindex{findall}, or
\code{best}\pindex{best}
are not defined in the \CYS prelude.
However, \CYS supports appropriate alternatives
to encapsulate non-deterministic computations:
\begin{description}
\item[Strong encapsulation:]
This means that all potential
non-determinism is encapsulated. Since this might
result in dependencies on the evaluation strategy
(see \cite{BrasselHanusHuch04JFLP} for a detailed discussion),
this kind of encapsulation is only available as I/O operations.
For instance, the library \code{AllSolutions}\pindex{AllSolutions}
(Section~\ref{Library:AllSolutions})
defines the operation
\begin{curry}
getAllValues :: a -> IO [a]
\end{curry}
to compute all values of a given argument expression.
There is also the library \code{SearchTree}\pindex{SearchTree}
(Section~\ref{Library:SearchTree})
which supports user-programmable search strategies
and contains some predefined strategies
like depth-first, breadth-first, iterative deepening search.
\item[Weak encapsulation:]
This means that only the non-determinism
defined inside an encapsulation operator is encapsulated.
Conceptually, these operators are offered as
\emph{set functions}\index{set functions} \cite{AntoyHanus09}
which compute the set of all results but do not encapsulate
non-determinism in the actual arguments.
See the library \code{SetFunctions}
(Section~\ref{Library:SetFunctions}) for more details.
\end{description}
\item
Concurrent computations\index{concurrency}
based on the suspension of expressions
containing free variables are not yet supported.
\CYS supports \emph{value generators} for free variables
so that a free variable is instantiated when its value is demanded.
For instance, the initial expression
\begin{curry}
x == True where x free
\end{curry}
is non-deterministically evaluated to \code{False} and \code{True}
by instantiating \code{x} to \code{False} and \code{True}, respectively.
Thus, a computation is never suspended due to free variables.
This behavior also applies to free variables of primitive types
like integers. For instance, the initial expression
\begin{curry}
x*y=:=1 where x,y free
\end{curry}
is non-deterministically evaluated to the two solutions
\begin{curry}
{x = -1, y = -1} Success
{x = 1, y = 1} Success
\end{curry}
\item
Unification is performed without an occur check.
\item
There is currently no general connection to external constraint solvers.
\end{itemize}

\subsection{Modules in \CYS}
\label{sec-modules}

The current implementation of \CYS supports only flat module names,
i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
In order to allow the structuring of modules in different directories,
\CYS searches for imported modules in various directories.
By default, imported modules are searched in the directory
of the main program and the system module directories
\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
This search path can be extended
by setting the environment variable \code{CURRYPATH}\pindex{CURRYPATH}
(which can be also set in a \CYS session by the option
\ccode{:set path}\pindex{path}\pindex{:set path},
see below)
to a list of directory names separated by colons (\ccode{:}).
In addition, a local standard search path
can be defined in the \ccode{.kics2rc} file
(see Section~\ref{sec-customization}).
Thus, modules to be loaded are searched in the following
directories (in this order, i.e., the first occurrence of a module file
in this search path is imported):
\begin{enumerate}
\item Current working directory (\ccode{.}) or directory prefix
of the main module (e.g., directory \ccode{/home/joe/curryprogs}
if one loads the Curry program \ccode{/home/joe/curryprogs/main}).
\item The directories enumerated in the environment variable \code{CURRYPATH}.
\item The directories enumerated in the \ccode{.kics2rc} variable
      \ccode{libraries}.
\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
\end{enumerate}
Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
will be always implicitly imported to all modules if a module
does not contain an explicit import declaration for the module
\code{Prelude}.

### 2014-04-08 docs/src/preface.tex

\addcontentsline{toc}{section}{Preface}
\section*{Preface}

This document describes \CYS (\textbf{Ki}el \textbf{C}urry \textbf{S}ystem
Version \textbf{2}),
an implementation of the multi-paradigm language Curry
\cite{Hanus97POPL,Hanus06Curry} that is
based on compiling Curry programs into Haskell programs.
Curry is a universal programming language aiming at the amalgamation
of the most important declarative programming paradigms,
namely functional programming and logic programming.
Curry combines in a seamless way features from functional programming
(nested expressions, lazy evaluation, higher-order functions),
logic programming (logical variables, partial data structures,
built-in search), and concurrent programming (concurrent evaluation
of constraints with synchronization on logical variables).
The current \CYS implementation does not support concurrent
constraints. Alternatively, one can write distributed applications
by the use of sockets that can be registered and accessed
with symbolic names.
Moreover, \CYS also supports
the high-level implementation of
graphical user interfaces and web services
(as described in more detail in
\cite{Hanus99PPDP,Hanus00PADL,Hanus01PADL,Hanus06PPDP}).

We assume familiarity with the ideas and features
of Curry as described in the Curry language definition \cite{Hanus12Curry}.
Therefore, this document only explains the use of the different
components of \CYS
and the differences and restrictions of \CYS
(see Section~\ref{sec-restrictions})
compared with the language Curry (Version 0.8.3).
The basic ideas of the implementation of \CYS
can be found in
\cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP}.

\bigskip

\subsection*{Acknowledgements}

This work has been supported in part by the
DFG grants Ha 2457/5-1 and Ha 2457/5-2.

### 2014-04-08 docs/src/problems.tex

\section{Technical Problems}

One can implement distributed systems with \CYS
by the use of the library \code{NamedSocket} (Section~\ref{Library:NamedSocket})
that supports a socket communication with symbolic names
rather than natural numbers.
For instance, this library is the basis of programming
dynamic web pages with the libraries
\code{HTML} (Section~\ref{Library:HTML})
or \code{WUI} (Section~\ref{Library:WUI}).
However, it might be possible that some technical problems
arise due to the use of named sockets.
Therefore, this section gives some information
about the technical requirements of \CYS and how to solve
problems due to these requirements.

There is one fixed port that is used by the implementation of \CYS:
\begin{description}
\item[Port 8767:] This port is used by the
{\bf Curry Port Name Server} (CPNS) to implement symbolic names for
named sockets in Curry.
If some other process uses this port on the machine,
the distribution facilities defined in the module \code{NamedSocket}
cannot be used.
\end{description}
If these features do not work, you can try to find out
whether this port is in use by the shell command
\ccode{netstat -a | fgrep 8767} (or similar).

The CPNS is implemented as a demon listening on its port 8767
in order to serve requests about registering a new symbolic
name for a named socket or asking the physical port number
of an registered named socket. The demon will be automatically started for
the first time on a machine when a user runs a program
using named sockets.
It can also be manually started and terminated by the
scripts \code{\cyshome/cpns/start} and
\code{\cyshome/cpns/stop}.
If the demon is already running, the command \code{\cyshome/cpns/start}
does nothing (so it can be always executed
before invoking a Curry program using named sockets).

If you detect any further technical problem,
please write to
\begin{center}
\code{mh@informatik.uni-kiel.de}
\end{center}

### 2014-04-08 docs/src/usage.tex

@@
 
-\item[\fbox{\code{+/-safe}}]\pindex{safe}
-Turn on/off the safe execution mode.
+\item[\fbox{\code{safe}}]\pindex{safe}
+Turn on the safe execution mode.
 In the safe execution mode, the initial goal is
@@
 This mode is useful to use \CYS in uncontrolled environments,
-like a computation service in a web page.
+like a computation service in a web page, where \CYS could
+be invoked by
+\begin{curry}
+kics2 :set safe
+\end{curry}
 

### 2014-04-08 talks/2013-padl/slides.tex

--- 2012-04-24 talks/wlp2011/slides.tex
@@
 
-\date{WLP 2011}
-\author[Michael Hanus]{Michael Hanus\\[2ex]
-{\small Joint work with Bernd Bra\ss{}el, Bj\"orn Peem\"oller and Fabian Reck}}
+\date{PADL 2013}
+\author[Fabian Reck]{Fabian Reck\\[2ex]
+{\small Joint work with Bernd Bra\ss{}el, Michael Hanus and Bj\"orn Peem\"oller}}
 \institute[CAU Kiel]{Christian-Albrechts-University of Kiel\\
@@
 (constraint solvers for specific domains)
-\item avoid non-declarative features of Prolog\\
-(arithmetic, cut, I/O, side-effects)
+\item avoid non-declarative features
 \end{itemize}
@@
 \item extension of Haskell (non-strict functional language)
-\item SWE advantage: better (high-level) APIs for various application domains\\
-(GUI programming, web programming, database programming,\ldots)
 \end{itemize}
@@
 
-\begin{exampleblock}{Example \hspace{4cm}\only<3>{(with \alert{equational constraint}~ \texttt{=:=})}\only<4>{(with \alert{functional pattern})}}
+\begin{exampleblock}{Example \hspace{4cm}\only<3>{(with \alert{equational constraint}~ \texttt{=:=})}}
 \begin{curry}
-(++) :: [a] -> [a] -> [a]    $\only<3-4>{\code{last :: [a] -> a}}$
-[]     ++ ys = ys              $\only<3>{\code{last xs | \alert{ys ++ [x] =:= xs}}}$$\only<4>{\code{last \alert{(ys ++ [x])} = x}}$
+(++) :: [a] -> [a] -> [a]    $\only<3>{\code{last :: [a] -> a}}$
+[]     ++ ys = ys              $\only<3>{\code{last xs | \alert{ys ++ [x] =:= xs}}}$
 (x:xs) ++ ys = x : xs ++ ys             $\only<3>{\code{= x  where ys,x free}}$
@@
 \begin{frame}[fragile]
-\frametitle{Implementing Curry}
-
-\begin{block}{Abstract machines}
-\begin{itemize}
-\item C (MCC, M\"unster Curry Compiler)
-\item Java (FLVM, Portland)
-\end{itemize}
-\end{block}
-\pause
-
-\begin{block}{Compiling to Prolog}
-\begin{itemize}
-\item PAKCS (Portland Aachen Kiel Curry System), TOY (Madrid)
-\item reuse non-determinism (backtracking), logic variables, constraint solvers
-\item implement laziness, higher-order
-\end{itemize}
-\end{block}
-\pause
+\frametitle{Implementing Curry: KiCS2}
 
@@
 \end{block}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}[fragile]
-\frametitle{Compiling Curry to Haskell: KiCS2}
-
+\pause
 \begin{block}{Principles of KiCS2}
@@
 \begin{frame}[fragile]
-\frametitle{Representing Non-deterministic Values}
+\frametitle{Representing Non-deterministic Results}
 
@@
 \begin{frame}[fragile]
-\frametitle{Depth-first Search}
-\vspace{-1ex}
-\begin{exampleblock}{Print all values of a choice structure in depth-first order}
-\vspace{-1ex}
-\begin{curry}
-  printValsDFS :: a -> IO ()$\listline$
-  printValsDFS Fail             = return ()$\listline$
-  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
-   where
-    follow ChooseLeft  = printValsDFS x1
-    follow ChooseRight = printValsDFS x2
-    follow NoDecision    = do newDecision ChooseLeft  x1
-                              newDecision ChooseRight x2
-
-    newDecision d x = do setDecision i d
-                         printValsDFS x
-                         setDecision i NoDecision$\listline$
-  printValsDFS v = print v
-\end{curry}
-\vfill\pause
-
-Main call: \code{initSupply >>= \bs{}s -> printValsDFS (main s)}\\
-Result:~~~~ \code{False}\\
-\hspace{1.62cm}\code{False}
-\end{exampleblock}
-
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Benchmarks {\lgrey [WFLP'11]}}
-
-\begin{block}{Deterministic Programs}
-\begin{center}
-\begin{tabular}{|@{~~}l@{~~}|l@{~~}|*{4}{@{~~}r@{~~}|}}
-\hline
-System  & Target   & ReverseUser & Reverse &  Tak  & TakPeano \\\hline
-KiCS2   & Haskell  &      0.12   &   0.12  &  0.21 &    0.79  \\[0.5ex]
-PAKCS   & SICStus-Prolog    &      2.05   &   1.88  & 39.80 &   62.43  \\[0.5ex]
-MCC     & VM/C     &      0.43   &   0.47  &  1.21 &    5.49  \\[0.5ex]
-\hline
-\end{tabular}
-\end{center}
-\end{block}
-\vfill
-
-\begin{block}{Non-deterministic Programs}
-\begin{center}
-\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
-\hline
-System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
-KiCS2   &    2.83  &        3.68   &  0.14 &   0.49 \\[0.5ex]
-PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\[0.5ex]
-MCC     &    1.46  &        5.74   &  0.09 &   0.57 \\[0.5ex]
-\hline
-\end{tabular}
-\end{center}
-\end{block}
-\vfill
-\begin{block}{}
-  \begin{itemize}
-  \item faster than Prolog-based implementations (PAKCS)
-  \item competitive with VM/C-based implementation (MCC)
-  \end{itemize}
-\end{block}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
-\begin{frame}[fragile]
 \frametitle{Logic Variables}
@@
 \begin{block}{}
-\citebib{Antoy/H. ICLP'06}:\\[1ex]
+\citebib{Antoy/Hanus ICLP'06}:\\[1ex]
 narrowing (resolution) on logic variables \\
@@
 
+\begin{block}{Comparing different representations for equations in KiCS2}
+\begin{center}
+\begin{tabular}{|l|r|r|r|}
+\hline
+Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
+\hline
+\code{last (map (inc 0) [1..10000])}   &  2.91   &    0.05   &    0.01   \\
+\code{simplify}                         & 10.30      & 6.77      & 7.07        \\
+\code{varInExp}                         & 2.34    & 0.24      & 0.21        \\
+\code{fromPeano (half (toPeano 10000))} & 26.67     & 5.95      & 11.19       \\
+\code{palindrome}                       & 30.86     & 14.05     & 20.26       \\
+\code{horseman}                         & 3.24      & 3.31      & n/a         \\
+\code{grep}                             & 1.06      & 0.10      & n/a         \\
+\hline
+\end{tabular}
+\end{center}
+\end{block}
+\vfill
+
+\begin{block}{}
+\begin{description}
+\item[\code{==}~:] Boolean equality without variable bindings
+\item[\code{=:=}~:] equational constraint with variable bindings
+\item[\code{=:<=}~:] lazy unification with unevaluated variable bindings
+\end{description}
+\end{block}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
+\begin{frame}
+\frametitle{Conclusions}
+\vspace{-1ex}
+\begin{block}{}
+\begin{itemize}
+\item novel implementation techniques for equational constraints
+\item add binding constraints to computed values
+\item process them when extracting values from search space
+\item only new constructors added, no overhead for existing implementation
+\item more efficient execution if features are used
+\end{itemize}
+\end{block}
+\vfill
+\begin{block}{Advantages of KiCS2}
+\begin{itemize}
+\item highly efficient for deterministic parts, competitive for non-det. programs
+\item flexible search strategies
+\item many opportunities for optimizations (determinism analysis)
+\item exploit future improvements of Haskell implementations
+\end{itemize}
+\end{block}
+\vfill
+\begin{block}{Future Work}
+\begin{itemize}
+\item Add further constraint structures
+\end{itemize}
+\end{block}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\frametitle{Benchmarks {\lgrey [WFLP'11]}}
+
+\begin{block}{Deterministic Programs}
+\begin{center}
+\begin{tabular}{|@{~~}l@{~~}|l@{~~}|*{4}{@{~~}r@{~~}|}}
+\hline
+System  & Target   & ReverseUser & Reverse &  Tak  & TakPeano \\\hline
+KiCS2   & Haskell  &      0.12   &   0.12  &  0.21 &    0.79  \\[0.5ex]
+PAKCS   & SICStus-Prolog    &      2.05   &   1.88  & 39.80 &   62.43  \\[0.5ex]
+MCC     & VM/C     &      0.43   &   0.47  &  1.21 &    5.49  \\[0.5ex]
+\hline
+\end{tabular}
+\end{center}
+\end{block}
+\vfill
+
+\begin{block}{Non-deterministic Programs}
+\begin{center}
+\begin{tabular}{|@{~~}l@{~~}|*{4}{@{~~}r@{~~}|}}
+\hline
+System  & PermSort & PermSortPeano &  Last & RegExp \\\hline
+KiCS2   &    2.83  &        3.68   &  0.14 &   0.49 \\[0.5ex]
+PAKCS   &   26.96  &       67.11   &  2.61 &  12.70 \\[0.5ex]
+MCC     &    1.46  &        5.74   &  0.09 &   0.57 \\[0.5ex]
+\hline
+\end{tabular}
+\end{center}
+\end{block}
+\vfill
+\begin{block}{}
+  \begin{itemize}
+  \item faster than Prolog-based implementations (PAKCS)
+  \item competitive with VM/C-based implementation (MCC)
+  \end{itemize}
+\end{block}
+\end{frame}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{frame}
+\frametitle{Benchmarks}
+
 \begin{block}{Equational constraints}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{frame}
-\frametitle{Benchmarks}
-
-\begin{block}{Comparing different representations for equations in KiCS2}
-\begin{center}
-\begin{tabular}{|l|r|r|r|}
-\hline
-Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
-\hline
-\code{last (map (inc 0) [1..10000])}   &  2.91   &    0.05   &    0.01   \\
-\code{simplify}                         & 10.30      & 6.77      & 7.07        \\
-\code{varInExp}                         & 2.34    & 0.24      & 0.21        \\
-\code{fromPeano (half (toPeano 10000))} & 26.67     & 5.95      & 11.19       \\
-\code{palindrome}                       & 30.86     & 14.05     & 20.26       \\
-\code{horseman}                         & 3.24      & 3.31      & n/a         \\
-\code{grep}                             & 1.06      & 0.10      & n/a         \\
-\hline
-\end{tabular}
-\end{center}
-\end{block}
-\vfill
-
-\begin{block}{}
-\begin{description}
-\item[\code{==}~:] Boolean equality without variable bindings
-\item[\code{=:=}~:] equational constraint with variable bindings
-\item[\code{=:<=}~:] lazy unification with unevaluated variable bindings
-\end{description}
-\end{block}
-\end{frame}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
-\begin{frame}
-\frametitle{Conclusions}
-\vspace{-1ex}
-\begin{block}{}
-\begin{itemize}
-\item novel implementation techniques for equational constraints
-\item add bindings constraints to computed values
-\item process them when extracting values from search space
-\item only new constructors added, no overhead for existing implementation
-\item more efficient execution if features are used
-\end{itemize}
-\end{block}
-\vfill
-\begin{block}{Advantages of KiCS2}
-\begin{itemize}
-\item highly efficient for deterministic parts, competitive for non-det. programs
-\item flexible search strategies
-\item many opportunities for optimizations (determinism analysis)
-\item exploit future improvements of Haskell implementations
-\end{itemize}
-\end{block}
-\vfill
-\begin{block}{Future Work}
-\begin{itemize}
-\item Add further constraint structures
-\end{itemize}
-\end{block}
-\end{frame}
-
 \end{document}

### 2014-05-26 docs/src/usage.tex
Manual updated

@@
 
+\item[\fbox{\code{parser $opts$}}]\pindex{parser}
+Define additional options passed to the \CYS front end, i.e.,
+the parser program \code{\cyshome/bin/cymake}.
+For instance, setting the option
+\begin{curry}
+:set parser -F --pgmF=transcurry
+\end{curry}
+has the effect that each Curry module to be compiled is
+transformed by the preprocessor command \code{transcurry}
+into a new Curry program which is actually compiled.
+
 \item[\fbox{\code{cmp $opts$}}]\pindex{cmp}

### 2014-08-07 docs/src/extensions.tex
Changed and documented the representation of numbers with free variables

@@
 part of the definition of Curry. These extensions are described below.
+
+\subsection{Narrowing on Int Literals}
+In addition to narrowing on algebraic data types,
+\CYS alos implements narrowing on values of the primitive type \code{Int}.
+For example, the goal \code{x == 3 where x free}
+can be narrowed to the solutions
+\begin{curry}
+Prelude> x == 3 where x free
+{x = (-_x2)       } False
+{x = 0            } False
+{x = 1            } False
+{x = (2 * _x3)    } False
+{x = 3            } True
+{x = (4 * _x4 + 1)} False
+{x = (4 * _x4 + 3)} False
+\end{curry}
+Note that the free variables occuring in the binding are restricted to
+positive numbers greater than $0$
+(the output has been indented to increase readability).
+This feature is realized by an internal binary representation
+of integer numbers. If necessary, this representation can be exposed to
+the user by setting the flag \code{BinaryInt} during installation:
+\begin{curry}
+make [kernel|install] RUNTIMEFLAGS=BinaryInt
+\end{curry}
+In an experimental (local) installation, the flag can also be set in the
+interpreter:
+\begin{curry}
+:set ghc -DBinaryInt
+\end{curry}
+The example above will then be evaluated (without indentation) to:
+\begin{curry}
+Prelude> x == 3 where x free
+{x = (Neg _x2)        } False
+{x = 0                } False
+{x = 1                } False
+{x = (Pos (O _x3))    } False
+{x = 3                } True
+{x = (Pos (I (O _x4)))} False
+{x = (Pos (I (I _x4)))} False
+\end{curry}
+In this output, values without free variables are presented like before.
+For value containing a free variable,
+the constructors \code{Neg} and \code{Pos} denote negative and
+positive numbers (without $0$), while the constructors \code{O} and \code{I}
+denote a $0$-- and $1$--bit where the \emph{least significant bit comes first}.
+That is,
+$\code{(Pos (I (O _x4)))} = +\code{(I (O _x4))} = + (2 * \code{(O _x4)}) + 1 = + (4 * \code{_x4}) + 1$
+which meets the output above.
 

### 2014-08-07 runtime/PrimTypes.hs

@@ -1,3 +1,4 @@
+{-# LANGUAGE CPP #-}
 {-# LANGUAGE FlexibleInstances, MultiParamTypeClasses #-}
 module PrimTypes where
 
@@ -21,11 +22,21 @@ data BinInt
 instance Show BinInt where
   showsPrec d (Choice_BinInt cd i x y) = showsChoice d cd i x y
   showsPrec d (Choices_BinInt cd i xs) = showsChoices d cd i xs
-  showsPrec d (Guard_BinInt cd c e) = showsGuard d cd c e
-  showsPrec _ (Fail_BinInt _ _) = showChar '!'
-  showsPrec _ (Neg x1) = (showString "(Neg") . ((showChar ' ') . ((shows x1) . (showChar ')')))
-  showsPrec _ Zero = showString "Zero"
-  showsPrec _ (Pos x1) = (showString "(Pos") . ((showChar ' ') . ((shows x1) . (showChar ')')))
+  showsPrec _ (Fail_BinInt        _ _) = showChar '!'
+  showsPrec d (Guard_BinInt    cd c e) = showsGuard d cd c e
+#ifdef BinaryInt
+  showsPrec _ (Neg                 x1) = (showString "(Neg") . ((showChar ' ')
+                                       . ((shows x1) . (showChar ')')))
+  showsPrec _ Zero                     = showString "Zero"
+  showsPrec _ (Pos                 x1) = (showString "(Pos") . ((showChar ' ')
+                                       . ((shows x1) . (showChar ')')))
+#else
+  showsPrec d (Neg                  x) = showString "(-" . showsPrec d x
+                                       . showChar ')'
+  showsPrec _ Zero                     = showChar '0'
+  showsPrec d (Pos                  x) = showsPrec d x
+#endif
+
 
 
 instance Read BinInt where
@@ -118,11 +129,28 @@ data Nat
 instance Show Nat where
   showsPrec d (Choice_Nat cd i x y) = showsChoice d cd i x y
   showsPrec d (Choices_Nat cd i xs) = showsChoices d cd i xs
-  showsPrec d (Guard_Nat cd c e) = showsGuard d cd c e
-  showsPrec _ (Fail_Nat _ _) = showChar '!'
-  showsPrec _ IHi = showString "IHi"
-  showsPrec _ (O x1) = (showString "(O") . ((showChar ' ') . ((shows x1) . (showChar ')')))
[…]

### 2014-08-22 docs/src/usage.tex
currytools updated, bin/curry added

@@
 
-\subsection{Command of \CYS}
+\subsection{Commands of \CYS}
 \label{sec:kics2-commands}
@@
 its value is used as a command to show the source text,
-otherwise the \ccode{cat} is used.
+otherwise the command \ccode{cat} is used.
 

### 2014-10-06 benchmarks/benchmark_papers/kics2/bench_kics2.tex
Benchmark suite updated and benchmark paper added

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This is a paper to benchmark various versions of the KiCS2 system.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}

\usepackage{currycode}
\usepackage{graphicx}

\begin{document}
\sloppy

\title{Benchmarking KiCS2}
\author{Michael Hanus}
\maketitle

\section{Introduction}

This document contains the result of benchmarking various
Curry systems.
%
All benchmarks were executed in the following environment:
\begin{itemize}
\item \runcurry{getOS} machine (named ``\runcurry{getHostName}'')
\item operating system: \runcurry{getSystemID} \runcurry{getSystemRelease}
\item CPU: ``\runcurry{getCPUModel}'' processor containing
      \runcurry{getCoreNumber} cores
\item time limit for each benchmark: \runcurry{return (show timeLimit)} seconds
\item number of iterations for each benchmark:
       \runcurry{return (show numberOfRuns)}
\end{itemize}

\section{Benchmarks}

%%% The actual benchmark programs are in a separate Curry module:
\begin{curryprog}
----------------------------------------------------------------------
import Benchmarks
import BenchKiCS2
----------------------------------------------------------------------
\end{curryprog}

\subsection{Deterministic Benchmarks}

The following diagram compares MCC, PAKCS, and KiCS2
on the classical naive reverse benchmark for different list lengths:

\begin{center}
\runcurry{benchNRevAndPlot 5.2 [500,1000..] "plot_all"}
\end{center}

Next we test various Curry systems on deterministic operations,
i.e., operations without overlapping rules or free variables:

\begin{center}
%\runcurry{mainBench (MCC:PAKCS:kics2Systems) detBench}
\end{center}

\subsection{Non-deterministic DFS Benchmarks}

The following benchmark programs contain non-deterministic operations
or free variables. We evaluate the computation of a
first solution/value via DFS in PAKCS vs. KiCS2:

\begin{center}
%\runcurry{mainBench [PAKCS, KiCS2 ""] nondetBenchDFS}
\end{center}

\subsection{Non-deterministic Benchmarks}

\begin{center}
%\runcurry{mainBench kics2Systems nondetBench}
\end{center}

\subsection{Encapsulated Search Benchmarks}

\begin{center}
%\runcurry{mainBench kics2Systems encapsBench}
\end{center}

\end{document}

### 2014-10-07 benchmarks/benchmark_papers/kics2/bench_kics2.tex
benchmark paper corrected
Parallel search benchmarks added.
Benchmarks: MCC handling improved, nondet benchmarks refined

@@
 \usepackage{graphicx}
+
+\setlength{\parindent}{0ex}
 
@@
 \begin{center}
-%\runcurry{mainBench (MCC:PAKCS:kics2Systems) detBench}
+\runcurry{mainBench (MCC:PAKCS:kics2Systems) detBench}
 \end{center}
@@
 The following benchmark programs contain non-deterministic operations
-or free variables. We evaluate the computation of a
-first solution/value via DFS in PAKCS vs. KiCS2:
+or free variables. We evaluate the computation of all
+solutions/values via DFS in various Curry systems:
 
 \begin{center}
-%\runcurry{mainBench [PAKCS, KiCS2 ""] nondetBenchDFS}
+\runcurry{mainBench [PAKCS, MCC, KiCS2 ""] nondetBenchDFS}
+\end{center}
+
+These benchmarks contain functional patterns which are not supported in MCC:
+
+\begin{center}
+\runcurry{mainBench [PAKCS, KiCS2 ""] funpatBenchDFS}
 \end{center}
@@
 
+The following benchmark programs compare various search strategies
+as they are supported in KiCS2.
+
+The first set of benchmarks computes all solutions/values:
+
 \begin{center}
-%\runcurry{mainBench kics2Systems nondetBench}
+\runcurry{mainBench kics2Systems nondetBench}
+\end{center}
+
+The following benchmarks computes the first solution/value
+only (since the search space is infinite):
+
+\begin{center}
+\runcurry{mainBench kics2Systems nondetBenchFirst}
+\end{center}
+
+\subsection{Parallel Non-deterministic Benchmarks}
+
+\begin{center}
+\runcurry{mainParBench (KiCS2 "") parSearchBench}
 \end{center}
@@
 \begin{center}
-%\runcurry{mainBench kics2Systems encapsBench}
+\runcurry{mainBench kics2Systems encapsBench}
 \end{center}

### 2014-10-08 docs/src/usage.tex

@@
 
-\item[\fbox{\code{:add} $m$}]\pindex{:add}
-Add module $m$ to the set of currently loaded modules
-so that its exported entities are available in the top-level environment.
+\item[\fbox{\code{:add} $m_1 \ldots m_n$}]\pindex{:add}
+Add modules $m_1,\ldots,m_n$ to the set of currently loaded modules
+so that their exported entities are available in the top-level environment.
 
@@
 \begin{curry}
-xs++ys =:= [1,2]  where xs,ys free
+not b where b free
 \end{curry}
-in order to obtain the following three possible bindings:
+in order to obtain the following bindings and results:
 \begin{curry}
-{xs = [], ys = [1,2]} Success
-{xs = [1], ys = [2]} Success
-{xs = [1,2], ys = []} Success
+{b = False} True
+{b = True} False
 \end{curry}

### 2014-10-08 benchmarks/benchmark_papers/kics2/bench_kics2.tex
PDF terminal for benchmark plots added

@@
 \begin{center}
-\runcurry{benchNRevAndPlot 5.2 [500,1000..] "plot_all"}
+\runcurry{benchNRevAndPlot 5.2 [500,1000..] "plot_all.pdf"}
 \end{center}
@@
 or free variables. We evaluate the computation of all
-solutions/values via DFS in various Curry systems:
+solutions/values via the depth-first strategy (DFS)
+with various Curry systems:
 
@@
 
-\subsection{Non-deterministic Benchmarks}
+\subsection{Search Strategy Benchmarks for KiCS2}
 
@@
 
+These benchmarks are executed with the option ``\texttt{:set +parallel}''
+and different run-time settings to influence the number
+of threads used for the execution.
+
 \begin{center}

### 2014-10-14 docs/src/extensions.tex
Manual extended and some typos fixed

@@
 \subsection{Narrowing on Int Literals}
+
 In addition to narrowing on algebraic data types,
-\CYS alos implements narrowing on values of the primitive type \code{Int}.
-For example, the goal \code{x == 3 where x free}
-can be narrowed to the solutions
+\CYS also implements narrowing on values of the primitive type \code{Int}.
+For example, the goal \ccode{x == 3 where x free}
+is evaluated to the solutions
 \begin{curry}
@@
 (the output has been indented to increase readability).
-This feature is realized by an internal binary representation
+This feature is implemented by an internal binary representation
 of integer numbers. If necessary, this representation can be exposed to
@@
 \end{curry}
-In this output, values without free variables are presented like before.
-For value containing a free variable,
+In this output, values without free variables are presented as before.
+For values containing a free variable,
 the constructors \code{Neg} and \code{Pos} denote negative and
@@
 \end{itemize}
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manual"
+%%% End:

### 2014-10-14 docs/src/usage.tex
Manual extended and some typos fixed

@@
 
+\subsection{Using \CYS in Batch Mode}
+
+Although \CYS is primarily designed as an interactive system,
+it can also be used to process data in batch mode.
+For example, consider a Curry program, say \code{myprocessor}, that
+reads argument strings from the command line and processes them.
+Suppose the entry point is a function called \code{just_doit}
+that takes no arguments. Such a processor can be invoked from
+the shell as follows:
+\begin{curry}
+> kics2 :set args string1 string2 :load myprocessor.curry :eval just_doit :quit
+\end{curry}
+The \ccode{:quit} directive in necessary to avoid \CYS going
+into interactive mode after the
+excution of the expression being evaluated.
+The actual run-time arguments (\code{string1}, \code{string2})
+are defined by setting the option \code{args} (see above).
+
+Here is an example to use \CYS in this way:
+\begin{curry}
+> kics2 :set args Hello World :add System :eval "getArgs >>= putStrLn . unwords" :quit
+Hello World
+>
+\end{curry}
+
 \subsection{Command Line Editing}
@@
 the use of this mode.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manual"
+%%% End:

### 2014-10-16 docs/src/usage.tex
:set path now stores absolute path names
Bug in ":add" command fixed, path handling improved

@@
 
-\item[\fbox{\code{:edit $file$}}]\pindex{:edit}
-Load file $file$ into a text editor which is defined
+\item[\fbox{\code{:edit $m$}}]\pindex{:edit}
+Load the source text of module $m$ (which must be accessible
+via the current load path if no path specification is given)
+into a text editor which is defined
 as in the command \ccode{:edit}.
@@
 Show the source text of module $m$ which must be accessible
-via the current load path.
+via the current load path if no path specification is given.
 
@@
 inside this invocation of \CYS.
+
+The path is a list of directories separated by \ccode{:}.
+The prefix \ccode{\char126} is replaced by the home directory as
+in the following example:
+\begin{curry}
+:set path aux:~/tests
+\end{curry}
+Relative directory names are replaced by absolute ones
+so that the path is independent of later changes of the
+current working directory.
 

### 2014-11-11 runtime/ID.hs
Refactorings in Runtime

@@ -84,9 +84,9 @@ makeStrictCList :: Constraint -> [Constraint]
 makeStrictCList (_ :=: (LazyBind cs)) = concatMap makeStrictCList cs
 makeStrictCList binding@(_ :=: _)     = [binding]
 makeStrictCList u@(Unsolvable _)      = [u]
-makeStrictCList (ConstraintChoice cd i csl csr) 
+makeStrictCList (ConstraintChoice cd i csl csr)
   = [ConstraintChoice cd i (concatMap makeStrictCList csl)(concatMap makeStrictCList csr)]
-makeStrictCList (ConstraintChoices cd i css) 
+makeStrictCList (ConstraintChoices cd i css)
   = [ConstraintChoices cd i (map (concatMap makeStrictCList) css)]
 -- ---------------------------------------------------------------------------
 -- Decision
@@ -152,7 +152,7 @@ data ID
 instance Show ID where
   show (ChoiceID          i) = "?" ++ showUnique i
   show (FreeID          _ i) = "_x" ++ show i
-  show (NarrowedID      _ i) = "Narrowed" ++ show i
+  show (NarrowedID      _ i) = "Narrowed_" ++ show i
 
 -- |Retrieve the 'IDSupply' from an 'ID'
 supply :: ID -> IDSupply
@@ -195,7 +195,6 @@ getUnique (NarrowedID      _ s) = unique s
 isNarrowed :: ID -> Bool
 isNarrowed (NarrowedID _ _) = True
 isNarrowed _                = False
-
 
 -- ---------------------------------------------------------------------------
 -- Tracing
@@ -245,30 +244,20 @@ lookupID i = snd `liftM` lookupDecisionID i
 
 -- |Lookup the 'Decision' and the 'ID' an 'ID' ultimately is bound to
 lookupDecisionID :: Store m => ID -> m (Decision, ID)
-lookupDecisionID i = getDecisionRaw (getUnique i) >>= unchain
+lookupDecisionID i = getDecisionRaw (getUnique i) >>= follow
   where
-    -- TODO: reactivate shortening of chains as soon as we know how
-    --       to do this correct and efficient
-    -- For BindTo, we shorten chains of multiple BindTos by directly binding
-    -- to the last ID in the chain.
-    unchain (BindTo j) = do
-      retVal@(c, _lastId) <- lookupDecisionID j
+    -- follow BindTo
+    follow (BindTo j) = do
[…]

### 2014-11-11 runtime/Search.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE FlexibleInstances, CPP#-}
+{-# LANGUAGE CPP#-}
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 module Search where
 
@@ -191,7 +191,7 @@ showChoiceTree n goal = showsTree n [] "" (try goal) []
   --            of the respective level (for drawing aesthetical corners)
   --   * @k@    is the key for the decision (L, R, constructor index)
   showsTree d l k ndVal
-    | d <= 0    = indent l k . showChar '_' . nl
+    | d <= 0    = indent l k . showChar '\x2026' . nl
     | otherwise = indent l k . case ndVal of
       Val v           -> showString "Val " . shows v . nl
       Fail _ _        -> showChar '!' . nl
@@ -649,18 +649,22 @@ encapsulatedSearch x cd store = searchMSearch cd $ ((\y _ _ -> y) $!! x) cd stor
 -- Generic search using MonadPlus instances for the result
 -- ---------------------------------------------------------------------------
 
-type DecisionMap = Map.Map Integer Decision
+newtype DecisionMap = DecisionMap { decisionMap :: Map.Map Integer Decision }
+emptyDecisionMap = DecisionMap Map.empty
+
+onDecisionMap f (DecisionMap m) = DecisionMap (f m)
 
 instance Monad m => Store (StateT DecisionMap m) where
   getDecisionRaw u        = gets
                           $ Map.findWithDefault defaultDecision (mkInteger u)
+                          . decisionMap
   setDecisionRaw u c
-    | isDefaultDecision c = modify $ Map.delete (mkInteger u)
-    | otherwise           = modify $ Map.insert (mkInteger u) c
-  unsetDecisionRaw u      = modify $ Map.delete (mkInteger u)
+    | isDefaultDecision c = unsetDecisionRaw u
+    | otherwise           = modify $ onDecisionMap $ Map.insert (mkInteger u) c
+  unsetDecisionRaw u      = modify $ onDecisionMap $ Map.delete (mkInteger u)
 
 searchMSearch :: (MonadSearch m, NormalForm a) => Cover -> a -> m a
-searchMSearch cd x = evalStateT (searchMSearch' cd return x) (Map.empty :: DecisionMap)
+searchMSearch cd x = evalStateT (searchMSearch' cd return x) emptyDecisionMap
 
[…]

### 2014-11-11 runtime/Solver.hs

@@ -1,7 +1,7 @@
 -- ---------------------------------------------------------------------------
 -- Constraint Solving
 -- ---------------------------------------------------------------------------
-module Solver (Solution, solve,mkSolution) where
+module Solver (Solution, solve) where
 
 import Types
 
@@ -57,7 +57,7 @@ solveOne cd (i :=: cc) e = lookupDecision i >>= follow cc
   -- 1st param: the (new) Choice which should be stored for i
   -- 2nd param: the (old) Choice for i in the store
   follow (LazyBind  _) NoDecision    = mkDecision i cc e
-  follow (LazyBind cs1)(LazyBind cs2)= mkDecision i NoDecision 
+  follow (LazyBind cs1)(LazyBind cs2)= mkDecision i NoDecision
                                      $ guardCons cd (StructConstr (concatMap makeStrictCList (cs1++cs2))) e
   follow _             (LazyBind cs) = mkDecision i cc
                                      $ guardCons cd (StructConstr cs) e

### 2014-11-11 runtime/idsupplyinteger/IDSupply.hs

@@ -7,10 +7,10 @@ module IDSupply
   , getDecisionRaw, setDecisionRaw, unsetDecisionRaw
   ) where
 
-import Control.Monad (liftM)
-import Data.IORef (IORef, newIORef, readIORef, modifyIORef)
-import qualified Data.Map as Map (Map, empty, delete, findWithDefault, insert)
-import System.IO.Unsafe (unsafePerformIO)
+import           Control.Monad    (liftM)
+import           Data.IORef       (IORef, newIORef, readIORef, modifyIORef)
+import qualified Data.Map as Map  (Map, empty, delete, findWithDefault, insert)
+import           System.IO.Unsafe (unsafePerformIO)
 
 -- SOURCE pragma to allow mutually recursive dependency
 import {-# SOURCE #-} ID (Decision, defaultDecision, isDefaultDecision)

### 2014-11-11 runtime/Search.hs
Added debugging output for top-level search

@@ -22,6 +22,20 @@ type DetExpr    a =             Cover -> ConstStore -> a
 type NonDetExpr a = IDSupply -> Cover -> ConstStore -> a
 type Strategy   a = NonDetExpr a -> IO (IOList a)
 
+
+debugSearch :: NormalForm a => (a -> IO ()) -> NonDetExpr a -> IO ()
+debugSearch _ goal = do
+  nf <- getNormalForm goal
+--   putStrLn $ header "Normalform"
+--   print nf
+--   putStrLn $ header "Search Tree"
+--   putStrLn $ showChoiceTree' nf
+  let res = evalStateT (searchMSearch' initCover return nf) emptyDecisionMap
+--   putStrLn $ header "Results"
+  mapM_ print (dfsSearch res)
+
+  where header s = s ++ '\n' : replicate (length s) '='
+
 -- ---------------------------------------------------------------------------
 -- Search combinators for top-level search in the IO monad
 -- ---------------------------------------------------------------------------
@@ -650,6 +664,8 @@ encapsulatedSearch x cd store = searchMSearch cd $ ((\y _ _ -> y) $!! x) cd stor
 -- ---------------------------------------------------------------------------
 
 newtype DecisionMap = DecisionMap { decisionMap :: Map.Map Integer Decision }
+  deriving Show
+
 emptyDecisionMap = DecisionMap Map.empty
 
 onDecisionMap f (DecisionMap m) = DecisionMap (f m)

### 2014-11-11 runtime/Search.hs
Improved search debugging

@@ -1,4 +1,4 @@
-{-# LANGUAGE CPP#-}
+{-# LANGUAGE CPP #-}
 {-# OPTIONS_GHC -fno-warn-orphans #-}
 module Search where
 
@@ -26,15 +26,20 @@ type Strategy   a = NonDetExpr a -> IO (IOList a)
 debugSearch :: NormalForm a => (a -> IO ()) -> NonDetExpr a -> IO ()
 debugSearch _ goal = do
   nf <- getNormalForm goal
---   putStrLn $ header "Normalform"
---   print nf
---   putStrLn $ header "Search Tree"
---   putStrLn $ showChoiceTree' nf
-  let res = evalStateT (searchMSearch' initCover return nf) emptyDecisionMap
---   putStrLn $ header "Results"
-  mapM_ print (dfsSearch res)
+  putStrLn $ header "Normalform"
+  print nf
+  putStrLn $ header "Search Tree"
+  putStrLn $ showChoiceTree' nf
+  let res = runStateT (searchMSearch' initCover return nf) emptyDecisionMap
+  putStrLn $ header "Results"
+  mapM_ showRes (dfsSearch res)
 
-  where header s = s ++ '\n' : replicate (length s) '='
+  where
+    header s = s ++ '\n' : replicate (length s) '='
+    showRes (r, dm) = do
+      putStr "Result   : " >> print r
+      putStr "#Bindings: " >> print (Map.size (decisionMap dm))
+      putStr "Bindings : " >> print (decisionMap dm)
 
 -- ---------------------------------------------------------------------------
 -- Search combinators for top-level search in the IO monad
@@ -205,7 +210,7 @@ showChoiceTree n goal = showsTree n [] "" (try goal) []
   --            of the respective level (for drawing aesthetical corners)
   --   * @k@    is the key for the decision (L, R, constructor index)
   showsTree d l k ndVal
-    | d <= 0    = indent l k . showChar '\x2026' . nl
+    | d <= 0    = indent l k . showChar elli . nl
     | otherwise = indent l k . case ndVal of
       Val v           -> showString "Val " . shows v . nl
       Fail _ _        -> showChar '!' . nl
@@ -227,10 +232,11 @@ showChoiceTree n goal = showsTree n [] "" (try goal) []
   showKey "" = id
   showKey k  = showString k . showString ": "
 
-  vbar = '\x2502'     -- vertical bar
-  hbar = '\x2500'     -- horizontal bar
[…]

### 2014-11-11 runtime/Types.hs
Made Generable a superclass of NormalForm

@@ -143,7 +143,7 @@ lazyBindOrNarrow _  _ _ j _ = internalError $ "Types.lazyBindOrNarrow: " ++ show
 --
 -- The NF/GNF computation is combined with a continuation to be applied to
 -- the NF/GNF.
-class (NonDet a, Show a) => NormalForm a where
+class (NonDet a, Generable a, Show a) => NormalForm a where
   -- |Apply a continuation to the normal form
   ($!!) :: NonDet b => (a -> Cover -> ConstStore -> b) -> a -> Cover -> ConstStore -> b
   -- |Apply a continuation to the ground normal form
@@ -156,24 +156,24 @@ class (NonDet a, Show a) => NormalForm a where
 
 -- |Auxiliary function to apply the continuation to the normal forms of the
 -- two alternatives of a binary choice.
-nfChoice :: (NormalForm a, NonDet b) 
+nfChoice :: (NormalForm a, NonDet b)
          => (a -> Cover -> ConstStore -> b) -> Cover -> ID -> a -> a -> Cover -> ConstStore -> b
 nfChoice cont d i x1 x2 cd cs = case i of
   ChoiceID      _ -> choiceCons d i ((cont $!! x1) cd cs) ((cont $!! x2) cd cs)
-  _               -> internalError "Basics.nfChoice: no ChoiceID" 
+  _               -> internalError "Basics.nfChoice: no ChoiceID"
 
 -- |Auxiliary function to apply the continuation to the normal forms of the
 -- n alternatives of a n-ary choice.
-nfChoices :: (NormalForm a, NonDet b) 
+nfChoices :: (NormalForm a, NonDet b)
           => (a -> Cover -> ConstStore -> b) -> Cover -> ID -> [a] -> Cover -> ConstStore -> b
 nfChoices cont d i xs cd cs = case i of
    ChoiceID _     -> internalError "Basics.nfChoices: ChoiceID"
    FreeID _ _     -> cont (choicesCons d i xs) cd cs
    NarrowedID _ _ -> choicesCons d i (map (\x -> (cont $!! x) cd cs) xs)
-   
+
 -- |Auxiliary function to apply the continuation to the ground normal forms of
 -- the two alternatives of a binary choice.
-gnfChoice :: (NormalForm a, NonDet b) 
+gnfChoice :: (NormalForm a, NonDet b)
[…]

### 2014-11-11 runtime/ID.hs
Swapping of choice structure for unified free variables

@@ -11,7 +11,7 @@ module ID
   , Decision (..), defaultDecision, isDefaultDecision
     -- * IDs
   , ID (..), leftID, rightID, narrowID, getKey, mkInteger, isNarrowed
-  , IDSupply, initSupply, leftSupply, rightSupply, thisID, freeID
+  , IDSupply, initSupply, leftSupply, rightSupply, thisID, freeID, supply
     -- * Decision management
   , traceLookup, traceDecision
   , lookupDecision, lookupID, lookupDecisionID, setDecision, setUnsetDecision

### 2014-11-11 runtime/Search.hs

@@ -285,10 +285,11 @@ printValsDFSMatch backTrack cont goal = do
       where decide bt c a = setDecision i c >> printValsDFSMatch bt cont a
     follow c           = internalError $ "Search.prChoice: " ++ show c
 
-  prFree _ i xs   = lookupDecisionID i >>= follow
+  prFree cd i xs   = lookupDecisionID i >>= follow
     where
     follow (LazyBind cs, _) = processLB backTrack cs i xs
-    follow (ChooseN c _, _) = printValsDFSMatch backTrack cont (xs !! c)
+    follow (ChooseN c _, j) = printValsDFSMatch backTrack cont (ys !! c)
+      where Free _ _ ys = try $ generate (supply j) cd
     follow (NoDecision , j) = cont $ choicesCons initCover j xs
     follow c                = internalError $ "Search.prFree: " ++ show c
 
@@ -344,10 +345,11 @@ printValsDFSTry backTrack cont goal = do
          where decide bt c a = setDecision i c >> printValsDFSTry bt cont a
        follow c           = error $ "Search.prChoice: " ++ show c
 
-    Free _ i xs   -> lookupDecisionID i >>= follow
+    Free cd i xs   -> lookupDecisionID i >>= follow
       where
       follow (LazyBind cs, _) = processLB backTrack cs i xs
-      follow (ChooseN c _, _) = printValsDFSTry backTrack cont (xs !! c)
+      follow (ChooseN c _, j) = printValsDFSTry backTrack cont (ys !! c)
+        where Free _ _ ys = try $ generate (supply j) cd
       follow (NoDecision , j) = cont $ choicesCons initCover j xs
       follow c                = error $ "Search.prFree: " ++ show c
 
@@ -437,7 +439,8 @@ searchDFS act goal = do
     dfsFree cd i xs = lookupDecisionID i >>= follow
       where
       follow (LazyBind cs, _) = processLB i cs xs
-      follow (ChooseN c _, _) = dfs cont (xs !! c)
+      follow (ChooseN c _, j) = dfs cont (ys !! c)
+        where Free _ _ ys = try $ generate (supply j) cd
       follow (NoDecision , j) = cont $ choicesCons cd j xs
       follow c                = internalError $ "Search.dfsFree: Bad choice " ++ show c
 
@@ -508,6 +511,14 @@ searchBFS act goal = do
[…]

### 2014-11-14 benchmarks/benchmark_papers/kics2/bench_kics2.tex
Comments updated
Benchmarks expanded with dist infos
Benchmarks extended with section on constraint store usage

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% This is a paper to benchmark various versions of the KiCS2 system.
+% This is a paper to benchmark various versions of the KiCS2 system
+% and compare its efficiency with other Curry systems (PAKCS, MCC).
+%
+% This can be executed and formatted by the command
+% > ../execpaper -f bench_kics2.tex
+%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{curryprog}
+
+The benchmarks are executed with various Curry systems.
+The following table summarizes some information about the
+actual back ends of these systems.
+
+\begin{center}
+\runcurry{distInfosAsTable (PAKCS:kics2Systems++[KiCS2Local []])}
+\end{center}
 
@@
 
+\subsection{Binding Benchmarks for KiCS2}
+
+The following benchmarks demonstrate the relevance of the
+constraint store of KiCS2. They compare PAKCS with a local installation
+of KiCS2 where KiCS2 runs either in standard mode and
+in a mode where the constraint store is disabled by the option
+\begin{quote}
+\texttt{:set ghc -DDISABLE\_CS -fforce-recomp}
+\end{quote}
+
+\begin{center}
+\runcurry{mainBench [PAKCS, KiCS2Local [], KiCS2Local [WithoutCS]] bindingBench}
+\end{center}
+
 \end{document}

### 2014-11-19 docs/src/usage.tex
Added explanation of (+/-)(trace/profile) to manual, changelog extended

@@
 
+\item[\fbox{\code{+/-trace}}]\pindex{trace}
+Turn on/off the trace mode. If the trace mode is on,
+it is possible to trace the sources of failing computations.
+
+\item[\fbox{\code{+/-profile}}]\pindex{profile}
+Turn on/off the profile mode. If the profile mode is on,
+expressions as well as programs are compiled with GHC's profiling
+capabilities enabled. For expressions, evaluation will automatically
+generate a file \code{Main.prof} containing the profiling information
+of the evaluation.
+For compiled programs, the profiling has to be manually activated
+using runtime options when executed:
+\begin{curry}
+kics2 :set +profile :load MyProgram.curry :save :quit
+./MyProgram +RTS -p -RTS [additional arguments]
+\end{curry}
+
 \item[\fbox{\code{+/-ghci}}]\pindex{ghci}

### 2014-12-04 docs/src/overview.tex
Added documentation for configurable profiling option

@@
 \end{description}
+
+Furthermore, \CYS can be installed with experimental support for profiling
+of executables. To use profiling, two requirements have to be met:
+\begin{itemize}
+\item The libraries that are shipped with the GHC that is used by \CYS
+have to be installed with profiling enabled. This is the default for the system
+libraries contained in the GHC release, but may not be the case for additional
+libraries.
+\item The \code{Makefile} of \CYS contains an option \code{PROFILING}
+which has to be set to \code{yes} to enable profiling support. You may
+either change the Makefile to
+\begin{lstlisting}
+PROFILING = yes
+\end{lstlisting}
+or specify this setting while starting the installation process using
+\begin{lstlisting}
+make <optional target> PROFILING=yes
+\end{lstlisting}
+
 In the following, \cyshome denotes the installation directory

### 2014-12-23 docs/src/auxiliary_files.tex
Manual extended w.r.t. hierarchical modules

@@
 local declarations.
-If $dir$ is the directory where the Curry program is stored,
+
+If the Curry module $M$ is stored in the directory $dir$,
 the corresponding FlatCurry program is stored in the directory
 \ccode{$dir$/.curry}.
+This is also the case for hierarchical module names:
+if the module $D1.D2.M$ is stored in the directory $dir$
+(i.e., the module is actually stored in $dir/D1/D2/M.curry$),
+then the corresponding FlatCurry program is stored in
+\ccode{$dir$/.curry/D1/D2/M.fcy}.
 
@@
 
-\item[\code{Curry_prog.nda}:] This file contains some information about
-the determinism behavior of operations that is used by the
-\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
-for more details about the use of this information).
-If $dir$ is the directory where the Curry program is stored,
-the corresponding Haskell program is stored in the directory
-\ccode{$dir$/.curry/.kics2}.
-
-\item[\code{Curry_prog.info}:] This file contains some
-information about the top-level functions of module \code{prog}
-that are used by the interactive environment,
-like determinism behavior or IO status.
-This file is stored in the same directory as \code{Curry_prog.nda}.
-
 \item[\code{Curry_prog.hs}:] This file contains a Haskell program
@@
 \CYS compiler.
-This file is stored in the same directory as \code{Curry_prog.nda}.
+
+If the Curry module $M$ is stored in the directory $dir$,
+the corresponding Haskell program is stored in the directory
+\ccode{$dir$/.curry/kics2}.
+This is also the case for hierarchical module names:
+if the module $D1.D2.M$ is stored in the directory $dir$
+(i.e., the module is actually stored in $dir/D1/D2/M.curry$),
+then the corresponding Haskell program is stored in
+\ccode{$dir$/.curry/kics2/D1/D2/Curry_prog.hs}.
+
+\item[\code{Curry_prog.hi}:] This file contains the interface
+of the Haskell program \code{Curry_prog.hs} when the latter program is
+compiled in order to execute it.
+This file is stored in the same directory as \code{Curry_prog.hs}.
 
@@
 
-\item[\code{Curry_prog.hi}:] This file contains the interface
-of the Haskell program \code{Curry_prog.hs} when the latter program is
-compiled in order to execute it.
+\item[\code{Curry_prog.nda}:] This file contains some information about
+the determinism behavior of operations that is used by the
+\CYS compiler (see \cite{BrasselHanusPeemoellerReck11}
+for more details about the use of this information).
+This file is stored in the same directory as \code{Curry_prog.hs}.
+
+\item[\code{Curry_prog.info}:] This file contains some
+information about the top-level functions of module \code{prog}
+that are used by the interactive environment,
+like determinism behavior or IO status.
 This file is stored in the same directory as \code{Curry_prog.hs}.
@@
 \end{description}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "manual"
+%%% End:

### 2014-12-23 docs/src/overview.tex
Added missing \end{itemize} to documentation
Manual extended w.r.t. hierarchical modules

@@
 \end{description}
-
-Furthermore, \CYS can be installed with experimental support for profiling
+%
+Furthermore, \CYS can be installed with experimental support
+for profiling\index{profiling}
 of executables. To use profiling, two requirements have to be met:
@@
 \end{lstlisting}
+\end{itemize}
 
@@
 
-The current implementation of \CYS supports only flat module names,
-i.e., the notation \code{Dir.Mod.f} is not supported.\index{modules}
-In order to allow the structuring of modules in different directories,
 \CYS searches for imported modules in various directories.
@@
 \end{enumerate}
+The same strategy also applies to modules with a hierarchical
+module name with the only difference that the hierarchy prefix
+of a module name corresponds to a directory prefix of the module.
+For instance, if the main module is stored in directory \code{MAINDIR}
+and imports the module \code{Test.Func}, then the module stored in
+\code{MAINDIR/Test/Func.curry} is imported (without setting
+any additional import path) according to the
+module search strategy described above.
+
 Note that the standard prelude (\code{\cyshome/lib/Prelude.curry})
@@
 \code{Prelude}.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "manual"
+%%% End:

### 2015-03-04 docs/src/extensions.tex
Updated KiCS2 manual regarding the new record syntax

@@
 
-\subsection {Records}
-\label{records}
-
-A record is a data structure for bundling several data of various types.
-It consists of typed data fields where each field is associated with
-a unique label. These labels can be used to construct, select or update
-fields in a record.
-
-Unlike labeled data fields in Haskell, records are
-not syntactic sugar but a real extension of the
-language\footnote{The current version allows to transform records
-  into abstract data types. Future extensions may not have
-  this facility.}.
-The basic concept is described in \cite{Leijen05} but the current
-version does not yet provide all features mentioned there.
-The restrictions are explained in Section~\ref{sec-restrinrecs}.
-
-\subsubsection{Record Type Declaration}
-\label{sec-recordtypedecl}
-
-It is necessary to declare a record type before a record
-can be constructed or used. The declaration has the following form:
-\begin{curry}
-type $R$ $\alpha_1$ $\ldots$ $\alpha_n$ = { $l_1$ :: $\tau_1$,$\ldots$, $l_m$ :: $\tau_m$ }
-\end{curry}
-It introduces a new $n$-ary record type $R$ which represents a
-record consisting of $m$ fields. Each field has a unique label $l_i$
-representing a value of the type $\tau_i$. Labels
-are identifiers which refer to the corresponding
-fields. The following examples define some record types:
-\begin{curry}
-type Person = {name :: String, age :: Int}
-type Address = {person :: Person, street :: String, city :: String}
-type Branch a b = {left :: a, right :: b}
-\end{curry}
-It is possible to summarize different labels which have the same
-type. For instance, the record \code{Address} can also be declared as follows:
-\begin{curry}
-type Address = {person :: Person, street,city :: String}
-\end{curry}
-The fields can occur in an arbitrary order. The example above
-can also be written as
-\begin{curry}
-type Address = {street,city :: String, person :: Person}
-\end{curry}
-The record type can be used in every type expression to represent
-the corresponding record, e.g.
-\begin{curry}
-data BiTree = Node (Branch BiTree BiTree) | Leaf Int
-\end{curry}
-\begin{curry}
-getName :: Person -> String
-getName $\ldots$
-\end{curry}
-
-Labels can only be used in the context of
-records. They do not share the name space with
-functions/constructors/variables or type constructors/type variables.
-For instance it is possible to use
-the same identifier for a label and a function at the same time. Label
-identifiers cannot be shadowed by other identifiers.
-
-Like in type synonym declarations, recursive or mutually
-dependent record declarations are not allowed. Records can only
-be declared at the top level. Further restrictions are described in
-section \ref{sec-restrinrecs}.
-
-\subsubsection{Record Construction}
-\label{sec-recordconstr}
-
-The record construction generates a record with initial values for
-each data field. It has the following form:
-\begin{curry}
-{ $l_1$ := $v_1$,$\ldots$, $l_m$ := $v_m$ }
-\end{curry}
-It generates a record where each label $l_i$ refers to the
-value $v_i$. The type of the record results from the record type
-declaration where the labels $l_i$ are defined.
-A mix of labels from different
-record types is not allowed. All labels must be specified with
-exactly one assignment. Examples for record constructions are
-\begin{curry}
-{name := "Johnson", age := 30}     -- generates a record of type 'Person'
-{left := True, right := 20}        -- generates a record of type 'Branch'
-\end{curry}
-Assignments to labels can occur in an arbitrary order. For instance a
-record of type \code{Person} can also be generated as follows:
-\begin{curry}
-{age := 30, name := "Johnson"}     -- generates a record of type 'Person'
-\end{curry}
-Unlike labeled fields in record type declarations,
-record constructions can be used in expressions without any restrictions
-(as well as all kinds of record expressions). For instance the following
-expression is valid:
-\begin{curry}
-{person := {name := "Smith", age := 20},   -- generates a record of
- street := "Main Street",                  -- type 'Address'
- city   := "Springfield"}
-\end{curry}
-
-\subsubsection{Field Selection}
-\label{sec-fieldsel}
-
-The field selection is used to extract data from records.
-It has the following form:
-\begin{curry}
-$r$ :> $l$
-\end{curry}
-It returns the value to which the label $l$ refers to from the
-record expression $r$. The label must occur in the declaration of
-the record type of $r$.
-An example for a field selection is:
-\begin{curry}
-pers :> name
-\end{curry}
-This returns the value of the label \code{name} from the record \code{pers}
-(which has the type \code{Person}).
-Sequential application of field selections are also possible:
-\begin{curry}
-addr :> person :> age
-\end{curry}
-The value of the label \code{age} is extracted from a record which itself
-is the value of the label \code{person} in the record \code{addr}
-(which has the type \code{Address}).
-
-\subsubsection{Field Update}
-\label{sec-fieldupd}
-
-Records can be updated by reassigning a new value to a label:
-\begin{curry}
-{$l_1$ := $v_1$,$\ldots$, $l_k$ := $v_k$ | $r$}
-\end{curry}
-The label $l_i$ is associated with the new value $v_i$ which
-replaces the current value in the record $r$.
-The labels must occur in the declaration
-of the record type of $r$. In contrast to record constructions,
-it is not necessary to specify all labels of a record.
-Assignments can occur in an arbitrary order. It is not allowed to
-specify more than one assignment for a label in a record update.
-Examples for record updates are:
-\begin{curry}
-{name := "Scott", age := 25 | pers}
-{person := {name := "Scott", age := 25 | pers} | addr}
-\end{curry}
-In these examples \code{pers} is a record of type \code{Person} and \code{addr}
-is a record of type \code{Address}.
-
-\subsubsection{Records in Pattern Matching}
-\label{sec-recsinpm}
-
-It is possible to apply pattern matching to records (e.g., in functions,
-let expressions or case branches). Two kinds of record patterns
-are available:
-\begin{curry}
-{$l_1$ = $p_1$,$\ldots$, $l_n$ = $p_n$}
-{$l_1$ = $p_1$,$\ldots$, $l_k$ = $p_k$ | _}
-\end{curry}
-In both cases each label $l_i$ is specified with a pattern $p_i$.
-All labels must occur only once in the record pattern.
-The first case is used to match the whole record. Thus, all labels
-of the record must occur in the pattern.
-The second case is used to match only a part of
-the record. Here it is not necessary to specify all labels.
-This case is represented by a vertical bar followed by the underscore
-(anonymous variable). It is
-not allowed to use a pattern term instead of the underscore.
-
-When trying to match a record against a record pattern, the
-patterns of the specified labels are matched against
-the corresponding values in the record expression. On success, all pattern
-variables occurring in the patterns are replaced by their actual expression.
-If none of the patterns matches, the computation fails.
-
-Here are some examples of pattern matching with records:
-\begin{curry}
-isSmith30 :: Person -> Bool
-isSmith30 {name = "Smith", age = 30} = True
-\end{curry}
-\begin{curry}
-startsWith :: Char -> Person -> Bool
-startsWith c {name = (d:_) | _} = c == d
-\end{curry}
-\begin{curry}
-getPerson :: Address -> Person
-getPerson {person = p | _} = p
-\end{curry}
-As shown in the last example, a field selection can also be obtained
-by pattern matching.
-
-\subsubsection{Export of Records}
-\label{sec-exprecs}
-
-Exporting record types and labels is very similar to exporting
-data types and constructors. There are three ways
-to specify an export:
-\begin{itemize}
-\item \code{module $M$ (\ldots, $R$, \ldots) where} \\
-  exports the record $R$ without any of its labels.
-\item \code{module $M$ (\ldots, $R$(..), \ldots) where} \\
-  exports the record $R$ together with all its labels.
-\item \code{module $M$ (\ldots, $R$($l_1$,\ldots,$l_k$), \ldots) where} \\
-  exports the record $R$ together with the labels $l_1$, \ldots, $l_k$.
-\end{itemize}
-%
-Note that imported labels cannot be overwritten in record declarations
-of the importing module. It is also not possible to import equal labels
-from different modules.
-
-\subsubsection{Restrictions in the Usage of Records}
-\label{sec-restrinrecs}
-
-In contrast to the basic concept in \cite{Leijen05}, \CYS/Curry provides a
-simpler version of records. Some of the features described there are
-currently not supported or even restricted.
-
-\begin{itemize}
-\item Labels must be unique within the whole scope of the program.
-  In particular, it is not allowed to define the same label within
-  different records, not even when they are imported from other
-  modules. However, it is possible to use equal identifiers for other
-  entities without restrictions, since labels have an independent
-  name space.
-\item The record type representation with labeled fields can only be
-  used as the right-hand-side of a record type declaration. It is
-  not allowed to use it in any other type annotation.
-\item Records are not extensible or reducible. The structure of a
-  record is specified in its record declaration and cannot be
-  modified at the runtime of the program.
-\item Empty records are not allowed.
-\item It is not allowed  to use a pattern term
-  at the right side of the vertical bar in a record pattern
-  except for the underscore (anonymous pattern variable).
-\item Labels cannot be sequentially associated with multiple values
-  (record fields do not behave like stacks).
-\end{itemize}
+\input{fieldlabels}
 

### 2015-04-02 docs/src/usage.tex
Added documentation for configurable profiling option
Manual extended w.r.t. hierarchical modules
Property definitions of the form "-Dprop=val" allowed as arguments

@@
 for systems implemented in Curry.
+
+There are also some specific options that can be used
+when invoking \CYS:
+\begin{description}
+\item[\code{--noreadline}:]
+Do not use input line editing (see Section~\ref{sec-readline}).
+\item[\code{-Dname=val}:]
+Overwrite values defined in the configuration file \ccode{.kics2rc}
+(see Section~\ref{sec-customization}),
+where \code{name} is a property defined in the configuration file
+and \code{val} its new value.
+\end{description}
 
@@
 
-\item[\fbox{\code{+/-profile}}]\pindex{profile}
+\item[\fbox{\code{+/-profile}}]\pindex{profile}\index{profiling}
+(only available when configured during installation, see Section~\ref{sec-install})
 Turn on/off the profile mode. If the profile mode is on,
@@
 \subsection{Command Line Editing}
+\label{sec-readline}
 

### 2015-09-17 docs/src/libraries.tex
Some libraries added to manual
Libraries updated
Library Prolog added to documentation
libs and tools updated
libs updated
libs and tools updated, src uses new AbstractCurry libs

@@
 \code{FlatCurry} (Section~\ref{Library:FlatCurry})
-and \code{AbstractCurry} (Section~\ref{Library:AbstractCurry}),
-stored in the directory \ccode{\cyshome/lib/meta},
+and \code{AbstractCurry.Types} (Section~\ref{Library:AbstractCurry.Types})
+%stored in the directory \ccode{\cyshome/lib/meta},
 which define datatypes for the representation
 of Curry programs.
-\code{AbstractCurry} is a more direct representation of a Curry program,
+\code{AbstractCurry.Types} is a more direct representation of a Curry program,
 whereas \code{FlatCurry} is a simplified representation
@@
 program manipulations (or, for writing new back ends for Curry),
-whereas \code{AbstractCurry} is intended for manipulations of
+whereas \code{AbstractCurry.Types} is intended for manipulations of
 programs that are more oriented towards the source program.
 
-Both modules contain predefined I/O actions to read programs
-in the \code{AbstractCurry} (\code{readCurry}\pindex{readCurry})
-or \code{FlatCurry}
-(\code{readFlatCurry}\pindex{readFlatCurry}) format.
+There are predefined I/O actions to read AbstractCurry and
+FlatCurry programs: \code{AbstractCurry.Files.readCurry}\pindex{readCurry})
+and \code{FlatCurry.readFlatCurry}\pindex{readFlatCurry}).
 These actions parse the corresponding source program and return
 a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry} and \code{FlatCurry}).
+in the modules \code{AbstractCurry.Types} and \code{FlatCurry}).
 
@@
 \url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.html}}
+\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.Types.html}}
 of these modules.
@@
 \input{lib/Constraint}
+\input{lib/CPNS}
 \input{lib/CSV}
+\input{lib/Debug}
 \input{lib/Directory}
+\input{lib/Distribution}
+\input{lib/Either}
 \input{lib/FileGoodies}
+\input{lib/FilePath}
+\input{lib/Findall}
 \input{lib/Float}
+\input{lib/Function}
+\input{lib/GetOpt}
 \input{lib/Global}
@@
 \input{lib/Profile}
+\input{lib/Prolog}
 \input{lib/PropertyFile}
@@
 \input{lib/ReadShowTerm}
+\input{lib/SearchTree}
 \input{lib/SetFunctions}
-\input{lib/SearchTree}
 \input{lib/Socket}
@@
 \input{lib/Unsafe}
+\input{lib/UnsafeSearchTree}
+\input{lib/ValueSequence}
 
@@
 \input{lib/HTML}
+\input{lib/HtmlCgi}
 \input{lib/HtmlParser}
@@
 \input{lib/Markdown}
+\input{lib/URL}
 \input{lib/WUI}
-\input{lib/URL}
+\input{lib/WUIjs}
 \input{lib/XML}
@@
 
-\input{lib/AbstractCurry}
-\input{lib/AbstractCurryPrinter}
+\input{lib/AbstractCurry.Types}
+\input{lib/AbstractCurry.Files}
+\input{lib/AbstractCurry.Select}
+\input{lib/AbstractCurry.Build}
+\input{lib/AbstractCurry.Pretty}
+\input{lib/AnnotatedFlatCurry}
+\input{lib/AnnotatedFlatCurryGoodies}
 \input{lib/CompactFlatCurry}
@@
 \input{lib/FlexRigid}
-\input{lib/PrettyAbstract}
 

### 2015-10-29 docs/src/libraries.tex
library docs updated

@@
 in Curry, there are system modules
-\code{FlatCurry} (Section~\ref{Library:FlatCurry})
+\code{FlatCurry.Types} (Section~\ref{Library:FlatCurry.Types})
 and \code{AbstractCurry.Types} (Section~\ref{Library:AbstractCurry.Types})
-%stored in the directory \ccode{\cyshome/lib/meta},
 which define datatypes for the representation
@@
 \code{AbstractCurry.Types} is a more direct representation of a Curry program,
-whereas \code{FlatCurry} is a simplified representation
+whereas \code{FlatCurry.Types} is a simplified representation
 where local function definitions are replaced by global definitions
@@
 is translated into explicit case/or expressions.
-Thus, \code{FlatCurry} can be used for more back-end oriented
+Thus, \code{FlatCurry.Types} can be used for more back-end oriented
 program manipulations (or, for writing new back ends for Curry),
@@
 FlatCurry programs: \code{AbstractCurry.Files.readCurry}\pindex{readCurry})
-and \code{FlatCurry.readFlatCurry}\pindex{readFlatCurry}).
+and \code{FlatCurry.Files.readFlatCurry}\pindex{readFlatCurry}).
 These actions parse the corresponding source program and return
 a data term representing this program (according to the definitions
-in the modules \code{AbstractCurry.Types} and \code{FlatCurry}).
+in the modules \code{AbstractCurry.Types} and \code{FlatCurry.Types}).
 
@@
 we refer to the online documentation\footnote{%
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/FlatCurry.html} and
-\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/CDOC/AbstractCurry.Types.html}}
+\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/FlatCurry.Types.html} and
+\url{http://www-ps.informatik.uni-kiel.de/kics2/lib/AbstractCurry.Types.html}}
 of these modules.
@@
 \end{curry}
-Then the I/O action \code{(FlatCurry.readFlatCurry "test")} returns the
+Then the I/O action \code{(FlatCurry.Files.readFlatCurry "test")} returns the
 following term:
@@
 \input{lib/AbstractCurry.Pretty}
-\input{lib/AnnotatedFlatCurry}
-\input{lib/AnnotatedFlatCurryGoodies}
-\input{lib/CompactFlatCurry}
+\input{lib/FlatCurry.Types}
+\input{lib/FlatCurry.Files}
+\input{lib/FlatCurry.Goodies}
+\input{lib/FlatCurry.Pretty}
+\input{lib/FlatCurry.Read}
+\input{lib/FlatCurry.Show}
+\input{lib/FlatCurry.XML}
+\input{lib/FlatCurry.FlexRigid}
+\input{lib/FlatCurry.Compact}
+\input{lib/FlatCurry.Annotated.Types}
+\input{lib/FlatCurry.Annotated.Pretty}
+\input{lib/FlatCurry.Annotated.Goodies}
 \input{lib/CurryStringClassifier}
-\input{lib/FlatCurry}
-\input{lib/FlatCurryGoodies}
-\input{lib/FlatCurryRead}
-\input{lib/FlatCurryShow}
-\input{lib/FlatCurryXML}
-\input{lib/FlexRigid}
 
 } % end setlength parindent
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "manual"
+%%% End:

### 2015-11-16 benchmarks/benchmark_papers/kics2/bench_kics2.tex
Benchmark implementation updated
Parallel benchmarks modified

@@
 
-\documentclass{article}
+\documentclass{scrartcl}
 
@@
 
-These benchmarks are executed with the option ``\texttt{:set +parallel}''
-and different run-time settings to influence the number
+These benchmarks are executed with different search strategy options
+to evaluate the parallel strategy w.r.t.\ the number
 of threads used for the execution.

### 2015-11-17 docs/src/auxiliary_files.tex
Manual updated, documentation of tools moved into currytools repo

@@
 and stored in different files which are shortly explained in this section.
-In general, it is not necessary to know about
+If you use \CYS, it is not necessary to know about
 these auxiliary files because they are automatically generated
@@
 generated when a program is compiled with \CYS.
+It can be also explicitly generated by the Curry front end\pindex{cymake}
+\begin{curry}
+cymake --flat -i$\cyshome$/lib prog
+\end{curry}
 The FlatCurry representation of a Curry program is usually

### 2015-11-17 docs/src/manual.tex
Added first version of syntax documentation
Updated KiCS2 manual regarding the new record syntax
Manual updated, documentation of tools moved into currytools repo

@@
 \input{version}
+
+% define home dir of the Curry system in order to locate tool manuals:
+\newcommand{\curryhome}{../..}
 
@@
 \newcommand{\cb}{CurryBrowser\xspace}
+\newcommand{\curryrc}{.kics2rc} % name of the system's rc file
+
+\usepackage{syntax}
 
@@
 \catcode`\_=12
+
+\newcommand{\translation}[1]{\bigskip\fbox{\begin{minipage}{15cm}\textbf{Translation: }{#1}\end{minipage}}\par\bigskip}
 
@@
 Bj\"orn Peem\"oller$^3$\\
-Fabian Reck$^4$ \\[4ex]
+Fabian Reck$^4$ \\
+Jan Rasmus Tikovsky$^5$ \\[4ex]
 \normalsize
@@
 (3) University of Kiel, Germany, {\tt bjp@informatik.uni-kiel.de} \\
-(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\[5ex]~
+(4) University of Kiel, Germany, {\tt fre@informatik.uni-kiel.de} \\
+(4) University of Kiel, Germany, {\tt jrt@informatik.uni-kiel.de} \\[5ex]~
 \end{center}
@@
 \include{extensions}
-\include{currydoc}
-\include{currybrowser}
-\include{currytest}
-\include{erd2curry}
+\include{syntax}
+\clearpage
+\input{\curryhome/currytools/currydoc/Docs/manual}
+\clearpage
+\input{\curryhome/currytools/browser/Docs/manual}
+\clearpage
+\input{\curryhome/currytools/currytest/Docs/manual}
+\clearpage
+\input{\curryhome/currytools/erd2curry/Docs/manual}
+\clearpage
+\input{\curryhome/currytools/spicey/Docs/manual}
 \include{problems}

### 2015-11-17 docs/src/overview.tex
Manual updated, documentation of tools moved into currytools repo

@@
 in the directory where you have stored your Curry programs.
-Moreover, the current implementation also recompiles
-system libraries according to the setting of some options.
-Therefore, the \CYS system should be locally installed
-in your user account.
+%Moreover, the current implementation also recompiles
+%system libraries according to the setting of some options.
+%Therefore, the \CYS system should be locally installed
+%in your user account.
 The auxiliary files for all Curry programs in the current
@@
 \begin{curry}
-{x = -1, y = -1} Success
-{x = 1, y = 1} Success
+{x = -1, y = -1} True
+{x = 1, y = 1} True
 \end{curry}
@@
 By default, imported modules are searched in the directory
-of the main program and the system module directories
-\ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
+of the main program and the system module directory
+\ccode{\cyshome/lib}.
 This search path can be extended
@@
       \ccode{libraries}.
-\item The directories \ccode{\cyshome/lib} and \ccode{\cyshome/lib/meta}.
+\item The directory \ccode{\cyshome/lib}.
 \end{enumerate}

### 2015-11-17 docs/src/preface.tex
Manual updated, documentation of tools moved into currytools repo

@@
 an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus06Curry} that is
+\cite{Hanus97POPL,Hanus15Curry} that is
 based on compiling Curry programs into Haskell programs.
@@
 We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus12Curry}.
+of Curry as described in the Curry language definition \cite{Hanus15Curry}.
 Therefore, this document only explains the use of the different
@@
 (see Section~\ref{sec-restrictions})
-compared with the language Curry (Version 0.8.3).
+compared with the language Curry (Version 0.9.0).
+%
 The basic ideas of the implementation of \CYS

### 2015-11-17 docs/src/usage.tex
Bug fix in reading of RC arguments

@@
 \begin{description}
-\item[\code{--noreadline}:]
+\item[\fbox{\code{--noreadline}}] (if used, this must be the first option):
 Do not use input line editing (see Section~\ref{sec-readline}).
-\item[\code{-Dname=val}:]
+\item[\fbox{\code{-Dname=val}}]
+(these options must come before any \CYS command):
 Overwrite values defined in the configuration file \ccode{.kics2rc}

### 2015-11-22 scripts/runcurry.sh
tools updated

@@ -1,71 +1,0 @@
-#!/bin/sh
-# run a Curry program without explicitly invoking the interactive environment
-
-# Define the main directory where KiCS2 is installed:
-KICS2HOME=`echo KICS2HOME must be defined here!`
-export KICS2HOME
-
-# usage function:
-show_usage () {
-    echo "Usage:"
-    echo
-    echo "As a shell command:"
-    echo "> runcurry [Curry system options] <Curry program name> <run-time arguments>"
-    echo
-    echo "As a shell script: start script with"
-    echo "#!/usr/bin/env runcurry"
-    echo "...your Curry program defining operation 'main'..."
-    echo
-    echo "In interactive mode:"
-    echo "> runcurry"
-    echo "...type your Curry program until end-of-file..."
-}
-
-# define a new Curry file name:
-NEWMOD=RUNCURRY$$
-while [ -f $NEWMOD.curry ] ; do
-    NEWMOD=$NEWMOD_0
-done
-NEWPROG=$NEWMOD.curry
-
-CURRYARGS=
-PROG=
-DELPROG=no
-while [ $# -gt 0 -a -z "$PROG" ] ; do
-    if [ "$1" = "-h" -o "$1" = "--help" -o "$1" = "-?" ] ; then
-	show_usage ; exit
-    fi
-    ARG=$1
-    shift
-    ARGwolcurry=`expr $ARG : '\(.*\)\.lcurry'`
-    ARGwocurry=`expr $ARG : '\(.*\)\.curry'`
-    # check whether runcurry is called in script mode, i.e., the argument
-    # is not a Curry program but an existing file:
-    if [ "$ARGwolcurry" != "" -o "$ARGwocurry" != "" ] ; then
-	# argument is a Curry program:
-	PROG=$ARG
-    elif [ -x $ARG ] ; then
-	# argument is not a Curry file but is executable, hence a script:
-	# store ARG in a Curry program, but remove lines starting with '#':
-	sed 's|^#.*$||' < $ARG > $NEWPROG
-	DELPROG=yes
-	PROG=$NEWPROG
-    else
-	# argument is a Curry system argument:
-	CURRYSYSARGS="$CURRYSYSARGS $ARG"
-    fi
-done
-
-if [ -z "$PROG" ] ; then
-    # no program argument provided, use remaining input as program:
-    PROG=$NEWPROG
-    cat > $PROG
[…]

### 2016-02-10 docs/src/preface.tex
tools and libs updated

@@
 an implementation of the multi-paradigm language Curry
-\cite{Hanus97POPL,Hanus15Curry} that is
+\cite{Hanus97POPL,Hanus16Curry} that is
 based on compiling Curry programs into Haskell programs.
@@
 We assume familiarity with the ideas and features
-of Curry as described in the Curry language definition \cite{Hanus15Curry}.
+of Curry as described in the Curry language definition \cite{Hanus16Curry}.
 Therefore, this document only explains the use of the different

### 2016-04-05 docs/src/usage.tex
REPL command ":compile" added

@@
 Recompile all currently loaded modules.
+
+\item[\fbox{\code{:compile $prog$}}]\pindex{:compile}
+Compile and load the program stored in \code{$prog$.curry},
+as with the command \ccode{:load $prog$},
+but compile also the generated Haskell modules.
+Usually, this is automatically done when an expression is evaluated.
+Hence, it is not necessary to use this command.
 

### 2016-06-06 docs/src/extensions.tex
Manual updated, documentation of tools moved into currytools repo
Updated description of functional patterns w.r.t. syntactical restrictions

@@
 Functional patterns \cite{AntoyHanus05LOPSTR} are a useful extension
-to code operations in a more readable way. Furthermore,
+to implement operations in a more readable way. Furthermore,
 defining operations with functional patterns avoids problems
@@
 
-In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} ($n>0$)
+In general, a pattern of the form \code{($f$ $t_1$\ldots$t_n$)} for $n>0$
+(or of the qualified form \code{($M.f$ $t_1$\ldots$t_n$)} for $n \geq 0$)
 is interpreted as a functional pattern if $f$ is not a visible constructor
 but a defined function that is visible in the scope of the pattern.
+Furthermore, for a functional pattern to be well-defined,
+there are two additional requirements to be satisfied:
+
+\begin{enumerate}
+
+\item
+If a function $f$ is defined by means of a functional pattern $fp$,
+then the evaluation of $fp$ must not depend on $f$, i.e.,
+the semantics of a function defined using functional patterns
+must not (transitively) depend on its own definition.
+This excludes definitions such as
+\begin{curry}
+(xs ++ ys) ++ zs = xs ++ (ys ++ zs)
+\end{curry}
+and is necessary to assign a semantics to funtions employing functional patterns
+(see \cite{AntoyHanus05LOPSTR} for more details).
+
+\item
+Only functions that are globally defined may occur inside a functional pattern.
+This restriction ensures that no local variable
+might occur in the value of a functional pattern,
+which might lead to a wrong semantics.
+Consider, for instance, the following (complicated) equality operation
+\begin{curry}
+eq :: a -> a -> Bool
+eq x y = h y
+  where
+  g True  = x
+  h (g a) = a
+\end{curry}
+where the locally defined function \code{g} occurs in the functional
+pattern \code{g a} of \code{h}.
+Since \code{g a} evaluates to the value of \code{x} whereas \code{a}
+is narrowed to \code{True}, the call \code{h y} now evaluates to \code{True}
+if the value of \code{y} equals the value of \code{x}.
+This requires a strict unification between \code{x} and \code{y},
+so that a equivalent definition without functional patterns would be:
+\begin{curry}
+eq :: a -> a -> Bool
+eq x y = h y
+  where
+  h x1 | x =:= x1 = True
+\end{curry}
+However, this implies that variables occuring in the value
+of a functional pattern imply a strict unification
+if they are defined in an outer scope,
+whereas variables defined \emph{inside} a functional pattern
+behave like pattern variables.
+In consequence, the occurrence of variables from an outer scope
+inside a functional pattern might lead to an unintuitive behavior,
+which is why this is excluded by a restriction to globally defined functions.
+Note that this does not exclude a functional pattern inside a local function,
+which is still perfectly reasonable.
+\end{enumerate}
 
@@
 
-\input{fieldlabels}
-
 %%% Local Variables:

### 2016-06-06 docs/src/extensions.tex
libs, tools, manual updated

@@
 but a defined function that is visible in the scope of the pattern.
-Furthermore, for a functional pattern to be well-defined,
+Furthermore, for a functional pattern to be well defined,
 there are two additional requirements to be satisfied:
@@
 might occur in the value of a functional pattern,
-which might lead to a wrong semantics.
+which might lead to an non-intuitive semantics.
 Consider, for instance, the following (complicated) equality operation
@@
 where the locally defined function \code{g} occurs in the functional
-pattern \code{g a} of \code{h}.
-Since \code{g a} evaluates to the value of \code{x} whereas \code{a}
-is narrowed to \code{True}, the call \code{h y} now evaluates to \code{True}
+pattern \code{(g a)} of \code{h}.
+Since \code{(g a)} evaluates to the value of \code{x} whereas \code{a}
+is instantiated to \code{True}, the call \code{h y} now evaluates to \code{True}
 if the value of \code{y} equals the value of \code{x}.
-This requires a strict unification between \code{x} and \code{y},
-so that a equivalent definition without functional patterns would be:
+In order to check this equality condition,
+a strict unification between \code{x} and \code{y} is required
+so that an equivalent definition without functional patterns would be:
 \begin{curry}
@@
 In consequence, the occurrence of variables from an outer scope
-inside a functional pattern might lead to an unintuitive behavior,
-which is why this is excluded by a restriction to globally defined functions.
+inside a functional pattern might lead to an non-intuitive behavior.
+To avoid such problems, locally defined functions are excluded
+as functional patterns.
 Note that this does not exclude a functional pattern inside a local function,
@@
 \end{enumerate}
-
+%
 It is also possible to combine functional patterns with

### 2016-08-25 docs/src/external_ops.tex
CASS manual added

@@
 non-deterministic or constrained values
-(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck11WLP})
+(see \cite{BrasselHanusPeemoellerReck11,BrasselHanusPeemoellerReck13PADL})
 that are meaningless in Haskell.
